### 2023-08-29 manuscript.Rmd
made rmarkdown doc, and many other things

---
title: "It's easier to move sideways"
author:
  - Mykle L. Hoban:
      institute:
        - himb
      equal_contributor: yes
      correspondence: yes
      email: "mhoban@hawaii.edu"
  - Kaleonani K.C. Hurley:
      institute: [himb, nerr]
      equal_contributor: yes
  - Molly A. Timmers:
      institute: [himb, natgeo]
  - Kerry Reardon:
      institute: noaa
  - Robert J. Toonen:
      institute: himb
institute:
  - himb: Hawai‘i Institute of Marine Biology
  - nerr: He‘eia National Estuarine Research Reserve
  - natgeo: National Geographic Pristine Seas
  - noaa: National Oceanic and Atmospheric Administration
abstract:
  Shallow coral reefs are extensively studied, and although scleractinian corals have been recorded to 165 m, little is known about other mesophotic coral reef ecosystem (MCE) inhabitants. Brachyuran crabs fill many ecological and trophic niches on reefs, making them ideal candidates for evaluating species composition among depths. Here we ask if MCEs host the same communities as the shallower reefs spread across the entire 2400 km Hawaiian Archipelago. We deployed Autonomous Reef Monitoring Structures (ARMS) for two years among shallow sites (12 m) across the latitudinal gradient of the Hawaiian Archipelago to compare directly with a depth gradient (12, 30, 60, and 90 m) south of Oʻahu island. A total of 2284 individual crabs representing 90 morphospecies (19 families) were found. Assemblage composition was not statistically significantly different among shallow sites spread across the archipelago but was significantly different and highly stratified across the depth gradient.  We show that 90 m of depth is a more potent determinant of brachyuran community composition than the latitudinal, geographic (high vs low islands), or anthropogenic impact (human population size) gradient across the archipelago. Deeper reefs host significantly different brachyuran communities than shallow ones in Hawai’i, and only 3 of 90 morphospecies (~3%) occurring across the entire depth range (mesophotic to shallow) were sampled. Further research on the cryptic reef-dwelling fauna is needed for identifying areas of unique biodiversity, as well as boundaries between depth-stratified reef ecosystems.
bibliography: resources/citations.bib
csl: resources/vancouver.csl
output:
  word_document:
    pandoc_args:
      - "--filter=resources/pandoc-crossref"
      - "--lua-filter=resources/scholarly-metadata.lua"
      - "--lua-filter=resources/author-info-blocks.lua"
      - "--reference-doc=resources/ref.docx"
    keep_md: yes
tblPrefix: Table
figPrefix:
  - "Fig."
  - "Figs."
---

\newpage

```{r setup, include=FALSE}
library(fs)
library(here)

knitr::opts_chunk$set(echo = FALSE)
source("crabs.R")
if (!exists("crabs_done")) {
  crabs_done <<- FALSE
}
if (!crabs_done) {
  setup_crabs()
  crabs_done <<- TRUE
}

format_range <- function(vals) {
  f <- range(vals)
  str_glue("{f[1]}--{f[2]}")
}

format_species <- function(spp) {
  if (spp$found) {
    str_glue("*{spp$species}* {spp$authority}")
  } else {
    s <- str_split(spp$species," ")[[1]]
    str_glue("*{s[1]}* {s[2]}")
  }
}

num <- function(x,accuracy=0.01) {
  scales::number(x,big.mark = ",",decimal.mark = ".",accuracy=accuracy)
}

dir_create(here("cache"))
if (file_exists(here("cache","all_summary.rds"))) {
  all_summary <<- read_rds(here("cache","all_summary.rds"))
} else {
  all_summary <<- sample_summary(crabs_untransformed)
  write_rds(all_summary,here("cache","all_summary.rds"))
}
if (file_exists(here("cache","deep_summary.rds"))) {
  deep_summary <<- read_rds(here("cache","deep_summary.rds"))
} else {
  deep_summary <<- sample_summary(crabs_untransformed %>% subset_samples(shallow_deep == "deep"))
  write_rds(deep_summary,here("cache","deep_summary.rds"))
}
```

# Introduction

## Sample Collection

Brachyuran crab assemblages were sampled from benthic coral reef habitat in the Hawaiian Archipelago using Autonomous Reef Monitoring Structures (ARMS). ARMS are standardized collection devices designed to mimic coral reef structural complexity to attract settlement of reef cryptofauna [@brainardAutonomousReefMonitoring2009; @zimmermanArtificialReefMatrix2004]. They are constructed from 10 gray type 1 PVC plates (23 cm x 23 cm) stacked in an alternating series of open and semi-enclosed layers (figure xxx, ARMS unit). The stack is fastened to a larger base plate (35 cm x 45 cm) which can be attached to the benthos upon deployment [@brainardAutonomousReefMonitoring2009; @hurleyAssessmentShallowMesophotic2016; @knowltonCoralReefBiodiversity2010; @zimmermanArtificialReefMatrix2004].

A total of 89 ARMS units were deployed over the course of six years (2010–2016; see Table xxx  for deployment and recovery schedules). Units were deployed by NOAA scuba divers during Pacific Reef Assessment and Monitoring Program (RAMP) research cruises in shallow coral reef habitat (12–16 m depth) in both the Main (MHI) and Northwest Hawaiian Islands (NWHI) (Table 1). The sites in the MHI were at Hawaiʻi Island, Maui, Oʻahu, and Kauaʻi. The sites in the NWHI were at Lalo (French Frigate Shoals), Kapou (Lisianski), Manawai (Pearl & Hermes), and Hōlanikū (Kure Atoll) (Figure xxx, site map). To explore the effects of depth on crab community structure, we also present data from a previous study by Hurley et al. [-@hurleyAssessmentShallowMesophotic2016], which included a total of 27 units deployed off O‘ahu from 2009–2012 in 12–90 m depth; nine units were deployed at 12 m, and six each at 30 m, 60 m, and 90 m. Shallow units were deployed according to standard NOAA methodology, whereas units ≥ 30 m were deployed and recovered using technical mixed-gas closed-circuit rebreather diving (Table 1).

ARMS units were retrieved by encapsulating them in-situ in a mesh-lined crate (100 μm pore size) to prevent escape of motile organisms. Once returned to ship or shore, they were placed in containers of aerated, filtered seawater, crates were removed, and units were systematically disassembled plate-by-plate. Seawater surrounding disassembled units was sieved through 2 mm, 500 μm, and 100 μm mesh-size geologic sieves [see @lerayDNABarcodingMetabarcoding2015 for processing details), and retained macroorganisms were preserved in 95% ethanol. All adult brachyuran crabs (≥ 5 mm) were photographed, identified to morphospecies, and preserved in 95% ethanol (Fig. 2d). Individual preserved specimens were accessioned into the invertebrate zoology collections at the Florida Museum of Natural History (catalog numbers xxx–xxx ).

## Data analysis

Unless otherwise specified, all analyses were performed using R version 4.1.2 [@rcoreteamLanguageEnvironmentStatistical2021]. Community data were loaded and encapsulated into data objects using the R package phyloseq [@mcmurdiePhyloseqPackageReproducible2013] and multivariate community analyses were performed using the package vegan [@oksanenVeganCommunityEcology2022] on Bray-Curtis dissimilarities of Hellinger-transformed species abundance matrices [@legendreEcologicallyMeaningfulTransformations2001; @raoReviewCanonicalCoordinates1995].

To examine patterns of crab diversity across the islands, we calculated alpha diversity statistics for ARMS units across the archipelago and examined how those values varied between the MHI and NWHI as well as the effects of environmental factors. We calculated species richness and Simpson diversity using the ‘estimate_richness’ function from phyloseq (Simpson = $1-\sum_{i=1}^{R}p_i^2$, where R is the total number of species, and pi is the proportional abundance of each species). Alpha diversity was compared between MHI and NWHI using two-sample t-tests, both with and without the inclusion of mesophotic ARMS units. To explore the effects of environmental factors on alpha diversity across the islands, we used a linear mixed-effect model, with Simpson diversity or richness as response variables, the environmental variables listed below as fixed effects, and island nested within region (MHI vs NWHI) as random effects. Models were generated using the ‘lmer’ function from the lme4 package [@batesFittingLinearMixedEffects2015].

We used the distance-based redundancy analysis method (dbRDA; a form of multivariate multiple regression) implemented in the ‘dbrda’ function from vegan to examine the relative importance of environmental variables on crab community variation across the Hawaiian Islands. We first assessed the importance of each environmental variable individually using marginal tests. Then, a stepwise forward selection procedure (using the function ‘ordiR2step’) was used to find the suite of environmental variables that best explained the dissimilarity among sites with regard to crab community composition. Colinearity between traits was assessed by calculating variance inflation factors (VIF) using the function ‘vif.cca’ from complete and reduced models and discarding variables where VIF > 10. Significance levels were assessed with the function ‘anova’ using permutation tests with 9,999 permutations. All continuous environmental variables used for dbRDA were scaled to unit variance. We used the permutational analysis of variance (PERMANOVA) method in the ‘adonis2’ function from vegan to assess variation among sites and/or depths across the islands, with 9,999 permutations to assess significance in both term-based and marginal tests.

## Environmental variables

Potential environmental factors affecting the community structure of brachyuran crabs on shallow reefs were assembled from published literature [@selkoeDNACoralReef2016] and the Pacific Islands Ocean Observing System (PacIOOS) (Table 2). Environmental factors were assembled only for shallow sites sampled during RAMP cruises.

# Results

## Diversity summary

### All ARMS

The `r nsamples(crabs_untransformed)` ARMS units deployed across the Hawaiian Islands (in `r format_range(crab_data$depth)` m depth) yielded a total of `r num(sum(sample_sums(crabs_untransformed)))` adult crabs. Sampling units contained `r num(all_summary$abundance_mean,0.1)` ± `r num(all_summary$abundance_sd,0.1)` (mean ± SD) individuals per unit, and abundance ranged from `r format_range(all_summary$abundance_range)` individuals. Crabs were identified to `r sum(unlist(all_summary$types_breakdown))` distinct types (including `r all_summary$types_breakdown$species` species and `r all_summary$types_breakdown$morphospecies` morphospecies) in `r all_summary$taxonomy_breakdown$genera` genera and `r all_summary$taxonomy_breakdown$families` families, which accounted for ~36% of the known crab diversity reported from Hawaiian waters [@castroCatalogAnomuranBrachyuran2011]. The four most common species were: Chlorodiella laevissima (Dana, 1852) (with 610 individuals on 59 units), Perinia tumida Dana, 1851 (268 individuals on 67 units), Percnon abbreviatum (Dana, 1851) (with 121 individuals on 47 units), and Dynomene hispida (Latreille in Milbert, 1812) (114 individuals on 49 units). Twenty-two species occurred in a single unit across all sites (singletons), accounting for ~21% of total recovered crab species.

### Mesophotic ARMS

The 27 units deployed at mesophotic depths ≥ 30 m on O‘ahu yielded both fewer species and fewer individuals compared to the complete dataset, with 271 crabs overall, a total of 39 distinct types (11 species, 28 morphospecies), 15.1 ± 6.2 individuals per unit, and abundance ranging from 7–31 crabs. The four most abundant species in the deep ARMS did not overlap with those from the complete dataset: Chlorodiella sp2 (46 individuals from 16 units), Epiactaea nodulosa (White, 1848) (40 individuals from 10 units), Catoptrus inaequalis (Rathbun, 1906) (17 individuals from 8 units), and Pilumnus sp2018 (12 individuals from 7 units). There were 11 singletons, which proportionally accounted for ~28% of species found on deep ARMS.

## Alpha diversity

### Summary

Overall species richness for the complete dataset was 7.8 ± 3.1 species per unit (Simpson diversity: 0.74 ± 0.14). Overall species richness for shallow ARMS < 30 m was 7.7 ± 3.2 species per unit (Simpson diversity: 0.73 ± 0.15). Overall species richness for mesophotic ARMS ≥ 30 m was 8.4 ± 2.5 species per unit (Simpson diversity: 0.82 ± 0.04).

### Comparisons

For the complete dataset, there was no significant difference in observed species richness between the MHI and NWHI (two-sample t-test: t = 1.06, DF = 106.43, p = 0.29), but Simpson diversity was significantly different, although mean values were similar with 0.78 ± 0.10 in the MHI and 0.70 ± 0.17 in the NWHI (two-sample t-test: t = 2.72, DF = 74.88, p = 0.008). When only shallow arms < 30 m were considered, neither comparison was significant (richness: t = 0.77, DF = 94.47, p = 0.44; Simpson: t  = 1.93, DF = 84.76, p = 0.06).

### Effects of environmental variables

# Discussion

# References

::: {#refs}
:::

### 2023-08-29 crabs.R

@@ -14,8 +14,16 @@ library(vegan)
 library(ggrepel)
 library(here)
 
+# to check model assumptions
+library(performance)
+
+# for linear mixed effects models
+library(lme4)
+
 # set the random seed
 set.seed(31337)
+
+run_things <- FALSE
 
 # functions ---------------------------------------------------------------
 
@@ -245,54 +253,215 @@ taxa_tibble <- function(ps,otu_col="otu") {
 
 
 # data prep ---------------------------------------------------------------
+
+setup_crabs <- function() {
+  # load our main phyloseq object and hellinger-transform it
+  crabs_untransformed <<- load_ps(
+    here("data","crabs.csv"),"unit",
+    here("data","taxonomy.csv"),"otu",
+    here("data","metadata.csv")
+  ) 
+    crabs <<- crabs_untransformed %>%
+    ps_standardize("hellinger")
+    # this is the shallow subset
+  crabs_shallow <<- crabs %>%
+    subset_samples(region != "mce") 
+    # we're gonna mess with the sample data a bit
+  # first, get rid of rows that have NAs
+  crab_data_shallow <<- crabs_shallow %>%
+    sample_tibble(sample_col = "unit") %>%
+    drop_na()
+    # make sure we drop samples that had NAs in their sample data
+  crabs_shallow <<- prune_samples(sample_names(crabs_shallow) %in% crab_data_shallow$unit,crabs_shallow)
+    # now we select the columns containing the environmental variables
+  # we care about and scale the numeric ones to unit variance
+  crab_data_shallow <<- crabs_shallow %>%
+    sample_tibble(sample_col = "unit") %>%
+    select(unit,region,island,depth,chl,sst,slope,coral_cover,closest_island,larval_connectivity,human_impact) %>%
+    mutate(across(where(is.numeric),~as.numeric(scale(.x)))) %>%
+    column_to_rownames("unit") 
+    # reassociate the new sample data
+  sample_data(crabs_shallow) <<- crab_data_shallow
+    # precalculate the bray-curtis distance for all crabs
+  crab_dist <<- distance(crabs,"bray")
+  # pull out the otu table
+  crab_otus <<- otu_table(crabs)
[…]

### 2023-08-29 map.R

@@ -1,12 +1,9 @@
 library(tidyverse)
 library(sf)
-library(mapdata)
 library(marmap)
-library(patchwork)
+library(fs)
 library(here)
 library(rnaturalearth)
-library(rnaturalearthhires)
-
 
 if (!exists("crab_data")) {
   crab_data <- read_csv(here("data","metadata.csv"),col_types = cols())
@@ -16,14 +13,9 @@ crab_sites <- crab_data %>%
   distinct(island,island_group,lat,lon,.keep_all = FALSE) %>%
   mutate( island = fct_reorder(island,-lat) ) %>%
   st_as_sf(coords=c("lon","lat"),crs=4326, remove=F) #%>%
-# st_shift_longitude()
 
 hawaii <- ne_states("united states of america",returnclass = "sf") %>%
   filter(name == "Hawaii")
-
-# hawaii <- rnaturalearthhires::states10 %>%
-#   st_as_sf() %>%
-#   filter(name == "Hawaii")
 
 # run this section --------------------------------------------------------
 trans_box <- function(b,crs) {
@@ -37,43 +29,76 @@ trans_box <- function(b,crs) {
     st_bbox()
 }
 
-
 box_before <- st_bbox(st_buffer(crab_sites,units::as_units(50,"km")))
-box_before <- st_bbox(crab_sites)
-# box_before[2] <- 18
-# box_before[4] <- 29
-bath <- getNOAA.bathy(lon1 = box_before[1], lon2 = box_before[3], lat1 = box_before[2], lat2 = box_before[4], resolution = 5) 
+bath <- getNOAA.bathy(lon1 = box_before[1], lon2 = box_before[3], lat1 = box_before[2], lat2 = box_before[4], resolution = 3) 
 transformer <- str_glue("+proj=leac +lat_1={box_before[2]} +lat_2={box_before[4]} +lon_0={mean(box_before[c(1,3)])}")
 
 box_after <- box_before %>%
   trans_box(transformer)
-mappr <- hawaii %>%
-  st_transform(transformer)
-trans <- crab_sites %>% 
-  st_transform(transformer)
-
-bath_latlon <- as.raster(bath) %>% 
-  as.bathy() %>% 
-  as.xyz() %>%
-  as_tibble() %>%
-  drop_na(V3)
 
[…]

### 2023-08-30 manuscript.Rmd
updates

@@
 author:
+  - Kaleonani K.C. Hurley:
+      institute: 
+        - himb
+        - nerr
+      equal_contributor: yes
+      correspondence: yes
+      email: "hurleyk@hawaii.edu"
   - Mykle L. Hoban:
@@
         - himb
-      equal_contributor: yes
-      correspondence: yes
-      email: "mhoban@hawaii.edu"
-  - Kaleonani K.C. Hurley:
-      institute: [himb, nerr]
       equal_contributor: yes
@@
 institute:
-  - himb: Hawai‘i Institute of Marine Biology
-  - nerr: He‘eia National Estuarine Research Reserve
+  - himb: Hawai&#x02BB;i Institute of Marine Biology
+  - nerr: He&#x02BB;eia National Estuarine Research Reserve
   - natgeo: National Geographic Pristine Seas
@@
 We show that 90 m of depth is a more potent determinant of brachyuran community composition than the latitudinal, geographic (high vs low islands), or anthropogenic impact (human population size) gradient across the archipelago.
-Deeper reefs host significantly different brachyuran communities than shallow ones in Hawai’i, and only 3 of 90 morphospecies (~3%) occurring across the entire depth range (mesophotic to shallow) were sampled.
+Deeper reefs host significantly different brachyuran communities than shallow ones in Hawai&#x02BB;i, and only 3 of 90 morphospecies (~3%) occurring across the entire depth range (mesophotic to shallow) were sampled.
 Further research on the cryptic reef-dwelling fauna is needed for identifying areas of unique biodiversity, as well as boundaries between depth-stratified reef ecosystems.
@@
       - "--reference-doc=resources/ref.docx"
-    keep_md: yes
+    keep_md: false
 tblPrefix: Table
@@
 
+### Keywords
+<div custom-style="mono">ARMs, mesophotic coral ecosystems, community structure, dispersal, Hawai&#x02BB;i, brachyuran crabs</div>
 \newpage
@@
 ```{r setup, include=FALSE}
+knitr::opts_chunk$set(
+  echo = FALSE,
+  fig.path="output/figures/fig-",
+  dev="svg",
+  dpi=300,
+  message=FALSE,
+  warning=FALSE
+)
+
 library(fs)
 library(here)
+library(english)
 
-knitr::opts_chunk$set(echo = FALSE)
 source("crabs.R")
@@
 if (!crabs_done) {
-  setup_crabs()
+  cc <<- setup_crabs()
   crabs_done <<- TRUE
@@
 
+t_test <- function(test,plain=FALSE) {
+  if (plain) {
+    str_glue("*t* = {num(test$statistic,0.01)}, DF = {num(test$parameter['df'],0.01)}, *p* = {num(test$p.value,0.01)}")
+  } else {
+    str_glue("(two-sample t-test: *t* = {num(test$statistic,0.01)}, DF = {num(test$parameter['df'],0.01)}, *p* = {num(test$p.value,0.01)})")
+  }
+}
+
+top_list <- function(species,n,units,len=4) {
+  s1 <- map_chr(species,format_species)
+  s2 <- map2_chr(n,units,~{
+    str_glue("(with {.x} individuals on {.y} units)")
+  })
+  ss <- str_c(s1," ",s2)
+  ss[1:(len-1)] %>%
+    str_c(collapse=", ") %>%
+    str_c(", and ",ss[len])
+}
+
+cc$crabs_deep <- cc$crabs_untransformed %>% subset_samples(shallow_deep == "deep")
+cc$crabs_deep <- prune_taxa(taxa_sums(cc$crabs_deep) > 0,cc$crabs_deep)
+cc$crabs_shallow <- cc$crabs_untransformed %>% subset_samples(shallow_deep == "shallow")
+cc$crabs_shallow <- prune_taxa(taxa_sums(cc$crabs_shallow) > 0,cc$crabs_shallow)
+
 dir_create(here("cache"))
@@
 } else {
-  all_summary <<- sample_summary(crabs_untransformed)
+  all_summary <<- sample_summary(cc$crabs_untransformed)
   write_rds(all_summary,here("cache","all_summary.rds"))
@@
 } else {
-  deep_summary <<- sample_summary(crabs_untransformed %>% subset_samples(shallow_deep == "deep"))
+  deep_summary <<- sample_summary(cc$crabs_deep)
   write_rds(deep_summary,here("cache","deep_summary.rds"))
 }
+
+all_alpha <- alpha_diversity(cc$crabs_untransformed)
+deep_alpha <- alpha_diversity(cc$crabs_deep)
+shallow_alpha <- alpha_diversity(cc$crabs_shallow)
 ```
@@
 # Introduction
+
+# Materials and Methods
 
@@
 ARMS are standardized collection devices designed to mimic coral reef structural complexity to attract settlement of reef cryptofauna [@brainardAutonomousReefMonitoring2009; @zimmermanArtificialReefMatrix2004].
-They are constructed from 10 gray type 1 PVC plates (23 cm x 23 cm) stacked in an alternating series of open and semi-enclosed layers (figure xxx, ARMS unit).
-The stack is fastened to a larger base plate (35 cm x 45 cm) which can be attached to the benthos upon deployment [@brainardAutonomousReefMonitoring2009; @hurleyAssessmentShallowMesophotic2016; @knowltonCoralReefBiodiversity2010; @zimmermanArtificialReefMatrix2004].
+They are constructed from 10 gray type 1 PVC plates (23 cm $\times$ 23 cm) stacked in an alternating series of open and semi-enclosed layers (figure xxx, ARMS unit).
+The stack is fastened to a larger base plate (35 cm $\times$ 45 cm) which can be attached to the benthos upon deployment [@brainardAutonomousReefMonitoring2009; @hurleyAssessmentShallowMesophotic2016; @knowltonCoralReefBiodiversity2010; @zimmermanArtificialReefMatrix2004].
 
@@
 Units were deployed by NOAA scuba divers during Pacific Reef Assessment and Monitoring Program (RAMP) research cruises in shallow coral reef habitat (12–16 m depth) in both the Main (MHI) and Northwest Hawaiian Islands (NWHI) (Table 1).
-The sites in the MHI were at Hawaiʻi Island, Maui, Oʻahu, and Kauaʻi.
-The sites in the NWHI were at Lalo (French Frigate Shoals), Kapou (Lisianski), Manawai (Pearl & Hermes), and Hōlanikū (Kure Atoll) (Figure xxx, site map).
+The sites in the MHI were at Hawai&#x02BB;i Island, Maui, O&#x02BB;ahu, and Kaua&#x02BB;i.
+The sites in the NWHI were at Lalo (French Frigate Shoals), Kapou (Lisianski), Manawai (Pearl & Hermes), and Hōlanikū (Kure Atoll) ([@fig:site_map]).
 To explore the effects of depth on crab community structure, we also present data from a previous study by Hurley et al.
-[-@hurleyAssessmentShallowMesophotic2016], which included a total of 27 units deployed off O‘ahu from 2009–2012 in 12–90 m depth; nine units were deployed at 12 m, and six each at 30 m, 60 m, and 90 m.
-Shallow units were deployed according to standard NOAA methodology, whereas units ≥ 30 m were deployed and recovered using technical mixed-gas closed-circuit rebreather diving (Table 1).
+[-@hurleyAssessmentShallowMesophotic2016], which included a total of 27 units deployed off O&#x02BB;ahu from 2009–2012 in 12–90 m depth; nine units were deployed at 12 m, and six each at 30 m, 60 m, and 90 m.
+Shallow units were deployed according to standard NOAA methodology, whereas units $\geq$ 30 m were deployed and recovered using technical mixed-gas closed-circuit rebreather diving (Table 1).
+
+```{r site-map, fig.cap='Sampling map (Papahānaumokuākea Marine National Monument shaded in light blue)', results="asis", dev="svg", fig.width=9, fig.height=6}
+library(sf)
+library(rnaturalearth)
+
+crab_sites <- cc$crab_data %>%
+  distinct(island,island_group,lat,lon,.keep_all = FALSE) %>%
+  mutate( island = fct_reorder(island,-lat) ) %>%
+  st_as_sf(coords=c("lon","lat"),crs=4326, remove=F) #%>%
+
+hawaii <- ne_states("united states of america",returnclass = "sf") %>%
+  filter(name == "Hawaii")
+
+trans_box <- function(b,crs) {
+  box_before <- st_sfc(
+    st_point(b[c(1,2)]),
+    st_point(b[c(3,4)]),
+    crs=4326
+  )
+  box_before %>%
+    st_transform(crs) %>%
+    st_bbox()
+}
+
+box_before <- st_bbox(st_buffer(crab_sites,units::as_units(50,"km")))
+
+transformer <- str_glue("+proj=leac +lat_1={box_before[2]} +lat_2={box_before[4]} +lon_0={mean(box_before[c(1,3)])}")
+
+box_after <- box_before %>%
+  trans_box(transformer)
+
+pmnm <- st_read(here("data","gis","pmnm.shp"),quiet = TRUE) %>%
+  st_transform(transformer)
+
+island_transform <- tribble(
+  ~island,~new_name,
+  "Kure Atoll","Hōlanikū",
+  "Pearl and Hermes Reef","Manawai",
+  "Lisianski","Kapou",
+  "French Frigate Shoals","Lalo",
+  "Kauai","Kaua‘i",
+  "Oahu","O‘ahu",
+  "Maui","Maui",
+  "Hawaii Island","Hawai‘i Island"
+)
+
+cs <- crab_sites %>%
+  left_join(island_transform,by="island") %>%
+  mutate(
+    new_name = fct_reorder(new_name,-lat)
+  )
+
+plotz <- ggplot() +
+  # geom_contour(data=bath_proj,mapping=aes(x=V1,y=V2,z = V3),  color="#cecece") +
+  # geom_raster(data=bpr,aes(x=x,y=y,fill=z)) +
+  # geom_tile(data=bath_proj,mapping=aes(x=V1,y=V2,fill = V3)) +
+  # scale_fill_gradient2(low="dodgerblue4", mid="gainsboro", high="darkgreen") +
+  geom_sf(data=hawaii %>% st_transform(transformer), color="black", fill="green", linewidth=0.4) +
+  scale_fill_brewer(palette="Accent",name="Island") +
+  geom_sf(data=pmnm,fill="#6ed8f6",alpha=0.1) +
+  geom_sf(data=cs %>% st_transform(transformer),aes(fill=new_name),size=3,shape=21,color="black") +
+  coord_sf(xlim=box_after[c(1,3)],ylim=box_after[c(2,4)]) +
+  scale_x_continuous(breaks = -seq(180, 155, -5)) +
+  theme_minimal() +
+  xlab("Longitude") +
+  ylab("Latitude") +
+  theme(legend.text = element_text(size=10),legend.title = element_text(size=12))
+# labela <- tribble(
+#   ~lon,~lat,~text,
+#   -170,25.4,"Paphānaumokuākea Marine\nNational Monument\n(NWHI)"
+# ) %>%
+#   st_as_sf(coords=c("lon","lat"),crs=4326, remove=F)  %>%
+#   st_transform(transformer)
+# 
+# plotz +
+#   geom_sf_text(aes(label=text),data=labela,size=3,angle=-17.0)
+plotz
+cat("{#fig:site_map}")
+```
 
@@
 Once returned to ship or shore, they were placed in containers of aerated, filtered seawater, crates were removed, and units were systematically disassembled plate-by-plate.
-Seawater surrounding disassembled units was sieved through 2 mm, 500 μm, and 100 μm mesh-size geologic sieves [see @lerayDNABarcodingMetabarcoding2015 for processing details), and retained macroorganisms were preserved in 95% ethanol.
-All adult brachyuran crabs (≥ 5 mm) were photographed, identified to morphospecies, and preserved in 95% ethanol (Fig.
+Seawater surrounding disassembled units was sieved through 2 mm, 500 μm, and 100 μm mesh-size geologic sieves [see @lerayDNABarcodingMetabarcoding2015 for processing details], and retained macroorganisms were preserved in 95% ethanol.
+All adult brachyuran crabs ($\geq$ 5 mm) were photographed, identified to morphospecies, and preserved in 95% ethanol (Fig.
 2d).
@@
 
-Unless otherwise specified, all analyses were performed using R version 4.1.2 [@rcoreteamLanguageEnvironmentStatistical2021].
-Community data were loaded and encapsulated into data objects using the R package phyloseq [@mcmurdiePhyloseqPackageReproducible2013] and multivariate community analyses were performed using the package vegan [@oksanenVeganCommunityEcology2022] on Bray-Curtis dissimilarities of Hellinger-transformed species abundance matrices [@legendreEcologicallyMeaningfulTransformations2001; @raoReviewCanonicalCoordinates1995].
+Unless otherwise specified, all analyses were performed using `R` version 4.1.2 [@rcoreteamLanguageEnvironmentStatistical2021].
+Community data were loaded and encapsulated into data objects using the R package `phyloseq` [@mcmurdiePhyloseqPackageReproducible2013] and multivariate community analyses were performed using the package `vegan` [@oksanenVeganCommunityEcology2022] on Bray-Curtis dissimilarities of Hellinger-transformed species abundance matrices [@legendreEcologicallyMeaningfulTransformations2001; @raoReviewCanonicalCoordinates1995].
 
 To examine patterns of crab diversity across the islands, we calculated alpha diversity statistics for ARMS units across the archipelago and examined how those values varied between the MHI and NWHI as well as the effects of environmental factors.
-We calculated species richness and Simpson diversity using the ‘estimate_richness’ function from phyloseq (Simpson = $1-\sum_{i=1}^{R}p_i^2$, where R is the total number of species, and pi is the proportional abundance of each species).
+We calculated species richness and Simpson diversity using the `estimate_richness` function from phyloseq (Simpson = $1-\sum_{i=1}^{R}p_i^2$, where R is the total number of species, and $\mathrm{p}_i$ is the proportional abundance of each species).
 Alpha diversity was compared between MHI and NWHI using two-sample t-tests, both with and without the inclusion of mesophotic ARMS units.
 To explore the effects of environmental factors on alpha diversity across the islands, we used a linear mixed-effect model, with Simpson diversity or richness as response variables, the environmental variables listed below as fixed effects, and island nested within region (MHI vs NWHI) as random effects.
-Models were generated using the ‘lmer’ function from the lme4 package [@batesFittingLinearMixedEffects2015].
+Models were generated using the `lmer` function from the `lme4` package [@batesFittingLinearMixedEffects2015].
 
-We used the distance-based redundancy analysis method (dbRDA; a form of multivariate multiple regression) implemented in the ‘dbrda’ function from vegan to examine the relative importance of environmental variables on crab community variation across the Hawaiian Islands.
+We used the distance-based redundancy analysis method (dbRDA; a form of multivariate multiple regression) implemented in the `dbrda` function from vegan to examine the relative importance of environmental variables on crab community variation across the Hawaiian Islands.
 We first assessed the importance of each environmental variable individually using marginal tests.
-Then, a stepwise forward selection procedure (using the function ‘ordiR2step’) was used to find the suite of environmental variables that best explained the dissimilarity among sites with regard to crab community composition.
-Colinearity between traits was assessed by calculating variance inflation factors (VIF) using the function ‘vif.cca’ from complete and reduced models and discarding variables where VIF > 10.
-Significance levels were assessed with the function ‘anova’ using permutation tests with 9,999 permutations.
+Then, a stepwise forward selection procedure (using the function `ordiR2step`) was used to find the suite of environmental variables that best explained the dissimilarity among sites with regard to crab community composition.
+Colinearity between traits was assessed by calculating variance inflation factors (VIF) using the function `vif.cca` from complete and reduced models and discarding variables where VIF > 10.
+Significance levels were assessed with the function `anova` using permutation tests with 9,999 permutations.
 All continuous environmental variables used for dbRDA were scaled to unit variance.
-We used the permutational analysis of variance (PERMANOVA) method in the ‘adonis2’ function from vegan to assess variation among sites and/or depths across the islands, with 9,999 permutations to assess significance in both term-based and marginal tests.
+We used the permutational analysis of variance (PERMANOVA) method in the `adonis2` function from vegan to assess variation among sites and/or depths across the islands, with 9,999 permutations to assess significance in both term-based and marginal tests.
 
@@
 
-The `r nsamples(crabs_untransformed)` ARMS units deployed across the Hawaiian Islands (in `r format_range(crab_data$depth)` m depth) yielded a total of `r num(sum(sample_sums(crabs_untransformed)))` adult crabs.
-Sampling units contained `r num(all_summary$abundance_mean,0.1)` ± `r num(all_summary$abundance_sd,0.1)` (mean ± SD) individuals per unit, and abundance ranged from `r format_range(all_summary$abundance_range)` individuals.
+The `r nsamples(cc$crabs_untransformed)` ARMS units deployed across the Hawaiian Islands (in `r format_range(cc$crab_data$depth)` m depth) yielded a total of `r num(sum(sample_sums(cc$crabs_untransformed)),1)` adult crabs.
+Sampling units contained `r num(all_summary$abundance_mean,0.1)` $\pm$ `r num(all_summary$abundance_sd,0.1)` (mean $\pm$ SD) individuals per unit, and abundance ranged from `r format_range(all_summary$abundance_range)` individuals.
 Crabs were identified to `r sum(unlist(all_summary$types_breakdown))` distinct types (including `r all_summary$types_breakdown$species` species and `r all_summary$types_breakdown$morphospecies` morphospecies) in `r all_summary$taxonomy_breakdown$genera` genera and `r all_summary$taxonomy_breakdown$families` families, which accounted for ~36% of the known crab diversity reported from Hawaiian waters [@castroCatalogAnomuranBrachyuran2011].
-The four most common species were: Chlorodiella laevissima (Dana, 1852) (with 610 individuals on 59 units), Perinia tumida Dana, 1851 (268 individuals on 67 units), Percnon abbreviatum (Dana, 1851) (with 121 individuals on 47 units), and Dynomene hispida (Latreille in Milbert, 1812) (114 individuals on 49 units).
-Twenty-two species occurred in a single unit across all sites (singletons), accounting for ~21% of total recovered crab species.
+The four most common species were: `r with(all_summary,top_list(top5,top5_n,top5_units))`.
+`r str_to_sentence(as.english(all_summary$singleton_count))` species occurred in a single unit across all sites (singletons), accounting for ~`r num(all_summary$singleton_count/ntaxa(cc$crabs_untransformed)*100,1)`% of total recovered crab species.
 
@@
 
-The 27 units deployed at mesophotic depths ≥ 30 m on O‘ahu yielded both fewer species and fewer individuals compared to the complete dataset, with 271 crabs overall, a total of 39 distinct types (11 species, 28 morphospecies), 15.1 ± 6.2 individuals per unit, and abundance ranging from 7–31 crabs.
-The four most abundant species in the deep ARMS did not overlap with those from the complete dataset: Chlorodiella sp2 (46 individuals from 16 units), Epiactaea nodulosa (White, 1848) (40 individuals from 10 units), Catoptrus inaequalis (Rathbun, 1906) (17 individuals from 8 units), and Pilumnus sp2018 (12 individuals from 7 units).
-There were 11 singletons, which proportionally accounted for ~28% of species found on deep ARMS.
+The `r nsamples(cc$crabs_deep)` units deployed at mesophotic depths $\geq$ 30 m on O&#x02BB;ahu yielded both fewer species and fewer individuals compared to the complete dataset, with `r num(sum(sample_sums(cc$crabs_deep)),1)` crabs overall, a total of `r sum(unlist(deep_summary$types_breakdown))` distinct types (`r deep_summary$types_breakdown$species` species and `r deep_summary$types_breakdown$morphospecies` morphospecies), `r num(deep_summary$abundance_mean,0.1)` $\pm$ `r num(deep_summary$abundance_sd,0.1)` individuals per unit, and abundance ranging from `r format_range(deep_summary$abundance_range)` crabs.
+The four most abundant species in the deep ARMS did not overlap with those from the complete dataset: `r with(deep_summary,top_list(top5,top5_n,top5_units))`.
+There were `r deep_summary$singleton_count` singletons, which proportionally accounted for ~`r num(deep_summary$singleton_count/ntaxa(cc$crabs_deep)*100,1)`% of species found on deep ARMS.
 
@@
 
-Overall species richness for the complete dataset was 7.8 ± 3.1 species per unit (Simpson diversity: 0.74 ± 0.14).
-Overall species richness for shallow ARMS < 30 m was 7.7 ± 3.2 species per unit (Simpson diversity: 0.73 ± 0.15).
-Overall species richness for mesophotic ARMS ≥ 30 m was 8.4 ± 2.5 species per unit (Simpson diversity: 0.82 ± 0.04).
+Overall mean species richness for the complete dataset was `r num(all_alpha$richness,0.1)` $\pm$ `r num(all_alpha$richness_sd,0.1)` species per unit (Simpson diversity: `r num(all_alpha$simpson,0.01)` $\pm$ `r num(all_alpha$simpson_sd,0.01)`).
+Overall mean species richness for shallow ARMS \< 30 m was `r num(shallow_alpha$richness,0.1)` $\pm$ `r num(shallow_alpha$richness_sd,0.1)` species per unit (Simpson diversity: `r num(shallow_alpha$simpson,0.01)` $\pm$ `r num(shallow_alpha$simpson_sd,0.01)`).
+Overall mean species richness for mesophotic ARMS $\geq$ 30 m was `r num(deep_alpha$richness,0.1)` $\pm$ `r num(deep_alpha$richness_sd,0.1)` species per unit (Simpson diversity: `r num(deep_alpha$simpson,0.01)` $\pm$ `r num(deep_alpha$simpson_sd,0.01)`).
 
@@
 
-For the complete dataset, there was no significant difference in observed species richness between the MHI and NWHI (two-sample t-test: t = 1.06, DF = 106.43, p = 0.29), but Simpson diversity was significantly different, although mean values were similar with 0.78 ± 0.10 in the MHI and 0.70 ± 0.17 in the NWHI (two-sample t-test: t = 2.72, DF = 74.88, p = 0.008).
-When only shallow arms < 30 m were considered, neither comparison was significant (richness: t = 0.77, DF = 94.47, p = 0.44; Simpson: t  = 1.93, DF = 84.76, p = 0.06).
+For the complete dataset, there was no significant difference in observed species richness between the MHI and NWHI `r t_test(all_alpha$t_r_island)`, but Simpson diversity was significantly different, although mean values were similar with `r num(all_alpha$richness_main$richness,0.1)` $\pm$ `r num(all_alpha$richness_main$richness_sd,0.1)` in the MHI and `r num(all_alpha$richness_nwhi$richness,0.1)` $\pm$ `r num(all_alpha$richness_nwhi$richness_sd,0.1)` in the NWHI ([@fig:alpha_box]; `r t_test(all_alpha$t_s_island,plain=TRUE)`).
+When only shallow arms \< 30 m were considered, neither comparison was significant (richness: `r t_test(shallow_alpha$t_r_island,plain=TRUE)`; Simpson: `r t_test(shallow_alpha$t_s_island,plain=TRUE)`)<!--(richness: t = 0.77, DF = 94.47, p = 0.44; Simpson: t  = 1.93, DF = 84.76, p = 0.06)-->.
+
+```{r alpha-box, fig.cap='Hello Dolly', results="asis", dev="svg", fig.width=10, fig.height=8}
+ggplot(all_alpha$richness_table) + 
+  geom_boxplot(aes(x=island_group,y=Simpson,fill=island_group)) +
+  scale_fill_brewer(palette="Set2") +
+  theme_bw() + 
+  theme(panel.grid = element_blank())
+cat("{#fig:alpha_box}")
+```
 

### 2023-08-30 crabs.R

@@ -13,6 +13,7 @@ library(phyloseq)
 library(vegan)
 library(ggrepel)
 library(here)
+library(worrms)
 
 # to check model assumptions
 library(performance)
@@ -255,43 +256,47 @@ taxa_tibble <- function(ps,otu_col="otu") {
 # data prep ---------------------------------------------------------------
 
 setup_crabs <- function() {
+  cc <- list()
   # load our main phyloseq object and hellinger-transform it
-  crabs_untransformed <<- load_ps(
+  cc$crabs_untransformed <- load_ps(
     here("data","crabs.csv"),"unit",
     here("data","taxonomy.csv"),"otu",
     here("data","metadata.csv")
   ) 
-    crabs <<- crabs_untransformed %>%
+  
+  cc$crabs <- cc$crabs_untransformed %>%
     ps_standardize("hellinger")
     # this is the shallow subset
-  crabs_shallow <<- crabs %>%
+  cc$crabs_shallow <- cc$crabs %>%
     subset_samples(region != "mce") 
     # we're gonna mess with the sample data a bit
   # first, get rid of rows that have NAs
-  crab_data_shallow <<- crabs_shallow %>%
+  cc$crab_data_shallow <- cc$crabs_shallow %>%
     sample_tibble(sample_col = "unit") %>%
     drop_na()
     # make sure we drop samples that had NAs in their sample data
-  crabs_shallow <<- prune_samples(sample_names(crabs_shallow) %in% crab_data_shallow$unit,crabs_shallow)
+  cc$crabs_shallow <- prune_samples(sample_names(cc$crabs_shallow) %in% cc$crab_data_shallow$unit,cc$crabs_shallow)
     # now we select the columns containing the environmental variables
   # we care about and scale the numeric ones to unit variance
-  crab_data_shallow <<- crabs_shallow %>%
+  cc$crab_data_shallow <- cc$crabs_shallow %>%
     sample_tibble(sample_col = "unit") %>%
     select(unit,region,island,depth,chl,sst,slope,coral_cover,closest_island,larval_connectivity,human_impact) %>%
     mutate(across(where(is.numeric),~as.numeric(scale(.x)))) %>%
     column_to_rownames("unit") 
     # reassociate the new sample data
-  sample_data(crabs_shallow) <<- crab_data_shallow
[…]

### 2023-08-31 manuscript.Rmd
updates

@@
 Further research on the cryptic reef-dwelling fauna is needed for identifying areas of unique biodiversity, as well as boundaries between depth-stratified reef ecosystems.
-bibliography: resources/citations.bib
-csl: resources/vancouver.csl
+bibliography: references/citations.bib
+csl: references/vancouver.csl
 output:
@@
       - "--reference-doc=resources/ref.docx"
-    keep_md: false
+    keep_md: true
 tblPrefix: Table
@@
 
+library(pander)
 library(fs)
@@
 
+print_table <- function(tbl,id="",caption="") {
+  if (!grepl("^: ",caption)) {
+    caption <- paste0(": ",caption)
+  }
+
+  if (inherits(tbl,"character")) {
+    if (id == "") {
+      id <- sub(pattern = "(.*?)\\..*$", replacement = "\\1", basename(file))
+    }
+    tbl <- read.csv(tbl,check.names=FALSE,stringsAsFactors=FALSE)
+  } else if (inherits(tbl,"data.frame")) {
+    if (id == "") stop("please specify table id")
+  } else {
+    stop("'tbl' must be a character or a data frame")
+  }
+
+  caption <- sprintf("%s {#tbl:%s}",caption,id)
+
+  table_str <- capture.output(pander::pandoc.table(tbl,keep.line.breaks=TRUE,caption=NULL,style="grid",split.cells=Inf,split.tables=Inf,justify='left',missing=''))
+  table_str <- table_str[table_str != ""]
+  cat(paste0(table_str,collapse="\n"),"\n\n",caption,"\n\n",sep="")
+}
+
 cc$crabs_deep <- cc$crabs_untransformed %>% subset_samples(shallow_deep == "deep")
@@
 
-A total of 89 ARMS units were deployed over the course of six years (2010–2016; see Table xxx  for deployment and recovery schedules).
-Units were deployed by NOAA scuba divers during Pacific Reef Assessment and Monitoring Program (RAMP) research cruises in shallow coral reef habitat (12–16 m depth) in both the Main (MHI) and Northwest Hawaiian Islands (NWHI) (Table 1).
+A total of 89 ARMS units were deployed over the course of six years (2010--2016).
+Units were deployed by NOAA scuba divers during Pacific Reef Assessment and Monitoring Program (RAMP) research cruises in shallow coral reef habitat (12–16 m depth) in both the Main (MHI) and Northwest Hawaiian Islands (NWHI) ([@tbl:arms_table]).
 The sites in the MHI were at Hawai&#x02BB;i Island, Maui, O&#x02BB;ahu, and Kaua&#x02BB;i.
@@
 To explore the effects of depth on crab community structure, we also present data from a previous study by Hurley et al.
-[-@hurleyAssessmentShallowMesophotic2016], which included a total of 27 units deployed off O&#x02BB;ahu from 2009–2012 in 12–90 m depth; nine units were deployed at 12 m, and six each at 30 m, 60 m, and 90 m.
-Shallow units were deployed according to standard NOAA methodology, whereas units $\geq$ 30 m were deployed and recovered using technical mixed-gas closed-circuit rebreather diving (Table 1).
+[-@hurleyAssessmentShallowMesophotic2016], which included a total of 27 units deployed off O&#x02BB;ahu from 2009–2012 in 12–90 m depth; nine units were deployed at 12 m, and six each at 30 m, 60 m, and 90 m ([@tbl:arms_table]).
+Shallow units were deployed according to standard NOAA methodology, whereas units $\geq$ 30 m were deployed and recovered using technical mixed-gas closed-circuit rebreather diving.
+
+```{r arms-table,results="asis"}
+arms_schedule <- cc$crabs_untransformed %>%
+  sample_tibble() %>%
+  separate_wider_regex(sample,patterns=c(recovery_year="^[0-9]+","_",sample=".*"))  %>%
+  select(sample,everything()) %>%
+  unite("range",deployment_year,recovery_year,sep="--") %>%
+  mutate(
+    island = case_when(
+      str_remove_all(hawaiian_name,'[^a-zA-Z ]') != island ~ as.character(str_glue("{hawaiian_name} ({island})")),
+      TRUE ~ as.character(hawaiian_name)
+    ),
+    island = fct_reorder(island,-lat)
+  ) %>%
+  count(island,range) %>%
+  pivot_wider(names_from="range",values_from="n")  %>%
+  select(island,order(colnames(.))) %>%
+  mutate(across(-island,~replace_na(.x,0)))
+print_table(arms_schedule,"arms_table","ARMS unit sampling schedule (number of units)")  
+# print_table(here("data","tables","arms_deployment.csv"),"arms_table","ARMS unit sampling schedule (number of units)")
+```
 
@@
 
-crab_sites <- cc$crab_data %>%
-  distinct(island,island_group,lat,lon,.keep_all = FALSE) %>%
-  mutate( island = fct_reorder(island,-lat) ) %>%
+crab_sites <- cc$crabs_untransformed %>%
+  sample_tibble() %>%
+  distinct(hawaiian_name,island_group,lat,lon,.keep_all = FALSE) %>%
+  # mutate( island = fct_reorder(island,-lat) ) %>%
   st_as_sf(coords=c("lon","lat"),crs=4326, remove=F) #%>%
@@
 
-island_transform <- tribble(
-  ~island,~new_name,
-  "Kure Atoll","Hōlanikū",
-  "Pearl and Hermes Reef","Manawai",
-  "Lisianski","Kapou",
-  "French Frigate Shoals","Lalo",
-  "Kauai","Kaua‘i",
-  "Oahu","O‘ahu",
-  "Maui","Maui",
-  "Hawaii Island","Hawai‘i Island"
-)
-
-cs <- crab_sites %>%
-  left_join(island_transform,by="island") %>%
-  mutate(
-    new_name = fct_reorder(new_name,-lat)
-  )
-
 plotz <- ggplot() +
@@
   geom_sf(data=pmnm,fill="#6ed8f6",alpha=0.1) +
-  geom_sf(data=cs %>% st_transform(transformer),aes(fill=new_name),size=3,shape=21,color="black") +
+  geom_sf(data=crab_sites %>% st_transform(transformer),aes(fill=hawaiian_name),size=3,shape=21,color="black") +
   coord_sf(xlim=box_after[c(1,3)],ylim=box_after[c(2,4)]) +
@@
 
-Potential environmental factors affecting the community structure of brachyuran crabs on shallow reefs were assembled from published literature [@selkoeDNACoralReef2016] and the Pacific Islands Ocean Observing System (PacIOOS) (Table 2).
+Potential environmental factors affecting the community structure of brachyuran crabs on shallow reefs were assembled from published literature [@selkoeDNACoralReef2016] and the Pacific Islands Ocean Observing System (PacIOOS) ([@tbl:env_table]).
 Environmental factors were assembled only for shallow sites sampled during RAMP cruises.
+
+```{r env-table,results="asis"}
+print_table(here("data","tables","environmental_variables.csv"),"env_table","Potential environmental drivers of crab community structure, assembled from PacIOOS and @selkoeDNACoralReef2016")
+```
 

### 2023-08-31 crabs.R

@@ -264,6 +264,28 @@ setup_crabs <- function() {
     here("data","metadata.csv")
   ) 
   
+  island_transform <- tribble(
+    ~island,~hawaiian_name,
+    "Kure Atoll","Hōlanikū",
+    "Pearl and Hermes Reef","Manawai",
+    "Lisianski","Kapou",
+    "French Frigate Shoals","Lalo",
+    "Kauai","Kaua‘i",
+    "Oahu","O‘ahu",
+    "Maui","Maui",
+    "Hawaii Island","Hawai‘i Island"
+  )
+  
+  cd <- sample_tibble(cc$crabs_untransformed) %>%
+    left_join(island_transform,by="island") %>%
+    mutate(
+      island = fct_reorder(island,-lat),
+      hawaiian_name = fct_reorder(hawaiian_name,-lat)
+    ) %>%
+    column_to_rownames("sample")
+  
+  sample_data(cc$crabs_untransformed) <- cd
+  
   cc$crabs <- cc$crabs_untransformed %>%
     ps_standardize("hellinger")
     # this is the shallow subset

### 2023-09-01 manuscript.Rmd
multiple updates, including caching code chunks

@@
   - Molly A. Timmers:
-      institute: [himb, natgeo]
+      institute:
+        - himb
+        - natgeo
   - Kerry Reardon:
@@
     pandoc_args:
+      - "--toc-depth=2"
       - "--filter=resources/pandoc-crossref"
@@
   - "Figs."
+chapters: True
+chaptersDepth: 1
+chapDelim: ""
 ---
 
-### Keywords
+### Keywords {-}
 <div custom-style="mono">ARMs, mesophotic coral ecosystems, community structure, dispersal, Hawai&#x02BB;i, brachyuran crabs</div>
+
 \newpage
@@
   message=FALSE,
-  warning=FALSE
+  warning=FALSE,
+  cache=TRUE,
+  cache.path = "cache/",
+  cache.comments=FALSE
 )
 
+library(tidyverse)
 library(pander)
@@
 library(english)
+library(phyloseq) 
+library(vegan)
+library(ggrepel)
+library(here)
+library(worrms)
+```
 
+```{r init, include=FALSE}
 source("crabs.R")
@@
 
-top_list <- function(species,n,units,len=4) {
+pval <- function(p) {
+  case_when(
+    p < 0.001 ~ "*p* \\< 0.001",
+    p < 0.01 ~ "*p* \\< 0.01",
+    TRUE ~ as.character(str_glue("*p* = {round(p,2)}"))
+  )
+}
+
+permanova <- function(modl,d=1) {
+  m <- suppressWarnings(broom::tidy(modl)) %>%
+    filter(!term %in% c("Total","Residual"))
+  if (nrow(m) > 1) {
+    m %>%
+      pmap_chr(~{
+        with(list(...), {
+          str_glue("{term}: DF = {round(df,1)}, SS = {round(SumOfSqs,1)}, $\\mathrm{{R}}^2$ = {round(R2,1)}, pseudo-F = {round(statistic,1)}, {pval(p.value)}")
+        })
+      }) %>%
+      str_c(collapse = "; ")
+  } else {
+    with(m %>% slice(1),{
+      str_glue("DF = {round(df,1)}, SS = {round(SumOfSqs,1)}, $\\mathrm{{R}}^2$ = {round(R2,1)}, pseudo-F = {round(statistic,1)}, {pval(p.value)}")
+    }) 
+  }
+}
+
+top_list <- function(species,n,units,len=4,fig=NULL) {
   s1 <- map_chr(species,format_species)
   s2 <- map2_chr(n,units,~{
+    if (is.null(fig)) {
       str_glue("(with {.x} individuals on {.y} units)")
+    } else {
+      str_glue("(with {.x} individuals on {.y} units; [@fig:{fig}[]{{{{}}}}{{.y}}])")
+    }
   })
+  if (!is.null(fig)) {
+    s2 <- map2_chr(s2,LETTERS[seq(length(s2))],~str_glue(.x))
+  }
   ss <- str_c(s1," ",s2)
@@
 
-print_table <- function(tbl,id="",caption="") {
+print_table <- function(tbl,id="",caption="",hdr_case="sentence") {
   if (!grepl("^: ",caption)) {
@@
   }
+
+  # can also use rename_with(str_to_sentence)
+  names(tbl) <- switch(
+    str_to_lower(hdr_case),
+    sentence = str_to_sentence(names(tbl)),
+    title = str_to_title(names(tbl)),
+    names(tbl)
+  )
 
@@
 shallow_alpha <- alpha_diversity(cc$crabs_shallow)
+
+# palette for island/group colors
+island_pal <- c(
+  "Hawaii Island" = "#8a3ffc",
+  "Hawai‘i Island" = "#8a3ffc",
+  "Maui" = "#007d79",
+  "Oahu" = "#ff7eb6",
+  "O‘ahu" = "#ff7eb6",
+  "Kauai" = "#fa4d56",
+  "Kaua‘i" = "#fa4d56",
+
+  "French Frigate Shoals" = "#fff1f1",
+  "Lalo" = "#fff1f1",
+  "Lisianski" = "#6fdc8c",
+  "Kapou" = "#6fdc8c",
+  "Pearl and Hermes Reef" = "#4589ff",
+  "Manawai" = "#4589ff",
+  "Kure Atoll" = "#d12771",
+  "Hōlanikū" = "#d12771"
+)
+island_group_pal <- c(
+  "main" = "#ffb400",
+  "Main" = "#ffb400",
+  "Northwest" = "#9080ff",
+  "northwest" = "#9080ff"
+)
+
 ```
 
-# Introduction
+# Introduction {-}
 
-# Materials and Methods
+# Materials and Methods {-}
 
-## Sample Collection
+## Sample Collection {-}
 
@@
 ARMS are standardized collection devices designed to mimic coral reef structural complexity to attract settlement of reef cryptofauna [@brainardAutonomousReefMonitoring2009; @zimmermanArtificialReefMatrix2004].
-They are constructed from 10 gray type 1 PVC plates (23 cm $\times$ 23 cm) stacked in an alternating series of open and semi-enclosed layers (figure xxx, ARMS unit).
+They are constructed from 10 gray type 1 PVC plates (23 cm $\times$ 23 cm) stacked in an alternating series of open and semi-enclosed layers ([@fig:fig_arms]).
 The stack is fastened to a larger base plate (35 cm $\times$ 45 cm) which can be attached to the benthos upon deployment [@brainardAutonomousReefMonitoring2009; @hurleyAssessmentShallowMesophotic2016; @knowltonCoralReefBiodiversity2010; @zimmermanArtificialReefMatrix2004].
+
+![An Autonomous reef monitoring structure (ARMS) unit deployed on a reef in the Red Sea (photo credit: Michael Berumen, used with permission)](data/images/berumen_uncropped.png){#fig:fig_arms}
 
@@
 
-```{r arms-table,results="asis"}
+```{r arms-table,results="asis", dependson="init"}
 arms_schedule <- cc$crabs_untransformed %>%
@@
 print_table(arms_schedule,"arms_table","ARMS unit sampling schedule (number of units)") 
-# print_table(here("data","tables","arms_deployment.csv"),"arms_table","ARMS unit sampling schedule (number of units)")
 ```
 
-```{r site-map, fig.cap='Sampling map (Papahānaumokuākea Marine National Monument shaded in light blue)', results="asis", dev="svg", fig.width=9, fig.height=6}
+```{r site-map, fig.cap='Sampling map (Papahānaumokuākea Marine National Monument shaded in light blue)', results="asis", dev="svg", fig.width=9, fig.height=6, dependson="init"}
 library(sf)
@@
 
-plotz <- ggplot() +
+ggplot() +
   # geom_contour(data=bath_proj,mapping=aes(x=V1,y=V2,z = V3),  color="#cecece") +
@@
   geom_sf(data=hawaii %>% st_transform(transformer), color="black", fill="green", linewidth=0.4) +
-  scale_fill_brewer(palette="Accent",name="Island") +
   geom_sf(data=pmnm,fill="#6ed8f6",alpha=0.1) +
   geom_sf(data=crab_sites %>% st_transform(transformer),aes(fill=hawaiian_name),size=3,shape=21,color="black") +
+  scale_fill_manual(values=island_pal,name="Island") +
   coord_sf(xlim=box_after[c(1,3)],ylim=box_after[c(2,4)]) +
@@
   theme(legend.text = element_text(size=10),legend.title = element_text(size=12))
-# labela <- tribble(
-#   ~lon,~lat,~text,
-#   -170,25.4,"Paphānaumokuākea Marine\nNational Monument\n(NWHI)"
-# ) %>%
-#   st_as_sf(coords=c("lon","lat"),crs=4326, remove=F)  %>%
-#   st_transform(transformer)
-# 
-# plotz +
-#   geom_sf_text(aes(label=text),data=labela,size=3,angle=-17.0)
-plotz
 cat("{#fig:site_map}")
@@
 ARMS units were retrieved by encapsulating them in-situ in a mesh-lined crate (100 μm pore size) to prevent escape of motile organisms.
-Once returned to ship or shore, they were placed in containers of aerated, filtered seawater, crates were removed, and units were systematically disassembled plate-by-plate.
+Once returned to the surface, they were placed in containers of aerated, filtered seawater, crates were removed, and units were systematically disassembled plate-by-plate.
 Seawater surrounding disassembled units was sieved through 2 mm, 500 μm, and 100 μm mesh-size geologic sieves [see @lerayDNABarcodingMetabarcoding2015 for processing details], and retained macroorganisms were preserved in 95% ethanol.
@@
 2d).
-Individual preserved specimens were accessioned into the invertebrate zoology collections at the Florida Museum of Natural History (catalog numbers xxx–xxx ).
+Individual preserved specimens were accessioned into the invertebrate zoology collections at the Florida Museum of Natural History (catalog numbers xxx–xxx).
 
-## Data analysis
+## Data analysis {-}
 
@@
 
-## Environmental variables
+## Environmental variables {-}
 
-Potential environmental factors affecting the community structure of brachyuran crabs on shallow reefs were assembled from published literature [@selkoeDNACoralReef2016] and the Pacific Islands Ocean Observing System (PacIOOS) ([@tbl:env_table]).
+Potential environmental factors affecting the community structure of brachyuran crabs on shallow reefs were assembled from published literature [@selkoeDNACoralReef2016], NOAA's National Coral Reef Monitoring Program (NCRMP), and the Pacific Islands Ocean Observing System (PacIOOS) ([@tbl:env_table]).
 Environmental factors were assembled only for shallow sites sampled during RAMP cruises.
 
-```{r env-table,results="asis"}
+```{r env-table,results="asis", dependson="init"}
 print_table(here("data","tables","environmental_variables.csv"),"env_table","Potential environmental drivers of crab community structure, assembled from PacIOOS and @selkoeDNACoralReef2016")
@@
 
-# Results
+# Results {-}
 
-## Diversity summary
+## Diversity summary {-}
 
-### All ARMS
+### All ARMS {-}
 
@@
 Crabs were identified to `r sum(unlist(all_summary$types_breakdown))` distinct types (including `r all_summary$types_breakdown$species` species and `r all_summary$types_breakdown$morphospecies` morphospecies) in `r all_summary$taxonomy_breakdown$genera` genera and `r all_summary$taxonomy_breakdown$families` families, which accounted for ~36% of the known crab diversity reported from Hawaiian waters [@castroCatalogAnomuranBrachyuran2011].
-The four most common species were: `r with(all_summary,top_list(top5,top5_n,top5_units))`.
+The four most common species were: `r with(all_summary,top_list(top5,top5_n,top5_units,fig="fig_crabs"))`.
 `r str_to_sentence(as.english(all_summary$singleton_count))` species occurred in a single unit across all sites (singletons), accounting for ~`r num(all_summary$singleton_count/ntaxa(cc$crabs_untransformed)*100,1)`% of total recovered crab species.
 
-### Mesophotic ARMS
+![The four most abundant crab species found across all deployed ARMS units.
+(A) *Chlorodiella laevissima*.
+(B) *Perinia tumida*.
+(C) *Percnon abbreviatum*.
+(D) *Dynomene hispida*.](data/images/4crabs.png){#fig:fig_crabs}
+
+### Mesophotic ARMS {-}
 
@@
 
-## Alpha diversity
+## Alpha diversity {-}
 
-### Summary
+### Summary {-}
 
@@
 
-### Comparisons
+### Comparisons {-}
 
 For the complete dataset, there was no significant difference in observed species richness between the MHI and NWHI `r t_test(all_alpha$t_r_island)`, but Simpson diversity was significantly different, although mean values were similar with `r num(all_alpha$richness_main$richness,0.1)` $\pm$ `r num(all_alpha$richness_main$richness_sd,0.1)` in the MHI and `r num(all_alpha$richness_nwhi$richness,0.1)` $\pm$ `r num(all_alpha$richness_nwhi$richness_sd,0.1)` in the NWHI ([@fig:alpha_box]; `r t_test(all_alpha$t_s_island,plain=TRUE)`).
-When only shallow arms \< 30 m were considered, neither comparison was significant (richness: `r t_test(shallow_alpha$t_r_island,plain=TRUE)`; Simpson: `r t_test(shallow_alpha$t_s_island,plain=TRUE)`)<!--(richness: t = 0.77, DF = 94.47, p = 0.44; Simpson: t  = 1.93, DF = 84.76, p = 0.06)-->.
+When only shallow arms \< 30 m were considered, neither comparison was significant (richness: `r t_test(shallow_alpha$t_r_island,plain=TRUE)`; Simpson: `r t_test(shallow_alpha$t_s_island,plain=TRUE)`).
 
-```{r alpha-box, fig.cap='Hello Dolly', results="asis", dev="svg", fig.width=10, fig.height=8}
-ggplot(all_alpha$richness_table) + 
-  geom_boxplot(aes(x=island_group,y=Simpson,fill=island_group)) +
-  scale_fill_brewer(palette="Set2") +
+```{r alpha-box, fig.cap='Box plots of Simpson diversity by island group.
+Notches show nonparametric confidence intervals about the median', results="asis", dev="svg", fig.width=10, fig.height=8, dependson="init"}
+ggplot(all_alpha$richness_table %>% mutate(island_group = str_to_sentence(island_group))) + 
+  geom_boxplot(aes(x=island_group,y=Simpson,fill=island_group),notch=TRUE) +
+  scale_fill_manual(values=island_group_pal) +
+  guides(fill="none") + 
   theme_bw() + 
-  theme(panel.grid = element_blank())
+  theme(panel.grid = element_blank()) + 
+  xlab("Island group") + 
+  ylab("Simpson's diversity index") 
 cat("{#fig:alpha_box}")
@@
 
-### Effects of environmental variables
+### Effects of environmental variables {-}
 
-# Discussion
+## Brachyuran community structure {-}
 
-# References
+```{r permanova, dependson="init"}
+# PERMANOVA of 
+anova_region <- adonis2(cc$crab_dist_shallow ~ island_group,data=cc$crab_data_shallow,permutations=9999)
+anova_island <- adonis2(cc$crab_dist_shallow ~ island,data=cc$crab_data_shallow,permutations=9999)
+```
+
+There were significant differences in shallow crab community structure between the Main (MHI) and Northwest Hawaiian Islands (`adonis2` PERMANOVA: `r permanova(anova_region)`) as well as among individual islands (`r permanova(anova_island)`; see [@fig:pairwise_island] for results of pairwise comparisons).
+
+# Discussion {-}
+
+# References {-}
 
@@
 :::
+
+# Supplemental material {label="S"}
+
+```{r pairwise-island, fig.cap='Results of pairwise PERMANOVA tests by island (FDR-adjusted p-values)', results="asis", dev="svg", fig.width=12, fig.height=9, dependson="init"}
+library(ggtext)
+pairwise_island <- pairwise_adonis(cc$crab_dist_shallow,cc$crab_data_shallow$island,permutations=999)
+
+pw <- pairwise_island %>%
+  filter(p_adj < 0.05)
+
+islands <- cc$crabs_shallow %>%
+  sample_tibble() %>%
+  pull(island) %>%
+  levels()
+
+pw1 <- pairwise_island %>%
+  select(left,right,p_adj) %>%
+  mutate(left=factor(left,levels=rev(islands)),right=factor(right,levels=rev(islands))) %>%
+  mutate(sig = p_adj < 0.05, p_adj = pval(p_adj)) %>%
+  mutate(
+    combo = map2_chr(left,right,~str_c(sort(c(.x,.y)),collapse="-"))
+  ) %>%
+  select(-c(left,right)) %>%
+  separate_wider_delim(combo,"-",names=c("left","right")) %>%
+  select(left,right,everything()) %>%
+  mutate(left=factor(left,levels=islands),right=factor(right,levels=islands))
+
+ggplot(pw1) + 
+  # geom_tile(aes(x=right,y=left,fill=sig),width=0.6,height=0.8) + 
+  # geom_tile(aes(x=right,y=left),width=0.6,height=0.8) + 
+  geom_richtext(
+    aes(
+      x=right,
+      y=left,
+      label=p_adj,
+      fontface = if_else(sig,"bold","plain")
+    ),
+    fill = NA,
+    label.color = NA
+  ) +
+  theme_bw() + 
+  theme(
+    panel.grid = element_blank(),
+    panel.border = element_blank(),
+    axis.line = element_line(color="black")
+  ) +
+  # scale_fill_manual(values = c("white","lightgoldenrod")) +
+  guides(fill="none") +
+  xlab("") + ylab("") +
+  scale_x_discrete(position="top") +
+  scale_y_discrete(limits=rev)
+cat("{#fig:pairwise_island}")
+```

### 2023-09-01 crabs.R

@@ -2,24 +2,24 @@
 # all of these can be directly installed except for phyloseq, which is a bioconductor package
 # phyloseq isn't directly necessary, but it's what I used for all my eDNA stuff and it conveniently
 # wraps the datasets, so I'm using it here
-library(tidyverse)
+# library(tidyverse)
 # to install phyloseq in regular R, do this:
 
 # if (!require("BiocManager", quietly = TRUE))
 #   install.packages("BiocManager")
 
 # BiocManager::install("phyloseq")
-library(phyloseq) 
-library(vegan)
-library(ggrepel)
-library(here)
-library(worrms)
+# library(phyloseq) 
+# library(vegan)
+# library(ggrepel)
+# library(here)
+# library(worrms)
 
 # to check model assumptions
-library(performance)
+# library(performance)
 
 # for linear mixed effects models
-library(lme4)
+# library(lme4)
 
 # set the random seed
 set.seed(31337)
@@ -117,7 +117,7 @@ plot_ord <- function(ps,ord,arrows=TRUE,sig=FALSE,by="term",perm=999,alpha=0.05,
 # do a pairwise PERMANOVA analysis by some factor
 pairwise_adonis <- function(comm, factors, permutations = 1000, correction = "fdr", method = "bray") {
   # get possible pairwise factor combinations
-  factor_combos <- combn(unique(factors), 2)
+  factor_combos <- combn(unique(as.character(factors)), 2)
   # map through factor combinations and run model for each pair,
   # dumping the output into a tibble
   model_output <- map_dfr(array_branch(factor_combos,2), ~{
@@ -125,7 +125,7 @@ pairwise_adonis <- function(comm, factors, permutations = 1000, correction = "fd
     
     # get our factor specific community or distance matrix
     if (inherits(comm,'dist')) {
-      dd <- as.matrix(comm)[factors %in% .x, factors %in% .x]
+      dd <- as.dist(as.matrix(comm)[factors %in% .x, factors %in% .x])
     } else {
       comm <- as(comm,'matrix')
       dd <- vegdist(comm[factors %in% .x,], method = method)
@@ -302,7 +302,7 @@ setup_crabs <- function() {
   # we care about and scale the numeric ones to unit variance
   cc$crab_data_shallow <- cc$crabs_shallow %>%
     sample_tibble(sample_col = "unit") %>%
[…]

### 2023-09-07 manuscript.Rmd
again, a ton of updaes

@@
 library(worrms)
+library(BiodiversityR)
 ```
@@
 
-format_species <- function(spp) {
+plotz_theme <- function() {
+  theme_bw() + 
+    theme(
+      legend.key = element_blank(),
+      panel.grid = element_blank()
+    )
+}
+
+format_species <- function(...) UseMethod("format_species")
+
+format_species.list <- function(spp,authority=TRUE,...) {
   if (spp$found) {
-    str_glue("*{spp$species}* {spp$authority}")
+    a <- ifelse(authority,str_glue(" {spp$authority}"),"")
+    str_glue("*{spp$species}*{a}")
   } else {
     s <- str_split(spp$species," ")[[1]]
-    str_glue("*{s[1]}* {s[2]}")
+    str_trim(str_glue("*{s[1]}* {s[2]}"))
   }
+}
+
+format_species.character <- function(species,found,authority,auth=TRUE,...) {
+  spp <- map2_chr(species,found,~{
+    .y <- ifelse(is.na(.y),FALSE,.y)
+    s <- list(species=.x,found=.y,authority="")
+    format_species(s,FALSE)
+  })
+  if (auth) {
+    authority <- if_else(!is.na(authority),str_c(" ",authority),"")
+    spp <- str_c(spp,authority)
+  }
+  return(spp)
+}
+
+format_species.data.frame <- function(spp,authority=TRUE,...) {
+  spp %>%
+    pwalk(~{
+      row <- list(...)
+      # s <- list(found = row$found, authority=row$authority,otu=row$otu)
+      s <- row[c('species','found','authority','otu')]
+      print(format_species(s,authority))
+    })
 }
@@
 
-permanova <- function(modl,d=1) {
+permdisp <- function(bd,d=1,perm=999) {
+  a <- anova(bd,permutations=perm) %>%
+    broom::tidy() %>%
+    filter(term == "Groups")
+  with(a,str_glue("DF = {round(df,d)}, SS = {round(sumsq,d)}, F = {round(statistic,d)}, {pval(p.value)}"))
+}
+
+permanova <- function(modl,d=1,stat="pseudo-F",by=NULL) {
+  has_r2 <- TRUE
+  if (is(modl,"dbrda")) {
+    modl <- anova(modl,by=by)
+    has_r2 <- FALSE
+  }
   m <- suppressWarnings(broom::tidy(modl)) %>%
@@
         with(list(...), {
-          str_glue("{term}: DF = {round(df,1)}, SS = {round(SumOfSqs,1)}, $\\mathrm{{R}}^2$ = {round(R2,1)}, pseudo-F = {round(statistic,1)}, {pval(p.value)}")
+          if (has_r2) {
+            r2 <- str_glue(" $\\mathrm{{R}}^2$ = {round(R2,d)},") 
+          } else {
+            r2 <- ""
+          }
+          str_glue("{term}: DF = {round(df,d)}, SS = {round(SumOfSqs,d)},{r2} {stat} = {round(statistic,d)}, {pval(p.value)}")
         })
@@
     with(m %>% slice(1),{
-      str_glue("DF = {round(df,1)}, SS = {round(SumOfSqs,1)}, $\\mathrm{{R}}^2$ = {round(R2,1)}, pseudo-F = {round(statistic,1)}, {pval(p.value)}")
+      if (has_r2) {
+        r2 <- str_glue(" $\\mathrm{{R}}^2$ = {round(R2,d)},") 
+      } else {
+        r2 <- ""
+      }
+      str_glue("{term}: DF = {round(df,d)}, SS = {round(SumOfSqs,d)},{r2} {stat} = {round(statistic,d)}, {pval(p.value)}")
+      # str_glue("DF = {round(df,d)}, SS = {round(SumOfSqs,d)}, $\\mathrm{{R}}^2$ = {round(R2,d)}, pseudo-F = {round(statistic,d)}, {pval(p.value)}")
     }) 
@@
 
-print_table <- function(tbl,id="",caption="",hdr_case="sentence") {
+print_table <- function(tbl,id="",caption="",hdr_case="sentence",hide=character(0)) {
   if (!grepl("^: ",caption)) {
@@
   }
+
+  tbl <- tbl %>%
+    select(-any_of(hide))
 
@@
 } else {
-  all_summary <<- sample_summary(cc$crabs_untransformed)
+  all_summary <<- sample_summary(cc$crabs_untransformed,4)
   write_rds(all_summary,here("cache","all_summary.rds"))
@@
 } else {
-  deep_summary <<- sample_summary(cc$crabs_deep)
+  deep_summary <<- sample_summary(cc$crabs_deep,4)
   write_rds(deep_summary,here("cache","deep_summary.rds"))
@@
 
+# colors for sample group / depth zone
+sample_depth_pal <- c(
+  "shallow" = "#ffb400",
+  "Shallow (MHI)" = "#ffb400",
+  "shallow_main" = "#ffb400",
+  "Shallow (NWHI)" = "#9080ff",
+  "shallow_nwhi" = "#9080ff",
+  "30 m" = "#d87939",
+  "30m" = "#d87939",
+  "60 m" = "#04c3c8",
+  "60m" = "#04c3c8",
+  "90 m" = "#04738d",
+  "90m" = "#04738d",
+  "deep" = "#04738d"
+)
+
+PERM <<- 999
+
 ```
@@
 The sites in the MHI were at Hawai&#x02BB;i Island, Maui, O&#x02BB;ahu, and Kaua&#x02BB;i.
-The sites in the NWHI were at Lalo (French Frigate Shoals), Kapou (Lisianski), Manawai (Pearl & Hermes), and Hōlanikū (Kure Atoll) ([@fig:site_map]).
+The sites in the NWHI were at Lalo (French Frigate Shoals), Kapou (Lisianski), Manawai (Pearl & Hermes), and Hōlanikū (Kure Atoll) ([@fig:fig_map]).
 To explore the effects of depth on crab community structure, we also present data from a previous study by Hurley et al.
@@
 
-```{r site-map, fig.cap='Sampling map (Papahānaumokuākea Marine National Monument shaded in light blue)', results="asis", dev="svg", fig.width=9, fig.height=6, dependson="init"}
+```{r site-map, fig.cap='Sampling map (Papahānaumokuākea Marine National Monument shaded in light blue)', results="asis", fig.width=9, fig.height=6, dependson="init"}
 library(sf)
@@
   theme(legend.text = element_text(size=10),legend.title = element_text(size=12))
-cat("{#fig:site_map}")
+cat("{#fig:fig_map}")
 ```
@@
 Colinearity between traits was assessed by calculating variance inflation factors (VIF) using the function `vif.cca` from complete and reduced models and discarding variables where VIF > 10.
-Significance levels were assessed with the function `anova` using permutation tests with 9,999 permutations.
+Significance levels were assessed with the function `anova` using permutation tests with `r num(PERM,1)` permutations.
 All continuous environmental variables used for dbRDA were scaled to unit variance.
-We used the permutational analysis of variance (PERMANOVA) method in the `adonis2` function from vegan to assess variation among sites and/or depths across the islands, with 9,999 permutations to assess significance in both term-based and marginal tests.
+We assessed community variation among groups using the `betadisper` and `adonis2` functions from vegan.
+These functions implement analysis of multivariate homogeneity of group dispersions (PERMDISP) and multivariate permutational analysis of variance (PERMANOVA), respectively.
+The PERMDISP method is used to assess variability within groups and PERMANOVA is used to assess differences among groups.
+We assessed significance with permutational tests, using `r num(PERM,1)` permutations.
+Crab community variation was further examined and visualized using canonical analysis of principal coordinates (CAP) based on discriminant analysis [@andersonCanonicalAnalysisPrincipal2003] as implemented in the `CAPdiscrim` function in the `BiodiversityR` package [@kindtTreeDiversityAnalysis2005].
 
@@
 ```{r env-table,results="asis", dependson="init"}
-print_table(here("data","tables","environmental_variables.csv"),"env_table","Potential environmental drivers of crab community structure, assembled from PacIOOS and @selkoeDNACoralReef2016")
+print_table(here("data","tables","environmental_variables.csv"),"env_table","Potential environmental drivers of crab community structure, assembled from PacIOOS and @selkoeDNACoralReef2016",hide="variable")
 ```
@@
 Crabs were identified to `r sum(unlist(all_summary$types_breakdown))` distinct types (including `r all_summary$types_breakdown$species` species and `r all_summary$types_breakdown$morphospecies` morphospecies) in `r all_summary$taxonomy_breakdown$genera` genera and `r all_summary$taxonomy_breakdown$families` families, which accounted for ~36% of the known crab diversity reported from Hawaiian waters [@castroCatalogAnomuranBrachyuran2011].
-The four most common species were: `r with(all_summary,top_list(top5,top5_n,top5_units,fig="fig_crabs"))`.
+The four most abundant species were: `r with(all_summary,top_list(top,top_n,top_units,fig="fig_crabs"))`.
 `r str_to_sentence(as.english(all_summary$singleton_count))` species occurred in a single unit across all sites (singletons), accounting for ~`r num(all_summary$singleton_count/ntaxa(cc$crabs_untransformed)*100,1)`% of total recovered crab species.
@@
 (C) *Percnon abbreviatum*.
-(D) *Dynomene hispida*.](data/images/4crabs.png){#fig:fig_crabs}
+(D) *Dynomene hispida*.
+Photo credit: Florida Museum of Natural History, licensed [CC BY-NC 3.0](https://creativecommons.org/licenses/by-nc/3.0/)](data/images/4crabs.png){#fig:fig_crabs}
 
@@
 The `r nsamples(cc$crabs_deep)` units deployed at mesophotic depths $\geq$ 30 m on O&#x02BB;ahu yielded both fewer species and fewer individuals compared to the complete dataset, with `r num(sum(sample_sums(cc$crabs_deep)),1)` crabs overall, a total of `r sum(unlist(deep_summary$types_breakdown))` distinct types (`r deep_summary$types_breakdown$species` species and `r deep_summary$types_breakdown$morphospecies` morphospecies), `r num(deep_summary$abundance_mean,0.1)` $\pm$ `r num(deep_summary$abundance_sd,0.1)` individuals per unit, and abundance ranging from `r format_range(deep_summary$abundance_range)` crabs.
-The four most abundant species in the deep ARMS did not overlap with those from the complete dataset: `r with(deep_summary,top_list(top5,top5_n,top5_units))`.
+The four most abundant species in the deep ARMS did not overlap with those from the complete dataset: `r with(deep_summary,top_list(top,top_n,top_units))`.
 There were `r deep_summary$singleton_count` singletons, which proportionally accounted for ~`r num(deep_summary$singleton_count/ntaxa(cc$crabs_deep)*100,1)`% of species found on deep ARMS.
@@
 
-For the complete dataset, there was no significant difference in observed species richness between the MHI and NWHI `r t_test(all_alpha$t_r_island)`, but Simpson diversity was significantly different, although mean values were similar with `r num(all_alpha$richness_main$richness,0.1)` $\pm$ `r num(all_alpha$richness_main$richness_sd,0.1)` in the MHI and `r num(all_alpha$richness_nwhi$richness,0.1)` $\pm$ `r num(all_alpha$richness_nwhi$richness_sd,0.1)` in the NWHI ([@fig:alpha_box]; `r t_test(all_alpha$t_s_island,plain=TRUE)`).
+For the complete dataset, there was no significant difference in observed species richness between the MHI and NWHI `r t_test(all_alpha$t_r_island)`, but Simpson diversity was significantly different, although mean values were similar with `r num(all_alpha$richness_main$richness,0.1)` $\pm$ `r num(all_alpha$richness_main$richness_sd,0.1)` in the MHI and `r num(all_alpha$richness_nwhi$richness,0.1)` $\pm$ `r num(all_alpha$richness_nwhi$richness_sd,0.1)` in the NWHI ([@fig:fig_alpha]; `r t_test(all_alpha$t_s_island,plain=TRUE)`).
 When only shallow arms \< 30 m were considered, neither comparison was significant (richness: `r t_test(shallow_alpha$t_r_island,plain=TRUE)`; Simpson: `r t_test(shallow_alpha$t_s_island,plain=TRUE)`).
@@
 ```{r alpha-box, fig.cap='Box plots of Simpson diversity by island group.
-Notches show nonparametric confidence intervals about the median', results="asis", dev="svg", fig.width=10, fig.height=8, dependson="init"}
+Notches represent nonparametric confidence intervals about the median', results="asis", dev="svg", fig.width=10, fig.height=8, dependson="init"}
 ggplot(all_alpha$richness_table %>% mutate(island_group = str_to_sentence(island_group))) + 
@@
   guides(fill="none") + 
-  theme_bw() + 
-  theme(panel.grid = element_blank()) + 
+  plotz_theme() + 
   xlab("Island group") + 
   ylab("Simpson's diversity index") 
-cat("{#fig:alpha_box}")
+cat("{#fig:fig_alpha}")
 ```
@@
 # PERMANOVA of 
-anova_region <- adonis2(cc$crab_dist_shallow ~ island_group,data=cc$crab_data_shallow,permutations=9999)
-anova_island <- adonis2(cc$crab_dist_shallow ~ island,data=cc$crab_data_shallow,permutations=9999)
+anova_region <- adonis2(cc$crab_dist_shallow ~ island_group,data=cc$crab_data_shallow,permutations=PERM)
+anova_island <- adonis2(cc$crab_dist_shallow ~ island,data=cc$crab_data_shallow,permutations=PERM)
+bd_region <- betadisper(cc$crab_dist_shallow,cc$crab_data_shallow$island_group)
+
+anova_depth <- adonis2(cc$crab_dist ~ depth_zone,data=cc$crab_data,permutations=PERM)
+bd_depth <- betadisper(cc$crab_dist,cc$crab_data$depth_zone)
+
+anova_both <- adonis2(cc$crab_dist ~ island_group + depth_zone, data=cc$crab_data,permutations = PERM, by="margin")
+effects <- broom::tidy(anova_both) %>%
+  filter(!term %in% c("Residual","Total")) %>%
+  pull(R2)
+effect_ratio <- max(effects)/min(effects)
+
+cd <- cc$crabs_deep %>%
+  sample_tibble()
+bd_deep <- betadisper(distance(cc$crabs_deep,"bray"),cd$depth_zone)
 ```
 
-There were significant differences in shallow crab community structure between the Main (MHI) and Northwest Hawaiian Islands (`adonis2` PERMANOVA: `r permanova(anova_region)`) as well as among individual islands (`r permanova(anova_island)`; see [@fig:pairwise_island] for results of pairwise comparisons).
+```{r rda-analysis, results="hide", dependson="init"}
+db_upr <- dbrda(cc$crab_otus_shallow ~ depth + chl + sst + slope + coral_cover + closest_island + larval_connectivity + human_impact,data=cc$crab_data_shallow,distance="bray")
+db_lwr <- dbrda(cc$crab_otus_shallow  ~ 1,data=cc$crab_data_shallow,distance="bray")
+
+# do our forward selection to find the best model
+# set trace to TRUE if you want to see a bunch of output
+db_modl <- ordiR2step(db_lwr,db_upr,trace=FALSE)
+
+# see marginal effects of model terms
+# anova(db_modl,by="margin")
+
+# vif.cca(db_upr)
+vif.cca(db_modl)
+
+# warning: the following *may* be sketchy:
+# dbRDA doesn't automatically give you species scores, so let's assign them
+# but the community data needs to be transformed to make sense
+# see https://github.com/vegandevs/vegan/issues/254 for information on the transformation below
+sppscores(db_modl) <- sqrt(decostand(cc$crab_otus_shallow, "total")/2)
+
+# this was to grab the 4 most abundant species overall
+top_species <- names(head(rev(sort(taxa_sums(cc$crabs_shallow))),n=4))
+
+# extract the species scores for the species we want
+species_scores <- scores(db_modl,display="species") %>%
+  as_tibble(rownames = "species") %>%
+  filter(species %in% top_species)
+
+model_terms <- read_csv(here("data","tables","environmental_variables.csv")) %>%
+  select(variable,driver=`Environmental Driver`)
+rda_table <- anova(db_modl,by="term") %>%
+  broom::tidy() %>%
+  inner_join(model_terms,by=c("term" = "variable")) %>%
+  # rename(Term=driver) %>%
+  select(-term) %>%
+  select(Term=driver,DF=df,`Sum of Squares`=SumOfSqs,`F-statistic`=statistic,`p-value`=p.value)
+
+abs_r2 <- diff(c(0,db_modl$anova$R2.adj))
+```
+### Environmental drivers {-}
+
+The best dbRDA model (`r permanova(db_modl,stat="F")`) explained `r scales::percent(RsquareAdj(db_modl)$adj.r.squared,0.01)` (adjusted $\mathrm{R}^2$) of the variability in crab communities across the archipelago ([@fig:fig_rda]).
+Seven of the 10 environmental variables tested were significant ([@tbl:rda_table]).
+The drivers contributing most to the explained variation were chlorophyll-A ($\mathrm{R}^2$ = `r num(abs_r2[1],0.01)`) and human impacts ($\mathrm{R}^2$ = `r num(abs_r2[2],0.01)`).
+
+```{r rda, fig.cap="Distance-based redundancy analysis (dbRDA) plot showing significant environmental drivers of crab community variation across the Hawaiian Islands.", results="asis", fig.width=9, fig.height=6, dependson="rda-analysis"}
+rda_plotz <- plot_ord(cc$crabs_shallow,db_modl,scale=2,color="island_group") + 
+  stat_ellipse(geom="polygon",aes(group=island_group,fill=island_group),color=NA,alpha=0.1) +
+  geom_point(aes(fill=island_group),shape=21,color="black",size=3) +
+  plotz_theme() + 
+  scale_fill_manual(values=island_group_pal)
+
+xvar <- eigenvals(db_modl)/sum(eigenvals(db_modl))
+r2 <- RsquareAdj(db_modl)$adj.r.squared
+constrained <- xvar*r2
+
+splained <- function(name,constrained,total,accuracy=0.1) {
+  str_glue("{name} ({scales::percent(constrained,accuracy)} constrained, {scales::percent(total,accuracy)} total)")
+}
+
+rda_plotz + 
+  xlab(splained("dbRDA1",xvar[1],constrained[1])) +
+  ylab(splained("dbRDA2",xvar[2],constrained[2])) 
+cat("{#fig:fig_rda}")
+```
+
+### Community variation {-}
+
+```{r cap-analysis,dependson="init",echo=FALSE,message=FALSE,warning=FALSE,results="hide"}
+# convert otu tables and sample data into appropriate format
+otus <- otu_table(cc$crabs) %>%
+  as.data.frame()
+cd <- cc$crab_data %>%
+  mutate(sample_group = factor(sample_group,ordered=FALSE)) %>%
+  as.data.frame()
+
+# run CAP analysis, taking care of negative eigenvalues where necessary
+cap_depth <- CAPdiscrim(otus ~ sample_group, data=cd, dist="bray", axes=2, m=0, add=TRUE)
+# # calculate proportion of explained variance
+cap_var <- cap_depth$manova$Eigenvalues / sum(cap_depth$manova$Eigenvalues)
+cap_var <- list(x = cap_var[1], y = cap_var[2])
+```
+
+Shallow crab community structure was significantly different between the MHI and NWHI (`r permanova(anova_region,2)`) as well as among individual islands (`r permanova(anova_island,2)`; see [@fig:fig_pairwise] for results of pairwise comparisons).
+There was no significant difference in community dispersion between the MHI and NWHI (`r permdisp(bd_region)`).
+When mesophotic ($\geq$ 30 m) units were included, there were significant differences among depth zones (\< 30 m, 30 m, 60 m, 90 m) (`r permanova(anova_depth,2)`) and community dispersion (`r permdisp(bd_depth)`).
+However, community dispersion was not different among depth zones $\geq$ 30 m (`r permdisp(bd_deep)`).
+Although crab communities varied significantly among both depth zones and island groups (`r permanova(anova_both,2)`), the effect size (model $\mathrm{R}^2$) of depth was `r num(effect_ratio,0.1)` times greater than that of island group.
+In the CAP analysis, samples were successfully assigned to their respective sampling groups `r scales::percent(cap_depth$percent/100,0.1)` of the time.
+The first two axes explained `r scales::percent(cap_var$x,0.1)` and `r scales::percent(cap_var$y,0.1)` of the variation, respectively.
+The samples from the three mesophotic groups on O&#x02BB;ahu grouped separately from the shallow samples collected across the entire ~2,500 km of the Hawaiian Islands ([@fig:fig_cap]).
+In addition to grouping apart from shallow samples, there was as much or more variation across the 30--90 m samples than across all of the shallow samples, although that variation was largely due to the differences between the 30 m samples and the 60 m and 90 m samples ([@fig:fig_cap]).
+
+```{r cap, fig.cap="Canonical analysis of principal coordinates (CAP) plot showing variation in brachyuran crab communities by sampling region and depth zone.
+Ellipses represent 95% confidence intervals for sample groups.", results="asis", fig.width=9, fig.height=6, dependson="cap-analysis"}
+# make the sample group look nice
+nice <- function(s) {
+  str_replace(s,"shallow","Shallow") %>%
+    str_replace("main","(MHI)") %>%
+    str_replace("nwhi","(NWHI)") %>%
+    str_split("((?<=[0-9])(?=[a-zA-Z]))|(_)") %>%
+    map_chr(~str_c(.x,collapse=" "))
+}
+
+# get all the stuff we need to plot
+pcoa <- cap_depth$x %>%
+  as_tibble(rownames="sample") %>%
+  left_join(cd,by="sample") %>%
+  mutate(
+    sample_group = nice(sample_group),
+    sample_group = fct_reorder(sample_group,depth)
+  )
+
+ggplot(pcoa,aes(x=LD1,y=LD2,fill=sample_group)) + 
+  stat_ellipse(geom="polygon",level=0.95,alpha=0.1,show.legend = FALSE) +
+  geom_point(color="black",shape=21,size=5)  +
+  # scale_color_manual(values=sample_depth_pal) + 
+  scale_fill_manual(values=sample_depth_pal,name="Sample group") +
+  xlab(str_glue("Linear discriminant 1 [{scales::percent(cap_var$x,0.1)}]")) +
+  ylab(str_glue("Linear discriminant 2 [{scales::percent(cap_var$y,0.1)}]")) +
+  plotz_theme()
+cat("{#fig:fig_cap}")
+```
+
+<!-- 
+Here's the thing: it's not that the differences between MHI & NWHI become non-significant with the inclusion of deep samples, 
+it's just that the deep samples are MORE different from all the shallow samples than all the shallow samples are from one another?
+We can think of PERMANOVA R^2 and ANOSIM R as effect sizes. Thus, it may be a bit weak but for shallow samples only, island group 
+R2 = 0.072, for all samples R2 = 0.065. For depth (all samples of course), R2 = 0.110. If we consider the effect sizes from ANOSIM, 
+for shallow samples, island group R = 0.175, for all samples, island group R = 0.161, and for shallow vs deep,  R = 0.509. 
+The apparent difference in effect size is much more pronounced in the ANOSIM, but it's different in each case, and it's smaller
+for all samples than it is for just shallow samples.
+-->
+
+### Effects of environmental variables {-}
 
@@
 
-```{r pairwise-island, fig.cap='Results of pairwise PERMANOVA tests by island (FDR-adjusted p-values)', results="asis", dev="svg", fig.width=12, fig.height=9, dependson="init"}
+```{r pairwise-island, fig.cap='Significance levels of pairwise PERMANOVA tests by island (FDR-adjusted p-values)', results="asis", dev="svg", fig.width=12, fig.height=9, dependson="init"}
 library(ggtext)
-pairwise_island <- pairwise_adonis(cc$crab_dist_shallow,cc$crab_data_shallow$island,permutations=999)
+pairwise_island <- pairwise_adonis(cc$crab_dist_shallow,cc$crab_data_shallow$island,permutations=PERM)
 
@@
   mutate(left=factor(left,levels=rev(islands)),right=factor(right,levels=rev(islands))) %>%
-  mutate(sig = p_adj < 0.05, p_adj = pval(p_adj)) %>%
+  mutate(sig = round(p_adj,2) < 0.05, p_adj = pval(p_adj)) %>%
   mutate(
@@
 ggplot(pw1) + 
-  # geom_tile(aes(x=right,y=left,fill=sig),width=0.6,height=0.8) + 
-  # geom_tile(aes(x=right,y=left),width=0.6,height=0.8) + 
   geom_richtext(
@@
     label.color = NA
+    # size=10
   ) +
-  theme_bw() + 
+  plotz_theme() +
   theme(
-    panel.grid = element_blank(),
     panel.border = element_blank(),
-    axis.line = element_line(color="black")
+    axis.line = element_line(color="black"),
+    axis.text = element_text(size = 14),
+    axis.text.x = element_text(angle=30,vjust=1,hjust=0)
   ) +
-  # scale_fill_manual(values = c("white","lightgoldenrod")) +
   guides(fill="none") +
@@
   scale_y_discrete(limits=rev)
-cat("{#fig:pairwise_island}")
+cat("{#fig:fig_pairwise}")
 ```
+
+\newpage
+
+```{r rda-table,results="asis",dependson="rda-analysis"}
+print_table(rda_table,"rda_table","Significant terms in the dbRDA analysis of crab community variation across the Hawaiian Archipelago",hdr_case="none") 
+```

### 2023-09-07 cap.R

@@ -1,65 +1,56 @@
 library(BiodiversityR)
 
-
-crabdata <- sample_data(crabs) %>%
-  as("data.frame")
-crabmatrix <- otu_table(crabs) %>%
-  as("matrix") %>%
+otus <- otu_table(cc$crabs) %>%
+  as.data.frame()
+cd <- cc$crab_data %>%
+  mutate(sample_group = factor(sample_group,ordered=FALSE)) %>%
   as.data.frame()
 
-crabdata <- crabdata %>%
-  dplyr::select(dplyr::where(~!any(is.na(.x)))) %>%
-  mutate(across(where(is.character),as.factor))
 
 
-cap <- CAPdiscrim(
-  crabmatrix ~ island_group + sample_group,
-  data=crabdata,
-  dist="bray",
-  axes=2,
-  m=10,
-  mmax=10,
-  add=FALSE#,
-  # permutations = 99
-)
+# cap <- CAPdiscrim(otus ~ sample_group, data=cd, dist="bray", axes=2, m=0, add=TRUE,permutations = 10)
+cap <- CAPdiscrim(otus ~ island_group + depth_zone, data=cd, dist="bray", axes=2, m=0, add=TRUE)
+cap <- CAPdiscrim(otus ~ sample_group, data=cd, dist="bray", axes=2, m=0, add=TRUE)
 
-cd <- crabdata %>%
-  as_tibble(rownames="sample")
+var <- cap$manova$Eigenvalues / sum(cap$manova$Eigenvalues)
+var <- list(x = var[1], y = var[2])
 
-capscores <- scores(cap) %>%
-  as_tibble(rownames="sample")
-
-capdata <- cap$PCoA %>%
-  as_tibble(rownames="sample",.name_repair = ~c("pcoa1","pcoa2")) %>%
-  left_join(cd,by="sample") %>%
-  left_join(capscores,by="sample")
-
-# quartz()
-ggplot(capdata,aes(x=pcoa1,y=pcoa2)) + 
-  geom_point(aes(shape=island_group,color=sample_group),size=3) +
-  stat_ellipse(aes(group=sample_group,color=sample_group),level=0.95)
-
-ggplot(capdata,aes(x=LD1,y=LD2)) + 
-  geom_point(aes(shape=island_group,color=sample_group)) +
-  stat_ellipse(aes(group=sample_group,color=sample_group),level=0.95) +
-  scale_y_reverse()
-
-
-plot1 <- ordiplot(cap, type="none")
[…]

### 2023-09-07 crabs.R

@@ -28,6 +28,7 @@ run_things <- FALSE
 
 # functions ---------------------------------------------------------------
 
+
 # plot an ordination with various options
 plot_ord <- function(ps,ord,arrows=TRUE,sig=FALSE,by="term",perm=999,alpha=0.05,scale=1,labelsize=5,pointsize=3,...) {
   
@@ -321,7 +322,7 @@ setup_crabs <- function() {
   return(cc)
 }
 
-sample_summary <- function(ps) {
+sample_summary <- function(ps,top_n=5) {
   ss <- list() 
   
   # basic numbers
@@ -342,52 +343,51 @@ sample_summary <- function(ps) {
     sd()
   
   # get 5 most common species
-  top5 <- ps %>%
+  top <- ps %>%
     taxa_sums() %>%
     sort() %>%
     rev() %>%
-    head(5) 
+    head(top_n) 
   
-  otus <- top5 %>%
+  otus <- top %>%
     names() %>%
-    str_replace_all("_"," ") 
+    set_names(.,str_replace_all(.,"_"," "))
   
-  ss$top5_n <- top5
+  ss$top_n <- top
   
   
   to_keep <- ps %>%
     taxa_tibble() %>%
-    filter(species %in% otus) %>%
+    filter(species %in% names(otus)) %>%
     pull(otu)
   
-  # ss$top5_units <- ps %>% 
-  #   subset_taxa(species %in% otus) %>%
-  ss$top5_units <- prune_taxa(to_keep,ps) %>%
+  ss$top_units <- prune_taxa(to_keep,ps) %>%
     ps_standardize("pa") %>%
     taxa_sums() %>% 
     sort() %>%
     rev()
   
   # make it the same order as the top 5 species
-  ss$top5_units <- ss$top5_units[match(names(ss$top5_units),names(top5))]
+  ss$top_units <- ss$top_units[match(names(ss$top_units),names(top))]
   
-  ss$top5 <- wm_records_taxamatch(otus) %>%
-    map2(otus,~{
+  ss$top <- wm_records_taxamatch(names(otus)) %>%
+    map2(names(otus),~{
       if (nrow(.x) > 0) {
         .x <- .x %>% 
           filter(rank == "Species") %>%
           select(species = scientificname, authority) %>%
           slice(1)  %>%
-          mutate(found = TRUE)
+          mutate(found = TRUE,otu = otus[.y])
       } 
       if (nrow(.x) == 0) {
[…]

### 2023-09-07 cap.R
deleting some unused files

@@ -1,56 +1,0 @@
-library(BiodiversityR)
-
-otus <- otu_table(cc$crabs) %>%
-  as.data.frame()
-cd <- cc$crab_data %>%
-  mutate(sample_group = factor(sample_group,ordered=FALSE)) %>%
-  as.data.frame()
-
-
-
-# cap <- CAPdiscrim(otus ~ sample_group, data=cd, dist="bray", axes=2, m=0, add=TRUE,permutations = 10)
-cap <- CAPdiscrim(otus ~ island_group + depth_zone, data=cd, dist="bray", axes=2, m=0, add=TRUE)
-cap <- CAPdiscrim(otus ~ sample_group, data=cd, dist="bray", axes=2, m=0, add=TRUE)
-
-var <- cap$manova$Eigenvalues / sum(cap$manova$Eigenvalues)
-var <- list(x = var[1], y = var[2])
-
-nice <- function(s) {
-  
-}
-
-pcoa <- cap$x %>%
-  as_tibble(rownames="sample") %>%
-  left_join(cd,by="sample") %>%
-  mutate(
-    sample_group = str_replace(sample_group,"shallow","Shallow"),
-    sample_group = str_replace(sample_group,"main","(MHI)"),
-    sample_group = str_replace(sample_group,"nwhi","(NWHI)"),
-    sample_group = str_split(sample_group,"((?<=[0-9])(?=[a-zA-Z]))|(_)") %>%
-      map_chr(~str_c(.x,collapse=" ")),
-    sample_group = fct_reorder(sample_group,depth)
-  )
-
-pal <- c("Shallow (MHI)" = "#ffb400", "Shallow (NWHI)" = "#9080ff","30 m" = "#d87939", "60 m" = "#04c3c8", "90 m" = "#04738d")
-
-# ggplot(pcoa,aes(x=LD1,y=LD2,color=sample_group)) + 
-#   stat_ellipse(geom="polygon",aes(fill=sample_group,color=NULL),level=0.95,alpha=0.1) +
-#   geom_point()  +
-#   scale_color_manual(values=pal) + 
-#   scale_fill_manual(values=pal)
-
-ggplot(pcoa,aes(x=LD1,y=LD2,fill=sample_group)) + 
-  stat_ellipse(geom="polygon",level=0.95,alpha=0.1,show.legend = FALSE) +
-  geom_point(color="black",shape=21,size=5)  +
-  # scale_color_manual(values=pal) + 
[…]

### 2023-09-07 map.R

@@ -1,104 +1,0 @@
-library(tidyverse)
-library(sf)
-library(marmap)
-library(fs)
-library(here)
-library(rnaturalearth)
-
-if (!exists("crab_data")) {
-  crab_data <- read_csv(here("data","metadata.csv"),col_types = cols())
-}
-
-crab_sites <- crab_data %>%
-  distinct(island,island_group,lat,lon,.keep_all = FALSE) %>%
-  mutate( island = fct_reorder(island,-lat) ) %>%
-  st_as_sf(coords=c("lon","lat"),crs=4326, remove=F) #%>%
-
-hawaii <- ne_states("united states of america",returnclass = "sf") %>%
-  filter(name == "Hawaii")
-
-# run this section --------------------------------------------------------
-trans_box <- function(b,crs) {
-  box_before <- st_sfc(
-    st_point(b[c(1,2)]),
-    st_point(b[c(3,4)]),
-    crs=4326
-  )
-  box_before %>%
-    st_transform(crs) %>%
-    st_bbox()
-}
-
-box_before <- st_bbox(st_buffer(crab_sites,units::as_units(50,"km")))
-bath <- getNOAA.bathy(lon1 = box_before[1], lon2 = box_before[3], lat1 = box_before[2], lat2 = box_before[4], resolution = 3) 
-transformer <- str_glue("+proj=leac +lat_1={box_before[2]} +lat_2={box_before[4]} +lon_0={mean(box_before[c(1,3)])}")
-
-box_after <- box_before %>%
-  trans_box(transformer)
-
-bath_proj <- as.raster(bath) %>% 
-  raster::projectRaster(crs = transformer) %>% 
-  as.bathy() %>% 
-  as.xyz() %>%
-  as_tibble() %>%
-  drop_na(V3) %>%
-  filter(V3 < 0)
-
-bpr <- as.raster(bath) %>% 
-  raster::projectRaster(crs = transformer) %>%
-  as.bathy() %>%
-  fortify.bathy()
-
-pmnm <- st_read(here("data","gis","pmnm.shp")) %>%
-  st_transform(transformer)
-
-
-# section break -----------------------------------------------------------
-
-island_transform <- tribble(
-  ~island,~new_name,
-  "Kure Atoll","Hōlanikū",
-  "Pearl and Hermes Reef","Manawai",
[…]

### 2023-09-07 prep.R

@@ -1,51 +1,0 @@
-library(tidyverse)
-library(here)
-
-
-all_arms <- read_csv(here("data","all_data.csv"),col_types=cols())
-
-islands <- tribble(
-  ~island,~island_name,
-  "OAH","Oahu",
-  "FFS","French Frigate Shoals",
-  "HAW","Hawaii Island",
-  "KAU","Kauai",
-  "KUR","Kure Atoll",
-  "LIS","Lisianski",
-  "MAI","Maui",
-  "PHR","Pearl and Hermes Reef"
-)
-
-metadata <- all_arms %>%
-  distinct(unit,.keep_all = T) %>%
-  select(unit,region,island,deployment_year,recovery_year,depth,lat,lon) %>%
-  inner_join(islands,by="island") %>%
-  select(-island) %>%
-  rename(island=island_name) %>% 
-  select(unit,island,everything()) 
-
-write_csv(metadata,here("data","metadata.csv"))
-  
-crabs_long <- read_csv(here("data","crabs_long.csv"),col_types = cols())
-crabs <- all_arms %>%
-  group_by(unit,species,family,genus,taxon_level,region,island,depth,lat,lon) %>%
-  summarise(count = sum(count)) %>%
-  ungroup() %>%
-  select(unit,species,count) %>%
-  arrange(unit,species) %>%
-  pivot_wider(names_from="species",values_from="count",values_fill = 0)
-
-write_csv(crabs,here("data","crabs.csv"))
-
-
-taxonomy <- all_arms %>%
-  select(species,family,genus,taxon_level) %>%
-  distinct(species,.keep_all = T) %>%
-  rename(otu = species) %>%
-  mutate(species = str_replace_all(otu,"_"," ")) %>%
-  select(otu,taxon_level,family,genus,species)
-  
-write_csv(taxonomy,here("data","taxonomy.csv"))
-
-
-metadata <- read_csv(here("data","all_data.csv"))

### 2023-09-07 manuscript.Rmd
updated arms table and fixed image link

@@
 
-![An Autonomous reef monitoring structure (ARMS) unit deployed on a reef in the Red Sea (photo credit: Michael Berumen, used with permission)](data/images/berumen_uncropped.png){#fig:fig_arms}
+![An autonomous reef monitoring structure (ARMS) unit deployed on a reef in the Red Sea (photo credit: Michael Berumen, used with permission)](data/images/arms_example.png){#fig:fig_arms}
 
@@
       TRUE ~ as.character(hawaiian_name)
+    ),
+    island = case_when(
+      island == "O‘ahu" ~ str_glue("{hawaiian_name} ({shallow_deep})"),
+      TRUE ~ island
     ),

### 2023-09-08 manuscript.Rmd
more updates

@@
 bibliography: references/citations.bib
-csl: references/vancouver.csl
+csl: references/style.csl
 output:
@@
 figPrefix:
-  - "Fig."
-  - "Figs."
+  - "figure"
+  - "figures"
 chapters: True
@@
 chapDelim: ""
+header-includes:
+  - \usepackage{bm}
 ---
@@
       row <- list(...)
-      # s <- list(found = row$found, authority=row$authority,otu=row$otu)
       s <- row[c('species','found','authority','otu')]
@@
 
-num <- function(x,accuracy=0.01) {
-  scales::number(x,big.mark = ",",decimal.mark = ".",accuracy=accuracy)
-}
+pc <- function(x,a=0.01) scales::percent_format(accuracy = a)(x)
+num <- function(x,a=0.01) scales::label_comma(accuracy=a)(x)
 
@@
   if (plain) {
-    str_glue("*t* = {num(test$statistic,0.01)}, DF = {num(test$parameter['df'],0.01)}, *p* = {num(test$p.value,0.01)}")
+    str_glue("*t* = {num(test$statistic,0.01)}, DF = {num(test$parameter['df'],0.01)}, {pval(test$p.value)}")
   } else {
-    str_glue("(two-sample t-test: *t* = {num(test$statistic,0.01)}, DF = {num(test$parameter['df'],0.01)}, *p* = {num(test$p.value,0.01)})")
+    str_glue("(two-sample t-test: *t* = {num(test$statistic,0.01)}, DF = {num(test$parameter['df'],0.01)}, {pval(test$p.value)})")
   }
@@
     p < 0.01 ~ "*p* \\< 0.01",
-    TRUE ~ as.character(str_glue("*p* = {round(p,2)}"))
+    TRUE ~ as.character(str_glue("*p* = {num(p,0.01)}"))
   )
@@
 
-permdisp <- function(bd,d=1,perm=999) {
+permdisp <- function(bd,accuracy=0.01,perm=999) {
   a <- anova(bd,permutations=perm) %>%
@@
     filter(term == "Groups")
-  with(a,str_glue("DF = {round(df,d)}, SS = {round(sumsq,d)}, F = {round(statistic,d)}, {pval(p.value)}"))
+  with(a,str_glue("DF = {num(df,accuracy)}, SS = {num(sumsq,accuracy)}, F = {num(statistic,accuracy)}, {pval(p.value)}"))
 }
 
-permanova <- function(modl,d=1,stat="pseudo-F",by=NULL) {
+permanova <- function(modl,accuracy=0.01,stat="pseudo-F",by=NULL,sentence=TRUE) {
   has_r2 <- TRUE
@@
           if (has_r2) {
-            r2 <- str_glue(" $\\mathrm{{R}}^2$ = {round(R2,d)},") 
+            r2 <- str_glue(" $\\mathrm{{R}}^2$ = {num(R2,0.01)},") 
           } else {
@@
           }
-          str_glue("{term}: DF = {round(df,d)}, SS = {round(SumOfSqs,d)},{r2} {stat} = {round(statistic,d)}, {pval(p.value)}")
+          if (sentence) {
+            term <- str_replace_all(term,"_"," ")
+          }
+          str_glue("{term}: DF = {num(df,accuracy)}, SS = {num(SumOfSqs,accuracy)},{r2} {stat} = {num(statistic,accuracy)}, {pval(p.value)}")
         })
@@
       if (has_r2) {
-        r2 <- str_glue(" $\\mathrm{{R}}^2$ = {round(R2,d)},") 
+        r2 <- str_glue(" $\\mathrm{{R}}^2$ = {num(R2,accuracy)},") 
       } else {
@@
       }
-      str_glue("{term}: DF = {round(df,d)}, SS = {round(SumOfSqs,d)},{r2} {stat} = {round(statistic,d)}, {pval(p.value)}")
-      # str_glue("DF = {round(df,d)}, SS = {round(SumOfSqs,d)}, $\\mathrm{{R}}^2$ = {round(R2,d)}, pseudo-F = {round(statistic,d)}, {pval(p.value)}")
+      if (sentence) {
+        term <- str_replace_all(term,"_"," ")
+      }
+      str_glue("{term}: DF = {num(df,accuracy)}, SS = {num(SumOfSqs,accuracy)},{r2} {stat} = {num(statistic,accuracy)}, {pval(p.value)}")
+      # str_glue("DF = {num(df,accuracy)}, SS = {num(SumOfSqs,accuracy)}, $\\mathrm{{R}}^2$ = {num(R2,accuracy)}, pseudo-F = {num(statistic,accuracy)}, {pval(p.value)}")
     }) 
@@
 
-print_table <- function(tbl,id="",caption="",hdr_case="sentence",hide=character(0)) {
+print_table <- function(tbl,id="",caption="",hdr_case="sentence",hide=character(0),accuracy = 0.01) {
   if (!grepl("^: ",caption)) {
@@
   tbl <- tbl %>%
-    select(-any_of(hide))
+    select(-any_of(hide)) %>%
+    mutate(
+      across(
+        where(is.numeric),
+        ~num(.x,accuracy)
+      )
+    )
 
@@
   "shallow" = "#ffb400",
+  "Shallow" = "#ffb400",
   "Shallow (MHI)" = "#ffb400",
@@
   "90m" = "#04738d",
-  "deep" = "#04738d"
+  "deep" = "#04738d",
+  "Deep" = "#04738d"
 )
@@
 
-# Introduction {-}
+# 1. Introduction {-}
 
-# Materials and Methods {-}
+# 2. Materials and Methods {-}
 
@@
 [-@hurleyAssessmentShallowMesophotic2016], which included a total of 27 units deployed off O&#x02BB;ahu from 2009–2012 in 12–90 m depth; nine units were deployed at 12 m, and six each at 30 m, 60 m, and 90 m ([@tbl:arms_table]).
-Shallow units were deployed according to standard NOAA methodology, whereas units $\geq$ 30 m were deployed and recovered using technical mixed-gas closed-circuit rebreather diving.
+Shallow units were deployed according to standard NOAA methodology, whereas units $\geq$ 30 m were deployed and recovered using technical mixed-gas closed-circuit rebreather diving [see @hurleyAssessmentShallowMesophotic2016].
 
@@
   mutate(across(-island,~replace_na(.x,0)))
-print_table(arms_schedule,"arms_table","ARMS unit sampling schedule (number of units)") 
+print_table(arms_schedule,"arms_table","ARMS unit deployment schedule (number of units)",accuracy=1) 
 ```
@@
 We assessed significance with permutational tests, using `r num(PERM,1)` permutations.
-Crab community variation was further examined and visualized using canonical analysis of principal coordinates (CAP) based on discriminant analysis [@andersonCanonicalAnalysisPrincipal2003] as implemented in the `CAPdiscrim` function in the `BiodiversityR` package [@kindtTreeDiversityAnalysis2005].
+Crab community variation among sampling groups (shallow MHI vs NWHI and O&#x02BB;ahu 30 m, 60 m, 90 m) was further examined and visualized using canonical analysis of principal coordinates (CAP) based on discriminant analysis [@andersonCanonicalAnalysisPrincipal2003] as implemented in the `CAPdiscrim` function in the `BiodiversityR` package [@kindtTreeDiversityAnalysis2005].
 
@@
 
-# Results {-}
+# 3. Results {-}
 
@@
 
-For the complete dataset, there was no significant difference in observed species richness between the MHI and NWHI `r t_test(all_alpha$t_r_island)`, but Simpson diversity was significantly different, although mean values were similar with `r num(all_alpha$richness_main$richness,0.1)` $\pm$ `r num(all_alpha$richness_main$richness_sd,0.1)` in the MHI and `r num(all_alpha$richness_nwhi$richness,0.1)` $\pm$ `r num(all_alpha$richness_nwhi$richness_sd,0.1)` in the NWHI ([@fig:fig_alpha]; `r t_test(all_alpha$t_s_island,plain=TRUE)`).
+For the complete dataset, there was no significant difference in observed species richness between the MHI and NWHI `r t_test(all_alpha$t_r_island)`, but Simpson diversity was significantly different, although mean values were similar with `r num(all_alpha$richness_main$richness,0.1)` $\pm$ `r num(all_alpha$richness_main$richness_sd,0.1)` in the MHI and `r num(all_alpha$richness_nwhi$richness,0.1)` $\pm$ `r num(all_alpha$richness_nwhi$richness_sd,0.1)` in the NWHI ([@fig:fig_alpha[]{}A]; `r t_test(all_alpha$t_s_island,plain=TRUE)`).
+Simpson diversity was significantly different between shallow and deep samples ([@fig:fig_alpha[]{}B]; `r t_test(all_alpha$t_s_depth,plain=TRUE)`) but richness was not (`r t_test(all_alpha$t_r_depth,plain=TRUE)`).
 When only shallow arms \< 30 m were considered, neither comparison was significant (richness: `r t_test(shallow_alpha$t_r_island,plain=TRUE)`; Simpson: `r t_test(shallow_alpha$t_s_island,plain=TRUE)`).
 
-```{r alpha-box, fig.cap='Box plots of Simpson diversity by island group.
-Notches represent nonparametric confidence intervals about the median', results="asis", dev="svg", fig.width=10, fig.height=8, dependson="init"}
-ggplot(all_alpha$richness_table %>% mutate(island_group = str_to_sentence(island_group))) + 
+```{r alpha-box, fig.cap='Box plots of Simpson diversity by (A) island group and (B) sampling depth.
+Notches represent nonparametric confidence intervals about the median', results="asis", dev="svg", fig.width=8, fig.height=4, dependson="init"}
+library(patchwork)
+
+p1 <- ggplot(all_alpha$richness_table %>% mutate(island_group = str_to_sentence(island_group))) + 
   geom_boxplot(aes(x=island_group,y=Simpson,fill=island_group),notch=TRUE) +
@@
   ylab("Simpson's diversity index") 
+p2 <- ggplot(all_alpha$richness_table %>% mutate(shallow_deep = str_to_sentence(shallow_deep),shallow_deep = fct_reorder(shallow_deep,depth))) + 
+  geom_boxplot(aes(x=shallow_deep,y=Simpson,fill=shallow_deep),notch=TRUE) +
+  scale_fill_manual(values=sample_depth_pal) +
+  guides(fill="none") + 
+  plotz_theme()  + 
+  xlab("Sampling depth") + 
+  ylab("Simpson's diversity index")
+
+p1 + p2 +
+  plot_annotation(tag_levels = "A")
 cat("{#fig:fig_alpha}")
@@
 
+```{r rda-analysis, results="hide", dependson="init"}
+db_upr <- dbrda(cc$crab_otus_shallow ~ depth + chl + sst + slope + coral_cover + closest_island + larval_connectivity + human_impact,data=cc$crab_data_shallow,distance="bray")
+db_lwr <- dbrda(cc$crab_otus_shallow  ~ 1,data=cc$crab_data_shallow,distance="bray")
+
+# do our forward selection to find the best model
+# set trace to TRUE if you want to see a bunch of output
+db_modl <- ordiR2step(db_lwr,db_upr,trace=FALSE)
+
+# see marginal effects of model terms
+# anova(db_modl,by="margin")
+
+# vif.cca(db_upr)
+vifs <- vif.cca(db_modl)
+
+# warning: the following *may* be sketchy:
+# dbRDA doesn't automatically give you species scores, so let's assign them
+# but the community data needs to be transformed to make sense
+# see https://github.com/vegandevs/vegan/issues/254 for information on the transformation below
+sppscores(db_modl) <- sqrt(decostand(cc$crab_otus_shallow, "total")/2)
+
+# this was to grab the 4 most abundant species overall
+top_species <- names(head(rev(sort(taxa_sums(cc$crabs_shallow))),n=4))
+
+# extract the species scores for the species we want
+species_scores <- scores(db_modl,display="species") %>%
+  as_tibble(rownames = "species") %>%
+  filter(species %in% top_species)
+
+model_terms <- read_csv(here("data","tables","environmental_variables.csv")) %>%
+  select(variable,driver=`Environmental Driver`)
+
+rda_table <- anova(db_modl,by="term") %>%
+  broom::tidy() %>%
+  inner_join(model_terms,by=c("term" = "variable")) %>%
+  select(-term) %>%
+  select(Term=driver,DF=df,`Sum of Squares`=SumOfSqs,`F-statistic`=statistic,`p-value`=p.value)
+
+abs_r2 <- diff(c(0,db_modl$anova$R2.adj))
+```
+### Environmental drivers {-}
+
+The best dbRDA model (`r permanova(db_modl,stat="F")`) explained `r pc(RsquareAdj(db_modl)$adj.r.squared,0.01)` (adjusted $\mathrm{R}^2$) of the variability in crab communities across the archipelago ([@fig:fig_rda]).
+Seven of the 10 environmental variables tested were significant ([@tbl:rda_table]).
+The drivers contributing most to the explained variation were chlorophyll-A ($\mathrm{R}^2$ = `r num(abs_r2[1],0.01)`) and human impacts ($\mathrm{R}^2$ = `r num(abs_r2[2],0.01)`).
+
+```{r rda, fig.cap="Distance-based redundancy analysis (dbRDA) plot showing significant environmental drivers of crab community variation across the Hawaiian Islands.", results="asis", fig.width=9, fig.height=6, dependson="rda-analysis"}
+rda_plotz <- plot_ord(cc$crabs_shallow,db_modl,scale=2,color="island_group") + 
+  stat_ellipse(geom="polygon",aes(group=island_group,fill=island_group),color=NA,alpha=0.1) +
+  geom_point(aes(fill=island_group),shape=21,color="black",size=3) +
+  plotz_theme() + 
+  scale_fill_manual(values=island_group_pal)
+
+xvar <- eigenvals(db_modl)/sum(eigenvals(db_modl))
+r2 <- RsquareAdj(db_modl)$adj.r.squared
+constrained <- xvar*r2
+
+splained <- function(name,constrained,total,accuracy=0.1) {
+  str_glue("{name} ({pc(constrained,accuracy)} constrained, {pc(total,accuracy)} total)")
+}
+
+# put the arrows on top
+rda_plotz$layers <- rda_plotz$layers[c(1,2,5,6,3,4)]
+
+# plot the plotz
+rda_plotz + 
+  xlab(splained("dbRDA1",xvar[1],constrained[1])) +
+  ylab(splained("dbRDA2",xvar[2],constrained[2])) 
+cat("{#fig:fig_rda}")
+```
+
+### Regional & depth comparisons {-}
+
 ```{r permanova, dependson="init"}
-# PERMANOVA of 
 anova_region <- adonis2(cc$crab_dist_shallow ~ island_group,data=cc$crab_data_shallow,permutations=PERM)
@@
 
-```{r rda-analysis, results="hide", dependson="init"}
-db_upr <- dbrda(cc$crab_otus_shallow ~ depth + chl + sst + slope + coral_cover + closest_island + larval_connectivity + human_impact,data=cc$crab_data_shallow,distance="bray")
-db_lwr <- dbrda(cc$crab_otus_shallow  ~ 1,data=cc$crab_data_shallow,distance="bray")
-
-# do our forward selection to find the best model
-# set trace to TRUE if you want to see a bunch of output
-db_modl <- ordiR2step(db_lwr,db_upr,trace=FALSE)
-
-# see marginal effects of model terms
-# anova(db_modl,by="margin")
-
-# vif.cca(db_upr)
-vif.cca(db_modl)
-
-# warning: the following *may* be sketchy:
-# dbRDA doesn't automatically give you species scores, so let's assign them
-# but the community data needs to be transformed to make sense
-# see https://github.com/vegandevs/vegan/issues/254 for information on the transformation below
-sppscores(db_modl) <- sqrt(decostand(cc$crab_otus_shallow, "total")/2)
-
-# this was to grab the 4 most abundant species overall
-top_species <- names(head(rev(sort(taxa_sums(cc$crabs_shallow))),n=4))
-
-# extract the species scores for the species we want
-species_scores <- scores(db_modl,display="species") %>%
-  as_tibble(rownames = "species") %>%
-  filter(species %in% top_species)
-
-model_terms <- read_csv(here("data","tables","environmental_variables.csv")) %>%
-  select(variable,driver=`Environmental Driver`)
-rda_table <- anova(db_modl,by="term") %>%
-  broom::tidy() %>%
-  inner_join(model_terms,by=c("term" = "variable")) %>%
-  # rename(Term=driver) %>%
-  select(-term) %>%
-  select(Term=driver,DF=df,`Sum of Squares`=SumOfSqs,`F-statistic`=statistic,`p-value`=p.value)
-
-abs_r2 <- diff(c(0,db_modl$anova$R2.adj))
-```
-### Environmental drivers {-}
-
-The best dbRDA model (`r permanova(db_modl,stat="F")`) explained `r scales::percent(RsquareAdj(db_modl)$adj.r.squared,0.01)` (adjusted $\mathrm{R}^2$) of the variability in crab communities across the archipelago ([@fig:fig_rda]).
-Seven of the 10 environmental variables tested were significant ([@tbl:rda_table]).
-The drivers contributing most to the explained variation were chlorophyll-A ($\mathrm{R}^2$ = `r num(abs_r2[1],0.01)`) and human impacts ($\mathrm{R}^2$ = `r num(abs_r2[2],0.01)`).
-
-```{r rda, fig.cap="Distance-based redundancy analysis (dbRDA) plot showing significant environmental drivers of crab community variation across the Hawaiian Islands.", results="asis", fig.width=9, fig.height=6, dependson="rda-analysis"}
-rda_plotz <- plot_ord(cc$crabs_shallow,db_modl,scale=2,color="island_group") + 
-  stat_ellipse(geom="polygon",aes(group=island_group,fill=island_group),color=NA,alpha=0.1) +
-  geom_point(aes(fill=island_group),shape=21,color="black",size=3) +
-  plotz_theme() + 
-  scale_fill_manual(values=island_group_pal)
-
-xvar <- eigenvals(db_modl)/sum(eigenvals(db_modl))
-r2 <- RsquareAdj(db_modl)$adj.r.squared
-constrained <- xvar*r2
-
-splained <- function(name,constrained,total,accuracy=0.1) {
-  str_glue("{name} ({scales::percent(constrained,accuracy)} constrained, {scales::percent(total,accuracy)} total)")
-}
-
-rda_plotz + 
-  xlab(splained("dbRDA1",xvar[1],constrained[1])) +
-  ylab(splained("dbRDA2",xvar[2],constrained[2])) 
-cat("{#fig:fig_rda}")
-```
-
-### Community variation {-}
-
 ```{r cap-analysis,dependson="init",echo=FALSE,message=FALSE,warning=FALSE,results="hide"}
@@
 ```
+<!-- 
+Here's the thing: it's not that the differences between MHI & NWHI become non-significant with the inclusion of deep samples, 
+it's just that the deep samples are MORE different from all the shallow samples than all the shallow samples are from one another?
+We can think of PERMANOVA R^2 and ANOSIM R as effect sizes. Thus, it may be a bit weak but for shallow samples only, island group 
+R2 = 0.072, for all samples R2 = 0.065. For depth (all samples of course), R2 = 0.110. If we consider the effect sizes from ANOSIM, 
+for shallow samples, island group R = 0.175, for all samples, island group R = 0.161, and for shallow vs deep,  R = 0.509. 
+The apparent difference in effect size is much more pronounced in the ANOSIM, but it's different in each case, and it's smaller
+for all samples than it is for just shallow samples.
+-->
 
-Shallow crab community structure was significantly different between the MHI and NWHI (`r permanova(anova_region,2)`) as well as among individual islands (`r permanova(anova_island,2)`; see [@fig:fig_pairwise] for results of pairwise comparisons).
+Shallow crab community structure was significantly different between island groups (MHI vs.
+NWHI) (`r permanova(anova_region,0.01)`) as well as among individual islands (`r permanova(anova_island,0.01)`; see [@fig:fig_pairwise] for results of pairwise comparisons).
 There was no significant difference in community dispersion between the MHI and NWHI (`r permdisp(bd_region)`).
-When mesophotic ($\geq$ 30 m) units were included, there were significant differences among depth zones (\< 30 m, 30 m, 60 m, 90 m) (`r permanova(anova_depth,2)`) and community dispersion (`r permdisp(bd_depth)`).
+When mesophotic ($\geq$ 30 m) ARMS units were included, there were significant differences among depth zones (\< 30 m, 30 m, 60 m, 90 m) (`r permanova(anova_depth,0.01)`) and community dispersion (`r permdisp(bd_depth)`).
 However, community dispersion was not different among depth zones $\geq$ 30 m (`r permdisp(bd_deep)`).
-Although crab communities varied significantly among both depth zones and island groups (`r permanova(anova_both,2)`), the effect size (model $\mathrm{R}^2$) of depth was `r num(effect_ratio,0.1)` times greater than that of island group.
-In the CAP analysis, samples were successfully assigned to their respective sampling groups `r scales::percent(cap_depth$percent/100,0.1)` of the time.
-The first two axes explained `r scales::percent(cap_var$x,0.1)` and `r scales::percent(cap_var$y,0.1)` of the variation, respectively.
-The samples from the three mesophotic groups on O&#x02BB;ahu grouped separately from the shallow samples collected across the entire ~2,500 km of the Hawaiian Islands ([@fig:fig_cap]).
-In addition to grouping apart from shallow samples, there was as much or more variation across the 30--90 m samples than across all of the shallow samples, although that variation was largely due to the differences between the 30 m samples and the 60 m and 90 m samples ([@fig:fig_cap]).
+Although crab communities varied significantly among both depth zones and island groups when examining the complete dataset (`r permanova(anova_both,0.01)`), depth explained `r num(effect_ratio,0.1)` times variability than island group (based on model $\mathrm{R}^2$).
+In the CAP analysis, ARMS units were correctly assigned to sampling groups `r pc(cap_depth$percent/100,0.1)` of the time.
+The first two axes explained `r pc(cap_var$x,0.1)` and `r pc(cap_var$y,0.1)` of the variation, respectively.
+The mesophotic samples from O&#x02BB;ahu grouped separately from the shallow samples collected across the entire ~2,500 km of the Hawaiian Islands ([@fig:fig_cap]).
+In addition, there was as much variability across the 30--90 m samples as across all shallow samples, with the 30 m samples spread out along the LD2 (y) axis and the 60 m and 90 m samples forming a tight group ([@fig:fig_cap]).
 
-```{r cap, fig.cap="Canonical analysis of principal coordinates (CAP) plot showing variation in brachyuran crab communities by sampling region and depth zone.
+```{r cap, fig.cap="Canonical analysis of principal coordinates (CAP) plot showing variation in brachyuran crab communities by island group and depth zone.
 Ellipses represent 95% confidence intervals for sample groups.", results="asis", fig.width=9, fig.height=6, dependson="cap-analysis"}
@@
   scale_fill_manual(values=sample_depth_pal,name="Sample group") +
-  xlab(str_glue("Linear discriminant 1 [{scales::percent(cap_var$x,0.1)}]")) +
-  ylab(str_glue("Linear discriminant 2 [{scales::percent(cap_var$y,0.1)}]")) +
+  xlab(str_glue("Linear discriminant 1 [{pc(cap_var$x,0.1)}]")) +
+  ylab(str_glue("Linear discriminant 2 [{pc(cap_var$y,0.1)}]")) +
   plotz_theme()
@@
 
-<!-- 
-Here's the thing: it's not that the differences between MHI & NWHI become non-significant with the inclusion of deep samples, 
-it's just that the deep samples are MORE different from all the shallow samples than all the shallow samples are from one another?
-We can think of PERMANOVA R^2 and ANOSIM R as effect sizes. Thus, it may be a bit weak but for shallow samples only, island group 
-R2 = 0.072, for all samples R2 = 0.065. For depth (all samples of course), R2 = 0.110. If we consider the effect sizes from ANOSIM, 
-for shallow samples, island group R = 0.175, for all samples, island group R = 0.161, and for shallow vs deep,  R = 0.509. 
-The apparent difference in effect size is much more pronounced in the ANOSIM, but it's different in each case, and it's smaller
-for all samples than it is for just shallow samples.
--->
-
 ### Effects of environmental variables {-}
 
-# Discussion {-}
+# 4. Discussion {-}
 
@@
 ```{r rda-table,results="asis",dependson="rda-analysis"}
-print_table(rda_table,"rda_table","Significant terms in the dbRDA analysis of crab community variation across the Hawaiian Archipelago",hdr_case="none") 
+print_table(rda_table,"rda_table","Significant terms in the dbRDA analysis of crab community variation across the Hawaiian Archipelago",hdr_case="none",accuracy=0.001) 
 ```

### 2023-09-08 crabs.R

@@ -104,7 +104,8 @@ plot_ord <- function(ps,ord,arrows=TRUE,sig=FALSE,by="term",perm=999,alpha=0.05,
         color = "black", 
         arrow = arrowhead
       ) + 
-      geom_text_repel(
+      # geom_text_repel(
+      geom_text(
         mapping = label_map, 
         size = labelsize,  
         data = arrowdf, 
@@ -451,6 +452,8 @@ alpha_diversity <- function(ps,measures=c("Observed","Simpson")) {
   
   try(ad$t_r_island <- t.test(Observed ~ island_group, data=all_richness),silent=TRUE)
   try(ad$t_s_island <- t.test(Simpson ~ island_group, data=all_richness),silent=TRUE)
+  try(ad$t_r_depth <- t.test(Observed ~ shallow_deep, data=all_richness),silent=TRUE)
+  try(ad$t_s_depth <- t.test(Simpson ~ shallow_deep, data=all_richness),silent=TRUE)
   
   # m <- lm(Observed ~ island_group, data=all_richness)
   # summary(m)

### 2023-09-12 manuscript.Rmd
a bunch of updates, including adding in some introduction text and getting cites figured out

@@
     if (is.null(fig)) {
-      str_glue("(with {.x} individuals on {.y} units)")
+      str_glue("({.x} individuals on {.y} units)")
     } else {
-      str_glue("(with {.x} individuals on {.y} units; [@fig:{fig}[]{{{{}}}}{{.y}}])")
+      str_glue("({.x} individuals on {.y} units; [@fig:{fig}[]{{{{}}}}{{.y}}])")
     }
@@
   "Lalo" = "#fff1f1",
+  "Lalo (French Frigate Shoals)" = "#fff1f1",
   "Lisianski" = "#6fdc8c",
   "Kapou" = "#6fdc8c",
+  "Kapou (Lisianski)" = "#6fdc8c",
   "Pearl and Hermes Reef" = "#4589ff",
   "Manawai" = "#4589ff",
+  "Manawai (Pearl and Hermes Reef)" = "#4589ff",
   "Kure Atoll" = "#d12771",
-  "Hōlanikū" = "#d12771"
+  "Hōlanikū" = "#d12771",
+  "Hōlanikū (Kure Atoll)" = "#d12771"
 )
@@
 
+model_terms <- read_csv(here("data","tables","environmental_variables.csv"))
 ```
@@
 # 1. Introduction {-}
+
+Natural communities are structured by a suite of biotic and abiotic factors including geography, habitat complexity, environment, productivity, and species interactions (among many others) [@begonEcologyIndividualsEcosystems2021].
+Such complexity makes direct comparisons of processes structuring communities over large geographic scales challenging.
+Marine communities experience unique influences and present particular research challenges, however studies have largely concluded that factors structuring marine communities differ from those in terrestrial habitats [@steeleComparisonTerrestrialMarine1985; @kinlanPropaguleDispersalMarine2003; @raffaelliMarineTerrestrialEcologists2005; @grosbergBiodiversityWaterLand2012; @knappDriversBiodiversityChange2017].
+Among the best studied of aquatic ecosystems is the littoral zone lying between the low-tide line and the upper limit of aquatic species on the shore where the habitat is exposed to air during low tide events that provide easy and consistent research access for humans.
+Classical studies of vertical zonation across the intertidal have contributed much of our understanding of the mechanisms of biotic zonation and community structure in marine environments [@raffaelliIntertidalEcology2012; @birdTideWaveMarks2013].
+Such studies point to an inseparable combination of physical and biological factors, including physiological tolerance [@someroThermalPhysiologyVertical2002] and species interactions [@brunoHabitatModificationFacilitation2001; @mengeRockyIntertidalCommunities2001] that together determine the structure of intertidal communities [@dennyPhysicalProcessesThat2001].
+Although there are exceptions [e.g., @woottonIndirectEffectsPrey1992; @bertnessTestingRelativeContribution1999], the generalization has emerged that biotic factors typically determine the lower limit whereas physical factors set the upper limit of the depth distribution of intertidal communities [@connellInfluenceInterspecificCompetition1961; @connellEffectsCompetitionPredation1961; @paineIntertidalCommunityStructure1974].
+
+No such generalization has emerged for the processes structuring coral reef communities [@moraEcologyFishesCoral2015; @sheppardBiologyCoralReefs2017], with hypotheses ranging from the Neutral Theory of Biodiversity [@hubbellUnifiedNeutralTheory2011] and Lottery Recruitment [@saleStockRecruitRelationshipsRegional1982] to competitive networks [@bussCompetitiveNetworksNontransitive1979] and fine-scale resource partitioning driven by predators and herbivores structuring coral reef communities [@abramsRecruitmentLotteriesCoexistence1984; @hixonPredationPreyRefuges1993].
+Zooxanthellate corals are the ecosystem engineers of reef habitats but are restricted to specific areas characterized by a relatively narrow range of chemical and physical environmental variables (i.e., tropical and subtropical warm waters, oligotrophic conditions, 32-40 PSU salinity, and sufficient incident light, which decreases rapidly below ~30 m depth).
+Shallow coral reef ecosystems cover less than 0.2% of the seafloor and in such a small area they hold the 32-38% of all marine species [@fisherSpeciesRichnessCoral2015].
+Yet the drivers of high biodiversity and the mechanisms by which diversity is maintained on coral reefs remains a subject of considerable debate [@carrBiodiversityPopulationRegulation2002; @connollyIndoPacificBiodiversityCoral2003; @almanyDoesIncreasedHabitat2004; @bellwoodEnvironmentalGeometricConstraints2005, 2006; @dornelasCoralReefDiversity2006; @grahamImportanceStructuralComplexity2013; @komyakovaRelativeImportanceCoral2013].
+
+Recent advances in underwater explorations have shown that light-dependent reefs extend well beyond the 30 m depth limit imposed by scuba exploration of coral reefs down to 150 m in some locations [@kahngDeepestZooxanthellateScleractinian2006].
+These mesophotic coral ecosystems (MCEs) are typically defined as those extending from ~40 to 150 m depth [@hindersteinThemeSectionMesophotic2010; @kahngCommunityEcologyMesophotic2010], and were believed to experience dampened disturbances in comparison to shallow reefs and could therefore act as deep water refugia for shallow reef species threatened by anthropogenic impacts [@bongaertsAssessingDeepReef2010; @smithDepthRefugiumCatastrophic2014; @holsteinFertileFathomsDeep2015; @muirSpeciesIdentityDepth2017].
+However, recent literature describing species compositions along depth gradients reveal a more complex picture among taxa and locations because some groups may use mesophotic reefs as a refuge in some places, while others show community turnover with depth such that MCE habitats are ecologically distinct from adjacent shallow water coral reefs [@hurleyAssessmentShallowMesophotic2016; @smithCaribbeanMesophoticCoral2016; @semmlerAreMesophoticCoral2017; @bongaertsDeepReefsAre2017; @rochaMesophoticCoralEcosystems2018].
+Further, there are a range of connections depending on the taxa – large, mobile organisms like fish tend to have more shared species across depth than sedentary organisms such as corals and algae, but even among the fishes the community structure of MCEs is often distinct [@kaneHighLevelsMesophotic2014; @pyleComprehensiveInvestigationMesophotic2016 ; @asherMesophoticDepthGradients2017; @fukunagaChangesMesophoticReef2017; @colemanShiftingReefFish2018; @pinheiroDeepReefFishes2019].
+Although the majority of coral reef biodiversity consists of the algal and invertebrate cryptobenthic organisms living on and within the reef structure [@fautinOverviewMarineBiodiversity2010; @knowltonCoralReefBiodiversity2010; @fisherSpeciesRichnessCoral2015], the studies listed above focus almost exclusively on the corals and fishes which are more easily surveyed.
+
+Studies comparing community structure either along a depth gradient or across a latitudinal gradients have been criticized because they may confound multiple aspects of physical environments with habitat area or habitat complexity that covary across the measured gradient in such surveys [e.g., @bellwoodRegionalScaleAssemblyRules2001; @grahamEffectsMarineReserve2003; @connollyCommunityStructureCorals2005; @ewersConfoundingFactorsDetection2006; @fletcherRoleHabitatArea2007a; ).
+Here, we use standardized habitat modules, autonomous reef monitoring structures or ARMS [@zimmermanArtificialReefMatrix2004; @brainardAutonomousReefMonitoring2009], to sample brachyuran crab communities collected from across the Hawaiian Archipelago and test which suite of environmental or anthropogenic drivers are most important in structuring coral reef infauna communities.
+The Hawaiian Archipelago is made up of both high islands to atolls stretching across 2500 km in the Pacific Ocean, and is typically regarded as two regions: the main Hawaiian Islands (MHI) are the 8 high volcanic islands, and the Northwestern Hawaiian Islands (NWHI) are the low islands and atolls north and west of Kauaʻi.
+We compared shallow reef crab communities gathered from across the archipelago based on the nested groupings of island versus region (MHI vs NWHI) spanning roughly 9 degrees of latitude and 24 degrees of longitude that encompass a wide range of physical and anthropogenic gradients [@selkoeMapHumanImpacts2009, 2016).
+Given the number and magnitude of differences between these regions, we hypothesized that species diversity would be different between regions (MHI vs NWHI).
+We therefore looked at community composition along the latitudinal gradient and investigated community composition in response to physical gradients (depth, temperature), biological gradients (coral cover, chlorophyll concentration), geological gradients (island slope, distances between neighboring islands), and anthropogenic gradients (human population count, human impacts).
+We hypothesized that the communities found across the latitudinal gradient encompassing all of these environmental variables would show fewer differences between individual islands or atolls, but would be greatest between the two regions, MHI and NWHI.
+Finally, to investigate whether a latitudinal gradient or depth gradient would be a stronger force in driving community composition, we compared the reef crab assemblage from across the shallow reefs in the Hawaiian Archipelago to reef crab assemblages across a mesophotic depth gradient.
 
@@
 A total of 89 ARMS units were deployed over the course of six years (2010--2016).
-Units were deployed by NOAA scuba divers during Pacific Reef Assessment and Monitoring Program (RAMP) research cruises in shallow coral reef habitat (12–16 m depth) in both the Main (MHI) and Northwest Hawaiian Islands (NWHI) ([@tbl:arms_table]).
-The sites in the MHI were at Hawai&#x02BB;i Island, Maui, O&#x02BB;ahu, and Kaua&#x02BB;i.
+Units were deployed by NOAA scuba divers during Pacific Reef Assessment and Monitoring Program (RAMP) research cruises in shallow coral reef habitat (12–16 m depth) in both the Main (MHI) and Northwest Hawaiian Islands (NWHI) ([@fig:fig_map; @tbl:arms_table]).
+<!--The sites in the MHI were at Hawai&#x02BB;i Island, Maui, O&#x02BB;ahu, and Kaua&#x02BB;i.
 The sites in the NWHI were at Lalo (French Frigate Shoals), Kapou (Lisianski), Manawai (Pearl & Hermes), and Hōlanikū (Kure Atoll) ([@fig:fig_map]).
-To explore the effects of depth on crab community structure, we also present data from a previous study by Hurley et al.
+-->To explore the effects of depth on crab community structure, we also present data from a previous study by Hurley et al.
 [-@hurleyAssessmentShallowMesophotic2016], which included a total of 27 units deployed off O&#x02BB;ahu from 2009–2012 in 12–90 m depth; nine units were deployed at 12 m, and six each at 30 m, 60 m, and 90 m ([@tbl:arms_table]).
@@
 
-```{r arms-table,results="asis", dependson="init"}
-arms_schedule <- cc$crabs_untransformed %>%
+```{r site-map, fig.cap='Sampling map (Papahānaumokuākea Marine National Monument is shaded in light blue)', results="asis", fig.width=9, fig.height=6, dependson="init"}
+library(sf)
+library(rnaturalearth)
+
+crab_sites <- cc$crabs_untransformed %>%
   sample_tibble() %>%
-  separate_wider_regex(sample,patterns=c(recovery_year="^[0-9]+","_",sample=".*"))  %>%
-  select(sample,everything()) %>%
-  unite("range",deployment_year,recovery_year,sep="--") %>%
+  distinct(hawaiian_name,island,island_group,lat,lon,.keep_all = FALSE) %>%
   mutate(
@@
     ),
-    island = case_when(
-      island == "O‘ahu" ~ str_glue("{hawaiian_name} ({shallow_deep})"),
-      TRUE ~ island
-    ),
     island = fct_reorder(island,-lat)
   ) %>%
-  count(island,range) %>%
-  pivot_wider(names_from="range",values_from="n")  %>%
-  select(island,order(colnames(.))) %>%
-  mutate(across(-island,~replace_na(.x,0)))
-print_table(arms_schedule,"arms_table","ARMS unit deployment schedule (number of units)",accuracy=1) 
-```
-
-```{r site-map, fig.cap='Sampling map (Papahānaumokuākea Marine National Monument shaded in light blue)', results="asis", fig.width=9, fig.height=6, dependson="init"}
-library(sf)
-library(rnaturalearth)
-
-crab_sites <- cc$crabs_untransformed %>%
-  sample_tibble() %>%
-  distinct(hawaiian_name,island_group,lat,lon,.keep_all = FALSE) %>%
   # mutate( island = fct_reorder(island,-lat) ) %>%
@@
   geom_sf(data=pmnm,fill="#6ed8f6",alpha=0.1) +
-  geom_sf(data=crab_sites %>% st_transform(transformer),aes(fill=hawaiian_name),size=3,shape=21,color="black") +
+  geom_sf(data=crab_sites %>% st_transform(transformer),aes(fill=island),size=3,shape=21,color="black") +
   scale_fill_manual(values=island_pal,name="Island") +
@@
 
+```{r arms-table,results="asis", dependson="init"}
+arms_schedule <- cc$crabs_untransformed %>%
+  sample_tibble() %>%
+  separate_wider_regex(sample,patterns=c(recovery_year="^[0-9]+","_",sample=".*"))  %>%
+  select(sample,everything()) %>%
+  unite("range",deployment_year,recovery_year,sep="--") %>%
+  mutate(
+    island = case_when(
+      str_remove_all(hawaiian_name,'[^a-zA-Z ]') != island ~ as.character(str_glue("{hawaiian_name} ({island})")),
+      TRUE ~ as.character(hawaiian_name)
+    ),
+    island = case_when(
+      island == "O‘ahu" ~ str_glue("{hawaiian_name} ({shallow_deep})"),
+      TRUE ~ island
+    ),
+    island = fct_reorder(island,-lat)
+  ) %>%
+  count(island,range) %>%
+  pivot_wider(names_from="range",values_from="n")  %>%
+  select(island,order(colnames(.))) %>%
+  mutate(across(-island,~replace_na(.x,0)))
+print_table(arms_schedule,"arms_table","ARMS unit deployment schedule (number of units)",accuracy=1) 
+```
+
 ARMS units were retrieved by encapsulating them in-situ in a mesh-lined crate (100 μm pore size) to prevent escape of motile organisms.
@@
 Seawater surrounding disassembled units was sieved through 2 mm, 500 μm, and 100 μm mesh-size geologic sieves [see @lerayDNABarcodingMetabarcoding2015 for processing details], and retained macroorganisms were preserved in 95% ethanol.
-All adult brachyuran crabs ($\geq$ 5 mm) were photographed, identified to morphospecies, and preserved in 95% ethanol (Fig.
-2d).
+All adult brachyuran crabs ($\geq$ 5 mm) were photographed, identified to morphospecies, and preserved in 95% ethanol.
 Individual preserved specimens were accessioned into the invertebrate zoology collections at the Florida Museum of Natural History (catalog numbers xxx–xxx).
@@
 
-To examine patterns of crab diversity across the islands, we calculated alpha diversity statistics for ARMS units across the archipelago and examined how those values varied between the MHI and NWHI as well as the effects of environmental factors.
-We calculated species richness and Simpson diversity using the `estimate_richness` function from phyloseq (Simpson = $1-\sum_{i=1}^{R}p_i^2$, where R is the total number of species, and $\mathrm{p}_i$ is the proportional abundance of each species).
-Alpha diversity was compared between MHI and NWHI using two-sample t-tests, both with and without the inclusion of mesophotic ARMS units.
-To explore the effects of environmental factors on alpha diversity across the islands, we used a linear mixed-effect model, with Simpson diversity or richness as response variables, the environmental variables listed below as fixed effects, and island nested within region (MHI vs NWHI) as random effects.
+To examine patterns of crab diversity across the islands, we used `estimate_richness` to calculate species richness and Simpson (Simpson = $1-\sum_{i=1}^{R}p_i^2$, where R is the total number of species, and $\mathrm{p}_i$ is the proportional abundance of each species).
+Diversity was compared between MHI and NWHI using two-sample t-tests, both with and without the inclusion of mesophotic samples.
+<!--To explore environmental influences on alpha diversity, we used a linear mixed-effect model with Simpson or richness as response variables, the environmental variables listed below as fixed effects, and island nested within island group (MHI vs NWHI) as random effects.
 Models were generated using the `lmer` function from the `lme4` package [@batesFittingLinearMixedEffects2015].
+-->
 
-We used the distance-based redundancy analysis method (dbRDA; a form of multivariate multiple regression) implemented in the `dbrda` function from vegan to examine the relative importance of environmental variables on crab community variation across the Hawaiian Islands.
-We first assessed the importance of each environmental variable individually using marginal tests.
-Then, a stepwise forward selection procedure (using the function `ordiR2step`) was used to find the suite of environmental variables that best explained the dissimilarity among sites with regard to crab community composition.
-Colinearity between traits was assessed by calculating variance inflation factors (VIF) using the function `vif.cca` from complete and reduced models and discarding variables where VIF > 10.
-Significance levels were assessed with the function `anova` using permutation tests with `r num(PERM,1)` permutations.
-All continuous environmental variables used for dbRDA were scaled to unit variance.
-We assessed community variation among groups using the `betadisper` and `adonis2` functions from vegan.
-These functions implement analysis of multivariate homogeneity of group dispersions (PERMDISP) and multivariate permutational analysis of variance (PERMANOVA), respectively.
-The PERMDISP method is used to assess variability within groups and PERMANOVA is used to assess differences among groups.
-We assessed significance with permutational tests, using `r num(PERM,1)` permutations.
-Crab community variation among sampling groups (shallow MHI vs NWHI and O&#x02BB;ahu 30 m, 60 m, 90 m) was further examined and visualized using canonical analysis of principal coordinates (CAP) based on discriminant analysis [@andersonCanonicalAnalysisPrincipal2003] as implemented in the `CAPdiscrim` function in the `BiodiversityR` package [@kindtTreeDiversityAnalysis2005].
+We used distance-based redundancy analysis (dbRDA; a form of multivariate multiple regression) with `dbrda` to examine the relative contribution of environmental variables to crab community variation across the islands.
+We first used a stepwise forward selection procedure (`ordiR2step`) to find the variables that best explained community dissimilarity.
+Then, we sequentially removed variables with the largest variance inflation factors (VIF; calculated using `vif.cca`) until all values were <3, indicating variables were sufficiently uncorrelated.
+Model significance was assessed using permutation tests with `r num(PERM,1)` permutations (function `anova`).
+Continuous variables were scaled to unit variance.
+We assessed variation among groups using the `betadisper` and `adonis2`.
+These functions implement analysis of multivariate homogeneity of group dispersions (PERMDISP) and multivariate permutational analysis of variance (PERMANOVA) respectively.
+PERMDISP is used to assess variability within groups and PERMANOVA is used to assess differences among groups.
+Significance was assessed with permutation tests in both cases.
+Variation among sampling groups (shallow MHI, shallow NWHI, O&#x02BB;ahu 30 m, 60 m, 90 m) was further examined and visualized using canonical analysis of principal coordinates (CAP) based on discriminant analysis [@andersonCanonicalAnalysisPrincipal2003] as implemented in the `CAPdiscrim` function in `BiodiversityR` [@kindtTreeDiversityAnalysis2005].
 
@@
 
-Potential environmental factors affecting the community structure of brachyuran crabs on shallow reefs were assembled from published literature [@selkoeDNACoralReef2016], NOAA's National Coral Reef Monitoring Program (NCRMP), and the Pacific Islands Ocean Observing System (PacIOOS) ([@tbl:env_table]).
-Environmental factors were assembled only for shallow sites sampled during RAMP cruises.
+Potential environmental drivers of shallow crab community structure were assembled from published literature [@selkoeDNACoralReef2016], NOAA's National Coral Reef Monitoring Program (NCRMP), and the Pacific Islands Ocean Observing System (PacIOOS) ([@tbl:env_table]).
+Due to logistical limitations, environmental factors are only available for shallow samples.
 
 ```{r env-table,results="asis", dependson="init"}
-print_table(here("data","tables","environmental_variables.csv"),"env_table","Potential environmental drivers of crab community structure, assembled from PacIOOS and @selkoeDNACoralReef2016",hide="variable")
+print_table(model_terms,"env_table","Potential environmental drivers of crab community structure, assembled from PacIOOS and @selkoeDNACoralReef2016",hide="variable")
 ```
@@
 The `r nsamples(cc$crabs_untransformed)` ARMS units deployed across the Hawaiian Islands (in `r format_range(cc$crab_data$depth)` m depth) yielded a total of `r num(sum(sample_sums(cc$crabs_untransformed)),1)` adult crabs.
-Sampling units contained `r num(all_summary$abundance_mean,0.1)` $\pm$ `r num(all_summary$abundance_sd,0.1)` (mean $\pm$ SD) individuals per unit, and abundance ranged from `r format_range(all_summary$abundance_range)` individuals.
+Units contained `r num(all_summary$abundance_mean,0.1)` $\pm$ `r num(all_summary$abundance_sd,0.1)` (mean $\pm$ SD) individuals, and abundance ranged from `r format_range(all_summary$abundance_range)`.
 Crabs were identified to `r sum(unlist(all_summary$types_breakdown))` distinct types (including `r all_summary$types_breakdown$species` species and `r all_summary$types_breakdown$morphospecies` morphospecies) in `r all_summary$taxonomy_breakdown$genera` genera and `r all_summary$taxonomy_breakdown$families` families, which accounted for ~36% of the known crab diversity reported from Hawaiian waters [@castroCatalogAnomuranBrachyuran2011].
 The four most abundant species were: `r with(all_summary,top_list(top,top_n,top_units,fig="fig_crabs"))`.
-`r str_to_sentence(as.english(all_summary$singleton_count))` species occurred in a single unit across all sites (singletons), accounting for ~`r num(all_summary$singleton_count/ntaxa(cc$crabs_untransformed)*100,1)`% of total recovered crab species.
+`r str_to_sentence(english(all_summary$singleton_count))` species occurred in a single unit across all sites (singletons), accounting for ~`r num(all_summary$singleton_count/ntaxa(cc$crabs_untransformed)*100,1)`% of total recovered crab species.
 
@@
 
-The `r nsamples(cc$crabs_deep)` units deployed at mesophotic depths $\geq$ 30 m on O&#x02BB;ahu yielded both fewer species and fewer individuals compared to the complete dataset, with `r num(sum(sample_sums(cc$crabs_deep)),1)` crabs overall, a total of `r sum(unlist(deep_summary$types_breakdown))` distinct types (`r deep_summary$types_breakdown$species` species and `r deep_summary$types_breakdown$morphospecies` morphospecies), `r num(deep_summary$abundance_mean,0.1)` $\pm$ `r num(deep_summary$abundance_sd,0.1)` individuals per unit, and abundance ranging from `r format_range(deep_summary$abundance_range)` crabs.
+The `r nsamples(cc$crabs_deep)` units deployed at mesophotic depths $\geq$ 30 m on O&#x02BB;ahu yielded fewer species and individuals compared to the complete dataset, with `r num(sum(sample_sums(cc$crabs_deep)),1)` crabs overall, `r sum(unlist(deep_summary$types_breakdown))` distinct types (`r deep_summary$types_breakdown$species` species and `r deep_summary$types_breakdown$morphospecies` morphospecies), `r num(deep_summary$abundance_mean,0.1)` $\pm$ `r num(deep_summary$abundance_sd,0.1)` individuals per unit, and abundance ranging from `r format_range(deep_summary$abundance_range)`.
 The four most abundant species in the deep ARMS did not overlap with those from the complete dataset: `r with(deep_summary,top_list(top,top_n,top_units))`.
@@
 For the complete dataset, there was no significant difference in observed species richness between the MHI and NWHI `r t_test(all_alpha$t_r_island)`, but Simpson diversity was significantly different, although mean values were similar with `r num(all_alpha$richness_main$richness,0.1)` $\pm$ `r num(all_alpha$richness_main$richness_sd,0.1)` in the MHI and `r num(all_alpha$richness_nwhi$richness,0.1)` $\pm$ `r num(all_alpha$richness_nwhi$richness_sd,0.1)` in the NWHI ([@fig:fig_alpha[]{}A]; `r t_test(all_alpha$t_s_island,plain=TRUE)`).
+When only shallow arms \< 30 m were considered, neither comparison was significant (richness: `r t_test(shallow_alpha$t_r_island,plain=TRUE)`; Simpson: `r t_test(shallow_alpha$t_s_island,plain=TRUE)`).
 Simpson diversity was significantly different between shallow and deep samples ([@fig:fig_alpha[]{}B]; `r t_test(all_alpha$t_s_depth,plain=TRUE)`) but richness was not (`r t_test(all_alpha$t_r_depth,plain=TRUE)`).
-When only shallow arms \< 30 m were considered, neither comparison was significant (richness: `r t_test(shallow_alpha$t_r_island,plain=TRUE)`; Simpson: `r t_test(shallow_alpha$t_s_island,plain=TRUE)`).
 
@@
 ```{r rda-analysis, results="hide", dependson="init"}
+library(formula.tools)
+
+reduce_model <- function(m,vif_limit=3) {
+  vifs <- vif.cca(m)
+  if (any(vifs > vif_limit)) {
+    v <- vifs[-which(vifs == max(vifs))[1]]
+    m_lhs <- capture.output(print(lhs(m$call$formula)))
+    f <- str_glue("{m_lhs} ~ {str_c(names(v),collapse=' + ')}") %>%
+      as.formula()
+    return(reduce_model(dbrda(f,data=cc$crab_data_shallow,distance = "bray")))
+  } else {
+    return(m)
+  }
+}
+
 db_upr <- dbrda(cc$crab_otus_shallow ~ depth + chl + sst + slope + coral_cover + closest_island + larval_connectivity + human_impact,data=cc$crab_data_shallow,distance="bray")
@@
 
-# vif.cca(db_upr)
-vifs <- vif.cca(db_modl)
+# get rid of variables with VIF > 3
+db_modl <- reduce_model(db_modl,vif_limit=3)
 
@@
 # see https://github.com/vegandevs/vegan/issues/254 for information on the transformation below
-sppscores(db_modl) <- sqrt(decostand(cc$crab_otus_shallow, "total")/2)
+# sppscores(db_modl) <- sqrt(decostand(cc$crab_otus_shallow, "total")/2) 
+# 
+# # this was to grab the 4 most abundant species overall
+# top_species <- names(head(rev(sort(taxa_sums(cc$crabs_shallow))),n=4))
+# 
+# # extract the species scores for the species we want
+# species_scores <- scores(db_modl,display="species") %>%
+#   as_tibble(rownames = "species") %>%
+#   filter(species %in% top_species)
 
-# this was to grab the 4 most abundant species overall
-top_species <- names(head(rev(sort(taxa_sums(cc$crabs_shallow))),n=4))
+modl_perm <- anova(db_modl,by="margin") %>%
+  broom::tidy() %>%
+  arrange(desc(statistic))
 
-# extract the species scores for the species we want
-species_scores <- scores(db_modl,display="species") %>%
-  as_tibble(rownames = "species") %>%
-  filter(species %in% top_species)
-
-model_terms <- read_csv(here("data","tables","environmental_variables.csv")) %>%
-  select(variable,driver=`Environmental Driver`)
-
-rda_table <- anova(db_modl,by="term") %>%
-  broom::tidy() %>%
-  inner_join(model_terms,by=c("term" = "variable")) %>%
+rda_table <- modl_perm %>%
+  inner_join(model_terms %>% select(variable,driver=`Environmental Driver`),by=c("term" = "variable")) %>%
   select(-term) %>%
@@
 
-abs_r2 <- diff(c(0,db_modl$anova$R2.adj))
+# TODO: check this, this might be wrong!
+nterm <- length(attr(db_modl$terms,"term.label"))
+
+drivers <- rda_table %>%
+  slice(1:2) %>%
+  select(term=Term,stat=`F-statistic`,pval=`p-value`)  %>%
+  mutate(term=str_to_lower(term))
 ```
@@
 The best dbRDA model (`r permanova(db_modl,stat="F")`) explained `r pc(RsquareAdj(db_modl)$adj.r.squared,0.01)` (adjusted $\mathrm{R}^2$) of the variability in crab communities across the archipelago ([@fig:fig_rda]).
-Seven of the 10 environmental variables tested were significant ([@tbl:rda_table]).
-The drivers contributing most to the explained variation were chlorophyll-A ($\mathrm{R}^2$ = `r num(abs_r2[1],0.01)`) and human impacts ($\mathrm{R}^2$ = `r num(abs_r2[2],0.01)`).
+`r str_to_sentence(english(nterm))` of the 10 environmental variables tested were significant.
+The drivers contributing most to explained variation were `r drivers$term[1]` (F = `r num(drivers$stat[1],0.01)`) and `r drivers$term[2]` (F = `r num(drivers$stat[2],0.01)`) ([@tbl:rda_table]).
 
 ```{r rda, fig.cap="Distance-based redundancy analysis (dbRDA) plot showing significant environmental drivers of crab community variation across the Hawaiian Islands.", results="asis", fig.width=9, fig.height=6, dependson="rda-analysis"}
-rda_plotz <- plot_ord(cc$crabs_shallow,db_modl,scale=2,color="island_group") + 
-  stat_ellipse(geom="polygon",aes(group=island_group,fill=island_group),color=NA,alpha=0.1) +
+termz <- model_terms %>%
+  select(variable,display=`Environmental Driver`)
+
+# rda_plotz <- plot_ord(cc$crabs_shallow,db_modl,scale=2,color="island_group",labelsize=3,term_map=termz) + 
+rda_plotz <- plot_ord(cc$crabs_shallow,db_modl,scale=2,color="island_group",labelsize=3,term_map=termz) + 
+  stat_ellipse(geom="polygon",aes(group=island_group,fill=island_group),color=NA,alpha=0.1,show.legend = FALSE) +
   geom_point(aes(fill=island_group),shape=21,color="black",size=3) +
   plotz_theme() + 
-  scale_fill_manual(values=island_group_pal)
+  scale_fill_manual(name="Island group",values=island_group_pal,labels = c("Main","Northwest")) +
+  guides(color = "none")
 
@@
 However, community dispersion was not different among depth zones $\geq$ 30 m (`r permdisp(bd_deep)`).
-Although crab communities varied significantly among both depth zones and island groups when examining the complete dataset (`r permanova(anova_both,0.01)`), depth explained `r num(effect_ratio,0.1)` times variability than island group (based on model $\mathrm{R}^2$).
+Although crab communities varied significantly among both depth zones and island groups when examining the complete dataset (`r permanova(anova_both,0.01)`), depth explained `r num(effect_ratio,0.1)` times more variability than island group (based on model $\mathrm{R}^2$).
 In the CAP analysis, ARMS units were correctly assigned to sampling groups `r pc(cap_depth$percent/100,0.1)` of the time.
 The first two axes explained `r pc(cap_var$x,0.1)` and `r pc(cap_var$y,0.1)` of the variation, respectively.
-The mesophotic samples from O&#x02BB;ahu grouped separately from the shallow samples collected across the entire ~2,500 km of the Hawaiian Islands ([@fig:fig_cap]).
-In addition, there was as much variability across the 30--90 m samples as across all shallow samples, with the 30 m samples spread out along the LD2 (y) axis and the 60 m and 90 m samples forming a tight group ([@fig:fig_cap]).
+The $\geq$ 30 m samples from O&#x02BB;ahu grouped separately from the shallow samples collected across the entire ~2,500 km of the Hawaiian Islands ([@fig:fig_cap]).
+In addition, there was as much variability across the 30--90 m samples as across all shallow samples, with the 30 m samples spread out primarily along the LD2 (y) axis and the 60 m and 90 m samples forming a tight group ([@fig:fig_cap]).
 
@@
 ```
-
-### Effects of environmental variables {-}
 
@@
 ```{r rda-table,results="asis",dependson="rda-analysis"}
-print_table(rda_table,"rda_table","Significant terms in the dbRDA analysis of crab community variation across the Hawaiian Archipelago",hdr_case="none",accuracy=0.001) 
+print_table(rda_table,"rda_table","Significant (marginal) terms in the dbRDA analysis of crab community variation across the Hawaiian Archipelago",hdr_case="none",accuracy=0.001) 
 ```

### 2023-09-12 crabs.R

@@ -30,7 +30,7 @@ run_things <- FALSE
 
 
 # plot an ordination with various options
-plot_ord <- function(ps,ord,arrows=TRUE,sig=FALSE,by="term",perm=999,alpha=0.05,scale=1,labelsize=5,pointsize=3,...) {
+plot_ord <- function(ps,ord,arrows=TRUE,sig=FALSE,by="term",perm=999,alpha=0.05,scale=1,labelsize=5,pointsize=3,term_map=NULL,...) {
   
   # map classnames to axis names
   axes <- c(
@@ -79,6 +79,19 @@ plot_ord <- function(ps,ord,arrows=TRUE,sig=FALSE,by="term",perm=999,alpha=0.05,
         filter(str_detect(labels,pattern))
     }
     
+    arrowdf <- arrowdf %>%
+      mutate(
+        hjust = if_else(.data[[str_c(axis,"1")]] < 0,1,0),
+        xnudge = if_else(.data[[str_c(axis,"1")]] < 0,-0.02,0.02)
+      )
+    
+    if (!is.null(term_map)) {
+      arrowdf <- arrowdf %>%
+        inner_join(term_map,by=c("labels" = "variable")) %>%
+        select(-labels) %>%
+        rename(labels=display)
+    }
+    
     # Define the arrow aesthetic mappings
     # map for arrowheads
     arrow_map <- aes(xend = .data[[str_c(axis,"1")]], # this is where we use our mapped axis name
@@ -88,11 +101,12 @@ plot_ord <- function(ps,ord,arrows=TRUE,sig=FALSE,by="term",perm=999,alpha=0.05,
                      shape = NULL, 
                      color = NULL)
     # map for arrow labels
-    label_map <- aes(x = 1.3 * .data[[str_c(axis,"1")]], 
-                     y = 1.3 * .data[[str_c(axis,"2")]], 
+    label_map <- aes(x = .data[[str_c(axis,"1")]]+xnudge, 
+                     y = .data[[str_c(axis,"2")]], 
                      shape = NULL, 
                      color = NULL, 
-                     label = labels)
+                     label = labels,
+                     hjust = hjust)
     # make a little arrowhead
     arrowhead = arrow(length = unit(0.02, "npc"))
     # add the arrows to the plot
@@ -105,12 +119,14 @@ plot_ord <- function(ps,ord,arrows=TRUE,sig=FALSE,by="term",perm=999,alpha=0.05,
         arrow = arrowhead
       ) + 
       # geom_text_repel(
-      geom_text(
[…]

### 2023-10-10 manuscript.Rmd
arbitrary commit to track progress

@@
   - \usepackage{bm}
+  - \usepackage{upgreek}
 ---
-
+<!--word count: Proceedings B publishes articles up to 12 pages, including figures and tables, with no more than 10 000 words.
+The word count includes the title page, author names, abstract, figure legends, tables and the reference list.
+As a rough guide, 10 000 words = 12 pages without additional figures; 850 words = 1 page and 2-4 small figures = 1 page.
+For example, an acceptable paper could be made up of 9 000 words and 4 small figures or 8500 words, 2 large figures and 4 small figures.-->
 ### Keywords {-}
@@
 num <- function(x,a=0.01) scales::label_comma(accuracy=a)(x)
+
+w_test <- function(test,plain=FALSE) {
+  if (plain) {
+    str_glue("*W* = {num(test$statistic,0.01)}, {pval(test$p.value)}")
+  } else {
+    str_glue("(Wilcoxon rank sum test: *W* = {num(test$statistic,0.01)}, {pval(test$p.value)})")
+  }
+}
+
+k_test <- function(test,plain=FALSE) {
+  if (plain) {
+    str_glue("*K-W $\\upchi^2$* = {num(test$statistic,0.01)}, DF = {num(test$parameter['df'],0.01)}, {pval(test$p.value)}")
+  } else {
+    str_glue("Kruskal-Wallis rank sum test: *K-W $\\upchi^2$* = {num(test$statistic,0.01)}, DF = {num(test$parameter['df'],0.01)}, {pval(test$p.value)}")
+  }
+}
 
@@
 # palette for island/group colors
+# island_pal <- c(
+#   "Hawaii Island" = "#8a3ffc",
+#   "Hawai‘i Island" = "#8a3ffc",
+#   "Maui" = "#007d79",
+#   "Oahu" = "#ff7eb6",
+#   "O‘ahu" = "#ff7eb6",
+#   "Kauai" = "#fa4d56",
+#   "Kaua‘i" = "#fa4d56",
+# 
+#   "French Frigate Shoals" = "#fff1f1",
+#   "Lalo" = "#fff1f1",
+#   "Lalo (French Frigate Shoals)" = "#fff1f1",
+#   "Lisianski" = "#6fdc8c",
+#   "Kapou" = "#6fdc8c",
+#   "Kapou (Lisianski)" = "#6fdc8c",
+#   "Pearl and Hermes Reef" = "#4589ff",
+#   "Manawai" = "#4589ff",
+#   "Manawai (Pearl and Hermes Reef)" = "#4589ff",
+#   "Kure Atoll" = "#d12771",
+#   "Hōlanikū" = "#d12771",
+#   "Hōlanikū (Kure Atoll)" = "#d12771"
+# )
 island_pal <- c(
-  "Hawaii Island" = "#8a3ffc",
-  "Hawai‘i Island" = "#8a3ffc",
-  "Maui" = "#007d79",
-  "Oahu" = "#ff7eb6",
-  "O‘ahu" = "#ff7eb6",
-  "Kauai" = "#fa4d56",
-  "Kaua‘i" = "#fa4d56",
+  "Hawaii Island" = "#5E4FA2",
+  "Hawai‘i Island" = "#5E4FA2",
+  "Maui" = "#3288BD",
+  "Oahu" = "#66C2A5",
+  "O‘ahu" = "#66C2A5",
+  "Kauai" = "#ABDDA4",
+  "Kaua‘i" = "#ABDDA4",
 
-  "French Frigate Shoals" = "#fff1f1",
-  "Lalo" = "#fff1f1",
-  "Lalo (French Frigate Shoals)" = "#fff1f1",
-  "Lisianski" = "#6fdc8c",
-  "Kapou" = "#6fdc8c",
-  "Kapou (Lisianski)" = "#6fdc8c",
-  "Pearl and Hermes Reef" = "#4589ff",
-  "Manawai" = "#4589ff",
-  "Manawai (Pearl and Hermes Reef)" = "#4589ff",
-  "Kure Atoll" = "#d12771",
-  "Hōlanikū" = "#d12771",
-  "Hōlanikū (Kure Atoll)" = "#d12771"
+  "French Frigate Shoals" = "#FDAE61",
+  "Lalo" = "#FDAE61",
+  "Lalo (French Frigate Shoals)" = "#FDAE61",
+  "Lisianski" = "#F46D43",
+  "Kapou" = "#F46D43",
+  "Kapou (Lisianski)" = "#F46D43",
+  "Pearl and Hermes Reef" = "#D53E4F",
+  "Manawai" = "#D53E4F",
+  "Manawai (Pearl and Hermes Reef)" = "#D53E4F",
+  "Kure Atoll" = "#9E0142",
+  "Hōlanikū" = "#9E0142",
+  "Hōlanikū (Kure Atoll)" = "#9E0142"
 )
@@
   "deep" = "#04738d",
-  "Deep" = "#04738d"
+  "Deep" = "#04738d",
+  "O‘ahu (shallow)" = "#e5ab52",
+  "O‘ahu 30 m" = "#d87939",
+  "O‘ahu 30m" = "#d87939",
+  "O‘ahu 60 m" = "#04c3c8",
+  "O‘ahu 60m" = "#04c3c8",
+  "O‘ahu 90 m" = "#04738d",
+  "O‘ahu 90m" = "#04738d",
+  "O‘ahu deep" = "#04738d",
+  "O‘ahu Deep" = "#04738d"
 )
@@
 
-# 1. Introduction {-}
+# Introduction {-}
+Natural communities are structured by a suite of biotic and abiotic factors including geography, habitat complexity, environment, productivity, and species interactions (among many others) [@begonEcologyIndividualsEcosystems2021] and such complexity makes direct comparisons over large geographic scales challenging.
+Marine communities in particular experience unique influences and present research challenges not found in terrestrial habitats [@steeleComparisonTerrestrialMarine1985; @kinlanPropaguleDispersalMarine2003; @raffaelliMarineTerrestrialEcologists2005; @grosbergBiodiversityWaterLand2012; @knappDriversBiodiversityChange2017].
+Much of our understanding of biotic zonation and community structure in the ocean comes from classic studies in the intertidal (some of the best-studied marine habitats) [e.g., @connellInfluenceInterspecificCompetition1961; @connellEffectsCompetitionPredation1961; @paineIntertidalCommunityStructure1974].
+Such studies point to a combination of physical and biological factors, including physiological tolerance [@dennyPhysicalProcessesThat2001; @someroThermalPhysiologyVertical2002] and species interactions [@brunoHabitatModificationFacilitation2001; @mengeRockyIntertidalCommunities2001], that contribute to the structure of intertidal communities.
+Although there are exceptions [e.g., @woottonIndirectEffectsPrey1992; @bertnessTestingRelativeContribution1999], a generalization has emerged for the intertidal that lower limits of species distributions are set by ecological interactions whereas upper limits are determined by abiotic factors.
 
-Natural communities are structured by a suite of biotic and abiotic factors including geography, habitat complexity, environment, productivity, and species interactions (among many others) [@begonEcologyIndividualsEcosystems2021].
-Such complexity makes direct comparisons of processes structuring communities over large geographic scales challenging.
-Marine communities experience unique influences and present particular research challenges, however studies have largely concluded that factors structuring marine communities differ from those in terrestrial habitats [@steeleComparisonTerrestrialMarine1985; @kinlanPropaguleDispersalMarine2003; @raffaelliMarineTerrestrialEcologists2005; @grosbergBiodiversityWaterLand2012; @knappDriversBiodiversityChange2017].
-Among the best studied of aquatic ecosystems is the littoral zone lying between the low-tide line and the upper limit of aquatic species on the shore where the habitat is exposed to air during low tide events that provide easy and consistent research access for humans.
-Classical studies of vertical zonation across the intertidal have contributed much of our understanding of the mechanisms of biotic zonation and community structure in marine environments [@raffaelliIntertidalEcology2012; @birdTideWaveMarks2013].
-Such studies point to an inseparable combination of physical and biological factors, including physiological tolerance [@someroThermalPhysiologyVertical2002] and species interactions [@brunoHabitatModificationFacilitation2001; @mengeRockyIntertidalCommunities2001] that together determine the structure of intertidal communities [@dennyPhysicalProcessesThat2001].
-Although there are exceptions [e.g., @woottonIndirectEffectsPrey1992; @bertnessTestingRelativeContribution1999], the generalization has emerged that biotic factors typically determine the lower limit whereas physical factors set the upper limit of the depth distribution of intertidal communities [@connellInfluenceInterspecificCompetition1961; @connellEffectsCompetitionPredation1961; @paineIntertidalCommunityStructure1974].
+Despite conclusions based on intertidal studies, similar generalizations have not emerged to describe coral reef communities [@moraEcologyFishesCoral2015; @sheppardBiologyCoralReefs2017], with existing hypotheses ranging from the neutral theory of biodiversity [@hubbellUnifiedNeutralTheory2011] and lottery recruitment [@saleStockRecruitRelationshipsRegional1982] to competitive networks [@bussCompetitiveNetworksNontransitive1979] and fine-scale resource partitioning driven by predators and herbivores [@abramsRecruitmentLotteriesCoexistence1984; @hixonPredationPreyRefuges1993].
+Studies that have investigated these topics on coral reefs have traditionally focused on conspicuous taxa such as fishes and corals [@cite], yet these groups constitute a minority of total reef biodiversity [@cite].
+The majority of coral reef biodiversity is cryptic, inhabiting interstitial spaces, and little is known about factors contributing to     Zooxanthellate corals are the ecosystem engineers of reef habitats but are restricted to specific areas characterized by a relatively narrow range of chemical and physical parameters<!-- (i.e., tropical and subtropical warm waters, oligotrophic conditions, 32-40 PSU salinity, and sufficient incident light, which decreases rapidly below ~30 m depth)-->.
+Shallow coral reef ecosystems cover less than 0.2% of the seafloor and in such a small area they hold the 32-38% of all marine species [@fisherSpeciesRichnessCoral2015].
+Yet the <!--drivers of high biodiversity and the -->mechanisms by which communities are structured on coral reefs remain a subject of debate [@carrBiodiversityPopulationRegulation2002; @connollyIndoPacificBiodiversityCoral2003; @almanyDoesIncreasedHabitat2004; @bellwoodEnvironmentalGeometricConstraints2005, 2006; @dornelasCoralReefDiversity2006; @grahamImportanceStructuralComplexity2013; @komyakovaRelativeImportanceCoral2013].
+*Add some text about community structuring across space, rather than just at smaller scales*.
 
-No such generalization has emerged for the processes structuring coral reef communities [@moraEcologyFishesCoral2015; @sheppardBiologyCoralReefs2017], with hypotheses ranging from the Neutral Theory of Biodiversity [@hubbellUnifiedNeutralTheory2011] and Lottery Recruitment [@saleStockRecruitRelationshipsRegional1982] to competitive networks [@bussCompetitiveNetworksNontransitive1979] and fine-scale resource partitioning driven by predators and herbivores structuring coral reef communities [@abramsRecruitmentLotteriesCoexistence1984; @hixonPredationPreyRefuges1993].
-Zooxanthellate corals are the ecosystem engineers of reef habitats but are restricted to specific areas characterized by a relatively narrow range of chemical and physical environmental variables (i.e., tropical and subtropical warm waters, oligotrophic conditions, 32-40 PSU salinity, and sufficient incident light, which decreases rapidly below ~30 m depth).
-Shallow coral reef ecosystems cover less than 0.2% of the seafloor and in such a small area they hold the 32-38% of all marine species [@fisherSpeciesRichnessCoral2015].
-Yet the drivers of high biodiversity and the mechanisms by which diversity is maintained on coral reefs remains a subject of considerable debate [@carrBiodiversityPopulationRegulation2002; @connollyIndoPacificBiodiversityCoral2003; @almanyDoesIncreasedHabitat2004; @bellwoodEnvironmentalGeometricConstraints2005, 2006; @dornelasCoralReefDiversity2006; @grahamImportanceStructuralComplexity2013; @komyakovaRelativeImportanceCoral2013].
+A<!--nother--> likely important but under-investigated aspect of coral reef community structure is depth.
+Most studies to date have focused on the upper ~30 m of the water column, yet coral reefs and associated communities are known to extend beyond 150 m [@pyleExploringDeepCoral1996; @pyleMesophoticCoralEcosystems2019], with scleractinian corals reported to 170 m [@rouzeSymbioticAssociationsDeepest2021].
+As much as two thirds of coral reef ecosystems lie in this under-explored zone [@hindersteinThemeSectionMesophotic2010].
+These habitats, known as mesophotic coral ecosystems (MCEs), are typically defined as those extending from ~30--150 m depth [@hindersteinThemeSectionMesophotic2010; @kahngCommunityEcologyMesophotic2010].
+<!--and were believed to experience dampened disturbances in comparison to shallow reefs and could therefore act as deep water refugia for shallow reef species threatened by anthropogenic impacts --> They have been hypothesized as possible refugia from disturbances on shallow reefs [@bongaertsAssessingDeepReef2010; @smithDepthRefugiumCatastrophic2014; @holsteinFertileFathomsDeep2015; @muirSpeciesIdentityDepth2017], although a number of studies have found evidence that deep reefs are unique and uniquely threatened, challenging this notion [@hobanPlumbingDepthsEnvironmental2023a; @hurleyAssessmentShallowMesophotic2016; @kaneHighLevelsMesophotic2014; @pinheiroDeepReefFishes2019; @rochaMesophoticCoralEcosystems2018].
+<!--However, recent literature describing species compositions along depth gradients reveal a more complex picture among taxa and locations because some groups may use mesophotic reefs as a refuge in some places, while others show community turnover with depth such that MCE habitats are ecologically distinct from adjacent shallow water coral reefs [@hurleyAssessmentShallowMesophotic2016; @smithCaribbeanMesophoticCoral2016; @semmlerAreMesophoticCoral2017; @bongaertsDeepReefsAre2017; @rochaMesophoticCoralEcosystems2018].
+Further, there are a range of connections depending on the taxa – large, mobile organisms like fish tend to have more shared species across depth than sedentary organisms such as corals and algae, but even among the fishes the community structure of MCEs is often distinct [@kaneHighLevelsMesophotic2014; @pyleComprehensiveInvestigationMesophotic2016 ; @asherMesophoticDepthGradients2017; @fukunagaChangesMesophoticReef2017; @colemanShiftingReefFish2018; @pinheiroDeepReefFishes2019].
+Although the majority of coral reef biodiversity consists of the algal and invertebrate cryptobenthic organisms living on and within the reef structure [@fautinOverviewMarineBiodiversity2010; @knowltonCoralReefBiodiversity2010; @fisherSpeciesRichnessCoral2015], the studies listed above focus almost exclusively on the corals and fishes which are more easily surveyed.-->
 
-Recent advances in underwater explorations have shown that light-dependent reefs extend well beyond the 30 m depth limit imposed by scuba exploration of coral reefs down to 150 m in some locations [@kahngDeepestZooxanthellateScleractinian2006].
-These mesophotic coral ecosystems (MCEs) are typically defined as those extending from ~40 to 150 m depth [@hindersteinThemeSectionMesophotic2010; @kahngCommunityEcologyMesophotic2010], and were believed to experience dampened disturbances in comparison to shallow reefs and could therefore act as deep water refugia for shallow reef species threatened by anthropogenic impacts [@bongaertsAssessingDeepReef2010; @smithDepthRefugiumCatastrophic2014; @holsteinFertileFathomsDeep2015; @muirSpeciesIdentityDepth2017].
-However, recent literature describing species compositions along depth gradients reveal a more complex picture among taxa and locations because some groups may use mesophotic reefs as a refuge in some places, while others show community turnover with depth such that MCE habitats are ecologically distinct from adjacent shallow water coral reefs [@hurleyAssessmentShallowMesophotic2016; @smithCaribbeanMesophoticCoral2016; @semmlerAreMesophoticCoral2017; @bongaertsDeepReefsAre2017; @rochaMesophoticCoralEcosystems2018].
-Further, there are a range of connections depending on the taxa – large, mobile organisms like fish tend to have more shared species across depth than sedentary organisms such as corals and algae, but even among the fishes the community structure of MCEs is often distinct [@kaneHighLevelsMesophotic2014; @pyleComprehensiveInvestigationMesophotic2016 ; @asherMesophoticDepthGradients2017; @fukunagaChangesMesophoticReef2017; @colemanShiftingReefFish2018; @pinheiroDeepReefFishes2019].
-Although the majority of coral reef biodiversity consists of the algal and invertebrate cryptobenthic organisms living on and within the reef structure [@fautinOverviewMarineBiodiversity2010; @knowltonCoralReefBiodiversity2010; @fisherSpeciesRichnessCoral2015], the studies listed above focus almost exclusively on the corals and fishes which are more easily surveyed.
-
-Studies comparing community structure either along a depth gradient or across a latitudinal gradients have been criticized because they may confound multiple aspects of physical environments with habitat area or habitat complexity that covary across the measured gradient in such surveys [e.g., @bellwoodRegionalScaleAssemblyRules2001; @grahamEffectsMarineReserve2003; @connollyCommunityStructureCorals2005; @ewersConfoundingFactorsDetection2006; @fletcherRoleHabitatArea2007a; ).
-Here, we use standardized habitat modules, autonomous reef monitoring structures or ARMS [@zimmermanArtificialReefMatrix2004; @brainardAutonomousReefMonitoring2009], to sample brachyuran crab communities collected from across the Hawaiian Archipelago and test which suite of environmental or anthropogenic drivers are most important in structuring coral reef infauna communities.
-The Hawaiian Archipelago is made up of both high islands to atolls stretching across 2500 km in the Pacific Ocean, and is typically regarded as two regions: the main Hawaiian Islands (MHI) are the 8 high volcanic islands, and the Northwestern Hawaiian Islands (NWHI) are the low islands and atolls north and west of Kauaʻi.
-We compared shallow reef crab communities gathered from across the archipelago based on the nested groupings of island versus region (MHI vs NWHI) spanning roughly 9 degrees of latitude and 24 degrees of longitude that encompass a wide range of physical and anthropogenic gradients [@selkoeMapHumanImpacts2009, 2016).
+Studies examining community structure along depth or latitudinal gradients have been criticized for confounding multiple potentially covarying aspects of the environment (such as habitat area or complexity) across the measured gradients [e.g., @bellwoodRegionalScaleAssemblyRules2001; @grahamEffectsMarineReserve2003; @connollyCommunityStructureCorals2005; @ewersConfoundingFactorsDetection2006; @fletcherRoleHabitatArea2007a].
+Here, we employed autonomous reef monitoring structures (ARMS) [@zimmermanArtificialReefMatrix2004; @brainardAutonomousReefMonitoring2009], collection devices designed to mimic reef structural complexity and enable sampling of cryptic fauna in a standardized way.
+Using ARMS, we sampled brachyuran crab communities across the Hawaiian Archipelago to test the effects of environmental and anthropogenic drivers on the structure of cryptic reef communities.
+The Hawaiian Archipelago comprises high volcanic islands and atolls stretching across ~2,500 km in the North Pacific Ocean, and is typically regarded as two regions: the main Hawaiian Islands (MHI) are the eight inhabited high islands, and the Northwestern Hawaiian Islands (NWHI) are the uninhabited atolls and banks northwest of Kaua&#x02BB;i.
+The islands are subject to anthropogenic and physical gradients across their span [@selkoeMapHumanImpacts2009], which we hypothesize to affect the variation in structure of sampled communities.
+We compared crab communities among islands and regions (MHI vs.
+NWHI) and examined the effects of a suite of possible drivers on community variation.
+Given the possibility for considerable separation between deep and shallow reefs [e.g., @hobanPlumbingDepthsEnvironmental2023a; @hurleyAssessmentShallowMesophotic2016; @rochaMesophoticCoralEcosystems2018], we also compared the relative effects of depth (shallow vs.
+mesophotic) and island region.
 Given the number and magnitude of differences between these regions, we hypothesized that species diversity would be different between regions (MHI vs NWHI).
@@
 
-# 2. Materials and Methods {-}
+The Hawaiian Archipelago is a volcanic island arc made up of high islands and atolls spanning ~9 degrees of latitude and ~2,500 km.
+It is typically regarded as two regions: the inhabited main Hawaiian Islands (MHI), which include the eight islands from Hawai&#x02BB;i Island to Kauai&#x02BB;i, and the largely uninhabited Northwestern Hawaiian Islands (NWHI), which constitute the 10 low islands and atolls to the northwest of Kaua&#x02BB;i.
+We compared shallow reef crab communities across the 2500 km archipelago and  we compared reef crab assemblages from the shallow reefs to the reef crab assemblages across a mesophotic depth gradient off of Kewalo, Oʻahu.
+By controlling for habitat area and complexity and by incorporating sites across various environmental and human impact gradients, we sought to better understand whether latitude, the environment, or depth is more influential in structuring reef crab communities.
+
+# Materials and Methods {-}
 
@@
 Brachyuran crab assemblages were sampled from benthic coral reef habitat in the Hawaiian Archipelago using Autonomous Reef Monitoring Structures (ARMS).
-ARMS are standardized collection devices designed to mimic coral reef structural complexity to attract settlement of reef cryptofauna [@brainardAutonomousReefMonitoring2009; @zimmermanArtificialReefMatrix2004].
+ARMS are standardized collection devices designed to mimic coral reef structural complexity to attract settlement of reef cryptofauna [@zimmermanArtificialReefMatrix2004; @brainardAutonomousReefMonitoring2009].
 They are constructed from 10 gray type 1 PVC plates (23 cm $\times$ 23 cm) stacked in an alternating series of open and semi-enclosed layers ([@fig:fig_arms]).
@@
 The sites in the NWHI were at Lalo (French Frigate Shoals), Kapou (Lisianski), Manawai (Pearl & Hermes), and Hōlanikū (Kure Atoll) ([@fig:fig_map]).
--->To explore the effects of depth on crab community structure, we also present data from a previous study by Hurley et al.
+-->To explore the effects of depth, we also present data from an earlier study by Hurley et al.
 [-@hurleyAssessmentShallowMesophotic2016], which included a total of 27 units deployed off O&#x02BB;ahu from 2009–2012 in 12–90 m depth; nine units were deployed at 12 m, and six each at 30 m, 60 m, and 90 m ([@tbl:arms_table]).
-Shallow units were deployed according to standard NOAA methodology, whereas units $\geq$ 30 m were deployed and recovered using technical mixed-gas closed-circuit rebreather diving [see @hurleyAssessmentShallowMesophotic2016].
+Shallow units were deployed according to standard NOAA methodology, whereas units $\geq$ 30 m were deployed and recovered using technical mixed gas closed-circuit rebreather diving [see @hurleyAssessmentShallowMesophotic2016].
 
-```{r site-map, fig.cap='Sampling map (Papahānaumokuākea Marine National Monument is shaded in light blue)', results="asis", fig.width=9, fig.height=6, dependson="init"}
+```{r site-map, fig.cap='Map of ARMS deployment site across the Hawaiian Archipelago (Papahānaumokuākea Marine National Monument is shaded in light blue)', results="asis", fig.width=9, fig.height=6, dependson="init"}
 library(sf)
@@
 
-To examine patterns of crab diversity across the islands, we used `estimate_richness` to calculate species richness and Simpson (Simpson = $1-\sum_{i=1}^{R}p_i^2$, where R is the total number of species, and $\mathrm{p}_i$ is the proportional abundance of each species).
-Diversity was compared between MHI and NWHI using two-sample t-tests, both with and without the inclusion of mesophotic samples.
+To examine patterns of crab diversity across the islands, we used `estimate_richness` to calculate species richness and Simpson diversity (Simpson = $1-\sum_{i=1}^{R}p_i^2$, where R is the total number of species, and $\mathrm{p}_i$ is the proportional abundance of each species).
+We used nonparametric Kruskal-Wallis ranked sum tests to compare diversity between shallow and mesophotic samples and between island groups (MHI vs.
+NWHI).
+Island group comparisons were performed both with and without mesophotic samples.
 <!--To explore environmental influences on alpha diversity, we used a linear mixed-effect model with Simpson or richness as response variables, the environmental variables listed below as fixed effects, and island nested within island group (MHI vs NWHI) as random effects.
@@
 Continuous variables were scaled to unit variance.
-We assessed variation among groups using the `betadisper` and `adonis2`.
+We assessed variation among groups using `betadisper` and `adonis2`.
 These functions implement analysis of multivariate homogeneity of group dispersions (PERMDISP) and multivariate permutational analysis of variance (PERMANOVA) respectively.
@@
 
-# 3. Results {-}
+# Results {-}
 
@@
 
-For the complete dataset, there was no significant difference in observed species richness between the MHI and NWHI `r t_test(all_alpha$t_r_island)`, but Simpson diversity was significantly different, although mean values were similar with `r num(all_alpha$richness_main$richness,0.1)` $\pm$ `r num(all_alpha$richness_main$richness_sd,0.1)` in the MHI and `r num(all_alpha$richness_nwhi$richness,0.1)` $\pm$ `r num(all_alpha$richness_nwhi$richness_sd,0.1)` in the NWHI ([@fig:fig_alpha[]{}A]; `r t_test(all_alpha$t_s_island,plain=TRUE)`).
+For the complete dataset, there was no significant difference in observed species richness between the MHI and NWHI ([@fig:fig_alpha[]{}A]; `r k_test(all_alpha$k_r_island)`), but Simpson diversity was significantly different ([@fig:fig_alpha[]{}C]; `r k_test(all_alpha$k_s_island,plain=TRUE)`).
+Richness was not significantly different between shallow and deep samples ([@fig:fig_alpha[]{}B]; `r k_test(all_alpha$k_r_depth,plain=TRUE)`) but Simpson diversity was ([@fig:fig_alpha[]{}D]; `r k_test(all_alpha$k_s_depth,plain=TRUE)`).
+When only shallow arms \< 30 m were considered, neither comparison was significant (richness: `r k_test(shallow_alpha$k_r_island,plain=TRUE)`, [@fig:fig_alpha2[]{}A]; Simpson: `r k_test(shallow_alpha$k_s_island,plain=TRUE)`, [@fig:fig_alpha2[]{}B]).
+
+<!--For the complete dataset, there was no significant difference in observed species richness between the MHI and NWHI `r w_test(all_alpha$w_r_island)`, but Simpson diversity was significantly different ([@fig:fig_alpha[]{}A]; `r w_test(all_alpha$w_s_island,plain=TRUE)`).
+When only shallow arms \< 30 m were considered, neither comparison was significant (richness: `r w_test(shallow_alpha$w_r_island,plain=TRUE)`; Simpson: `r w_test(shallow_alpha$w_s_island,plain=TRUE)`).
+Simpson diversity was significantly different between shallow and deep samples ([@fig:fig_alpha[]{}B]; `r w_test(all_alpha$w_s_depth,plain=TRUE)`) but richness was not (`r w_test(all_alpha$w_r_depth,plain=TRUE)`).
+-->
+
+<!--For the complete dataset, there was no significant difference in observed species richness between the MHI and NWHI `r t_test(all_alpha$t_r_island)`, but Simpson diversity was significantly different ([@fig:fig_alpha[]{}A]; `r t_test(all_alpha$t_s_island,plain=TRUE)`).
 When only shallow arms \< 30 m were considered, neither comparison was significant (richness: `r t_test(shallow_alpha$t_r_island,plain=TRUE)`; Simpson: `r t_test(shallow_alpha$t_s_island,plain=TRUE)`).
 Simpson diversity was significantly different between shallow and deep samples ([@fig:fig_alpha[]{}B]; `r t_test(all_alpha$t_s_depth,plain=TRUE)`) but richness was not (`r t_test(all_alpha$t_r_depth,plain=TRUE)`).
+-->
 
-```{r alpha-box, fig.cap='Box plots of Simpson diversity by (A) island group and (B) sampling depth.
-Notches represent nonparametric confidence intervals about the median', results="asis", dev="svg", fig.width=8, fig.height=4, dependson="init"}
-library(patchwork)
-
-p1 <- ggplot(all_alpha$richness_table %>% mutate(island_group = str_to_sentence(island_group))) + 
-  geom_boxplot(aes(x=island_group,y=Simpson,fill=island_group),notch=TRUE) +
-  scale_fill_manual(values=island_group_pal) +
-  guides(fill="none") + 
-  plotz_theme() + 
-  xlab("Island group") + 
-  ylab("Simpson's diversity index") 
-p2 <- ggplot(all_alpha$richness_table %>% mutate(shallow_deep = str_to_sentence(shallow_deep),shallow_deep = fct_reorder(shallow_deep,depth))) + 
-  geom_boxplot(aes(x=shallow_deep,y=Simpson,fill=shallow_deep),notch=TRUE) +
-  scale_fill_manual(values=sample_depth_pal) +
-  guides(fill="none") + 
-  plotz_theme()  + 
-  xlab("Sampling depth") + 
-  ylab("Simpson's diversity index")
-
-p1 + p2 +
-  plot_annotation(tag_levels = "A")
-cat("{#fig:fig_alpha}")
-```
-
-### Effects of environmental variables {-}
+<!-- ### Effects of environmental variables {-} -->
 
@@
 The best dbRDA model (`r permanova(db_modl,stat="F")`) explained `r pc(RsquareAdj(db_modl)$adj.r.squared,0.01)` (adjusted $\mathrm{R}^2$) of the variability in crab communities across the archipelago ([@fig:fig_rda]).
-`r str_to_sentence(english(nterm))` of the 10 environmental variables tested were significant.
+`r str_to_sentence(english(nterm))` of the 10 environmental variables tested were significant and sufficiently uncorrelated.
 The drivers contributing most to explained variation were `r drivers$term[1]` (F = `r num(drivers$stat[1],0.01)`) and `r drivers$term[2]` (F = `r num(drivers$stat[2],0.01)`) ([@tbl:rda_table]).
@@
 
-# rda_plotz <- plot_ord(cc$crabs_shallow,db_modl,scale=2,color="island_group",labelsize=3,term_map=termz) + 
+# make the initial ordination plot
 rda_plotz <- plot_ord(cc$crabs_shallow,db_modl,scale=2,color="island_group",labelsize=3,term_map=termz) + 
@@
 
-xvar <- eigenvals(db_modl)/sum(eigenvals(db_modl))
-r2 <- RsquareAdj(db_modl)$adj.r.squared
-constrained <- xvar*r2
+# calculate variation explained (constrained & unconstrained)
+eigs <- eigenvals(db_modl)
+constrained <- eigs[str_detect(names(eigs),"dbRDA")]
+constrained <- constrained / sum(constrained)
+unconstrained <- eigs/sum(eigs)
 
+# function to format axis titles
 splained <- function(name,constrained,total,accuracy=0.1) {
@@
 }
+
+axes <- str_glue_data(list(num=1:2),"dbRDA{num}")
+axis_var <- splained(axes,constrained[1:2],unconstrained[1:2])
 
@@
 rda_plotz + 
-  xlab(splained("dbRDA1",xvar[1],constrained[1])) +
-  ylab(splained("dbRDA2",xvar[2],constrained[2])) 
+  xlab(axis_var[1]) + ylab(axis_var[2])
 cat("{#fig:fig_rda}")
@@
 ```{r cap, fig.cap="Canonical analysis of principal coordinates (CAP) plot showing variation in brachyuran crab communities by island group and depth zone.
-Ellipses represent 95% confidence intervals for sample groups.", results="asis", fig.width=9, fig.height=6, dependson="cap-analysis"}
+Ellipses represent 95% confidence intervals around sample groups.", results="asis", fig.width=9, fig.height=6, dependson="cap-analysis"}
 # make the sample group look nice
@@
     str_replace("nwhi","(NWHI)") %>%
+    str_replace("(^[0-9]+)","O‘ahu \\1") %>%
     str_split("((?<=[0-9])(?=[a-zA-Z]))|(_)") %>%
@@
 # get all the stuff we need to plot
-pcoa <- cap_depth$x %>%
+pcap <- cap_depth$x %>%
   as_tibble(rownames="sample") %>%
@@
   mutate(
+    # sample_group = case_when(
+    #   island == "Oahu" & depth_zone == "shallow" ~ "O‘ahu (shallow)",
+    #   .default = sample_group
+    # ),
     sample_group = nice(sample_group),
@@
   )
+ centroids <- pcap %>% group_by(sample_group) %>%
+  summarise(LD1=mean(LD1),LD2=mean(LD2))
 
-ggplot(pcoa,aes(x=LD1,y=LD2,fill=sample_group)) + 
+# complement <- function(pal) {
+#   s <- str_replace(pal,"^#","") %>%
+#     str_match_all("..") %>%
+#     map_chr(~{
+#       ss <- .x[,1]
+#       ss <- str_c("0x",ss) 
+#       ss <- sprintf("%02x",255 - as.numeric(ss))
+#       ss <- str_c("#",str_c(ss,collapse=""))
+#     })
+#   names(s) <- names(pal)
+#   return(s)
+# }
+
+ggplot(pcap,aes(x=LD1,y=LD2,fill=sample_group)) + 
   stat_ellipse(geom="polygon",level=0.95,alpha=0.1,show.legend = FALSE) +
-  geom_point(color="black",shape=21,size=5)  +
+  # geom_point(color="black",shape=21,size=5)  +
+  geom_point(color="black",size=5)  +
+  geom_label_repel(data=centroids,aes(label=sample_group), color="white", show.legend = FALSE) + 
+  # scale_color_manual(values = complement(sample_depth_pal)) + 
   # scale_color_manual(values=sample_depth_pal) + 
@@
 
-# 4. Discussion {-}
+# Discussion {-}
 
@@
 ```
+
+\newpage
+
+```{r alpha-box, fig.cap='Box plots of Simpson diversity by (A) island group and (B) sampling depth for complete dataset (all depths).
+Notches represent nonparametric confidence intervals about the median', results="asis", dev="svg", fig.width=10, fig.height=10, dependson="init"}
+library(patchwork)
+
+art <- all_alpha$richness_table %>% 
+  mutate(island_group = str_to_sentence(island_group)) %>%
+  mutate(shallow_deep = str_to_sentence(shallow_deep),shallow_deep = fct_reorder(shallow_deep,depth))
+p1 <- ggplot(art) + 
+  geom_boxplot(aes(x=island_group,y=Observed,fill=island_group),notch=TRUE) +
+  scale_fill_manual(values=island_group_pal) +
+  guides(fill="none") + 
+  plotz_theme() + 
+  xlab("Island group") + 
+  ylab("Richness") 
+p2 <- ggplot(art) + 
+  geom_boxplot(aes(x=shallow_deep,y=Observed,fill=shallow_deep),notch=TRUE) +
+  scale_fill_manual(values=sample_depth_pal) +
+  guides(fill="none") + 
+  plotz_theme()  + 
+  xlab("Sampling depth") + 
+  ylab("Richness")
+
+p3 <- ggplot(art) + 
+  geom_boxplot(aes(x=island_group,y=Simpson,fill=island_group),notch=TRUE) +
+  scale_fill_manual(values=island_group_pal) +
+  guides(fill="none") + 
+  plotz_theme() + 
+  xlab("Island group") + 
+  ylab("Simpson's diversity index") 
+p4 <- ggplot(art) + 
+  geom_boxplot(aes(x=shallow_deep,y=Simpson,fill=shallow_deep),notch=TRUE) +
+  scale_fill_manual(values=sample_depth_pal) +
+  guides(fill="none") + 
+  plotz_theme()  + 
+  xlab("Sampling depth") + 
+  ylab("Simpson's diversity index")
+
+(p1 + p2) / (p3 + p4) & 
+  plot_annotation(tag_levels = "A")
+cat("{#fig:fig_alpha}")
+```
+
+```{r alpha-box2, fig.cap='Box plots of (A) species richness and (B) Simpson diversity and by island group for shallow samples only.', results="asis", dev="svg", fig.width=8, fig.height=4, dependson="init"}
+art <- shallow_alpha$richness_table %>% 
+  mutate(island_group = str_to_sentence(island_group)) %>%
+  mutate(shallow_deep = str_to_sentence(shallow_deep),shallow_deep = fct_reorder(shallow_deep,depth))
+
+p1 <- ggplot(art) + 
+  geom_boxplot(aes(x=island_group,y=Observed,fill=island_group),notch=TRUE) +
+  scale_fill_manual(values=island_group_pal) +
+  guides(fill="none") + 
+  plotz_theme() + 
+  xlab("Island group") + 
+  ylab("Richness")
+
+p2 <- ggplot(art) + 
+  geom_boxplot(aes(x=island_group,y=Simpson,fill=island_group),notch=TRUE) +
+  scale_fill_manual(values=island_group_pal) +
+  guides(fill="none") + 
+  plotz_theme() + 
+  xlab("Island group") + 
+  ylab("Simpson's diversity index")
+
+(p1 + p2) & 
+  plot_annotation(tag_levels = "A")
+cat("{#fig:fig_alpha2}")
+```

### 2023-10-10 crabs.R

@@ -471,6 +471,16 @@ alpha_diversity <- function(ps,measures=c("Observed","Simpson")) {
   try(ad$t_r_depth <- t.test(Observed ~ shallow_deep, data=all_richness),silent=TRUE)
   try(ad$t_s_depth <- t.test(Simpson ~ shallow_deep, data=all_richness),silent=TRUE)
   
+  try(ad$k_r_island <- kruskal.test(Observed ~ island_group, data=all_richness),silent=TRUE)
+  try(ad$k_s_island <- kruskal.test(Simpson ~ island_group, data=all_richness),silent=TRUE)
+  try(ad$k_r_depth <- kruskal.test(Observed ~ shallow_deep, data=all_richness),silent=TRUE)
+  try(ad$k_s_depth <- kruskal.test(Simpson ~ shallow_deep, data=all_richness),silent=TRUE)
+  
+  try(ad$w_r_island <- wilcox.test(Observed ~ island_group, data=all_richness),silent=TRUE)
+  try(ad$w_s_island <- wilcox.test(Simpson ~ island_group, data=all_richness),silent=TRUE)
+  try(ad$w_r_depth <- wilcox.test(Observed ~ shallow_deep, data=all_richness),silent=TRUE)
+  try(ad$w_s_depth <- wilcox.test(Simpson ~ shallow_deep, data=all_richness),silent=TRUE)
+  
   # m <- lm(Observed ~ island_group, data=all_richness)
   # summary(m)
   # check_model(m)

### 2023-10-19 manuscript.Rmd
some updates

@@
 ---
-title: "It's easier to move sideways"
+title: "Depth over distance: Hawaiian cryptic reef communities are more distinct across ~90 m depth than ~2500 km of shallow habitat"
 author:
@@
 
-\newpage
-
 ```{r setup, include=FALSE}
@@
 # Introduction {-}
-Natural communities are structured by a suite of biotic and abiotic factors including geography, habitat complexity, environment, productivity, and species interactions (among many others) [@begonEcologyIndividualsEcosystems2021] and such complexity makes direct comparisons over large geographic scales challenging.
-Marine communities in particular experience unique influences and present research challenges not found in terrestrial habitats [@steeleComparisonTerrestrialMarine1985; @kinlanPropaguleDispersalMarine2003; @raffaelliMarineTerrestrialEcologists2005; @grosbergBiodiversityWaterLand2012; @knappDriversBiodiversityChange2017].
-Much of our understanding of biotic zonation and community structure in the ocean comes from classic studies in the intertidal (some of the best-studied marine habitats) [e.g., @connellInfluenceInterspecificCompetition1961; @connellEffectsCompetitionPredation1961; @paineIntertidalCommunityStructure1974].
-Such studies point to a combination of physical and biological factors, including physiological tolerance [@dennyPhysicalProcessesThat2001; @someroThermalPhysiologyVertical2002] and species interactions [@brunoHabitatModificationFacilitation2001; @mengeRockyIntertidalCommunities2001], that contribute to the structure of intertidal communities.
-Although there are exceptions [e.g., @woottonIndirectEffectsPrey1992; @bertnessTestingRelativeContribution1999], a generalization has emerged for the intertidal that lower limits of species distributions are set by ecological interactions whereas upper limits are determined by abiotic factors.
+
+The structure and composition of coral reef communities is influenced by a suite of environmental and anthropogenic factors, yet their nature and makeup remain poorly resolved.
+Coral reefs are among the most diverse habitats in the world, yet they face increasing threats of degradation, including climate-driven bleaching and phase shifts to alternate ecological regimes [@grahamManagingResilienceReverse2013; @norstromAlternativeStatesCoral2009; @pandolfiGlobalTrajectoriesLongTerm2003].
+In the face of these and other threats, understanding drivers of community structure is critically important [@hughesRisingChallengeSustaining2010; @pageSeekingResistanceCoral2019].
+To date, the majority of efforts in this area have focused on conspicuous taxa such as corals and fishes [e.g., @benfieldComparisonCoralReef2008; @caballero-aragonWaveExposureTemperature2023; @kaneTrophicDesignationLive2017; @langeWaveExposureShapes2021; @vanlierHabitatConnectivityComplexity2018] or microbes [e.g., @desouzaImportanceDepthTemperature2023; @morrowEnvironmentalDriversMicrobial2016].
+However, these taxa make up a minority of coral reef biodiversity and thus may not be indicative of community patterns overall [[cite?]]{.mark}.
+Here, we investigate the factors influencing the community structure of a group of cryptic coral reef taxa across the Hawaiian Archipelago.
+<!--Say something about depth here, but I'm not sure what or how-->
+
+Much of our understanding of biotic zonation and community structure in the ocean originates with classic studies from the intertidal [e.g., @connellInfluenceInterspecificCompetition1961; @connellEffectsCompetitionPredation1961; @paineIntertidalCommunityStructure1974].
+Such studies point to a combination of physical and biological factors<!--, including phyiological tolerance [@dennyPhysicalProcessesThat2001; @someroThermalPhysiologyVertical2002] and species interactions [@brunoHabitatModificationFacilitation2001; @mengeRockyIntertidalCommunities2001], that contribute to the structure of intertidal communities.
+Although there are exceptions [e.g., @woottonIndirectEffectsPrey1992; @bertnessTestingRelativeContribution1999]-->, but (with exceptions) a generalization has emerged <!--for the intertidal -->that communities' lower bathymetric limits are regulated by ecological interactions whereas upper limits are determined by abiotic factors.
+However, these models describe zonation at individual localities rather than community variation over larger spatial scales and may not apply to coral reefs, which are subject to more complex community dynamics [@abramsRecruitmentLotteriesCoexistence1984; @bussCompetitiveNetworksNontransitive1979; @hixonPredationPreyRefuges1993; @hubbellUnifiedNeutralTheory2011; @moraEcologyFishesCoral2015; @saleStockRecruitRelationshipsRegional1982; @sheppardBiologyCoralReefs2017].
+In any case, it is likely that while ecological interactions are important for structuring communities at the local scale, external factors may exert stronger influence across greater distances [[cite?]]{.mark}.
+
+The Hawaiian Archipelago comprises high volcanic islands and atolls stretching across ~2,500 km in the North Pacific Ocean, and is typically regarded as two regions: the main Hawaiian Islands (MHI), which consist of the eight inhabited high islands, and the Northwestern Hawaiian Islands (NWHI), which are the uninhabited atolls and banks northwest of Kaua&#x02BB;i and Ni&#x02BB;ihau.
+The islands are subject to anthropogenic and physical gradients across their span [@selkoeMapHumanImpacts2009], which we hypothesize to influence coral reef communities.
+Studies examining community structure across environmental gradients have been criticized for confounding multiple potentially covarying factors (such as habitat area and/or complexity) across the measured gradients [e.g., @bellwoodRegionalScaleAssemblyRules2001; @grahamEffectsMarineReserve2003; @connollyCommunityStructureCorals2005; @ewersConfoundingFactorsDetection2006; @fletcherRoleHabitatArea2007a].
+Thus, we used autonomous reef monitoring structures (ARMS)---standardized collection devices designed to mimic reef structural complexity---to sample cryptic communities while controlling for habitat<!--differences in habitat variables--> [@zimmermanArtificialReefMatrix2004; @brainardAutonomousReefMonitoring2009].
+Specifically, we used ARMS to sample brachyuran crabs, a diverse group of arthropods that comprise a considerable proportion of coral reef cryptofauna [289 species occur in Hawai&#x02BB;i, @castroCatalogAnomuranBrachyuran2011].
+We compared crab communities among islands and regions (MHI vs.
+NWHI) and examined the effects of a suite of possible environmental and anthropogenic drivers.
+Additionally, given the strong reported differences between deep and shallow reefs across taxa [@hobanPlumbingDepthsEnvironmental2023; @hurleyAssessmentShallowMesophotic2016; @rochaMesophoticCoralEcosystems2018], we also compared the relative contributions of depth (shallow vs.
+mesophotic) and island region to crab community variation.
 
@@
 These habitats, known as mesophotic coral ecosystems (MCEs), are typically defined as those extending from ~30--150 m depth [@hindersteinThemeSectionMesophotic2010; @kahngCommunityEcologyMesophotic2010].
-<!--and were believed to experience dampened disturbances in comparison to shallow reefs and could therefore act as deep water refugia for shallow reef species threatened by anthropogenic impacts --> They have been hypothesized as possible refugia from disturbances on shallow reefs [@bongaertsAssessingDeepReef2010; @smithDepthRefugiumCatastrophic2014; @holsteinFertileFathomsDeep2015; @muirSpeciesIdentityDepth2017], although a number of studies have found evidence that deep reefs are unique and uniquely threatened, challenging this notion [@hobanPlumbingDepthsEnvironmental2023a; @hurleyAssessmentShallowMesophotic2016; @kaneHighLevelsMesophotic2014; @pinheiroDeepReefFishes2019; @rochaMesophoticCoralEcosystems2018].
+<!--and were believed to experience dampened disturbances in comparison to shallow reefs and could therefore act as deep water refugia for shallow reef species threatened by anthropogenic impacts --> They have been hypothesized as possible refugia from disturbances on shallow reefs [@bongaertsAssessingDeepReef2010; @smithDepthRefugiumCatastrophic2014; @holsteinFertileFathomsDeep2015; @muirSpeciesIdentityDepth2017], although a number of studies have found evidence that deep reefs are unique and uniquely threatened, challenging this notion [@hobanPlumbingDepthsEnvironmental2023; @hurleyAssessmentShallowMesophotic2016; @kaneHighLevelsMesophotic2014; @pinheiroDeepReefFishes2019; @rochaMesophoticCoralEcosystems2018].
 <!--However, recent literature describing species compositions along depth gradients reveal a more complex picture among taxa and locations because some groups may use mesophotic reefs as a refuge in some places, while others show community turnover with depth such that MCE habitats are ecologically distinct from adjacent shallow water coral reefs [@hurleyAssessmentShallowMesophotic2016; @smithCaribbeanMesophoticCoral2016; @semmlerAreMesophoticCoral2017; @bongaertsDeepReefsAre2017; @rochaMesophoticCoralEcosystems2018].
@@
 
-Studies examining community structure along depth or latitudinal gradients have been criticized for confounding multiple potentially covarying aspects of the environment (such as habitat area or complexity) across the measured gradients [e.g., @bellwoodRegionalScaleAssemblyRules2001; @grahamEffectsMarineReserve2003; @connollyCommunityStructureCorals2005; @ewersConfoundingFactorsDetection2006; @fletcherRoleHabitatArea2007a].
-Here, we employed autonomous reef monitoring structures (ARMS) [@zimmermanArtificialReefMatrix2004; @brainardAutonomousReefMonitoring2009], collection devices designed to mimic reef structural complexity and enable sampling of cryptic fauna in a standardized way.
-Using ARMS, we sampled brachyuran crab communities across the Hawaiian Archipelago to test the effects of environmental and anthropogenic drivers on the structure of cryptic reef communities.
-The Hawaiian Archipelago comprises high volcanic islands and atolls stretching across ~2,500 km in the North Pacific Ocean, and is typically regarded as two regions: the main Hawaiian Islands (MHI) are the eight inhabited high islands, and the Northwestern Hawaiian Islands (NWHI) are the uninhabited atolls and banks northwest of Kaua&#x02BB;i.
-The islands are subject to anthropogenic and physical gradients across their span [@selkoeMapHumanImpacts2009], which we hypothesize to affect the variation in structure of sampled communities.
-We compared crab communities among islands and regions (MHI vs.
-NWHI) and examined the effects of a suite of possible drivers on community variation.
-Given the possibility for considerable separation between deep and shallow reefs [e.g., @hobanPlumbingDepthsEnvironmental2023a; @hurleyAssessmentShallowMesophotic2016; @rochaMesophoticCoralEcosystems2018], we also compared the relative effects of depth (shallow vs.
-mesophotic) and island region.
+Here, we employed autonomous reef monitoring structures (ARMS)  Using ARMS, we sampled brachyuran crab communities across the Hawaiian Archipelago to test the effects of environmental and anthropogenic drivers on the structure of cryptic reef communities.
 Given the number and magnitude of differences between these regions, we hypothesized that species diversity would be different between regions (MHI vs NWHI).
@@
 
-```{r site-map, fig.cap='Map of ARMS deployment site across the Hawaiian Archipelago (Papahānaumokuākea Marine National Monument is shaded in light blue)', results="asis", fig.width=9, fig.height=6, dependson="init"}
+```{r fig-map, fig.cap='Map of ARMS deployment site across the Hawaiian Archipelago (Papahānaumokuākea Marine National Monument is shaded in light blue)', results="asis", fig.width=9, fig.height=6, dependson="init"}
 library(sf)
@@
   ) %>%
-  # mutate( island = fct_reorder(island,-lat) ) %>%
   st_as_sf(coords=c("lon","lat"),crs=4326, remove=F) #%>%
@@
 
-box_before <- st_bbox(st_buffer(crab_sites,units::as_units(50,"km")))
+shift_box <- function(b) {
+  if (b['xmin'] < 0) 
+    b['xmin'] <- b['xmin'] + 360
+  if (b['xmax'] < 0) 
+    b['xmax'] <- b['xmax'] + 360
+  return(b)
+}
+
+pmnm <- st_read(here("data","gis","hi_noaa_nwhi_papahanaumokuakea.shp"),quiet = TRUE) %>%
+  st_shift_longitude()
+
+pbox <- st_bbox(pmnm)
+cbox <- st_bbox(st_buffer(crab_sites,units::as_units(50,"km")) %>% st_shift_longitude())
+hbox <- hawaii %>%
+  st_shift_longitude() %>%
+  st_bbox()
+
+box_before <- c(
+  xmin = min(c(pbox['xmin'],hbox['xmin'],hbox['xmin'])),
+  ymin = min(c(pbox['ymin'],cbox['ymin'],hbox['ymin'])),
+  xmax = max(c(pbox['xmax'],cbox['xmax'],hbox['xmax'])) + 1.5,
+  ymax = max(c(pbox['ymax'],cbox['ymax'],hbox['ymax']))
+)
 
@@
 
-pmnm <- st_read(here("data","gis","pmnm.shp"),quiet = TRUE) %>%
+pmnm <- pmnm %>%
   st_transform(transformer)
@@
 ggplot() +
-  # geom_contour(data=bath_proj,mapping=aes(x=V1,y=V2,z = V3),  color="#cecece") +
-  # geom_raster(data=bpr,aes(x=x,y=y,fill=z)) +
-  # geom_tile(data=bath_proj,mapping=aes(x=V1,y=V2,fill = V3)) +
-  # scale_fill_gradient2(low="dodgerblue4", mid="gainsboro", high="darkgreen") +
   geom_sf(data=hawaii %>% st_transform(transformer), color="black", fill="green", linewidth=0.4) +
-  geom_sf(data=pmnm,fill="#6ed8f6",alpha=0.1) +
+  geom_sf(data=pmnm,color=NA,fill="#6ed8f6",alpha=0.2) +
   geom_sf(data=crab_sites %>% st_transform(transformer),aes(fill=island),size=3,shape=21,color="black") +
@@
 
+Environmental variables clustered strongly according to island group (MHI vs NWHI), with the difference driven primarily by human impact, coral cover, and island slope ([@fig:fig_pca]).
 The best dbRDA model (`r permanova(db_modl,stat="F")`) explained `r pc(RsquareAdj(db_modl)$adj.r.squared,0.01)` (adjusted $\mathrm{R}^2$) of the variability in crab communities across the archipelago ([@fig:fig_rda]).
@@
 
-```{r rda, fig.cap="Distance-based redundancy analysis (dbRDA) plot showing significant environmental drivers of crab community variation across the Hawaiian Islands.", results="asis", fig.width=9, fig.height=6, dependson="rda-analysis"}
+```{r fig-rda, fig.cap="Distance-based redundancy analysis (dbRDA) plot showing significant environmental drivers of crab community variation across the Hawaiian Islands.", results="asis", fig.width=9, fig.height=6, dependson="rda-analysis"}
 termz <- model_terms %>%
@@
 
-```{r cap, fig.cap="Canonical analysis of principal coordinates (CAP) plot showing variation in brachyuran crab communities by island group and depth zone.
+```{r fig-cap, fig.cap="Canonical analysis of principal coordinates (CAP) plot showing variation in brachyuran crab communities by island group and depth zone.
 Ellipses represent 95% confidence intervals around sample groups.", results="asis", fig.width=9, fig.height=6, dependson="cap-analysis"}
@@
   mutate(
-    # sample_group = case_when(
-    #   island == "Oahu" & depth_zone == "shallow" ~ "O‘ahu (shallow)",
-    #   .default = sample_group
-    # ),
     sample_group = nice(sample_group),
@@
 
-# complement <- function(pal) {
-#   s <- str_replace(pal,"^#","") %>%
-#     str_match_all("..") %>%
-#     map_chr(~{
-#       ss <- .x[,1]
-#       ss <- str_c("0x",ss) 
-#       ss <- sprintf("%02x",255 - as.numeric(ss))
-#       ss <- str_c("#",str_c(ss,collapse=""))
-#     })
-#   names(s) <- names(pal)
-#   return(s)
-# }
-
 ggplot(pcap,aes(x=LD1,y=LD2,fill=sample_group)) + 
   stat_ellipse(geom="polygon",level=0.95,alpha=0.1,show.legend = FALSE) +
-  # geom_point(color="black",shape=21,size=5)  +
-  geom_point(color="black",size=5)  +
+  geom_point(color="black",shape=21,size=5)  +
   geom_label_repel(data=centroids,aes(label=sample_group), color="white", show.legend = FALSE) + 
-  # scale_color_manual(values = complement(sample_depth_pal)) + 
-  # scale_color_manual(values=sample_depth_pal) + 
   scale_fill_manual(values=sample_depth_pal,name="Sample group") +
@@
 # Supplemental material {label="S"}
+
+```{r fig-alpha, fig.cap='Box plots of Simpson diversity by (A) island group and (B) sampling depth for complete dataset (all depths).
+Notches represent nonparametric confidence intervals about the median', results="asis", dev="svg", fig.width=10, fig.height=10, dependson="init"}
+library(patchwork)
+
+art <- all_alpha$richness_table %>% 
+  mutate(island_group = str_to_sentence(island_group)) %>%
+  mutate(shallow_deep = str_to_sentence(shallow_deep),shallow_deep = fct_reorder(shallow_deep,depth))
+p1 <- ggplot(art) + 
+  geom_boxplot(aes(x=island_group,y=Observed,fill=island_group),notch=TRUE) +
+  scale_fill_manual(values=island_group_pal) +
+  guides(fill="none") + 
+  plotz_theme() + 
+  xlab("Island group") + 
+  ylab("Richness") 
+p2 <- ggplot(art) + 
+  geom_boxplot(aes(x=shallow_deep,y=Observed,fill=shallow_deep),notch=TRUE) +
+  scale_fill_manual(values=sample_depth_pal) +
+  guides(fill="none") + 
+  plotz_theme()  + 
+  xlab("Sampling depth") + 
+  ylab("Richness")
+
+p3 <- ggplot(art) + 
+  geom_boxplot(aes(x=island_group,y=Simpson,fill=island_group),notch=TRUE) +
+  scale_fill_manual(values=island_group_pal) +
+  guides(fill="none") + 
+  plotz_theme() + 
+  xlab("Island group") + 
+  ylab("Simpson's diversity index") 
+p4 <- ggplot(art) + 
+  geom_boxplot(aes(x=shallow_deep,y=Simpson,fill=shallow_deep),notch=TRUE) +
+  scale_fill_manual(values=sample_depth_pal) +
+  guides(fill="none") + 
+  plotz_theme()  + 
+  xlab("Sampling depth") + 
+  ylab("Simpson's diversity index")
+
+(p1 + p2) / (p3 + p4) & 
+  plot_annotation(tag_levels = "A")
+cat("{#fig:fig_alpha}")
+```
+
+\newpage
+
+```{r fig-alpha2, fig.cap='Box plots of (A) species richness and (B) Simpson diversity and by island group for shallow samples only.', results="asis", dev="svg", fig.width=8, fig.height=4, dependson="init"}
+art <- shallow_alpha$richness_table %>% 
+  mutate(island_group = str_to_sentence(island_group)) %>%
+  mutate(shallow_deep = str_to_sentence(shallow_deep),shallow_deep = fct_reorder(shallow_deep,depth))
+
+p1 <- ggplot(art) + 
+  geom_boxplot(aes(x=island_group,y=Observed,fill=island_group),notch=TRUE) +
+  scale_fill_manual(values=island_group_pal) +
+  guides(fill="none") + 
+  plotz_theme() + 
+  xlab("Island group") + 
+  ylab("Richness")
+
+p2 <- ggplot(art) + 
+  geom_boxplot(aes(x=island_group,y=Simpson,fill=island_group),notch=TRUE) +
+  scale_fill_manual(values=island_group_pal) +
+  guides(fill="none") + 
+  plotz_theme() + 
+  xlab("Island group") + 
+  ylab("Simpson's diversity index")
+
+(p1 + p2) & 
+  plot_annotation(tag_levels = "A")
+cat("{#fig:fig_alpha2}")
+```
+
+\newpage
+
+```{r fig-pca, fig.cap='Principal component analysis (PCA) of environmental variables by sampling site.
+Ellipses represent 95% confidence intervals around island groups', results='asis', dev='svg', fig.width=8,fig.height=8, dependson='init'}
+cds <- cc$crab_data_shallow
+
+pcdata <- cds %>%
+  select(where(is.numeric))
+
+pca <- prcomp(pcdata)
+explained <- (pca$sdev^2)/sum(pca$sdev^2)
+
+bp <- cds %>%
+  bind_cols(scores(pca))
+
+loadings <- pca$rotation %>%
+  as_tibble(rownames = "label") %>%
+  mutate(
+    PC1 = PC1*3,
+    PC2 = PC2*3,
+    hjust = if_else(PC1 < 0,1,0),
+    xnudge = if_else(PC1 < 0,-0.02,0.02)
+  )
+
+ggplot(bp,aes(x=PC1,y=PC2)) + 
+  geom_point(aes(fill=island_group),shape=21,color="black",size=4) +
+  stat_ellipse(geom="polygon",aes(group=island_group,fill=island_group),color="black",alpha=0.1,show.legend = FALSE, size=0.07) +
+  # geom_mark_ellipse(color="black",alpha=0.2,show.legend=FALSE, linewidth=0.07) +
+  scale_fill_manual(values=island_group_pal) + 
+  geom_segment(data=loadings,aes(x=0,y=0,xend=PC1,yend=PC2),color="black",size=0.5,arrow=arrow(length=unit(0.02,"npc"))) + 
+  geom_label(data=loadings,aes(x=PC1+xnudge,y=PC2,label=label,hjust=hjust)) + 
+  xlab(str_glue("PC1 [{pc(explained[1])}]")) + 
+  ylab(str_glue("PC2 [{pc(explained[2])}]"))
+cat("{#fig:fig_pca}")
+```
+
+\newpage
 
@@
 ```{r rda-table,results="asis",dependson="rda-analysis"}
-print_table(rda_table,"rda_table","Significant (marginal) terms in the dbRDA analysis of crab community variation across the Hawaiian Archipelago",hdr_case="none",accuracy=0.001) 
-```
-
-\newpage
-
-```{r alpha-box, fig.cap='Box plots of Simpson diversity by (A) island group and (B) sampling depth for complete dataset (all depths).
-Notches represent nonparametric confidence intervals about the median', results="asis", dev="svg", fig.width=10, fig.height=10, dependson="init"}
-library(patchwork)
-
-art <- all_alpha$richness_table %>% 
-  mutate(island_group = str_to_sentence(island_group)) %>%
-  mutate(shallow_deep = str_to_sentence(shallow_deep),shallow_deep = fct_reorder(shallow_deep,depth))
-p1 <- ggplot(art) + 
-  geom_boxplot(aes(x=island_group,y=Observed,fill=island_group),notch=TRUE) +
-  scale_fill_manual(values=island_group_pal) +
-  guides(fill="none") + 
-  plotz_theme() + 
-  xlab("Island group") + 
-  ylab("Richness") 
-p2 <- ggplot(art) + 
-  geom_boxplot(aes(x=shallow_deep,y=Observed,fill=shallow_deep),notch=TRUE) +
-  scale_fill_manual(values=sample_depth_pal) +
-  guides(fill="none") + 
-  plotz_theme()  + 
-  xlab("Sampling depth") + 
-  ylab("Richness")
-
-p3 <- ggplot(art) + 
-  geom_boxplot(aes(x=island_group,y=Simpson,fill=island_group),notch=TRUE) +
-  scale_fill_manual(values=island_group_pal) +
-  guides(fill="none") + 
-  plotz_theme() + 
-  xlab("Island group") + 
-  ylab("Simpson's diversity index") 
-p4 <- ggplot(art) + 
-  geom_boxplot(aes(x=shallow_deep,y=Simpson,fill=shallow_deep),notch=TRUE) +
-  scale_fill_manual(values=sample_depth_pal) +
-  guides(fill="none") + 
-  plotz_theme()  + 
-  xlab("Sampling depth") + 
-  ylab("Simpson's diversity index")
-
-(p1 + p2) / (p3 + p4) & 
-  plot_annotation(tag_levels = "A")
-cat("{#fig:fig_alpha}")
-```
-
-```{r alpha-box2, fig.cap='Box plots of (A) species richness and (B) Simpson diversity and by island group for shallow samples only.', results="asis", dev="svg", fig.width=8, fig.height=4, dependson="init"}
-art <- shallow_alpha$richness_table %>% 
-  mutate(island_group = str_to_sentence(island_group)) %>%
-  mutate(shallow_deep = str_to_sentence(shallow_deep),shallow_deep = fct_reorder(shallow_deep,depth))
-
-p1 <- ggplot(art) + 
-  geom_boxplot(aes(x=island_group,y=Observed,fill=island_group),notch=TRUE) +
-  scale_fill_manual(values=island_group_pal) +
-  guides(fill="none") + 
-  plotz_theme() + 
-  xlab("Island group") + 
-  ylab("Richness")
-
-p2 <- ggplot(art) + 
-  geom_boxplot(aes(x=island_group,y=Simpson,fill=island_group),notch=TRUE) +
-  scale_fill_manual(values=island_group_pal) +
-  guides(fill="none") + 
-  plotz_theme() + 
-  xlab("Island group") + 
-  ylab("Simpson's diversity index")
-
-(p1 + p2) & 
-  plot_annotation(tag_levels = "A")
-cat("{#fig:fig_alpha2}")
-```
+print_table(rda_table,"rda_table","Significant marginal terms in the dbRDA analysis of crab community variation across the Hawaiian Archipelago",hdr_case="none",accuracy=0.001) 
+```

### 2023-10-19 crabs.R

@@ -320,7 +320,7 @@ setup_crabs <- function() {
   # we care about and scale the numeric ones to unit variance
   cc$crab_data_shallow <- cc$crabs_shallow %>%
     sample_tibble(sample_col = "unit") %>%
-    select(unit,region,island_group,island,depth,chl,sst,slope,coral_cover,closest_island,larval_connectivity,human_impact) %>%
+    select(unit,region,island_group,island,lat,lon,depth,chl,sst,slope,coral_cover,closest_island,larval_connectivity,human_impact) %>%
     mutate(across(where(is.numeric),~as.numeric(scale(.x)))) %>%
     column_to_rownames("unit") 
     # reassociate the new sample data

### 2023-10-20 manuscript.Rmd
updates

@@
     keep_md: true
-tblPrefix: Table
+tblPrefix: table
 figPrefix:
@@
 
-The Hawaiian Archipelago comprises high volcanic islands and atolls stretching across ~2,500 km in the North Pacific Ocean, and is typically regarded as two regions: the main Hawaiian Islands (MHI), which consist of the eight inhabited high islands, and the Northwestern Hawaiian Islands (NWHI), which are the uninhabited atolls and banks northwest of Kaua&#x02BB;i and Ni&#x02BB;ihau.
+The Hawaiian Archipelago comprises high volcanic islands and atolls stretching across ~2,500 km in the North Pacific Ocean, and is typically regarded as two regions: the main Hawaiian Islands (MHI), which include the eight inhabited high islands, and the Northwestern Hawaiian Islands (NWHI), the uninhabited atolls and banks northwest of Kaua&#x02BB;i and Ni&#x02BB;ihau.
 The islands are subject to anthropogenic and physical gradients across their span [@selkoeMapHumanImpacts2009], which we hypothesize to influence coral reef communities.
@@
 Thus, we used autonomous reef monitoring structures (ARMS)---standardized collection devices designed to mimic reef structural complexity---to sample cryptic communities while controlling for habitat<!--differences in habitat variables--> [@zimmermanArtificialReefMatrix2004; @brainardAutonomousReefMonitoring2009].
-Specifically, we used ARMS to sample brachyuran crabs, a diverse group of arthropods that comprise a considerable proportion of coral reef cryptofauna [289 species occur in Hawai&#x02BB;i, @castroCatalogAnomuranBrachyuran2011].
+Specifically, we used ARMS to sample brachyuran crabs, a group comprising a notable proportion of the coral reef cryptofauna [289 species in Hawai&#x02BB;i, @castroCatalogAnomuranBrachyuran2011].
 We compared crab communities among islands and regions (MHI vs.
 NWHI) and examined the effects of a suite of possible environmental and anthropogenic drivers.
-Additionally, given the strong reported differences between deep and shallow reefs across taxa [@hobanPlumbingDepthsEnvironmental2023; @hurleyAssessmentShallowMesophotic2016; @rochaMesophoticCoralEcosystems2018], we also compared the relative contributions of depth (shallow vs.
-mesophotic) and island region to crab community variation.
-
-Despite conclusions based on intertidal studies, similar generalizations have not emerged to describe coral reef communities [@moraEcologyFishesCoral2015; @sheppardBiologyCoralReefs2017], with existing hypotheses ranging from the neutral theory of biodiversity [@hubbellUnifiedNeutralTheory2011] and lottery recruitment [@saleStockRecruitRelationshipsRegional1982] to competitive networks [@bussCompetitiveNetworksNontransitive1979] and fine-scale resource partitioning driven by predators and herbivores [@abramsRecruitmentLotteriesCoexistence1984; @hixonPredationPreyRefuges1993].
-Studies that have investigated these topics on coral reefs have traditionally focused on conspicuous taxa such as fishes and corals [@cite], yet these groups constitute a minority of total reef biodiversity [@cite].
-The majority of coral reef biodiversity is cryptic, inhabiting interstitial spaces, and little is known about factors contributing to     Zooxanthellate corals are the ecosystem engineers of reef habitats but are restricted to specific areas characterized by a relatively narrow range of chemical and physical parameters<!-- (i.e., tropical and subtropical warm waters, oligotrophic conditions, 32-40 PSU salinity, and sufficient incident light, which decreases rapidly below ~30 m depth)-->.
-Shallow coral reef ecosystems cover less than 0.2% of the seafloor and in such a small area they hold the 32-38% of all marine species [@fisherSpeciesRichnessCoral2015].
-Yet the <!--drivers of high biodiversity and the -->mechanisms by which communities are structured on coral reefs remain a subject of debate [@carrBiodiversityPopulationRegulation2002; @connollyIndoPacificBiodiversityCoral2003; @almanyDoesIncreasedHabitat2004; @bellwoodEnvironmentalGeometricConstraints2005, 2006; @dornelasCoralReefDiversity2006; @grahamImportanceStructuralComplexity2013; @komyakovaRelativeImportanceCoral2013].
-*Add some text about community structuring across space, rather than just at smaller scales*.
-
-A<!--nother--> likely important but under-investigated aspect of coral reef community structure is depth.
+
+A likely important but under-investigated aspect of coral reef community structure is depth.
 Most studies to date have focused on the upper ~30 m of the water column, yet coral reefs and associated communities are known to extend beyond 150 m [@pyleExploringDeepCoral1996; @pyleMesophoticCoralEcosystems2019], with scleractinian corals reported to 170 m [@rouzeSymbioticAssociationsDeepest2021].
@@
 These habitats, known as mesophotic coral ecosystems (MCEs), are typically defined as those extending from ~30--150 m depth [@hindersteinThemeSectionMesophotic2010; @kahngCommunityEcologyMesophotic2010].
-<!--and were believed to experience dampened disturbances in comparison to shallow reefs and could therefore act as deep water refugia for shallow reef species threatened by anthropogenic impacts --> They have been hypothesized as possible refugia from disturbances on shallow reefs [@bongaertsAssessingDeepReef2010; @smithDepthRefugiumCatastrophic2014; @holsteinFertileFathomsDeep2015; @muirSpeciesIdentityDepth2017], although a number of studies have found evidence that deep reefs are unique and uniquely threatened, challenging this notion [@hobanPlumbingDepthsEnvironmental2023; @hurleyAssessmentShallowMesophotic2016; @kaneHighLevelsMesophotic2014; @pinheiroDeepReefFishes2019; @rochaMesophoticCoralEcosystems2018].
+<!--and were believed to experience dampened disturbances in comparison to shallow reefs and could therefore act as deep water refugia for shallow reef species threatened by anthropogenic impacts --> They have been hypothesized as possible refugia from disturbances on shallow reefs [@bongaertsAssessingDeepReef2010; @smithDepthRefugiumCatastrophic2014; @holsteinFertileFathomsDeep2015; @muirSpeciesIdentityDepth2017], although a number of studies have found evidence that deep reefs are unique and uniquely threatened, challenging this notion [@diazLightTemperatureDrive; @diazMesophoticCoralBleaching2023; @hobanPlumbingDepthsEnvironmental2023; @hurleyAssessmentShallowMesophotic2016; @kaneHighLevelsMesophotic2014; @pinheiroDeepReefFishes2019; @rochaMesophoticCoralEcosystems2018].
 <!--However, recent literature describing species compositions along depth gradients reveal a more complex picture among taxa and locations because some groups may use mesophotic reefs as a refuge in some places, while others show community turnover with depth such that MCE habitats are ecologically distinct from adjacent shallow water coral reefs [@hurleyAssessmentShallowMesophotic2016; @smithCaribbeanMesophoticCoral2016; @semmlerAreMesophoticCoral2017; @bongaertsDeepReefsAre2017; @rochaMesophoticCoralEcosystems2018].
 Further, there are a range of connections depending on the taxa – large, mobile organisms like fish tend to have more shared species across depth than sedentary organisms such as corals and algae, but even among the fishes the community structure of MCEs is often distinct [@kaneHighLevelsMesophotic2014; @pyleComprehensiveInvestigationMesophotic2016 ; @asherMesophoticDepthGradients2017; @fukunagaChangesMesophoticReef2017; @colemanShiftingReefFish2018; @pinheiroDeepReefFishes2019].
-Although the majority of coral reef biodiversity consists of the algal and invertebrate cryptobenthic organisms living on and within the reef structure [@fautinOverviewMarineBiodiversity2010; @knowltonCoralReefBiodiversity2010; @fisherSpeciesRichnessCoral2015], the studies listed above focus almost exclusively on the corals and fishes which are more easily surveyed.-->
-
-Here, we employed autonomous reef monitoring structures (ARMS)  Using ARMS, we sampled brachyuran crab communities across the Hawaiian Archipelago to test the effects of environmental and anthropogenic drivers on the structure of cryptic reef communities.
+Although the majority of coral reef biodiversity consists of the algal and invertebrate cryptobenthic organisms living on and within the reef structure [@fautinOverviewMarineBiodiversity2010; @knowltonCoralReefBiodiversity2010; @fisherSpeciesRichnessCoral2015], the studies listed above focus almost exclusively on the corals and fishes which are more easily surveyed.--> Given the strong reported differences between deep and shallow reefs across taxa and the paucity of research on drivers of cryptic coral reef community variation, we sought to answer the following research questions: (1) How do shallow crab communities vary along the natural and anthropogenic gradients of the Hawaiian Archipelago?
+(2) Do shallow crab communities differ between the Main and Northwestern Hawaiian Islands?
+(3) What environmental and/or anthropogenic factors drive crab community structure in shallow coral reefs across the Hawaiian Archipelago?
+(4) What are the relative contributions of depth (shallow vs.
+mesophotic) and island region to crab community variation across the Hawaiian Archipelago?
+
+<!-- Here, we employed autonomous reef monitoring structures (ARMS)  Using ARMS, we sampled brachyuran crab communities across the Hawaiian Archipelago to test the effects of environmental and anthropogenic drivers on the structure of cryptic reef communities.
 Given the number and magnitude of differences between these regions, we hypothesized that species diversity would be different between regions (MHI vs NWHI).
@@
 Finally, to investigate whether a latitudinal gradient or depth gradient would be a stronger force in driving community composition, we compared the reef crab assemblage from across the shallow reefs in the Hawaiian Archipelago to reef crab assemblages across a mesophotic depth gradient.
-
-The Hawaiian Archipelago is a volcanic island arc made up of high islands and atolls spanning ~9 degrees of latitude and ~2,500 km.
+-->
+
+<!-- The Hawaiian Archipelago is a volcanic island arc made up of high islands and atolls spanning ~9 degrees of latitude and ~2,500 km.
 It is typically regarded as two regions: the inhabited main Hawaiian Islands (MHI), which include the eight islands from Hawai&#x02BB;i Island to Kauai&#x02BB;i, and the largely uninhabited Northwestern Hawaiian Islands (NWHI), which constitute the 10 low islands and atolls to the northwest of Kaua&#x02BB;i.
@@
 By controlling for habitat area and complexity and by incorporating sites across various environmental and human impact gradients, we sought to better understand whether latitude, the environment, or depth is more influential in structuring reef crab communities.
+-->
 
@@
 
-```{r fig-map, fig.cap='Map of ARMS deployment site across the Hawaiian Archipelago (Papahānaumokuākea Marine National Monument is shaded in light blue)', results="asis", fig.width=9, fig.height=6, dependson="init"}
+```{r fig-map, fig.cap='Map of ARMS deployment sites across the Hawaiian Archipelago (Papahānaumokuākea Marine National Monument is shaded in light blue).', results="asis", fig.width=9, fig.height=6, dependson="init"}
 library(sf)
@@
 All adult brachyuran crabs ($\geq$ 5 mm) were photographed, identified to morphospecies, and preserved in 95% ethanol.
-Individual preserved specimens were accessioned into the invertebrate zoology collections at the Florida Museum of Natural History (catalog numbers xxx–xxx).
+Individual preserved specimens were accessioned into the invertebrate zoology collections at the Florida Museum of Natural History ([catalog numbers xxx--xxx]{.mark}).
 
@@
 Then, we sequentially removed variables with the largest variance inflation factors (VIF; calculated using `vif.cca`) until all values were <3, indicating variables were sufficiently uncorrelated.
-Model significance was assessed using permutation tests with `r num(PERM,1)` permutations (function `anova`).
+Model significance was assessed using permutation tests with `r num(PERM,1)` permutations in the function `anova`.
 Continuous variables were scaled to unit variance.
@@
 Potential environmental drivers of shallow crab community structure were assembled from published literature [@selkoeDNACoralReef2016], NOAA's National Coral Reef Monitoring Program (NCRMP), and the Pacific Islands Ocean Observing System (PacIOOS) ([@tbl:env_table]).
-Due to logistical limitations, environmental factors are only available for shallow samples.
+Due to logistical limitations, environmental factors were only available for shallow samples.
 
@@
 Units contained `r num(all_summary$abundance_mean,0.1)` $\pm$ `r num(all_summary$abundance_sd,0.1)` (mean $\pm$ SD) individuals, and abundance ranged from `r format_range(all_summary$abundance_range)`.
-Crabs were identified to `r sum(unlist(all_summary$types_breakdown))` distinct types (including `r all_summary$types_breakdown$species` species and `r all_summary$types_breakdown$morphospecies` morphospecies) in `r all_summary$taxonomy_breakdown$genera` genera and `r all_summary$taxonomy_breakdown$families` families, which accounted for ~36% of the known crab diversity reported from Hawaiian waters [@castroCatalogAnomuranBrachyuran2011].
+Crabs were identified to `r sum(unlist(all_summary$types_breakdown))` distinct types (`r all_summary$types_breakdown$species` species and `r all_summary$types_breakdown$morphospecies` morphospecies) in `r all_summary$taxonomy_breakdown$genera` genera and `r all_summary$taxonomy_breakdown$families` families, which accounted for ~36% of the known crab diversity reported from Hawaiian waters [@castroCatalogAnomuranBrachyuran2011].
 The four most abundant species were: `r with(all_summary,top_list(top,top_n,top_units,fig="fig_crabs"))`.
@@
 In the CAP analysis, ARMS units were correctly assigned to sampling groups `r pc(cap_depth$percent/100,0.1)` of the time.
-The first two axes explained `r pc(cap_var$x,0.1)` and `r pc(cap_var$y,0.1)` of the variation, respectively.
-The $\geq$ 30 m samples from O&#x02BB;ahu grouped separately from the shallow samples collected across the entire ~2,500 km of the Hawaiian Islands ([@fig:fig_cap]).
-In addition, there was as much variability across the 30--90 m samples as across all shallow samples, with the 30 m samples spread out primarily along the LD2 (y) axis and the 60 m and 90 m samples forming a tight group ([@fig:fig_cap]).
+The first two linear discriminant axes explained `r pc(cap_var$x,0.1)` and `r pc(cap_var$y,0.1)` of the variation, respectively.
+The mesophotic samples ($\geq$ 30 m) from O&#x02BB;ahu grouped separately from the shallow samples collected across the entire ~2,500 km of the Hawaiian Islands ([@fig:fig_cap]).
+In addition, there was as much variability across the 30--90 m samples as across all shallow samples, with the 30 m samples spread along the LD2 (vertical) axis and the 60 m and 90 m samples grouped together ([@fig:fig_cap]).
 
@@
 
-```{r fig-alpha, fig.cap='Box plots of Simpson diversity by (A) island group and (B) sampling depth for complete dataset (all depths).
-Notches represent nonparametric confidence intervals about the median', results="asis", dev="svg", fig.width=10, fig.height=10, dependson="init"}
+```{r fig-alpha, fig.cap='Box plots of (A,B) species richness and (C,D) Simpson diversity for complete dataset (all sampled depths).
+(A) Species richness by island group.
+(B) Species richness by sampling depth.
+(C) Simpson diversity by island group.
+(D) Simpson diversity by sampling depth.
+Notches represent nonparametric confidence intervals about the median.', results="asis", dev="svg", fig.width=10, fig.height=10, dependson="init"}
 library(patchwork)
@@
 
-```{r fig-alpha2, fig.cap='Box plots of (A) species richness and (B) Simpson diversity and by island group for shallow samples only.', results="asis", dev="svg", fig.width=8, fig.height=4, dependson="init"}
+```{r fig-alpha2, fig.cap='Box plots of (A) species richness and (B) Simpson diversity by island group for shallow samples only.
+Notches as in [@fig:fig_alpha].', results="asis", dev="svg", fig.width=8, fig.height=4, dependson="init"}
 art <- shallow_alpha$richness_table %>% 
@@
 ```{r fig-pca, fig.cap='Principal component analysis (PCA) of environmental variables by sampling site.
-Ellipses represent 95% confidence intervals around island groups', results='asis', dev='svg', fig.width=8,fig.height=8, dependson='init'}
+Ellipses represent 95% confidence intervals around island groups.', results='asis', dev='svg', fig.width=8,fig.height=8, dependson='init'}
 cds <- cc$crab_data_shallow
@@
 
-```{r pairwise-island, fig.cap='Significance levels of pairwise PERMANOVA tests by island (FDR-adjusted p-values)', results="asis", dev="svg", fig.width=12, fig.height=9, dependson="init"}
+```{r pairwise-island, fig.cap='Significance levels of pairwise PERMANOVA tests by island for shallow crab communities (p-values adjusted according to the false discovery rate).', results="asis", dev="svg", fig.width=12, fig.height=9, dependson="init"}
 library(ggtext)

### 2023-11-06 manuscript.Rmd
various updates including integration of real chl/sst data

@@
 library(BiodiversityR)
+library(ggtext)
 ```
@@
 
-The Hawaiian Archipelago comprises high volcanic islands and atolls stretching across ~2,500 km in the North Pacific Ocean, and is typically regarded as two regions: the main Hawaiian Islands (MHI), which include the eight inhabited high islands, and the Northwestern Hawaiian Islands (NWHI), the uninhabited atolls and banks northwest of Kaua&#x02BB;i and Ni&#x02BB;ihau.
+The Hawaiian Archipelago comprises high volcanic islands and atolls stretching across ~2,500 km in the North Pacific Ocean, and is typically regarded as two regions: the main Hawaiian Islands (MHI), which include the eight inhabited high islands, and the Northwestern Hawaiian Islands (NWHI), the uninhabited atolls and banks northwest of Kaua&#x02BB;i and Ni&#x02BB;ihau that make up the Papahānaumokuākea Marine National Monument.
 The islands are subject to anthropogenic and physical gradients across their span [@selkoeMapHumanImpacts2009], which we hypothesize to influence coral reef communities.
@@
 Further, there are a range of connections depending on the taxa – large, mobile organisms like fish tend to have more shared species across depth than sedentary organisms such as corals and algae, but even among the fishes the community structure of MCEs is often distinct [@kaneHighLevelsMesophotic2014; @pyleComprehensiveInvestigationMesophotic2016 ; @asherMesophoticDepthGradients2017; @fukunagaChangesMesophoticReef2017; @colemanShiftingReefFish2018; @pinheiroDeepReefFishes2019].
-Although the majority of coral reef biodiversity consists of the algal and invertebrate cryptobenthic organisms living on and within the reef structure [@fautinOverviewMarineBiodiversity2010; @knowltonCoralReefBiodiversity2010; @fisherSpeciesRichnessCoral2015], the studies listed above focus almost exclusively on the corals and fishes which are more easily surveyed.--> Given the strong reported differences between deep and shallow reefs across taxa and the paucity of research on drivers of cryptic coral reef community variation, we sought to answer the following research questions: (1) How do shallow crab communities vary along the natural and anthropogenic gradients of the Hawaiian Archipelago?
-(2) Do shallow crab communities differ between the Main and Northwestern Hawaiian Islands?
-(3) What environmental and/or anthropogenic factors drive crab community structure in shallow coral reefs across the Hawaiian Archipelago?
-(4) What are the relative contributions of depth (shallow vs.
+Although the majority of coral reef biodiversity consists of the algal and invertebrate cryptobenthic organisms living on and within the reef structure [@fautinOverviewMarineBiodiversity2010; @knowltonCoralReefBiodiversity2010; @fisherSpeciesRichnessCoral2015], the studies listed above focus almost exclusively on the corals and fishes which are more easily surveyed.--> Given the strong reported differences between deep and shallow reefs across taxa and the paucity of research on drivers of cryptic coral reef community variation, we sought to answer the following research questions: (i) How do shallow crab communities vary along the natural and anthropogenic gradients of the Hawaiian Archipelago?
+(ii) Do shallow crab communities differ between the Main and Northwestern Hawaiian Islands?
+(iii) What environmental and/or anthropogenic factors drive crab community structure in shallow coral reefs across the Hawaiian Archipelago?
+(iv) What are the relative contributions of depth (shallow vs.
 mesophotic) and island region to crab community variation across the Hawaiian Archipelago?
@@
 
-```{r fig-map, fig.cap='Map of ARMS deployment sites across the Hawaiian Archipelago (Papahānaumokuākea Marine National Monument is shaded in light blue).', results="asis", fig.width=9, fig.height=6, dependson="init"}
+```{r fig-map, fig.cap='Map of ARMS deployment sites across the Hawaiian Archipelago (Papahānaumokuākea Marine National Monument (NWHI) is shaded in light blue).', results="asis", fig.width=9, fig.height=6, dependson="init"}
 library(sf)
@@
 All adult brachyuran crabs ($\geq$ 5 mm) were photographed, identified to morphospecies, and preserved in 95% ethanol.
-Individual preserved specimens were accessioned into the invertebrate zoology collections at the Florida Museum of Natural History ([catalog numbers xxx--xxx]{.mark}).
+Specimens were collected under permits [#xxx]{.mark} (NWHI) and [#xxx]{.mark} (MHI) and accessioned into the invertebrate zoology collections at the Florida Museum of Natural History ([catalog numbers xxx--xxx]{.mark}).
 
@@
 
-Potential environmental drivers of shallow crab community structure were assembled from published literature [@selkoeDNACoralReef2016], NOAA's National Coral Reef Monitoring Program (NCRMP), and the Pacific Islands Ocean Observing System (PacIOOS) ([@tbl:env_table]).
-Due to logistical limitations, environmental factors were only available for shallow samples.
+Potential environmental drivers of shallow crab community structure were assembled from a study on factors affecting population genetics in Hawai&#x02BB;i, [@selkoeDNACoralReef2016], NOAA's National Coral Reef Monitoring Program (NCRMP), the Pacific Islands Ocean Observing System (PacIOOS) ([@tbl:env_table]), NASA Physical Oceanograpy Distributed Active Archive Center (PO.DAAC), and NASA Ocean Biology Distributed Active Archive Center (OB.DAAC).
+For details on how sea-surface temperature and chlorophyll data were compiled and aggregated, see Supplemental meterial.
+Due to logistical limitations, environmental factors were collected only for shallow samples ($\le$ 30 m).
 
 ```{r env-table,results="asis", dependson="init"}
-print_table(model_terms,"env_table","Potential environmental drivers of crab community structure, assembled from PacIOOS and @selkoeDNACoralReef2016",hide="variable")
+print_table(model_terms,"env_table","Potential environmental drivers of crab community structure across the Hawaiian Archipelago",hide="variable")
 ```
@@
 ![The four most abundant crab species found across all deployed ARMS units.
-(A) *Chlorodiella laevissima*.
-(B) *Perinia tumida*.
-(C) *Percnon abbreviatum*.
-(D) *Dynomene hispida*.
+(a) *Chlorodiella laevissima*.
+(b) *Perinia tumida*.
+(c) *Percnon abbreviatum*.
+(d) *Dynomene hispida*.
 Photo credit: Florida Museum of Natural History, licensed [CC BY-NC 3.0](https://creativecommons.org/licenses/by-nc/3.0/)](data/images/4crabs.png){#fig:fig_crabs}
@@
 
-db_upr <- dbrda(cc$crab_otus_shallow ~ depth + chl + sst + slope + coral_cover + closest_island + larval_connectivity + human_impact,data=cc$crab_data_shallow,distance="bray")
+# db_upr <- dbrda(cc$crab_otus_shallow ~ depth + chl + sst + slope + coral_cover + closest_island + larval_connectivity + human_impact,data=cc$crab_data_shallow,distance="bray")
+db_upr <- dbrda(cc$crab_otus_shallow ~ depth + chl + sst + slope + coral_cover + larval_connectivity + human_impact,data=cc$crab_data_shallow,distance="bray")
 db_lwr <- dbrda(cc$crab_otus_shallow  ~ 1,data=cc$crab_data_shallow,distance="bray")
@@
 # get rid of variables with VIF > 3
+# db_modl <- reduce_model(db_modl,vif_limit=8)
 db_modl <- reduce_model(db_modl,vif_limit=3)
@@
 Environmental variables clustered strongly according to island group (MHI vs NWHI), with the difference driven primarily by human impact, coral cover, and island slope ([@fig:fig_pca]).
-The best dbRDA model (`r permanova(db_modl,stat="F")`) explained `r pc(RsquareAdj(db_modl)$adj.r.squared,0.01)` (adjusted $\mathrm{R}^2$) of the variability in crab communities across the archipelago ([@fig:fig_rda]).
+The best dbRDA model (`r permanova(db_modl,stat="F")`) explained `r pc(RsquareAdj(db_modl)$adj.r.squared,0.01)` (adjusted $\mathrm{R}^2$) of the variability in crab communities across the archipelago ([@fig:fig_dbrda]).
 `r str_to_sentence(english(nterm))` of the 10 environmental variables tested were significant and sufficiently uncorrelated.
@@
 
-```{r fig-rda, fig.cap="Distance-based redundancy analysis (dbRDA) plot showing significant environmental drivers of crab community variation across the Hawaiian Islands.", results="asis", fig.width=9, fig.height=6, dependson="rda-analysis"}
+```{r fig-dbrda, fig.cap="dbRDA plot showing significant drivers of crab community variation across the Hawaiian Islands.
+Ellipses represent 95% confidence intervals around groups.", results="asis", fig.width=9, fig.height=6, dependson="rda-analysis"}
 termz <- model_terms %>%
@@
   xlab(axis_var[1]) + ylab(axis_var[2])
-cat("{#fig:fig_rda}")
+cat("{#fig:fig_dbrda}")
 ```
@@
 
-```{r fig-cap, fig.cap="Canonical analysis of principal coordinates (CAP) plot showing variation in brachyuran crab communities by island group and depth zone.
-Ellipses represent 95% confidence intervals around sample groups.", results="asis", fig.width=9, fig.height=6, dependson="cap-analysis"}
+```{r fig-cap, fig.cap="CAP plot of crab communities by island group and depth zone.", results="asis", fig.width=9, fig.height=6, dependson="cap-analysis"}
 # make the sample group look nice
@@
 
-```{r fig-alpha, fig.cap='Box plots of (A,B) species richness and (C,D) Simpson diversity for complete dataset (all sampled depths).
-(A) Species richness by island group.
-(B) Species richness by sampling depth.
-(C) Simpson diversity by island group.
-(D) Simpson diversity by sampling depth.
+```{r fig-alpha, fig.cap='Box plots of (a,b) species richness and (c,d) Simpson diversity for complete dataset (all sampled depths).
+(a) Species richness by island group.
+(b) Species richness by sampling depth.
+(c) Simpson diversity by island group.
+(d) Simpson diversity by sampling depth.
 Notches represent nonparametric confidence intervals about the median.', results="asis", dev="svg", fig.width=10, fig.height=10, dependson="init"}
@@
 
-```{r fig-alpha2, fig.cap='Box plots of (A) species richness and (B) Simpson diversity by island group for shallow samples only.
+```{r fig-alpha2, fig.cap='Box plots of (a) species richness and (b) Simpson diversity by island group for shallow samples only.
 Notches as in [@fig:fig_alpha].', results="asis", dev="svg", fig.width=8, fig.height=4, dependson="init"}
@@
 pcdata <- cds %>%
-  select(where(is.numeric))
+  select(where(is.numeric),-c(lon,lat))
 
@@
   xlab(str_glue("PC1 [{pc(explained[1])}]")) + 
-  ylab(str_glue("PC2 [{pc(explained[2])}]"))
+  ylab(str_glue("PC2 [{pc(explained[2])}]")) + 
+  plotz_theme()
 cat("{#fig:fig_pca}")

### 2023-11-06 chl-sst.R

@@ -1,0 +1,171 @@
+library(tidyverse)
+library(sf)
+library(raster)
+library(here)
+library(lubridate)
+library(httr)
+library(fs)
+
+
+## download SST and CHL data from NASA and aggregate it by ARMS sampling site
+
+
+# pre-parse metadata and get deployment & recovery years as well as deployment range
+arms <- read_csv(here("data","metadata.csv")) %>%
+  rename(sample=unit) %>%
+  st_as_sf(coords=c("lon","lat"),crs=4326,remove=FALSE) %>%
+  separate_wider_regex(sample,patterns=c(recovery_year="^[0-9]+","_",sample=".*"),cols_remove=FALSE)  %>%
+  mutate(recovery_year=as.numeric(recovery_year)) %>%
+  dplyr::select(sample,everything()) %>%
+  unite("range",deployment_year,recovery_year,sep="--",remove=FALSE) 
+
+
+# sample 30 months from each date range
+dates <- arms %>%
+  group_by(range) %>%
+  group_map(\(data,grp) {
+    start_date <- str_glue("{min(data$deployment_year)}-01-01") %>%
+      as.Date()
+    end_date <- str_glue("{max(data$recovery_year)}-12-31") %>%
+      as.Date()
+    
+    tibble(
+      range = grp$range,
+      date =  seq(start_date,end_date,by="month") %>%
+        sample(30) %>%
+        as.character()
+    )
+  }) %>%
+  bind_rows()
+
+# chlorophyll and sst datasets
+datasets <- c(chl="MY1DMM_CHLORA", sst="MYD28M")
+
+
+# url templates for getting image IDs and GeoTIFF data
+id_url="https://neo.gsfc.nasa.gov/lib/imageSizes.php?datasetId={dataset}&curDate={timestamp}"
+tiff_url="https://neo.gsfc.nasa.gov/servlet/RenderData?si={tiff_id}&cs=rgb&format=FLOAT.TIFF&width=3600&height=1800"
+
+# get unix timestamp
+ts <- function(d) {
+  as.numeric(as.POSIXct(d,tz="HST"))
+}
+
+# retrieve a URL
+get_url <- function(u) {
+  r <- GET(u)
+  if (r$status_code == 200) {
+    return(content(r,as="text",encoding="UTF-8"))
+  } else {
+    return(NULL)
+  }
+}
+
+# make sure directories exist
+datasets %>%
[…]

### 2023-11-06 crabs.R

@@ -120,7 +120,8 @@ plot_ord <- function(ps,ord,arrows=TRUE,sig=FALSE,by="term",perm=999,alpha=0.05,
       ) + 
       # geom_text_repel(
       # geom_text(
-      geom_label(
+      # geom_label(
+      geom_richtext(
         mapping = label_map, 
         size = labelsize,  
         data = arrowdf, 
@@ -320,7 +321,8 @@ setup_crabs <- function() {
   # we care about and scale the numeric ones to unit variance
   cc$crab_data_shallow <- cc$crabs_shallow %>%
     sample_tibble(sample_col = "unit") %>%
-    select(unit,region,island_group,island,lat,lon,depth,chl,sst,slope,coral_cover,closest_island,larval_connectivity,human_impact) %>%
+    select(-sst,-chl) %>%
+    select(unit,region,island_group,island,lat,lon,depth,chl=chl_sat,sst=sst_sat,slope,coral_cover,closest_island,larval_connectivity,human_impact) %>%
     mutate(across(where(is.numeric),~as.numeric(scale(.x)))) %>%
     column_to_rownames("unit") 
     # reassociate the new sample data

### 2023-11-07 manuscript.Rmd
various updates

@@
 library(ggtext)
+
+# set random seed for consistency
+set.seed(31337)
 ```
@@
 Seawater surrounding disassembled units was sieved through 2 mm, 500 μm, and 100 μm mesh-size geologic sieves [see @lerayDNABarcodingMetabarcoding2015 for processing details], and retained macroorganisms were preserved in 95% ethanol.
-All adult brachyuran crabs ($\geq$ 5 mm) were photographed, identified to morphospecies, and preserved in 95% ethanol.
+All adult brachyuran crabs ($\geq$ 5 mm) were photographed, identified (to species where possible, otherwise morphospecies), and preserved in 95% ethanol.
 Specimens were collected under permits [#xxx]{.mark} (NWHI) and [#xxx]{.mark} (MHI) and accessioned into the invertebrate zoology collections at the Florida Museum of Natural History ([catalog numbers xxx--xxx]{.mark}).
@@
 
-Potential environmental drivers of shallow crab community structure were assembled from a study on factors affecting population genetics in Hawai&#x02BB;i, [@selkoeDNACoralReef2016], NOAA's National Coral Reef Monitoring Program (NCRMP), the Pacific Islands Ocean Observing System (PacIOOS) ([@tbl:env_table]), NASA Physical Oceanograpy Distributed Active Archive Center (PO.DAAC), and NASA Ocean Biology Distributed Active Archive Center (OB.DAAC).
-For details on how sea-surface temperature and chlorophyll data were compiled and aggregated, see Supplemental meterial.
-Due to logistical limitations, environmental factors were collected only for shallow samples ($\le$ 30 m).
+Potential environmental drivers of shallow crab community structure were assembled a previous study on environmental drivers of genetic diversity [@selkoeDNACoralReef2016], NOAA's National Coral Reef Monitoring Program (NCRMP), the Pacific Islands Ocean Observing System (PacIOOS), NASA Physical Oceanography Distributed Active Archive Center (PO.DAAC), and NASA Ocean Biology Distributed Active Archive Center (OB.DAAC) ([@tbl:env_table]).
+For details on how sea-surface temperature and chlorophyll data were compiled and aggregated, see Supplemental material.
+Due to logistical limitations, environmental data were collected only for shallow samples ($\le$ 30 m).
 
@@
 
-For the complete dataset, there was no significant difference in observed species richness between the MHI and NWHI ([@fig:fig_alpha[]{}A]; `r k_test(all_alpha$k_r_island)`), but Simpson diversity was significantly different ([@fig:fig_alpha[]{}C]; `r k_test(all_alpha$k_s_island,plain=TRUE)`).
-Richness was not significantly different between shallow and deep samples ([@fig:fig_alpha[]{}B]; `r k_test(all_alpha$k_r_depth,plain=TRUE)`) but Simpson diversity was ([@fig:fig_alpha[]{}D]; `r k_test(all_alpha$k_s_depth,plain=TRUE)`).
-When only shallow arms \< 30 m were considered, neither comparison was significant (richness: `r k_test(shallow_alpha$k_r_island,plain=TRUE)`, [@fig:fig_alpha2[]{}A]; Simpson: `r k_test(shallow_alpha$k_s_island,plain=TRUE)`, [@fig:fig_alpha2[]{}B]).
-
-<!--For the complete dataset, there was no significant difference in observed species richness between the MHI and NWHI `r w_test(all_alpha$w_r_island)`, but Simpson diversity was significantly different ([@fig:fig_alpha[]{}A]; `r w_test(all_alpha$w_s_island,plain=TRUE)`).
+For the complete dataset, there was no significant difference in observed species richness between the MHI and NWHI ([@fig:fig_alpha[]{}a]; `r k_test(all_alpha$k_r_island)`), but Simpson diversity was significantly different ([@fig:fig_alpha[]{}c]; `r k_test(all_alpha$k_s_island,plain=TRUE)`).
+Richness was not significantly different between shallow and deep samples ([@fig:fig_alpha[]{}b]; `r k_test(all_alpha$k_r_depth,plain=TRUE)`) but Simpson diversity was ([@fig:fig_alpha[]{}d]; `r k_test(all_alpha$k_s_depth,plain=TRUE)`).
+When only shallow arms \< 30 m were considered, neither comparison was significant (richness: `r k_test(shallow_alpha$k_r_island,plain=TRUE)`, [@fig:fig_alpha2[]{}a]; Simpson: `r k_test(shallow_alpha$k_s_island,plain=TRUE)`, [@fig:fig_alpha2[]{}b]).
+
+<!--For the complete dataset, there was no significant difference in observed species richness between the MHI and NWHI `r w_test(all_alpha$w_r_island)`, but Simpson diversity was significantly different ([@fig:fig_alpha[]{}a]; `r w_test(all_alpha$w_s_island,plain=TRUE)`).
 When only shallow arms \< 30 m were considered, neither comparison was significant (richness: `r w_test(shallow_alpha$w_r_island,plain=TRUE)`; Simpson: `r w_test(shallow_alpha$w_s_island,plain=TRUE)`).
-Simpson diversity was significantly different between shallow and deep samples ([@fig:fig_alpha[]{}B]; `r w_test(all_alpha$w_s_depth,plain=TRUE)`) but richness was not (`r w_test(all_alpha$w_r_depth,plain=TRUE)`).
+Simpson diversity was significantly different between shallow and deep samples ([@fig:fig_alpha[]{}b]; `r w_test(all_alpha$w_s_depth,plain=TRUE)`) but richness was not (`r w_test(all_alpha$w_r_depth,plain=TRUE)`).
 -->
 
-<!--For the complete dataset, there was no significant difference in observed species richness between the MHI and NWHI `r t_test(all_alpha$t_r_island)`, but Simpson diversity was significantly different ([@fig:fig_alpha[]{}A]; `r t_test(all_alpha$t_s_island,plain=TRUE)`).
+<!--For the complete dataset, there was no significant difference in observed species richness between the MHI and NWHI `r t_test(all_alpha$t_r_island)`, but Simpson diversity was significantly different ([@fig:fig_alpha[]{}a]; `r t_test(all_alpha$t_s_island,plain=TRUE)`).
 When only shallow arms \< 30 m were considered, neither comparison was significant (richness: `r t_test(shallow_alpha$t_r_island,plain=TRUE)`; Simpson: `r t_test(shallow_alpha$t_s_island,plain=TRUE)`).
-Simpson diversity was significantly different between shallow and deep samples ([@fig:fig_alpha[]{}B]; `r t_test(all_alpha$t_s_depth,plain=TRUE)`) but richness was not (`r t_test(all_alpha$t_r_depth,plain=TRUE)`).
+Simpson diversity was significantly different between shallow and deep samples ([@fig:fig_alpha[]{}b]; `r t_test(all_alpha$t_s_depth,plain=TRUE)`) but richness was not (`r t_test(all_alpha$t_r_depth,plain=TRUE)`).
 -->
@@
 
-```{r rda-analysis, results="hide", dependson="init"}
+```{r dbrda-analysis, results="hide", dependson="init"}
 library(formula.tools)
@@
 
-Environmental variables clustered strongly according to island group (MHI vs NWHI), with the difference driven primarily by human impact, coral cover, and island slope ([@fig:fig_pca]).
+Environmental variables were strongly clustered according to island group (MHI vs NWHI), with the separation primarily along the direction influenced by human impact, coral cover, sea-surface temperature, and chlorophyll ([@fig:fig_pca]).
 The best dbRDA model (`r permanova(db_modl,stat="F")`) explained `r pc(RsquareAdj(db_modl)$adj.r.squared,0.01)` (adjusted $\mathrm{R}^2$) of the variability in crab communities across the archipelago ([@fig:fig_dbrda]).
-`r str_to_sentence(english(nterm))` of the 10 environmental variables tested were significant and sufficiently uncorrelated.
-The drivers contributing most to explained variation were `r drivers$term[1]` (F = `r num(drivers$stat[1],0.01)`) and `r drivers$term[2]` (F = `r num(drivers$stat[2],0.01)`) ([@tbl:rda_table]).
+`r str_to_sentence(english(nterm))` of the seven environmental variables tested were significant and sufficiently uncorrelated.
+The two drivers contributing most to explained variation were `r drivers$term[1]` (F = `r num(drivers$stat[1],0.01)`) and `r drivers$term[2]` (F = `r num(drivers$stat[2],0.01)`) ([@tbl:rda_table]).
 
 ```{r fig-dbrda, fig.cap="dbRDA plot showing significant drivers of crab community variation across the Hawaiian Islands.
-Ellipses represent 95% confidence intervals around groups.", results="asis", fig.width=9, fig.height=6, dependson="rda-analysis"}
+Colored ellipses represent 95% confidence intervals around island groups.", results="asis", fig.width=9, fig.height=6, dependson="rda-analysis"}
 termz <- model_terms %>%
@@
 ```
-<!-- 
-Here's the thing: it's not that the differences between MHI & NWHI become non-significant with the inclusion of deep samples, 
-it's just that the deep samples are MORE different from all the shallow samples than all the shallow samples are from one another?
-We can think of PERMANOVA R^2 and ANOSIM R as effect sizes. Thus, it may be a bit weak but for shallow samples only, island group 
-R2 = 0.072, for all samples R2 = 0.065. For depth (all samples of course), R2 = 0.110. If we consider the effect sizes from ANOSIM, 
-for shallow samples, island group R = 0.175, for all samples, island group R = 0.161, and for shallow vs deep,  R = 0.509. 
-The apparent difference in effect size is much more pronounced in the ANOSIM, but it's different in each case, and it's smaller
-for all samples than it is for just shallow samples.
--->
-
 Shallow crab community structure was significantly different between island groups (MHI vs.
@@
 However, community dispersion was not different among depth zones $\geq$ 30 m (`r permdisp(bd_deep)`).
-Although crab communities varied significantly among both depth zones and island groups when examining the complete dataset (`r permanova(anova_both,0.01)`), depth explained `r num(effect_ratio,0.1)` times more variability than island group (based on model $\mathrm{R}^2$).
-In the CAP analysis, ARMS units were correctly assigned to sampling groups `r pc(cap_depth$percent/100,0.1)` of the time.
-The first two linear discriminant axes explained `r pc(cap_var$x,0.1)` and `r pc(cap_var$y,0.1)` of the variation, respectively.
-The mesophotic samples ($\geq$ 30 m) from O&#x02BB;ahu grouped separately from the shallow samples collected across the entire ~2,500 km of the Hawaiian Islands ([@fig:fig_cap]).
-In addition, there was as much variability across the 30--90 m samples as across all shallow samples, with the 30 m samples spread along the LD2 (vertical) axis and the 60 m and 90 m samples grouped together ([@fig:fig_cap]).
+Although crab communities were significantly different among both depth zones and island groups for the complete dataset (`r permanova(anova_both,0.01)`), depth explained `r num(effect_ratio,0.1)` times more variability than island group (based on model $\mathrm{R}^2$).
+
+In the CAP analysis of island group and depth zone, ARMS units were correctly reassigned to groups `r pc(cap_depth$percent/100,0.1)` of the time.
+The first two axes explained `r pc(cap_var$x,0.1)` and `r pc(cap_var$y,0.1)` of the variation, respectively.
+Mesophotic samples ($\geq$ 30 m) from O&#x02BB;ahu grouped separately from shallow samples collected across the entire ~2,500 km expanse of the archipelago ([@fig:fig_cap]).
+Additionally, variability among 30--90 m samples was equivalent to that among all shallow samples, although that was largely due to differences between 30 m and deeper sample groups ([@fig:fig_cap]).
 
@@
 # Discussion {-}
+The Hawaiian Archipelago is among the most remote on the planet and encompasses a range of gradients (anthropogenic, biotic, abiotic), making it an ideal area to study the impacts of these factors on species assemblages across space.
+Human impacts are considerably higher in the populated MHI than in the protected and remote NWHI [@selkoeEvaluatingAnthropogenicThreats2008; @selkoeMapHumanImpacts2009; @selkoeDNACoralReef2016].
+Although we did not consider individual effects here, an index representing cumulative anthropogenic impacts was the top driver of brachyuran community structure, similar to previous work examining genetic diversity across the archipelago [@selkoeDNACoralReef2016].
+The next-most significant driver, chlorophyll a, can be used as a proxy for primary productivity.
+Productivity gradients have been reported across the Hawaiian Archipelago, which coincide with differences in coral reef community composition [@selkoeDNACoralReef2016], endemic reef fish dominance [@friedlanderDominanceEndemicsReef2020], and survival rates of top predators [@bakerEffectVariableOceanic2007].
+It is not surprising, therefore, that we find differences in the brachyuran crab communities are driven by these same factors across the Hawaiian archipelago.
 
@@
 # Supplemental material {label="S"}
+
+## Processing of satellite-derived SST and chl-A data
+Sea-surface temperature (SST) and chlorophyll A (chl-A) data were retrieved from NASA's Physical Oceanography Distributed Active Archive Center [PO.DAAC; @NASAJPL_2020] and Ocean Biology Distributed Active Archive Center [OB.DAAC; @NASAOceanBiologyProcessingGroup_2022] respectively.
+Monthly summary datasets were downloaded for 30 randomly subsampled months within each ARMs unit's three-year soaking period, averaged, and values (SST or chl-A) were extracted for the geographic coordinates of each sampling site, with a 300 m surrounding buffer zone.
 
@@
 pcdata <- cds %>%
-  select(where(is.numeric),-c(lon,lat))
+  select(where(is.numeric),-c(lon,lat),-closest_island)
 
@@
 
-```{r pairwise-island, fig.cap='Significance levels of pairwise PERMANOVA tests by island for shallow crab communities (p-values adjusted according to the false discovery rate).', results="asis", dev="svg", fig.width=12, fig.height=9, dependson="init"}
+```{r pairwise-island, fig.cap='Significance levels of pairwise PERMANOVA tests by island for shallow crab communities (p-values adjusted according to the false discovery rate [@benjaminiControllingFalseDiscovery1995]).', results="asis", dev="svg", fig.width=12, fig.height=9, dependson="init"}
 library(ggtext)

### 2023-11-07 chl-sst.R

@@ -6,6 +6,9 @@ library(lubridate)
 library(httr)
 library(fs)
 
+# set random seed for consistency
+set.seed(31337)
+
 
 ## download SST and CHL data from NASA and aggregate it by ARMS sampling site
 

### 2023-11-08 manuscript.Rmd
updates

@@
 
+Despite the fact that shallow ARMS units were deployed within a narrow depth range (8--17 m), depth still emerged as a significant driver of community variation.
+Depth in and of itself does not affect anything [[cite?]]{.mark}, but at the same time, the factors that are known to covary most consistently with it (viz.
+temparature and light) do not show much variation across such a narrow range.
+However, we also did not sample these values for the specific depths where units where deployed.
+So, depth is clearly doing something.
+We only have environmental data from units deployed \< 30 m, but community ordinations show that there are considerable differences between all the shallow units across the entire span of the archipelago and the $\geq$ 30 m units collected from a single island.
+
 # References {-}

### 2023-11-08 crabs.R

@@ -317,7 +317,7 @@ setup_crabs <- function() {
     drop_na()
     # make sure we drop samples that had NAs in their sample data
   cc$crabs_shallow <- prune_samples(sample_names(cc$crabs_shallow) %in% cc$crab_data_shallow$unit,cc$crabs_shallow)
-    # now we select the columns containing the environmental variables
+  # now we select the columns containing the environmental variables
   # we care about and scale the numeric ones to unit variance
   cc$crab_data_shallow <- cc$crabs_shallow %>%
     sample_tibble(sample_col = "unit") %>%

### 2024-01-05 manuscript.Rmd
a bunch of updates

@@
   if (!is.null(fig)) {
-    s2 <- map2_chr(s2,LETTERS[seq(length(s2))],~str_glue(.x))
+    s2 <- map2_chr(s2,letters[seq(length(s2))],~str_glue(.x))
   }
@@
 
-The structure and composition of coral reef communities is influenced by a suite of environmental and anthropogenic factors, yet their nature and makeup remain poorly resolved.
-Coral reefs are among the most diverse habitats in the world, yet they face increasing threats of degradation, including climate-driven bleaching and phase shifts to alternate ecological regimes [@grahamManagingResilienceReverse2013; @norstromAlternativeStatesCoral2009; @pandolfiGlobalTrajectoriesLongTerm2003].
+The structure and composition of coral reef communities is influenced by a suite of environmental and anthropogenic factors, yet the nature and makeup of those factors remain poorly resolved.
+Coral reefs are among the most diverse habitats in the world, and they face increasing threats of degradation, including climate-driven bleaching and phase shifts to alternate ecological regimes [@grahamManagingResilienceReverse2013; @norstromAlternativeStatesCoral2009; @pandolfiGlobalTrajectoriesLongTerm2003], among others.
 In the face of these and other threats, understanding drivers of community structure is critically important [@hughesRisingChallengeSustaining2010; @pageSeekingResistanceCoral2019].
@@
 These habitats, known as mesophotic coral ecosystems (MCEs), are typically defined as those extending from ~30--150 m depth [@hindersteinThemeSectionMesophotic2010; @kahngCommunityEcologyMesophotic2010].
-<!--and were believed to experience dampened disturbances in comparison to shallow reefs and could therefore act as deep water refugia for shallow reef species threatened by anthropogenic impacts --> They have been hypothesized as possible refugia from disturbances on shallow reefs [@bongaertsAssessingDeepReef2010; @smithDepthRefugiumCatastrophic2014; @holsteinFertileFathomsDeep2015; @muirSpeciesIdentityDepth2017], although a number of studies have found evidence that deep reefs are unique and uniquely threatened, challenging this notion [@diazLightTemperatureDrive; @diazMesophoticCoralBleaching2023; @hobanPlumbingDepthsEnvironmental2023; @hurleyAssessmentShallowMesophotic2016; @kaneHighLevelsMesophotic2014; @pinheiroDeepReefFishes2019; @rochaMesophoticCoralEcosystems2018].
+<!--and were believed to experience dampened disturbances in comparison to shallow reefs and could therefore act as deep water refugia for shallow reef species threatened by anthropogenic impacts --> They have been hypothesized as possible refugia from disturbances on shallow reefs [@glynnCoralReefBleaching1996; @bongaertsAssessingDeepReef2010; @smithDepthRefugiumCatastrophic2014; @holsteinFertileFathomsDeep2015; @muirSpeciesIdentityDepth2017], although studies have found evidence that deep reefs are unique and uniquely threatened, challenging this notion [@diazLightTemperatureDrive; @diazMesophoticCoralBleaching2023; @hobanPlumbingDepthsEnvironmental2023; @hurleyAssessmentShallowMesophotic2016; @kaneHighLevelsMesophotic2014; @pinheiroDeepReefFishes2019; @rochaMesophoticCoralEcosystems2018].
 <!--However, recent literature describing species compositions along depth gradients reveal a more complex picture among taxa and locations because some groups may use mesophotic reefs as a refuge in some places, while others show community turnover with depth such that MCE habitats are ecologically distinct from adjacent shallow water coral reefs [@hurleyAssessmentShallowMesophotic2016; @smithCaribbeanMesophoticCoral2016; @semmlerAreMesophoticCoral2017; @bongaertsDeepReefsAre2017; @rochaMesophoticCoralEcosystems2018].
@@
 ![An autonomous reef monitoring structure (ARMS) unit deployed on a reef in the Red Sea (photo credit: Michael Berumen, used with permission)](data/images/arms_example.png){#fig:fig_arms}
+
+<br><br>
 
@@
 ggplot() +
-  geom_sf(data=hawaii %>% st_transform(transformer), color="black", fill="green", linewidth=0.4) +
+  geom_sf(data=hawaii %>% st_transform(transformer), color="black", fill="#A3866A", linewidth=0.4) +
   geom_sf(data=pmnm,color=NA,fill="#6ed8f6",alpha=0.2) +
@@
 ```
+
+<br><br>
 
@@
 
+<br><br>
+
 ARMS units were retrieved by encapsulating them in-situ in a mesh-lined crate (100 μm pore size) to prevent escape of motile organisms.
@@
 Then, we sequentially removed variables with the largest variance inflation factors (VIF; calculated using `vif.cca`) until all values were <3, indicating variables were sufficiently uncorrelated.
-Model significance was assessed using permutation tests with `r num(PERM,1)` permutations in the function `anova`.
-Continuous variables were scaled to unit variance.
+Model significance was assessed using permutation tests with `r num(PERM,1)` permutations with the function `anova`.
+All continuous variables were scaled to unit variance.
+
 We assessed variation among groups using `betadisper` and `adonis2`.
@@
 
-Potential environmental drivers of shallow crab community structure were assembled a previous study on environmental drivers of genetic diversity [@selkoeDNACoralReef2016], NOAA's National Coral Reef Monitoring Program (NCRMP), the Pacific Islands Ocean Observing System (PacIOOS), NASA Physical Oceanography Distributed Active Archive Center (PO.DAAC), and NASA Ocean Biology Distributed Active Archive Center (OB.DAAC) ([@tbl:env_table]).
+Potential environmental drivers of shallow crab community structure were assembled from a previous study on environmental drivers of genetic diversity [@selkoeDNACoralReef2016], NOAA's National Coral Reef Monitoring Program (NCRMP), the Pacific Islands Ocean Observing System (PacIOOS), NASA Physical Oceanography Distributed Active Archive Center (PO.DAAC), and NASA Ocean Biology Distributed Active Archive Center (OB.DAAC) ([@tbl:env_table]).
 For details on how sea-surface temperature and chlorophyll data were compiled and aggregated, see Supplemental material.
@@
 Photo credit: Florida Museum of Natural History, licensed [CC BY-NC 3.0](https://creativecommons.org/licenses/by-nc/3.0/)](data/images/4crabs.png){#fig:fig_crabs}
+
+<br><br>
 
@@
 The two drivers contributing most to explained variation were `r drivers$term[1]` (F = `r num(drivers$stat[1],0.01)`) and `r drivers$term[2]` (F = `r num(drivers$stat[2],0.01)`) ([@tbl:rda_table]).
+<!--TODO: perhaps include discussion of the vector lengths here as well, since chl-a is the shortest vector despite being the second most-significant factor.-->
 
@@
 
+<br><br>
+
 ### Regional & depth comparisons {-}
@@
 The first two axes explained `r pc(cap_var$x,0.1)` and `r pc(cap_var$y,0.1)` of the variation, respectively.
-Mesophotic samples ($\geq$ 30 m) from O&#x02BB;ahu grouped separately from shallow samples collected across the entire ~2,500 km expanse of the archipelago ([@fig:fig_cap]).
-Additionally, variability among 30--90 m samples was equivalent to that among all shallow samples, although that was largely due to differences between 30 m and deeper sample groups ([@fig:fig_cap]).
+Mesophotic samples (30--90 m) from O&#x02BB;ahu grouped separately from shallow samples collected across the entire ~2,500 km expanse of the archipelago ([@fig:fig_cap]).
+Additionally, variability among 30--90 m samples was visually equivalent to that among all shallow samples, although that was largely due to differences between 30 m and deeper sample groups ([@fig:fig_cap]).
 
@@
 
+<br><br>
+
 # Discussion {-}
+Marine communities are structured by a suite of factors, but there remains no consensus for generalized mechanisms for coral reefs<!-- structuring communities in shallow coral reefs-->.
+Classic studies from the intertidal point to biotic factors determining lower limits and abiotic factors setting upper limits of vertical species distributions [@connellEffectsCompetitionPredation1961; @connellInfluenceInterspecificCompetition1961; @dennyPhysicalProcessesThat2001; @paineIntertidalCommunityStructure1974].
+However, these conclusions describe distributions at individual sites and generalizations about the mechanisms and ecological processes that structure coral reef communities remain a subject of debate [@bellwoodEnvironmentalGeometricConstraints2005; @bellwoodSleepingFunctionalGroup2006; @carrBiodiversityPopulationRegulation2002; @connollyIndoPacificBiodiversityCoral2003; @dornelasCoralReefDiversity2006; @grahamImportanceStructuralComplexity2013; @komyakovaRelativeImportanceCoral2013].
+In the Hawaiian archipelago, we found significant differences in the composition of cryptic crab communities between the populated high islands of the MHI and the uninhabited atolls of the NWHI.
+Moreover, these differences were driven by a combination of environmental (sea-surface temperature, chlorophyll concentration, depth), biotic (potential larval immigration), and anthropogenic (cumulative human impacts) factors.
+<!--Alpha diversity metrics show slight differences in the number of species and overall abundance of crabs in each sampled assemblage between the Main and Northwestern Hawaiian Islands, but CAP and PERMANOVA highlight significant differences in species assemblages between the regions.
+Differences in the species assemblages across regions are driven primarily by human impacts, cholorophyll a and sea surface temperature, although other factors such as larval connectivity, distance to nearest island, mean coral cover, and depth are also significant.-->
+
 The Hawaiian Archipelago is among the most remote on the planet and encompasses a range of gradients (anthropogenic, biotic, abiotic), making it an ideal area to study the impacts of these factors on species assemblages across space.
 Human impacts are considerably higher in the populated MHI than in the protected and remote NWHI [@selkoeEvaluatingAnthropogenicThreats2008; @selkoeMapHumanImpacts2009; @selkoeDNACoralReef2016].
-Although we did not consider individual effects here, an index representing cumulative anthropogenic impacts was the top driver of brachyuran community structure, similar to previous work examining genetic diversity across the archipelago [@selkoeDNACoralReef2016].
+Although we did not consider individual effects here, an index representing cumulative anthropogenic impacts was the top driver of brachyuran community structure, similar to findings in previous work examining genetic diversity across the archipelago [@selkoeDNACoralReef2016].
 The next-most significant driver, chlorophyll a, can be used as a proxy for primary productivity.
@@
 
-## Processing of satellite-derived SST and chl-A data
-Sea-surface temperature (SST) and chlorophyll A (chl-A) data were retrieved from NASA's Physical Oceanography Distributed Active Archive Center [PO.DAAC; @NASAJPL_2020] and Ocean Biology Distributed Active Archive Center [OB.DAAC; @NASAOceanBiologyProcessingGroup_2022] respectively.
-Monthly summary datasets were downloaded for 30 randomly subsampled months within each ARMs unit's three-year soaking period, averaged, and values (SST or chl-A) were extracted for the geographic coordinates of each sampling site, with a 300 m surrounding buffer zone.
+## Processing of satellite-derived environmental data
+Level-3 mapped (9km) sea-surface temperature (SST) and chlorophyll A (chl-A) data were retrieved from NASA's Physical Oceanography Distributed Active Archive Center [PO.DAAC; @NASAJPL_2020] and Ocean Biology Distributed Active Archive Center [OB.DAAC; @NASAOceanBiologyProcessingGroup_2022] respectively.
+Monthly summary datasets were downloaded for the period of 2009--2016 and averaged over each of the ARMS unit soaking periods (e.g.
+2009--2011, 2010--2012, etc.).
+Values (SST or chl-A) were extracted for the geographic coordinates of each sampling site from the averaged dataset for the relevant soaking period.
+Specific monthly-average dataset names can be found in @tbl:nasa_table.
 
@@
 (p1 + p2) / (p3 + p4) & 
-  plot_annotation(tag_levels = "A")
+  plot_annotation(tag_levels = list(str_glue("({letters[1:4]})"))) & 
+  theme(plot.tag = element_text(face="italic"))
 cat("{#fig:fig_alpha}")
@@
 (p1 + p2) & 
-  plot_annotation(tag_levels = "A")
+  plot_annotation(tag_levels = list(str_glue("({letters[1:2]})"))) & 
+  theme(plot.tag = element_text(face="italic"))
 cat("{#fig:fig_alpha2}")
@@
 ```
+
+\newpage
+
+```{r nasa-table,results="asis",depenson="init"}
+nasa_table <- read_csv(here("data","tables","nasa.csv"))
+print_table(nasa_table,"nasa_table","NASA data repository filenames for CHL and SST satellite data",accuracy=1) 
+```

### 2024-01-05 chl-sst.R

@@ -9,166 +9,100 @@ library(fs)
 # set random seed for consistency
 set.seed(31337)
 
-
-## download SST and CHL data from NASA and aggregate it by ARMS sampling site
-
-
-# pre-parse metadata and get deployment & recovery years as well as deployment range
-arms <- read_csv(here("data","metadata.csv")) %>%
-  rename(sample=unit) %>%
-  st_as_sf(coords=c("lon","lat"),crs=4326,remove=FALSE) %>%
-  separate_wider_regex(sample,patterns=c(recovery_year="^[0-9]+","_",sample=".*"),cols_remove=FALSE)  %>%
-  mutate(recovery_year=as.numeric(recovery_year)) %>%
-  dplyr::select(sample,everything()) %>%
-  unite("range",deployment_year,recovery_year,sep="--",remove=FALSE) 
-
-
-# sample 30 months from each date range
-dates <- arms %>%
-  group_by(range) %>%
-  group_map(\(data,grp) {
-    start_date <- str_glue("{min(data$deployment_year)}-01-01") %>%
-      as.Date()
-    end_date <- str_glue("{max(data$recovery_year)}-12-31") %>%
-      as.Date()
-    
-    tibble(
-      range = grp$range,
-      date =  seq(start_date,end_date,by="month") %>%
-        sample(30) %>%
-        as.character()
-    )
-  }) %>%
-  bind_rows()
-
-# chlorophyll and sst datasets
-datasets <- c(chl="MY1DMM_CHLORA", sst="MYD28M")
-
-
-# url templates for getting image IDs and GeoTIFF data
-id_url="https://neo.gsfc.nasa.gov/lib/imageSizes.php?datasetId={dataset}&curDate={timestamp}"
-tiff_url="https://neo.gsfc.nasa.gov/servlet/RenderData?si={tiff_id}&cs=rgb&format=FLOAT.TIFF&width=3600&height=1800"
-
-# get unix timestamp
-ts <- function(d) {
-  as.numeric(as.POSIXct(d,tz="HST"))
-}
-
-# retrieve a URL
-get_url <- function(u) {
-  r <- GET(u)
-  if (r$status_code == 200) {
-    return(content(r,as="text",encoding="UTF-8"))
-  } else {
-    return(NULL)
-  }
-}
-
-# make sure directories exist
-datasets %>%
-  iwalk(\(ds,n) {
[…]

### 2024-01-05 crabs.R

@@ -322,7 +322,9 @@ setup_crabs <- function() {
   cc$crab_data_shallow <- cc$crabs_shallow %>%
     sample_tibble(sample_col = "unit") %>%
     select(-sst,-chl) %>%
-    select(unit,region,island_group,island,lat,lon,depth,chl=chl_sat,sst=sst_sat,slope,coral_cover,closest_island,larval_connectivity,human_impact) %>%
+    # select(unit,region,island_group,island,lat,lon,depth,chl=chl_sat,sst=sst_sat,slope,coral_cover,closest_island,larval_connectivity,human_impact) %>%
+    select(unit,region,island_group,island,lat,lon,depth,chl=chlorophyll_oracle,sst=sst_sat,slope,coral_cover,closest_island,larval_connectivity,human_impact) %>%
+    # select(unit,region,island_group,island,lat,lon,depth,chl=chl_new,sst=sst_new,slope,coral_cover,closest_island,larval_connectivity,human_impact) %>%
     mutate(across(where(is.numeric),~as.numeric(scale(.x)))) %>%
     column_to_rownames("unit") 
     # reassociate the new sample data

### 2024-02-01 manuscript.Rmd
updates before revision

@@
 The structure and composition of coral reef communities is influenced by a suite of environmental and anthropogenic factors, yet the nature and makeup of those factors remain poorly resolved.
-Coral reefs are among the most diverse habitats in the world, and they face increasing threats of degradation, including climate-driven bleaching and phase shifts to alternate ecological regimes [@grahamManagingResilienceReverse2013; @norstromAlternativeStatesCoral2009; @pandolfiGlobalTrajectoriesLongTerm2003], among others.
+Coral reefs are among the most diverse habitats in the world, and they face increasing threats of degradation, including climate-driven bleaching and phase shifts to alternate ecological regimes [@grahamManagingResilienceReverse2013; @norstromAlternativeStatesCoral2009; @pandolfiGlobalTrajectoriesLongTerm2003].
 In the face of these and other threats, understanding drivers of community structure is critically important [@hughesRisingChallengeSustaining2010; @pageSeekingResistanceCoral2019].
@@
 Here, we investigate the factors influencing the community structure of a group of cryptic coral reef taxa across the Hawaiian Archipelago.
-<!--Say something about depth here, but I'm not sure what or how-->
 
@@
 A likely important but under-investigated aspect of coral reef community structure is depth.
-Most studies to date have focused on the upper ~30 m of the water column, yet coral reefs and associated communities are known to extend beyond 150 m [@pyleExploringDeepCoral1996; @pyleMesophoticCoralEcosystems2019], with scleractinian corals reported to 170 m [@rouzeSymbioticAssociationsDeepest2021].
+Most studies to date have focused on the upper ~30 m of the water column, yet coral reefs and associated communities are known to extend beyond 150 m [@pyleExploringDeepCoral1996; @pyleMesophoticCoralEcosystems2019], with photosymbiotic corals reported to 170 m [@rouzeSymbioticAssociationsDeepest2021].
 As much as two thirds of coral reef ecosystems lie in this under-explored zone [@hindersteinThemeSectionMesophotic2010].
@@
 Further, there are a range of connections depending on the taxa – large, mobile organisms like fish tend to have more shared species across depth than sedentary organisms such as corals and algae, but even among the fishes the community structure of MCEs is often distinct [@kaneHighLevelsMesophotic2014; @pyleComprehensiveInvestigationMesophotic2016 ; @asherMesophoticDepthGradients2017; @fukunagaChangesMesophoticReef2017; @colemanShiftingReefFish2018; @pinheiroDeepReefFishes2019].
-Although the majority of coral reef biodiversity consists of the algal and invertebrate cryptobenthic organisms living on and within the reef structure [@fautinOverviewMarineBiodiversity2010; @knowltonCoralReefBiodiversity2010; @fisherSpeciesRichnessCoral2015], the studies listed above focus almost exclusively on the corals and fishes which are more easily surveyed.--> Given the strong reported differences between deep and shallow reefs across taxa and the paucity of research on drivers of cryptic coral reef community variation, we sought to answer the following research questions: (i) How do shallow crab communities vary along the natural and anthropogenic gradients of the Hawaiian Archipelago?
+Although the majority of coral reef biodiversity consists of the algal and invertebrate cryptobenthic organisms living on and within the reef structure [@fautinOverviewMarineBiodiversity2010; @knowltonCoralReefBiodiversity2010; @fisherSpeciesRichnessCoral2015], the studies listed above focus almost exclusively on the corals and fishes which are more easily surveyed.-->Given the strong reported differences between deep and shallow reefs across taxa and the paucity of research on drivers of cryptic coral reef community variation, we sought to answer the following research questions: (i) How do shallow crab communities vary along the natural and anthropogenic gradients of the Hawaiian Archipelago?
 (ii) Do shallow crab communities differ between the Main and Northwestern Hawaiian Islands?
@@
 A total of 89 ARMS units were deployed over the course of six years (2010--2016).
-Units were deployed by NOAA scuba divers during Pacific Reef Assessment and Monitoring Program (RAMP) research cruises in shallow coral reef habitat (12–16 m depth) in both the Main (MHI) and Northwest Hawaiian Islands (NWHI) ([@fig:fig_map; @tbl:arms_table]).
-<!--The sites in the MHI were at Hawai&#x02BB;i Island, Maui, O&#x02BB;ahu, and Kaua&#x02BB;i.
-The sites in the NWHI were at Lalo (French Frigate Shoals), Kapou (Lisianski), Manawai (Pearl & Hermes), and Hōlanikū (Kure Atoll) ([@fig:fig_map]).
--->To explore the effects of depth, we also present data from an earlier study by Hurley et al.
-[-@hurleyAssessmentShallowMesophotic2016], which included a total of 27 units deployed off O&#x02BB;ahu from 2009–2012 in 12–90 m depth; nine units were deployed at 12 m, and six each at 30 m, 60 m, and 90 m ([@tbl:arms_table]).
-Shallow units were deployed according to standard NOAA methodology, whereas units $\geq$ 30 m were deployed and recovered using technical mixed gas closed-circuit rebreather diving [see @hurleyAssessmentShallowMesophotic2016].
+Units were deployed by NOAA divers during Pacific Reef Assessment and Monitoring Program (RAMP) cruises in shallow coral reef habitat (12--16 m depth) in both the Main (MHI) and Northwest Hawaiian Islands (NWHI) ([@fig:fig_map; @tbl:arms_table]).
+To explore the effects of depth, we also present data from an earlier study by Hurley et al.
+[-@hurleyAssessmentShallowMesophotic2016], in which 27 units were deployed off O&#x02BB;ahu from 2009--2012 in 12--90 m depth; nine units were deployed at 12 m, and six each at 30 m, 60 m, and 90 m ([@tbl:arms_table]).
+Shallow units were deployed according to NOAA-RAMP methodology, whereas units $\geq$ 30 m were deployed and recovered using technical mixed gas closed-circuit rebreather diving [see @hurleyAssessmentShallowMesophotic2016].
 
@@
 
-ARMS units were retrieved by encapsulating them in-situ in a mesh-lined crate (100 μm pore size) to prevent escape of motile organisms.
+ARMS units were retrieved by encapsulating them in-situ in a 100 μm mesh-lined crate to prevent escape of motile organisms.
 Once returned to the surface, they were placed in containers of aerated, filtered seawater, crates were removed, and units were systematically disassembled plate-by-plate.
-Seawater surrounding disassembled units was sieved through 2 mm, 500 μm, and 100 μm mesh-size geologic sieves [see @lerayDNABarcodingMetabarcoding2015 for processing details], and retained macroorganisms were preserved in 95% ethanol.
+Seawater surrounding disassembled units was sieved through 2 mm, 500 μm, and 100 μm mesh geologic sieves [see @lerayDNABarcodingMetabarcoding2015 for processing details], and retained macroorganisms were preserved in 95% ethanol.
 All adult brachyuran crabs ($\geq$ 5 mm) were photographed, identified (to species where possible, otherwise morphospecies), and preserved in 95% ethanol.
@@
 Unless otherwise specified, all analyses were performed using `R` version 4.1.2 [@rcoreteamLanguageEnvironmentStatistical2021].
-Community data were loaded and encapsulated into data objects using the R package `phyloseq` [@mcmurdiePhyloseqPackageReproducible2013] and multivariate community analyses were performed using the package `vegan` [@oksanenVeganCommunityEcology2022] on Bray-Curtis dissimilarities of Hellinger-transformed species abundance matrices [@legendreEcologicallyMeaningfulTransformations2001; @raoReviewCanonicalCoordinates1995].
+Community matrices were loaded and encapsulated into data objects using the R package `phyloseq` [@mcmurdiePhyloseqPackageReproducible2013] and multivariate community analyses were performed using the package `vegan` [@oksanenVeganCommunityEcology2022] on Bray-Curtis dissimilarities of Hellinger-transformed species abundance matrices [@legendreEcologicallyMeaningfulTransformations2001; @raoReviewCanonicalCoordinates1995].
 
@@
 Model significance was assessed using permutation tests with `r num(PERM,1)` permutations with the function `anova`.
-All continuous variables were scaled to unit variance.
+Continuous variables were scaled to unit variance.
 
@@
 Significance was assessed with permutation tests in both cases.
-Variation among sampling groups (shallow MHI, shallow NWHI, O&#x02BB;ahu 30 m, 60 m, 90 m) was further examined and visualized using canonical analysis of principal coordinates (CAP) based on discriminant analysis [@andersonCanonicalAnalysisPrincipal2003] as implemented in the `CAPdiscrim` function in `BiodiversityR` [@kindtTreeDiversityAnalysis2005].
+Variation among sampling groups (island group and depth) was further examined and visualized using canonical analysis of principal coordinates (CAP) based on discriminant analysis [@andersonCanonicalAnalysisPrincipal2003] as implemented in the `CAPdiscrim` function in `BiodiversityR` [@kindtTreeDiversityAnalysis2005].
 
@@
 
-Potential environmental drivers of shallow crab community structure were assembled from a previous study on environmental drivers of genetic diversity [@selkoeDNACoralReef2016], NOAA's National Coral Reef Monitoring Program (NCRMP), the Pacific Islands Ocean Observing System (PacIOOS), NASA Physical Oceanography Distributed Active Archive Center (PO.DAAC), and NASA Ocean Biology Distributed Active Archive Center (OB.DAAC) ([@tbl:env_table]).
+Potential drivers of shallow crab community structure were assembled from a previous study examining environmental effects on genetic diversity [@selkoeDNACoralReef2016], NOAA's National Coral Reef Monitoring Program (NCRMP), the Pacific Islands Ocean Observing System (PacIOOS), and NASA's Physical Oceanography/Ocean Biology Distributed Active Archive Centers (PO.DAAC, OB.DAAC) ([@tbl:env_table]).
 For details on how sea-surface temperature and chlorophyll data were compiled and aggregated, see Supplemental material.
-Due to logistical limitations, environmental data were collected only for shallow samples ($\le$ 30 m).
+Due to logistical limitations, environmental data were collected only for shallow samples (\< 30 m).
 
@@
 ```{r pairwise-island, fig.cap='Significance levels of pairwise PERMANOVA tests by island for shallow crab communities (p-values adjusted according to the false discovery rate [@benjaminiControllingFalseDiscovery1995]).', results="asis", dev="svg", fig.width=12, fig.height=9, dependson="init"}
-library(ggtext)
-pairwise_island <- pairwise_adonis(cc$crab_dist_shallow,cc$crab_data_shallow$island,permutations=PERM)
+pairwise_island <- pairwise_adonis(cc$crab_dist_shallow,cc$crab_data_shallow$hawaiian_name,permutations=PERM)
 
@@
   sample_tibble() %>%
-  pull(island) %>%
+  pull(hawaiian_name) %>%
   levels()
@@
 pw1 <- pairwise_island %>%
-  select(left,right,p_adj) %>%
+  select(left,right,p_adj,pseudo_f) %>%
   mutate(left=factor(left,levels=rev(islands)),right=factor(right,levels=rev(islands))) %>%
-  mutate(sig = round(p_adj,2) < 0.05, p_adj = pval(p_adj)) %>%
+  mutate(sig = round(p_adj,2) < 0.05, p_adj = pval(p_adj), f = str_glue("F = {num(pseudo_f)}")) %>%
   mutate(
@@
   select(left,right,everything()) %>%
-  mutate(left=factor(left,levels=islands),right=factor(right,levels=islands))
+  mutate(left=factor(left,levels=islands),right=factor(right,levels=islands)) %>%
+  mutate(txt = str_glue("{p_adj}<br>{f}"))
 
@@
       y=left,
-      label=p_adj,
+      label=txt,
       fontface = if_else(sig,"bold","plain")

### 2024-02-01 crabs.R

@@ -323,7 +323,7 @@ setup_crabs <- function() {
     sample_tibble(sample_col = "unit") %>%
     select(-sst,-chl) %>%
     # select(unit,region,island_group,island,lat,lon,depth,chl=chl_sat,sst=sst_sat,slope,coral_cover,closest_island,larval_connectivity,human_impact) %>%
-    select(unit,region,island_group,island,lat,lon,depth,chl=chlorophyll_oracle,sst=sst_sat,slope,coral_cover,closest_island,larval_connectivity,human_impact) %>%
+    select(unit,region,island_group,island,hawaiian_name,lat,lon,depth,chl=chlorophyll_oracle,sst=sst_sat,slope,coral_cover,closest_island,larval_connectivity,human_impact) %>%
     # select(unit,region,island_group,island,lat,lon,depth,chl=chl_new,sst=sst_new,slope,coral_cover,closest_island,larval_connectivity,human_impact) %>%
     mutate(across(where(is.numeric),~as.numeric(scale(.x)))) %>%
     column_to_rownames("unit") 

### 2024-03-18 manuscript.Rmd
two rounds of comments from RJT

@@
 ---
-title: "Depth over distance: Hawaiian cryptic reef communities are more distinct across ~90 m depth than ~2500 km of shallow habitat"
+title: "Hawaiian reef communities are more distinct across a 90 m depth gradient than 2,500 km of shallow habitat"
 author:
@@
         - natgeo
+  - Derek J. Skillings:
+      institute: 
+        - uncg
   - Kerry Reardon:
@@
   - natgeo: National Geographic Pristine Seas
+  - uncg: 'University of North Carolina, Greensboro'
   - noaa: National Oceanic and Atmospheric Administration
 abstract:
-  Shallow coral reefs are extensively studied, and although scleractinian corals have been recorded to 165 m, little is known about other mesophotic coral reef ecosystem (MCE) inhabitants.
+  Coral reef communities are influenced by various environmental and anthropogenic factors, yet those factors have not been consistently resolved.
+Furthermore, due to logistical difficulties, investigations of the role of depth (particularly mesophotic coral ecosystems) in influencing coral reef community structure have been taxonomically limited.
+Here, we examined the effects of a suite of environmental, ecological, and anthropogenic factors on the community structure of shallow brachyuran crabs across the ~2,500 km span of the Hawaiian Archipelago.
+In addition, we compared shallow crab communities with those sampled on a depth transect from 12--90 m on O&#x02BB;ahu.
+We deployed Autonomous Reef Monitoring Structures (ARMS) on shallow reefs (12--16 m) across the latitudinal gradient of the Hawaiian Archipelago to compare with a depth gradient (12, 30, 60, and 90 m) on the south shore of O&#x02BB;ahu.
+We found a total of 2,284 individual crabs representing 90 morphospecies and 19 families.
+Shallow crab communities were significantly different between the Main and Northwest Hawaiian Islands and variation in community structure was significantly associated with six biotic, environmental, and anthropogenic factors (cumulative human impacts, chlorophyll-A, sea-surface temperature, site depth, island slope, and potential larval connectivity).
+In depth comparisons, we found more variation among mesophotic samples from O&#x02BB;ahu (30--90 m) than among all shallow samples across the entire archipelago.
+We show that 90 m of depth is a more potent determinant of brachyuran community composition than the latitudinal, environmental, or anthropogenic gradients across the Hawaiian Islands.<!--Deeper reefs host significantly different brachyuran communities than shallow ones in Hawai&#x02BB;i, and only 3 of 90 morphospecies (~3%) occurring across the entire depth range (mesophotic to shallow) were sampled.
+Further research on the cryptic reef-dwelling fauna is needed for identifying areas of unique biodiversity, as well as boundaries between depth-stratified reef ecosystems.
+Shallow coral reefs are extensively studied, and although scleractinian corals have been recorded to ~170 m, little is known about other mesophotic coral reef ecosystem (MCE) inhabitants.
 Brachyuran crabs fill many ecological and trophic niches on reefs, making them ideal candidates for evaluating species composition among depths.
-Here we ask if MCEs host the same communities as the shallower reefs spread across the entire 2400 km Hawaiian Archipelago.
-We deployed Autonomous Reef Monitoring Structures (ARMS) for two years among shallow sites (12 m) across the latitudinal gradient of the Hawaiian Archipelago to compare directly with a depth gradient (12, 30, 60, and 90 m) south of Oʻahu island.
-A total of 2284 individual crabs representing 90 morphospecies (19 families) were found.
-Assemblage composition was not statistically significantly different among shallow sites spread across the archipelago but was significantly different and highly stratified across the depth gradient.
-We show that 90 m of depth is a more potent determinant of brachyuran community composition than the latitudinal, geographic (high vs low islands), or anthropogenic impact (human population size) gradient across the archipelago.
-Deeper reefs host significantly different brachyuran communities than shallow ones in Hawai&#x02BB;i, and only 3 of 90 morphospecies (~3%) occurring across the entire depth range (mesophotic to shallow) were sampled.
-Further research on the cryptic reef-dwelling fauna is needed for identifying areas of unique biodiversity, as well as boundaries between depth-stratified reef ecosystems.
+Here we ask if MCEs host the same communities as the shallower reefs spread across the entire 2400 km Hawaiian Archipelago.-->
 bibliography: references/citations.bib
@@
 
-print_table <- function(tbl,id="",caption="",hdr_case="sentence",hide=character(0),accuracy = 0.01) {
+print_table <- function(tbl,id="",caption="",hdr_case="sentence",hide=character(0),accuracy = 0.01,format_numbers=TRUE) {
   if (!grepl("^: ",caption)) {
@@
       across(
-        where(is.numeric),
+        where(~format_numbers & is.numeric(.x)),
         ~num(.x,accuracy)
@@
     title = str_to_title(names(tbl)),
+    upper = str_to_upper(names(tbl)),
+    lower = str_to_lower(names(tbl)),
     names(tbl)
@@
 
-# palette for island/group colors
-# island_pal <- c(
-#   "Hawaii Island" = "#8a3ffc",
-#   "Hawai‘i Island" = "#8a3ffc",
-#   "Maui" = "#007d79",
-#   "Oahu" = "#ff7eb6",
-#   "O‘ahu" = "#ff7eb6",
-#   "Kauai" = "#fa4d56",
-#   "Kaua‘i" = "#fa4d56",
-# 
-#   "French Frigate Shoals" = "#fff1f1",
-#   "Lalo" = "#fff1f1",
-#   "Lalo (French Frigate Shoals)" = "#fff1f1",
-#   "Lisianski" = "#6fdc8c",
-#   "Kapou" = "#6fdc8c",
-#   "Kapou (Lisianski)" = "#6fdc8c",
-#   "Pearl and Hermes Reef" = "#4589ff",
-#   "Manawai" = "#4589ff",
-#   "Manawai (Pearl and Hermes Reef)" = "#4589ff",
-#   "Kure Atoll" = "#d12771",
-#   "Hōlanikū" = "#d12771",
-#   "Hōlanikū (Kure Atoll)" = "#d12771"
-# )
 island_pal <- c(
@@
 
-The structure and composition of coral reef communities is influenced by a suite of environmental and anthropogenic factors, yet the nature and makeup of those factors remain poorly resolved.
+The structure and composition of coral reef communities is influenced by a suite of environmental and anthropogenic factors, yet the nature and makeup of those factors remain poorly resolved [@hughesCommunityStructureDiversity1989; @smithEffectsTopBottom2010; @brandlCoralReefEcosystem2019].
 Coral reefs are among the most diverse habitats in the world, and they face increasing threats of degradation, including climate-driven bleaching and phase shifts to alternate ecological regimes [@grahamManagingResilienceReverse2013; @norstromAlternativeStatesCoral2009; @pandolfiGlobalTrajectoriesLongTerm2003].
-In the face of these and other threats, understanding drivers of community structure is critically important [@hughesRisingChallengeSustaining2010; @pageSeekingResistanceCoral2019].
-To date, the majority of efforts in this area have focused on conspicuous taxa such as corals and fishes [e.g., @benfieldComparisonCoralReef2008; @caballero-aragonWaveExposureTemperature2023; @kaneTrophicDesignationLive2017; @langeWaveExposureShapes2021; @vanlierHabitatConnectivityComplexity2018] or microbes [e.g., @desouzaImportanceDepthTemperature2023; @morrowEnvironmentalDriversMicrobial2016].
-However, these taxa make up a minority of coral reef biodiversity and thus may not be indicative of community patterns overall [[cite?]]{.mark}.
-Here, we investigate the factors influencing the community structure of a group of cryptic coral reef taxa across the Hawaiian Archipelago.
+In the face of these threats, understanding drivers of metazoan community structure in coral reef ecosystems is critically important [@hughesRisingChallengeSustaining2010; @pageSeekingResistanceCoral2019], particularly as the search for criteria to prioritize climate refugia such as upwelling zone or mesophotic reefs intensifies.
+To date, the majority of such efforts have focused on conspicuous taxa such as corals and fishes [e.g., @benfieldComparisonCoralReef2008; @caballero-aragonWaveExposureTemperature2023; @kaneTrophicDesignationLive2017; @langeWaveExposureShapes2021; @vanlierHabitatConnectivityComplexity2018].
+However, these taxa make up a minority of coral reef biodiversity and thus may not be indicative of community patterns overall [@fautinOverviewMarineBiodiversity2010].
+Here, we investigate the factors influencing the community structure of brachyuran crabs across a suite of anthropogenic and environmental gradients using the Hawaiian Archipelago as a natural laboratory.
+
+The Hawaiian Archipelago comprises high volcanic islands and atolls stretching across ~2,500 km in the North Pacific Ocean and is the most isolated island chain on the planet.
+Typically regarded as two regions, the main Hawaiian Islands (MHI) include the eight inhabited high islands, and the Northwestern Hawaiian Islands (NWHI) comprise the uninhabited islands, atolls, and banks northwest of Kaua&#x02BB;i and Ni&#x02BB;ihau that make up the Papahānaumokuākea Marine National Monument.
+The islands are subject to considerable anthropogenic and physical gradients across their span [@selkoeMapHumanImpacts2009], which are widely believed to influence coral reef communities.
+These gradients include differences in temperature, human population, and anthropogenic impact, among others.
+Studies examining community structure across broad environmental gradients have been criticized for confounding multiple potentially covarying factors (such as habitat area and/or complexity) across the measured gradients [e.g., @bellwoodRegionalScaleAssemblyRules2001; @grahamEffectsMarineReserve2003; @connollyCommunityStructureCorals2005; @ewersConfoundingFactorsDetection2006; @fletcherRoleHabitatArea2007a].
+Thus, we used autonomous reef monitoring structures (ARMS)---standardized collection devices designed to mimic reef structural complexity---to sample cryptic communities while controlling for habitat [@zimmermanArtificialReefMatrix2004; @brainardAutonomousReefMonitoring2009].
+We then sampled brachyuran crabs, a group comprising a notable proportion of the coral reef cryptofauna [289 species in Hawai&#x02BB;i, @castroCatalogAnomuranBrachyuran2011] from these structures to examine drivers of community composition across these gradients.
+We compared crab communities among islands and regions (MHI vs.
+NWHI) and examined the effects of a suite of possible environmental and anthropogenic drivers of community structure to test hypotheses about their relative importance.
 
 Much of our understanding of biotic zonation and community structure in the ocean originates with classic studies from the intertidal [e.g., @connellInfluenceInterspecificCompetition1961; @connellEffectsCompetitionPredation1961; @paineIntertidalCommunityStructure1974].
-Such studies point to a combination of physical and biological factors<!--, including phyiological tolerance [@dennyPhysicalProcessesThat2001; @someroThermalPhysiologyVertical2002] and species interactions [@brunoHabitatModificationFacilitation2001; @mengeRockyIntertidalCommunities2001], that contribute to the structure of intertidal communities.
-Although there are exceptions [e.g., @woottonIndirectEffectsPrey1992; @bertnessTestingRelativeContribution1999]-->, but (with exceptions) a generalization has emerged <!--for the intertidal -->that communities' lower bathymetric limits are regulated by ecological interactions whereas upper limits are determined by abiotic factors.
-However, these models describe zonation at individual localities rather than community variation over larger spatial scales and may not apply to coral reefs, which are subject to more complex community dynamics [@abramsRecruitmentLotteriesCoexistence1984; @bussCompetitiveNetworksNontransitive1979; @hixonPredationPreyRefuges1993; @hubbellUnifiedNeutralTheory2011; @moraEcologyFishesCoral2015; @saleStockRecruitRelationshipsRegional1982; @sheppardBiologyCoralReefs2017].
-In any case, it is likely that while ecological interactions are important for structuring communities at the local scale, external factors may exert stronger influence across greater distances [[cite?]]{.mark}.
-
-The Hawaiian Archipelago comprises high volcanic islands and atolls stretching across ~2,500 km in the North Pacific Ocean, and is typically regarded as two regions: the main Hawaiian Islands (MHI), which include the eight inhabited high islands, and the Northwestern Hawaiian Islands (NWHI), the uninhabited atolls and banks northwest of Kaua&#x02BB;i and Ni&#x02BB;ihau that make up the Papahānaumokuākea Marine National Monument.
-The islands are subject to anthropogenic and physical gradients across their span [@selkoeMapHumanImpacts2009], which we hypothesize to influence coral reef communities.
-Studies examining community structure across environmental gradients have been criticized for confounding multiple potentially covarying factors (such as habitat area and/or complexity) across the measured gradients [e.g., @bellwoodRegionalScaleAssemblyRules2001; @grahamEffectsMarineReserve2003; @connollyCommunityStructureCorals2005; @ewersConfoundingFactorsDetection2006; @fletcherRoleHabitatArea2007a].
-Thus, we used autonomous reef monitoring structures (ARMS)---standardized collection devices designed to mimic reef structural complexity---to sample cryptic communities while controlling for habitat<!--differences in habitat variables--> [@zimmermanArtificialReefMatrix2004; @brainardAutonomousReefMonitoring2009].
-Specifically, we used ARMS to sample brachyuran crabs, a group comprising a notable proportion of the coral reef cryptofauna [289 species in Hawai&#x02BB;i, @castroCatalogAnomuranBrachyuran2011].
-We compared crab communities among islands and regions (MHI vs.
-NWHI) and examined the effects of a suite of possible environmental and anthropogenic drivers.
-
-A likely important but under-investigated aspect of coral reef community structure is depth.
+While there are no shortages of exceptions, a long history of studies point towards a generalization that the lower bathymetric limits are generally regulated by ecological interactions whereas upper limits of intertidal communities are more often determined by abiotic factors [@stephenson1972life; @mengeRockyIntertidalCommunities2001; @bertnessTestingRelativeContribution1999; @dennyPhysicalProcessesThat2001].
+However, most of these models describe zonation at specific localities rather than community variation over larger spatial scales [but see @birdTideWaveMarks2013; @bishop-taylorTidesModellingElevation2019] and may not apply to coral reefs, which are subject to more complex community dynamics [@abramsRecruitmentLotteriesCoexistence1984; @bussCompetitiveNetworksNontransitive1979; @hixonPredationPreyRefuges1993; @hubbellUnifiedNeutralTheory2011; @moraEcologyFishesCoral2015; @saleStockRecruitRelationshipsRegional1982; @sheppardBiologyCoralReefs2017].
+While there is little doubt that ecological interactions are important on coral reefs, the suite of factors and which play the most important role in structuring coral reef communities remains a subject of considerable debate [@bellwoodRegionalScaleAssemblyRules2001; @connollyUnifiedModelExplains2017; @hubbellUnifiedNeutralTheory2011; @jacksonAdaptationDiversityReef1991; @mumbyCoralReefHabitats2008; @saleMaintenanceHighDiversity1977].
+
+One aspect of coral reef community structure that is thought to be important but remains under-investigatedis depth.
 Most studies to date have focused on the upper ~30 m of the water column, yet coral reefs and associated communities are known to extend beyond 150 m [@pyleExploringDeepCoral1996; @pyleMesophoticCoralEcosystems2019], with photosymbiotic corals reported to 170 m [@rouzeSymbioticAssociationsDeepest2021].
@@
 These habitats, known as mesophotic coral ecosystems (MCEs), are typically defined as those extending from ~30--150 m depth [@hindersteinThemeSectionMesophotic2010; @kahngCommunityEcologyMesophotic2010].
-<!--and were believed to experience dampened disturbances in comparison to shallow reefs and could therefore act as deep water refugia for shallow reef species threatened by anthropogenic impacts --> They have been hypothesized as possible refugia from disturbances on shallow reefs [@glynnCoralReefBleaching1996; @bongaertsAssessingDeepReef2010; @smithDepthRefugiumCatastrophic2014; @holsteinFertileFathomsDeep2015; @muirSpeciesIdentityDepth2017], although studies have found evidence that deep reefs are unique and uniquely threatened, challenging this notion [@diazLightTemperatureDrive; @diazMesophoticCoralBleaching2023; @hobanPlumbingDepthsEnvironmental2023; @hurleyAssessmentShallowMesophotic2016; @kaneHighLevelsMesophotic2014; @pinheiroDeepReefFishes2019; @rochaMesophoticCoralEcosystems2018].
-<!--However, recent literature describing species compositions along depth gradients reveal a more complex picture among taxa and locations because some groups may use mesophotic reefs as a refuge in some places, while others show community turnover with depth such that MCE habitats are ecologically distinct from adjacent shallow water coral reefs [@hurleyAssessmentShallowMesophotic2016; @smithCaribbeanMesophoticCoral2016; @semmlerAreMesophoticCoral2017; @bongaertsDeepReefsAre2017; @rochaMesophoticCoralEcosystems2018].
-Further, there are a range of connections depending on the taxa – large, mobile organisms like fish tend to have more shared species across depth than sedentary organisms such as corals and algae, but even among the fishes the community structure of MCEs is often distinct [@kaneHighLevelsMesophotic2014; @pyleComprehensiveInvestigationMesophotic2016 ; @asherMesophoticDepthGradients2017; @fukunagaChangesMesophoticReef2017; @colemanShiftingReefFish2018; @pinheiroDeepReefFishes2019].
-Although the majority of coral reef biodiversity consists of the algal and invertebrate cryptobenthic organisms living on and within the reef structure [@fautinOverviewMarineBiodiversity2010; @knowltonCoralReefBiodiversity2010; @fisherSpeciesRichnessCoral2015], the studies listed above focus almost exclusively on the corals and fishes which are more easily surveyed.-->Given the strong reported differences between deep and shallow reefs across taxa and the paucity of research on drivers of cryptic coral reef community variation, we sought to answer the following research questions: (i) How do shallow crab communities vary along the natural and anthropogenic gradients of the Hawaiian Archipelago?
+They have been hypothesized as possible refugia from climate change impacts on shallow reefs [@glynnCoralReefBleaching1996; @bongaertsAssessingDeepReef2010; @smithDepthRefugiumCatastrophic2014; @holsteinFertileFathomsDeep2015; @muirSpeciesIdentityDepth2017], although a growing number of studies have found evidence that mesophotic reefs are unique and uniquely threatened, challenging this notion [@diazMesophoticCoralBleaching2023; @diazMesophoticCoralBleaching2023; @hobanPlumbingDepthsEnvironmental2023; @hurleyAssessmentShallowMesophotic2016; @kaneHighLevelsMesophotic2014; @pinheiroDeepReefFishes2019; @rochaMesophoticCoralEcosystems2018; @montgomeryCommunitySimilaritySpecies2021].
+Given the strong reported differences between deep and shallow reefs across taxa and the paucity of research on drivers of cryptic coral reef community variation, we sought to answer the following four research questions: (i) How do shallow crab communities vary along the natural and anthropogenic gradients of the Hawaiian Archipelago?
 (ii) Do shallow crab communities differ between the Main and Northwestern Hawaiian Islands?
-(iii) What environmental and/or anthropogenic factors drive crab community structure in shallow coral reefs across the Hawaiian Archipelago?
+(iii) What environmental and anthropogenic factors drive crab community structure in shallow coral reefs across the Hawaiian Archipelago?
 (iv) What are the relative contributions of depth (shallow vs.
 mesophotic) and island region to crab community variation across the Hawaiian Archipelago?
-
-<!-- Here, we employed autonomous reef monitoring structures (ARMS)  Using ARMS, we sampled brachyuran crab communities across the Hawaiian Archipelago to test the effects of environmental and anthropogenic drivers on the structure of cryptic reef communities.
-Given the number and magnitude of differences between these regions, we hypothesized that species diversity would be different between regions (MHI vs NWHI).
-We therefore looked at community composition along the latitudinal gradient and investigated community composition in response to physical gradients (depth, temperature), biological gradients (coral cover, chlorophyll concentration), geological gradients (island slope, distances between neighboring islands), and anthropogenic gradients (human population count, human impacts).
-We hypothesized that the communities found across the latitudinal gradient encompassing all of these environmental variables would show fewer differences between individual islands or atolls, but would be greatest between the two regions, MHI and NWHI.
-Finally, to investigate whether a latitudinal gradient or depth gradient would be a stronger force in driving community composition, we compared the reef crab assemblage from across the shallow reefs in the Hawaiian Archipelago to reef crab assemblages across a mesophotic depth gradient.
--->
-
-<!-- The Hawaiian Archipelago is a volcanic island arc made up of high islands and atolls spanning ~9 degrees of latitude and ~2,500 km.
-It is typically regarded as two regions: the inhabited main Hawaiian Islands (MHI), which include the eight islands from Hawai&#x02BB;i Island to Kauai&#x02BB;i, and the largely uninhabited Northwestern Hawaiian Islands (NWHI), which constitute the 10 low islands and atolls to the northwest of Kaua&#x02BB;i.
-We compared shallow reef crab communities across the 2500 km archipelago and  we compared reef crab assemblages from the shallow reefs to the reef crab assemblages across a mesophotic depth gradient off of Kewalo, Oʻahu.
-By controlling for habitat area and complexity and by incorporating sites across various environmental and human impact gradients, we sought to better understand whether latitude, the environment, or depth is more influential in structuring reef crab communities.
--->
 
@@
 ARMS units were retrieved by encapsulating them in-situ in a 100 μm mesh-lined crate to prevent escape of motile organisms.
-Once returned to the surface, they were placed in containers of aerated, filtered seawater, crates were removed, and units were systematically disassembled plate-by-plate.
-Seawater surrounding disassembled units was sieved through 2 mm, 500 μm, and 100 μm mesh geologic sieves [see @lerayDNABarcodingMetabarcoding2015 for processing details], and retained macroorganisms were preserved in 95% ethanol.
-All adult brachyuran crabs ($\geq$ 5 mm) were photographed, identified (to species where possible, otherwise morphospecies), and preserved in 95% ethanol.
-Specimens were collected under permits [#xxx]{.mark} (NWHI) and [#xxx]{.mark} (MHI) and accessioned into the invertebrate zoology collections at the Florida Museum of Natural History ([catalog numbers xxx--xxx]{.mark}).
+Once returned to the surface, they were placed in containers of aerated, filtered seawater, crates were removed, and units were systematically disassembled plate-by-plate [see @lerayDNABarcodingMetabarcoding2015 for processing details].
+All adult brachyuran crabs ($\geq$ 5 mm) were photographed, identified (to species where possible, otherwise morphospecies), and preserved in 95% ethanol.<!-- Specimens were collected under permits #xxx (NWHI) and #xxx (MHI) and accessioned into the invertebrate zoology collections at the Florida Museum of Natural History (catalog numbers xxx--xxx).-->
 
@@
 NWHI).
-Island group comparisons were performed both with and without mesophotic samples.
+Island group comparisons were performed both including and excluding mesophotic samples.
 <!--To explore environmental influences on alpha diversity, we used a linear mixed-effect model with Simpson or richness as response variables, the environmental variables listed below as fixed effects, and island nested within island group (MHI vs NWHI) as random effects.
@@
 Significance was assessed with permutation tests in both cases.
-Variation among sampling groups (island group and depth) was further examined and visualized using canonical analysis of principal coordinates (CAP) based on discriminant analysis [@andersonCanonicalAnalysisPrincipal2003] as implemented in the `CAPdiscrim` function in `BiodiversityR` [@kindtTreeDiversityAnalysis2005].
+Variation among sampling groups (island group and sampling depth) was further examined and visualized using canonical analysis of principal coordinates (CAP) based on discriminant analysis [@andersonCanonicalAnalysisPrincipal2003] as implemented in the `CAPdiscrim` function in `BiodiversityR` [@kindtTreeDiversityAnalysis2005].
 
@@
 The four most abundant species were: `r with(all_summary,top_list(top,top_n,top_units,fig="fig_crabs"))`.
-`r str_to_sentence(english(all_summary$singleton_count))` species occurred in a single unit across all sites (singletons), accounting for ~`r num(all_summary$singleton_count/ntaxa(cc$crabs_untransformed)*100,1)`% of total recovered crab species.
+There were `r all_summary$singleton_count` singleton taxa (morpho/species occurring in a single unit across all sites), accounting for ~`r num(all_summary$singleton_count/ntaxa(cc$crabs_untransformed)*100,1)`% of total recovered crab species.
+<!--`r str_to_sentence(english(all_summary$singleton_count))` species occurred in a single unit across all sites (singletons), accounting for ~`r num(all_summary$singleton_count/ntaxa(cc$crabs_untransformed)*100,1)`% of total recovered crab species.-->
 
@@
 The `r nsamples(cc$crabs_deep)` units deployed at mesophotic depths $\geq$ 30 m on O&#x02BB;ahu yielded fewer species and individuals compared to the complete dataset, with `r num(sum(sample_sums(cc$crabs_deep)),1)` crabs overall, `r sum(unlist(deep_summary$types_breakdown))` distinct types (`r deep_summary$types_breakdown$species` species and `r deep_summary$types_breakdown$morphospecies` morphospecies), `r num(deep_summary$abundance_mean,0.1)` $\pm$ `r num(deep_summary$abundance_sd,0.1)` individuals per unit, and abundance ranging from `r format_range(deep_summary$abundance_range)`.
-The four most abundant species in the deep ARMS did not overlap with those from the complete dataset: `r with(deep_summary,top_list(top,top_n,top_units))`.
+The four most abundant taxa in the deep ARMS did not overlap with those from the complete dataset: `r with(deep_summary,top_list(top,top_n,top_units))`.
 There were `r deep_summary$singleton_count` singletons, which proportionally accounted for ~`r num(deep_summary$singleton_count/ntaxa(cc$crabs_deep)*100,1)`% of species found on deep ARMS.
@@
 Richness was not significantly different between shallow and deep samples ([@fig:fig_alpha[]{}b]; `r k_test(all_alpha$k_r_depth,plain=TRUE)`) but Simpson diversity was ([@fig:fig_alpha[]{}d]; `r k_test(all_alpha$k_s_depth,plain=TRUE)`).
-When only shallow arms \< 30 m were considered, neither comparison was significant (richness: `r k_test(shallow_alpha$k_r_island,plain=TRUE)`, [@fig:fig_alpha2[]{}a]; Simpson: `r k_test(shallow_alpha$k_s_island,plain=TRUE)`, [@fig:fig_alpha2[]{}b]).
+When only shallow arms \< 30 m were considered, there were no significant differences in alpha diversity between the MHI and NWHI for either index (richness: `r k_test(shallow_alpha$k_r_island,plain=TRUE)`, [@fig:fig_alpha2[]{}a]; Simpson: `r k_test(shallow_alpha$k_s_island,plain=TRUE)`, [@fig:fig_alpha2[]{}b]).
 
@@
 <!-- ### Effects of environmental variables {-} -->
-
 ## Brachyuran community structure {-}
@@
 db_modl <- reduce_model(db_modl,vif_limit=3)
+
+# calculate biplot vectors and magnitudes
+vectors <- db_modl %>%
+  scores(display="bp") %>%
+  as_tibble(rownames="term") %>%
+  inner_join(model_terms %>% select(variable,driver=`Environmental Driver`),by=c("term" = "variable")) %>%
+  mutate(size = sqrt(dbRDA1^2 + dbRDA2^2), driver=str_to_lower(driver)) %>%
+  arrange(desc(size))
 
@@
 
-Environmental variables were strongly clustered according to island group (MHI vs NWHI), with the separation primarily along the direction influenced by human impact, coral cover, sea-surface temperature, and chlorophyll ([@fig:fig_pca]).
+Environmental variables were strongly clustered according to island group (MHI vs NWHI) ([@fig:fig_pca]).
 The best dbRDA model (`r permanova(db_modl,stat="F")`) explained `r pc(RsquareAdj(db_modl)$adj.r.squared,0.01)` (adjusted $\mathrm{R}^2$) of the variability in crab communities across the archipelago ([@fig:fig_dbrda]).
-`r str_to_sentence(english(nterm))` of the seven environmental variables tested were significant and sufficiently uncorrelated.
-The two drivers contributing most to explained variation were `r drivers$term[1]` (F = `r num(drivers$stat[1],0.01)`) and `r drivers$term[2]` (F = `r num(drivers$stat[2],0.01)`) ([@tbl:rda_table]).
-<!--TODO: perhaps include discussion of the vector lengths here as well, since chl-a is the shortest vector despite being the second most-significant factor.-->
+`r str_to_sentence(english(nterm))` of the seven environmental variables tested were significant and sufficiently uncorrelated.<!--TODO: make "chlorophyll-a" look like "chlorophyll-A"--> The two drivers contributing most to explained variation were `r drivers$term[1]` (F = `r num(drivers$stat[1],0.01)`) and `r drivers$term[2]` (F = `r num(drivers$stat[2],0.01)`) ([@tbl:rda_table]), although the two factors with the longest effect vectors on the dBRDA plot were `r vectors$driver[1]` and `r vectors$driver[2]` ([@fig:fig_dbrda]).
 
@@
 cap_var <- list(x = cap_var[1], y = cap_var[2])
-```
-Shallow crab community structure was significantly different between island groups (MHI vs.
-NWHI) (`r permanova(anova_region,0.01)`) as well as among individual islands (`r permanova(anova_island,0.01)`; see [@fig:fig_pairwise] for results of pairwise comparisons).
-There was no significant difference in community dispersion between the MHI and NWHI (`r permdisp(bd_region)`).
-When mesophotic ($\geq$ 30 m) ARMS units were included, there were significant differences among depth zones (\< 30 m, 30 m, 60 m, 90 m) (`r permanova(anova_depth,0.01)`) and community dispersion (`r permdisp(bd_depth)`).
-However, community dispersion was not different among depth zones $\geq$ 30 m (`r permdisp(bd_deep)`).
-Although crab communities were significantly different among both depth zones and island groups for the complete dataset (`r permanova(anova_both,0.01)`), depth explained `r num(effect_ratio,0.1)` times more variability than island group (based on model $\mathrm{R}^2$).
-
-In the CAP analysis of island group and depth zone, ARMS units were correctly reassigned to groups `r pc(cap_depth$percent/100,0.1)` of the time.
-The first two axes explained `r pc(cap_var$x,0.1)` and `r pc(cap_var$y,0.1)` of the variation, respectively.
-Mesophotic samples (30--90 m) from O&#x02BB;ahu grouped separately from shallow samples collected across the entire ~2,500 km expanse of the archipelago ([@fig:fig_cap]).
-Additionally, variability among 30--90 m samples was visually equivalent to that among all shallow samples, although that was largely due to differences between 30 m and deeper sample groups ([@fig:fig_cap]).
-
-```{r fig-cap, fig.cap="CAP plot of crab communities by island group and depth zone.", results="asis", fig.width=9, fig.height=6, dependson="cap-analysis"}
+
 # make the sample group look nice
@@
   )
+
+shallow_dist <- pcap %>%
+  filter(shallow_deep == "shallow") %>%
+  select(sample,LD1,LD2) %>%
+  column_to_rownames("sample") %>%
+  dist()
+
+deep_dist <- pcap %>%
+  filter(shallow_deep == "deep") %>%
+  select(sample,LD1,LD2) %>%
+  column_to_rownames("sample") %>%
+  dist()
+```
+Shallow crab community structure was significantly different between island groups (MHI vs.
+NWHI) (`r permanova(anova_region,0.01)`) as well as among individual islands (`r permanova(anova_island,0.01)`; see [@fig:fig_pairwise] for results of pairwise comparisons).
+There was no significant difference in community dispersion between the MHI and NWHI (`r permdisp(bd_region)`).
+When mesophotic ($\geq$ 30 m) ARMS units were included, there were significant differences among depth zones (\< 30 m, 30 m, 60 m, 90 m) (`r permanova(anova_depth,0.01)`) and community dispersion (`r permdisp(bd_depth)`).
+However, community dispersion was not different among depth zones $\geq$ 30 m (`r permdisp(bd_deep)`).
+Although crab communities were significantly different among both depth zones and island groups for the complete dataset (`r permanova(anova_both,0.01)`), depth explained `r num(effect_ratio,0.1)` times more variability than island group (based on model $\mathrm{R}^2$).
+
+In the CAP analysis of island group and depth zone, ARMS units were correctly reassigned to groups `r pc(cap_depth$percent/100,0.1)` of the time.
+The first two axes explained `r pc(cap_var$x,0.1)` and `r pc(cap_var$y,0.1)` of the variation, respectively.
+Mesophotic samples (30--90 m) from O&#x02BB;ahu grouped separately from shallow samples collected across the entire ~2,500 km expanse of the archipelago ([@fig:fig_cap]).
+Additionally, variability among 30--90 m samples was greater than among all shallow samples, (max.
+Euclidean distance between shallow samples in the CAP ordination: `r num(max(shallow_dist),0.01)`, max.
+Euclidean distance between deep samples: `r num(max(deep_dist),0.01)`) ([@fig:fig_cap]).
+
+```{r fig-cap, fig.cap="CAP plot of crab communities by island group and depth zone.", results="asis", fig.width=9, fig.height=6, dependson="cap-analysis"}
 centroids <- pcap %>% group_by(sample_group) %>%
@@
 # Discussion {-}
-Marine communities are structured by a suite of factors, but there remains no consensus for generalized mechanisms for coral reefs<!-- structuring communities in shallow coral reefs-->.
-Classic studies from the intertidal point to biotic factors determining lower limits and abiotic factors setting upper limits of vertical species distributions [@connellEffectsCompetitionPredation1961; @connellInfluenceInterspecificCompetition1961; @dennyPhysicalProcessesThat2001; @paineIntertidalCommunityStructure1974].
-However, these conclusions describe distributions at individual sites and generalizations about the mechanisms and ecological processes that structure coral reef communities remain a subject of debate [@bellwoodEnvironmentalGeometricConstraints2005; @bellwoodSleepingFunctionalGroup2006; @carrBiodiversityPopulationRegulation2002; @connollyIndoPacificBiodiversityCoral2003; @dornelasCoralReefDiversity2006; @grahamImportanceStructuralComplexity2013; @komyakovaRelativeImportanceCoral2013].
-In the Hawaiian archipelago, we found significant differences in the composition of cryptic crab communities between the populated high islands of the MHI and the uninhabited atolls of the NWHI.
+Marine communities are structured by a suite of factors, but there remains no consensus for either generalized mechanisms or the relative importance of drivers of community structure for coral reefs.
+Our best knowledge of the role of depth in structuring marine communities comes from the intertidal, and points to biotic factors (e.g., predation, competition) determining lower limits and abiotic factors (e.g., desiccation, thermal stress) setting upper limits of vertical species distributions [@connellEffectsCompetitionPredation1961; @connellInfluenceInterspecificCompetition1961; @dennyPhysicalProcessesThat2001; @paineIntertidalCommunityStructure1974].
+However, abiotic stressors on intertidal communities are obviously different from those on coral reefs, and generalizations about the mechanisms and ecological processes that structure coral reef communities remain a subject of considerable debate [@bellwoodEnvironmentalGeometricConstraints2005; @bellwoodSleepingFunctionalGroup2006; @carrBiodiversityPopulationRegulation2002; @connollyIndoPacificBiodiversityCoral2003; @dornelasCoralReefDiversity2006; @grahamImportanceStructuralComplexity2013; @komyakovaRelativeImportanceCoral2013].
+In the Hawaiian archipelago, we found significant differences in the composition of coral reef crab communities between the populated high islands of the MHI and the uninhabited atolls of the NWHI.
 Moreover, these differences were driven by a combination of environmental (sea-surface temperature, chlorophyll concentration, depth), biotic (potential larval immigration), and anthropogenic (cumulative human impacts) factors.
-<!--Alpha diversity metrics show slight differences in the number of species and overall abundance of crabs in each sampled assemblage between the Main and Northwestern Hawaiian Islands, but CAP and PERMANOVA highlight significant differences in species assemblages between the regions.
-Differences in the species assemblages across regions are driven primarily by human impacts, cholorophyll a and sea surface temperature, although other factors such as larval connectivity, distance to nearest island, mean coral cover, and depth are also significant.-->
-
-The Hawaiian Archipelago is among the most remote on the planet and encompasses a range of gradients (anthropogenic, biotic, abiotic), making it an ideal area to study the impacts of these factors on species assemblages across space.
-Human impacts are considerably higher in the populated MHI than in the protected and remote NWHI [@selkoeEvaluatingAnthropogenicThreats2008; @selkoeMapHumanImpacts2009; @selkoeDNACoralReef2016].
-Although we did not consider individual effects here, an index representing cumulative anthropogenic impacts was the top driver of brachyuran community structure, similar to findings in previous work examining genetic diversity across the archipelago [@selkoeDNACoralReef2016].
-The next-most significant driver, chlorophyll a, can be used as a proxy for primary productivity.
-Productivity gradients have been reported across the Hawaiian Archipelago, which coincide with differences in coral reef community composition [@selkoeDNACoralReef2016], endemic reef fish dominance [@friedlanderDominanceEndemicsReef2020], and survival rates of top predators [@bakerEffectVariableOceanic2007].
-It is not surprising, therefore, that we find differences in the brachyuran crab communities are driven by these same factors across the Hawaiian archipelago.
-
-Despite the fact that shallow ARMS units were deployed within a narrow depth range (8--17 m), depth still emerged as a significant driver of community variation.
-Depth in and of itself does not affect anything [[cite?]]{.mark}, but at the same time, the factors that are known to covary most consistently with it (viz.
-temparature and light) do not show much variation across such a narrow range.
-However, we also did not sample these values for the specific depths where units where deployed.
-So, depth is clearly doing something.
-We only have environmental data from units deployed \< 30 m, but community ordinations show that there are considerable differences between all the shallow units across the entire span of the archipelago and the $\geq$ 30 m units collected from a single island.
+
+The Hawaiian Archipelago is among the most remote oceanic island chains on the planet and encompasses a range of gradients (anthropogenic, biotic, abiotic), making it an ideal natural laboratory to study the impacts of various drivers of species assemblages across space.
+Human activities are considerably higher in the populated MHI than in the protected and remote NWHI [@selkoeEvaluatingAnthropogenicThreats2008; @selkoeMapHumanImpacts2009; @selkoeDNACoralReef2016].
+Although we did not consider the effects of individual drivers here, an index representing cumulative anthropogenic impacts was the top driver of brachyuran community structure, similar to findings in previous work examining genetic diversity, species richness and habitat area, quality and stability across the archipelago [@selkoeDNACoralReef2016].
+The next-most significant driver, chlorophyll A, can be used as a proxy for primary productivity.
+Productivity gradients across the Hawaiian Archipelago coincide with differences in coral reef community composition [@selkoeDNACoralReef2016], endemic reef fish dominance [@friedlanderDominanceEndemicsReef2020], and survival rates of top predators [@bakerEffectVariableOceanic2007].
+Thus, it is not surprising that we find the same top drivers of community structure for one of the primary taxonomic groups inhabiting coral reefs, the brachyuran crabs, across the Hawaiian archipelago.
+What is surprising is the consistent significance of depth as a primary driver of community composition across these gradients.
+
+Even though our shallow-water ARMS units were deployed within a narrow depth range (8--17 m), depth still emerged as a significant and consistent driver of community variation among these units.
+Depth in and of itself is unlikely to affect marine community composition.
+At the same time, oceanographic factors with well-known depth profiles (e.g., temperature, light, oxygen, nutrients, etc.) are not expected to vary dramatically across such a narrow depth range (\< 10 m).
+Unfortunately, we lack environmental data from the specific deployment depths of our shallow-water ARMS sites, and cannot unravel the co-varying factor(s) associated with the depth effects we observed.
+Regardless, given its significance across both narrow (~10 m) and wider (~100 m) depth ranges sampled here, depth (or an associated factor) is clearly an important driver for reef crab communities in across the Hawaiian Archipelago.
+Given brachyurans are a ubiquitous and important component of coral reef communities [@reynsPatternsProcessesBrachyuran1999; @lasleyjrPhylogeneticRelationshipsUbiquitous2015; @hazeriLatitudinalSpeciesDiversity2019], these findings are likely to apply more broadly as well.
+
+While depth was a significant driver of community composition among shallow-water ARMS units, its influence was even more apparent when comparing shallow to mesophotic samples.
+Due to logistical constraints (and the limitations of satellite-derived data products), our environmental data only covers shallow-water deployments, but CAP ordinations show considerable differences between the grouping of all shallow communities and the units deployed $\geq$ 30 m at the single island of O&#x02BB;ahu ([@fig:fig_cap]).
+In other words, there was more community variation across 90 m of depth on a single island than across 2,500 km of shallow habitat experiencing multiple gradients of environmental, ecological, and anthropogenic influence.
+The ordination revealed two visible clusters for mesophotic units: one containing the 30 m units and the other containing the 60 m and 90 m units.
+These two clusters coincide with what is regarded as the upper and lower mesophotic zones, respectively [@kahngCommunityEcologyMesophotic2010; @rochaMesophoticCoralEcosystems2018; but see @pyleMesophoticCoralEcosystems2019].
+The upper mesophotic (~30--60 m) has often been considered as a continuation of shallow reefs while the lower mesophotic (~60--150 m) is primarily thought to constitute a unique community.
+This pattern has held true across various taxonomic groups [@lesserGlobalCommunityBreaks2019; @rochaMesophoticCoralEcosystems2018], but our data show more distinct separations among shallow, upper, and lower mesophotic depths (although 30 m appears to represent an upper-to-lower transition zone).
+
+In addition to community composition, mesophotic samples also contributed significantly to alpha diversity differences across the archipelago.
+We found significant differences in Simpson diversity between the MHI and NWHI, but only when mesophotic samples were included in the comparison; when deep ARMS units were left out, these same alpha diversity comparisons were non-significant.
+When comparing shallow to mesophotic, Simpson diversity was significantly (if slightly) higher in deep communities.
+Thus, the apparent differences in alpha diversity between the MHI and NWHI can be attributed entirely to differences between shallow and deep samples on the island of O&#x02BB;ahu.
+This pattern is likely driven by species unique to mesophotic habitat [@hurleyAssessmentShallowMesophotic2016].
+
+Our depth transect was limited to O&#x02BB;ahu, but community and alpha diversity differences among depth zones at that single island appear to overwhelm differences observed among shallow-water coral reef habitat across the archipelago.
+Similar patterns were found at a local scale (tens of kilometers) on Hawai&#x02BB;i Island using environmental DNA (eDNA) [@hobanPlumbingDepthsEnvironmental2023].
+In that study, communities were found to have more affinity across sites within depth zones than among depth zones within sites.
+These results and the patterns we observed suggest that depth-associated factors exert stronger ecological influence than those that vary across the latitudinal, physical, and anthropogenic gradients throughout the Hawaiian Archipelago.
+
+Since we lack environmental data for mesophotic samples, we cannot make conclusions about the specific influences driving the differences we found.
+However, light and temperature have been proposed to be important with regard to depth zonation in benthic communities [@kahngCommunityEcologyMesophotic2010; @lesserGlobalCommunityBreaks2019; @pyleComprehensiveInvestigationMesophotic2016; @pyleMesophoticCoralEcosystems2019].
+Oceanographic conditions may be inconsistent close to shore, but the average mixed-layer depth calculated (using 0.8 &deg;C $\upDelta$T [@karaOptimalDefinitionOcean2000]) from CTD casts taken at station ALOHA (22.75&deg;N, 158.00&deg;W; within the latitudinal expanse of the MHI) was 84.6 $\pm$ 26 m over the course of the year (based on monthly averages from casts taken across the duration of ARMS unit deployments [@HawaiiOceanTimeseries]).
+Thus, given the assumption of similar temperature profiles near the O&#x02BB;ahu shoreline, temperature alone appears insufficient to explain the zonation patterns we observed.
+In contrast, the lower depth boundaries of mesophotic reef communities are largely determined by surface irradiance [@kahngCommunityEcologyMesophotic2010; @hindersteinThemeSectionMesophotic2010], so light seems a possible candidate to be a significant abiotic driver of community structure.
+
+Further work to investigate drivers of depth zonation could take advantage of established gradients and use the Hawaiian Archipelago as a natural laboratory.
+Coral reef biodiversity is generally thought to be structured at depth by a combination of the attenuation of light, a decrease in temperature, increasing salinity, and a reduction in trophic resources [@lesserEcologyMesophoticCoral2009; @kahngCommunityEcologyMesophotic2010; @slatteryFunctionStabilityMesophotic2024a].
+If patterns of community structure are driven primarily by light, that may explain why even our shallow-water sites show a significant depth effect over less than 10m.
+The importance of light could be tested by sampling communities from the same irradiance level but different depths due to factors altering light penetration to determine if we see consistent patterns of community zonation.
+In contrast, other factors (such as dissolved inorganic nitrogen, particulate organic carbon, or productivity) expected to have more strongly-defined gradients across the regions [@slatteryFunctionStabilityMesophotic2024a], would lead to variable zonation patterns that should be detected in our comparison of the MHI & NWHI.
+Such differences should be tested in future studies to understand the specific physical and ecological drivers of coral reef depth zonation.
+
+Regardless of the exact environmental and ecological drivers at play, our findings suggest that the classical model of intertidal zonation with biotic and abiotic drivers could likely be reversed for coral reef ecosystems.
+*Lower* limits of mesophotic coral reef ecosystems are likely determined by abiotic factors that correlate with depth, whereas *upper* limits may be driven by ecological interactions.
+By comparing shallow-water ARMS across the latitudinal, physical and anthropogenic gradients of the Hawaiian Archipelago, we can test the relative importance of the significant drivers of community structure.
+Our analysis corroborates previous work and shows depth is the strongest driver of community structure.
+We find that a 90 m depth transect at a single site shows greater divergence in canonical space than the full 2,500 km gradient of the Hawaiian Archipelago for shallow-water ARMS.
+We conclude that some factor(s) that covary with depth exert more influence on the cryptobenthic coral reef community structure than those biotic and abiotic gradients that span the entire Hawaiian Archipelago in shallow-water environments.
 
@@
 
+\newpage
+
 # Supplemental material {label="S"}
@@
 ## Processing of satellite-derived environmental data
-Level-3 mapped (9km) sea-surface temperature (SST) and chlorophyll A (chl-A) data were retrieved from NASA's Physical Oceanography Distributed Active Archive Center [PO.DAAC; @NASAJPL_2020] and Ocean Biology Distributed Active Archive Center [OB.DAAC; @NASAOceanBiologyProcessingGroup_2022] respectively.
+Level-3 mapped (9km) sea-surface temperature (SST) and chlorophyll A (chl-A) data were retrieved from NASA's Physical Oceanography Distributed Active Archive Center [PO.DAAC; @https://doi.org/10.5067/modsa-mo9d9] and Ocean Biology Distributed Active Archive Center [OB.DAAC; @NASAOceanBiologyProcessingGroup_2022] respectively.
 Monthly summary datasets were downloaded for the period of 2009--2016 and averaged over each of the ARMS unit soaking periods (e.g.
@@
 ```{r fig-alpha, fig.cap='Box plots of (a,b) species richness and (c,d) Simpson diversity for complete dataset (all sampled depths).
+Asterisk denotes significant difference (K-W test).
 (a) Species richness by island group.
@@
   mutate(shallow_deep = str_to_sentence(shallow_deep),shallow_deep = fct_reorder(shallow_deep,depth))
+
+if (all_alpha$k_r_island$p.value < 0.05) {
+  which_max <- art %>% 
+    group_by(island_group) %>% 
+    summarise(r = mean(Observed)) %>%
+    slice(which.max(r)) %>%
+    pull(island_group) %>%
+    as.character()
+  max_r <- max(art$Observed) + (diff(range(art$Observed)))/10
+  sig <- tribble(
+    ~x,~y,
+    which_max,max_r
+  ) 
+} else {
+  sig <- tribble(~x,~y)
+}
+
+pointsize <- diff(range(art$Observed))
+
 p1 <- ggplot(art) + 
   geom_boxplot(aes(x=island_group,y=Observed,fill=island_group),notch=TRUE) +
+  geom_point(aes(x,y),data=sig,shape=pointsize,size=1,stroke=1.5) +
   scale_fill_manual(values=island_group_pal) +
@@
   ylab("Richness")
+
+if (all_alpha$k_r_depth$p.value < 0.05) {
+  which_max <- art %>% 
+    group_by(shallow_deep) %>% 
+    summarise(r = mean(Observed)) %>%
+    slice(which.max(r)) %>%
+    pull(shallow_deep) %>%
+    as.character()
+  max_r <- max(art$Observed) + (diff(range(art$Observed)))/10
+  sig <- tribble(
+    ~x,~y,
+    which_max,max_r
+  ) 
+} else {
+  sig <- tribble(~x,~y)
+}
+
 p2 <- ggplot(art) + 
   geom_boxplot(aes(x=shallow_deep,y=Observed,fill=shallow_deep),notch=TRUE) +
+  geom_point(aes(x,y),data=sig,shape=8,size=pointsize,stroke=1.5) + 
   scale_fill_manual(values=sample_depth_pal) +
@@
 
+if (all_alpha$k_s_island$p.value < 0.05) {
+  which_max <- art %>% 
+    group_by(island_group) %>% 
+    summarise(r = mean(Simpson)) %>%
+    slice(which.max(r)) %>%
+    pull(island_group) %>%
+    as.character()
+  max_r <- max(art$Simpson) + (diff(range(art$Simpson)))/10
+  sig <- tribble(
+    ~x,~y,
+    which_max,max_r
+  ) 
+} else {
+  sig <- tribble(~x,~y)
+}
+
+pointsize <- diff(range(art$Simpson))*5
+
 p3 <- ggplot(art) + 
   geom_boxplot(aes(x=island_group,y=Simpson,fill=island_group),notch=TRUE) +
+  geom_point(aes(x,y),data=sig,shape=8,size=pointsize,stroke=1.5) + 
   scale_fill_manual(values=island_group_pal) +
@@
   ylab("Simpson's diversity index")
+
+if (all_alpha$k_s_depth$p.value < 0.05) {
+  which_max <- art %>% 
+    group_by(shallow_deep) %>% 
+    summarise(r = mean(Simpson)) %>%
+    slice(which.max(r)) %>%
+    pull(shallow_deep) %>%
+    as.character()
+  max_r <- max(art$Simpson) + (diff(range(art$Simpson)))/10
+  sig <- tribble(
+    ~x,~y,
+    which_max,max_r
+  ) 
+} else {
+  sig <- tribble(~x,~y)
+}
+
 p4 <- ggplot(art) + 
   geom_boxplot(aes(x=shallow_deep,y=Simpson,fill=shallow_deep),notch=TRUE) +
+  geom_point(aes(x,y),data=sig,shape=8,size=pointsize,stroke=1.5) + 
   scale_fill_manual(values=sample_depth_pal) +
@@
 ```{r fig-alpha2, fig.cap='Box plots of (a) species richness and (b) Simpson diversity by island group for shallow samples only.
-Notches as in [@fig:fig_alpha].', results="asis", dev="svg", fig.width=8, fig.height=4, dependson="init"}
+Notches as in [@fig:fig_alpha].
+Both comparisons are non-significant.', results="asis", dev="svg", fig.width=8, fig.height=4, dependson="init"}
 art <- shallow_alpha$richness_table %>% 
@@
 
+if (shallow_alpha$k_r_island$p.value < 0.05) {
+  which_max <- art %>% 
+    group_by(island_group) %>% 
+    summarise(r = mean(Observed)) %>%
+    slice(which.max(r)) %>%
+    pull(island_group) %>%
+    as.character()
+  max_r <- max(art$Observed) + (diff(range(art$Observed)))/10
+  sig <- tribble(
+    ~x,~y,
+    which_max,max_r
+  ) 
+} else {
+  sig <- tribble(~x,~y)
+}
+
+pointsize <- diff(range(art$Observed))
+
 p1 <- ggplot(art) + 
   geom_boxplot(aes(x=island_group,y=Observed,fill=island_group),notch=TRUE) +
+  geom_point(aes(x,y),data=sig,shape=8,size=pointsize,stroke=1.5) + 
   scale_fill_manual(values=island_group_pal) +
@@
 
+if (shallow_alpha$k_s_island$p.value < 0.05) {
+  which_max <- art %>% 
+    group_by(island_group) %>% 
+    summarise(r = mean(Simpson)) %>%
+    slice(which.max(r)) %>%
+    pull(island_group) %>%
+    as.character()
+  max_r <- max(art$Simpson) + (diff(range(art$Simpson)))/10
+  sig <- tribble(
+    ~x,~y,
+    which_max,max_r
+  ) 
+} else {
+  sig <- tribble(~x,~y)
+}
+
+pointsize <- diff(range(art$Simpson))*5
+
 p2 <- ggplot(art) + 
   geom_boxplot(aes(x=island_group,y=Simpson,fill=island_group),notch=TRUE) +
+  geom_point(aes(x,y),data=sig,shape=8,size=pointsize,stroke=1.5) + 
   scale_fill_manual(values=island_group_pal) +
@@
   geom_point(aes(fill=island_group),shape=21,color="black",size=4) +
-  stat_ellipse(geom="polygon",aes(group=island_group,fill=island_group),color="black",alpha=0.1,show.legend = FALSE, size=0.07) +
+  stat_ellipse(geom="polygon",aes(group=island_group,fill=island_group),color=NA,alpha=0.1,show.legend = FALSE) +
   # geom_mark_ellipse(color="black",alpha=0.2,show.legend=FALSE, linewidth=0.07) +
@@
 
-```{r pairwise-island, fig.cap='Significance levels of pairwise PERMANOVA tests by island for shallow crab communities (p-values adjusted according to the false discovery rate [@benjaminiControllingFalseDiscovery1995]).', results="asis", dev="svg", fig.width=12, fig.height=9, dependson="init"}
+```{r pairwise-island, fig.cap='Results of pairwise PERMANOVA tests by island for shallow crab communities (p-values adjusted according to the false discovery rate [@benjaminiControllingFalseDiscovery1995]).
+Significant comparisons presented in bold.', results="asis", dev="svg", fig.width=12, fig.height=9, dependson="init"}
 pairwise_island <- pairwise_adonis(cc$crab_dist_shallow,cc$crab_data_shallow$hawaiian_name,permutations=PERM)
@@
 nasa_table <- read_csv(here("data","tables","nasa.csv"))
-print_table(nasa_table,"nasa_table","NASA data repository filenames for CHL and SST satellite data",accuracy=1) 
-```
+print_table(nasa_table,"nasa_table","NASA data repository filenames for CHL and SST satellite data",accuracy=1,hdr_case=NA,format_numbers=FALSE) 
+```

### 2024-03-18 crabs.R

@@ -1,26 +1,3 @@
-# required libraries
-# all of these can be directly installed except for phyloseq, which is a bioconductor package
-# phyloseq isn't directly necessary, but it's what I used for all my eDNA stuff and it conveniently
-# wraps the datasets, so I'm using it here
-# library(tidyverse)
-# to install phyloseq in regular R, do this:
-
-# if (!require("BiocManager", quietly = TRUE))
-#   install.packages("BiocManager")
-
-# BiocManager::install("phyloseq")
-# library(phyloseq) 
-# library(vegan)
-# library(ggrepel)
-# library(here)
-# library(worrms)
-
-# to check model assumptions
-# library(performance)
-
-# for linear mixed effects models
-# library(lme4)
-
 # set the random seed
 set.seed(31337)
 

### 2024-05-16 manuscript.Rmd
Converted citations to csl-json, minor text updates

@@
 ---
-title: "Hawaiian reef communities are more distinct across a 90 m depth gradient than 2,500 km of shallow habitat"
+title: 'Cryptobenthic reef assemblages are more distinct across a 90 m depth gradient than 2,500 km of shallow marine habitat in the Hawaiian Archipelago'
 author:
@@
         - himb
-        - nerr
       equal_contributor: yes
@@
       equal_contributor: yes
+  - Kerry Reardon:
+      institute: noaa
+  - Derek J. Skillings:
+      institute: 
+        - uncg
   - Molly A. Timmers:
@@
         - natgeo
-  - Derek J. Skillings:
-      institute: 
-        - uncg
-  - Kerry Reardon:
-      institute: noaa
   - Robert J. Toonen:
@@
 institute:
-  - himb: Hawai&#x02BB;i Institute of Marine Biology
-  - nerr: He&#x02BB;eia National Estuarine Research Reserve
+  - himb: Hawai&#x02BB;i Institute of Marine Biology, University of Hawai&#x02BB;i, Mānoa
+  - noaa: National Oceanic and Atmospheric Administration
+  - uncg: 'University of North Carolina, Greensboro'
   - natgeo: National Geographic Pristine Seas
-  - uncg: 'University of North Carolina, Greensboro'
-  - noaa: National Oceanic and Atmospheric Administration
 abstract:
@@
 Furthermore, due to logistical difficulties, investigations of the role of depth (particularly mesophotic coral ecosystems) in influencing coral reef community structure have been taxonomically limited.
-Here, we examined the effects of a suite of environmental, ecological, and anthropogenic factors on the community structure of shallow brachyuran crabs across the ~2,500 km span of the Hawaiian Archipelago.
-In addition, we compared shallow crab communities with those sampled on a depth transect from 12--90 m on O&#x02BB;ahu.
+Here, we examined the effects of a suite of environmental, ecological, and anthropogenic factors on the assemblage structure of shallow brachyuran crabs across the ~2,500 km span of the Hawaiian Archipelago.
+In addition, we compared shallow crab assemblages with those sampled on a depth transect from 12--90 m on O&#x02BB;ahu.
 We deployed Autonomous Reef Monitoring Structures (ARMS) on shallow reefs (12--16 m) across the latitudinal gradient of the Hawaiian Archipelago to compare with a depth gradient (12, 30, 60, and 90 m) on the south shore of O&#x02BB;ahu.
-We found a total of 2,284 individual crabs representing 90 morphospecies and 19 families.
-Shallow crab communities were significantly different between the Main and Northwest Hawaiian Islands and variation in community structure was significantly associated with six biotic, environmental, and anthropogenic factors (cumulative human impacts, chlorophyll-A, sea-surface temperature, site depth, island slope, and potential larval connectivity).
-In depth comparisons, we found more variation among mesophotic samples from O&#x02BB;ahu (30--90 m) than among all shallow samples across the entire archipelago.
-We show that 90 m of depth is a more potent determinant of brachyuran community composition than the latitudinal, environmental, or anthropogenic gradients across the Hawaiian Islands.<!--Deeper reefs host significantly different brachyuran communities than shallow ones in Hawai&#x02BB;i, and only 3 of 90 morphospecies (~3%) occurring across the entire depth range (mesophotic to shallow) were sampled.
+We found a total of 2,284 individual crabs representing 103 (morpho)species in 60 genera and 22 families.
+Shallow crab assemblages were significantly different between the Main and Northwest Hawaiian Islands and variation in community structure was significantly associated with six biotic, environmental, and anthropogenic factors (cumulative human impacts, chlorophyll-A, sea-surface temperature, site depth, island slope, and potential larval connectivity).
+In depth comparisons, we found more variation among mesophotic samples from O&#x02BB;ahu (30--90 m) than among all shallow samples across the complete span of the archipelago.
+We show that 90 m of depth is a more potent determinant of brachyuran assemblage composition than the latitudinal, environmental, or anthropogenic gradients across the Hawaiian Islands.<!--Deeper reefs host significantly different brachyuran communities than shallow ones in Hawai&#x02BB;i, and only 3 of 90 morphospecies (~3%) occurring across the entire depth range (mesophotic to shallow) were sampled.
 Further research on the cryptic reef-dwelling fauna is needed for identifying areas of unique biodiversity, as well as boundaries between depth-stratified reef ecosystems.
@@
 Here we ask if MCEs host the same communities as the shallower reefs spread across the entire 2400 km Hawaiian Archipelago.-->
-bibliography: references/citations.bib
+bibliography: references/citations.json
 csl: references/style.csl
@@
 
-print_table <- function(tbl,id="",caption="",hdr_case="sentence",hide=character(0),accuracy = 0.01,format_numbers=TRUE) {
+print_table <- function(tbl,id="",caption="",hdr_case="sentence",hide=character(0),accuracy = 0.01,format_numbers=TRUE, cell_width = Inf) {
   if (!grepl("^: ",caption)) {
@@
 
-  table_str <- capture.output(pander::pandoc.table(tbl,keep.line.breaks=TRUE,caption=NULL,style="grid",split.cells=Inf,split.tables=Inf,justify='left',missing=''))
+  table_str <- pander::pandoc.table.return(
+    tbl,
+    keep.line.breaks=TRUE,
+    caption=NULL,
+    style="grid",
+    split.cells=cell_width,
+    split.tables=Inf,
+    justify='left',
+    missing=''
+  )
   table_str <- table_str[table_str != ""]
@@
 The structure and composition of coral reef communities is influenced by a suite of environmental and anthropogenic factors, yet the nature and makeup of those factors remain poorly resolved [@hughesCommunityStructureDiversity1989; @smithEffectsTopBottom2010; @brandlCoralReefEcosystem2019].
-Coral reefs are among the most diverse habitats in the world, and they face increasing threats of degradation, including climate-driven bleaching and phase shifts to alternate ecological regimes [@grahamManagingResilienceReverse2013; @norstromAlternativeStatesCoral2009; @pandolfiGlobalTrajectoriesLongTerm2003].
+Coral reefs are among the most diverse habitats in the world, and they face increasing threats of degradation, including climate-driven bleaching and phase shifts to alternate ecological regimes [@grahamManagingResilienceReverse2013; @norstromAlternativeStatesCoral2009; @pandolfiGlobalTrajectoriesLongterm2003].
 In the face of these threats, understanding drivers of metazoan community structure in coral reef ecosystems is critically important [@hughesRisingChallengeSustaining2010; @pageSeekingResistanceCoral2019], particularly as the search for criteria to prioritize climate refugia such as upwelling zone or mesophotic reefs intensifies.
@@
 The Hawaiian Archipelago comprises high volcanic islands and atolls stretching across ~2,500 km in the North Pacific Ocean and is the most isolated island chain on the planet.
-Typically regarded as two regions, the main Hawaiian Islands (MHI) include the eight inhabited high islands, and the Northwestern Hawaiian Islands (NWHI) comprise the uninhabited islands, atolls, and banks northwest of Kaua&#x02BB;i and Ni&#x02BB;ihau that make up the Papahānaumokuākea Marine National Monument.
+Typically regarded as two regions, the main Hawaiian Islands (MHI) include the eight inhabited high islands, and the Northwestern Hawaiian Islands (NWHI) comprise the uninhabited islands, atolls, and banks northwest of Kaua&#x02BB;i and Ni&#x02BB;ihau.
+The NWHI are wholly enclosed by the Papahānaumokuākea Marine National Monument, one of the largest marine protected areas in the world.
 The islands are subject to considerable anthropogenic and physical gradients across their span [@selkoeMapHumanImpacts2009], which are widely believed to influence coral reef communities.
-These gradients include differences in temperature, human population, and anthropogenic impact, among others.
-Studies examining community structure across broad environmental gradients have been criticized for confounding multiple potentially covarying factors (such as habitat area and/or complexity) across the measured gradients [e.g., @bellwoodRegionalScaleAssemblyRules2001; @grahamEffectsMarineReserve2003; @connollyCommunityStructureCorals2005; @ewersConfoundingFactorsDetection2006; @fletcherRoleHabitatArea2007a].
+These gradients include differences in temperature, coral cover, human population, and anthropogenic impact, among others.
+Studies examining community structure across broad environmental gradients have been criticized for confounding multiple potentially covarying factors (such as habitat area and/or complexity) across the measured gradients [e.g., @bellwoodRegionalscaleAssemblyRules2001; @grahamEffectsMarineReserve2003; @connollyCommunityStructureCorals2005; @ewersConfoundingFactorsDetection2006; @fletcherRoleHabitatArea2007].
 Thus, we used autonomous reef monitoring structures (ARMS)---standardized collection devices designed to mimic reef structural complexity---to sample cryptic communities while controlling for habitat [@zimmermanArtificialReefMatrix2004; @brainardAutonomousReefMonitoring2009].
-We then sampled brachyuran crabs, a group comprising a notable proportion of the coral reef cryptofauna [289 species in Hawai&#x02BB;i, @castroCatalogAnomuranBrachyuran2011] from these structures to examine drivers of community composition across these gradients.
+We then sampled brachyuran crabs, a group comprising a notable proportion of the coral reef cryptofauna [284 species in Hawai&#x02BB;i, @castroCatalogAnomuranBrachyuran2011] from these structures to examine drivers of community composition across these gradients.
 We compared crab communities among islands and regions (MHI vs.
@@
 While there are no shortages of exceptions, a long history of studies point towards a generalization that the lower bathymetric limits are generally regulated by ecological interactions whereas upper limits of intertidal communities are more often determined by abiotic factors [@stephenson1972life; @mengeRockyIntertidalCommunities2001; @bertnessTestingRelativeContribution1999; @dennyPhysicalProcessesThat2001].
-However, most of these models describe zonation at specific localities rather than community variation over larger spatial scales [but see @birdTideWaveMarks2013; @bishop-taylorTidesModellingElevation2019] and may not apply to coral reefs, which are subject to more complex community dynamics [@abramsRecruitmentLotteriesCoexistence1984; @bussCompetitiveNetworksNontransitive1979; @hixonPredationPreyRefuges1993; @hubbellUnifiedNeutralTheory2011; @moraEcologyFishesCoral2015; @saleStockRecruitRelationshipsRegional1982; @sheppardBiologyCoralReefs2017].
-While there is little doubt that ecological interactions are important on coral reefs, the suite of factors and which play the most important role in structuring coral reef communities remains a subject of considerable debate [@bellwoodRegionalScaleAssemblyRules2001; @connollyUnifiedModelExplains2017; @hubbellUnifiedNeutralTheory2011; @jacksonAdaptationDiversityReef1991; @mumbyCoralReefHabitats2008; @saleMaintenanceHighDiversity1977].
-
-One aspect of coral reef community structure that is thought to be important but remains under-investigatedis depth.
+However, most of these models describe zonation at specific localities rather than community variation over larger spatial scales [but see @birdTideWaveMarks2013; @bishop-taylorTidesModellingElevation2019] and may not apply to coral reefs, which are subject to more complex community dynamics [@abramsRecruitmentLotteriesCoexistence1984; @bussCompetitiveNetworksNontransitive1979; @hixonPredationPreyRefuges1993; @hubbellUnifiedNeutralTheory2011; @moraEcologyFishesCoral2015; @saleStockrecruitRelationshipsRegional1982; @sheppardBiologyCoralReefs2017].
+While there is little doubt that ecological interactions are important on coral reefs, the suite of factors and which play the most important role in structuring coral reef communities remains a subject of considerable debate [@bellwoodRegionalscaleAssemblyRules2001; @connollyUnifiedModelExplains2017; @hubbellUnifiedNeutralTheory2011; @jacksonAdaptationDiversityReef1991; @mumbyCoralReefHabitats2008; @saleMaintenanceHighDiversity1977].
+
+One aspect of coral reef community structure that is thought to be important but remains under-investigated is depth.
 Most studies to date have focused on the upper ~30 m of the water column, yet coral reefs and associated communities are known to extend beyond 150 m [@pyleExploringDeepCoral1996; @pyleMesophoticCoralEcosystems2019], with photosymbiotic corals reported to 170 m [@rouzeSymbioticAssociationsDeepest2021].
@@
 These habitats, known as mesophotic coral ecosystems (MCEs), are typically defined as those extending from ~30--150 m depth [@hindersteinThemeSectionMesophotic2010; @kahngCommunityEcologyMesophotic2010].
-They have been hypothesized as possible refugia from climate change impacts on shallow reefs [@glynnCoralReefBleaching1996; @bongaertsAssessingDeepReef2010; @smithDepthRefugiumCatastrophic2014; @holsteinFertileFathomsDeep2015; @muirSpeciesIdentityDepth2017], although a growing number of studies have found evidence that mesophotic reefs are unique and uniquely threatened, challenging this notion [@diazMesophoticCoralBleaching2023; @diazMesophoticCoralBleaching2023; @hobanPlumbingDepthsEnvironmental2023; @hurleyAssessmentShallowMesophotic2016; @kaneHighLevelsMesophotic2014; @pinheiroDeepReefFishes2019; @rochaMesophoticCoralEcosystems2018; @montgomeryCommunitySimilaritySpecies2021].
-Given the strong reported differences between deep and shallow reefs across taxa and the paucity of research on drivers of cryptic coral reef community variation, we sought to answer the following four research questions: (i) How do shallow crab communities vary along the natural and anthropogenic gradients of the Hawaiian Archipelago?
-(ii) Do shallow crab communities differ between the Main and Northwestern Hawaiian Islands?
-(iii) What environmental and anthropogenic factors drive crab community structure in shallow coral reefs across the Hawaiian Archipelago?
+They have been hypothesized as possible refugia from climate change impacts on shallow reefs [@glynnCoralReefBleaching1996; @bongaertsAssessingDeepReef2010; @smithDepthRefugiumCatastrophic2014; @holsteinFertileFathomsDeep2015; @muirSpeciesIdentityDepth2017], although a growing number of studies have found evidence that mesophotic reefs are unique and uniquely threatened, challenging this notion [@diazLightTemperatureDrive2023; @diazMesophoticCoralBleaching2023; @hobanPlumbingDepthsEnvironmental2023; @hurleyAssessmentShallowMesophotic2016; @kaneHighLevelsMesophotic2014; @pinheiroDeepReefFishes2019; @rochaMesophoticCoralEcosystems2018; @montgomeryCommunitySimilaritySpecies2021].
+Given the strong reported differences between deep and shallow reefs across taxa and the paucity of research on drivers of cryptic coral reef community variation, we sought to answer the following four research questions: (i) How do shallow reef brachyuran crab assemblages vary along the natural and anthropogenic gradients of the Hawaiian Archipelago?
+(ii) Do shallow reef brachyuran crab assemblages differ between the Main and Northwestern Hawaiian Islands?
+(iii) What environmental and anthropogenic factors drive crab assemblage structure in shallow coral reefs across the Hawaiian Archipelago?
 (iv) What are the relative contributions of depth (shallow vs.
@@
 Once returned to the surface, they were placed in containers of aerated, filtered seawater, crates were removed, and units were systematically disassembled plate-by-plate [see @lerayDNABarcodingMetabarcoding2015 for processing details].
-All adult brachyuran crabs ($\geq$ 5 mm) were photographed, identified (to species where possible, otherwise morphospecies), and preserved in 95% ethanol.<!-- Specimens were collected under permits #xxx (NWHI) and #xxx (MHI) and accessioned into the invertebrate zoology collections at the Florida Museum of Natural History (catalog numbers xxx--xxx).-->
+All adult brachyuran crabs ($\geq$ 5 mm) were photographed, identified, and preserved in 95% ethanol.
+Identifications were made to species level where possible, otherwise crabs were classified into morphospecies represented by suites of shared morphological characters.<!-- Specimens were collected under permits #xxx (NWHI) and #xxx (MHI) and accessioned into the invertebrate zoology collections at the Florida Museum of Natural History (catalog numbers xxx--xxx).-->
 

### 2024-06-05 manuscript.Rmd
a bunch of updates, now its ready to submit

@@
       equal_contributor: yes
-      correspondence: yes
-      email: "hurleyk@hawaii.edu"
   - Mykle L. Hoban:
@@
       equal_contributor: yes
+      correspondence: yes
+      email: "mhoban@hawaii.edu"
   - Kerry Reardon:
@@
 abstract:
-  Coral reef communities are influenced by various environmental and anthropogenic factors, yet those factors have not been consistently resolved.
-Furthermore, due to logistical difficulties, investigations of the role of depth (particularly mesophotic coral ecosystems) in influencing coral reef community structure have been taxonomically limited.
-Here, we examined the effects of a suite of environmental, ecological, and anthropogenic factors on the assemblage structure of shallow brachyuran crabs across the ~2,500 km span of the Hawaiian Archipelago.
-In addition, we compared shallow crab assemblages with those sampled on a depth transect from 12--90 m on O&#x02BB;ahu.
-We deployed Autonomous Reef Monitoring Structures (ARMS) on shallow reefs (12--16 m) across the latitudinal gradient of the Hawaiian Archipelago to compare with a depth gradient (12, 30, 60, and 90 m) on the south shore of O&#x02BB;ahu.
-We found a total of 2,284 individual crabs representing 103 (morpho)species in 60 genera and 22 families.
-Shallow crab assemblages were significantly different between the Main and Northwest Hawaiian Islands and variation in community structure was significantly associated with six biotic, environmental, and anthropogenic factors (cumulative human impacts, chlorophyll-A, sea-surface temperature, site depth, island slope, and potential larval connectivity).
-In depth comparisons, we found more variation among mesophotic samples from O&#x02BB;ahu (30--90 m) than among all shallow samples across the complete span of the archipelago.
-We show that 90 m of depth is a more potent determinant of brachyuran assemblage composition than the latitudinal, environmental, or anthropogenic gradients across the Hawaiian Islands.<!--Deeper reefs host significantly different brachyuran communities than shallow ones in Hawai&#x02BB;i, and only 3 of 90 morphospecies (~3%) occurring across the entire depth range (mesophotic to shallow) were sampled.
-Further research on the cryptic reef-dwelling fauna is needed for identifying areas of unique biodiversity, as well as boundaries between depth-stratified reef ecosystems.
-Shallow coral reefs are extensively studied, and although scleractinian corals have been recorded to ~170 m, little is known about other mesophotic coral reef ecosystem (MCE) inhabitants.
-Brachyuran crabs fill many ecological and trophic niches on reefs, making them ideal candidates for evaluating species composition among depths.
-Here we ask if MCEs host the same communities as the shallower reefs spread across the entire 2400 km Hawaiian Archipelago.-->
+  An inability to consistently resolve the environmental and anthropogenic factors that shape coral reefs impacts our understanding of how those factors interact and how coral reef communities may respond to change.
+Logistical challenges have additionally limited the taxonomic resolution and depth ranges investigated, with the focus largely on fishes and corals in the shallowest ~30 m of coral reefs.
+Here, we examine the effects of a suite of environmental, ecological, and anthropogenic factors on the assemblage structure of brachyuran crabs recovered from Autonomous Reef Monitoring Structures on shallow reefs (8--17 m) across the ~2,500 km span of the Hawaiian Archipelago.
+We compare these assemblages with those sampled along a depth gradient (12--90 m) on O&#x02BB;ahu.
+We found 2,284 individuals representing 103 (morpho)species from 22 families.
+Shallow assemblages were significantly different between the Main and Northwest Hawaiian Islands and significantly associated with human impacts, chlorophyll-A, sea-surface temperature, slope, potential larval connectivity, and depth.
+However, we found more variation among mesophotic samples from O&#x02BB;ahu (30--90 m) than among shallow samples across the entire archipelago, suggesting that 90 m of depth is a stronger determinant of brachyuran assemblage structure than the latitudinal, environmental, or anthropogenic gradients across the Hawaiian Islands.
 bibliography: references/citations.json
@@
       - "--reference-doc=resources/ref.docx"
-    keep_md: true
+    keep_md: false
 tblPrefix: table
@@
 source("crabs.R")
+
 if (!exists("crabs_done")) {
@@
 
-The structure and composition of coral reef communities is influenced by a suite of environmental and anthropogenic factors, yet the nature and makeup of those factors remain poorly resolved [@hughesCommunityStructureDiversity1989; @smithEffectsTopBottom2010; @brandlCoralReefEcosystem2019].
 Coral reefs are among the most diverse habitats in the world, and they face increasing threats of degradation, including climate-driven bleaching and phase shifts to alternate ecological regimes [@grahamManagingResilienceReverse2013; @norstromAlternativeStatesCoral2009; @pandolfiGlobalTrajectoriesLongterm2003].
-In the face of these threats, understanding drivers of metazoan community structure in coral reef ecosystems is critically important [@hughesRisingChallengeSustaining2010; @pageSeekingResistanceCoral2019], particularly as the search for criteria to prioritize climate refugia such as upwelling zone or mesophotic reefs intensifies.
-To date, the majority of such efforts have focused on conspicuous taxa such as corals and fishes [e.g., @benfieldComparisonCoralReef2008; @caballero-aragonWaveExposureTemperature2023; @kaneTrophicDesignationLive2017; @langeWaveExposureShapes2021; @vanlierHabitatConnectivityComplexity2018].
-However, these taxa make up a minority of coral reef biodiversity and thus may not be indicative of community patterns overall [@fautinOverviewMarineBiodiversity2010].
-Here, we investigate the factors influencing the community structure of brachyuran crabs across a suite of anthropogenic and environmental gradients using the Hawaiian Archipelago as a natural laboratory.
-
-The Hawaiian Archipelago comprises high volcanic islands and atolls stretching across ~2,500 km in the North Pacific Ocean and is the most isolated island chain on the planet.
-Typically regarded as two regions, the main Hawaiian Islands (MHI) include the eight inhabited high islands, and the Northwestern Hawaiian Islands (NWHI) comprise the uninhabited islands, atolls, and banks northwest of Kaua&#x02BB;i and Ni&#x02BB;ihau.
-The NWHI are wholly enclosed by the Papahānaumokuākea Marine National Monument, one of the largest marine protected areas in the world.
-The islands are subject to considerable anthropogenic and physical gradients across their span [@selkoeMapHumanImpacts2009], which are widely believed to influence coral reef communities.
-These gradients include differences in temperature, coral cover, human population, and anthropogenic impact, among others.
-Studies examining community structure across broad environmental gradients have been criticized for confounding multiple potentially covarying factors (such as habitat area and/or complexity) across the measured gradients [e.g., @bellwoodRegionalscaleAssemblyRules2001; @grahamEffectsMarineReserve2003; @connollyCommunityStructureCorals2005; @ewersConfoundingFactorsDetection2006; @fletcherRoleHabitatArea2007].
-Thus, we used autonomous reef monitoring structures (ARMS)---standardized collection devices designed to mimic reef structural complexity---to sample cryptic communities while controlling for habitat [@zimmermanArtificialReefMatrix2004; @brainardAutonomousReefMonitoring2009].
-We then sampled brachyuran crabs, a group comprising a notable proportion of the coral reef cryptofauna [284 species in Hawai&#x02BB;i, @castroCatalogAnomuranBrachyuran2011] from these structures to examine drivers of community composition across these gradients.
-We compared crab communities among islands and regions (MHI vs.
-NWHI) and examined the effects of a suite of possible environmental and anthropogenic drivers of community structure to test hypotheses about their relative importance.
-
-Much of our understanding of biotic zonation and community structure in the ocean originates with classic studies from the intertidal [e.g., @connellInfluenceInterspecificCompetition1961; @connellEffectsCompetitionPredation1961; @paineIntertidalCommunityStructure1974].
-While there are no shortages of exceptions, a long history of studies point towards a generalization that the lower bathymetric limits are generally regulated by ecological interactions whereas upper limits of intertidal communities are more often determined by abiotic factors [@stephenson1972life; @mengeRockyIntertidalCommunities2001; @bertnessTestingRelativeContribution1999; @dennyPhysicalProcessesThat2001].
-However, most of these models describe zonation at specific localities rather than community variation over larger spatial scales [but see @birdTideWaveMarks2013; @bishop-taylorTidesModellingElevation2019] and may not apply to coral reefs, which are subject to more complex community dynamics [@abramsRecruitmentLotteriesCoexistence1984; @bussCompetitiveNetworksNontransitive1979; @hixonPredationPreyRefuges1993; @hubbellUnifiedNeutralTheory2011; @moraEcologyFishesCoral2015; @saleStockrecruitRelationshipsRegional1982; @sheppardBiologyCoralReefs2017].
-While there is little doubt that ecological interactions are important on coral reefs, the suite of factors and which play the most important role in structuring coral reef communities remains a subject of considerable debate [@bellwoodRegionalscaleAssemblyRules2001; @connollyUnifiedModelExplains2017; @hubbellUnifiedNeutralTheory2011; @jacksonAdaptationDiversityReef1991; @mumbyCoralReefHabitats2008; @saleMaintenanceHighDiversity1977].
-
+In the face of these threats, understanding drivers of metazoan community structure in coral reef ecosystems is critically important [@hughesRisingChallengeSustaining2010; @pageSeekingResistanceCoral2019], particularly as the search for criteria to prioritize climate refugia such as upwelling zones or mesophotic reefs intensifies.
+While the structure and composition of coral reef communities is influenced by a suite of environmental and anthropogenic factors, the nature and makeup of those factors remain poorly resolved [@hughesCommunityStructureDiversity1989; @smithEffectsTopBottom2010; @brandlCoralReefEcosystem2019].
 One aspect of coral reef community structure that is thought to be important but remains under-investigated is depth.
@@
 They have been hypothesized as possible refugia from climate change impacts on shallow reefs [@glynnCoralReefBleaching1996; @bongaertsAssessingDeepReef2010; @smithDepthRefugiumCatastrophic2014; @holsteinFertileFathomsDeep2015; @muirSpeciesIdentityDepth2017], although a growing number of studies have found evidence that mesophotic reefs are unique and uniquely threatened, challenging this notion [@diazLightTemperatureDrive2023; @diazMesophoticCoralBleaching2023; @hobanPlumbingDepthsEnvironmental2023; @hurleyAssessmentShallowMesophotic2016; @kaneHighLevelsMesophotic2014; @pinheiroDeepReefFishes2019; @rochaMesophoticCoralEcosystems2018; @montgomeryCommunitySimilaritySpecies2021].
+
+The Hawaiian Archipelago, stretching across ~2,500 km in the North Pacific Ocean, provides a natural laboratory to examine drivers of metazoan community structure.
+This chain of islands comprises two regions, the inhabited high islands of the main Hawaiian Islands (MHI) and the uninhabited low islands, atolls, and banks of the Northwestern Hawaiian Islands (NWHI) which reside within one of the world's largest marine protected areas---the Papahānaumokuākea Marine National Monument.
+The island chain is subject to considerable anthropogenic and physical gradients driving differences in temperature, coral cover, human population, and anthropogenic impact, among others, all of which influence coral reef communities [@selkoeMapHumanImpacts2009].
+Nearshore drop-offs along the island chain also allow access to deeper reef habitat, enabling the examination of depth influences on community structure.
+
+To date, the majority of studies that have examined community structure across the Hawaiian archipelago and along a depth gradient have focused on conspicuous taxa such as corals and fishes [e.g., @benfieldComparisonCoralReef2008; @caballero-aragonWaveExposureTemperature2023; @kaneTrophicDesignationLive2017; @langeWaveExposureShapes2021; @vanlierHabitatConnectivityComplexity2018].
+However, these taxa make up a minority of coral reef biodiversity and thus may not be indicative of community patterns overall [@fautinOverviewMarineBiodiversity2010].
+Furthermore, studies on these taxa have also been criticized for confounding multiple potentially covarying factors (such as habitat area and/or complexity) across the measured gradients [e.g., @bellwoodRegionalscaleAssemblyRules2001; @grahamEffectsMarineReserve2003; @connollyCommunityStructureCorals2005; @ewersConfoundingFactorsDetection2006; @fletcherRoleHabitatArea2007].
+
+Here, we used autonomous reef monitoring structures (ARMS)---standardized collection devices designed to mimic reef structural complexity---to sample cryptic communities while controlling for habitat [@zimmermanArtificialReefMatrix2004; @brainardAutonomousReefMonitoring2009].
+We targeted brachyuran crabs, a group comprising a notable proportion of the coral reef cryptofauna [284 species in Hawai&#x02BB;i, @castroCatalogAnomuranBrachyuran2011], recovered from these structures to examine drivers of community composition across the Hawaiian Archipelago.
+We compared crab communities among islands and regions (MHI vs.
+NWHI) and examined the effects of a suite of possible environmental and anthropogenic drivers of community structure to test hypotheses about their relative importance.
+
+Much of our understanding of biotic zonation and community structure in the ocean originates with classic studies from the intertidal [e.g., @connellInfluenceInterspecificCompetition1961; @connellEffectsCompetitionPredation1961; @paineIntertidalCommunityStructure1974].
+While there are no shortages of exceptions, a long history of studies point towards a generalization that the lower bathymetric limits are generally regulated by ecological interactions whereas upper limits of intertidal communities are more often determined by abiotic factors [@stephenson1972life; @mengeRockyIntertidalCommunities2001; @bertnessTestingRelativeContribution1999; @dennyPhysicalProcessesThat2001].
+However, most of these models describe zonation at specific localities rather than community variation over larger spatial scales [but see @birdTideWaveMarks2013; @bishop-taylorTidesModellingElevation2019] and may not apply to coral reefs, which are subject to more complex community dynamics [@abramsRecruitmentLotteriesCoexistence1984; @bussCompetitiveNetworksNontransitive1979; @hixonPredationPreyRefuges1993; @hubbellUnifiedNeutralTheory2011; @moraEcologyFishesCoral2015; @saleStockrecruitRelationshipsRegional1982; @sheppardBiologyCoralReefs2017].
+While there is little doubt that ecological interactions are important on coral reefs, the suite of factors and which play the most important role in structuring coral reef communities remains a subject of considerable debate [@bellwoodRegionalscaleAssemblyRules2001; @connollyUnifiedModelExplains2017; @hubbellUnifiedNeutralTheory2011; @jacksonAdaptationDiversityReef1991; @mumbyCoralReefHabitats2008; @saleMaintenanceHighDiversity1977].
+
 Given the strong reported differences between deep and shallow reefs across taxa and the paucity of research on drivers of cryptic coral reef community variation, we sought to answer the following four research questions: (i) How do shallow reef brachyuran crab assemblages vary along the natural and anthropogenic gradients of the Hawaiian Archipelago?
@@
 
-Brachyuran crab assemblages were sampled from benthic coral reef habitat in the Hawaiian Archipelago using Autonomous Reef Monitoring Structures (ARMS).
+Brachyuran crab assemblages were sampled from benthic coral reef habitat in the Hawaiian Archipelago using Autonomous Reef Monitoring Structures (ARMS) collected by NOAA's Ecosystem Science Division as part of its Pacific Reef Assessment and Monitoring Program.
 ARMS are standardized collection devices designed to mimic coral reef structural complexity to attract settlement of reef cryptofauna [@zimmermanArtificialReefMatrix2004; @brainardAutonomousReefMonitoring2009].
@@
 
-![An autonomous reef monitoring structure (ARMS) unit deployed on a reef in the Red Sea (photo credit: Michael Berumen, used with permission)](data/images/arms_example.png){#fig:fig_arms}
+![Autonomous reef monitoring structures (ARMS) deployed in the Hawaiian Archipelago.](data/images/arms_example.png){#fig:fig_arms}
 
@@
 
-A total of 89 ARMS units were deployed over the course of six years (2010--2016).
-Units were deployed by NOAA divers during Pacific Reef Assessment and Monitoring Program (RAMP) cruises in shallow coral reef habitat (12--16 m depth) in both the Main (MHI) and Northwest Hawaiian Islands (NWHI) ([@fig:fig_map; @tbl:arms_table]).
+A total of 89 ARMS units were deployed over the course of six years (2009--2016) in shallow coral reef habitat (8--17 m depth) in both the Main (MHI) and Northwest Hawaiian Islands (NWHI) ([@fig:fig_map; @tbl:arms_table]).
 To explore the effects of depth, we also present data from an earlier study by Hurley et al.
@@
   sample_tibble() %>%
-  separate_wider_regex(sample,patterns=c(recovery_year="^[0-9]+","_",sample=".*"))  %>%
-  select(sample,everything()) %>%
+  # separate_wider_regex(sample,patterns=c(recovery_year="^[0-9]+","_",sample=".*"))  %>%
+  # select(sample,everything()) %>%
   unite("range",deployment_year,recovery_year,sep="--") %>%
@@
   mutate(across(-island,~replace_na(.x,0)))
-print_table(arms_schedule,"arms_table","ARMS unit deployment schedule (number of units)",accuracy=1) 
+print_table(arms_schedule,"arms_table","Recovered ARMS unit soak times across sites (number of units)",accuracy=1) 
 ```
@@
 Once returned to the surface, they were placed in containers of aerated, filtered seawater, crates were removed, and units were systematically disassembled plate-by-plate [see @lerayDNABarcodingMetabarcoding2015 for processing details].
-All adult brachyuran crabs ($\geq$ 5 mm) were photographed, identified, and preserved in 95% ethanol.
-Identifications were made to species level where possible, otherwise crabs were classified into morphospecies represented by suites of shared morphological characters.<!-- Specimens were collected under permits #xxx (NWHI) and #xxx (MHI) and accessioned into the invertebrate zoology collections at the Florida Museum of Natural History (catalog numbers xxx--xxx).-->
+All adult brachyuran crabs ($\geq$ 2 mm) were photographed, identified, and preserved in 95% ethanol.
+Identifications were made to species level where possible, otherwise crabs were classified into morphospecies represented by suites of shared morphological characters according to expert advice and Poore & Ahyong [-@pooreMarineDecapodCrustacea2023].<!-- Specimens were collected under permits #xxx (NWHI) and #xxx (MHI) and accessioned into the invertebrate zoology collections at the Florida Museum of Natural History (catalog numbers xxx--xxx).-->
 
@@
 
-To examine patterns of crab diversity across the islands, we used `estimate_richness` to calculate species richness and Simpson diversity (Simpson = $1-\sum_{i=1}^{R}p_i^2$, where R is the total number of species, and $\mathrm{p}_i$ is the proportional abundance of each species).
+To examine patterns of crab diversity across the islands, we used `estimate_richness` to calculate species richness and effective Simpson diversity (effective Simpson = $\frac{1}{\sum_{i=1}^{R}p_i^2}$, where R is the total number of species, and $\mathrm{p}_i$ is the proportional abundance of each species).
+The effective Simpson index (the inverse of the "classic" Simpson entropy), which has units of species, is easier to interpret in an ecological context than an entropic non-linear index [@jostEntropyDiversity2006; @tuomistoUpdatedConsumerGuide2012].
 We used nonparametric Kruskal-Wallis ranked sum tests to compare diversity between shallow and mesophotic samples and between island groups (MHI vs.
@@
 Island group comparisons were performed both including and excluding mesophotic samples.
-<!--To explore environmental influences on alpha diversity, we used a linear mixed-effect model with Simpson or richness as response variables, the environmental variables listed below as fixed effects, and island nested within island group (MHI vs NWHI) as random effects.
-Models were generated using the `lmer` function from the `lme4` package [@batesFittingLinearMixedEffects2015].
--->
-
-We used distance-based redundancy analysis (dbRDA; a form of multivariate multiple regression) with `dbrda` to examine the relative contribution of environmental variables to crab community variation across the islands.
-We first used a stepwise forward selection procedure (`ordiR2step`) to find the variables that best explained community dissimilarity.
-Then, we sequentially removed variables with the largest variance inflation factors (VIF; calculated using `vif.cca`) until all values were <3, indicating variables were sufficiently uncorrelated.
+
+We used distance-based redundancy analysis (dbRDA; a form of multivariate multiple regression) with `dbrda` to examine the relative contribution of environmental variables ([@tbl:env_table]) to crab community variation across the islands, conditioning the recovery year as a random effect.
+From a model incorporating all potential drivers, we sequentially removed variables with the largest variance inflation factors (VIF; calculated using `vif.cca`) until all values were <3, indicating variables were sufficiently uncorrelated.
 Model significance was assessed using permutation tests with `r num(PERM,1)` permutations with the function `anova`.
@@
 
-Potential drivers of shallow crab community structure were assembled from a previous study examining environmental effects on genetic diversity [@selkoeDNACoralReef2016], NOAA's National Coral Reef Monitoring Program (NCRMP), the Pacific Islands Ocean Observing System (PacIOOS), and NASA's Physical Oceanography/Ocean Biology Distributed Active Archive Centers (PO.DAAC, OB.DAAC) ([@tbl:env_table]).
-For details on how sea-surface temperature and chlorophyll data were compiled and aggregated, see Supplemental material.
+Potential drivers of shallow crab community structure were assembled from a previous study examining environmental effects on genetic diversity [@selkoeDNACoralReef2016], NOAA's National Coral Reef Monitoring Program (NCRMP), the Pacific Islands Ocean Observing System (PacIOOS), and MODIS satellite data products compiled by NASA's Physical Oceanography/Ocean Biology Distributed Active Archive Centers (PO.DAAC, OB.DAAC) ([@tbl:env_table; @tbl:arms_metadata_table]).
+Sea-surface temparature (SST) and chlorophyll (Chl-A) values were calculated by averaging monthly means for the duration of the soaking period at each sampling site.
+For details on satellite data preparation and data sources, see Supplemental material.
 Due to logistical limitations, environmental data were collected only for shallow samples (\< 30 m).
@@
 
-Overall mean species richness for the complete dataset was `r num(all_alpha$richness,0.1)` $\pm$ `r num(all_alpha$richness_sd,0.1)` species per unit (Simpson diversity: `r num(all_alpha$simpson,0.01)` $\pm$ `r num(all_alpha$simpson_sd,0.01)`).
-Overall mean species richness for shallow ARMS \< 30 m was `r num(shallow_alpha$richness,0.1)` $\pm$ `r num(shallow_alpha$richness_sd,0.1)` species per unit (Simpson diversity: `r num(shallow_alpha$simpson,0.01)` $\pm$ `r num(shallow_alpha$simpson_sd,0.01)`).
-Overall mean species richness for mesophotic ARMS $\geq$ 30 m was `r num(deep_alpha$richness,0.1)` $\pm$ `r num(deep_alpha$richness_sd,0.1)` species per unit (Simpson diversity: `r num(deep_alpha$simpson,0.01)` $\pm$ `r num(deep_alpha$simpson_sd,0.01)`).
+Overall mean species richness for the complete dataset was `r num(all_alpha$richness,0.1)` $\pm$ `r num(all_alpha$richness_sd,0.1)` species per unit (effective Simpson diversity: `r num(all_alpha$simpson,0.01)` $\pm$ `r num(all_alpha$simpson_sd,0.01)`).
+Overall mean species richness for shallow ARMS \< 30 m was `r num(shallow_alpha$richness,0.1)` $\pm$ `r num(shallow_alpha$richness_sd,0.1)` species per unit (effective Simpson diversity: `r num(shallow_alpha$simpson,0.01)` $\pm$ `r num(shallow_alpha$simpson_sd,0.01)`).
+Overall mean species richness for mesophotic ARMS $\geq$ 30 m was `r num(deep_alpha$richness,0.1)` $\pm$ `r num(deep_alpha$richness_sd,0.1)` species per unit (effective Simpson diversity: `r num(deep_alpha$simpson,0.01)` $\pm$ `r num(deep_alpha$simpson_sd,0.01)`).
 
@@
 
-For the complete dataset, there was no significant difference in observed species richness between the MHI and NWHI ([@fig:fig_alpha[]{}a]; `r k_test(all_alpha$k_r_island)`), but Simpson diversity was significantly different ([@fig:fig_alpha[]{}c]; `r k_test(all_alpha$k_s_island,plain=TRUE)`).
-Richness was not significantly different between shallow and deep samples ([@fig:fig_alpha[]{}b]; `r k_test(all_alpha$k_r_depth,plain=TRUE)`) but Simpson diversity was ([@fig:fig_alpha[]{}d]; `r k_test(all_alpha$k_s_depth,plain=TRUE)`).
-When only shallow arms \< 30 m were considered, there were no significant differences in alpha diversity between the MHI and NWHI for either index (richness: `r k_test(shallow_alpha$k_r_island,plain=TRUE)`, [@fig:fig_alpha2[]{}a]; Simpson: `r k_test(shallow_alpha$k_s_island,plain=TRUE)`, [@fig:fig_alpha2[]{}b]).
-
-<!--For the complete dataset, there was no significant difference in observed species richness between the MHI and NWHI `r w_test(all_alpha$w_r_island)`, but Simpson diversity was significantly different ([@fig:fig_alpha[]{}a]; `r w_test(all_alpha$w_s_island,plain=TRUE)`).
-When only shallow arms \< 30 m were considered, neither comparison was significant (richness: `r w_test(shallow_alpha$w_r_island,plain=TRUE)`; Simpson: `r w_test(shallow_alpha$w_s_island,plain=TRUE)`).
-Simpson diversity was significantly different between shallow and deep samples ([@fig:fig_alpha[]{}b]; `r w_test(all_alpha$w_s_depth,plain=TRUE)`) but richness was not (`r w_test(all_alpha$w_r_depth,plain=TRUE)`).
+For the complete dataset, there was no significant difference in observed species richness between the MHI and NWHI ([@fig:fig_alpha[]{}a]; `r k_test(all_alpha$k_r_island)`), but effective Simpson diversity was significantly different ([@fig:fig_alpha[]{}c]; `r k_test(all_alpha$k_s_island,plain=TRUE)`).
+Richness was not significantly different between shallow and deep samples ([@fig:fig_alpha[]{}b]; `r k_test(all_alpha$k_r_depth,plain=TRUE)`) but effective Simpson diversity was ([@fig:fig_alpha[]{}d]; `r k_test(all_alpha$k_s_depth,plain=TRUE)`).
+When only shallow arms \< 30 m were considered, there were no significant differences in alpha diversity between the MHI and NWHI for either index (richness: `r k_test(shallow_alpha$k_r_island,plain=TRUE)`, [@fig:fig_alpha2[]{}a]; effective Simpson: `r k_test(shallow_alpha$k_s_island,plain=TRUE)`, [@fig:fig_alpha2[]{}b]).
+
+<!--For the complete dataset, there was no significant difference in observed species richness between the MHI and NWHI `r w_test(all_alpha$w_r_island)`, but effective Simpson diversity was significantly different ([@fig:fig_alpha[]{}a]; `r w_test(all_alpha$w_s_island,plain=TRUE)`).
+When only shallow arms \< 30 m were considered, neither comparison was significant (richness: `r w_test(shallow_alpha$w_r_island,plain=TRUE)`; effective Simpson: `r w_test(shallow_alpha$w_s_island,plain=TRUE)`).
+Effective Simpson diversity was significantly different between shallow and deep samples ([@fig:fig_alpha[]{}b]; `r w_test(all_alpha$w_s_depth,plain=TRUE)`) but richness was not (`r w_test(all_alpha$w_r_depth,plain=TRUE)`).
 -->
 
-<!--For the complete dataset, there was no significant difference in observed species richness between the MHI and NWHI `r t_test(all_alpha$t_r_island)`, but Simpson diversity was significantly different ([@fig:fig_alpha[]{}a]; `r t_test(all_alpha$t_s_island,plain=TRUE)`).
-When only shallow arms \< 30 m were considered, neither comparison was significant (richness: `r t_test(shallow_alpha$t_r_island,plain=TRUE)`; Simpson: `r t_test(shallow_alpha$t_s_island,plain=TRUE)`).
-Simpson diversity was significantly different between shallow and deep samples ([@fig:fig_alpha[]{}b]; `r t_test(all_alpha$t_s_depth,plain=TRUE)`) but richness was not (`r t_test(all_alpha$t_r_depth,plain=TRUE)`).
+<!--For the complete dataset, there was no significant difference in observed species richness between the MHI and NWHI `r t_test(all_alpha$t_r_island)`, but effective Simpson diversity was significantly different ([@fig:fig_alpha[]{}a]; `r t_test(all_alpha$t_s_island,plain=TRUE)`).
+When only shallow arms \< 30 m were considered, neither comparison was significant (richness: `r t_test(shallow_alpha$t_r_island,plain=TRUE)`; effective Simpson: `r t_test(shallow_alpha$t_s_island,plain=TRUE)`).
+Effective Simpson diversity was significantly different between shallow and deep samples ([@fig:fig_alpha[]{}b]; `r t_test(all_alpha$t_s_depth,plain=TRUE)`) but richness was not (`r t_test(all_alpha$t_r_depth,plain=TRUE)`).
 -->
@@
 
-reduce_model <- function(m,vif_limit=3) {
+reduce_model <- function(m,o,d,vif_limit=3,condition=character()) {
   vifs <- vif.cca(m)
   if (any(vifs > vif_limit)) {
-    v <- vifs[-which(vifs == max(vifs))[1]]
-    m_lhs <- capture.output(print(lhs(m$call$formula)))
-    f <- str_glue("{m_lhs} ~ {str_c(names(v),collapse=' + ')}") %>%
-      as.formula()
-    return(reduce_model(dbrda(f,data=cc$crab_data_shallow,distance = "bray")))
+    oldterms <- rhs.vars(m$call$formula)
+    v <- names(vifs)[which.max(vifs)]
+    newterms <- grep(v,oldterms,value=TRUE,invert=TRUE)
+    f <- reformulate(newterms,response=quote(o))
+    newmod <- dbrda(f,data=d,distance="bray")
+    reduce_model(newmod,o,d,vif_limit,condition)
   } else {
@@
 
-# db_upr <- dbrda(cc$crab_otus_shallow ~ depth + chl + sst + slope + coral_cover + closest_island + larval_connectivity + human_impact,data=cc$crab_data_shallow,distance="bray")
-db_upr <- dbrda(cc$crab_otus_shallow ~ depth + chl + sst + slope + coral_cover + larval_connectivity + human_impact,data=cc$crab_data_shallow,distance="bray")
-db_lwr <- dbrda(cc$crab_otus_shallow  ~ 1,data=cc$crab_data_shallow,distance="bray")
-
-# do our forward selection to find the best model
-# set trace to TRUE if you want to see a bunch of output
-db_modl <- ordiR2step(db_lwr,db_upr,trace=FALSE)
-
-# see marginal effects of model terms
-# anova(db_modl,by="margin")
+# calculate full dbRDA model
+db_modl <- dbrda(
+  cc$crab_otus_shallow ~ depth +
+    chl + sst + slope + coral_cover +
+    larval_connectivity + human_impact +
+    Condition(recovery_year),
+  data=cc$crab_data_shallow,
+  distance="bray"
+)
 
 # get rid of variables with VIF > 3
-# db_modl <- reduce_model(db_modl,vif_limit=8)
-db_modl <- reduce_model(db_modl,vif_limit=3)
+db_modl <- reduce_model(db_modl,cc$crab_otus_shallow,cc$crab_data_shallow)
 
@@
 
-# warning: the following *may* be sketchy:
-# dbRDA doesn't automatically give you species scores, so let's assign them
-# but the community data needs to be transformed to make sense
-# see https://github.com/vegandevs/vegan/issues/254 for information on the transformation below
-# sppscores(db_modl) <- sqrt(decostand(cc$crab_otus_shallow, "total")/2) 
-# 
-# # this was to grab the 4 most abundant species overall
-# top_species <- names(head(rev(sort(taxa_sums(cc$crabs_shallow))),n=4))
-# 
-# # extract the species scores for the species we want
-# species_scores <- scores(db_modl,display="species") %>%
-#   as_tibble(rownames = "species") %>%
-#   filter(species %in% top_species)
-
+# get marginal term information and sort by descending f-statistic
 modl_perm <- anova(db_modl,by="margin") %>%
@@
 
+# make a nicer display-worthy table for significant terms
 rda_table <- modl_perm %>%
@@
 
-# TODO: check this, this might be wrong!
-nterm <- length(attr(db_modl$terms,"term.label"))
-
+# get number of model terms
+tt <- attr(db_modl$terms,"term.label")
+nterm <- length(grep("Condition",tt,ignore.case = FALSE,value=TRUE,invert = TRUE))
+
+# get top drivers for the text
 drivers <- rda_table %>%
@@
 ### Environmental drivers {-}
-
 Environmental variables were strongly clustered according to island group (MHI vs NWHI) ([@fig:fig_pca]).
@@
 Additionally, variability among 30--90 m samples was greater than among all shallow samples, (max.
-Euclidean distance between shallow samples in the CAP ordination: `r num(max(shallow_dist),0.01)`, max.
-Euclidean distance between deep samples: `r num(max(deep_dist),0.01)`) ([@fig:fig_cap]).
+Euclidean distance between shallow samples in the CAP ordination: `r num(max(shallow_dist),0.01)`, avg.
+`r num(mean(shallow_dist),0.01)`; max.
+Euclidean distance between deep samples: `r num(max(deep_dist),0.01)`, avg.
+`r num(mean(deep_dist),0.01)`) ([@fig:fig_cap]).
 
@@
 The next-most significant driver, chlorophyll A, can be used as a proxy for primary productivity.
-Productivity gradients across the Hawaiian Archipelago coincide with differences in coral reef community composition [@selkoeDNACoralReef2016], endemic reef fish dominance [@friedlanderDominanceEndemicsReef2020], and survival rates of top predators [@bakerEffectVariableOceanic2007].
+Productivity gradients across the Hawaiian Archipelago coincide with differences in coral reef community composition [@selkoeDNACoralReef2016], endemic reef fish dominance [@friedlanderDominanceEndemicsReef2020], survival rates of top predators [@bakerEffectVariableOceanic2007], and island geomorphology [@goveNearislandBiologicalHotspots2016].
 Thus, it is not surprising that we find the same top drivers of community structure for one of the primary taxonomic groups inhabiting coral reefs, the brachyuran crabs, across the Hawaiian archipelago.
@@
 Depth in and of itself is unlikely to affect marine community composition.
-At the same time, oceanographic factors with well-known depth profiles (e.g., temperature, light, oxygen, nutrients, etc.) are not expected to vary dramatically across such a narrow depth range (\< 10 m).
-Unfortunately, we lack environmental data from the specific deployment depths of our shallow-water ARMS sites, and cannot unravel the co-varying factor(s) associated with the depth effects we observed.
-Regardless, given its significance across both narrow (~10 m) and wider (~100 m) depth ranges sampled here, depth (or an associated factor) is clearly an important driver for reef crab communities in across the Hawaiian Archipelago.
+At the same time, oceanographic factors with well-known depth profiles (e.g., temperature, light, oxygen, nutrients, etc.) are not expected to vary dramatically across such a narrow depth range (\< 10 m), although fine-scale habitat variability among sites may also be driving the community differences we found [@steyaertRemoteReefCryptobenthic2022].
+Unfortunately, we lack environmental data at depth for each ARMS site, and thus cannot unravel the co-varying factor(s) associated with the depth effects we observed.
+Regardless, given its significance across both narrow (\< 10 m) and wider (90 m) depth ranges sampled here, depth (or an associated factor) is clearly an important driver for cryptobenthic crab communities across the Hawaiian Archipelago.
 Given brachyurans are a ubiquitous and important component of coral reef communities [@reynsPatternsProcessesBrachyuran1999; @lasleyjrPhylogeneticRelationshipsUbiquitous2015; @hazeriLatitudinalSpeciesDiversity2019], these findings are likely to apply more broadly as well.
@@
 
-In addition to community composition, mesophotic samples also contributed significantly to alpha diversity differences across the archipelago.
-We found significant differences in Simpson diversity between the MHI and NWHI, but only when mesophotic samples were included in the comparison; when deep ARMS units were left out, these same alpha diversity comparisons were non-significant.
-When comparing shallow to mesophotic, Simpson diversity was significantly (if slightly) higher in deep communities.
-Thus, the apparent differences in alpha diversity between the MHI and NWHI can be attributed entirely to differences between shallow and deep samples on the island of O&#x02BB;ahu.
+In addition to community composition, mesophotic samples also contributed to alpha diversity differences across the archipelago.
+We found significant differences in effective Simpson diversity between the MHI and NWHI, but only when mesophotic samples were included in the comparison; when deep ARMS units were left out, these comparisons were non-significant.
+When comparing shallow to mesophotic, effective Simpson diversity was significantly (if slightly) higher in deep communities.
+Thus, the apparent overall differences in alpha diversity between the MHI and NWHI can be attributed entirely to the deep samples from the island of O&#x02BB;ahu.
 This pattern is likely driven by species unique to mesophotic habitat [@hurleyAssessmentShallowMesophotic2016].
@@
 *Lower* limits of mesophotic coral reef ecosystems are likely determined by abiotic factors that correlate with depth, whereas *upper* limits may be driven by ecological interactions.
-By comparing shallow-water ARMS across the latitudinal, physical and anthropogenic gradients of the Hawaiian Archipelago, we can test the relative importance of the significant drivers of community structure.
+By comparing shallow-water ARMS across the latitudinal, physical and anthropogenic gradients of the Hawaiian Archipelago, we were able to test the relative importance of the significant drivers of community structure.
 Our analysis corroborates previous work and shows depth is the strongest driver of community structure.
@@
 We conclude that some factor(s) that covary with depth exert more influence on the cryptobenthic coral reef community structure than those biotic and abiotic gradients that span the entire Hawaiian Archipelago in shallow-water environments.
+
+# Acknowledgements
+We thank the Coral Reef Conservation Program for supporting the deployment and recovery of the shallow ARMS.
+We thank the UH Mānoa Dive Safety Office (D.
+Pence, K.
+Stender, J.
+Jones, and C.J.
+Bradley) for installation and retrieval of the mesophotic ARMS.
+We thank Joshua Copus and Kevin Flanagan for additional assistance with retrieval and processing of mesophotic ARMS units.
+We thank the ToBo laboratory members, particularly M.
+Iacchei, C.
+Ka&#x02BB;apu-Lyons, and Z.
+Hee, for discussion, support, and their efforts in the disassembly and sorting of ARMS samples on retrieval.
+This is contribution #xxx from the Hawai&#x02BB;i Institute of Marine Biology and #xxx from the School of Earth Science and Technology at the University of Hawai&#x02BB;i.
+
+# Funding
+This work was funded through a combination of grants including NSF OCE 12-60169, NSF GRFP DGE-1329626, and NSF-2048457.
+In addition, the NOAA Coral Reef Conservation Program funded the Pacific Reef Assessment and Monitoring Program, which included the shallow ARMS deployment in the Hawaiian Archipelago.
 
@@
 
-```{r fig-alpha, fig.cap='Box plots of (a,b) species richness and (c,d) Simpson diversity for complete dataset (all sampled depths).
-Asterisk denotes significant difference (K-W test).
+```{r fig-alpha, fig.cap='Box plots of (a,b) species richness and (c,d) effective Simpson diversity for complete dataset (all sampled depths).
+Asterisks indicate significant differences (K-W test).
+Notches represent nonparametric confidence intervals about the median.
 (a) Species richness by island group.
 (b) Species richness by sampling depth.
-(c) Simpson diversity by island group.
-(d) Simpson diversity by sampling depth.
-Notches represent nonparametric confidence intervals about the median.', results="asis", dev="svg", fig.width=10, fig.height=10, dependson="init"}
+(c) Effective Simpson diversity by island group.
+(d) Effective Simpson diversity by sampling depth.', results="asis", dev="svg", fig.width=10, fig.height=10, dependson="init"}
 library(patchwork)
@@
 
-pointsize <- diff(range(art$Observed))
+pointsize <- 5
 
@@
   geom_boxplot(aes(x=island_group,y=Observed,fill=island_group),notch=TRUE) +
-  geom_point(aes(x,y),data=sig,shape=pointsize,size=1,stroke=1.5) +
+  geom_point(aes(x,y),data=sig,shape=8,size=pointsize,stroke=1.5) + 
   scale_fill_manual(values=island_group_pal) +
@@
 
-pointsize <- diff(range(art$Simpson))*5
+pointsize <- 5
 
@@
   xlab("Island group") + 
-  ylab("Simpson's diversity index")
+  ylab("Effective Simpson diversity")
 
@@
   xlab("Sampling depth") + 
-  ylab("Simpson's diversity index")
+  ylab("Effective Simpson diversity")
 
@@
 
-```{r fig-alpha2, fig.cap='Box plots of (a) species richness and (b) Simpson diversity by island group for shallow samples only.
+```{r fig-alpha2, fig.cap='Box plots of (a) species richness and (b) Effective Simpson diversity by island group for shallow samples only.
 Notches as in [@fig:fig_alpha].
@@
 
+```{r arms-metadata-table, results="asis", dependson="init"}
+region_map <- c(
+  'main' = 'MHI',
+  'northwest' = 'NWHI',
+  'mce' = 'MHI (mesophotic)'
+)
+
+arms_metadata <- cc$crabs_untransformed %>%
+  sample_tibble() %>%
+  # separate_wider_regex(sample,patterns=c(recovery_year="^[0-9]+","_",sample=".*"))  %>%
+  # select(sample,everything()) %>%
+  mutate(
+    island = case_when(
+      str_remove_all(hawaiian_name,'[^a-zA-Z ]') != island ~ as.character(str_glue("{hawaiian_name} ({island})")),
+      TRUE ~ as.character(hawaiian_name)
+    ),
+    island = case_when(
+      island == "O‘ahu" ~ str_glue("{hawaiian_name} ({shallow_deep})"),
+      TRUE ~ island
+    ),
+    island = fct_reorder(island,-lat),
+    region = region_map[region],
+    lat = num(lat,0.0001),
+    lon = num(lon,0.0001)
+  ) %>%
+  select(sample,region,island,depth,deployment_year,recovery_year,lat,lon) %>%
+  janitor::clean_names(case="sentence")
+print_table(arms_metadata,"arms_metadata_table","ARMS unit metadata",accuracy=1,hdr_case=NA,format_numbers=FALSE) 
+```
+
+\newpage
+
 ```{r rda-table,results="asis",dependson="rda-analysis"}

### 2024-06-05 crabs.R

@@ -300,7 +300,7 @@ setup_crabs <- function() {
     sample_tibble(sample_col = "unit") %>%
     select(-sst,-chl) %>%
     # select(unit,region,island_group,island,lat,lon,depth,chl=chl_sat,sst=sst_sat,slope,coral_cover,closest_island,larval_connectivity,human_impact) %>%
-    select(unit,region,island_group,island,hawaiian_name,lat,lon,depth,chl=chlorophyll_oracle,sst=sst_sat,slope,coral_cover,closest_island,larval_connectivity,human_impact) %>%
+    select(unit,region,island_group,recovery_year,island,hawaiian_name,lat,lon,depth,chl=chlorophyll_oracle,sst=sst_sat,slope,coral_cover,closest_island,larval_connectivity,human_impact) %>%
     # select(unit,region,island_group,island,lat,lon,depth,chl=chl_new,sst=sst_new,slope,coral_cover,closest_island,larval_connectivity,human_impact) %>%
     mutate(across(where(is.numeric),~as.numeric(scale(.x)))) %>%
     column_to_rownames("unit") 
@@ -412,7 +412,7 @@ sample_summary <- function(ps,top_n=5) {
   return(ss)
 }
 
-alpha_diversity <- function(ps,measures=c("Observed","Simpson")) {
+alpha_diversity <- function(ps,measures=c("Observed","InvSimpson")) {
   
   ad <- list()
   
@@ -420,6 +420,7 @@ alpha_diversity <- function(ps,measures=c("Observed","Simpson")) {
   richness <- ps %>%
     estimate_richness(measures = measures) %>%
     as_tibble(rownames="sample") %>%
+    rename(Simpson = InvSimpson) %>%
     mutate(sample=str_replace(sample,"^X",""))
   
   cd <- ps %>%

### 2024-07-10 manuscript.Rmd
acknowledgments/keywords and updating renv

@@
 ### Keywords {-}
-<div custom-style="mono">ARMs, mesophotic coral ecosystems, community structure, dispersal, Hawai&#x02BB;i, brachyuran crabs</div>
+<div custom-style="mono">Mesophotic coral ecosystems, community structure, depth gradient, environmental drivers, Hawaii, Autonomous Reef Monitoring Structures</div>
 
@@
 # Acknowledgements
-We thank the Coral Reef Conservation Program for supporting the deployment and recovery of the shallow ARMS.
+We thank the NOAA Coral Reef Conservation Program for supporting the deployment and recovery of the shallow ARMS.
 We thank the UH Mānoa Dive Safety Office (D.
@@
 Bradley) for installation and retrieval of the mesophotic ARMS.
-We thank Joshua Copus and Kevin Flanagan for additional assistance with retrieval and processing of mesophotic ARMS units.
-We thank the ToBo laboratory members, particularly M.
+We specially thank K.
+Flanagan from the UH dive program, whose support was instrumental in the development of the mesophotic ARMS project, and J.
+Copus, who provided expert mesophotic field support; both were taken too soon.
+May their memories be a blessing.
+We thank the members of the ToBo lab, particularly M.
 Iacchei, C.
@@
 Hee, for discussion, support, and their efforts in the disassembly and sorting of ARMS samples on retrieval.
-This is contribution #xxx from the Hawai&#x02BB;i Institute of Marine Biology and #xxx from the School of Earth Science and Technology at the University of Hawai&#x02BB;i.
+This is contribution #xxx from the Hawai&#x02BB;i Institute of Marine Biology and #xxx from the School of Ocean and Earth Science and Technology at the University of Hawai&#x02BB;i.
 
 # Funding
-This work was funded through a combination of grants including NSF OCE 12-60169, NSF GRFP DGE-1329626, and NSF-2048457.
+This work was funded through a combination of grants including NSF OCE-1260169, NSF GRFP DGE-1329626, and NSF OCE-2048457.
 In addition, the NOAA Coral Reef Conservation Program funded the Pacific Reef Assessment and Monitoring Program, which included the shallow ARMS deployment in the Hawaiian Archipelago.
+<!--
+Specimens used in this study were collected under Hawai&#x02BB;i Division of Aquatic Resources Special Activity Permit numbers SAP 2013-47, SAP 2012-63, SAP 2011-01, and SAP 2008-99.
+-->
 

### 2024-08-27 manuscript.Rmd
beginning proc b revisions

@@
 library(worrms)
-library(BiodiversityR)
 library(ggtext)
+
+# can't get BiodiversityR to work on new mac, but that's ok because
+# all we need is this one function, which miraculously works on its own!
+source(here("lib","CAPdiscrim.R"))
 
@@
 
+<!-- TODO: update number of different types to reflect family/genus/species level rather than species/morphospecies level -->
 The `r nsamples(cc$crabs_untransformed)` ARMS units deployed across the Hawaiian Islands (in `r format_range(cc$crab_data$depth)` m depth) yielded a total of `r num(sum(sample_sums(cc$crabs_untransformed)),1)` adult crabs.
@@
 There were `r all_summary$singleton_count` singleton taxa (morpho/species occurring in a single unit across all sites), accounting for ~`r num(all_summary$singleton_count/ntaxa(cc$crabs_untransformed)*100,1)`% of total recovered crab species.
-<!--`r str_to_sentence(english(all_summary$singleton_count))` species occurred in a single unit across all sites (singletons), accounting for ~`r num(all_summary$singleton_count/ntaxa(cc$crabs_untransformed)*100,1)`% of total recovered crab species.-->
 

### 2024-08-27 collapse_taxonomy.R

@@ -1,0 +1,48 @@
+# load taxonomy table and transform it
+taxonomy <- read_csv(here("data","taxonomy.csv"))
+taxonomy_collapsed <- taxonomy %>%
+  mutate(
+    species = case_when(
+      # fix two misspellings
+      species == "Charybdis (Charybdis) hawaiensis" ~ "Charybdis Charybdis hawaiensis",
+      species == "Catoptrus inequalis" ~ "Catoptrus inaequalis",
+      # make morphospecies into just "genus/family sp."
+      str_detect(species,"sp[0-9]+") ~ str_c(coalesce(genus,family)," sp."),
+      .default = species
+    ),
+    # reset known taxon level
+    taxon_level = case_when(
+      species == str_glue("{genus} sp.") ~ "genus",
+      species == str_glue("{family} sp.") ~ "family",
+      .default = taxon_level
+    )
+  ) %>%
+  # assign them new otu IDs
+  group_by(species) %>%
+  mutate( old_otu = otu,  otu = str_replace_all(str_replace(species,fixed(" sp."),"")," ","_")) %>%
+  ungroup() %>%
+  select(-ends_with("otu"),ends_with("otu")) %>%
+  arrange(family,genus,species)
+
+# make a lookup table
+tax_lookup <- taxonomy_collapsed %>% select(ends_with("otu"))
+
+
+# now remap the community matrix
+crabs <- read_csv(here("data","crabs.csv"))
+
+crabs_collapsed <- crabs %>%
+  # pivot long and get the new otu IDs
+  pivot_longer(-unit,names_to = "otu",values_to = "count") %>%
+  inner_join(tax_lookup,by=c("otu" = "old_otu")) %>%
+  select(-otu,unit,otu=otu.y,count) %>%
+  # pivot back to wide and sum values by new otu ID
+  pivot_wider(names_from = "otu", values_from = "count", values_fn = sum)
+
+# make sure the taxonomy table is unique
+taxonomy_collapsed <- taxonomy_collapsed %>% select(otu,taxon_level:species) %>%
+  distinct(otu,taxon_level,family,genus,species)
+
+# save the new collapsed tables
+write_csv(taxonomy_collapsed,here("data","taxonomy_collapsed.csv"))
+write_csv(crabs_collapsed,here("data","crabs_collapsed.csv"))

### 2024-08-27 crabs.R

@@ -1,8 +1,8 @@
 # set the random seed
 set.seed(31337)
 
-run_things <- FALSE
-
+# collapse taxonomy and crabs tables by unknowns
+source("collapse_taxonomy.R")
 # functions ---------------------------------------------------------------
 
 
@@ -255,8 +255,10 @@ setup_crabs <- function() {
   cc <- list()
   # load our main phyloseq object and hellinger-transform it
   cc$crabs_untransformed <- load_ps(
-    here("data","crabs.csv"),"unit",
-    here("data","taxonomy.csv"),"otu",
+    here("data","crabs_collapsed.csv"),"unit",
+    here("data","taxonomy_collapsed.csv"),"otu",
+    # here("data","crabs.csv"),"unit",
+    # here("data","taxonomy.csv"),"otu",
     here("data","metadata.csv")
   ) 
   
@@ -323,19 +325,31 @@ setup_crabs <- function() {
 sample_summary <- function(ps,top_n=5) {
   ss <- list() 
   
+  # make otu lookup table
+  otu_lookup <- tax_table(ps) %>%
+    as.data.frame() %>%
+    as_tibble(rownames="otu") %>%
+    select(otu,species) %>%
+    deframe()
+  
   # basic numbers
+  
+  # total number of individuals
   ss$total <- ps %>%
     sample_sums() %>%
     sum()
   
+  # overall range of abundance 
   ss$abundance_range <- ps %>%
     sample_sums() %>%
     range()
   
+  # overall abundance mean
   ss$abundance_mean <- ps %>%
     sample_sums() %>%
     mean()
   
+  # overall abundace sd
   ss$abundance_sd <- ps %>%
     sample_sums() %>%
     sd()
@@ -345,40 +359,37 @@ sample_summary <- function(ps,top_n=5) {
     taxa_sums() %>%
     sort() %>%
     rev() %>%
-    head(top_n) 
-  
-  otus <- top %>%
-    names() %>%
-    set_names(.,str_replace_all(.,"_"," "))
+    head(top_n) %>%
+    set_names(otu_lookup[names(.)])
   
   ss$top_n <- top
   
-  
   to_keep <- ps %>%
     taxa_tibble() %>%
-    filter(species %in% names(otus)) %>%
+    filter(species %in% names(top)) %>%
     pull(otu)
   
   ss$top_units <- prune_taxa(to_keep,ps) %>%
     ps_standardize("pa") %>%
     taxa_sums() %>% 
     sort() %>%
[…]

### 2024-08-30 manuscript.Rmd
revision updates

@@
   - natgeo: National Geographic Pristine Seas
-abstract:
-  An inability to consistently resolve the environmental and anthropogenic factors that shape coral reefs impacts our understanding of how those factors interact and how coral reef communities may respond to change.
-Logistical challenges have additionally limited the taxonomic resolution and depth ranges investigated, with the focus largely on fishes and corals in the shallowest ~30 m of coral reefs.
-Here, we examine the effects of a suite of environmental, ecological, and anthropogenic factors on the assemblage structure of brachyuran crabs recovered from Autonomous Reef Monitoring Structures on shallow reefs (8--17 m) across the ~2,500 km span of the Hawaiian Archipelago.
-We compare these assemblages with those sampled along a depth gradient (12--90 m) on O&#x02BB;ahu.
-We found 2,284 individuals representing 103 (morpho)species from 22 families.
-Shallow assemblages were significantly different between the Main and Northwest Hawaiian Islands and significantly associated with human impacts, chlorophyll-A, sea-surface temperature, slope, potential larval connectivity, and depth.
-However, we found more variation among mesophotic samples from O&#x02BB;ahu (30--90 m) than among shallow samples across the entire archipelago, suggesting that 90 m of depth is a stronger determinant of brachyuran assemblage structure than the latitudinal, environmental, or anthropogenic gradients across the Hawaiian Islands.
 bibliography: references/citations.json
@@
       - "--lua-filter=resources/author-info-blocks.lua"
+      - "--lua-filter=resources/abstract-section.lua"
       - "--reference-doc=resources/ref.docx"
@@
 # can't get BiodiversityR to work on new mac, but that's ok because
-# all we need is this one function, which miraculously works on its own!
+# all we need is the CAPdiscrim function, which miraculously works on its own!
+# retrieved from: https://github.com/cran/BiodiversityR
 source(here("lib","CAPdiscrim.R"))
@@
 ```{r init, include=FALSE}
+# source utility functions
 source("crabs.R")
 
+# setup community matrices and data summaries
 if (!exists("crabs_done")) {
@@
 
+# some utility funtions
+
+# format a range of numbers
 format_range <- function(vals) {
@@
 
+# basic reusable ggplot theme
 plotz_theme <- function() {
@@
 
+# abstract method to format species name
 format_species <- function(...) UseMethod("format_species")
 
+# format species in a list
 format_species.list <- function(spp,authority=TRUE,...) {
@@
 
+# format species in a character vector
 format_species.character <- function(species,found,authority,auth=TRUE,...) {
@@
 
+# format species in a data frame
 format_species.data.frame <- function(spp,authority=TRUE,...) {
   spp %>%
-    pwalk(~{
+    pmap_chr(~{
       row <- list(...)
       s <- row[c('species','found','authority','otu')]
-      print(format_species(s,authority))
+      format_species(s,authority)
     })
@@
 
+# format percents and numbers
 pc <- function(x,a=0.01) scales::percent_format(accuracy = a)(x)
@@
 
+# format wilcoxon test results
 w_test <- function(test,plain=FALSE) {
@@
 
+# format kruskal-wallis test results
 k_test <- function(test,plain=FALSE) {
@@
 
+# format t-test results
 t_test <- function(test,plain=FALSE) {
@@
 
+# format a p-value nicely
 pval <- function(p) {
@@
 
+# format permdisp test results
 permdisp <- function(bd,accuracy=0.01,perm=999) {
@@
 
+# format permanova test results
 permanova <- function(modl,accuracy=0.01,stat="pseudo-F",by=NULL,sentence=TRUE) {
@@
 
+# format a top-n list of species
 top_list <- function(species,n,units,len=4,fig=NULL) {
@@
 
+# format a markdown grid table from a data frame or file
 print_table <- function(tbl,id="",caption="",hdr_case="sentence",hide=character(0),accuracy = 0.01,format_numbers=TRUE, cell_width = Inf) {
@@
 
-cc$crabs_deep <- cc$crabs_untransformed %>% subset_samples(shallow_deep == "deep")
+# create deep crabs subset
+cc$crabs_deep <- cc$crabs_untransformed %>%
+ subset_samples(shallow_deep == "deep")
+ # prune deep crabs taxa
 cc$crabs_deep <- prune_taxa(taxa_sums(cc$crabs_deep) > 0,cc$crabs_deep)
-cc$crabs_shallow <- cc$crabs_untransformed %>% subset_samples(shallow_deep == "shallow")
+# create shallow crabs subset
+cc$crabs_shallow <- cc$crabs_untransformed %>%
+ subset_samples(shallow_deep == "shallow")
+ # prune its taxa
 cc$crabs_shallow <- prune_taxa(taxa_sums(cc$crabs_shallow) > 0,cc$crabs_shallow)
 
-dir_create(here("cache"))
-if (file_exists(here("cache","all_summary.rds"))) {
-  all_summary <<- read_rds(here("cache","all_summary.rds"))
-} else {
-  all_summary <<- sample_summary(cc$crabs_untransformed,4)
-  write_rds(all_summary,here("cache","all_summary.rds"))
-}
-if (file_exists(here("cache","deep_summary.rds"))) {
-  deep_summary <<- read_rds(here("cache","deep_summary.rds"))
-} else {
-  deep_summary <<- sample_summary(cc$crabs_deep,4)
-  write_rds(deep_summary,here("cache","deep_summary.rds"))
-}
-
+# get average distance from a centroid
+centroid_dist <- function(data) {
+  pts <- t(as.matrix(data))
+  centers <- matrix(apply(pts,1,mean),nrow=2)
+  dists <- sqrt(apply(centers, 2, function(center) {
+    colSums((pts - center)^2)
+  }))
+  return(mean(dists))
+}
+
+# get alpha diversity statistics
 all_alpha <- alpha_diversity(cc$crabs_untransformed)
@@
 
+# color palette for islands
 island_pal <- c(
@@
 )
+
+# color palette for island groups
 island_group_pal <- c(
@@
 
+# number of permutations for adonis tests
 PERM <<- 999
-
+# total number of brachyuran species in Hawai‘i
+hawaiian_crabs <- 284
+# load model terms table from file
 model_terms <- read_csv(here("data","tables","environmental_variables.csv"))
 ```
+
+```{r summaries, include=FALSE, dependson="init"}
+# generate sample summaries. block should be cached 
+# so worms API calls don't happen every times
+all_summary <- sample_summary(cc$crabs_untransformed,4)
+deep_summary <- sample_summary(cc$crabs_deep,4)
+```
+
+```{r taxonomy-validation, include=FALSE, dependson="init"}
+# validate taxonomy using worms / get authorities
+crab_taxonomy <- cc$crabs %>%
+  taxa_tibble() %>%
+  pull(species) %>%
+  split(ceiling(seq_along(.)/50)) %>%
+  map(\(sp) {
+    wm_records_taxamatch(sp) %>%
+      map2(sp,~.x %>% mutate(query = .y)) %>%
+      list_rbind()
+  }) %>% 
+  list_rbind()
+
+crab_taxonomy_validated <- cc$crabs %>%
+  taxa_tibble() %>%
+  left_join(crab_taxonomy,by=c("species" = "query")) %>%
+  mutate(
+    found = rank == "Species",
+    species = case_when(
+      valid_name != scientificname ~ valid_name,
+      .default = species
+    ),
+    authority = case_when(
+      valid_name != scientificname ~ valid_authority,
+      .default = authority
+    )
+  ) %>%
+  mutate(display_species = format_species(.))
+```
+
+<!-- 
+the abstract-section.lua filter allows us to put our abstract down here 
+which allows us to include computed values that couldn't easily go in the yaml header
+-->
+# Abstract
+
+An inability to consistently resolve the environmental and anthropogenic factors that shape coral reefs impacts our understanding of how those factors interact and how coral reef communities may respond to change.
+Logistical challenges have additionally limited the taxonomic resolution and depth ranges investigated, with the focus largely on fishes and corals in the shallowest ~30 m of coral reefs.
+Here, we examine the effects of a suite of environmental, ecological, and anthropogenic factors on the assemblage structure of brachyuran crabs recovered from Autonomous Reef Monitoring Structures on shallow reefs (`r format_range(range(cc$crab_data_shallow_unscaled$depth))` m) across the ~2,500 km span of the Hawaiian Archipelago.
+We compare these assemblages with those sampled along a depth gradient (`r format_range(range(cc$crab_data$depth[cc$crab_data$island == "Oahu"]))` m) on O&#x02BB;ahu.
+We recovered `r num(sum(sample_sums(cc$crabs_untransformed)),1)` individuals representing `r sum(unlist(all_summary$types_breakdown))` taxa from `r all_summary$taxonomy_breakdown$families` families.
+Shallow assemblages were significantly different between the Main and Northwest Hawaiian Islands and significantly associated with human impacts, chlorophyll-A, sea-surface temperature, slope, potential larval connectivity, and depth.
+However, we found more variation among mesophotic samples from O&#x02BB;ahu (30--90 m) than among shallow samples across the entire archipelago, suggesting that 90 m of depth is a stronger determinant of brachyuran assemblage structure than the latitudinal, environmental, or anthropogenic gradients across the Hawaiian Islands.
 
@@
 All adult brachyuran crabs ($\geq$ 2 mm) were photographed, identified, and preserved in 95% ethanol.
-Identifications were made to species level where possible, otherwise crabs were classified into morphospecies represented by suites of shared morphological characters according to expert advice and Poore & Ahyong [-@pooreMarineDecapodCrustacea2023].<!-- Specimens were collected under permits #xxx (NWHI) and #xxx (MHI) and accessioned into the invertebrate zoology collections at the Florida Museum of Natural History (catalog numbers xxx--xxx).-->
+Specimens were initially identified to family or genus using Poore & Ahyong [-@pooreMarineDecapodCrustacea2023] and additionally to species where possible according to expert advice from Scott Godwin (former curator of invertebrate zoology at Bishop Museum) and Gustav Paulay (curator of invertebrate zoology, Florida Museum of Natural History).
+Where species-level identifications were not possible, taxa were collapsed to family or genus (depending on individual identification).
+We recognize that the species-level taxonomy of Indo-Pacific brachyurans remains poorly resolved, particularly for abundant groups like the Chlorodiellinae [@lasleyjrIntegrativeSystematicRevision2014; @lasleyjrPhylogeneticRelationshipsUbiquitous2015], and some "species" are known to comprise complexes [e.g.
+*Perinia tumida*, @servisCharacterizingCoralReef2020; @plaisanceReefassociatedCrustaceanFauna2009].
+However, the prevalence of complexes and cryptic diversity suggests that our assignments are likely to under- rather than overrepresent the true number of taxa.
+Thus, any significant patterns detected likely persist in spite rather than because of cryptic diversity.
 
@@
 Potential drivers of shallow crab community structure were assembled from a previous study examining environmental effects on genetic diversity [@selkoeDNACoralReef2016], NOAA's National Coral Reef Monitoring Program (NCRMP), the Pacific Islands Ocean Observing System (PacIOOS), and MODIS satellite data products compiled by NASA's Physical Oceanography/Ocean Biology Distributed Active Archive Centers (PO.DAAC, OB.DAAC) ([@tbl:env_table; @tbl:arms_metadata_table]).
-Sea-surface temparature (SST) and chlorophyll (Chl-A) values were calculated by averaging monthly means for the duration of the soaking period at each sampling site.
+Sea-surface temparature (SST) and chlorophyll-A (Chl-A) values were calculated by averaging monthly means for the duration of the soaking period at each sampling site.
 For details on satellite data preparation and data sources, see Supplemental material.
@@
 Units contained `r num(all_summary$abundance_mean,0.1)` $\pm$ `r num(all_summary$abundance_sd,0.1)` (mean $\pm$ SD) individuals, and abundance ranged from `r format_range(all_summary$abundance_range)`.
-Crabs were identified to `r sum(unlist(all_summary$types_breakdown))` distinct types (`r all_summary$types_breakdown$species` species and `r all_summary$types_breakdown$morphospecies` morphospecies) in `r all_summary$taxonomy_breakdown$genera` genera and `r all_summary$taxonomy_breakdown$families` families, which accounted for ~36% of the known crab diversity reported from Hawaiian waters [@castroCatalogAnomuranBrachyuran2011].
+Crabs were identified to `r sum(unlist(all_summary$types_breakdown))` distinct taxa (`r all_summary$types_breakdown$species` species-level identifications, `r all_summary$types_breakdown$genus` genus, `r all_summary$types_breakdown$family` family) encompassing `r all_summary$taxonomy_breakdown$genera` genera in `r all_summary$taxonomy_breakdown$families` families, which accounted for ~`r pc(all_summary$types_breakdown$species / hawaiian_crabs,0.1)` of the known crab diversity reported from Hawaiian waters for species-level IDs [@castroCatalogAnomuranBrachyuran2011].
 The four most abundant species were: `r with(all_summary,top_list(top,top_n,top_units,fig="fig_crabs"))`.
-There were `r all_summary$singleton_count` singleton taxa (morpho/species occurring in a single unit across all sites), accounting for ~`r num(all_summary$singleton_count/ntaxa(cc$crabs_untransformed)*100,1)`% of total recovered crab species.
-
-![The four most abundant crab species found across all deployed ARMS units.
-(a) *Chlorodiella laevissima*.
-(b) *Perinia tumida*.
-(c) *Percnon abbreviatum*.
-(d) *Dynomene hispida*.
-Photo credit: Florida Museum of Natural History, licensed [CC BY-NC 3.0](https://creativecommons.org/licenses/by-nc/3.0/)](data/images/4crabs.png){#fig:fig_crabs}
+There were `r all_summary$singleton_count` singleton taxa (taxa occurring in a single unit across all sites), accounting for ~`r pc(all_summary$singleton_count/ntaxa(cc$crabs_untransformed))` of total recovered crab taxa.
+For a breakdown of crab taxa recovered by island and deployment depth, see [@tbl:species_table].
+
+![Representative photographs of the four most abundant crab species found across all deployed ARMS units (specimens shown were not collected in this study).
+(a) *Chlorodiella laevissima* (FLMNH 45490, Kāne&#x02BB;ohe, Hawai&#x02BB;i).
+(b) *Perinia tumida* (FLMNH 45652, Kāne&#x02BB;ohe, Hawai&#x02BB;i).
+(c) *Percnon abbreviatum* (FLMNH 45628, Kāne&#x02BB;ohe, Hawai&#x02BB;i).
+(d) *Dynomene hispida* (FLMNH 69204, Manawai, Hawai&#x02BB;i).
+Photo credits: Florida Museum of Natural History, licensed [CC BY-NC 3.0](https://creativecommons.org/licenses/by-nc/3.0/)](data/images/4crabs.png){#fig:fig_crabs}
 
@@
 
-The `r nsamples(cc$crabs_deep)` units deployed at mesophotic depths $\geq$ 30 m on O&#x02BB;ahu yielded fewer species and individuals compared to the complete dataset, with `r num(sum(sample_sums(cc$crabs_deep)),1)` crabs overall, `r sum(unlist(deep_summary$types_breakdown))` distinct types (`r deep_summary$types_breakdown$species` species and `r deep_summary$types_breakdown$morphospecies` morphospecies), `r num(deep_summary$abundance_mean,0.1)` $\pm$ `r num(deep_summary$abundance_sd,0.1)` individuals per unit, and abundance ranging from `r format_range(deep_summary$abundance_range)`.
-The four most abundant taxa in the deep ARMS did not overlap with those from the complete dataset: `r with(deep_summary,top_list(top,top_n,top_units))`.
-There were `r deep_summary$singleton_count` singletons, which proportionally accounted for ~`r num(deep_summary$singleton_count/ntaxa(cc$crabs_deep)*100,1)`% of species found on deep ARMS.
+The `r nsamples(cc$crabs_deep)` units deployed at mesophotic depths $\geq$ 30 m on O&#x02BB;ahu yielded fewer species and individuals compared to the complete dataset, with `r num(sum(sample_sums(cc$crabs_deep)),1)` crabs overall, `r sum(unlist(deep_summary$types_breakdown))` distinct taxa (`r deep_summary$types_breakdown$species` species-level identifications, `r deep_summary$types_breakdown$genus` genus, `r deep_summary$types_breakdown$family` family), `r num(deep_summary$abundance_mean,0.1)` $\pm$ `r num(deep_summary$abundance_sd,0.1)` individuals per unit, and abundance ranging from `r format_range(deep_summary$abundance_range)`.
+The four most abundant taxa in the deep ARMS did not overlap with the most abundant taxa from the complete dataset: `r with(deep_summary,top_list(top,top_n,top_units))`.
+There were `r deep_summary$singleton_count` singletons, which proportionally accounted for ~`r pc(deep_summary$singleton_count/ntaxa(cc$crabs_deep))` of species found on deep ARMS.
 
@@
     larval_connectivity + human_impact +
+    lat + lon +
     Condition(recovery_year),
@@
   select(term=Term,stat=`F-statistic`,pval=`p-value`)  %>%
-  mutate(term=str_to_lower(term))
+  mutate(
+    term=str_to_lower(term),
+    term = str_replace_all(term,r'(\b[a-z]\b)',replacement=str_to_upper)
+  )
+
+all_terms <- english::english(nrow(model_terms))
 ```
@@
 The best dbRDA model (`r permanova(db_modl,stat="F")`) explained `r pc(RsquareAdj(db_modl)$adj.r.squared,0.01)` (adjusted $\mathrm{R}^2$) of the variability in crab communities across the archipelago ([@fig:fig_dbrda]).
-`r str_to_sentence(english(nterm))` of the seven environmental variables tested were significant and sufficiently uncorrelated.<!--TODO: make "chlorophyll-a" look like "chlorophyll-A"--> The two drivers contributing most to explained variation were `r drivers$term[1]` (F = `r num(drivers$stat[1],0.01)`) and `r drivers$term[2]` (F = `r num(drivers$stat[2],0.01)`) ([@tbl:rda_table]), although the two factors with the longest effect vectors on the dBRDA plot were `r vectors$driver[1]` and `r vectors$driver[2]` ([@fig:fig_dbrda]).
+`r str_to_sentence(english(nterm))` of the `r str_to_lower(english(nrow(model_terms)))` environmental variables tested were significant and sufficiently uncorrelated.
+The two drivers contributing most to explained variation were `r drivers$term[1]` (F = `r num(drivers$stat[1],0.01)`) and `r drivers$term[2]` (F = `r num(drivers$stat[2],0.01)`) ([@tbl:rda_table]), although the two factors with the longest effect vectors on the dBRDA plot were `r vectors$driver[1]` and `r vectors$driver[2]` ([@fig:fig_dbrda]).
 
@@
 
+shallow_cd <- pcap %>%
+  filter(shallow_deep == "shallow") %>%
+  select(sample,LD1,LD2) %>%
+  column_to_rownames("sample") %>%
+  centroid_dist()
+
+deep_cd <- pcap %>%
+  filter(shallow_deep == "deep") %>%
+  select(sample,LD1,LD2) %>%
+  column_to_rownames("sample") %>%
+  centroid_dist()
+
 shallow_dist <- pcap %>%
@@
 Euclidean distance between shallow samples in the CAP ordination: `r num(max(shallow_dist),0.01)`, avg.
-`r num(mean(shallow_dist),0.01)`; max.
+`r num(mean(shallow_dist),0.01)`, mean distance to centroid among shallow samples: `r num(mean(shallow_cd,0.01))`; max.
 Euclidean distance between deep samples: `r num(max(deep_dist),0.01)`, avg.
-`r num(mean(deep_dist),0.01)`) ([@fig:fig_cap]).
+`r num(mean(deep_dist),0.01)`, mean distance to centroid among deep samples: `r num(mean(deep_cd,0.01))`) ([@fig:fig_cap]).
 
@@
 In the Hawaiian archipelago, we found significant differences in the composition of coral reef crab communities between the populated high islands of the MHI and the uninhabited atolls of the NWHI.
-Moreover, these differences were driven by a combination of environmental (sea-surface temperature, chlorophyll concentration, depth), biotic (potential larval immigration), and anthropogenic (cumulative human impacts) factors.
+Moreover, these differences were driven by a combination of environmental (sea-surface temperature, chlorophyll-A concentration, depth), biotic (potential larval immigration), and anthropogenic (cumulative human impacts) factors.
 
@@
 Although we did not consider the effects of individual drivers here, an index representing cumulative anthropogenic impacts was the top driver of brachyuran community structure, similar to findings in previous work examining genetic diversity, species richness and habitat area, quality and stability across the archipelago [@selkoeDNACoralReef2016].
-The next-most significant driver, chlorophyll A, can be used as a proxy for primary productivity.
+The next-most significant driver, chlorophyll-A, can be used as a proxy for primary productivity.
 Productivity gradients across the Hawaiian Archipelago coincide with differences in coral reef community composition [@selkoeDNACoralReef2016], endemic reef fish dominance [@friedlanderDominanceEndemicsReef2020], survival rates of top predators [@bakerEffectVariableOceanic2007], and island geomorphology [@goveNearislandBiologicalHotspots2016].
 Thus, it is not surprising that we find the same top drivers of community structure for one of the primary taxonomic groups inhabiting coral reefs, the brachyuran crabs, across the Hawaiian archipelago.
-What is surprising is the consistent significance of depth as a primary driver of community composition across these gradients.
+What is surprising, given the narrow depth range within which shallow ARMS units were deployed, is the consistent significance of depth as a primary driver of community composition across these gradients.
 
@@
 Since we lack environmental data for mesophotic samples, we cannot make conclusions about the specific influences driving the differences we found.
-However, light and temperature have been proposed to be important with regard to depth zonation in benthic communities [@kahngCommunityEcologyMesophotic2010; @lesserGlobalCommunityBreaks2019; @pyleComprehensiveInvestigationMesophotic2016; @pyleMesophoticCoralEcosystems2019].
+However, light and temperature have been proposed to be important bottom-up drivers of depth zonation in benthic communities [@kahngCommunityEcologyMesophotic2010; @lesserGlobalCommunityBreaks2019; @pyleComprehensiveInvestigationMesophotic2016; @pyleMesophoticCoralEcosystems2019].
 Oceanographic conditions may be inconsistent close to shore, but the average mixed-layer depth calculated (using 0.8 &deg;C $\upDelta$T [@karaOptimalDefinitionOcean2000]) from CTD casts taken at station ALOHA (22.75&deg;N, 158.00&deg;W; within the latitudinal expanse of the MHI) was 84.6 $\pm$ 26 m over the course of the year (based on monthly averages from casts taken across the duration of ARMS unit deployments [@HawaiiOceanTimeseries]).
 Thus, given the assumption of similar temperature profiles near the O&#x02BB;ahu shoreline, temperature alone appears insufficient to explain the zonation patterns we observed.
-In contrast, the lower depth boundaries of mesophotic reef communities are largely determined by surface irradiance [@kahngCommunityEcologyMesophotic2010; @hindersteinThemeSectionMesophotic2010], so light seems a possible candidate to be a significant abiotic driver of community structure.
+In contrast, the lower depth boundaries of mesophotic reef communities are largely defined by surface irradiance [@kahngCommunityEcologyMesophotic2010; @hindersteinThemeSectionMesophotic2010], so light seems a possible candidate as an abiotic factor influencing community structure.
 
@@
 ## Processing of satellite-derived environmental data
-Level-3 mapped (9km) sea-surface temperature (SST) and chlorophyll A (chl-A) data were retrieved from NASA's Physical Oceanography Distributed Active Archive Center [PO.DAAC; @https://doi.org/10.5067/modsa-mo9d9] and Ocean Biology Distributed Active Archive Center [OB.DAAC; @NASAOceanBiologyProcessingGroup_2022] respectively.
+Level-3 mapped (9km) sea-surface temperature (SST) and chlorophyll-A (chl-A) data were retrieved from NASA's Physical Oceanography Distributed Active Archive Center [PO.DAAC; @https://doi.org/10.5067/modsa-mo9d9] and Ocean Biology Distributed Active Archive Center [OB.DAAC; @NASAOceanBiologyProcessingGroup_2022] respectively.
 Monthly summary datasets were downloaded for the period of 2009--2016 and averaged over each of the ARMS unit soaking periods (e.g.
@@
   # geom_mark_ellipse(color="black",alpha=0.2,show.legend=FALSE, linewidth=0.07) +
-  scale_fill_manual(values=island_group_pal) + 
+  scale_fill_manual(values=island_group_pal,name="Island group") + 
   geom_segment(data=loadings,aes(x=0,y=0,xend=PC1,yend=PC2),color="black",size=0.5,arrow=arrow(length=unit(0.02,"npc"))) + 
@@
 
+```{r species-table, results="asis", dependson="taxonomy-validation"}
+ species_tbl <- cc$crabs %>%
+  psmelt() %>%
+  # mutate(dz = str_glue("{depth_zone}_{island_group}")) %>%
+  group_by(OTU,island_group,hawaiian_name,depth_zone,depth,family,genus,species) %>%
+  summarise(n=sum(Abundance)) %>%
+  ungroup() %>%
+  filter(n > 0) %>%
+  # distinct(dz,family,genus,species) %>%
+  group_by(species) %>%
+  mutate(zones = str_c(depth_zone,collapse="|")) %>%
+  ungroup() %>%
+  mutate(depth_exclusive = depth_zone == zones) %>%
+  arrange(island_group,hawaiian_name,depth_zone,depth,family,genus,species) %>%
+  left_join(crab_taxonomy_validated,by=c("OTU" = "otu")) %>%
+  select(hawaiian_name,island_group,depth_zone,depth,family.y,display_species,depth_exclusive) %>%
+  rename(island = hawaiian_name,valid_species=display_species) %>%
+  mutate(
+    across(c(island_group,depth_zone),str_to_sentence),
+    depth_exclusive = if_else(depth_exclusive,"Yes","No")
+  ) %>%
+  rename_with(\(n) {
+    n %>%
+      str_replace_all("depth$","depth (m)") %>%
+      str_replace_all("_"," ") %>%
+      str_replace_all("\\.y$","") %>%
+      str_replace_all("^display.","") %>%
+      str_to_sentence()
+  })
+print_table(species_tbl,"species_table","List of crab taxa detected by island and ARMS unit deployment depth",hdr_case="none",accuracy=1)
+```
+
+\newpage
+
 ```{r rda-table,results="asis",dependson="rda-analysis"}

### 2024-08-30 crabs.R

@@ -289,7 +289,12 @@ setup_crabs <- function() {
     # this is the shallow subset
   cc$crabs_shallow <- cc$crabs %>%
     subset_samples(region != "mce") 
-    # we're gonna mess with the sample data a bit
+  
+  # save an unscaled version of the shallow crab metadata
+  cc$crab_data_shallow_unscaled <- cc$crabs_shallow %>%
+    sample_tibble(sample_col = "sample")
+  
+  # we're gonna mess with the sample data a bit
   # first, get rid of rows that have NAs
   cc$crab_data_shallow <- cc$crabs_shallow %>%
     sample_tibble(sample_col = "unit") %>%

### 2024-08-30 manuscript.Rmd
change "connectivity" to more-accurate "immigration"

@@
 -->
-# Abstract
+# Abstract {-}
 
@@
 We recovered `r num(sum(sample_sums(cc$crabs_untransformed)),1)` individuals representing `r sum(unlist(all_summary$types_breakdown))` taxa from `r all_summary$taxonomy_breakdown$families` families.
-Shallow assemblages were significantly different between the Main and Northwest Hawaiian Islands and significantly associated with human impacts, chlorophyll-A, sea-surface temperature, slope, potential larval connectivity, and depth.
+Shallow assemblages were significantly different between the Main and Northwest Hawaiian Islands and significantly associated with human impacts, chlorophyll-A, sea-surface temperature, slope, potential larval immigration, and depth.
 However, we found more variation among mesophotic samples from O&#x02BB;ahu (30--90 m) than among shallow samples across the entire archipelago, suggesting that 90 m of depth is a stronger determinant of brachyuran assemblage structure than the latitudinal, environmental, or anthropogenic gradients across the Hawaiian Islands.
@@
 The four most abundant species were: `r with(all_summary,top_list(top,top_n,top_units,fig="fig_crabs"))`.
-There were `r all_summary$singleton_count` singleton taxa (taxa occurring in a single unit across all sites), accounting for ~`r pc(all_summary$singleton_count/ntaxa(cc$crabs_untransformed))` of total recovered crab taxa.
-For a breakdown of crab taxa recovered by island and deployment depth, see [@tbl:species_table].
-
-![Representative photographs of the four most abundant crab species found across all deployed ARMS units (specimens shown were not collected in this study).
+There were `r all_summary$singleton_count` singleton taxa (taxa occurring in a single unit across all sites), accounting for ~`r pc(all_summary$singleton_count/ntaxa(cc$crabs_untransformed),0.1)` of total recovered crab taxa.
+For a breakdown of specific crab taxa recovered by island and deployment depth, see [@tbl:species_table].
+
+![Representative photographs of the four most abundant crab species found across all deployed ARMS units (specimens shown were not collected as part of this study).
 (a) *Chlorodiella laevissima* (FLMNH 45490, Kāne&#x02BB;ohe, Hawai&#x02BB;i).
@@
     chl + sst + slope + coral_cover +
-    larval_connectivity + human_impact +
+    larval_immigration + human_impact +
     lat + lon +

### 2024-08-30 crabs.R

@@ -306,9 +306,9 @@ setup_crabs <- function() {
   cc$crab_data_shallow <- cc$crabs_shallow %>%
     sample_tibble(sample_col = "unit") %>%
     select(-sst,-chl) %>%
-    # select(unit,region,island_group,island,lat,lon,depth,chl=chl_sat,sst=sst_sat,slope,coral_cover,closest_island,larval_connectivity,human_impact) %>%
-    select(unit,region,island_group,recovery_year,island,hawaiian_name,lat,lon,depth,chl=chlorophyll_oracle,sst=sst_sat,slope,coral_cover,closest_island,larval_connectivity,human_impact) %>%
-    # select(unit,region,island_group,island,lat,lon,depth,chl=chl_new,sst=sst_new,slope,coral_cover,closest_island,larval_connectivity,human_impact) %>%
+    # select(unit,region,island_group,island,lat,lon,depth,chl=chl_sat,sst=sst_sat,slope,coral_cover,closest_island,larval_immigration,human_impact) %>%
+    select(unit,region,island_group,recovery_year,island,hawaiian_name,lat,lon,depth,chl=chlorophyll_oracle,sst=sst_sat,slope,coral_cover,closest_island,larval_immigration,human_impact) %>%
+    # select(unit,region,island_group,island,lat,lon,depth,chl=chl_new,sst=sst_new,slope,coral_cover,closest_island,larval_immigration,human_impact) %>%
     mutate(across(where(is.numeric),~as.numeric(scale(.x)))) %>%
     column_to_rownames("unit") 
     # reassociate the new sample data
@@ -499,9 +499,9 @@ alpha_diversity <- function(ps,measures=c("Observed","InvSimpson")) {
   # 
   # scaled_richness <- all_richness %>%
   #   mutate(
-  #     across(c(lat,depth,chl,sst,slope,coral_cover,closest_island,larval_connectivity,human_impact),~as.numeric(scale(.x)))
+  #     across(c(lat,depth,chl,sst,slope,coral_cover,closest_island,larval_immigration,human_impact),~as.numeric(scale(.x)))
   #   )
-  # # lme_richness <- lmer(Observed ~ lat + depth + chl + sst + slope + coral_cover + closest_island + larval_connectivity + human_impact + (1|region/island), data=scaled_richness)
[…]

### 2024-08-30 collapse_taxonomy.R
moved generated stuff to generated directory

@@ -44,5 +44,6 @@ taxonomy_collapsed <- taxonomy_collapsed %>% select(otu,taxon_level:species) %>%
   distinct(otu,taxon_level,family,genus,species)
 
 # save the new collapsed tables
-write_csv(taxonomy_collapsed,here("data","taxonomy_collapsed.csv"))
-write_csv(crabs_collapsed,here("data","crabs_collapsed.csv"))
+dir_create(here("data","generated"))
+write_csv(taxonomy_collapsed,here("data","generated","taxonomy_collapsed.csv"))
+write_csv(crabs_collapsed,here("data","generated","crabs_collapsed.csv"))

### 2024-08-30 crabs.R

@@ -255,8 +255,8 @@ setup_crabs <- function() {
   cc <- list()
   # load our main phyloseq object and hellinger-transform it
   cc$crabs_untransformed <- load_ps(
-    here("data","crabs_collapsed.csv"),"unit",
-    here("data","taxonomy_collapsed.csv"),"otu",
+    here("data","generated","crabs_collapsed.csv"),"unit",
+    here("data","generated","taxonomy_collapsed.csv"),"otu",
     # here("data","crabs.csv"),"unit",
     # here("data","taxonomy.csv"),"otu",
     here("data","metadata.csv")

### 2024-08-31 manuscript.Rmd
streamlined processing

@@
 
-```{r summaries, include=FALSE, dependson="init"}
-# generate sample summaries. block should be cached 
-# so worms API calls don't happen every times
-all_summary <- sample_summary(cc$crabs_untransformed,4)
-deep_summary <- sample_summary(cc$crabs_deep,4)
-```
-
 ```{r taxonomy-validation, include=FALSE, dependson="init"}
@@
   mutate(display_species = format_species(.))
+```
+
+```{r summaries, include=FALSE, dependson="init"}
+# generate sample summaries. block should be cached 
+# so worms API calls don't happen every times
+all_summary <- sample_summary(cc$crabs_untransformed,4,crab_taxonomy_validated)
+deep_summary <- sample_summary(cc$crabs_deep,4,crab_taxonomy_validated)
 ```
@@
 The four most abundant taxa in the deep ARMS did not overlap with the most abundant taxa from the complete dataset: `r with(deep_summary,top_list(top,top_n,top_units))`.
-There were `r deep_summary$singleton_count` singletons, which proportionally accounted for ~`r pc(deep_summary$singleton_count/ntaxa(cc$crabs_deep))` of species found on deep ARMS.
+There were `r deep_summary$singleton_count` singletons, which proportionally accounted for ~`r pc(deep_summary$singleton_count/ntaxa(cc$crabs_deep),0.1)` of species found on deep ARMS.
 

### 2024-08-31 crabs.R

@@ -327,7 +327,7 @@ setup_crabs <- function() {
   return(cc)
 }
 
-sample_summary <- function(ps,top_n=5) {
+sample_summary <- function(ps,top_n=5,taxonomy=NULL) {
   ss <- list() 
   
   # make otu lookup table
@@ -384,20 +384,30 @@ sample_summary <- function(ps,top_n=5) {
   # make it the same order as the top 5 species
   ss$top_units <- ss$top_units[match(names(ss$top_units),names(top))]
   
-  ss$top <- wm_records_taxamatch(names(top)) %>%
-    map2(names(top),~{
-      if (nrow(.x) > 0) {
-        .x <- .x %>% 
-          filter(rank == "Species") %>%
-          select(species = scientificname, authority) %>%
-          slice(1)  %>%
-          mutate(found = TRUE,otu = otu_lookup[.y])
-      } 
-      if (nrow(.x) == 0) {
-        .x <- list(species = .y, authority = NA, found = FALSE, otu=otu_lookup[.y])
-      }
-      return(as.list(.x))
-    })
+  if (is.null(taxonomy)) {
+    ss$top <- wm_records_taxamatch(names(top)) %>%
+      map2(names(top),~{
+        if (nrow(.x) > 0) {
+          .x <- .x %>% 
+            filter(rank == "Species") %>%
+            select(species = scientificname, authority) %>%
+            slice(1)  %>%
+            mutate(found = TRUE,otu = otu_lookup[.y])
+        } 
+        if (nrow(.x) == 0) {
+          .x <- list(species = .y, authority = NA, found = FALSE, otu=otu_lookup[.y])
+        }
+        return(as.list(.x))
+      })
+  } else {
+    ss$top <- taxonomy %>%
+      filter(species %in% names(top)) %>%
+      mutate(found = species == valid_name) %>%
+      select(species,authority=valid_authority,found,otu) %>%
+      slice(match(names(top),species)) %>%
+      split(seq(nrow(.))) %>%
+      map(as.list)
+  }
   
   
   ts <- ps %>% taxa_sums()

### 2024-09-17 manuscript.Rmd
human impacts code and other updates

@@
 
-# set random seed for consistency
-set.seed(31337)
 ```
@@
 # validate taxonomy using worms / get authorities
+# worms API call results are cached
 crab_taxonomy <- cc$crabs %>%
@@
   pull(species) %>%
+  # split into subsets of 50 taxa
   split(ceiling(seq_along(.)/50)) %>%
   map(\(sp) {
+    # validate taxa using fuzzy matching
     wm_records_taxamatch(sp) %>%
+      # smash results into a single tibble
       map2(sp,~.x %>% mutate(query = .y)) %>%
@@
 
+# max validated taxonomy table
 crab_taxonomy_validated <- cc$crabs %>%
@@
     found = rank == "Species",
+    # substitute valid name if ours isn't
     species = case_when(
@@
 ```{r summaries, include=FALSE, dependson="init"}
-# generate sample summaries. block should be cached 
-# so worms API calls don't happen every times
+# generate sample summaries. code block is cached 
+# so worms API calls don't happen every time
 all_summary <- sample_summary(cc$crabs_untransformed,4,crab_taxonomy_validated)
@@
 A total of 89 ARMS units were deployed over the course of six years (2009--2016) in shallow coral reef habitat (8--17 m depth) in both the Main (MHI) and Northwest Hawaiian Islands (NWHI) ([@fig:fig_map; @tbl:arms_table]).
+Only units deployed on outer reefs were included in this study.
 To explore the effects of depth, we also present data from an earlier study by Hurley et al.
@@
 
-Potential drivers of shallow crab community structure were assembled from a previous study examining environmental effects on genetic diversity [@selkoeDNACoralReef2016], NOAA's National Coral Reef Monitoring Program (NCRMP), the Pacific Islands Ocean Observing System (PacIOOS), and MODIS satellite data products compiled by NASA's Physical Oceanography/Ocean Biology Distributed Active Archive Centers (PO.DAAC, OB.DAAC) ([@tbl:env_table; @tbl:arms_metadata_table]).
-Sea-surface temparature (SST) and chlorophyll-A (Chl-A) values were calculated by averaging monthly means for the duration of the soaking period at each sampling site.
-For details on satellite data preparation and data sources, see Supplemental material.
+Potential drivers of brachyuran crab community structure were assembled from previous studies [@selkoeEvaluatingAnthropogenicThreats2008; @halpernGlobalMapHuman2008; @selkoeDNACoralReef2016; @selkoeMapHumanImpacts2009], NOAA's National Coral Reef Monitoring Program (NCRMP), the Pacific Islands Ocean Observing System (PacIOOS), and MODIS satellite data products compiled by NASA's Physical Oceanography/Ocean Biology Distributed Active Archive Centers [@NASAOceanBiologyProcessingGroup_2022; @https://doi.org/10.5067/modsa-mo9d9].
+Variables were either intrinsic (site depth, latitude, longitude, island slope) or derived from remote sensing and/or modeling (chlorophyll-A, sea-surface temperature, coral cover, potential larval immigration, cumulative human impacts) ([@tbl:env_table; @tbl:arms_metadata_table]).
+Sea-surface temperature (SST) and chlorophyll-A values were calculated by averaging monthly means across the duration of the soaking period at each sampling site using AQUA MODIS 4 km-resolution gridded datasets.
+The cumulative human impacts index, a dimensionless numerical value representing spatially-explicit cumulative anthropogenic impacts on various marine ecosystems, is based on a synthesis of a combination of "high resolution, annual data on the intensity of 14 human stressors and their impact on 21 marine ecosystems over 11 years (2003--2013)" [@halpernGlobalMapHuman2008; @halpernRecentPaceChange2019].
+For details regarding data import, preparation, and synthesis, see Supplemental material.
 Due to logistical limitations, environmental data were collected only for shallow samples (\< 30 m).
@@
 library(formula.tools)
+
+# set random seed for consistency
+set.seed(99917)
 
@@
 
+# dd <- cc$crab_data_shallow %>%
+#   # select(-c(chl,sst,human_impact)) %>%
+#   rename(chl=X4km_chl,sst=X4km_sst,human_impact=cumulative_impacts)
+
 # calculate full dbRDA model
@@
     chl + sst + slope + coral_cover +
+    # X4km_chl + X4km_sst + slope + coral_cover +
+    # chl_4km_terra + sst_4km_terra + slope + coral_cover +
+    # chl_9km_terra + sst_9km_terra + slope + coral_cover +
     larval_immigration + human_impact +
+    # larval_immigration + cumulative_impacts +
     lat + lon +
@@
   data=cc$crab_data_shallow,
+  # data=dd,
   distance="bray"
@@
 
-# get rid of variables with VIF > 3
-db_modl <- reduce_model(db_modl,cc$crab_otus_shallow,cc$crab_data_shallow)
+# reduce model to terms with VIF <= 3
+db_modl <- reduce_model(db_modl,cc$crab_otus_shallow,cc$crab_data_shallow,vif_limit = 4)
+# db_modl <- reduce_model(db_modl,cc$crab_otus_shallow,dd,vif_limit = 3)
 
@@
 # get marginal term information and sort by descending f-statistic
-modl_perm <- anova(db_modl,by="margin") %>%
+modl_perm <- anova(db_modl,by="margin",permutations = PERM) %>%
+  # as_tibble(rownames="term") %>%
+  # janitor::clean_names() %>%
   broom::tidy() %>%
@@
 `r str_to_sentence(english(nterm))` of the `r str_to_lower(english(nrow(model_terms)))` environmental variables tested were significant and sufficiently uncorrelated.
-The two drivers contributing most to explained variation were `r drivers$term[1]` (F = `r num(drivers$stat[1],0.01)`) and `r drivers$term[2]` (F = `r num(drivers$stat[2],0.01)`) ([@tbl:rda_table]), although the two factors with the longest effect vectors on the dBRDA plot were `r vectors$driver[1]` and `r vectors$driver[2]` ([@fig:fig_dbrda]).
+The two drivers contributing most to explained variation were `r drivers$term[1]` (F = `r num(drivers$stat[1],0.01)`) and `r drivers$term[2]` (F = `r num(drivers$stat[2],0.01)`) ([@tbl:rda_table]).
+These factors also resulted in the longest effect vectors in the dbRDA plot ([@fig:fig_dbrda]).
 
@@
 Thus, it is not surprising that we find the same top drivers of community structure for one of the primary taxonomic groups inhabiting coral reefs, the brachyuran crabs, across the Hawaiian archipelago.
-What is surprising, given the narrow depth range within which shallow ARMS units were deployed, is the consistent significance of depth as a primary driver of community composition across these gradients.
+What is surprising---given the narrow depth range within which shallow ARMS units were deployed---is the consistent significance of depth as a primary driver of community composition across these gradients.
 
@@
 ## Processing of satellite-derived environmental data
-Level-3 mapped (9km) sea-surface temperature (SST) and chlorophyll-A (chl-A) data were retrieved from NASA's Physical Oceanography Distributed Active Archive Center [PO.DAAC; @https://doi.org/10.5067/modsa-mo9d9] and Ocean Biology Distributed Active Archive Center [OB.DAAC; @NASAOceanBiologyProcessingGroup_2022] respectively.
+Level-3 mapped (4 km grid-size) sea-surface temperature (SST) and chlorophyll-A (chl-A) data were retrieved from NASA's Physical Oceanography Distributed Active Archive Center [PO.DAAC; @https://doi.org/10.5067/modsa-mo9d9] and Ocean Biology Distributed Active Archive Center [OB.DAAC; @NASAOceanBiologyProcessingGroup_2022] respectively.
 Monthly summary datasets were downloaded for the period of 2009--2016 and averaged over each of the ARMS unit soaking periods (e.g.
@@
 Specific monthly-average dataset names can be found in @tbl:nasa_table.
+
+## Processing of cumulative human impacts data
+Gridded cumulative human impacts raster data calcuated by Halpern et al [-@halpernRecentPaceChange2019] was downloaded from the Knowledge Network for Biocomplexity (KNB) repository [@frazierRecentPaceChange2019].
+Human impacts datasets were downloaded for each available year in this study (2009--2013) and averaged over ARMS unit soaking periods (as above).
+Since cumulative impacts data was only available up until 2013, we used the combined average of all available years for the 2013--2016 soaking period.
+Since the availability of data points at individual coordinates was highly variable, we averaged impact scores for each sampled island in the study.
 
@@
 ```{r species-table, results="asis", dependson="taxonomy-validation"}
+# make table of all sampled taxa
 species_tbl <- cc$crabs %>%
   psmelt() %>%
-  # mutate(dz = str_glue("{depth_zone}_{island_group}")) %>%
+  # sum counts by island group, depth zone, and species
   group_by(OTU,island_group,hawaiian_name,depth_zone,depth,family,genus,species) %>%
@@
   ungroup() %>%
+  # keep species we actually found
   filter(n > 0) %>%
-  # distinct(dz,family,genus,species) %>%
+  # figure out if a species is exclusive to a depth zone
   group_by(species) %>%
@@
   arrange(island_group,hawaiian_name,depth_zone,depth,family,genus,species) %>%
+  # join in validated taxonomy
   left_join(crab_taxonomy_validated,by=c("OTU" = "otu")) %>%

### 2024-09-17 crabs.R

@@ -305,11 +305,15 @@ setup_crabs <- function() {
   # we care about and scale the numeric ones to unit variance
   cc$crab_data_shallow <- cc$crabs_shallow %>%
     sample_tibble(sample_col = "unit") %>%
-    select(-sst,-chl) %>%
+    # select(-sst,-chl) %>%
     # select(unit,region,island_group,island,lat,lon,depth,chl=chl_sat,sst=sst_sat,slope,coral_cover,closest_island,larval_immigration,human_impact) %>%
-    select(unit,region,island_group,recovery_year,island,hawaiian_name,lat,lon,depth,chl=chlorophyll_oracle,sst=sst_sat,slope,coral_cover,closest_island,larval_immigration,human_impact) %>%
-    # select(unit,region,island_group,island,lat,lon,depth,chl=chl_new,sst=sst_new,slope,coral_cover,closest_island,larval_immigration,human_impact) %>%
-    mutate(across(where(is.numeric),~as.numeric(scale(.x)))) %>%
+    
+    select(unit,region,island_group,deployment_year,recovery_year,island,hawaiian_name,lat,lon,depth,contains("chl"),contains("sst"),slope,coral_cover,closest_island,larval_immigration,contains('impact')) %>%
+    # select(unit,region,island_group,recovery_year,island,hawaiian_name,lat,lon,depth,chl=chl_sat,sst=sst_sat,slope,coral_cover,closest_island,larval_immigration,human_impact) %>%
+    
+    # select(unit,region,island_group,recovery_year,island,hawaiian_name,lat,lon,depth,chl=chlorophyll_oracle,sst=sst_sat,slope,coral_cover,closest_island,larval_immigration,human_impact) %>%
+    # select(unit,region,island_group,recovery_year,island,lat,lon,depth,chl=chl_new,sst=sst_new,slope,coral_cover,closest_island,larval_immigration,human_impact) %>%
+    mutate(across(c(where(is.numeric),-recovery_year,-deployment_year),~as.numeric(scale(.x)))) %>%
     column_to_rownames("unit") 
     # reassociate the new sample data
   sample_data(cc$crabs_shallow) <- cc$crab_data_shallow

### 2024-09-17 get-data.R

@@ -1,23 +1,17 @@
+# prep/load libraries ---------------------------------------------------------------------------------------------
 library(tidyverse)
 library(sf)
-library(raster)
 library(here)
 library(lubridate)
 library(httr)
 library(fs)
 library(furrr)
-
-
+library(httr)
+library(terra)
+source("extract_raster.R")
 
 # set random seed for consistency
 set.seed(31337)
-
-# setup parallel workers
-if (rstudioapi::isAvailable()) {
-  plan(multisession,workers=8)
-} else {
-  plan(multicore,workers=8)
-}
 
 # output directory
 outdir <- here("output","modis")
@@ -27,19 +21,48 @@ dir_create(outdir)
 monthly <- here("data","chlorophyll","monthly")
 dir_create(monthly)
 
+# human impacts directory/file
+impacts_data <- here("data","impact")
+impacts_dir <- here("output","impact")
+dir_create(impacts_data)
+dir_create(impacts_dir)
+
 
 if (!exists("skip_dl")) {
   skip_dl <<- TRUE
 }
-# download code -----------------------------------------------------------
-library(httr)
+
+# load metadata and get deployment range
+arms <- read_csv(here("data","metadata.csv")) %>%
+  rename(sample=unit) %>%
+  unite("range",deployment_year,recovery_year,sep="--",remove=FALSE) 
+
+ranges <- arms %>%
+  distinct(range) %>%
+  pull(range)
+
+# create sf object for arms deployment sites
+arms_sf <- read_csv("data/metadata.csv") %>%
+  st_as_sf(coords=c("lon","lat"),crs=4326,remove=FALSE)
+  
+
+# initial point search radius
+initial_radius <- 10
+
+# download chl/sst datasets ---------------------------------------------------------------------------------------
 library(raster)
 
+# setup parallel workers
+if (rstudioapi::isAvailable()) {
+  plan(multisession,workers=8)
+} else {
+  plan(multicore,workers=8)
+}
+
 if (!skip_dl) {
-    # netrc file must exist with login credentials to nasa data warehouse
+  # netrc file must exist with login credentials to nasa data warehouse
   netrc <- here("data","chlorophyll","netrc")
   # load data urls
[…]

### 2024-09-17 manuscript.Rmd
do separate ref lists for main and supp, fix nasa cites

@@
     pandoc_args:
+      - "--filter=resources/pandoc-crossref"
+      - "--citeproc"
       - "--toc-depth=2"
-      - "--filter=resources/pandoc-crossref"
       - "--lua-filter=resources/scholarly-metadata.lua"
@@
       - "--reference-doc=resources/ref.docx"
-    keep_md: false
+      - "--lua-filter=resources/multirefs.lua"
+    keep_md: true
 tblPrefix: table
@@
   - \usepackage{upgreek}
+multiref_no_duplicates: true
 ---
@@
 
-Potential drivers of brachyuran crab community structure were assembled from previous studies [@selkoeEvaluatingAnthropogenicThreats2008; @halpernGlobalMapHuman2008; @selkoeDNACoralReef2016; @selkoeMapHumanImpacts2009], NOAA's National Coral Reef Monitoring Program (NCRMP), the Pacific Islands Ocean Observing System (PacIOOS), and MODIS satellite data products compiled by NASA's Physical Oceanography/Ocean Biology Distributed Active Archive Centers [@NASAOceanBiologyProcessingGroup_2022; @https://doi.org/10.5067/modsa-mo9d9].
+Potential drivers of brachyuran crab community structure were assembled from previous studies [@selkoeEvaluatingAnthropogenicThreats2008; @halpernGlobalMapHuman2008; @selkoeDNACoralReef2016; @selkoeMapHumanImpacts2009], NOAA's National Coral Reef Monitoring Program (NCRMP), the Pacific Islands Ocean Observing System (PacIOOS), and MODIS satellite data products compiled by NASA's Physical Oceanography/Ocean Biology Distributed Active Archive Centers [@nasagoddardspaceflightcenterModerateresolutionImagingSpectroradiometer2022; @nasaobpgMODISAquaGlobal2020].
 Variables were either intrinsic (site depth, latitude, longitude, island slope) or derived from remote sensing and/or modeling (chlorophyll-A, sea-surface temperature, coral cover, potential larval immigration, cumulative human impacts) ([@tbl:env_table; @tbl:arms_metadata_table]).
@@
 
-::: {#refs}
-:::
+<div class="multi-refs"></div>
 
@@
 ## Processing of satellite-derived environmental data
-Level-3 mapped (4 km grid-size) sea-surface temperature (SST) and chlorophyll-A (chl-A) data were retrieved from NASA's Physical Oceanography Distributed Active Archive Center [PO.DAAC; @https://doi.org/10.5067/modsa-mo9d9] and Ocean Biology Distributed Active Archive Center [OB.DAAC; @NASAOceanBiologyProcessingGroup_2022] respectively.
+Level-3 mapped (4 km grid-size) sea-surface temperature (SST) and chlorophyll-A (chl-A) data were retrieved from NASA's Physical Oceanography Distributed Active Archive Center [PO.DAAC; @nasaobpgMODISAquaGlobal2020] and Ocean Biology Distributed Active Archive Center [OB.DAAC; @nasagoddardspaceflightcenterModerateresolutionImagingSpectroradiometer2022] respectively.
 Monthly summary datasets were downloaded for the period of 2009--2016 and averaged over each of the ARMS unit soaking periods (e.g.
@@
 pcdata <- cds %>%
-  select(where(is.numeric),-c(lon,lat),-closest_island)
+  select(where(is.numeric),-c(lon,lat),-closest_island) %>%
+  select(deployment_year,depth,chl,sst,slope:human_impact,-contains('radius'))
 
@@
 ```
+
+## References
+
+<div class="multi-refs"></div>

### 2024-09-17 crabs.R

@@ -313,7 +313,7 @@ setup_crabs <- function() {
     
     # select(unit,region,island_group,recovery_year,island,hawaiian_name,lat,lon,depth,chl=chlorophyll_oracle,sst=sst_sat,slope,coral_cover,closest_island,larval_immigration,human_impact) %>%
     # select(unit,region,island_group,recovery_year,island,lat,lon,depth,chl=chl_new,sst=sst_new,slope,coral_cover,closest_island,larval_immigration,human_impact) %>%
-    mutate(across(c(where(is.numeric),-recovery_year,-deployment_year),~as.numeric(scale(.x)))) %>%
+    mutate(across(where(is.numeric),~as.numeric(scale(.x)))) %>%
     column_to_rownames("unit") 
     # reassociate the new sample data
   sample_data(cc$crabs_shallow) <- cc$crab_data_shallow

### 2024-09-18 manuscript.Rmd
some updates

@@
 
-A total of 89 ARMS units were deployed over the course of six years (2009--2016) in shallow coral reef habitat (8--17 m depth) in both the Main (MHI) and Northwest Hawaiian Islands (NWHI) ([@fig:fig_map; @tbl:arms_table]).
+A total of 89 ARMS units were deployed over the course of six years (2009--2016) in shallow coral reef habitat (`r format_range(cc$crabs_shallow %>% sample_tibble() %>% pull(depth))` m depth) in both the Main (MHI) and Northwest Hawaiian Islands (NWHI) ([@fig:fig_map; @tbl:arms_table]).
 Only units deployed on outer reefs were included in this study.
 To explore the effects of depth, we also present data from an earlier study by Hurley et al.
-[-@hurleyAssessmentShallowMesophotic2016], in which 27 units were deployed off O&#x02BB;ahu from 2009--2012 in 12--90 m depth; nine units were deployed at 12 m, and six each at 30 m, 60 m, and 90 m ([@tbl:arms_table]).
+[-@hurleyAssessmentShallowMesophotic2016], in which 27 units were deployed off O&#x02BB;ahu from 2009--2012 in `r format_range(cc$crab_data %>% sample_tibble() %>% filter(island == "Oahu") %>% pull(depth))` m depth; nine units were deployed at 12 m, and six each at 30 m, 60 m, and 90 m ([@tbl:arms_table]).
 Shallow units were deployed according to NOAA-RAMP methodology, whereas units $\geq$ 30 m were deployed and recovered using technical mixed gas closed-circuit rebreather diving [see @hurleyAssessmentShallowMesophotic2016].
+Unless referring to specific depth zones, we will refer to any units deployed \< 30 m as "shallow" and those deployed $\geq$ 30 m as "deep".
 
@@
 library(rnaturalearth)
+library(rnaturalearthhires)
 
@@
 All adult brachyuran crabs ($\geq$ 2 mm) were photographed, identified, and preserved in 95% ethanol.
-Specimens were initially identified to family or genus using Poore & Ahyong [-@pooreMarineDecapodCrustacea2023] and additionally to species where possible according to expert advice from Scott Godwin (former curator of invertebrate zoology at Bishop Museum) and Gustav Paulay (curator of invertebrate zoology, Florida Museum of Natural History).
+Specimens were initially identified to family or genus using Poore & Ahyong [-@pooreMarineDecapodCrustacea2023] and additionally to species where possible according to expert advice from Scott Godwin (former curator of invertebrate zoology, Bishop Museum) and Gustav Paulay (curator of invertebrate zoology, Florida Museum of Natural History).
 Where species-level identifications were not possible, taxa were collapsed to family or genus (depending on individual identification).
@@
 However, the prevalence of complexes and cryptic diversity suggests that our assignments are likely to under- rather than overrepresent the true number of taxa.
-Thus, any significant patterns detected likely persist in spite rather than because of cryptic diversity.
+Thus, any significant patterns detected likely persist in spite of rather than because of cryptic diversity.
 
@@
 
-Unless otherwise specified, all analyses were performed using `R` version 4.1.2 [@rcoreteamLanguageEnvironmentStatistical2021].
+Unless otherwise specified, all analyses were performed using `R` version `r R.version$major`.`r R.version$minor` [@rcoreteamLanguageEnvironmentStatistical2021].
 Community matrices were loaded and encapsulated into data objects using the R package `phyloseq` [@mcmurdiePhyloseqPackageReproducible2013] and multivariate community analyses were performed using the package `vegan` [@oksanenVeganCommunityEcology2022] on Bray-Curtis dissimilarities of Hellinger-transformed species abundance matrices [@legendreEcologicallyMeaningfulTransformations2001; @raoReviewCanonicalCoordinates1995].
@@
 
-Potential drivers of brachyuran crab community structure were assembled from previous studies [@selkoeEvaluatingAnthropogenicThreats2008; @halpernGlobalMapHuman2008; @selkoeDNACoralReef2016; @selkoeMapHumanImpacts2009], NOAA's National Coral Reef Monitoring Program (NCRMP), the Pacific Islands Ocean Observing System (PacIOOS), and MODIS satellite data products compiled by NASA's Physical Oceanography/Ocean Biology Distributed Active Archive Centers [@nasagoddardspaceflightcenterModerateresolutionImagingSpectroradiometer2022; @nasaobpgMODISAquaGlobal2020].
+Potential drivers of brachyuran crab community structure were assembled from previous studies [@selkoeDNACoralReef2016; @halpernRecentPaceChange2019], NOAA's National Coral Reef Monitoring Program (NCRMP), the Pacific Islands Ocean Observing System (PacIOOS), and MODIS satellite data products compiled by NASA's Physical Oceanography/Ocean Biology Distributed Active Archive Centers [@nasagoddardspaceflightcenterModerateresolutionImagingSpectroradiometer2022; @nasaobpgMODISAquaGlobal2020].
 Variables were either intrinsic (site depth, latitude, longitude, island slope) or derived from remote sensing and/or modeling (chlorophyll-A, sea-surface temperature, coral cover, potential larval immigration, cumulative human impacts) ([@tbl:env_table; @tbl:arms_metadata_table]).
-Sea-surface temperature (SST) and chlorophyll-A values were calculated by averaging monthly means across the duration of the soaking period at each sampling site using AQUA MODIS 4 km-resolution gridded datasets.
+Sea-surface temperature (SST) and chlorophyll-A values were calculated by averaging monthly means across the duration of the soaking period for each deployment using AQUA MODIS 4 km-resolution gridded datasets.
 The cumulative human impacts index, a dimensionless numerical value representing spatially-explicit cumulative anthropogenic impacts on various marine ecosystems, is based on a synthesis of a combination of "high resolution, annual data on the intensity of 14 human stressors and their impact on 21 marine ecosystems over 11 years (2003--2013)" [@halpernGlobalMapHuman2008; @halpernRecentPaceChange2019].
-For details regarding data import, preparation, and synthesis, see Supplemental material.
+The index of potential larval immigration was imported from [@selkoeDNACoralReef2016], and represents an "estimate of in-closeness centrality" from a biophysical model simulating larval dispersal in the Hawaiian Islands [@selkoeDNACoralReef2016 supp.
+mat.] For details regarding data import, preparation, and synthesis, see Supplemental material.
 Due to logistical limitations, environmental data were collected only for shallow samples (\< 30 m).
@@
 
-For the complete dataset, there was no significant difference in observed species richness between the MHI and NWHI ([@fig:fig_alpha[]{}a]; `r k_test(all_alpha$k_r_island)`), but effective Simpson diversity was significantly different ([@fig:fig_alpha[]{}c]; `r k_test(all_alpha$k_s_island,plain=TRUE)`).
-Richness was not significantly different between shallow and deep samples ([@fig:fig_alpha[]{}b]; `r k_test(all_alpha$k_r_depth,plain=TRUE)`) but effective Simpson diversity was ([@fig:fig_alpha[]{}d]; `r k_test(all_alpha$k_s_depth,plain=TRUE)`).
+For the complete dataset, there was no significant difference in observed species richness (`r k_test(all_alpha$k_r_island)`; [@fig:fig_alpha[]{}a]) or effective Simpson diversity (`r k_test(all_alpha$k_s_island,plain=TRUE)`; [@fig:fig_alpha[]{}c]) between the MHI and NWHI.
+Richness was not significantly different between shallow and deep samples (`r k_test(all_alpha$k_r_depth,plain=TRUE)`; [@fig:fig_alpha[]{}b]) but effective Simpson diversity was (`r k_test(all_alpha$k_s_depth,plain=TRUE)`; [@fig:fig_alpha[]{}d]).
 When only shallow arms \< 30 m were considered, there were no significant differences in alpha diversity between the MHI and NWHI for either index (richness: `r k_test(shallow_alpha$k_r_island,plain=TRUE)`, [@fig:fig_alpha2[]{}a]; effective Simpson: `r k_test(shallow_alpha$k_s_island,plain=TRUE)`, [@fig:fig_alpha2[]{}b]).
@@
 
+<!-- show main bibliography -->
 <div class="multi-refs"></div>
@@
 
-## Processing of satellite-derived environmental data
+## Data import and processing
+
+All data import and processing code for environmental variables used in the dbRDA model can be found in the file `get-data.R` from the associated github repository.
+Details for specific datasets are given below.
+
+### Processing of satellite-derived environmental data
 Level-3 mapped (4 km grid-size) sea-surface temperature (SST) and chlorophyll-A (chl-A) data were retrieved from NASA's Physical Oceanography Distributed Active Archive Center [PO.DAAC; @nasaobpgMODISAquaGlobal2020] and Ocean Biology Distributed Active Archive Center [OB.DAAC; @nasagoddardspaceflightcenterModerateresolutionImagingSpectroradiometer2022] respectively.
@@
 Values (SST or chl-A) were extracted for the geographic coordinates of each sampling site from the averaged dataset for the relevant soaking period.
-Specific monthly-average dataset names can be found in @tbl:nasa_table.
-
-## Processing of cumulative human impacts data
+Original raw monthly-average dataset names can be found in @tbl:nasa_table.
+
+### Processing of cumulative human impacts data
 Gridded cumulative human impacts raster data calcuated by Halpern et al [-@halpernRecentPaceChange2019] was downloaded from the Knowledge Network for Biocomplexity (KNB) repository [@frazierRecentPaceChange2019].
 Human impacts datasets were downloaded for each available year in this study (2009--2013) and averaged over ARMS unit soaking periods (as above).
-Since cumulative impacts data was only available up until 2013, we used the combined average of all available years for the 2013--2016 soaking period.
+Since cumulative impacts data was only available through 2013, we used the combined average of all available years for the 2013--2016 soaking period.
 Since the availability of data points at individual coordinates was highly variable, we averaged impact scores for each sampled island in the study.
+
+### Other datasets
+Coral cover and potential larval immigration were imported from the online supplemtal material accompanying Selkoe et al. [-@selkoeDNACoralReef2016].
+
+## Figures and tables
 
@@
 Notches as in [@fig:fig_alpha].
-Both comparisons are non-significant.', results="asis", dev="svg", fig.width=8, fig.height=4, dependson="init"}
+Both comparisons were non-significant.', results="asis", dev="svg", fig.width=8, fig.height=4, dependson="init"}
 art <- shallow_alpha$richness_table %>% 
@@
 
+<!-- show supplemental bibliography -->
 <div class="multi-refs"></div>

### 2024-09-19 manuscript.Rmd
updates

@@
 
-# create deep crabs subset
-cc$crabs_deep <- cc$crabs_untransformed %>%
- subset_samples(shallow_deep == "deep")
- # prune deep crabs taxa
-cc$crabs_deep <- prune_taxa(taxa_sums(cc$crabs_deep) > 0,cc$crabs_deep)
-# create shallow crabs subset
-cc$crabs_shallow <- cc$crabs_untransformed %>%
- subset_samples(shallow_deep == "shallow")
- # prune its taxa
-cc$crabs_shallow <- prune_taxa(taxa_sums(cc$crabs_shallow) > 0,cc$crabs_shallow)
-
 # get average distance from a centroid
@@
 all_alpha <- alpha_diversity(cc$crabs_untransformed)
-deep_alpha <- alpha_diversity(cc$crabs_deep)
-shallow_alpha <- alpha_diversity(cc$crabs_shallow)
+deep_alpha <- alpha_diversity(cc$crabs_deep_untransformed)
+shallow_alpha <- alpha_diversity(cc$crabs_shallow_untransformed)
 
@@
 all_summary <- sample_summary(cc$crabs_untransformed,4,crab_taxonomy_validated)
-deep_summary <- sample_summary(cc$crabs_deep,4,crab_taxonomy_validated)
+deep_summary <- sample_summary(cc$crabs_deep_untransformed,4,crab_taxonomy_validated)
 ```
@@
 
-A total of 89 ARMS units were deployed over the course of six years (2009--2016) in shallow coral reef habitat (`r format_range(cc$crabs_shallow %>% sample_tibble() %>% pull(depth))` m depth) in both the Main (MHI) and Northwest Hawaiian Islands (NWHI) ([@fig:fig_map; @tbl:arms_table]).
+A total of 89 ARMS units were deployed over the course of six years (2009--2016) in shallow coral reef habitat (`r format_range(cc$crabs_shallow_untransformed %>% sample_tibble() %>% pull(depth))` m depth) in both the Main (MHI) and Northwest Hawaiian Islands (NWHI) ([@fig:fig_map; @tbl:arms_table]).
 Only units deployed on outer reefs were included in this study.
@@
 Shallow units were deployed according to NOAA-RAMP methodology, whereas units $\geq$ 30 m were deployed and recovered using technical mixed gas closed-circuit rebreather diving [see @hurleyAssessmentShallowMesophotic2016].
-Unless referring to specific depth zones, we will refer to any units deployed \< 30 m as "shallow" and those deployed $\geq$ 30 m as "deep".
+Except when discussing specific depth zones, we will refer to any units deployed \< 30 m as "shallow" and those deployed $\geq$ 30 m as "deep".
 
@@
 *Perinia tumida*, @servisCharacterizingCoralReef2020; @plaisanceReefassociatedCrustaceanFauna2009].
-However, the prevalence of complexes and cryptic diversity suggests that our assignments are likely to under- rather than overrepresent the true number of taxa.
+However, the prevalence of species complexes and cryptic diversity suggests that our assignments are likely to under- rather than overrepresent the true number of taxa.
 Thus, any significant patterns detected likely persist in spite of rather than because of cryptic diversity.
@@
 
-The `r nsamples(cc$crabs_deep)` units deployed at mesophotic depths $\geq$ 30 m on O&#x02BB;ahu yielded fewer species and individuals compared to the complete dataset, with `r num(sum(sample_sums(cc$crabs_deep)),1)` crabs overall, `r sum(unlist(deep_summary$types_breakdown))` distinct taxa (`r deep_summary$types_breakdown$species` species-level identifications, `r deep_summary$types_breakdown$genus` genus, `r deep_summary$types_breakdown$family` family), `r num(deep_summary$abundance_mean,0.1)` $\pm$ `r num(deep_summary$abundance_sd,0.1)` individuals per unit, and abundance ranging from `r format_range(deep_summary$abundance_range)`.
+The `r nsamples(cc$crabs_deep_untransformed)` units deployed at mesophotic depths $\geq$ 30 m on O&#x02BB;ahu yielded fewer species and individuals compared to the complete dataset, with `r num(sum(sample_sums(cc$crabs_deep_untransformed)),1)` crabs overall, `r sum(unlist(deep_summary$types_breakdown))` distinct taxa (`r deep_summary$types_breakdown$species` species-level identifications, `r deep_summary$types_breakdown$genus` genus, `r deep_summary$types_breakdown$family` family), `r num(deep_summary$abundance_mean,0.1)` $\pm$ `r num(deep_summary$abundance_sd,0.1)` individuals per unit, and abundance ranging from `r format_range(deep_summary$abundance_range)`.
 The four most abundant taxa in the deep ARMS did not overlap with the most abundant taxa from the complete dataset: `r with(deep_summary,top_list(top,top_n,top_units))`.
-There were `r deep_summary$singleton_count` singletons, which proportionally accounted for ~`r pc(deep_summary$singleton_count/ntaxa(cc$crabs_deep),0.1)` of species found on deep ARMS.
+There were `r deep_summary$singleton_count` singletons, which proportionally accounted for ~`r pc(deep_summary$singleton_count/ntaxa(cc$crabs_deep_untransformed),0.1)` of species found on deep ARMS.
 
@@
 Richness was not significantly different between shallow and deep samples (`r k_test(all_alpha$k_r_depth,plain=TRUE)`; [@fig:fig_alpha[]{}b]) but effective Simpson diversity was (`r k_test(all_alpha$k_s_depth,plain=TRUE)`; [@fig:fig_alpha[]{}d]).
+When comparing diversity across sampling depths for only sites in the MHI, richness was not significantly different (`r k_test(all_alpha$k_r_depth_main,plain=TRUE)`; [@fig:fig_alpha_main[]{}a]) but effective Simposn diversity was (`r k_test(all_alpha$k_s_depth_main,plain=TRUE)`; [@fig:fig_alpha_main[]{}b]).
 When only shallow arms \< 30 m were considered, there were no significant differences in alpha diversity between the MHI and NWHI for either index (richness: `r k_test(shallow_alpha$k_r_island,plain=TRUE)`, [@fig:fig_alpha2[]{}a]; effective Simpson: `r k_test(shallow_alpha$k_s_island,plain=TRUE)`, [@fig:fig_alpha2[]{}b]).
@@
 # set random seed for consistency
-set.seed(99917)
+set.seed(32767)
 
@@
 
-# dd <- cc$crab_data_shallow %>%
-#   # select(-c(chl,sst,human_impact)) %>%
-#   rename(chl=X4km_chl,sst=X4km_sst,human_impact=cumulative_impacts)
-
 # calculate full dbRDA model
@@
     chl + sst + slope + coral_cover +
-    # X4km_chl + X4km_sst + slope + coral_cover +
-    # chl_4km_terra + sst_4km_terra + slope + coral_cover +
-    # chl_9km_terra + sst_9km_terra + slope + coral_cover +
     larval_immigration + human_impact +
-    # larval_immigration + cumulative_impacts +
     lat + lon +
@@
 # reduce model to terms with VIF <= 3
-db_modl <- reduce_model(db_modl,cc$crab_otus_shallow,cc$crab_data_shallow,vif_limit = 4)
-# db_modl <- reduce_model(db_modl,cc$crab_otus_shallow,dd,vif_limit = 3)
+db_modl <- reduce_model(db_modl,cc$crab_otus_shallow,cc$crab_data_shallow,vif_limit = 3)
 
@@
 
+\newpage
+
 ## Figures and tables
@@
 
+```{r fig-alpha-main, fig.cap='Box plots of species richness and effective Simpson diversity per depth zone for Main Hawaiian Islands only.
+Asterisks and notches as above.
+(a) Species richness by sampling depth.
+(b) Effective Simpson diversity by sampling depth.', results="asis", dev="svg", fig.width=8, fig.height=4, dependson="init"}
+# depth comparison for main islands only
+art <- all_alpha$richness_table %>% 
+  filter(island_group == "main") %>%
+  mutate(island_group = str_to_sentence(island_group)) %>%
+  mutate(shallow_deep = str_to_sentence(shallow_deep),shallow_deep = fct_reorder(shallow_deep,depth))
+
+if (all_alpha$k_r_depth$p.value < 0.05) {
+  which_max <- art %>% 
+    group_by(shallow_deep) %>% 
+    summarise(r = mean(Observed)) %>%
+    slice(which.max(r)) %>%
+    pull(shallow_deep) %>%
+    as.character()
+  max_r <- max(art$Observed) + (diff(range(art$Observed)))/10
+  sig <- tribble(
+    ~x,~y,
+    which_max,max_r
+  ) 
+} else {
+  sig <- tribble(~x,~y)
+}
+
+pointsize <- 5
+
+p1 <- ggplot(art) + 
+  geom_boxplot(aes(x=shallow_deep,y=Observed,fill=shallow_deep),notch=TRUE) +
+  geom_point(aes(x,y),data=sig,shape=8,size=pointsize,stroke=1.5) + 
+  scale_fill_manual(values=sample_depth_pal) +
+  guides(fill="none") + 
+  plotz_theme()  + 
+  xlab("Sampling depth") + 
+  ylab("Richness")
+
+if (all_alpha$k_s_depth$p.value < 0.05) {
+  which_max <- art %>% 
+    group_by(shallow_deep) %>% 
+    summarise(r = mean(Simpson)) %>%
+    slice(which.max(r)) %>%
+    pull(shallow_deep) %>%
+    as.character()
+  max_r <- max(art$Simpson) + (diff(range(art$Simpson)))/10
+  sig <- tribble(
+    ~x,~y,
+    which_max,max_r
+  ) 
+} else {
+  sig <- tribble(~x,~y)
+}
+
+p2 <- ggplot(art) + 
+  geom_boxplot(aes(x=shallow_deep,y=Simpson,fill=shallow_deep),notch=TRUE) +
+  geom_point(aes(x,y),data=sig,shape=8,size=pointsize,stroke=1.5) + 
+  scale_fill_manual(values=sample_depth_pal) +
+  guides(fill="none") + 
+  plotz_theme()  + 
+  xlab("Sampling depth") + 
+  ylab("Effective Simpson diversity")
+
+(p1 + p2) & 
+  plot_annotation(tag_levels = list(str_glue("({letters[1:2]})"))) & 
+  theme(plot.tag = element_text(face="italic"))
+cat("{#fig:fig_alpha_main}")
+```
+
+\newpage
+
 ```{r fig-alpha2, fig.cap='Box plots of (a) species richness and (b) Effective Simpson diversity by island group for shallow samples only.
@@
 
-islands <- cc$crabs_shallow %>%
+islands <- cc$crabs_shallow_untransformed %>%
   sample_tibble() %>%
@@
 
+\newpage
+
 ## References

### 2024-09-19 crabs.R

@@ -286,9 +286,39 @@ setup_crabs <- function() {
   
   cc$crabs <- cc$crabs_untransformed %>%
     ps_standardize("hellinger")
-    # this is the shallow subset
-  cc$crabs_shallow <- cc$crabs %>%
-    subset_samples(region != "mce") 
+  
+  # shallow subset
+  cc$crabs_shallow <- prune_samples(
+    sample_tibble(cc$crabs)$shallow_deep == "shallow",
+    cc$crabs
+  )
+  # prune taxa
+  cc$crabs_shallow <- prune_taxa(taxa_sums(cc$crabs_shallow) > 0,cc$crabs_shallow)
+  
+  # deep subset
+  cc$crabs_deep <- prune_samples(
+    sample_tibble(cc$crabs)$shallow_deep == "deep",
+    cc$crabs
+  )
+  # prune taxa
+  cc$crabs_deep <- prune_taxa(taxa_sums(cc$crabs_deep) > 0,cc$crabs_deep)
+  
+  # untransformed deep subset
+  cc$crabs_deep_untransformed <- prune_samples(
+    sample_tibble(cc$crabs_untransformed)$shallow_deep == "deep",
+    cc$crabs_untransformed
+  ) 
+  # prune deep crabs taxa
+  cc$crabs_deep_untransformed <- prune_taxa(taxa_sums(cc$crabs_deep_untransformed) > 0,cc$crabs_deep_untransformed)
+  
+  # untransformed shallow subset
+  cc$crabs_shallow_untransformed <- prune_samples(
+    sample_tibble(cc$crabs_untransformed)$shallow_deep == "shallow",
+    cc$crabs_untransformed
+  ) 
+  # prune shallow crabs taxa
+  cc$crabs_shallow_untransformed <- prune_taxa(taxa_sums(cc$crabs_shallow_untransformed) > 0,cc$crabs_shallow_untransformed)
+  
   
   # save an unscaled version of the shallow crab metadata
   cc$crab_data_shallow_unscaled <- cc$crabs_shallow %>%
@@ -486,6 +516,8 @@ alpha_diversity <- function(ps,measures=c("Observed","InvSimpson")) {
   try(ad$k_s_island <- kruskal.test(Simpson ~ island_group, data=all_richness),silent=TRUE)
   try(ad$k_r_depth <- kruskal.test(Observed ~ shallow_deep, data=all_richness),silent=TRUE)
[…]

### 2024-09-21 manuscript.Rmd
more updates

@@
 # format t-test results
-t_test <- function(test,plain=FALSE) {
+t_test <- function(test,plain=FALSE,paren=FALSE) {
   if (plain) {
@@
   } else {
-    str_glue("(two-sample t-test: *t* = {num(test$statistic,0.01)}, DF = {num(test$parameter['df'],0.01)}, {pval(test$p.value)})")
+    str_glue("{ifelse(paren,'(','')}{test$method}: *t* = {num(test$statistic,0.01)}, DF = {num(test$parameter['df'],0.01)}, {pval(test$p.value)}{ifelse(paren,')','')}")
+  }
+}
+
+# format f-test results
+f_test <- function(test,plain=FALSE,paren=FALSE) {
+  if (plain) {
+    str_glue("*F* = {num(test$statistic,0.01)}, DF = $\\frac{{{test$parameter[1]}}}{{{test$parameter[2]}}}$, {pval(test$p.value)}")
+  } else {
+    str_glue("{ifelse(paren,'(','')}F-test: *F* = {num(test$statistic,0.01)}, DF = $\\frac{{{test$parameter[1]}}}{{{test$parameter[2]}}}$, {pval(test$p.value)}{ifelse(paren,')','')}")
   }
@@
     if (is.null(fig)) {
-      str_glue("({.x} individuals on {.y} units)")
-    } else {
-      str_glue("({.x} individuals on {.y} units; [@fig:{fig}[]{{{{}}}}{{.y}}])")
+      str_glue("({.x} individuals from {.y} units)")
+    } else {
+      str_glue("({.x} individuals from {.y} units; [@fig:{fig}[]{{{{}}}}{{.y}}])")
     }
@@
 # get average distance from a centroid
-centroid_dist <- function(data) {
+centroid_dist <- function(data,mean=TRUE) {
   pts <- t(as.matrix(data))
@@
   }))
+  if (mean) {
     return(mean(dists))
+  } else {
+    return(c(dists))
+  }
 }
@@
 Here, we used autonomous reef monitoring structures (ARMS)---standardized collection devices designed to mimic reef structural complexity---to sample cryptic communities while controlling for habitat [@zimmermanArtificialReefMatrix2004; @brainardAutonomousReefMonitoring2009].
+Such standardization can reduce variability in recovered communities that may arise due to local variation in habitat composition and availability [(cite?)]{.mark}.
 We targeted brachyuran crabs, a group comprising a notable proportion of the coral reef cryptofauna [284 species in Hawai&#x02BB;i, @castroCatalogAnomuranBrachyuran2011], recovered from these structures to examine drivers of community composition across the Hawaiian Archipelago.
@@
 
-A total of 89 ARMS units were deployed over the course of six years (2009--2016) in shallow coral reef habitat (`r format_range(cc$crabs_shallow_untransformed %>% sample_tibble() %>% pull(depth))` m depth) in both the Main (MHI) and Northwest Hawaiian Islands (NWHI) ([@fig:fig_map; @tbl:arms_table]).
+A total of 89 ARMS units were deployed over the course of six years (2009--2016) in shallow (\< 30 m) coral reef habitat (`r format_range(cc$crabs_shallow_untransformed %>% sample_tibble() %>% pull(depth))` m depth) in both the Main (MHI) and Northwest Hawaiian Islands (NWHI) ([@fig:fig_map; @tbl:arms_table]).
 Only units deployed on outer reefs were included in this study.
@@
   mutate(across(-island,~replace_na(.x,0)))
-print_table(arms_schedule,"arms_table","Recovered ARMS unit soak times across sites (number of units)",accuracy=1) 
+print_table(arms_schedule,"arms_table","ARMS unit deployment schedule by island (number of units)",accuracy=1) 
 ```
@@
 
-We used distance-based redundancy analysis (dbRDA; a form of multivariate multiple regression) with `dbrda` to examine the relative contribution of environmental variables ([@tbl:env_table]) to crab community variation across the islands, conditioning the recovery year as a random effect.
+We used distance-based redundancy analysis (dbRDA; a form of multivariate multiple regression) with `dbrda` to examine the relative contribution of environmental variables ([@tbl:env_table]) to shallow crab community variation across the islands, conditioning the recovery year as a random effect.
 From a model incorporating all potential drivers, we sequentially removed variables with the largest variance inflation factors (VIF; calculated using `vif.cca`) until all values were <3, indicating variables were sufficiently uncorrelated.
@@
 Continuous variables were scaled to unit variance.
+dbRDA analyses were conducted only for communities sampled from shallow (\< 30 m) ARMS units, since detailed driver variables were not available for deep ($\geq$ 30 m) samples.
 
@@
 Significance was assessed with permutation tests in both cases.
-Variation among sampling groups (island group and sampling depth) was further examined and visualized using canonical analysis of principal coordinates (CAP) based on discriminant analysis [@andersonCanonicalAnalysisPrincipal2003] as implemented in the `CAPdiscrim` function in `BiodiversityR` [@kindtTreeDiversityAnalysis2005].
+Variation among island groups and sampling depths was further examined and visualized using canonical analysis of principal coordinates (CAP) based on discriminant analysis [@andersonCanonicalAnalysisPrincipal2003] as implemented in the `CAPdiscrim` function in `BiodiversityR` [@kindtTreeDiversityAnalysis2005].
+To explore differences in community variability among groups (shallow/deep, O&#x02BB;ahu/all shallow samples), we compared Euclidean distances among samples in the CAP ordination using Welch two-sample t-tests, assuming unequal variances.
 
@@
 The cumulative human impacts index, a dimensionless numerical value representing spatially-explicit cumulative anthropogenic impacts on various marine ecosystems, is based on a synthesis of a combination of "high resolution, annual data on the intensity of 14 human stressors and their impact on 21 marine ecosystems over 11 years (2003--2013)" [@halpernGlobalMapHuman2008; @halpernRecentPaceChange2019].
-The index of potential larval immigration was imported from [@selkoeDNACoralReef2016], and represents an "estimate of in-closeness centrality" from a biophysical model simulating larval dispersal in the Hawaiian Islands [@selkoeDNACoralReef2016 supp.
-mat.] For details regarding data import, preparation, and synthesis, see Supplemental material.
+The index of potential larval immigration was imported from [@selkoeDNACoralReef2016], and represents the potential of a given island to act as a destination for larval settlement.
+It is an "estimate of in-closeness centrality" from a biophysical model simulating larval dispersal in the Hawaiian Islands [@selkoeDNACoralReef2016 supp.
+mat.].
+For details regarding data import, preparation, and synthesis, see Supplemental material.
 Due to logistical limitations, environmental data were collected only for shallow samples (\< 30 m).
@@
 When only shallow arms \< 30 m were considered, there were no significant differences in alpha diversity between the MHI and NWHI for either index (richness: `r k_test(shallow_alpha$k_r_island,plain=TRUE)`, [@fig:fig_alpha2[]{}a]; effective Simpson: `r k_test(shallow_alpha$k_s_island,plain=TRUE)`, [@fig:fig_alpha2[]{}b]).
-
-<!--For the complete dataset, there was no significant difference in observed species richness between the MHI and NWHI `r w_test(all_alpha$w_r_island)`, but effective Simpson diversity was significantly different ([@fig:fig_alpha[]{}a]; `r w_test(all_alpha$w_s_island,plain=TRUE)`).
-When only shallow arms \< 30 m were considered, neither comparison was significant (richness: `r w_test(shallow_alpha$w_r_island,plain=TRUE)`; effective Simpson: `r w_test(shallow_alpha$w_s_island,plain=TRUE)`).
-Effective Simpson diversity was significantly different between shallow and deep samples ([@fig:fig_alpha[]{}b]; `r w_test(all_alpha$w_s_depth,plain=TRUE)`) but richness was not (`r w_test(all_alpha$w_r_depth,plain=TRUE)`).
--->
-
-<!--For the complete dataset, there was no significant difference in observed species richness between the MHI and NWHI `r t_test(all_alpha$t_r_island)`, but effective Simpson diversity was significantly different ([@fig:fig_alpha[]{}a]; `r t_test(all_alpha$t_s_island,plain=TRUE)`).
-When only shallow arms \< 30 m were considered, neither comparison was significant (richness: `r t_test(shallow_alpha$t_r_island,plain=TRUE)`; effective Simpson: `r t_test(shallow_alpha$t_s_island,plain=TRUE)`).
-Effective Simpson diversity was significantly different between shallow and deep samples ([@fig:fig_alpha[]{}b]; `r t_test(all_alpha$t_s_depth,plain=TRUE)`) but richness was not (`r t_test(all_alpha$t_r_depth,plain=TRUE)`).
--->
 
@@
   column_to_rownames("sample") %>%
-  centroid_dist()
+  centroid_dist(mean=FALSE)
 
@@
   column_to_rownames("sample") %>%
-  centroid_dist()
+  centroid_dist(mean=FALSE)
+
+centroid_t <- t.test(deep_cd,shallow_cd,alternative = "g",var.equal = FALSE)
+
+oahu_dist <- pcap %>%
+  filter(island == "Oahu") %>%
+  select(sample,LD1,LD2) %>%
+  column_to_rownames("sample") %>%
+  dist() %>% 
+  c()
 
@@
   column_to_rownames("sample") %>%
-  dist()
+  dist() %>% 
+  c()
 
@@
   column_to_rownames("sample") %>%
-  dist()
+  dist() %>%
+  c()
+
+distance_t <- t.test(deep_dist,shallow_dist,alternative = "g",var.equal = FALSE)
+oahu_t <- t.test(oahu_dist,shallow_dist,alternative = "g",var.equal = FALSE)
 ```
@@
 
-In the CAP analysis of island group and depth zone, ARMS units were correctly reassigned to groups `r pc(cap_depth$percent/100,0.1)` of the time.
-The first two axes explained `r pc(cap_var$x,0.1)` and `r pc(cap_var$y,0.1)` of the variation, respectively.
-Mesophotic samples (30--90 m) from O&#x02BB;ahu grouped separately from shallow samples collected across the entire ~2,500 km expanse of the archipelago ([@fig:fig_cap]).
-Additionally, variability among 30--90 m samples was greater than among all shallow samples, (max.
-Euclidean distance between shallow samples in the CAP ordination: `r num(max(shallow_dist),0.01)`, avg.
-`r num(mean(shallow_dist),0.01)`, mean distance to centroid among shallow samples: `r num(mean(shallow_cd,0.01))`; max.
-Euclidean distance between deep samples: `r num(max(deep_dist),0.01)`, avg.
-`r num(mean(deep_dist),0.01)`, mean distance to centroid among deep samples: `r num(mean(deep_cd,0.01))`) ([@fig:fig_cap]).
-
-```{r fig-cap, fig.cap="CAP plot of crab communities by island group and depth zone.", results="asis", fig.width=9, fig.height=6, dependson="cap-analysis"}
+In the CAP analysis of island group and depth zone, ARMS units were correctly reassigned to sample groups `r pc(cap_depth$percent/100,0.1)` of the time.
+The first two linear discriminant axes explained `r pc(cap_var$x,0.1)` and `r pc(cap_var$y,0.1)` of the variation, respectively.
+Mesophotic samples (30--90 m) from O&#x02BB;ahu grouped separately from the shallow samples collected across the entire ~2,500 km expanse of the archipelago ([@fig:fig_cap]), whereas shallow samples from O&#x02BB;ahu primarily clustered with shallow samples from other islands ([@fig:fig_cap]).
+Variability among 30--90 m samples was apparently greater than among all shallow samples, with mean Euclidean distances between deep samples in the CAP ordination significantly greater than between shallow samples (`r t_test(distance_t)`; [@fig:fig_cap; @fig:fig_euclid]).
+Similarly, variability among samples from O&#x02BB;ahu (including mesophotic depths) exceeded variability among all shallow samples combined (based on Euclidean distance among CAP points, as above; `r t_test(oahu_t,plain=TRUE)`).
+
+```{r fig-cap, fig.cap="CAP ordination of crab communities by island group and depth zone.", results="asis", fig.width=9, fig.height=6, dependson="cap-analysis"}
 centroids <- pcap %>% group_by(sample_group) %>%
   summarise(LD1=mean(LD1),LD2=mean(LD2))
+pcap <- pcap %>%
+  mutate(
+    oahu = factor(
+      if_else(island == "Oahu","oahu","noahu"),
+      levels = c("oahu","noahu")
+    )
+  )
 
@@
   stat_ellipse(geom="polygon",level=0.95,alpha=0.1,show.legend = FALSE) +
-  geom_point(color="black",shape=21,size=5)  +
+  # geom_point(aes(shape=oahu),color="black",shape=21,size=5)  +
+  geom_point(mapping=aes(shape=oahu),color="black",size=5)  +
   geom_label_repel(data=centroids,aes(label=sample_group), color="white", show.legend = FALSE) + 
+  scale_shape_manual(values=c("oahu" = 23, "noahu" = 21),name="Island", labels=c("noahu"="Other islands","oahu"="O‘ahu")) +
   scale_fill_manual(values=sample_depth_pal,name="Sample group") +
+  guides( 
+    fill=guide_legend(override.aes=list(shape=21)), 
+    shape=guide_legend(override.aes=list(fill="black"))
+  ) +
   xlab(str_glue("Linear discriminant 1 [{pc(cap_var$x,0.1)}]")) +
@@
   plotz_theme()
+
+# ggplot(pcap,aes(x=LD1,y=LD2,fill=sample_group)) + 
+#   stat_ellipse(geom="polygon",level=0.95,alpha=0.1,show.legend = FALSE) +
+#   geom_point(color="black",shape=21,size=5)  +
+#   geom_label_repel(data=centroids,aes(label=sample_group), color="white", show.legend = FALSE) + 
+#   scale_fill_manual(values=sample_depth_pal,name="Sample group") +
+#   xlab(str_glue("Linear discriminant 1 [{pc(cap_var$x,0.1)}]")) +
+#   ylab(str_glue("Linear discriminant 2 [{pc(cap_var$y,0.1)}]")) +
+#   plotz_theme()
 cat("{#fig:fig_cap}")
@@
 Our best knowledge of the role of depth in structuring marine communities comes from the intertidal, and points to biotic factors (e.g., predation, competition) determining lower limits and abiotic factors (e.g., desiccation, thermal stress) setting upper limits of vertical species distributions [@connellEffectsCompetitionPredation1961; @connellInfluenceInterspecificCompetition1961; @dennyPhysicalProcessesThat2001; @paineIntertidalCommunityStructure1974].
+There are numerous examples from other study systems in which the relative contributions of biotic versus abiotic factors to structuring communities run in opposing directions across some habitat gradient [e.g., climate, soil type, elevation, etc.
+@bradyEvolutionaryEcologyPlant2005; @moore10PlantCompetition2011; @wishingradTemperateZoneIsolation2023].
 However, abiotic stressors on intertidal communities are obviously different from those on coral reefs, and generalizations about the mechanisms and ecological processes that structure coral reef communities remain a subject of considerable debate [@bellwoodEnvironmentalGeometricConstraints2005; @bellwoodSleepingFunctionalGroup2006; @carrBiodiversityPopulationRegulation2002; @connollyIndoPacificBiodiversityCoral2003; @dornelasCoralReefDiversity2006; @grahamImportanceStructuralComplexity2013; @komyakovaRelativeImportanceCoral2013].
@@
 Human activities are considerably higher in the populated MHI than in the protected and remote NWHI [@selkoeEvaluatingAnthropogenicThreats2008; @selkoeMapHumanImpacts2009; @selkoeDNACoralReef2016].
-Although we did not consider the effects of individual drivers here, an index representing cumulative anthropogenic impacts was the top driver of brachyuran community structure, similar to findings in previous work examining genetic diversity, species richness and habitat area, quality and stability across the archipelago [@selkoeDNACoralReef2016].
+Although we did not consider the effects of individual impacts here, an index representing *cumulative* anthropogenic impacts was among the top drivers of brachyuran community structure, similar to findings in previous work examining genetic diversity as it relates to species richness and habitat area, quality, and stability across the archipelago [@selkoeDNACoralReef2016].
 The next-most significant driver, chlorophyll-A, can be used as a proxy for primary productivity.
@@
 Thus, it is not surprising that we find the same top drivers of community structure for one of the primary taxonomic groups inhabiting coral reefs, the brachyuran crabs, across the Hawaiian archipelago.
-What is surprising---given the narrow depth range within which shallow ARMS units were deployed---is the consistent significance of depth as a primary driver of community composition across these gradients.
+What is surprising---given the maximum vertical separation among shallow ARMS units of `r diff(range(cc$crab_data_shallow_unscaled$depth))` m---is the consistent significance of depth as a driver of community composition across these gradients.
 
@@
 Depth in and of itself is unlikely to affect marine community composition.
-At the same time, oceanographic factors with well-known depth profiles (e.g., temperature, light, oxygen, nutrients, etc.) are not expected to vary dramatically across such a narrow depth range (\< 10 m), although fine-scale habitat variability among sites may also be driving the community differences we found [@steyaertRemoteReefCryptobenthic2022].
-Unfortunately, we lack environmental data at depth for each ARMS site, and thus cannot unravel the co-varying factor(s) associated with the depth effects we observed.
-Regardless, given its significance across both narrow (\< 10 m) and wider (90 m) depth ranges sampled here, depth (or an associated factor) is clearly an important driver for cryptobenthic crab communities across the Hawaiian Archipelago.
-Given brachyurans are a ubiquitous and important component of coral reef communities [@reynsPatternsProcessesBrachyuran1999; @lasleyjrPhylogeneticRelationshipsUbiquitous2015; @hazeriLatitudinalSpeciesDiversity2019], these findings are likely to apply more broadly as well.
+At the same time, oceanographic factors with well-known depth profiles (e.g., temperature, light, oxygen, nutrients, etc.) are not expected to vary dramatically across such a narrow depth range (\< 10 m), although fine-scale habitat variability among sites may also influence the community differences we found [@steyaertRemoteReefCryptobenthic2022].
+Unfortunately, we lack environmental data at depth for individual deployments, and thus cannot unravel the co-varying factor(s) associated with the depth effects we observed.
+Regardless, given its significance across both narrow (\< 10 m) and wider (90 m) depth ranges sampled here, a factor associated with depth is clearly an important driver for cryptobenthic crab communities across the Hawaiian Archipelago.
+As brachyurans are a ubiquitous and important component of coral reef communities [@reynsPatternsProcessesBrachyuran1999; @lasleyjrPhylogeneticRelationshipsUbiquitous2015; @hazeriLatitudinalSpeciesDiversity2019], these findings are likely to apply more broadly as well.
 
@@
 Due to logistical constraints (and the limitations of satellite-derived data products), our environmental data only covers shallow-water deployments, but CAP ordinations show considerable differences between the grouping of all shallow communities and the units deployed $\geq$ 30 m at the single island of O&#x02BB;ahu ([@fig:fig_cap]).
-In other words, there was more community variation across 90 m of depth on a single island than across 2,500 km of shallow habitat experiencing multiple gradients of environmental, ecological, and anthropogenic influence.
-The ordination revealed two visible clusters for mesophotic units: one containing the 30 m units and the other containing the 60 m and 90 m units.
+According to our results, there was more community variation among samples spanning 90 m of depth on a single island than across 2,500 km of shallow habitat experiencing multiple gradients of environmental, ecological, and anthropogenic influence.
+The ordination revealed two distinct clusters for mesophotic samples: one comprising the 30 m samples and the other the 60 m and 90 m samples.
 These two clusters coincide with what is regarded as the upper and lower mesophotic zones, respectively [@kahngCommunityEcologyMesophotic2010; @rochaMesophoticCoralEcosystems2018; but see @pyleMesophoticCoralEcosystems2019].
-The upper mesophotic (~30--60 m) has often been considered as a continuation of shallow reefs while the lower mesophotic (~60--150 m) is primarily thought to constitute a unique community.
+The upper mesophotic (~30--60 m) has often been considered an extension of shallow reefs while the lower mesophotic (~60--150 m) is primarily thought to constitute a unique community.
 This pattern has held true across various taxonomic groups [@lesserGlobalCommunityBreaks2019; @rochaMesophoticCoralEcosystems2018], but our data show more distinct separations among shallow, upper, and lower mesophotic depths (although 30 m appears to represent an upper-to-lower transition zone).
 
-In addition to community composition, mesophotic samples also contributed to alpha diversity differences across the archipelago.
-We found significant differences in effective Simpson diversity between the MHI and NWHI, but only when mesophotic samples were included in the comparison; when deep ARMS units were left out, these comparisons were non-significant.
-When comparing shallow to mesophotic, effective Simpson diversity was significantly (if slightly) higher in deep communities.
-Thus, the apparent overall differences in alpha diversity between the MHI and NWHI can be attributed entirely to the deep samples from the island of O&#x02BB;ahu.
-This pattern is likely driven by species unique to mesophotic habitat [@hurleyAssessmentShallowMesophotic2016].
+In addition to community composition, mesophotic samples also exhibited alpha diversity differences.
+Effective Simpson diversity was significantly (if slightly) higher in deep communities (for comparisons both among all islands and solely within the MHI) ([@fig:fig_alpha; @fig:fig_alpha_main]).
+This pattern is in contrast to observations that biodiversity often decreases with depth [@pyleComprehensiveInvestigationMesophotic2016; @kahngCommunityEcologyMesophotic2010], although it may be influenced by differences in habitat availability across depth zones (especially of the habitat types ARMS are designed to mimic).
 
@@
 Since we lack environmental data for mesophotic samples, we cannot make conclusions about the specific influences driving the differences we found.
-However, light and temperature have been proposed to be important bottom-up drivers of depth zonation in benthic communities [@kahngCommunityEcologyMesophotic2010; @lesserGlobalCommunityBreaks2019; @pyleComprehensiveInvestigationMesophotic2016; @pyleMesophoticCoralEcosystems2019].
+However, light and temperature have been proposed as important bottom-up drivers of depth zonation in benthic communities [@kahngCommunityEcologyMesophotic2010; @lesserGlobalCommunityBreaks2019; @pyleComprehensiveInvestigationMesophotic2016; @pyleMesophoticCoralEcosystems2019].
 Oceanographic conditions may be inconsistent close to shore, but the average mixed-layer depth calculated (using 0.8 &deg;C $\upDelta$T [@karaOptimalDefinitionOcean2000]) from CTD casts taken at station ALOHA (22.75&deg;N, 158.00&deg;W; within the latitudinal expanse of the MHI) was 84.6 $\pm$ 26 m over the course of the year (based on monthly averages from casts taken across the duration of ARMS unit deployments [@HawaiiOceanTimeseries]).
 Thus, given the assumption of similar temperature profiles near the O&#x02BB;ahu shoreline, temperature alone appears insufficient to explain the zonation patterns we observed.
-In contrast, the lower depth boundaries of mesophotic reef communities are largely defined by surface irradiance [@kahngCommunityEcologyMesophotic2010; @hindersteinThemeSectionMesophotic2010], so light seems a possible candidate as an abiotic factor influencing community structure.
-
-Further work to investigate drivers of depth zonation could take advantage of established gradients and use the Hawaiian Archipelago as a natural laboratory.
+In contrast, the lower depth boundaries of mesophotic reef communities are largely defined by surface irradiance [@kahngCommunityEcologyMesophotic2010; @hindersteinThemeSectionMesophotic2010], so light seems a possible candidate as bottom-up abiotic factor influencing community structure.
+
+Further work to investigate drivers of depth zonation could take advantage of established gradients and use the Hawaiian Archipelago as a natural laboratory, by sampling depth transects across the archipelago.
+The logistical complexity and time investment required for ARMS deployments make this challenging, but emerging approaches such as environmental DNA (eDNA) may offer a suitable alternative.
 Coral reef biodiversity is generally thought to be structured at depth by a combination of the attenuation of light, a decrease in temperature, increasing salinity, and a reduction in trophic resources [@lesserEcologyMesophoticCoral2009; @kahngCommunityEcologyMesophotic2010; @slatteryFunctionStabilityMesophotic2024a].
-If patterns of community structure are driven primarily by light, that may explain why even our shallow-water sites show a significant depth effect over less than 10m.
+Crab communities are unlikely to be directly affected by light variability.
+However, habitat-building species such as corals, algae, and sponges may [@kahngCommunityEcologyMesophotic2010; @slatteryFunctionStabilityMesophotic2024a], and may in turn affect the structure of the communities dependent upon them.
+If patterns of community structure are driven primarily from the bottom up by light, that may explain our observation of a significant depth effect over less than 10 m in our shallow samples.
 The importance of light could be tested by sampling communities from the same irradiance level but different depths due to factors altering light penetration to determine if we see consistent patterns of community zonation.
-In contrast, other factors (such as dissolved inorganic nitrogen, particulate organic carbon, or productivity) expected to have more strongly-defined gradients across the regions [@slatteryFunctionStabilityMesophotic2024a], would lead to variable zonation patterns that should be detected in our comparison of the MHI & NWHI.
+In contrast, other factors (such as temperature, dissolved inorganic nitrogen, particulate organic carbon, or productivity) expected to have more strongly-defined gradients across the regions [@slatteryFunctionStabilityMesophotic2024a], would lead to variable zonation patterns that should be detected in our comparison of the MHI & NWHI.
 Such differences should be tested in future studies to understand the specific physical and ecological drivers of coral reef depth zonation.
-
-Regardless of the exact environmental and ecological drivers at play, our findings suggest that the classical model of intertidal zonation with biotic and abiotic drivers could likely be reversed for coral reef ecosystems.
-*Lower* limits of mesophotic coral reef ecosystems are likely determined by abiotic factors that correlate with depth, whereas *upper* limits may be driven by ecological interactions.
+Regardless of the exact environmental and ecological drivers at play, it may be possible that the gradients driving depth distributions on coral reefs run opposite to those in the intertidal.
+Given the rapid changes in important factors like light, temperature, and nutrients (among others) throughout the upper ~100--150 m of the water column, the *lower* limits of the coral reef mesophotic zone may be determined by abiotic factors that correlate with depth, whereas *upper* limits may be driven by ecological interactions.
+
 By comparing shallow-water ARMS across the latitudinal, physical and anthropogenic gradients of the Hawaiian Archipelago, we were able to test the relative importance of the significant drivers of community structure.
 Our analysis corroborates previous work and shows depth is the strongest driver of community structure.
-We find that a 90 m depth transect at a single site shows greater divergence in canonical space than the full 2,500 km gradient of the Hawaiian Archipelago for shallow-water ARMS.
+We find that a 90 m depth transect at a single island shows greater divergence in canonical space than the full 2,500 km gradient of the Hawaiian Archipelago for shallow-water ARMS.
 We conclude that some factor(s) that covary with depth exert more influence on the cryptobenthic coral reef community structure than those biotic and abiotic gradients that span the entire Hawaiian Archipelago in shallow-water environments.
@@
 Hee, for discussion, support, and their efforts in the disassembly and sorting of ARMS samples on retrieval.
+Finally, we are grateful to the two anonymous reviewers for constructive comments and suggestions on an earlier version of this paper.
 This is contribution #xxx from the Hawai&#x02BB;i Institute of Marine Biology and #xxx from the School of Ocean and Earth Science and Technology at the University of Hawai&#x02BB;i.
@@
 cat("{#fig:fig_pairwise}")
+```
+
+\newpage
+
+```{r fig-euclid, fig.cap='Box plot of Euclidean distances between points in a CAP ordination by depth zone', results="asis", dev="svg", fig.width=6, fig.height=4, dependson="cap-analysis"}
+dists <- bind_rows(
+  tibble(dist=shallow_dist,depth="Shallow"),
+  tibble(dist=deep_dist,depth="Deep")
+) %>%
+  mutate(depth = factor(depth,levels=c("Shallow","Deep")))
+
+ggplot(dists,aes(x=depth,y=dist,fill=depth)) + 
+  geom_boxplot(notch=TRUE) + 
+  scale_fill_manual(values=sample_depth_pal) + 
+  guides(fill="none") + 
+  plotz_theme() + 
+  xlab("Sampling depth") + 
+  ylab("Euclidean distance between CAP points")
+cat("{#fig:fig_euclid}")
 ```

### 2024-09-23 manuscript.Rmd
lil typo

@@
 Richness was not significantly different between shallow and deep samples (`r k_test(all_alpha$k_r_depth,plain=TRUE)`; [@fig:fig_alpha[]{}b]) but effective Simpson diversity was (`r k_test(all_alpha$k_s_depth,plain=TRUE)`; [@fig:fig_alpha[]{}d]).
-When comparing diversity across sampling depths for only sites in the MHI, richness was not significantly different (`r k_test(all_alpha$k_r_depth_main,plain=TRUE)`; [@fig:fig_alpha_main[]{}a]) but effective Simposn diversity was (`r k_test(all_alpha$k_s_depth_main,plain=TRUE)`; [@fig:fig_alpha_main[]{}b]).
+When comparing diversity across sampling depths for only sites in the MHI, richness was not significantly different (`r k_test(all_alpha$k_r_depth_main,plain=TRUE)`; [@fig:fig_alpha_main[]{}a]) but effective Simpson diversity was (`r k_test(all_alpha$k_s_depth_main,plain=TRUE)`; [@fig:fig_alpha_main[]{}b]).
 When only shallow arms \< 30 m were considered, there were no significant differences in alpha diversity between the MHI and NWHI for either index (richness: `r k_test(shallow_alpha$k_r_island,plain=TRUE)`, [@fig:fig_alpha2[]{}a]; effective Simpson: `r k_test(shallow_alpha$k_s_island,plain=TRUE)`, [@fig:fig_alpha2[]{}b]).

### 2024-10-16 manuscript.Rmd
more revision updates

@@
 
+# quickie to just give us a depth range
+dr <- function(ps) {
+  ps %>%
+    sample_tibble() %>%
+    pull(depth) %>%
+    format_range()
+}
+
 # basic reusable ggplot theme
@@
 Logistical challenges have additionally limited the taxonomic resolution and depth ranges investigated, with the focus largely on fishes and corals in the shallowest ~30 m of coral reefs.
-Here, we examine the effects of a suite of environmental, ecological, and anthropogenic factors on the assemblage structure of brachyuran crabs recovered from Autonomous Reef Monitoring Structures on shallow reefs (`r format_range(range(cc$crab_data_shallow_unscaled$depth))` m) across the ~2,500 km span of the Hawaiian Archipelago.
+Here, we examine the effects of a suite of environmental, ecological, and anthropogenic factors on the assemblage structure of brachyuran crabs recovered from Autonomous Reef Monitoring Structures on shallow reefs (`r dr(cc$crabs_shallow_untransformed)` m) across the ~2,500 km span of the Hawaiian Archipelago.
 We compare these assemblages with those sampled along a depth gradient (`r format_range(range(cc$crab_data$depth[cc$crab_data$island == "Oahu"]))` m) on O&#x02BB;ahu.
@@
 Here, we used autonomous reef monitoring structures (ARMS)---standardized collection devices designed to mimic reef structural complexity---to sample cryptic communities while controlling for habitat [@zimmermanArtificialReefMatrix2004; @brainardAutonomousReefMonitoring2009].
-Such standardization can reduce variability in recovered communities that may arise due to local variation in habitat composition and availability [(cite?)]{.mark}.
-We targeted brachyuran crabs, a group comprising a notable proportion of the coral reef cryptofauna [284 species in Hawai&#x02BB;i, @castroCatalogAnomuranBrachyuran2011], recovered from these structures to examine drivers of community composition across the Hawaiian Archipelago.
-We compared crab communities among islands and regions (MHI vs.
-NWHI) and examined the effects of a suite of possible environmental and anthropogenic drivers of community structure to test hypotheses about their relative importance.
+Such standardization can reduce variability in recovered communities that may arise due to local variation in habitat composition and availability [@ransomeImportanceStandardizationBiodiversity2017].
+We targeted brachyuran crabs, a ubiquitous group often used as an exemplar to assess coral reef biodiversity [@knowltonCoralReefBiodiversity2010; @plaisanceDiversityCoralReefs2011; @servisCharacterizingCoralReef2020; @plaisanceReefassociatedCrustaceanFauna2009] and comprising a notable proportion of the coral reef cryptofauna [284 species in Hawai&#x02BB;i, @castroCatalogAnomuranBrachyuran2011].
+We used assemblages recovered from ARMS units to examine drivers of community composition across the Hawaiian Archipelago, compared crab communities among islands and regions (MHI vs.
+NWHI), and examined the effects of a suite of possible environmental and anthropogenic drivers of community structure to test hypotheses about their relative importance.
 
@@
 
-A total of 89 ARMS units were deployed over the course of six years (2009--2016) in shallow (\< 30 m) coral reef habitat (`r format_range(cc$crabs_shallow_untransformed %>% sample_tibble() %>% pull(depth))` m depth) in both the Main (MHI) and Northwest Hawaiian Islands (NWHI) ([@fig:fig_map; @tbl:arms_table]).
+A total of 89 ARMS units were deployed over the course of six years (2009--2016) in shallow (`r dr(cc$crabs_shallow_untransformed)` m depth) coral reef habitat in both the Main (MHI) and Northwest Hawaiian Islands (NWHI) ([@fig:fig_map; @tbl:arms_table]).
 Only units deployed on outer reefs were included in this study.
 To explore the effects of depth, we also present data from an earlier study by Hurley et al.
-[-@hurleyAssessmentShallowMesophotic2016], in which 27 units were deployed off O&#x02BB;ahu from 2009--2012 in `r format_range(cc$crab_data %>% sample_tibble() %>% filter(island == "Oahu") %>% pull(depth))` m depth; nine units were deployed at 12 m, and six each at 30 m, 60 m, and 90 m ([@tbl:arms_table]).
+[-@hurleyAssessmentShallowMesophotic2016], in which 27 units were deployed off O&#x02BB;ahu from 2009--2012 in `r format_range(cc$crab_data %>% sample_tibble() %>% filter(island == "Oahu") %>% pull(depth))` m depth; nine shallow units were deployed at 12 m, and six each at depths of 30 m, 60 m, and 90 m ([@tbl:arms_table]).
 Shallow units were deployed according to NOAA-RAMP methodology, whereas units $\geq$ 30 m were deployed and recovered using technical mixed gas closed-circuit rebreather diving [see @hurleyAssessmentShallowMesophotic2016].
-Except when discussing specific depth zones, we will refer to any units deployed \< 30 m as "shallow" and those deployed $\geq$ 30 m as "deep".
+Except when discussing specific depth zones, we will refer to any units deployed at `r dr(cc$crabs_shallow_untransformed)` m as "shallow" and those deployed $\geq$ 30 m as "deep".
 
@@
 The cumulative human impacts index, a dimensionless numerical value representing spatially-explicit cumulative anthropogenic impacts on various marine ecosystems, is based on a synthesis of a combination of "high resolution, annual data on the intensity of 14 human stressors and their impact on 21 marine ecosystems over 11 years (2003--2013)" [@halpernGlobalMapHuman2008; @halpernRecentPaceChange2019].
-The index of potential larval immigration was imported from [@selkoeDNACoralReef2016], and represents the potential of a given island to act as a destination for larval settlement.
+The index of potential larval immigration was imported from [@selkoeDNACoralReef2016] and represents the potential of a given island to act as a destination for larval settlement.
 It is an "estimate of in-closeness centrality" from a biophysical model simulating larval dispersal in the Hawaiian Islands [@selkoeDNACoralReef2016 supp.
@@
 # Discussion {-}
+The Hawaiian Archipelago is among the most remote oceanic island chains on the planet and encompasses a range of gradients (anthropogenic, biotic, abiotic), making it an ideal natural laboratory to study the impacts of various drivers of species assemblages across space.
+Human activities are considerably higher in the populated MHI than in the protected and remote NWHI [@selkoeEvaluatingAnthropogenicThreats2008; @selkoeMapHumanImpacts2009; @selkoeDNACoralReef2016].
+Although we did not consider the effects of individual impacts here, an index representing *cumulative* anthropogenic impacts was among the top drivers of shallow brachyuran community structure, similar to findings in previous work examining genetic diversity as it relates to species richness and habitat area, quality, and stability across the archipelago [@selkoeDNACoralReef2016].
+The next-most significant driver, chlorophyll-A, can be used as a proxy for primary productivity.
+Productivity gradients across the Hawaiian Archipelago coincide with differences in coral reef community composition [@selkoeDNACoralReef2016], endemic reef fish dominance [@friedlanderDominanceEndemicsReef2020], survival rates of top predators [@bakerEffectVariableOceanic2007], and island geomorphology [@goveNearislandBiologicalHotspots2016].
+Thus, it is not surprising that we find the same top drivers of community structure for one of the primary taxonomic groups inhabiting coral reefs, the brachyuran crabs, across the Hawaiian archipelago.
+What is surprising---given the maximum vertical separation among shallow ARMS units of `r diff(range(cc$crab_data_shallow_unscaled$depth))` m---is the consistent significance of depth as a driver of community composition across these gradients.
+
+Even though our shallow-water ARMS units were deployed within a narrow depth range (`r dr(cc$crabs_shallow_untransformed)` m), depth still emerged as a significant and consistent driver of community variation among these units.
+Depth in and of itself is unlikely to affect marine community composition.
+At the same time, oceanographic factors with well-known depth profiles (e.g., temperature, oxygen, nutrients, etc.) are not expected to vary dramatically across such a narrow depth range (\< 10 m), although fine-scale habitat variability among sites may also influence the community differences we found [@steyaertRemoteReefCryptobenthic2022].
+Unfortunately, we lack environmental data at depth for individual deployments, and thus cannot unravel the co-varying factor(s) associated with the depth effects we observed.
+Regardless, given its significance across both narrow (\< 10 m) and wider (90 m) depth ranges sampled here, a factor associated with depth is clearly an important driver for cryptobenthic crab communities across the Hawaiian Archipelago.
+As brachyurans are a ubiquitous and important component of coral reef communities [@reynsPatternsProcessesBrachyuran1999; @lasleyjrPhylogeneticRelationshipsUbiquitous2015; @hazeriLatitudinalSpeciesDiversity2019], these findings are likely to apply more broadly as well.
+
+While depth was a significant driver of community composition among shallow-water ARMS units ([@fig:fig_dbrda]), its influence was even more striking when comparing shallow to mesophotic samples.
+Due to logistical constraints (and the limitations of satellite-derived data products), our environmental data only covers shallow-water deployments, but CAP ordinations show considerable differences between the grouping of all shallow communities and the units deployed $\geq$ 30 m at the single island of O&#x02BB;ahu ([@fig:fig_cap]).
+According to these analyses, there was more community variation among samples spanning 90 m of depth on a single island than across 2,500 km of shallow habitat experiencing multiple gradients of environmental, ecological, and anthropogenic influence.
+The ordination revealed two distinct clusters for mesophotic samples: one comprising the 30 m samples and the other the 60 m and 90 m samples.
+These two clusters coincide with what is regarded as the upper and lower mesophotic zones, respectively [@kahngCommunityEcologyMesophotic2010; @rochaMesophoticCoralEcosystems2018; but see @pyleMesophoticCoralEcosystems2019].
+The upper mesophotic (~30--60 m) has often been considered an extension of shallow reefs while the lower mesophotic (~60--150 m) is primarily thought to constitute a unique community.
+This pattern has held true across various taxonomic groups [@lesserGlobalCommunityBreaks2019; @rochaMesophoticCoralEcosystems2018], but our data show more distinct separations among shallow, upper, and lower mesophotic depths (although 30 m appears to encompass a transition zone).
+
+In addition to community composition, mesophotic samples also exhibited alpha diversity differences.
+Effective Simpson diversity was significantly higher in deep communities (for comparisons both among all islands and solely within the MHI) ([@fig:fig_alpha; @fig:fig_alpha_main]).
+This pattern is in contrast to observations that biodiversity often decreases with depth [@pyleComprehensiveInvestigationMesophotic2016; @kahngCommunityEcologyMesophotic2010], although the differences we found were slight and our findings may be influenced by differences in habitat availability across depth zones (especially of the habitat types ARMS are designed to mimic).
+
+Our depth transect was limited to O&#x02BB;ahu, but community and alpha diversity differences among depth zones at that single island appear to overwhelm differences observed among shallow-water coral reef habitat across the archipelago.
+Similar patterns were found at a local scale (tens of kilometers) on Hawai&#x02BB;i Island using environmental DNA (eDNA) [@hobanPlumbingDepthsEnvironmental2023].
+In that study, communities were found to have more affinity within depth zones across sites than among depth zones within sites.
+These results and the patterns we observed suggest that depth-associated factors exert stronger ecological influence than those that vary across the latitudinal, physical, and anthropogenic gradients throughout the Hawaiian Archipelago.
+
+Since we lack environmental data for mesophotic samples, we cannot make conclusions about the specific influences driving the differences we found.
+However, light and temperature have been proposed as important bottom-up drivers of depth zonation in benthic communities [@kahngCommunityEcologyMesophotic2010; @lesserGlobalCommunityBreaks2019; @pyleComprehensiveInvestigationMesophotic2016; @pyleMesophoticCoralEcosystems2019].
+Oceanographic conditions may be inconsistent close to shore, but the average mixed-layer depth calculated (using 0.8 &deg;C $\upDelta$T [@karaOptimalDefinitionOcean2000]) from CTD casts taken at station ALOHA (22.75&deg;N, 158.00&deg;W; within the latitudinal expanse of the MHI) was 84.6 $\pm$ 26 m over the course of the year (based on monthly averages from casts taken across the duration of ARMS unit deployments [@HawaiiOceanTimeseries]).
+Thus, given the assumption of similar temperature profiles near the O&#x02BB;ahu shoreline, temperature alone appears insufficient to explain the zonation patterns we observed---although due to its significant role as a driver of shallow community variation, it cannot be ruled out.
+In contrast, the lower depth boundaries of mesophotic ecosystems are largely defined by surface irradiance [@kahngCommunityEcologyMesophotic2010; @hindersteinThemeSectionMesophotic2010], so light seems a possible candidate as bottom-up abiotic factor influencing community structure across depths.
+Crab communities are unlikely to be directly affected by light variability.
+However, habitat-building species such as corals, algae, and sponges may be [@kahngCommunityEcologyMesophotic2010; @slatteryFunctionStabilityMesophotic2024a], and may in turn affect the structure of the communities dependent upon them [@hustonCompetitionCoexistenceEffects1994; @smithRoleHabitatComplexity2014; @sebensHabitatStructureCommunity1991].
+If patterns of community structure are driven primarily from the bottom up by light, that may contribute to our observation of a significant depth effect over less than 10 m in our shallow samples.
+
+Further work to investigate drivers of depth zonation could take advantage of established gradients and use the Hawaiian Archipelago as a natural laboratory, by sampling depth transects across the archipelago.
+Coral reef biodiversity is generally thought to be structured at depth by a combination of the attenuation of light, a decrease in temperature, increasing salinity, and a reduction in trophic resources [@lesserEcologyMesophoticCoral2009; @kahngCommunityEcologyMesophotic2010; @slatteryFunctionStabilityMesophotic2024a].
+If mesophotic communities are found to be different from their shallow counterparts across multiple islands (ideally comparing sites experiencing similar irradiance), this suggests that they are vertically structured by a factor (e.g., light) outside of the numerous natural gradients that exist across the Hawaiian Islands.
+In contrast, other factors (such as temperature, dissolved inorganic nitrogen, particulate organic carbon, or productivity) expected to have more strongly-defined gradients across the regions [@slatteryFunctionStabilityMesophotic2024a], would lead to variable zonation patterns that should be detected in our spatial comparison of the MHI & NWHI.
+Such differences should be tested in future studies to understand the specific physical and ecological drivers of coral reef depth zonation.
+The logistical complexity and time investment required for ARMS deployments make this challenging, but emerging approaches such as environmental DNA (eDNA) may offer a suitable alternative.
+
 Marine communities are structured by a suite of factors, but there remains no consensus for either generalized mechanisms or the relative importance of drivers of community structure for coral reefs.
@@
 However, abiotic stressors on intertidal communities are obviously different from those on coral reefs, and generalizations about the mechanisms and ecological processes that structure coral reef communities remain a subject of considerable debate [@bellwoodEnvironmentalGeometricConstraints2005; @bellwoodSleepingFunctionalGroup2006; @carrBiodiversityPopulationRegulation2002; @connollyIndoPacificBiodiversityCoral2003; @dornelasCoralReefDiversity2006; @grahamImportanceStructuralComplexity2013; @komyakovaRelativeImportanceCoral2013].
-In the Hawaiian archipelago, we found significant differences in the composition of coral reef crab communities between the populated high islands of the MHI and the uninhabited atolls of the NWHI.
+In the Hawaiian archipelago, we found significant differences in the composition of shallow coral reef crab communities between the populated high islands of the MHI and the uninhabited atolls of the NWHI.
 Moreover, these differences were driven by a combination of environmental (sea-surface temperature, chlorophyll-A concentration, depth), biotic (potential larval immigration), and anthropogenic (cumulative human impacts) factors.
-
-The Hawaiian Archipelago is among the most remote oceanic island chains on the planet and encompasses a range of gradients (anthropogenic, biotic, abiotic), making it an ideal natural laboratory to study the impacts of various drivers of species assemblages across space.
-Human activities are considerably higher in the populated MHI than in the protected and remote NWHI [@selkoeEvaluatingAnthropogenicThreats2008; @selkoeMapHumanImpacts2009; @selkoeDNACoralReef2016].
-Although we did not consider the effects of individual impacts here, an index representing *cumulative* anthropogenic impacts was among the top drivers of brachyuran community structure, similar to findings in previous work examining genetic diversity as it relates to species richness and habitat area, quality, and stability across the archipelago [@selkoeDNACoralReef2016].
-The next-most significant driver, chlorophyll-A, can be used as a proxy for primary productivity.
-Productivity gradients across the Hawaiian Archipelago coincide with differences in coral reef community composition [@selkoeDNACoralReef2016], endemic reef fish dominance [@friedlanderDominanceEndemicsReef2020], survival rates of top predators [@bakerEffectVariableOceanic2007], and island geomorphology [@goveNearislandBiologicalHotspots2016].
-Thus, it is not surprising that we find the same top drivers of community structure for one of the primary taxonomic groups inhabiting coral reefs, the brachyuran crabs, across the Hawaiian archipelago.
-What is surprising---given the maximum vertical separation among shallow ARMS units of `r diff(range(cc$crab_data_shallow_unscaled$depth))` m---is the consistent significance of depth as a driver of community composition across these gradients.
-
-Even though our shallow-water ARMS units were deployed within a narrow depth range (8--17 m), depth still emerged as a significant and consistent driver of community variation among these units.
-Depth in and of itself is unlikely to affect marine community composition.
-At the same time, oceanographic factors with well-known depth profiles (e.g., temperature, light, oxygen, nutrients, etc.) are not expected to vary dramatically across such a narrow depth range (\< 10 m), although fine-scale habitat variability among sites may also influence the community differences we found [@steyaertRemoteReefCryptobenthic2022].
-Unfortunately, we lack environmental data at depth for individual deployments, and thus cannot unravel the co-varying factor(s) associated with the depth effects we observed.
-Regardless, given its significance across both narrow (\< 10 m) and wider (90 m) depth ranges sampled here, a factor associated with depth is clearly an important driver for cryptobenthic crab communities across the Hawaiian Archipelago.
-As brachyurans are a ubiquitous and important component of coral reef communities [@reynsPatternsProcessesBrachyuran1999; @lasleyjrPhylogeneticRelationshipsUbiquitous2015; @hazeriLatitudinalSpeciesDiversity2019], these findings are likely to apply more broadly as well.
-
-While depth was a significant driver of community composition among shallow-water ARMS units, its influence was even more apparent when comparing shallow to mesophotic samples.
-Due to logistical constraints (and the limitations of satellite-derived data products), our environmental data only covers shallow-water deployments, but CAP ordinations show considerable differences between the grouping of all shallow communities and the units deployed $\geq$ 30 m at the single island of O&#x02BB;ahu ([@fig:fig_cap]).
-According to our results, there was more community variation among samples spanning 90 m of depth on a single island than across 2,500 km of shallow habitat experiencing multiple gradients of environmental, ecological, and anthropogenic influence.
-The ordination revealed two distinct clusters for mesophotic samples: one comprising the 30 m samples and the other the 60 m and 90 m samples.
-These two clusters coincide with what is regarded as the upper and lower mesophotic zones, respectively [@kahngCommunityEcologyMesophotic2010; @rochaMesophoticCoralEcosystems2018; but see @pyleMesophoticCoralEcosystems2019].
-The upper mesophotic (~30--60 m) has often been considered an extension of shallow reefs while the lower mesophotic (~60--150 m) is primarily thought to constitute a unique community.
-This pattern has held true across various taxonomic groups [@lesserGlobalCommunityBreaks2019; @rochaMesophoticCoralEcosystems2018], but our data show more distinct separations among shallow, upper, and lower mesophotic depths (although 30 m appears to represent an upper-to-lower transition zone).
-
-In addition to community composition, mesophotic samples also exhibited alpha diversity differences.
-Effective Simpson diversity was significantly (if slightly) higher in deep communities (for comparisons both among all islands and solely within the MHI) ([@fig:fig_alpha; @fig:fig_alpha_main]).
-This pattern is in contrast to observations that biodiversity often decreases with depth [@pyleComprehensiveInvestigationMesophotic2016; @kahngCommunityEcologyMesophotic2010], although it may be influenced by differences in habitat availability across depth zones (especially of the habitat types ARMS are designed to mimic).
-
-Our depth transect was limited to O&#x02BB;ahu, but community and alpha diversity differences among depth zones at that single island appear to overwhelm differences observed among shallow-water coral reef habitat across the archipelago.
-Similar patterns were found at a local scale (tens of kilometers) on Hawai&#x02BB;i Island using environmental DNA (eDNA) [@hobanPlumbingDepthsEnvironmental2023].
-In that study, communities were found to have more affinity across sites within depth zones than among depth zones within sites.
-These results and the patterns we observed suggest that depth-associated factors exert stronger ecological influence than those that vary across the latitudinal, physical, and anthropogenic gradients throughout the Hawaiian Archipelago.
-
-Since we lack environmental data for mesophotic samples, we cannot make conclusions about the specific influences driving the differences we found.
-However, light and temperature have been proposed as important bottom-up drivers of depth zonation in benthic communities [@kahngCommunityEcologyMesophotic2010; @lesserGlobalCommunityBreaks2019; @pyleComprehensiveInvestigationMesophotic2016; @pyleMesophoticCoralEcosystems2019].
-Oceanographic conditions may be inconsistent close to shore, but the average mixed-layer depth calculated (using 0.8 &deg;C $\upDelta$T [@karaOptimalDefinitionOcean2000]) from CTD casts taken at station ALOHA (22.75&deg;N, 158.00&deg;W; within the latitudinal expanse of the MHI) was 84.6 $\pm$ 26 m over the course of the year (based on monthly averages from casts taken across the duration of ARMS unit deployments [@HawaiiOceanTimeseries]).
-Thus, given the assumption of similar temperature profiles near the O&#x02BB;ahu shoreline, temperature alone appears insufficient to explain the zonation patterns we observed.
-In contrast, the lower depth boundaries of mesophotic reef communities are largely defined by surface irradiance [@kahngCommunityEcologyMesophotic2010; @hindersteinThemeSectionMesophotic2010], so light seems a possible candidate as bottom-up abiotic factor influencing community structure.
-
-Further work to investigate drivers of depth zonation could take advantage of established gradients and use the Hawaiian Archipelago as a natural laboratory, by sampling depth transects across the archipelago.
-The logistical complexity and time investment required for ARMS deployments make this challenging, but emerging approaches such as environmental DNA (eDNA) may offer a suitable alternative.
-Coral reef biodiversity is generally thought to be structured at depth by a combination of the attenuation of light, a decrease in temperature, increasing salinity, and a reduction in trophic resources [@lesserEcologyMesophoticCoral2009; @kahngCommunityEcologyMesophotic2010; @slatteryFunctionStabilityMesophotic2024a].
-Crab communities are unlikely to be directly affected by light variability.
-However, habitat-building species such as corals, algae, and sponges may [@kahngCommunityEcologyMesophotic2010; @slatteryFunctionStabilityMesophotic2024a], and may in turn affect the structure of the communities dependent upon them.
-If patterns of community structure are driven primarily from the bottom up by light, that may explain our observation of a significant depth effect over less than 10 m in our shallow samples.
-The importance of light could be tested by sampling communities from the same irradiance level but different depths due to factors altering light penetration to determine if we see consistent patterns of community zonation.
-In contrast, other factors (such as temperature, dissolved inorganic nitrogen, particulate organic carbon, or productivity) expected to have more strongly-defined gradients across the regions [@slatteryFunctionStabilityMesophotic2024a], would lead to variable zonation patterns that should be detected in our comparison of the MHI & NWHI.
-Such differences should be tested in future studies to understand the specific physical and ecological drivers of coral reef depth zonation.
-Regardless of the exact environmental and ecological drivers at play, it may be possible that the gradients driving depth distributions on coral reefs run opposite to those in the intertidal.
-Given the rapid changes in important factors like light, temperature, and nutrients (among others) throughout the upper ~100--150 m of the water column, the *lower* limits of the coral reef mesophotic zone may be determined by abiotic factors that correlate with depth, whereas *upper* limits may be driven by ecological interactions.
+Despite the significance of these factors in driving shallow community variation across the archipelago, mesophotic assemblages were found to be distinct ([@fig:fig_cap]).
+Although there were variables we did not consider in our model, this suggests that vertical distributions are influenced by factors other than those driving spatial variation aross the island chain.
+In light of this, and since mesophotic communities are generally less biodiverse than shallow reefs [@kahngCommunityEcologyMesophotic2010; @lesserEcologyMesophoticCoral2009; @pyleComprehensiveInvestigationMesophotic2016], it may be possible that the gradients driving depth distributions on coral reefs run opposite to those in the intertidal.
+Given the rapid changes in important factors like temperature, nutrients, and light throughout the upper ~100--150 m of the water column (particularly the diminishment of photosynthetically active radiation below 100 m), the *lower* limits of the coral reef mesophotic zone may be determined by abiotic factors that correlate with depth.
+Conversely, the *upper* limits of the mesophotic zone (the transition from shallow to mesophotic habitat), which occur at the higher end of the depth-biodiversity gradient [@kahngCommunityEcologyMesophotic2010; @lesserEcologyMesophoticCoral2009] may be driven by ecological interactions [@vellendEffectsDiversityDiversity2008].
 
 By comparing shallow-water ARMS across the latitudinal, physical and anthropogenic gradients of the Hawaiian Archipelago, we were able to test the relative importance of the significant drivers of community structure.
-Our analysis corroborates previous work and shows depth is the strongest driver of community structure.
-We find that a 90 m depth transect at a single island shows greater divergence in canonical space than the full 2,500 km gradient of the Hawaiian Archipelago for shallow-water ARMS.
-We conclude that some factor(s) that covary with depth exert more influence on the cryptobenthic coral reef community structure than those biotic and abiotic gradients that span the entire Hawaiian Archipelago in shallow-water environments.
+We additionally find that a 90 m depth transect at a single island shows greater divergence in canonical space than the full 2,500 km gradient of the Hawaiian Archipelago for shallow-water ARMS.
+Our analyses corroborate previous work and show that depth is among the strongest drivers of community structure.
+We conclude that some factor(s) covarying with depth exert more influence on cryptobenthic coral reef community structure than those biotic and abiotic gradients that span the entire Hawaiian Archipelago in shallow-water environments.
 
@@
 
-```{r fig-euclid, fig.cap='Box plot of Euclidean distances between points in a CAP ordination by depth zone', results="asis", dev="svg", fig.width=6, fig.height=4, dependson="cap-analysis"}
+```{r fig-euclid, fig.cap='Box plot of Euclidean distances between points in CAP ordination by depth zone', results="asis", dev="svg", fig.width=6, fig.height=4, dependson="cap-analysis"}
 dists <- bind_rows(

### 2024-10-16 get-data.R

@@ -50,8 +50,6 @@ arms_sf <- read_csv("data/metadata.csv") %>%
 initial_radius <- 10
 
 # download chl/sst datasets ---------------------------------------------------------------------------------------
-library(raster)
-
 # setup parallel workers
 if (rstudioapi::isAvailable()) {
   plan(multisession,workers=8)
@@ -326,8 +324,8 @@ aa <- arms %>%
   group_by(range) %>%
   group_modify(\(data,grp) {
     # get date range
-    y1 <- min(arms$deployment_year)
-    y2 <- max(arms$recovery_year)
+    y1 <- min(data$deployment_year)
+    y2 <- max(data$recovery_year)
     # if our range is outside of the available range,
     # take the mean across the whole time period we have data for
     if (!all(c(y1,y2) %in% all_years)) {

### 2024-10-16 manuscript.Rmd
fixing biodiversityR problem in original branch

--- 2024-09-23 manuscript.Rmd
@@
   - natgeo: National Geographic Pristine Seas
+abstract:
+  An inability to consistently resolve the environmental and anthropogenic factors that shape coral reefs impacts our understanding of how those factors interact and how coral reef communities may respond to change.
+Logistical challenges have additionally limited the taxonomic resolution and depth ranges investigated, with the focus largely on fishes and corals in the shallowest ~30 m of coral reefs.
+Here, we examine the effects of a suite of environmental, ecological, and anthropogenic factors on the assemblage structure of brachyuran crabs recovered from Autonomous Reef Monitoring Structures on shallow reefs (8--17 m) across the ~2,500 km span of the Hawaiian Archipelago.
+We compare these assemblages with those sampled along a depth gradient (12--90 m) on O&#x02BB;ahu.
+We found 2,284 individuals representing 103 (morpho)species from 22 families.
+Shallow assemblages were significantly different between the Main and Northwest Hawaiian Islands and significantly associated with human impacts, chlorophyll-A, sea-surface temperature, slope, potential larval connectivity, and depth.
+However, we found more variation among mesophotic samples from O&#x02BB;ahu (30--90 m) than among shallow samples across the entire archipelago, suggesting that 90 m of depth is a stronger determinant of brachyuran assemblage structure than the latitudinal, environmental, or anthropogenic gradients across the Hawaiian Islands.
 bibliography: references/citations.json
@@
     pandoc_args:
+      - "--toc-depth=2"
       - "--filter=resources/pandoc-crossref"
-      - "--citeproc"
-      - "--toc-depth=2"
       - "--lua-filter=resources/scholarly-metadata.lua"
       - "--lua-filter=resources/author-info-blocks.lua"
-      - "--lua-filter=resources/abstract-section.lua"
       - "--reference-doc=resources/ref.docx"
-      - "--lua-filter=resources/multirefs.lua"
-    keep_md: true
+    keep_md: false
 tblPrefix: table
@@
   - \usepackage{upgreek}
-multiref_no_duplicates: true
 ---
@@
 library(ggtext)
-
-# can't get BiodiversityR to work on new mac, but that's ok because
-# all we need is the CAPdiscrim function, which miraculously works on its own!
-# retrieved from: https://github.com/cran/BiodiversityR
-source(here("lib","CAPdiscrim.R"))
-
+source("lib/CAPdiscrim.R")
+
+# set random seed for consistency
+set.seed(31337)
 ```
@@
 ```{r init, include=FALSE}
-# source utility functions
 source("crabs.R")
 
-# setup community matrices and data summaries
 if (!exists("crabs_done")) {
@@
 
-# some utility funtions
-
-# format a range of numbers
 format_range <- function(vals) {
@@
 
-# basic reusable ggplot theme
 plotz_theme <- function() {
@@
 
-# abstract method to format species name
 format_species <- function(...) UseMethod("format_species")
 
-# format species in a list
 format_species.list <- function(spp,authority=TRUE,...) {
@@
 
-# format species in a character vector
 format_species.character <- function(species,found,authority,auth=TRUE,...) {
@@
 
-# format species in a data frame
 format_species.data.frame <- function(spp,authority=TRUE,...) {
   spp %>%
-    pmap_chr(~{
+    pwalk(~{
       row <- list(...)
       s <- row[c('species','found','authority','otu')]
-      format_species(s,authority)
+      print(format_species(s,authority))
     })
@@
 
-# format percents and numbers
 pc <- function(x,a=0.01) scales::percent_format(accuracy = a)(x)
@@
 
-# format wilcoxon test results
 w_test <- function(test,plain=FALSE) {
@@
 
-# format kruskal-wallis test results
 k_test <- function(test,plain=FALSE) {
@@
 
-# format t-test results
-t_test <- function(test,plain=FALSE,paren=FALSE) {
+t_test <- function(test,plain=FALSE) {
   if (plain) {
@@
   } else {
-    str_glue("{ifelse(paren,'(','')}{test$method}: *t* = {num(test$statistic,0.01)}, DF = {num(test$parameter['df'],0.01)}, {pval(test$p.value)}{ifelse(paren,')','')}")
-  }
-}
-
-# format f-test results
-f_test <- function(test,plain=FALSE,paren=FALSE) {
-  if (plain) {
-    str_glue("*F* = {num(test$statistic,0.01)}, DF = $\\frac{{{test$parameter[1]}}}{{{test$parameter[2]}}}$, {pval(test$p.value)}")
-  } else {
-    str_glue("{ifelse(paren,'(','')}F-test: *F* = {num(test$statistic,0.01)}, DF = $\\frac{{{test$parameter[1]}}}{{{test$parameter[2]}}}$, {pval(test$p.value)}{ifelse(paren,')','')}")
-  }
-}
-
-# format a p-value nicely
+    str_glue("(two-sample t-test: *t* = {num(test$statistic,0.01)}, DF = {num(test$parameter['df'],0.01)}, {pval(test$p.value)})")
+  }
+}
+
 pval <- function(p) {
@@
 
-# format permdisp test results
 permdisp <- function(bd,accuracy=0.01,perm=999) {
@@
 
-# format permanova test results
 permanova <- function(modl,accuracy=0.01,stat="pseudo-F",by=NULL,sentence=TRUE) {
@@
 
-# format a top-n list of species
 top_list <- function(species,n,units,len=4,fig=NULL) {
@@
     if (is.null(fig)) {
-      str_glue("({.x} individuals from {.y} units)")
-    } else {
-      str_glue("({.x} individuals from {.y} units; [@fig:{fig}[]{{{{}}}}{{.y}}])")
+      str_glue("({.x} individuals on {.y} units)")
+    } else {
+      str_glue("({.x} individuals on {.y} units; [@fig:{fig}[]{{{{}}}}{{.y}}])")
     }
@@
 
-# format a markdown grid table from a data frame or file
 print_table <- function(tbl,id="",caption="",hdr_case="sentence",hide=character(0),accuracy = 0.01,format_numbers=TRUE, cell_width = Inf) {
@@
 
-# get average distance from a centroid
-centroid_dist <- function(data,mean=TRUE) {
-  pts <- t(as.matrix(data))
-  centers <- matrix(apply(pts,1,mean),nrow=2)
-  dists <- sqrt(apply(centers, 2, function(center) {
-    colSums((pts - center)^2)
-  }))
-  if (mean) {
-    return(mean(dists))
-  } else {
-    return(c(dists))
-  }
-}
-
-# get alpha diversity statistics
+cc$crabs_deep <- cc$crabs_untransformed %>% subset_samples(shallow_deep == "deep")
+cc$crabs_deep <- prune_taxa(taxa_sums(cc$crabs_deep) > 0,cc$crabs_deep)
+cc$crabs_shallow <- cc$crabs_untransformed %>% subset_samples(shallow_deep == "shallow")
+cc$crabs_shallow <- prune_taxa(taxa_sums(cc$crabs_shallow) > 0,cc$crabs_shallow)
+
+dir_create(here("cache"))
+if (file_exists(here("cache","all_summary.rds"))) {
+  all_summary <<- read_rds(here("cache","all_summary.rds"))
+} else {
+  all_summary <<- sample_summary(cc$crabs_untransformed,4)
+  write_rds(all_summary,here("cache","all_summary.rds"))
+}
+if (file_exists(here("cache","deep_summary.rds"))) {
+  deep_summary <<- read_rds(here("cache","deep_summary.rds"))
+} else {
+  deep_summary <<- sample_summary(cc$crabs_deep,4)
+  write_rds(deep_summary,here("cache","deep_summary.rds"))
+}
+
 all_alpha <- alpha_diversity(cc$crabs_untransformed)
-deep_alpha <- alpha_diversity(cc$crabs_deep_untransformed)
-shallow_alpha <- alpha_diversity(cc$crabs_shallow_untransformed)
-
-# color palette for islands
+deep_alpha <- alpha_diversity(cc$crabs_deep)
+shallow_alpha <- alpha_diversity(cc$crabs_shallow)
+
 island_pal <- c(
@@
 )
-
-# color palette for island groups
 island_group_pal <- c(
@@
 
-# number of permutations for adonis tests
 PERM <<- 999
-# total number of brachyuran species in Hawai‘i
-hawaiian_crabs <- 284
-# load model terms table from file
+
 model_terms <- read_csv(here("data","tables","environmental_variables.csv"))
 ```
-
-```{r taxonomy-validation, include=FALSE, dependson="init"}
-# validate taxonomy using worms / get authorities
-# worms API call results are cached
-crab_taxonomy <- cc$crabs %>%
-  taxa_tibble() %>%
-  pull(species) %>%
-  # split into subsets of 50 taxa
-  split(ceiling(seq_along(.)/50)) %>%
-  map(\(sp) {
-    # validate taxa using fuzzy matching
-    wm_records_taxamatch(sp) %>%
-      # smash results into a single tibble
-      map2(sp,~.x %>% mutate(query = .y)) %>%
-      list_rbind()
-  }) %>% 
-  list_rbind()
-
-# max validated taxonomy table
-crab_taxonomy_validated <- cc$crabs %>%
-  taxa_tibble() %>%
-  left_join(crab_taxonomy,by=c("species" = "query")) %>%
-  mutate(
-    found = rank == "Species",
-    # substitute valid name if ours isn't
-    species = case_when(
-      valid_name != scientificname ~ valid_name,
-      .default = species
-    ),
-    authority = case_when(
-      valid_name != scientificname ~ valid_authority,
-      .default = authority
-    )
-  ) %>%
-  mutate(display_species = format_species(.))
-```
-
-```{r summaries, include=FALSE, dependson="init"}
-# generate sample summaries. code block is cached 
-# so worms API calls don't happen every time
-all_summary <- sample_summary(cc$crabs_untransformed,4,crab_taxonomy_validated)
-deep_summary <- sample_summary(cc$crabs_deep_untransformed,4,crab_taxonomy_validated)
-```
-
-<!-- 
-the abstract-section.lua filter allows us to put our abstract down here 
-which allows us to include computed values that couldn't easily go in the yaml header
--->
-# Abstract {-}
-
-An inability to consistently resolve the environmental and anthropogenic factors that shape coral reefs impacts our understanding of how those factors interact and how coral reef communities may respond to change.
-Logistical challenges have additionally limited the taxonomic resolution and depth ranges investigated, with the focus largely on fishes and corals in the shallowest ~30 m of coral reefs.
-Here, we examine the effects of a suite of environmental, ecological, and anthropogenic factors on the assemblage structure of brachyuran crabs recovered from Autonomous Reef Monitoring Structures on shallow reefs (`r format_range(range(cc$crab_data_shallow_unscaled$depth))` m) across the ~2,500 km span of the Hawaiian Archipelago.
-We compare these assemblages with those sampled along a depth gradient (`r format_range(range(cc$crab_data$depth[cc$crab_data$island == "Oahu"]))` m) on O&#x02BB;ahu.
-We recovered `r num(sum(sample_sums(cc$crabs_untransformed)),1)` individuals representing `r sum(unlist(all_summary$types_breakdown))` taxa from `r all_summary$taxonomy_breakdown$families` families.
-Shallow assemblages were significantly different between the Main and Northwest Hawaiian Islands and significantly associated with human impacts, chlorophyll-A, sea-surface temperature, slope, potential larval immigration, and depth.
-However, we found more variation among mesophotic samples from O&#x02BB;ahu (30--90 m) than among shallow samples across the entire archipelago, suggesting that 90 m of depth is a stronger determinant of brachyuran assemblage structure than the latitudinal, environmental, or anthropogenic gradients across the Hawaiian Islands.
 
@@
 Here, we used autonomous reef monitoring structures (ARMS)---standardized collection devices designed to mimic reef structural complexity---to sample cryptic communities while controlling for habitat [@zimmermanArtificialReefMatrix2004; @brainardAutonomousReefMonitoring2009].
-Such standardization can reduce variability in recovered communities that may arise due to local variation in habitat composition and availability [(cite?)]{.mark}.
 We targeted brachyuran crabs, a group comprising a notable proportion of the coral reef cryptofauna [284 species in Hawai&#x02BB;i, @castroCatalogAnomuranBrachyuran2011], recovered from these structures to examine drivers of community composition across the Hawaiian Archipelago.
@@
 
-A total of 89 ARMS units were deployed over the course of six years (2009--2016) in shallow (\< 30 m) coral reef habitat (`r format_range(cc$crabs_shallow_untransformed %>% sample_tibble() %>% pull(depth))` m depth) in both the Main (MHI) and Northwest Hawaiian Islands (NWHI) ([@fig:fig_map; @tbl:arms_table]).
-Only units deployed on outer reefs were included in this study.
+A total of 89 ARMS units were deployed over the course of six years (2009--2016) in shallow coral reef habitat (8--17 m depth) in both the Main (MHI) and Northwest Hawaiian Islands (NWHI) ([@fig:fig_map; @tbl:arms_table]).
 To explore the effects of depth, we also present data from an earlier study by Hurley et al.
-[-@hurleyAssessmentShallowMesophotic2016], in which 27 units were deployed off O&#x02BB;ahu from 2009--2012 in `r format_range(cc$crab_data %>% sample_tibble() %>% filter(island == "Oahu") %>% pull(depth))` m depth; nine units were deployed at 12 m, and six each at 30 m, 60 m, and 90 m ([@tbl:arms_table]).
+[-@hurleyAssessmentShallowMesophotic2016], in which 27 units were deployed off O&#x02BB;ahu from 2009--2012 in 12--90 m depth; nine units were deployed at 12 m, and six each at 30 m, 60 m, and 90 m ([@tbl:arms_table]).
 Shallow units were deployed according to NOAA-RAMP methodology, whereas units $\geq$ 30 m were deployed and recovered using technical mixed gas closed-circuit rebreather diving [see @hurleyAssessmentShallowMesophotic2016].
-Except when discussing specific depth zones, we will refer to any units deployed \< 30 m as "shallow" and those deployed $\geq$ 30 m as "deep".
 
@@
 library(rnaturalearth)
-library(rnaturalearthhires)
 
@@
   mutate(across(-island,~replace_na(.x,0)))
-print_table(arms_schedule,"arms_table","ARMS unit deployment schedule by island (number of units)",accuracy=1) 
+print_table(arms_schedule,"arms_table","Recovered ARMS unit soak times across sites (number of units)",accuracy=1) 
 ```
@@
 All adult brachyuran crabs ($\geq$ 2 mm) were photographed, identified, and preserved in 95% ethanol.
-Specimens were initially identified to family or genus using Poore & Ahyong [-@pooreMarineDecapodCrustacea2023] and additionally to species where possible according to expert advice from Scott Godwin (former curator of invertebrate zoology, Bishop Museum) and Gustav Paulay (curator of invertebrate zoology, Florida Museum of Natural History).
-Where species-level identifications were not possible, taxa were collapsed to family or genus (depending on individual identification).
-We recognize that the species-level taxonomy of Indo-Pacific brachyurans remains poorly resolved, particularly for abundant groups like the Chlorodiellinae [@lasleyjrIntegrativeSystematicRevision2014; @lasleyjrPhylogeneticRelationshipsUbiquitous2015], and some "species" are known to comprise complexes [e.g.
-*Perinia tumida*, @servisCharacterizingCoralReef2020; @plaisanceReefassociatedCrustaceanFauna2009].
-However, the prevalence of species complexes and cryptic diversity suggests that our assignments are likely to under- rather than overrepresent the true number of taxa.
-Thus, any significant patterns detected likely persist in spite of rather than because of cryptic diversity.
+Identifications were made to species level where possible, otherwise crabs were classified into morphospecies represented by suites of shared morphological characters according to expert advice and Poore & Ahyong [-@pooreMarineDecapodCrustacea2023].<!-- Specimens were collected under permits #xxx (NWHI) and #xxx (MHI) and accessioned into the invertebrate zoology collections at the Florida Museum of Natural History (catalog numbers xxx--xxx).-->
 
@@
 
-Unless otherwise specified, all analyses were performed using `R` version `r R.version$major`.`r R.version$minor` [@rcoreteamLanguageEnvironmentStatistical2021].
+Unless otherwise specified, all analyses were performed using `R` version 4.1.2 [@rcoreteamLanguageEnvironmentStatistical2021].
 Community matrices were loaded and encapsulated into data objects using the R package `phyloseq` [@mcmurdiePhyloseqPackageReproducible2013] and multivariate community analyses were performed using the package `vegan` [@oksanenVeganCommunityEcology2022] on Bray-Curtis dissimilarities of Hellinger-transformed species abundance matrices [@legendreEcologicallyMeaningfulTransformations2001; @raoReviewCanonicalCoordinates1995].
@@
 
-We used distance-based redundancy analysis (dbRDA; a form of multivariate multiple regression) with `dbrda` to examine the relative contribution of environmental variables ([@tbl:env_table]) to shallow crab community variation across the islands, conditioning the recovery year as a random effect.
+We used distance-based redundancy analysis (dbRDA; a form of multivariate multiple regression) with `dbrda` to examine the relative contribution of environmental variables ([@tbl:env_table]) to crab community variation across the islands, conditioning the recovery year as a random effect.
 From a model incorporating all potential drivers, we sequentially removed variables with the largest variance inflation factors (VIF; calculated using `vif.cca`) until all values were <3, indicating variables were sufficiently uncorrelated.
@@
 Continuous variables were scaled to unit variance.
-dbRDA analyses were conducted only for communities sampled from shallow (\< 30 m) ARMS units, since detailed driver variables were not available for deep ($\geq$ 30 m) samples.
 
@@
 Significance was assessed with permutation tests in both cases.
-Variation among island groups and sampling depths was further examined and visualized using canonical analysis of principal coordinates (CAP) based on discriminant analysis [@andersonCanonicalAnalysisPrincipal2003] as implemented in the `CAPdiscrim` function in `BiodiversityR` [@kindtTreeDiversityAnalysis2005].
-To explore differences in community variability among groups (shallow/deep, O&#x02BB;ahu/all shallow samples), we compared Euclidean distances among samples in the CAP ordination using Welch two-sample t-tests, assuming unequal variances.
+Variation among sampling groups (island group and sampling depth) was further examined and visualized using canonical analysis of principal coordinates (CAP) based on discriminant analysis [@andersonCanonicalAnalysisPrincipal2003] as implemented in the `CAPdiscrim` function in `BiodiversityR` [@kindtTreeDiversityAnalysis2005].
 
@@
 
-Potential drivers of brachyuran crab community structure were assembled from previous studies [@selkoeDNACoralReef2016; @halpernRecentPaceChange2019], NOAA's National Coral Reef Monitoring Program (NCRMP), the Pacific Islands Ocean Observing System (PacIOOS), and MODIS satellite data products compiled by NASA's Physical Oceanography/Ocean Biology Distributed Active Archive Centers [@nasagoddardspaceflightcenterModerateresolutionImagingSpectroradiometer2022; @nasaobpgMODISAquaGlobal2020].
-Variables were either intrinsic (site depth, latitude, longitude, island slope) or derived from remote sensing and/or modeling (chlorophyll-A, sea-surface temperature, coral cover, potential larval immigration, cumulative human impacts) ([@tbl:env_table; @tbl:arms_metadata_table]).
-Sea-surface temperature (SST) and chlorophyll-A values were calculated by averaging monthly means across the duration of the soaking period for each deployment using AQUA MODIS 4 km-resolution gridded datasets.
-The cumulative human impacts index, a dimensionless numerical value representing spatially-explicit cumulative anthropogenic impacts on various marine ecosystems, is based on a synthesis of a combination of "high resolution, annual data on the intensity of 14 human stressors and their impact on 21 marine ecosystems over 11 years (2003--2013)" [@halpernGlobalMapHuman2008; @halpernRecentPaceChange2019].
-The index of potential larval immigration was imported from [@selkoeDNACoralReef2016], and represents the potential of a given island to act as a destination for larval settlement.
-It is an "estimate of in-closeness centrality" from a biophysical model simulating larval dispersal in the Hawaiian Islands [@selkoeDNACoralReef2016 supp.
-mat.].
-For details regarding data import, preparation, and synthesis, see Supplemental material.
+Potential drivers of shallow crab community structure were assembled from a previous study examining environmental effects on genetic diversity [@selkoeDNACoralReef2016], NOAA's National Coral Reef Monitoring Program (NCRMP), the Pacific Islands Ocean Observing System (PacIOOS), and MODIS satellite data products compiled by NASA's Physical Oceanography/Ocean Biology Distributed Active Archive Centers (PO.DAAC, OB.DAAC) ([@tbl:env_table; @tbl:arms_metadata_table]).
+Sea-surface temparature (SST) and chlorophyll (Chl-A) values were calculated by averaging monthly means for the duration of the soaking period at each sampling site.
+For details on satellite data preparation and data sources, see Supplemental material.
 Due to logistical limitations, environmental data were collected only for shallow samples (\< 30 m).
@@
 
-<!-- TODO: update number of different types to reflect family/genus/species level rather than species/morphospecies level -->
 The `r nsamples(cc$crabs_untransformed)` ARMS units deployed across the Hawaiian Islands (in `r format_range(cc$crab_data$depth)` m depth) yielded a total of `r num(sum(sample_sums(cc$crabs_untransformed)),1)` adult crabs.
 Units contained `r num(all_summary$abundance_mean,0.1)` $\pm$ `r num(all_summary$abundance_sd,0.1)` (mean $\pm$ SD) individuals, and abundance ranged from `r format_range(all_summary$abundance_range)`.
-Crabs were identified to `r sum(unlist(all_summary$types_breakdown))` distinct taxa (`r all_summary$types_breakdown$species` species-level identifications, `r all_summary$types_breakdown$genus` genus, `r all_summary$types_breakdown$family` family) encompassing `r all_summary$taxonomy_breakdown$genera` genera in `r all_summary$taxonomy_breakdown$families` families, which accounted for ~`r pc(all_summary$types_breakdown$species / hawaiian_crabs,0.1)` of the known crab diversity reported from Hawaiian waters for species-level IDs [@castroCatalogAnomuranBrachyuran2011].
+Crabs were identified to `r sum(unlist(all_summary$types_breakdown))` distinct types (`r all_summary$types_breakdown$species` species and `r all_summary$types_breakdown$morphospecies` morphospecies) in `r all_summary$taxonomy_breakdown$genera` genera and `r all_summary$taxonomy_breakdown$families` families, which accounted for ~36% of the known crab diversity reported from Hawaiian waters [@castroCatalogAnomuranBrachyuran2011].
 The four most abundant species were: `r with(all_summary,top_list(top,top_n,top_units,fig="fig_crabs"))`.
-There were `r all_summary$singleton_count` singleton taxa (taxa occurring in a single unit across all sites), accounting for ~`r pc(all_summary$singleton_count/ntaxa(cc$crabs_untransformed),0.1)` of total recovered crab taxa.
-For a breakdown of specific crab taxa recovered by island and deployment depth, see [@tbl:species_table].
-
-![Representative photographs of the four most abundant crab species found across all deployed ARMS units (specimens shown were not collected as part of this study).
-(a) *Chlorodiella laevissima* (FLMNH 45490, Kāne&#x02BB;ohe, Hawai&#x02BB;i).
-(b) *Perinia tumida* (FLMNH 45652, Kāne&#x02BB;ohe, Hawai&#x02BB;i).
-(c) *Percnon abbreviatum* (FLMNH 45628, Kāne&#x02BB;ohe, Hawai&#x02BB;i).
-(d) *Dynomene hispida* (FLMNH 69204, Manawai, Hawai&#x02BB;i).
-Photo credits: Florida Museum of Natural History, licensed [CC BY-NC 3.0](https://creativecommons.org/licenses/by-nc/3.0/)](data/images/4crabs.png){#fig:fig_crabs}
+There were `r all_summary$singleton_count` singleton taxa (morpho/species occurring in a single unit across all sites), accounting for ~`r num(all_summary$singleton_count/ntaxa(cc$crabs_untransformed)*100,1)`% of total recovered crab species.
+<!--`r str_to_sentence(english(all_summary$singleton_count))` species occurred in a single unit across all sites (singletons), accounting for ~`r num(all_summary$singleton_count/ntaxa(cc$crabs_untransformed)*100,1)`% of total recovered crab species.-->
+
+![The four most abundant crab species found across all deployed ARMS units.
+(a) *Chlorodiella laevissima*.
+(b) *Perinia tumida*.
+(c) *Percnon abbreviatum*.
+(d) *Dynomene hispida*.
+Photo credit: Florida Museum of Natural History, licensed [CC BY-NC 3.0](https://creativecommons.org/licenses/by-nc/3.0/)](data/images/4crabs.png){#fig:fig_crabs}
 
@@
 
-The `r nsamples(cc$crabs_deep_untransformed)` units deployed at mesophotic depths $\geq$ 30 m on O&#x02BB;ahu yielded fewer species and individuals compared to the complete dataset, with `r num(sum(sample_sums(cc$crabs_deep_untransformed)),1)` crabs overall, `r sum(unlist(deep_summary$types_breakdown))` distinct taxa (`r deep_summary$types_breakdown$species` species-level identifications, `r deep_summary$types_breakdown$genus` genus, `r deep_summary$types_breakdown$family` family), `r num(deep_summary$abundance_mean,0.1)` $\pm$ `r num(deep_summary$abundance_sd,0.1)` individuals per unit, and abundance ranging from `r format_range(deep_summary$abundance_range)`.
-The four most abundant taxa in the deep ARMS did not overlap with the most abundant taxa from the complete dataset: `r with(deep_summary,top_list(top,top_n,top_units))`.
-There were `r deep_summary$singleton_count` singletons, which proportionally accounted for ~`r pc(deep_summary$singleton_count/ntaxa(cc$crabs_deep_untransformed),0.1)` of species found on deep ARMS.
+The `r nsamples(cc$crabs_deep)` units deployed at mesophotic depths $\geq$ 30 m on O&#x02BB;ahu yielded fewer species and individuals compared to the complete dataset, with `r num(sum(sample_sums(cc$crabs_deep)),1)` crabs overall, `r sum(unlist(deep_summary$types_breakdown))` distinct types (`r deep_summary$types_breakdown$species` species and `r deep_summary$types_breakdown$morphospecies` morphospecies), `r num(deep_summary$abundance_mean,0.1)` $\pm$ `r num(deep_summary$abundance_sd,0.1)` individuals per unit, and abundance ranging from `r format_range(deep_summary$abundance_range)`.
+The four most abundant taxa in the deep ARMS did not overlap with those from the complete dataset: `r with(deep_summary,top_list(top,top_n,top_units))`.
+There were `r deep_summary$singleton_count` singletons, which proportionally accounted for ~`r num(deep_summary$singleton_count/ntaxa(cc$crabs_deep)*100,1)`% of species found on deep ARMS.
 
@@
 
-For the complete dataset, there was no significant difference in observed species richness (`r k_test(all_alpha$k_r_island)`; [@fig:fig_alpha[]{}a]) or effective Simpson diversity (`r k_test(all_alpha$k_s_island,plain=TRUE)`; [@fig:fig_alpha[]{}c]) between the MHI and NWHI.
-Richness was not significantly different between shallow and deep samples (`r k_test(all_alpha$k_r_depth,plain=TRUE)`; [@fig:fig_alpha[]{}b]) but effective Simpson diversity was (`r k_test(all_alpha$k_s_depth,plain=TRUE)`; [@fig:fig_alpha[]{}d]).
-When comparing diversity across sampling depths for only sites in the MHI, richness was not significantly different (`r k_test(all_alpha$k_r_depth_main,plain=TRUE)`; [@fig:fig_alpha_main[]{}a]) but effective Simpson diversity was (`r k_test(all_alpha$k_s_depth_main,plain=TRUE)`; [@fig:fig_alpha_main[]{}b]).
+For the complete dataset, there was no significant difference in observed species richness between the MHI and NWHI ([@fig:fig_alpha[]{}a]; `r k_test(all_alpha$k_r_island)`), but effective Simpson diversity was significantly different ([@fig:fig_alpha[]{}c]; `r k_test(all_alpha$k_s_island,plain=TRUE)`).
+Richness was not significantly different between shallow and deep samples ([@fig:fig_alpha[]{}b]; `r k_test(all_alpha$k_r_depth,plain=TRUE)`) but effective Simpson diversity was ([@fig:fig_alpha[]{}d]; `r k_test(all_alpha$k_s_depth,plain=TRUE)`).
 When only shallow arms \< 30 m were considered, there were no significant differences in alpha diversity between the MHI and NWHI for either index (richness: `r k_test(shallow_alpha$k_r_island,plain=TRUE)`, [@fig:fig_alpha2[]{}a]; effective Simpson: `r k_test(shallow_alpha$k_s_island,plain=TRUE)`, [@fig:fig_alpha2[]{}b]).
+
+<!--For the complete dataset, there was no significant difference in observed species richness between the MHI and NWHI `r w_test(all_alpha$w_r_island)`, but effective Simpson diversity was significantly different ([@fig:fig_alpha[]{}a]; `r w_test(all_alpha$w_s_island,plain=TRUE)`).
+When only shallow arms \< 30 m were considered, neither comparison was significant (richness: `r w_test(shallow_alpha$w_r_island,plain=TRUE)`; effective Simpson: `r w_test(shallow_alpha$w_s_island,plain=TRUE)`).
+Effective Simpson diversity was significantly different between shallow and deep samples ([@fig:fig_alpha[]{}b]; `r w_test(all_alpha$w_s_depth,plain=TRUE)`) but richness was not (`r w_test(all_alpha$w_r_depth,plain=TRUE)`).
+-->
+
+<!--For the complete dataset, there was no significant difference in observed species richness between the MHI and NWHI `r t_test(all_alpha$t_r_island)`, but effective Simpson diversity was significantly different ([@fig:fig_alpha[]{}a]; `r t_test(all_alpha$t_s_island,plain=TRUE)`).
+When only shallow arms \< 30 m were considered, neither comparison was significant (richness: `r t_test(shallow_alpha$t_r_island,plain=TRUE)`; effective Simpson: `r t_test(shallow_alpha$t_s_island,plain=TRUE)`).
+Effective Simpson diversity was significantly different between shallow and deep samples ([@fig:fig_alpha[]{}b]; `r t_test(all_alpha$t_s_depth,plain=TRUE)`) but richness was not (`r t_test(all_alpha$t_r_depth,plain=TRUE)`).
+-->
 
@@
 library(formula.tools)
-
-# set random seed for consistency
-set.seed(32767)
 
@@
     chl + sst + slope + coral_cover +
-    larval_immigration + human_impact +
-    lat + lon +
+    larval_connectivity + human_impact +
     Condition(recovery_year),
   data=cc$crab_data_shallow,
-  # data=dd,
   distance="bray"
@@
 
-# reduce model to terms with VIF <= 3
-db_modl <- reduce_model(db_modl,cc$crab_otus_shallow,cc$crab_data_shallow,vif_limit = 3)
+# get rid of variables with VIF > 3
+db_modl <- reduce_model(db_modl,cc$crab_otus_shallow,cc$crab_data_shallow)
 
@@
 # get marginal term information and sort by descending f-statistic
-modl_perm <- anova(db_modl,by="margin",permutations = PERM) %>%
-  # as_tibble(rownames="term") %>%
-  # janitor::clean_names() %>%
+modl_perm <- anova(db_modl,by="margin") %>%
   broom::tidy() %>%
@@
   select(term=Term,stat=`F-statistic`,pval=`p-value`)  %>%
-  mutate(
-    term=str_to_lower(term),
-    term = str_replace_all(term,r'(\b[a-z]\b)',replacement=str_to_upper)
-  )
-
-all_terms <- english::english(nrow(model_terms))
+  mutate(term=str_to_lower(term))
 ```
@@
 The best dbRDA model (`r permanova(db_modl,stat="F")`) explained `r pc(RsquareAdj(db_modl)$adj.r.squared,0.01)` (adjusted $\mathrm{R}^2$) of the variability in crab communities across the archipelago ([@fig:fig_dbrda]).
-`r str_to_sentence(english(nterm))` of the `r str_to_lower(english(nrow(model_terms)))` environmental variables tested were significant and sufficiently uncorrelated.
-The two drivers contributing most to explained variation were `r drivers$term[1]` (F = `r num(drivers$stat[1],0.01)`) and `r drivers$term[2]` (F = `r num(drivers$stat[2],0.01)`) ([@tbl:rda_table]).
-These factors also resulted in the longest effect vectors in the dbRDA plot ([@fig:fig_dbrda]).
+`r str_to_sentence(english(nterm))` of the seven environmental variables tested were significant and sufficiently uncorrelated.<!--TODO: make "chlorophyll-a" look like "chlorophyll-A"--> The two drivers contributing most to explained variation were `r drivers$term[1]` (F = `r num(drivers$stat[1],0.01)`) and `r drivers$term[2]` (F = `r num(drivers$stat[2],0.01)`) ([@tbl:rda_table]), although the two factors with the longest effect vectors on the dBRDA plot were `r vectors$driver[1]` and `r vectors$driver[2]` ([@fig:fig_dbrda]).
 
@@
 
-shallow_cd <- pcap %>%
-  filter(shallow_deep == "shallow") %>%
-  select(sample,LD1,LD2) %>%
-  column_to_rownames("sample") %>%
-  centroid_dist(mean=FALSE)
-
-deep_cd <- pcap %>%
-  filter(shallow_deep == "deep") %>%
-  select(sample,LD1,LD2) %>%
-  column_to_rownames("sample") %>%
-  centroid_dist(mean=FALSE)
-
-centroid_t <- t.test(deep_cd,shallow_cd,alternative = "g",var.equal = FALSE)
-
-oahu_dist <- pcap %>%
-  filter(island == "Oahu") %>%
-  select(sample,LD1,LD2) %>%
-  column_to_rownames("sample") %>%
-  dist() %>% 
-  c()
-
 shallow_dist <- pcap %>%
@@
   column_to_rownames("sample") %>%
-  dist() %>% 
-  c()
+  dist()
 
@@
   column_to_rownames("sample") %>%
-  dist() %>%
-  c()
-
-distance_t <- t.test(deep_dist,shallow_dist,alternative = "g",var.equal = FALSE)
-oahu_t <- t.test(oahu_dist,shallow_dist,alternative = "g",var.equal = FALSE)
+  dist()
 ```
@@
 
-In the CAP analysis of island group and depth zone, ARMS units were correctly reassigned to sample groups `r pc(cap_depth$percent/100,0.1)` of the time.
-The first two linear discriminant axes explained `r pc(cap_var$x,0.1)` and `r pc(cap_var$y,0.1)` of the variation, respectively.
-Mesophotic samples (30--90 m) from O&#x02BB;ahu grouped separately from the shallow samples collected across the entire ~2,500 km expanse of the archipelago ([@fig:fig_cap]), whereas shallow samples from O&#x02BB;ahu primarily clustered with shallow samples from other islands ([@fig:fig_cap]).
-Variability among 30--90 m samples was apparently greater than among all shallow samples, with mean Euclidean distances between deep samples in the CAP ordination significantly greater than between shallow samples (`r t_test(distance_t)`; [@fig:fig_cap; @fig:fig_euclid]).
-Similarly, variability among samples from O&#x02BB;ahu (including mesophotic depths) exceeded variability among all shallow samples combined (based on Euclidean distance among CAP points, as above; `r t_test(oahu_t,plain=TRUE)`).
-
-```{r fig-cap, fig.cap="CAP ordination of crab communities by island group and depth zone.", results="asis", fig.width=9, fig.height=6, dependson="cap-analysis"}
+In the CAP analysis of island group and depth zone, ARMS units were correctly reassigned to groups `r pc(cap_depth$percent/100,0.1)` of the time.
+The first two axes explained `r pc(cap_var$x,0.1)` and `r pc(cap_var$y,0.1)` of the variation, respectively.
+Mesophotic samples (30--90 m) from O&#x02BB;ahu grouped separately from shallow samples collected across the entire ~2,500 km expanse of the archipelago ([@fig:fig_cap]).
+Additionally, variability among 30--90 m samples was greater than among all shallow samples, (max.
+Euclidean distance between shallow samples in the CAP ordination: `r num(max(shallow_dist),0.01)`, avg.
+`r num(mean(shallow_dist),0.01)`; max.
+Euclidean distance between deep samples: `r num(max(deep_dist),0.01)`, avg.
+`r num(mean(deep_dist),0.01)`) ([@fig:fig_cap]).
+
+```{r fig-cap, fig.cap="CAP plot of crab communities by island group and depth zone.", results="asis", fig.width=9, fig.height=6, dependson="cap-analysis"}
 centroids <- pcap %>% group_by(sample_group) %>%
   summarise(LD1=mean(LD1),LD2=mean(LD2))
-pcap <- pcap %>%
-  mutate(
-    oahu = factor(
-      if_else(island == "Oahu","oahu","noahu"),
-      levels = c("oahu","noahu")
-    )
-  )
 
@@
   stat_ellipse(geom="polygon",level=0.95,alpha=0.1,show.legend = FALSE) +
-  # geom_point(aes(shape=oahu),color="black",shape=21,size=5)  +
-  geom_point(mapping=aes(shape=oahu),color="black",size=5)  +
+  geom_point(color="black",shape=21,size=5)  +
   geom_label_repel(data=centroids,aes(label=sample_group), color="white", show.legend = FALSE) + 
-  scale_shape_manual(values=c("oahu" = 23, "noahu" = 21),name="Island", labels=c("noahu"="Other islands","oahu"="O‘ahu")) +
   scale_fill_manual(values=sample_depth_pal,name="Sample group") +
-  guides( 
-    fill=guide_legend(override.aes=list(shape=21)), 
-    shape=guide_legend(override.aes=list(fill="black"))
-  ) +
   xlab(str_glue("Linear discriminant 1 [{pc(cap_var$x,0.1)}]")) +
@@
   plotz_theme()
-
-# ggplot(pcap,aes(x=LD1,y=LD2,fill=sample_group)) + 
-#   stat_ellipse(geom="polygon",level=0.95,alpha=0.1,show.legend = FALSE) +
-#   geom_point(color="black",shape=21,size=5)  +
-#   geom_label_repel(data=centroids,aes(label=sample_group), color="white", show.legend = FALSE) + 
-#   scale_fill_manual(values=sample_depth_pal,name="Sample group") +
-#   xlab(str_glue("Linear discriminant 1 [{pc(cap_var$x,0.1)}]")) +
-#   ylab(str_glue("Linear discriminant 2 [{pc(cap_var$y,0.1)}]")) +
-#   plotz_theme()
 cat("{#fig:fig_cap}")
@@
 Our best knowledge of the role of depth in structuring marine communities comes from the intertidal, and points to biotic factors (e.g., predation, competition) determining lower limits and abiotic factors (e.g., desiccation, thermal stress) setting upper limits of vertical species distributions [@connellEffectsCompetitionPredation1961; @connellInfluenceInterspecificCompetition1961; @dennyPhysicalProcessesThat2001; @paineIntertidalCommunityStructure1974].
-There are numerous examples from other study systems in which the relative contributions of biotic versus abiotic factors to structuring communities run in opposing directions across some habitat gradient [e.g., climate, soil type, elevation, etc.
-@bradyEvolutionaryEcologyPlant2005; @moore10PlantCompetition2011; @wishingradTemperateZoneIsolation2023].
 However, abiotic stressors on intertidal communities are obviously different from those on coral reefs, and generalizations about the mechanisms and ecological processes that structure coral reef communities remain a subject of considerable debate [@bellwoodEnvironmentalGeometricConstraints2005; @bellwoodSleepingFunctionalGroup2006; @carrBiodiversityPopulationRegulation2002; @connollyIndoPacificBiodiversityCoral2003; @dornelasCoralReefDiversity2006; @grahamImportanceStructuralComplexity2013; @komyakovaRelativeImportanceCoral2013].
 In the Hawaiian archipelago, we found significant differences in the composition of coral reef crab communities between the populated high islands of the MHI and the uninhabited atolls of the NWHI.
-Moreover, these differences were driven by a combination of environmental (sea-surface temperature, chlorophyll-A concentration, depth), biotic (potential larval immigration), and anthropogenic (cumulative human impacts) factors.
+Moreover, these differences were driven by a combination of environmental (sea-surface temperature, chlorophyll concentration, depth), biotic (potential larval immigration), and anthropogenic (cumulative human impacts) factors.
 
@@
 Human activities are considerably higher in the populated MHI than in the protected and remote NWHI [@selkoeEvaluatingAnthropogenicThreats2008; @selkoeMapHumanImpacts2009; @selkoeDNACoralReef2016].
-Although we did not consider the effects of individual impacts here, an index representing *cumulative* anthropogenic impacts was among the top drivers of brachyuran community structure, similar to findings in previous work examining genetic diversity as it relates to species richness and habitat area, quality, and stability across the archipelago [@selkoeDNACoralReef2016].
-The next-most significant driver, chlorophyll-A, can be used as a proxy for primary productivity.
+Although we did not consider the effects of individual drivers here, an index representing cumulative anthropogenic impacts was the top driver of brachyuran community structure, similar to findings in previous work examining genetic diversity, species richness and habitat area, quality and stability across the archipelago [@selkoeDNACoralReef2016].
+The next-most significant driver, chlorophyll A, can be used as a proxy for primary productivity.
 Productivity gradients across the Hawaiian Archipelago coincide with differences in coral reef community composition [@selkoeDNACoralReef2016], endemic reef fish dominance [@friedlanderDominanceEndemicsReef2020], survival rates of top predators [@bakerEffectVariableOceanic2007], and island geomorphology [@goveNearislandBiologicalHotspots2016].
 Thus, it is not surprising that we find the same top drivers of community structure for one of the primary taxonomic groups inhabiting coral reefs, the brachyuran crabs, across the Hawaiian archipelago.
-What is surprising---given the maximum vertical separation among shallow ARMS units of `r diff(range(cc$crab_data_shallow_unscaled$depth))` m---is the consistent significance of depth as a driver of community composition across these gradients.
+What is surprising is the consistent significance of depth as a primary driver of community composition across these gradients.
 
@@
 Depth in and of itself is unlikely to affect marine community composition.
-At the same time, oceanographic factors with well-known depth profiles (e.g., temperature, light, oxygen, nutrients, etc.) are not expected to vary dramatically across such a narrow depth range (\< 10 m), although fine-scale habitat variability among sites may also influence the community differences we found [@steyaertRemoteReefCryptobenthic2022].
-Unfortunately, we lack environmental data at depth for individual deployments, and thus cannot unravel the co-varying factor(s) associated with the depth effects we observed.
-Regardless, given its significance across both narrow (\< 10 m) and wider (90 m) depth ranges sampled here, a factor associated with depth is clearly an important driver for cryptobenthic crab communities across the Hawaiian Archipelago.
-As brachyurans are a ubiquitous and important component of coral reef communities [@reynsPatternsProcessesBrachyuran1999; @lasleyjrPhylogeneticRelationshipsUbiquitous2015; @hazeriLatitudinalSpeciesDiversity2019], these findings are likely to apply more broadly as well.
+At the same time, oceanographic factors with well-known depth profiles (e.g., temperature, light, oxygen, nutrients, etc.) are not expected to vary dramatically across such a narrow depth range (\< 10 m), although fine-scale habitat variability among sites may also be driving the community differences we found [@steyaertRemoteReefCryptobenthic2022].
+Unfortunately, we lack environmental data at depth for each ARMS site, and thus cannot unravel the co-varying factor(s) associated with the depth effects we observed.
+Regardless, given its significance across both narrow (\< 10 m) and wider (90 m) depth ranges sampled here, depth (or an associated factor) is clearly an important driver for cryptobenthic crab communities across the Hawaiian Archipelago.
+Given brachyurans are a ubiquitous and important component of coral reef communities [@reynsPatternsProcessesBrachyuran1999; @lasleyjrPhylogeneticRelationshipsUbiquitous2015; @hazeriLatitudinalSpeciesDiversity2019], these findings are likely to apply more broadly as well.
 
@@
 Due to logistical constraints (and the limitations of satellite-derived data products), our environmental data only covers shallow-water deployments, but CAP ordinations show considerable differences between the grouping of all shallow communities and the units deployed $\geq$ 30 m at the single island of O&#x02BB;ahu ([@fig:fig_cap]).
-According to our results, there was more community variation among samples spanning 90 m of depth on a single island than across 2,500 km of shallow habitat experiencing multiple gradients of environmental, ecological, and anthropogenic influence.
-The ordination revealed two distinct clusters for mesophotic samples: one comprising the 30 m samples and the other the 60 m and 90 m samples.
+In other words, there was more community variation across 90 m of depth on a single island than across 2,500 km of shallow habitat experiencing multiple gradients of environmental, ecological, and anthropogenic influence.
+The ordination revealed two visible clusters for mesophotic units: one containing the 30 m units and the other containing the 60 m and 90 m units.
 These two clusters coincide with what is regarded as the upper and lower mesophotic zones, respectively [@kahngCommunityEcologyMesophotic2010; @rochaMesophoticCoralEcosystems2018; but see @pyleMesophoticCoralEcosystems2019].
-The upper mesophotic (~30--60 m) has often been considered an extension of shallow reefs while the lower mesophotic (~60--150 m) is primarily thought to constitute a unique community.
+The upper mesophotic (~30--60 m) has often been considered as a continuation of shallow reefs while the lower mesophotic (~60--150 m) is primarily thought to constitute a unique community.
 This pattern has held true across various taxonomic groups [@lesserGlobalCommunityBreaks2019; @rochaMesophoticCoralEcosystems2018], but our data show more distinct separations among shallow, upper, and lower mesophotic depths (although 30 m appears to represent an upper-to-lower transition zone).
 
-In addition to community composition, mesophotic samples also exhibited alpha diversity differences.
-Effective Simpson diversity was significantly (if slightly) higher in deep communities (for comparisons both among all islands and solely within the MHI) ([@fig:fig_alpha; @fig:fig_alpha_main]).
-This pattern is in contrast to observations that biodiversity often decreases with depth [@pyleComprehensiveInvestigationMesophotic2016; @kahngCommunityEcologyMesophotic2010], although it may be influenced by differences in habitat availability across depth zones (especially of the habitat types ARMS are designed to mimic).
+In addition to community composition, mesophotic samples also contributed to alpha diversity differences across the archipelago.
+We found significant differences in effective Simpson diversity between the MHI and NWHI, but only when mesophotic samples were included in the comparison; when deep ARMS units were left out, these comparisons were non-significant.
+When comparing shallow to mesophotic, effective Simpson diversity was significantly (if slightly) higher in deep communities.
+Thus, the apparent overall differences in alpha diversity between the MHI and NWHI can be attributed entirely to the deep samples from the island of O&#x02BB;ahu.
+This pattern is likely driven by species unique to mesophotic habitat [@hurleyAssessmentShallowMesophotic2016].
 
@@
 Since we lack environmental data for mesophotic samples, we cannot make conclusions about the specific influences driving the differences we found.
-However, light and temperature have been proposed as important bottom-up drivers of depth zonation in benthic communities [@kahngCommunityEcologyMesophotic2010; @lesserGlobalCommunityBreaks2019; @pyleComprehensiveInvestigationMesophotic2016; @pyleMesophoticCoralEcosystems2019].
+However, light and temperature have been proposed to be important with regard to depth zonation in benthic communities [@kahngCommunityEcologyMesophotic2010; @lesserGlobalCommunityBreaks2019; @pyleComprehensiveInvestigationMesophotic2016; @pyleMesophoticCoralEcosystems2019].
 Oceanographic conditions may be inconsistent close to shore, but the average mixed-layer depth calculated (using 0.8 &deg;C $\upDelta$T [@karaOptimalDefinitionOcean2000]) from CTD casts taken at station ALOHA (22.75&deg;N, 158.00&deg;W; within the latitudinal expanse of the MHI) was 84.6 $\pm$ 26 m over the course of the year (based on monthly averages from casts taken across the duration of ARMS unit deployments [@HawaiiOceanTimeseries]).
 Thus, given the assumption of similar temperature profiles near the O&#x02BB;ahu shoreline, temperature alone appears insufficient to explain the zonation patterns we observed.
-In contrast, the lower depth boundaries of mesophotic reef communities are largely defined by surface irradiance [@kahngCommunityEcologyMesophotic2010; @hindersteinThemeSectionMesophotic2010], so light seems a possible candidate as bottom-up abiotic factor influencing community structure.
-
-Further work to investigate drivers of depth zonation could take advantage of established gradients and use the Hawaiian Archipelago as a natural laboratory, by sampling depth transects across the archipelago.
-The logistical complexity and time investment required for ARMS deployments make this challenging, but emerging approaches such as environmental DNA (eDNA) may offer a suitable alternative.
+In contrast, the lower depth boundaries of mesophotic reef communities are largely determined by surface irradiance [@kahngCommunityEcologyMesophotic2010; @hindersteinThemeSectionMesophotic2010], so light seems a possible candidate to be a significant abiotic driver of community structure.
+
+Further work to investigate drivers of depth zonation could take advantage of established gradients and use the Hawaiian Archipelago as a natural laboratory.
 Coral reef biodiversity is generally thought to be structured at depth by a combination of the attenuation of light, a decrease in temperature, increasing salinity, and a reduction in trophic resources [@lesserEcologyMesophoticCoral2009; @kahngCommunityEcologyMesophotic2010; @slatteryFunctionStabilityMesophotic2024a].
-Crab communities are unlikely to be directly affected by light variability.
-However, habitat-building species such as corals, algae, and sponges may [@kahngCommunityEcologyMesophotic2010; @slatteryFunctionStabilityMesophotic2024a], and may in turn affect the structure of the communities dependent upon them.
-If patterns of community structure are driven primarily from the bottom up by light, that may explain our observation of a significant depth effect over less than 10 m in our shallow samples.
+If patterns of community structure are driven primarily by light, that may explain why even our shallow-water sites show a significant depth effect over less than 10m.
 The importance of light could be tested by sampling communities from the same irradiance level but different depths due to factors altering light penetration to determine if we see consistent patterns of community zonation.
-In contrast, other factors (such as temperature, dissolved inorganic nitrogen, particulate organic carbon, or productivity) expected to have more strongly-defined gradients across the regions [@slatteryFunctionStabilityMesophotic2024a], would lead to variable zonation patterns that should be detected in our comparison of the MHI & NWHI.
+In contrast, other factors (such as dissolved inorganic nitrogen, particulate organic carbon, or productivity) expected to have more strongly-defined gradients across the regions [@slatteryFunctionStabilityMesophotic2024a], would lead to variable zonation patterns that should be detected in our comparison of the MHI & NWHI.
 Such differences should be tested in future studies to understand the specific physical and ecological drivers of coral reef depth zonation.
-Regardless of the exact environmental and ecological drivers at play, it may be possible that the gradients driving depth distributions on coral reefs run opposite to those in the intertidal.
-Given the rapid changes in important factors like light, temperature, and nutrients (among others) throughout the upper ~100--150 m of the water column, the *lower* limits of the coral reef mesophotic zone may be determined by abiotic factors that correlate with depth, whereas *upper* limits may be driven by ecological interactions.
-
+
+Regardless of the exact environmental and ecological drivers at play, our findings suggest that the classical model of intertidal zonation with biotic and abiotic drivers could likely be reversed for coral reef ecosystems.
+*Lower* limits of mesophotic coral reef ecosystems are likely determined by abiotic factors that correlate with depth, whereas *upper* limits may be driven by ecological interactions.
 By comparing shallow-water ARMS across the latitudinal, physical and anthropogenic gradients of the Hawaiian Archipelago, we were able to test the relative importance of the significant drivers of community structure.
 Our analysis corroborates previous work and shows depth is the strongest driver of community structure.
-We find that a 90 m depth transect at a single island shows greater divergence in canonical space than the full 2,500 km gradient of the Hawaiian Archipelago for shallow-water ARMS.
+We find that a 90 m depth transect at a single site shows greater divergence in canonical space than the full 2,500 km gradient of the Hawaiian Archipelago for shallow-water ARMS.
 We conclude that some factor(s) that covary with depth exert more influence on the cryptobenthic coral reef community structure than those biotic and abiotic gradients that span the entire Hawaiian Archipelago in shallow-water environments.
@@
 Hee, for discussion, support, and their efforts in the disassembly and sorting of ARMS samples on retrieval.
-Finally, we are grateful to the two anonymous reviewers for constructive comments and suggestions on an earlier version of this paper.
 This is contribution #xxx from the Hawai&#x02BB;i Institute of Marine Biology and #xxx from the School of Ocean and Earth Science and Technology at the University of Hawai&#x02BB;i.
@@
 
-<!-- show main bibliography -->
-<div class="multi-refs"></div>
+::: {#refs}
+:::
 
@@
 
-## Data import and processing
-
-All data import and processing code for environmental variables used in the dbRDA model can be found in the file `get-data.R` from the associated github repository.
-Details for specific datasets are given below.
-
-### Processing of satellite-derived environmental data
-Level-3 mapped (4 km grid-size) sea-surface temperature (SST) and chlorophyll-A (chl-A) data were retrieved from NASA's Physical Oceanography Distributed Active Archive Center [PO.DAAC; @nasaobpgMODISAquaGlobal2020] and Ocean Biology Distributed Active Archive Center [OB.DAAC; @nasagoddardspaceflightcenterModerateresolutionImagingSpectroradiometer2022] respectively.
+## Processing of satellite-derived environmental data
+Level-3 mapped (9km) sea-surface temperature (SST) and chlorophyll A (chl-A) data were retrieved from NASA's Physical Oceanography Distributed Active Archive Center [PO.DAAC; @https://doi.org/10.5067/modsa-mo9d9] and Ocean Biology Distributed Active Archive Center [OB.DAAC; @NASAOceanBiologyProcessingGroup_2022] respectively.
 Monthly summary datasets were downloaded for the period of 2009--2016 and averaged over each of the ARMS unit soaking periods (e.g.
@@
 Values (SST or chl-A) were extracted for the geographic coordinates of each sampling site from the averaged dataset for the relevant soaking period.
-Original raw monthly-average dataset names can be found in @tbl:nasa_table.
-
-### Processing of cumulative human impacts data
-Gridded cumulative human impacts raster data calcuated by Halpern et al [-@halpernRecentPaceChange2019] was downloaded from the Knowledge Network for Biocomplexity (KNB) repository [@frazierRecentPaceChange2019].
-Human impacts datasets were downloaded for each available year in this study (2009--2013) and averaged over ARMS unit soaking periods (as above).
-Since cumulative impacts data was only available through 2013, we used the combined average of all available years for the 2013--2016 soaking period.
-Since the availability of data points at individual coordinates was highly variable, we averaged impact scores for each sampled island in the study.
-
-### Other datasets
-Coral cover and potential larval immigration were imported from the online supplemtal material accompanying Selkoe et al. [-@selkoeDNACoralReef2016].
-
-\newpage
-
-## Figures and tables
+Specific monthly-average dataset names can be found in @tbl:nasa_table.
 
@@
 
-```{r fig-alpha-main, fig.cap='Box plots of species richness and effective Simpson diversity per depth zone for Main Hawaiian Islands only.
-Asterisks and notches as above.
-(a) Species richness by sampling depth.
-(b) Effective Simpson diversity by sampling depth.', results="asis", dev="svg", fig.width=8, fig.height=4, dependson="init"}
-# depth comparison for main islands only
-art <- all_alpha$richness_table %>% 
-  filter(island_group == "main") %>%
-  mutate(island_group = str_to_sentence(island_group)) %>%
-  mutate(shallow_deep = str_to_sentence(shallow_deep),shallow_deep = fct_reorder(shallow_deep,depth))
-
-if (all_alpha$k_r_depth$p.value < 0.05) {
-  which_max <- art %>% 
-    group_by(shallow_deep) %>% 
-    summarise(r = mean(Observed)) %>%
-    slice(which.max(r)) %>%
-    pull(shallow_deep) %>%
-    as.character()
-  max_r <- max(art$Observed) + (diff(range(art$Observed)))/10
-  sig <- tribble(
-    ~x,~y,
-    which_max,max_r
-  ) 
-} else {
-  sig <- tribble(~x,~y)
-}
-
-pointsize <- 5
-
-p1 <- ggplot(art) + 
-  geom_boxplot(aes(x=shallow_deep,y=Observed,fill=shallow_deep),notch=TRUE) +
-  geom_point(aes(x,y),data=sig,shape=8,size=pointsize,stroke=1.5) + 
-  scale_fill_manual(values=sample_depth_pal) +
-  guides(fill="none") + 
-  plotz_theme()  + 
-  xlab("Sampling depth") + 
-  ylab("Richness")
-
-if (all_alpha$k_s_depth$p.value < 0.05) {
-  which_max <- art %>% 
-    group_by(shallow_deep) %>% 
-    summarise(r = mean(Simpson)) %>%
-    slice(which.max(r)) %>%
-    pull(shallow_deep) %>%
-    as.character()
-  max_r <- max(art$Simpson) + (diff(range(art$Simpson)))/10
-  sig <- tribble(
-    ~x,~y,
-    which_max,max_r
-  ) 
-} else {
-  sig <- tribble(~x,~y)
-}
-
-p2 <- ggplot(art) + 
-  geom_boxplot(aes(x=shallow_deep,y=Simpson,fill=shallow_deep),notch=TRUE) +
-  geom_point(aes(x,y),data=sig,shape=8,size=pointsize,stroke=1.5) + 
-  scale_fill_manual(values=sample_depth_pal) +
-  guides(fill="none") + 
-  plotz_theme()  + 
-  xlab("Sampling depth") + 
-  ylab("Effective Simpson diversity")
-
-(p1 + p2) & 
-  plot_annotation(tag_levels = list(str_glue("({letters[1:2]})"))) & 
-  theme(plot.tag = element_text(face="italic"))
-cat("{#fig:fig_alpha_main}")
-```
-
-\newpage
-
 ```{r fig-alpha2, fig.cap='Box plots of (a) species richness and (b) Effective Simpson diversity by island group for shallow samples only.
 Notches as in [@fig:fig_alpha].
-Both comparisons were non-significant.', results="asis", dev="svg", fig.width=8, fig.height=4, dependson="init"}
+Both comparisons are non-significant.', results="asis", dev="svg", fig.width=8, fig.height=4, dependson="init"}
 art <- shallow_alpha$richness_table %>% 
@@
 pcdata <- cds %>%
-  select(where(is.numeric),-c(lon,lat),-closest_island) %>%
-  select(deployment_year,depth,chl,sst,slope:human_impact,-contains('radius'))
+  select(where(is.numeric),-c(lon,lat),-closest_island)
 
@@
   # geom_mark_ellipse(color="black",alpha=0.2,show.legend=FALSE, linewidth=0.07) +
-  scale_fill_manual(values=island_group_pal,name="Island group") + 
+  scale_fill_manual(values=island_group_pal) + 
   geom_segment(data=loadings,aes(x=0,y=0,xend=PC1,yend=PC2),color="black",size=0.5,arrow=arrow(length=unit(0.02,"npc"))) + 
@@
 
-islands <- cc$crabs_shallow_untransformed %>%
+islands <- cc$crabs_shallow %>%
   sample_tibble() %>%
@@
 
-```{r fig-euclid, fig.cap='Box plot of Euclidean distances between points in a CAP ordination by depth zone', results="asis", dev="svg", fig.width=6, fig.height=4, dependson="cap-analysis"}
-dists <- bind_rows(
-  tibble(dist=shallow_dist,depth="Shallow"),
-  tibble(dist=deep_dist,depth="Deep")
-) %>%
-  mutate(depth = factor(depth,levels=c("Shallow","Deep")))
-
-ggplot(dists,aes(x=depth,y=dist,fill=depth)) + 
-  geom_boxplot(notch=TRUE) + 
-  scale_fill_manual(values=sample_depth_pal) + 
-  guides(fill="none") + 
-  plotz_theme() + 
-  xlab("Sampling depth") + 
-  ylab("Euclidean distance between CAP points")
-cat("{#fig:fig_euclid}")
-```
-
-\newpage
-
 ```{r arms-metadata-table, results="asis", dependson="init"}
@@
 
-```{r species-table, results="asis", dependson="taxonomy-validation"}
-# make table of all sampled taxa
-species_tbl <- cc$crabs %>%
-  psmelt() %>%
-  # sum counts by island group, depth zone, and species
-  group_by(OTU,island_group,hawaiian_name,depth_zone,depth,family,genus,species) %>%
-  summarise(n=sum(Abundance)) %>%
-  ungroup() %>%
-  # keep species we actually found
-  filter(n > 0) %>%
-  # figure out if a species is exclusive to a depth zone
-  group_by(species) %>%
-  mutate(zones = str_c(depth_zone,collapse="|")) %>%
-  ungroup() %>%
-  mutate(depth_exclusive = depth_zone == zones) %>%
-  arrange(island_group,hawaiian_name,depth_zone,depth,family,genus,species) %>%
-  # join in validated taxonomy
-  left_join(crab_taxonomy_validated,by=c("OTU" = "otu")) %>%
-  select(hawaiian_name,island_group,depth_zone,depth,family.y,display_species,depth_exclusive) %>%
-  rename(island = hawaiian_name,valid_species=display_species) %>%
-  mutate(
-    across(c(island_group,depth_zone),str_to_sentence),
-    depth_exclusive = if_else(depth_exclusive,"Yes","No")
-  ) %>%
-  rename_with(\(n) {
-    n %>%
-      str_replace_all("depth$","depth (m)") %>%
-      str_replace_all("_"," ") %>%
-      str_replace_all("\\.y$","") %>%
-      str_replace_all("^display.","") %>%
-      str_to_sentence()
-  })
-print_table(species_tbl,"species_table","List of crab taxa detected by island and ARMS unit deployment depth",hdr_case="none",accuracy=1)
-```
-
-\newpage
-
 ```{r rda-table,results="asis",dependson="rda-analysis"}
@@
 ```
-
-\newpage
-
-## References
-
-<!-- show supplemental bibliography -->
-<div class="multi-refs"></div>

### 2024-10-16 lib/CAPdiscrim.R

@@ -1,0 +1,104 @@
+`CAPdiscrim` <-
+function(formula, data, dist="bray", axes=4, m=0, mmax=10,
+    add=FALSE, permutations=0, aitchison_pseudocount=1) {
+#    if (!require(MASS)) {stop("Requires package MASS")}
+    CAPresult=function(points, y, group, axes=axes, m=1, eig) {
+        lda1 <- MASS::lda(y[,group]~points[,1:m], CV=T, tol=1.0e-25)
+        lda2 <- MASS::lda(y[,group]~points[,1:m], tol=1.0e-25)
+        matches <- (lda1$class == y[,group])
+        correct <- sum(matches) / length(matches) * 100
+        lda3 <- predict(lda2, y[, group])
+        rownames(lda3$x) <- rownames(points)
+        tot <- sum(eig)
+        varm <- sum(eig[1:m])/tot*100
+        result <- list(PCoA=points[,1:axes], m=m, tot=tot, varm=varm, group=y[,group], CV=lda1$class, percent=correct,
+            x=lda3$x, F=(lda2$svd)^2, lda.CV=lda1, lda.other=lda2)
+        return(result)
+    }
+    x <- eval(as.name((all.vars(formula)[1])))
+    group <- all.vars(formula)[2]
+    y <- data
+    if (inherits(x, "dist")) {
+        distmatrix <- x
+        x <- data.frame(as.matrix(x))
+    }else{
+        distmatrix <- vegdist(x, method = dist, pseudocount=aitchison_pseudocount)
+    }
+    distmatrix <- as.dist(distmatrix, diag=F, upper=F)
+    pcoa <- cmdscale(distmatrix, k=nrow(x)-1, eig=T, add=add)    
+    points <- pcoa$points
+    mmax <- min(ncol(points), mmax)
+    rownames(points) <- rownames(x)
+    eig <- pcoa$eig
+    if (m==0) {
+        correct <- -1
+        for (i in 1:mmax) {
+            if (eig[i] > 0) {
+                result1 <- CAPresult(points=points, y=y, group=group, axes=axes, m=i, eig=eig)
+                if (result1$percent > correct) {
+                    correct <- result1$percent
+                    result <- result1
+                } 
+            }            
+        }
+    }else{
+        result <- CAPresult(points=points, y=y, group=group, axes=axes, m=m, eig=eig)
+    }
+    if (permutations>0) {
+        permutations <- permutations-1
+        permresult <- numeric(permutations)
[…]

### 2024-10-16 manuscript.Rmd
revision updates

--- 2024-10-16 manuscript.Rmd
@@
 When comparing diversity across sampling depths for only sites in the MHI, richness was not significantly different (`r k_test(all_alpha$k_r_depth_main,plain=TRUE)`; [@fig:fig_alpha_main[]{}a]) but effective Simpson diversity was (`r k_test(all_alpha$k_s_depth_main,plain=TRUE)`; [@fig:fig_alpha_main[]{}b]).
+Neither richness nor Simpson diversity were significantly different between shallow and deep when comparisons were constrained to the island of O&#x02BB;ahu (Richness: `r k_test(all_alpha$k_r_depth_oahu,plain=TRUE)`; Simpson: `r k_test(all_alpha$k_s_depth_oahu,plain=TRUE)`).
 When only shallow arms \< 30 m were considered, there were no significant differences in alpha diversity between the MHI and NWHI for either index (richness: `r k_test(shallow_alpha$k_r_island,plain=TRUE)`, [@fig:fig_alpha2[]{}a]; effective Simpson: `r k_test(shallow_alpha$k_s_island,plain=TRUE)`, [@fig:fig_alpha2[]{}b]).
@@
 The two drivers contributing most to explained variation were `r drivers$term[1]` (F = `r num(drivers$stat[1],0.01)`) and `r drivers$term[2]` (F = `r num(drivers$stat[2],0.01)`) ([@tbl:rda_table]).
-These factors also resulted in the longest effect vectors in the dbRDA plot ([@fig:fig_dbrda]).
+These factors also produced the longest effect vectors in the dbRDA plot ([@fig:fig_dbrda]).
 
@@
 In addition to community composition, mesophotic samples also exhibited alpha diversity differences.
-Effective Simpson diversity was significantly higher in deep communities (for comparisons both among all islands and solely within the MHI) ([@fig:fig_alpha; @fig:fig_alpha_main]).
+Effective Simpson diversity was significantly higher in deep communities (for comparisons among all islands and solely within the MHI, but not exclusively at O&#x02BB;ahu) ([@fig:fig_alpha; @fig:fig_alpha_main]).
 This pattern is in contrast to observations that biodiversity often decreases with depth [@pyleComprehensiveInvestigationMesophotic2016; @kahngCommunityEcologyMesophotic2010], although the differences we found were slight and our findings may be influenced by differences in habitat availability across depth zones (especially of the habitat types ARMS are designed to mimic).
@@
 ### Other datasets
-Coral cover and potential larval immigration were imported from the online supplemtal material accompanying Selkoe et al. [-@selkoeDNACoralReef2016].
+Coral cover and potential larval immigration were imported from the online supplemental material accompanying Selkoe et al. [-@selkoeDNACoralReef2016].
 

### 2024-10-16 crabs.R

@@ -518,6 +518,8 @@ alpha_diversity <- function(ps,measures=c("Observed","InvSimpson")) {
   try(ad$k_s_depth <- kruskal.test(Simpson ~ shallow_deep, data=all_richness),silent=TRUE)
   try(ad$k_r_depth_main <- kruskal.test(Observed ~ shallow_deep, data=all_richness %>% filter(island_group == "main")),silent=TRUE)
   try(ad$k_s_depth_main <- kruskal.test(Simpson ~ shallow_deep, data=all_richness  %>% filter(island_group == "main")),silent=TRUE)
+  try(ad$k_r_depth_oahu <- kruskal.test(Observed ~ shallow_deep, data=all_richness %>% filter(island == "Oahu")),silent=TRUE)
+  try(ad$k_s_depth_oahu <- kruskal.test(Simpson ~ shallow_deep, data=all_richness  %>% filter(island == "Oahu")),silent=TRUE)
   
   try(ad$w_r_island <- wilcox.test(Observed ~ island_group, data=all_richness),silent=TRUE)
   try(ad$w_s_island <- wilcox.test(Simpson ~ island_group, data=all_richness),silent=TRUE)

### 2024-10-18 manuscript.Rmd
tiny tweak
revision updates

@@
 author:
-  - Kaleonani K.C. Hurley:
-      institute:
-        - himb
-      equal_contributor: yes
   - Mykle L. Hoban:
@@
       email: "mhoban@hawaii.edu"
+  - Kaleonani K.C. Hurley:
+      institute:
+        - himb
+      equal_contributor: yes
   - Kerry Reardon:
@@
 
+# do an english comma-separated list
+format_list <- function(l) {
+  l[length(l)] <- str_c("and ",l[length(l)])
+  if (length(l) > 2) { 
+    str_c(l,collapse=", ")
+  } else {
+    str_c(l,collapse=" ")
+  }
+}
+
 # format a range of numbers
@@
 Significance was assessed with permutation tests in both cases.
-Variation among island groups and sampling depths was further examined and visualized using canonical analysis of principal coordinates (CAP) based on discriminant analysis [@andersonCanonicalAnalysisPrincipal2003] as implemented in the `CAPdiscrim` function in `BiodiversityR` [@kindtTreeDiversityAnalysis2005].
+We used PERMANOVA to make the following comparisons: island group (MHI vs NWHI, shallow samples only), island (shallow samples only), depth zone only (all samples), depth zone only (O&#x02BB;ahu samples), island group and depth zone (all samples).
+Variation among island groups and sampling depths was further examined and visualized using canonical analysis of principal coordinates (CAP) based on discriminant analysis [@andersonCanonicalAnalysisPrincipal2003] as implemented in the `CAPdiscrim` function from `BiodiversityR` [@kindtTreeDiversityAnalysis2005].
 To explore differences in community variability among groups (shallow/deep, O&#x02BB;ahu/all shallow samples), we compared Euclidean distances among samples in the CAP ordination using Welch two-sample t-tests, assuming unequal variances.
@@
 There were `r all_summary$singleton_count` singleton taxa (taxa occurring in a single unit across all sites), accounting for ~`r pc(all_summary$singleton_count/ntaxa(cc$crabs_untransformed),0.1)` of total recovered crab taxa.
-For a breakdown of specific crab taxa recovered by island and deployment depth, see [@tbl:species_table].
+For a breakdown of specific crab taxa recovered by island and deployment depth, see [@tbl:species_table_s].
 
@@
 
+### Comparisons {-}
+```{r depth-ranges, results="hide", dependson="taxonomy-validation"}
+f <- cc$crabs %>%
+  psmelt() %>%
+  group_by(depth_zone) %>%
+  filter(Abundance > 0) %>%
+  ungroup() %>%
+  group_by(family) %>%
+  filter(n_distinct(shallow_deep) == 1) %>%
+  ungroup()
+
+fam <- f %>%
+  distinct(family,shallow_deep) %>%
+  count(shallow_deep) %>%
+  deframe()
+
+deep_fams <- f %>%
+  filter(shallow_deep == "deep") %>%
+  pull(family) %>%
+  unique() %>%
+  sort()
+
+g <- cc$crabs %>%
+  psmelt() %>%
+  filter(!is.na(genus)) %>%
+  group_by(depth_zone) %>%
+  filter(Abundance > 0) %>%
+  ungroup() %>%
+  group_by(genus) %>%
+  filter(n_distinct(shallow_deep) == 1) %>%
+  ungroup()
+
+gen <- g %>%
+  distinct(genus,shallow_deep) %>%
+  count(shallow_deep) %>%
+  deframe()
+
+deep_gens <- g %>%
+  filter(shallow_deep == "deep") %>%
+  pull(genus) %>%
+  unique() %>%
+  sort()
+
+s <- cc$crabs %>%
+  psmelt() %>%
+  group_by(depth_zone) %>%
+  filter(Abundance > 0) %>%
+  ungroup() %>%
+  group_by(species) %>%
+  filter(n_distinct(shallow_deep) == 1) %>%
+  ungroup()
+
+spp <- s %>%
+  distinct(species,shallow_deep) %>%
+  count(shallow_deep) %>%
+  deframe()
+
+deep_spp <- s %>%
+  filter(shallow_deep == "deep") %>%
+  pull(species) %>%
+  unique() %>%
+  sort()
+```
+Out of the `r all_summary$taxonomy_breakdown$families` families recovered, `r fam['shallow']` were exclusive to shallow ARMS recovered from `r dr(cc$crabs_shallow_untransformed)` m and `r fam['deep']` were exclusive to units $\geq$ 30 m (`r format_list(deep_fams)`; [@tbl:family_depth_table_s]).
+`r str_to_sentence(english(gen['shallow']))` genera were found exclusively shallow and `r gen['deep']` exclusively deep (`r format_list(deep_gens)`; [@tbl:genus_depth_table_s]).
+`r str_to_sentence(english(spp['shallow']))` distinct taxa (including those collapsed to genus or family) were exclusive to shallow ARMS units and `r spp['deep']` were found only deep (`r format_list(deep_spp)`; [@tbl:species_depth_table_s]).
+
 ## Alpha diversity {-}
@@
 Richness was not significantly different between shallow and deep samples (`r k_test(all_alpha$k_r_depth,plain=TRUE)`; [@fig:fig_alpha[]{}b]) but effective Simpson diversity was (`r k_test(all_alpha$k_s_depth,plain=TRUE)`; [@fig:fig_alpha[]{}d]).
-When comparing diversity across sampling depths for only sites in the MHI, richness was not significantly different (`r k_test(all_alpha$k_r_depth_main,plain=TRUE)`; [@fig:fig_alpha_main[]{}a]) but effective Simpson diversity was (`r k_test(all_alpha$k_s_depth_main,plain=TRUE)`; [@fig:fig_alpha_main[]{}b]).
-Neither richness nor Simpson diversity were significantly different between shallow and deep when comparisons were constrained to the island of O&#x02BB;ahu (Richness: `r k_test(all_alpha$k_r_depth_oahu,plain=TRUE)`; Simpson: `r k_test(all_alpha$k_s_depth_oahu,plain=TRUE)`).
+When comparing diversity across sampling depths for only sites in the MHI, richness was not significantly different (`r k_test(all_alpha$k_r_depth_main,plain=TRUE)`; [@fig:fig_alpha_main[]{}a]) but effective Simpson diversity was (`r k_test(all_alpha$k_s_depth_main,plain=TRUE)`; [@fig:fig_alpha_main[]{}c]).
+Neither richness nor Simpson diversity were significantly different between shallow and deep when comparisons were constrained to the island of O&#x02BB;ahu (Richness: `r k_test(all_alpha$k_r_depth_oahu,plain=TRUE)`, [@fig:fig_alpha_main[]{}b]; Simpson: `r k_test(all_alpha$k_s_depth_oahu,plain=TRUE)`, [@fig:fig_alpha_main[]{}d]).
 When only shallow arms \< 30 m were considered, there were no significant differences in alpha diversity between the MHI and NWHI for either index (richness: `r k_test(shallow_alpha$k_r_island,plain=TRUE)`, [@fig:fig_alpha2[]{}a]; effective Simpson: `r k_test(shallow_alpha$k_s_island,plain=TRUE)`, [@fig:fig_alpha2[]{}b]).
@@
 `r str_to_sentence(english(nterm))` of the `r str_to_lower(english(nrow(model_terms)))` environmental variables tested were significant and sufficiently uncorrelated.
-The two drivers contributing most to explained variation were `r drivers$term[1]` (F = `r num(drivers$stat[1],0.01)`) and `r drivers$term[2]` (F = `r num(drivers$stat[2],0.01)`) ([@tbl:rda_table]).
+The two drivers contributing most to explained variation were `r drivers$term[1]` (F = `r num(drivers$stat[1],0.01)`) and `r drivers$term[2]` (F = `r num(drivers$stat[2],0.01)`) ([@tbl:rda_table_s]).
 These factors also produced the longest effect vectors in the dbRDA plot ([@fig:fig_dbrda]).
@@
 anova_depth <- adonis2(cc$crab_dist ~ depth_zone,data=cc$crab_data,permutations=PERM)
+anova_depth_oahu <- adonis2(cc$crab_dist_oahu ~ depth_zone,data=sample_tibble(cc$crabs_oahu),permutations=PERM)
+anova_depth2_oahu <- adonis2(cc$crab_dist_oahu ~ shallow_deep,data=sample_tibble(cc$crabs_oahu),permutations=PERM)
 bd_depth <- betadisper(cc$crab_dist,cc$crab_data$depth_zone)
@@
 There was no significant difference in community dispersion between the MHI and NWHI (`r permdisp(bd_region)`).
-When mesophotic ($\geq$ 30 m) ARMS units were included, there were significant differences among depth zones (\< 30 m, 30 m, 60 m, 90 m) (`r permanova(anova_depth,0.01)`) and community dispersion (`r permdisp(bd_depth)`).
+When mesophotic ($\geq$ 30 m) ARMS units were included, there were significant differences among depth zones (\< 30 m, 30 m, 60 m, 90 m) both for all samples (`r permanova(anova_depth,0.01)`) and samples restricted to O&#x02BB;ahu (`r permanova(anova_depth_oahu,0.01)`), and community dispersion (`r permdisp(bd_depth)`).
 However, community dispersion was not different among depth zones $\geq$ 30 m (`r permdisp(bd_deep)`).
@@
     oahu = factor(
-      if_else(island == "Oahu","oahu","noahu"),
-      levels = c("oahu","noahu")
+      # if_else(island == "Oahu","oahu","noahu"),
+      case_when(
+        island == "Oahu" ~ "oahu",
+        island == "French Frigate Shoals" ~ "ffs",
+        .default = "noahu"
+      ),
+      levels = c("oahu","ffs","noahu")
     )
@@
   geom_label_repel(data=centroids,aes(label=sample_group), color="white", show.legend = FALSE) + 
-  scale_shape_manual(values=c("oahu" = 23, "noahu" = 21),name="Island", labels=c("noahu"="Other islands","oahu"="O‘ahu")) +
+  scale_shape_manual(values=c("oahu" = 23, "ffs" = 22, "noahu" = 21),name="Island", labels=c("noahu"="Other islands","oahu"="O‘ahu","ffs"="Lalo (French Frigate Shoals)" )) +
   scale_fill_manual(values=sample_depth_pal,name="Sample group") +
@@
 
+# centroids <- pcap %>% group_by(sample_group) %>%
+#   summarise(LD1=mean(LD1),LD2=mean(LD2))
+# pcap <- pcap %>%
+#   mutate(
+#     oahu = factor(
+#       if_else(island == "Oahu","oahu","noahu"),
+#       levels = c("oahu","noahu")
+#     )
+#   )
+
 # ggplot(pcap,aes(x=LD1,y=LD2,fill=sample_group)) + 
 #   stat_ellipse(geom="polygon",level=0.95,alpha=0.1,show.legend = FALSE) +
-#   geom_point(color="black",shape=21,size=5)  +
+#   # geom_point(aes(shape=oahu),color="black",shape=21,size=5)  +
+#   geom_point(mapping=aes(shape=oahu),color="black",size=5)  +
 #   geom_label_repel(data=centroids,aes(label=sample_group), color="white", show.legend = FALSE) + 
+#   scale_shape_manual(values=c("oahu" = 23, "noahu" = 21),name="Island", labels=c("noahu"="Other islands","oahu"="O‘ahu")) +
 #   scale_fill_manual(values=sample_depth_pal,name="Sample group") +
+#   guides( 
+#     fill=guide_legend(override.aes=list(shape=21)), 
+#     shape=guide_legend(override.aes=list(fill="black"))
+#   ) +
 #   xlab(str_glue("Linear discriminant 1 [{pc(cap_var$x,0.1)}]")) +
@@
 Productivity gradients across the Hawaiian Archipelago coincide with differences in coral reef community composition [@selkoeDNACoralReef2016], endemic reef fish dominance [@friedlanderDominanceEndemicsReef2020], survival rates of top predators [@bakerEffectVariableOceanic2007], and island geomorphology [@goveNearislandBiologicalHotspots2016].
-Thus, it is not surprising that we find the same top drivers of community structure for one of the primary taxonomic groups inhabiting coral reefs, the brachyuran crabs, across the Hawaiian archipelago.
+Although not a significant factor in our model, coral cover and species composition are thought to affect patterns of benthic assemblages.
+Due to the limited availability of datasets covering our entire study area, our model only included coral cover at the island scale, but previous work has shown that coral species composition does not vary considerably across the archipelago [@griggCommunityStructureSuccession1983].
+Crab communities from Lalo (French Frigate Shoals), which was found to harbor the highest coral species diversity [@griggCommunityStructureSuccession1983], did not group separately from the rest of our samples ([@fig:fig_cap]).
+Given the prevalence in our model of significant factors known to be important influencers of marine communities in Hawai&#x02BB;i, it is not surprising that we find the same top drivers of community structure for brachyuran crabs, one of the primary taxonomic groups inhabiting coral reefs.
 What is surprising---given the maximum vertical separation among shallow ARMS units of `r diff(range(cc$crab_data_shallow_unscaled$depth))` m---is the consistent significance of depth as a driver of community composition across these gradients.
@@
 
-While depth was a significant driver of community composition among shallow-water ARMS units ([@fig:fig_dbrda]), its influence was even more striking when comparing shallow to mesophotic samples.
+While depth was a significant driver of community composition among shallow-water ARMS units ([@fig:fig_dbrda; @tbl:rda_table_s]), its influence was even more striking when comparing shallow to mesophotic samples.
 Due to logistical constraints (and the limitations of satellite-derived data products), our environmental data only covers shallow-water deployments, but CAP ordinations show considerable differences between the grouping of all shallow communities and the units deployed $\geq$ 30 m at the single island of O&#x02BB;ahu ([@fig:fig_cap]).
@@
 Despite the significance of these factors in driving shallow community variation across the archipelago, mesophotic assemblages were found to be distinct ([@fig:fig_cap]).
-Although there were variables we did not consider in our model, this suggests that vertical distributions are influenced by factors other than those driving spatial variation aross the island chain.
+Although there were variables we did not consider in our model, this suggests that vertical distributions are influenced by factors other than those driving spatial variation across the island chain.
 In light of this, and since mesophotic communities are generally less biodiverse than shallow reefs [@kahngCommunityEcologyMesophotic2010; @lesserEcologyMesophoticCoral2009; @pyleComprehensiveInvestigationMesophotic2016], it may be possible that the gradients driving depth distributions on coral reefs run opposite to those in the intertidal.
@@
 Hee, for discussion, support, and their efforts in the disassembly and sorting of ARMS samples on retrieval.
-Finally, we are grateful to the two anonymous reviewers for constructive comments and suggestions on an earlier version of this paper.
+Finally, we are grateful to two anonymous reviewers for constructive comments and suggestions on an earlier version of this paper.
 This is contribution #xxx from the Hawai&#x02BB;i Institute of Marine Biology and #xxx from the School of Ocean and Earth Science and Technology at the University of Hawai&#x02BB;i.
@@
 Values (SST or chl-A) were extracted for the geographic coordinates of each sampling site from the averaged dataset for the relevant soaking period.
-Original raw monthly-average dataset names can be found in @tbl:nasa_table.
+Original raw monthly-average dataset names can be found in @tbl:nasa_table_s.
 
 ### Processing of cumulative human impacts data
-Gridded cumulative human impacts raster data calcuated by Halpern et al [-@halpernRecentPaceChange2019] was downloaded from the Knowledge Network for Biocomplexity (KNB) repository [@frazierRecentPaceChange2019].
+Gridded cumulative human impacts raster data calculated by Halpern et al [-@halpernRecentPaceChange2019] was downloaded from the Knowledge Network for Biocomplexity (KNB) repository [@frazierRecentPaceChange2019].
 Human impacts datasets were downloaded for each available year in this study (2009--2013) and averaged over ARMS unit soaking periods (as above).
@@
 
-```{r fig-alpha-main, fig.cap='Box plots of species richness and effective Simpson diversity per depth zone for Main Hawaiian Islands only.
+```{r fig-alpha-main, fig.cap='Box plots of species richness and effective Simpson diversity per depth zone for Main Hawaiian Islands (a,c) and O&#x02BB;ahu (b,d) only.
 Asterisks and notches as above.
-(a) Species richness by sampling depth.
-(b) Effective Simpson diversity by sampling depth.', results="asis", dev="svg", fig.width=8, fig.height=4, dependson="init"}
+(a) Species richness by sampling depth, MHI.
+(b) Species richness by sampling depth, O&#x02BB;ahu.
+(c) Effective Simpson diversity by sampling depth, MHI.
+(d) Effective Simpson diversity by sampling depth, O&#x02BB;ahu.', results="asis", dev="svg", fig.width=10, fig.height=10, dependson="init"}
 # depth comparison for main islands only
@@
   plotz_theme()  + 
-  xlab("Sampling depth") + 
+  xlab("Sampling depth - MHI") + 
   ylab("Richness")
@@
   plotz_theme()  + 
-  xlab("Sampling depth") + 
+  xlab("Sampling depth - MHI") + 
   ylab("Effective Simpson diversity")
 
-(p1 + p2) & 
-  plot_annotation(tag_levels = list(str_glue("({letters[1:2]})"))) & 
+art <- all_alpha$richness_table %>% 
+  filter(island == "Oahu") %>%
+  mutate(island_group = str_to_sentence(island_group)) %>%
+  mutate(shallow_deep = str_to_sentence(shallow_deep),shallow_deep = fct_reorder(shallow_deep,depth))
+
+if (all_alpha$k_r_depth_oahu$p.value < 0.05) {
+  which_max <- art %>% 
+    group_by(shallow_deep) %>% 
+    summarise(r = mean(Observed)) %>%
+    slice(which.max(r)) %>%
+    pull(shallow_deep) %>%
+    as.character()
+  max_r <- max(art$Observed) + (diff(range(art$Observed)))/10
+  sig <- tribble(
+    ~x,~y,
+    which_max,max_r
+  ) 
+} else {
+  sig <- tribble(~x,~y)
+}
+
+pointsize <- 5
+
+p3 <- ggplot(art) + 
+  geom_boxplot(aes(x=shallow_deep,y=Observed,fill=shallow_deep),notch=TRUE) +
+  geom_point(aes(x,y),data=sig,shape=8,size=pointsize,stroke=1.5) + 
+  scale_fill_manual(values=sample_depth_pal) +
+  guides(fill="none") + 
+  plotz_theme()  + 
+  xlab("Sampling depth - O‘ahu") + 
+  ylab("Richness")
+
+if (all_alpha$k_s_depth_oahu$p.value < 0.05) {
+  which_max <- art %>% 
+    group_by(shallow_deep) %>% 
+    summarise(r = mean(Simpson)) %>%
+    slice(which.max(r)) %>%
+    pull(shallow_deep) %>%
+    as.character()
+  max_r <- max(art$Simpson) + (diff(range(art$Simpson)))/10
+  sig <- tribble(
+    ~x,~y,
+    which_max,max_r
+  ) 
+} else {
+  sig <- tribble(~x,~y)
+}
+
+p4 <- ggplot(art) + 
+  geom_boxplot(aes(x=shallow_deep,y=Simpson,fill=shallow_deep),notch=TRUE) +
+  geom_point(aes(x,y),data=sig,shape=8,size=pointsize,stroke=1.5) + 
+  scale_fill_manual(values=sample_depth_pal) +
+  guides(fill="none") + 
+  plotz_theme()  + 
+  xlab("Sampling depth - O‘ahu") + 
+  ylab("Effective Simpson diversity")
+
+(p1 + p3) / (p2 + p4) & 
+  plot_annotation(tag_levels = list(str_glue("({letters[1:4]})"))) & 
   theme(plot.tag = element_text(face="italic"))
@@
 
-```{r fig-euclid, fig.cap='Box plot of Euclidean distances between points in CAP ordination by depth zone', results="asis", dev="svg", fig.width=6, fig.height=4, dependson="cap-analysis"}
+```{r fig-euclid, fig.cap='Box plot of Euclidean distances between points in CAP ordination by depth zone.
+Asterisk indicates significant difference (Welch two-sample t-test).', results="asis", dev="svg", fig.width=6, fig.height=4, dependson="cap-analysis"}
 dists <- bind_rows(
@@
 
+if (distance_t$p.value < 0.05) {
+  which_max <- dists %>%
+    group_by(depth) %>%
+    summarise(d = mean(dist)) %>%
+    slice(which.max(d)) %>%
+    pull(depth) %>%
+    as.character()
+  max_d <- max(dists$dist) + (diff(range(dists$dist)))/10
+  sig <- tribble(
+    ~x,~y,
+    which_max,max_d
+  ) 
+} else {
+  sig <- tribble(~x,~y)
+}
+
+pointsize <- 5
 ggplot(dists,aes(x=depth,y=dist,fill=depth)) + 
   geom_boxplot(notch=TRUE) + 
+  geom_point(aes(x,y),data=sig,shape=8,size=pointsize,stroke=1.5,fill="black") +
   scale_fill_manual(values=sample_depth_pal) + 
@@
   })
-print_table(species_tbl,"species_table","List of crab taxa detected by island and ARMS unit deployment depth",hdr_case="none",accuracy=1)
+print_table(species_tbl,"species_table_s","List of crab taxa detected by island and ARMS unit deployment depth",hdr_case="none",accuracy=1)
 ```
@@
 
+```{r family-depth-table, results="asis", dependson="taxonomy-validation"}
+deep_map <- c(
+  "shallow" = str_glue("Shallow ({dr(cc$crabs_shallow_untransformed)} m)"),
+  "deep" = "Deep ($\\geq$ 30 m)"
+)
+crabbo <- cc$crabs %>%
+  psmelt() %>%
+  mutate(depth_zone = replace(depth_zone,which(depth_zone == "shallow"),str_c(dr(cc$crabs_shallow_untransformed)," m"))) %>%
+  group_by(depth_zone) %>%
+  filter(Abundance > 0) %>%
+  ungroup() %>%
+  group_by(family) %>%
+  filter(n_distinct(shallow_deep) == 1) %>%
+  summarise(zones = str_c(unique(depth_zone),collapse=", "), deeps = str_c(unique(shallow_deep),collapse=", "),species=str_c(unique(species),collapse=", ")) %>%
+  ungroup() %>%
+  arrange(deeps,zones,family,species) %>%
+  select(deeps,family,species,zones) %>%
+  mutate( deeps = deep_map[deeps] ) %>%
+  rename(Family=family,`"Species"`=species,`Depth range`=deeps,`Depth zones`=zones)
+
+print_table(crabbo,"family_depth_table_s","Brachyuran families exclusive to shallow or mesophotic habitat",hdr_case="none",accuracy=1)
+```
+
+\newpage
+
+```{r genus-depth-table, results="asis", dependson="taxonomy-validation"}
+deep_map <- c(
+  "shallow" = str_glue("Shallow ({dr(cc$crabs_shallow_untransformed)} m)"),
+  "deep" = "Deep ($\\geq$ 30 m)"
+)
+crabbo <- cc$crabs %>%
+  psmelt() %>%
+  mutate(depth_zone = replace(depth_zone,which(depth_zone == "shallow"),str_c(dr(cc$crabs_shallow_untransformed)," m"))) %>%
+  group_by(depth_zone) %>%
+  filter(Abundance > 0) %>%
+  ungroup() %>%
+  group_by(genus) %>%
+  filter(n_distinct(shallow_deep) == 1) %>%
+  summarise(zones = str_c(unique(depth_zone),collapse=", "), deeps = str_c(unique(shallow_deep),collapse=", "),species=str_c(unique(species),collapse=", ")) %>%
+  ungroup() %>%
+  arrange(deeps,zones,genus,species) %>%
+  select(deeps,genus,species,zones) %>%
+  mutate( deeps = deep_map[deeps] ) %>%
+  rename(Genus=genus,`"Species"`=species,`Depth range`=deeps,`Depth zones`=zones)
+
+print_table(crabbo,"genus_depth_table_s","Brachyuran genera exclusive to shallow or mesophotic habitat",hdr_case="none",accuracy=1)
+```
+
+\newpage
+
+```{r species-depth-table, results="asis", dependson="taxonomy-validation"}
+deep_map <- c(
+  "shallow" = str_glue("Shallow ({dr(cc$crabs_shallow_untransformed)} m)"),
+  "deep" = "Deep ($\\geq$ 30 m)"
+)
+crabbo <- cc$crabs %>%
+  psmelt() %>%
+  mutate(depth_zone = replace(depth_zone,which(depth_zone == "shallow"),str_c(dr(cc$crabs_shallow_untransformed)," m"))) %>%
+  group_by(depth_zone) %>%
+  filter(Abundance > 0) %>%
+  ungroup() %>%
+  group_by(species) %>%
+  filter(n_distinct(shallow_deep) == 1) %>%
+  summarise(zones = str_c(unique(depth_zone),collapse=", "), deeps = str_c(unique(shallow_deep),collapse=", ")) %>%
+  ungroup() %>%
+  arrange(deeps,zones,species) %>%
+  select(deeps,species,zones) %>%
+  mutate( deeps = deep_map[deeps] ) %>%
+  rename(`"Species"`=species,`Depth range`=deeps,`Depth zones`=zones)
+
+print_table(crabbo,"species_depth_table_s","Brachyuran taxa exclusive to shallow or mesophotic habitat",hdr_case="none",accuracy=1)
+```
+
+\newpage
+
 ```{r rda-table,results="asis",dependson="rda-analysis"}
-print_table(rda_table,"rda_table","Significant marginal terms in the dbRDA analysis of crab community variation across the Hawaiian Archipelago",hdr_case="none",accuracy=0.001) 
+print_table(rda_table,"rda_table_s","Significant marginal terms in the dbRDA analysis of crab community variation across the Hawaiian Archipelago",hdr_case="none",accuracy=0.001) 
 ```
@@
 nasa_table <- read_csv(here("data","tables","nasa.csv"))
-print_table(nasa_table,"nasa_table","NASA data repository filenames for CHL and SST satellite data",accuracy=1,hdr_case=NA,format_numbers=FALSE) 
+print_table(nasa_table,"nasa_table_s","NASA data repository filenames for CHL and SST satellite data",accuracy=1,hdr_case=NA,format_numbers=FALSE) 
 ```

### 2024-10-18 crabs.R

@@ -358,6 +358,15 @@ setup_crabs <- function() {
   cc$crab_otus_shallow <- otu_table(cc$crabs_shallow)
   cc$crab_data_shallow <- sample_tibble(cc$crabs_shallow)
   
+  
+  cc$crabs_oahu <- prune_samples(
+    sample_tibble(cc$crabs)$island == "Oahu",
+    cc$crabs
+  )
+  # prune taxa
+  cc$crabs_oahu <- prune_taxa(taxa_sums(cc$crabs_oahu) > 0,cc$crabs_oahu)
+  cc$crab_dist_oahu <- distance(cc$crabs_oahu,"bray")
+  
   return(cc)
 }
 

### 2024-11-19 manuscript.Rmd
some small updates

@@
 ---
+<!--
+TODO: italics in tables S4/S5
+-->
 <!--word count: Proceedings B publishes articles up to 12 pages, including figures and tables, with no more than 10 000 words.
@@
 # do an english comma-separated list
-format_list <- function(l) {
-  l[length(l)] <- str_c("and ",l[length(l)])
+format_list <- function(l,fmt=function(x) x) {
+  l <- fmt(l)
+  if(length(l) > 1) l[length(l)] <- str_c("and ",l[length(l)])
   if (length(l) > 2) { 
@@
     )
-}
-
-# abstract method to format species name
-format_species <- function(...) UseMethod("format_species")
-
-# format species in a list
-format_species.list <- function(spp,authority=TRUE,...) {
-  if (spp$found) {
-    a <- ifelse(authority,str_glue(" {spp$authority}"),"")
-    str_glue("*{spp$species}*{a}")
-  } else {
-    s <- str_split(spp$species," ")[[1]]
-    str_trim(str_glue("*{s[1]}* {s[2]}"))
-  }
-}
-
-# format species in a character vector
-format_species.character <- function(species,found,authority,auth=TRUE,...) {
-  spp <- map2_chr(species,found,~{
-    .y <- ifelse(is.na(.y),FALSE,.y)
-    s <- list(species=.x,found=.y,authority="")
-    format_species(s,FALSE)
-  })
-  if (auth) {
-    authority <- if_else(!is.na(authority),str_c(" ",authority),"")
-    spp <- str_c(spp,authority)
-  }
-  return(spp)
-}
-
-# format species in a data frame
-format_species.data.frame <- function(spp,authority=TRUE,...) {
-  spp %>%
-    pmap_chr(~{
-      row <- list(...)
-      s <- row[c('species','found','authority','otu')]
-      format_species(s,authority)
-    })
 }
@@
 # format a top-n list of species
-top_list <- function(species,n,units,len=4,fig=NULL) {
-  s1 <- map_chr(species,format_species)
+top_list <- function(species,n,units,len=4,fig=NULL,taxonomy) {
+  s1 <- species 
   s2 <- map2_chr(n,units,~{
@@
       .default = authority
-    )
-  ) %>%
-  mutate(display_species = format_species(.))
-```
-
-```{r summaries, include=FALSE, dependson="init"}
+    ),
+    genus = coalesce(genus.y,genus.x),
+    family = coalesce(family.y,family.x),
+    display_species = case_when(
+      taxon_level == "species" ~ str_glue("*{species}*"),
+      taxon_level == "genus" ~ str_glue("*{genus}* sp."),
+      taxon_level == "family" ~ str_glue("{family} sp."),
+      .default = species 
+    )
+  )
+```
+
+```{r summaries, include=FALSE, dependson="taxonomy-validation"}
 # generate sample summaries. code block is cached 
@@
 
-To date, the majority of studies that have examined community structure across the Hawaiian archipelago and along a depth gradient have focused on conspicuous taxa such as corals and fishes [e.g., @benfieldComparisonCoralReef2008; @caballero-aragonWaveExposureTemperature2023; @kaneTrophicDesignationLive2017; @langeWaveExposureShapes2021; @vanlierHabitatConnectivityComplexity2018].
+To date, the majority of studies that have examined community structure across the Hawaiian archipelago and/or along a depth gradient have focused on conspicuous taxa such as corals and fishes [e.g., @benfieldComparisonCoralReef2008; @caballero-aragonWaveExposureTemperature2023; @kaneTrophicDesignationLive2017; @langeWaveExposureShapes2021; @vanlierHabitatConnectivityComplexity2018].
 However, these taxa make up a minority of coral reef biodiversity and thus may not be indicative of community patterns overall [@fautinOverviewMarineBiodiversity2010].
@@
   sort()
+deep_gens <- str_glue("*{deep_gens}*")
 
@@
   filter(shallow_deep == "deep") %>%
-  pull(species) %>%
-  unique() %>%
-  sort()
+  arrange(species) %>%
+  select(OTU) %>%
+  left_join(crab_taxonomy_validated,by=c("OTU" = "otu")) %>%
+  pull(display_species) %>%
+  unique()
 ```
@@
   ungroup() %>%
+  left_join(crab_taxonomy_validated %>% select(otu,display_species),by=c("OTU" = "otu")) %>%
   group_by(family) %>%
   filter(n_distinct(shallow_deep) == 1) %>%
-  summarise(zones = str_c(unique(depth_zone),collapse=", "), deeps = str_c(unique(shallow_deep),collapse=", "),species=str_c(unique(species),collapse=", ")) %>%
+  summarise(
+    zones = str_c(unique(depth_zone),collapse=", "),
+    deeps = str_c(unique(shallow_deep),collapse=", "),
+    species = format_list( unique(display_species) )
+  ) %>%
   ungroup() %>%
@@
   ungroup() %>%
+  left_join(crab_taxonomy_validated %>% select(otu,display_species),by=c("OTU" = "otu")) %>%
   group_by(genus) %>%
   filter(n_distinct(shallow_deep) == 1) %>%
-  summarise(zones = str_c(unique(depth_zone),collapse=", "), deeps = str_c(unique(shallow_deep),collapse=", "),species=str_c(unique(species),collapse=", ")) %>%
+  summarise(
+    zones = str_c(unique(depth_zone),collapse=", "),
+    deeps = str_c(unique(shallow_deep),collapse=", "),
+    species=format_list(unique(display_species))
+  ) %>%
   ungroup() %>%
@@
   ungroup() %>%
-  group_by(species) %>%
+  left_join(crab_taxonomy_validated %>% select(otu,display_species),by=c("OTU" = "otu")) %>%
+  group_by(display_species) %>%
   filter(n_distinct(shallow_deep) == 1) %>%
@@
   ungroup() %>%
-  arrange(deeps,zones,species) %>%
-  select(deeps,species,zones) %>%
+  arrange(deeps,zones,display_species) %>%
+  select(deeps,species=display_species,zones) %>%
   mutate(  deeps = deep_map[deeps] ) %>%

### 2024-11-19 crabs.R

@@ -397,7 +397,7 @@ sample_summary <- function(ps,top_n=5,taxonomy=NULL) {
     sample_sums() %>%
     mean()
   
-  # overall abundace sd
+  # overall abundance sd
   ss$abundance_sd <- ps %>%
     sample_sums() %>%
     sd()
@@ -407,17 +407,17 @@ sample_summary <- function(ps,top_n=5,taxonomy=NULL) {
     taxa_sums() %>%
     sort() %>%
     rev() %>%
-    head(top_n) %>%
+    head(top_n) 
+  
+  ss$top_otus <- names(top) 
+  
+  top <- top %>%
     set_names(otu_lookup[names(.)])
   
+  ss$top_taxa <- names(top)
   ss$top_n <- top
   
-  to_keep <- ps %>%
-    taxa_tibble() %>%
-    filter(species %in% names(top)) %>%
-    pull(otu)
-  
-  ss$top_units <- prune_taxa(to_keep,ps) %>%
+  ss$top_units <- prune_taxa(ss$top_otus,ps) %>%
     ps_standardize("pa") %>%
     taxa_sums() %>% 
     sort() %>%
@@ -427,31 +427,16 @@ sample_summary <- function(ps,top_n=5,taxonomy=NULL) {
   # make it the same order as the top 5 species
   ss$top_units <- ss$top_units[match(names(ss$top_units),names(top))]
   
-  if (is.null(taxonomy)) {
-    ss$top <- wm_records_taxamatch(names(top)) %>%
-      map2(names(top),~{
-        if (nrow(.x) > 0) {
-          .x <- .x %>% 
-            filter(rank == "Species") %>%
-            select(species = scientificname, authority) %>%
-            slice(1)  %>%
-            mutate(found = TRUE,otu = otu_lookup[.y])
-        } 
-        if (nrow(.x) == 0) {
-          .x <- list(species = .y, authority = NA, found = FALSE, otu=otu_lookup[.y])
-        }
-        return(as.list(.x))
-      })
-  } else {
-    ss$top <- taxonomy %>%
-      filter(species %in% names(top)) %>%
-      mutate(found = species == valid_name) %>%
-      select(species,authority=valid_authority,found,otu) %>%
-      slice(match(names(top),species)) %>%
-      split(seq(nrow(.))) %>%
-      map(as.list)
-  }
-  
+  ss$top <- taxonomy %>%
+    filter(otu %in% ss$top_otus) %>%
[…]

### 2024-11-22 manuscript.Rmd
molly's suggestions

@@
 An inability to consistently resolve the environmental and anthropogenic factors that shape coral reefs impacts our understanding of how those factors interact and how coral reef communities may respond to change.
-Logistical challenges have additionally limited the taxonomic resolution and depth ranges investigated, with the focus largely on fishes and corals in the shallowest ~30 m of coral reefs.
+Logistical challenges have additionally limited the taxonomic resolution and depth ranges investigated, with previous studies focusing largely on fishes and corals in the shallowest ~30 m of coral reefs.
 Here, we examine the effects of a suite of environmental, ecological, and anthropogenic factors on the assemblage structure of brachyuran crabs recovered from Autonomous Reef Monitoring Structures on shallow reefs (`r dr(cc$crabs_shallow_untransformed)` m) across the ~2,500 km span of the Hawaiian Archipelago.
 We compare these assemblages with those sampled along a depth gradient (`r format_range(range(cc$crab_data$depth[cc$crab_data$island == "Oahu"]))` m) on O&#x02BB;ahu.
-We recovered `r num(sum(sample_sums(cc$crabs_untransformed)),1)` individuals representing `r sum(unlist(all_summary$types_breakdown))` taxa from `r all_summary$taxonomy_breakdown$families` families.
-Shallow assemblages were significantly different between the Main and Northwest Hawaiian Islands and significantly associated with human impacts, chlorophyll-A, sea-surface temperature, slope, potential larval immigration, and depth.
+<!--We recovered `r num(sum(sample_sums(cc$crabs_untransformed)),1)` individuals representing `r sum(unlist(all_summary$types_breakdown))` taxa from `r all_summary$taxonomy_breakdown$families` families.-->Shallow assemblages were significantly different between the Main and Northwest Hawaiian Islands and significantly associated with human impacts, chlorophyll-A, sea-surface temperature, slope, potential larval immigration, and depth.
 However, we found more variation among mesophotic samples from O&#x02BB;ahu (30--90 m) than among shallow samples across the entire archipelago, suggesting that 90 m of depth is a stronger determinant of brachyuran assemblage structure than the latitudinal, environmental, or anthropogenic gradients across the Hawaiian Islands.
@@
 Such standardization can reduce variability in recovered communities that may arise due to local variation in habitat composition and availability [@ransomeImportanceStandardizationBiodiversity2017].
-We targeted brachyuran crabs, a ubiquitous group often used as an exemplar to assess coral reef biodiversity [@knowltonCoralReefBiodiversity2010; @plaisanceDiversityCoralReefs2011; @servisCharacterizingCoralReef2020; @plaisanceReefassociatedCrustaceanFauna2009] and comprising a notable proportion of the coral reef cryptofauna [284 species in Hawai&#x02BB;i, @castroCatalogAnomuranBrachyuran2011].
+We targeted brachyuran crabs, a ubiquitous group often used as an exemplar to assess coral reef biodiversity [@knowltonCoralReefBiodiversity2010; @plaisanceDiversityCoralReefs2011; @servisCharacterizingCoralReef2020; @plaisanceReefassociatedCrustaceanFauna2009] and comprising a notable proportion of the Hawaiian coral reef cryptofauna [284 reported species, @castroCatalogAnomuranBrachyuran2011].
 We used assemblages recovered from ARMS units to examine drivers of community composition across the Hawaiian Archipelago, compared crab communities among islands and regions (MHI vs.
-NWHI), and examined the effects of a suite of possible environmental and anthropogenic drivers of community structure to test hypotheses about their relative importance.
-
+NWHI), and examined the effects of a suite of possible environmental and anthropogenic drivers of community structure, including depth, to test hypotheses about their relative importance.
+
+<!--
 Much of our understanding of biotic zonation and community structure in the ocean originates with classic studies from the intertidal [e.g., @connellInfluenceInterspecificCompetition1961; @connellEffectsCompetitionPredation1961; @paineIntertidalCommunityStructure1974].
-While there are no shortages of exceptions, a long history of studies point towards a generalization that the lower bathymetric limits are generally regulated by ecological interactions whereas upper limits of intertidal communities are more often determined by abiotic factors [@stephenson1972life; @mengeRockyIntertidalCommunities2001; @bertnessTestingRelativeContribution1999; @dennyPhysicalProcessesThat2001].
-However, most of these models describe zonation at specific localities rather than community variation over larger spatial scales [but see @birdTideWaveMarks2013; @bishop-taylorTidesModellingElevation2019] and may not apply to coral reefs, which are subject to more complex community dynamics [@abramsRecruitmentLotteriesCoexistence1984; @bussCompetitiveNetworksNontransitive1979; @hixonPredationPreyRefuges1993; @hubbellUnifiedNeutralTheory2011; @moraEcologyFishesCoral2015; @saleStockrecruitRelationshipsRegional1982; @sheppardBiologyCoralReefs2017].
+While there are no shortages of exceptions, a long history of studies point towards a generalization that lower bathymetric limits are regulated by ecological interactions whereas upper limits of intertidal communities are more often determined by abiotic factors [@stephenson1972life; @mengeRockyIntertidalCommunities2001; @bertnessTestingRelativeContribution1999; @dennyPhysicalProcessesThat2001].
+However, most of these models describe vertical zonation at specific localities rather than community variation over larger spatial scales [but see @birdTideWaveMarks2013; @bishop-taylorTidesModellingElevation2019] and may not apply to coral reefs, which are subject to more complex community dynamics [@abramsRecruitmentLotteriesCoexistence1984; @bussCompetitiveNetworksNontransitive1979; @hixonPredationPreyRefuges1993; @hubbellUnifiedNeutralTheory2011; @moraEcologyFishesCoral2015; @saleStockrecruitRelationshipsRegional1982; @sheppardBiologyCoralReefs2017].
 While there is little doubt that ecological interactions are important on coral reefs, the suite of factors and which play the most important role in structuring coral reef communities remains a subject of considerable debate [@bellwoodRegionalscaleAssemblyRules2001; @connollyUnifiedModelExplains2017; @hubbellUnifiedNeutralTheory2011; @jacksonAdaptationDiversityReef1991; @mumbyCoralReefHabitats2008; @saleMaintenanceHighDiversity1977].
+-->
 
@@
 
-A total of 89 ARMS units were deployed over the course of six years (2009--2016) in shallow (`r dr(cc$crabs_shallow_untransformed)` m depth) coral reef habitat in both the Main (MHI) and Northwest Hawaiian Islands (NWHI) ([@fig:fig_map; @tbl:arms_table]).
-Only units deployed on outer reefs were included in this study.
+A total of 89 ARMS units were deployed over the course of six years (2009--2016) in shallow (`r dr(cc$crabs_shallow_untransformed)` m depth) coral reefs in both the Main (MHI) and Northwest Hawaiian Islands (NWHI) ([@fig:fig_map; @tbl:arms_table]).
+All units included in this study were deployed in forereef habitat.
 To explore the effects of depth, we also present data from an earlier study by Hurley et al.
@@
 However, the prevalence of species complexes and cryptic diversity suggests that our assignments are likely to under- rather than overrepresent the true number of taxa.
-Thus, any significant patterns detected likely persist in spite of rather than because of cryptic diversity.
+Thus, any significant patterns we detected likely persist in spite of the potential presence of cryptic diversity.
 
@@
 Even though our shallow-water ARMS units were deployed within a narrow depth range (`r dr(cc$crabs_shallow_untransformed)` m), depth still emerged as a significant and consistent driver of community variation among these units.
-Depth in and of itself is unlikely to affect marine community composition.
-At the same time, oceanographic factors with well-known depth profiles (e.g., temperature, oxygen, nutrients, etc.) are not expected to vary dramatically across such a narrow depth range (\< 10 m), although fine-scale habitat variability among sites may also influence the community differences we found [@steyaertRemoteReefCryptobenthic2022].
-Unfortunately, we lack environmental data at depth for individual deployments, and thus cannot unravel the co-varying factor(s) associated with the depth effects we observed.
+Depth in and of itself is unlikely to affect marine community composition and oceanographic factors with well-known depth profiles (e.g., temperature, oxygen, nutrients, etc.) are not expected to vary dramatically across such a narrow depth range (\< 10 m), although internal waves can deliver variable temperatures across shallow reefs [@storlazziInternalTidesCan2020; @wyattHeatAccumulationCoral2020] and fine-scale habitat variability among sites may also influence the community differences we found [@steyaertRemoteReefCryptobenthic2022].
+Unfortunately, we lack high resolution environmental data at depth for individual deployments, and thus cannot unravel the co-varying factor(s) associated with the depth effects we observed for shallow assemblages.
 Regardless, given its significance across both narrow (\< 10 m) and wider (90 m) depth ranges sampled here, a factor associated with depth is clearly an important driver for cryptobenthic crab communities across the Hawaiian Archipelago.
@@
 Level-3 mapped (4 km grid-size) sea-surface temperature (SST) and chlorophyll-A (chl-A) data were retrieved from NASA's Physical Oceanography Distributed Active Archive Center [PO.DAAC; @nasaobpgMODISAquaGlobal2020] and Ocean Biology Distributed Active Archive Center [OB.DAAC; @nasagoddardspaceflightcenterModerateresolutionImagingSpectroradiometer2022] respectively.
-Monthly summary datasets were downloaded for the period of 2009--2016 and averaged over each of the ARMS unit soaking periods (e.g.
+Monthly summary datasets were downloaded for the period of 2009--2016 and averaged over ARMS unit soaking periods (e.g.
 2009--2011, 2010--2012, etc.).
 Values (SST or chl-A) were extracted for the geographic coordinates of each sampling site from the averaged dataset for the relevant soaking period.
-Original raw monthly-average dataset names can be found in @tbl:nasa_table_s.
+Original raw monthly-average datasets can be retrieved from the NASA ocean data repository at [https://oceandata.sci.gsfc.nasa.gov/cgi/getfile/](https://oceandata.sci.gsfc.nasa.gov/cgi/getfile/)`<data_filename>`, where data filenames are formatted as `'AQUA_MODIS.<YYYYmmdd_start>_<YYYYmmdd_end>.L3m.MO.CHL.chlor_a.4km.nc'` or `'AQUA_MODIS.<YYYYmmdd_start>_<YYYYmmdd_end>.L3m.MO.SST.sst.4km.nc'` (NASA ocean data repository login credentials are required).
 
@@
 Human impacts datasets were downloaded for each available year in this study (2009--2013) and averaged over ARMS unit soaking periods (as above).
-Since cumulative impacts data was only available through 2013, we used the combined average of all available years for the 2013--2016 soaking period.
-Since the availability of data points at individual coordinates was highly variable, we averaged impact scores for each sampled island in the study.
+Cumulative impacts data was only available through 2013, so we used the combined average of all available years for the 2013--2016 soaking period.
+Additionally, as the availability of data points at individual sampling site coordinates was highly variable, we averaged impact scores across sites within islands.
 
@@
 
-```{r nasa-table,results="asis",depenson="init"}
-nasa_table <- read_csv(here("data","tables","nasa.csv"))
-print_table(nasa_table,"nasa_table_s","NASA data repository filenames for CHL and SST satellite data",accuracy=1,hdr_case=NA,format_numbers=FALSE) 
-```
-
-\newpage
-
 ## References

### 2024-11-22 manuscript.Rmd
remove TODO

@@
 ---
-<!--
-TODO: italics in tables S4/S5
--->
 <!--word count: Proceedings B publishes articles up to 12 pages, including figures and tables, with no more than 10 000 words.

### 2025-03-12 manuscript.Rmd
initial revisions, abstract overhaul

@@
 
-An inability to consistently resolve the environmental and anthropogenic factors that shape coral reefs impacts our understanding of how those factors interact and how coral reef communities may respond to change.
-Logistical challenges have additionally limited the taxonomic resolution and depth ranges investigated, with previous studies focusing largely on fishes and corals in the shallowest ~30 m of coral reefs.
-Here, we examine the effects of a suite of environmental, ecological, and anthropogenic factors on the assemblage structure of brachyuran crabs recovered from Autonomous Reef Monitoring Structures on shallow reefs (`r dr(cc$crabs_shallow_untransformed)` m) across the ~2,500 km span of the Hawaiian Archipelago.
-We compare these assemblages with those sampled along a depth gradient (`r format_range(range(cc$crab_data$depth[cc$crab_data$island == "Oahu"]))` m) on O&#x02BB;ahu.
-<!--We recovered `r num(sum(sample_sums(cc$crabs_untransformed)),1)` individuals representing `r sum(unlist(all_summary$types_breakdown))` taxa from `r all_summary$taxonomy_breakdown$families` families.-->Shallow assemblages were significantly different between the Main and Northwest Hawaiian Islands and significantly associated with human impacts, chlorophyll-A, sea-surface temperature, slope, potential larval immigration, and depth.
-However, we found more variation among mesophotic samples from O&#x02BB;ahu (30--90 m) than among shallow samples across the entire archipelago, suggesting that 90 m of depth is a stronger determinant of brachyuran assemblage structure than the latitudinal, environmental, or anthropogenic gradients across the Hawaiian Islands.
+Investigations into the environmental and anthropogenic factors that shape coral reef communities have been limited in both their depth range and taxonomic scope.
+Despite high biodiversity and the known importance of mesophotic habitats, due to logistical challenges, previous studies have focused largely on fishes and corals and the shallowest ~30 m of their depth ranges.
+Here, we examine the variability patterns of brachyuran crabs recovered from Autonomous Reef Monitoring Structures across the ~2,500 km Hawaiian Archipelago and ~90 m of depth.
+We examine the effects of a suite of environmental, ecological, and anthropogenic factors on shallow (8--17 m) crab assemblages and compare them with those sampled along a shallow-to-mesophotic depth gradient (12--90 m) on O&#x02BB;ahu.
+Shallow assemblages were significantly different between the Main and Northwest Hawaiian Islands and significantly associated with human impacts, chlorophyll-A, sea-surface temperature, island slope, the potential for larval immigration, and depth.
+However, we found more variation along the depth gradient on O&#x02BB;ahu than among the shallow assemblages across the entire archipelago.
+This finding suggests that 90 m of depth is a stronger determinant of brachyuran assemblage structure than the latitudinal, environmental, or anthropogenic gradients across the entire span of the Hawaiian Islands.
 
@@
 We targeted brachyuran crabs, a ubiquitous group often used as an exemplar to assess coral reef biodiversity [@knowltonCoralReefBiodiversity2010; @plaisanceDiversityCoralReefs2011; @servisCharacterizingCoralReef2020; @plaisanceReefassociatedCrustaceanFauna2009] and comprising a notable proportion of the Hawaiian coral reef cryptofauna [284 reported species, @castroCatalogAnomuranBrachyuran2011].
-We used assemblages recovered from ARMS units to examine drivers of community composition across the Hawaiian Archipelago, compared crab communities among islands and regions (MHI vs.
-NWHI), and examined the effects of a suite of possible environmental and anthropogenic drivers of community structure, including depth, to test hypotheses about their relative importance.
+We used assemblages recovered from ARMS units to examine drivers of community composition across the Hawaiian Archipelago.
+We compared crab communities among islands and regions (MHI vs.
+NWHI), among major depth zones (shallow vs.
+mesophotic), and examined the effects of a suite of possible environmental and anthropogenic drivers of community structure on shallow reefs (including fine-scale depth differences), to test hypotheses about their relative importance.
 
@@
 Once returned to the surface, they were placed in containers of aerated, filtered seawater, crates were removed, and units were systematically disassembled plate-by-plate [see @lerayDNABarcodingMetabarcoding2015 for processing details].
-All adult brachyuran crabs ($\geq$ 2 mm) were photographed, identified, and preserved in 95% ethanol.
+All adult brachyuran crabs ($\geq$ 2 mm) were photographed, identified, preserved in 95% ethanol, and retained in a specimen archive at the Hawai&#x02BB;i Institute of Marine Biology.
 Specimens were initially identified to family or genus using Poore & Ahyong [-@pooreMarineDecapodCrustacea2023] and additionally to species where possible according to expert advice from Scott Godwin (former curator of invertebrate zoology, Bishop Museum) and Gustav Paulay (curator of invertebrate zoology, Florida Museum of Natural History).
@@
 However, the prevalence of species complexes and cryptic diversity suggests that our assignments are likely to under- rather than overrepresent the true number of taxa.
-Thus, any significant patterns we detected likely persist in spite of the potential presence of cryptic diversity.
+Thus, any significant patterns we detected likely persist in spite of the potential presence of cryptic diversity, and we have retained species names in cases where complexes are suspected.
 
@@
 
-Even though our shallow-water ARMS units were deployed within a narrow depth range (`r dr(cc$crabs_shallow_untransformed)` m), depth still emerged as a significant and consistent driver of community variation among these units.
+Although our shallow-water ARMS units were deployed within a narrow depth range (`r dr(cc$crabs_shallow_untransformed)` m), depth still emerged as a significant and consistent driver of community variation among these units.
+Similar fine-scale depth patterns have been found in previous studies, both using molecular [e.g., @desouzaImportanceDepthTemperature2023; @westEDNAMetabarcodingSurvey2020] and traditional survey methods [e.g., @williamsBenthicCommunitiesTwo2013; @srinivasanDepthDistributionsCoral2003], but our results recover them alongside multiple other geographic and environmental gradients.
 Depth in and of itself is unlikely to affect marine community composition and oceanographic factors with well-known depth profiles (e.g., temperature, oxygen, nutrients, etc.) are not expected to vary dramatically across such a narrow depth range (\< 10 m), although internal waves can deliver variable temperatures across shallow reefs [@storlazziInternalTidesCan2020; @wyattHeatAccumulationCoral2020] and fine-scale habitat variability among sites may also influence the community differences we found [@steyaertRemoteReefCryptobenthic2022].
@@
 
-While depth was a significant driver of community composition among shallow-water ARMS units ([@fig:fig_dbrda; @tbl:rda_table_s]), its influence was even more striking when comparing shallow to mesophotic samples.
-Due to logistical constraints (and the limitations of satellite-derived data products), our environmental data only covers shallow-water deployments, but CAP ordinations show considerable differences between the grouping of all shallow communities and the units deployed $\geq$ 30 m at the single island of O&#x02BB;ahu ([@fig:fig_cap]).
-According to these analyses, there was more community variation among samples spanning 90 m of depth on a single island than across 2,500 km of shallow habitat experiencing multiple gradients of environmental, ecological, and anthropogenic influence.
-The ordination revealed two distinct clusters for mesophotic samples: one comprising the 30 m samples and the other the 60 m and 90 m samples.
-These two clusters coincide with what is regarded as the upper and lower mesophotic zones, respectively [@kahngCommunityEcologyMesophotic2010; @rochaMesophoticCoralEcosystems2018; but see @pyleMesophoticCoralEcosystems2019].
-The upper mesophotic (~30--60 m) has often been considered an extension of shallow reefs while the lower mesophotic (~60--150 m) is primarily thought to constitute a unique community.
-This pattern has held true across various taxonomic groups [@lesserGlobalCommunityBreaks2019; @rochaMesophoticCoralEcosystems2018], but our data show more distinct separations among shallow, upper, and lower mesophotic depths (although 30 m appears to encompass a transition zone).
+While depth was a significant driver of community composition among shallow-water ARMS units ([@fig:fig_dbrda; @tbl:rda_table_s]), its influence was considerably more striking when comparing shallow to mesophotic samples.
+In our CAP ordination, all shallow communities across the archipelago clustered separately from the deep samples $\geq$ 30 m at the single island of O&#x02BB;ahu ([@fig:fig_cap]).
+Additionally, the two deepest sampled depths (60 m and 90 m) clustered apart from 30 m.
+According to this analysis, there was more community variation among samples spanning 90 m of depth on a single island than across ~2,500 km of shallow habitat experiencing multiple gradients of environmental, ecological, and anthropogenic influence.
+Shallow communities from O&#x02BB;ahu share more affinity with Hōlanikū (~2,200 km away) than they do with O&#x02BB;ahu communities at 90 m depth ([@fig:fig_cap]).
+The ordination also revealed two distinct groups within the mesophotic samples: one comprising the 30 m samples and the other the 60 m and 90 m samples.
+These two groups correspond with the upper and lower mesophotic zones, respectively [@kahngCommunityEcologyMesophotic2010; @rochaMesophoticCoralEcosystems2018; but see @pyleMesophoticCoralEcosystems2019].
+ARMS units from the upper mesophotic zone (~30--45 m), which has often been considered an extension of shallow reefs, lie closer to shallow O&#x02BB;ahu units.
+Conversely, the lower mesophotic (~60--150 m), which is primarily thought to constitute a unique community, can be seen in the more tightly- and separately-grouped cluster of units from 60 m and 90 m.
+This pattern of community transition from shallow to mesophotic reefs is consistent with what has been shown for various other taxonomic groups [@lesserGlobalCommunityBreaks2019; @rochaMesophoticCoralEcosystems2018], lending support to the hypothesis that there may be broad-scale depth-based assembly rules for coral reef communities.
 
@@
 In that study, communities were found to have more affinity within depth zones across sites than among depth zones within sites.
-These results and the patterns we observed suggest that depth-associated factors exert stronger ecological influence than those that vary across the latitudinal, physical, and anthropogenic gradients throughout the Hawaiian Archipelago.
+These results and the patterns we observed suggest that factors associated with the shallow-to-mesophotic depth gradient exert stronger ecological influence than those (including depth) that vary across the latitudinal, physical, and anthropogenic gradients throughout the Hawaiian Archipelago.
 
@@
 By comparing shallow-water ARMS across the latitudinal, physical and anthropogenic gradients of the Hawaiian Archipelago, we were able to test the relative importance of the significant drivers of community structure.
-We additionally find that a 90 m depth transect at a single island shows greater divergence in canonical space than the full 2,500 km gradient of the Hawaiian Archipelago for shallow-water ARMS.
+We additionally find that a 90 m depth transect at a single island shows greater divergence in canonical space than the full ~2,500 km gradient of the Hawaiian Archipelago for shallow-water ARMS.
 Our analyses corroborate previous work and show that depth is among the strongest drivers of community structure.

### 2025-03-12 manuscript.Rmd
little abstract tweaks

@@
 
-Investigations into the environmental and anthropogenic factors that shape coral reef communities have been limited in both their depth range and taxonomic scope.
-Despite high biodiversity and the known importance of mesophotic habitats, due to logistical challenges, previous studies have focused largely on fishes and corals and the shallowest ~30 m of their depth ranges.
+Investigations into the environmental and anthropogenic factors that shape coral reef communities have historically been limited in both depth range and taxonomic scope.
+Despite high biodiversity and the recognized importance of mesophotic habitats and due to logistical challenges, previous studies have focused largely on fishes and corals and the shallowest ~30 m of their depth ranges.
 Here, we examine the variability patterns of brachyuran crabs recovered from Autonomous Reef Monitoring Structures across the ~2,500 km Hawaiian Archipelago and ~90 m of depth.

### 2025-03-12 manuscript.Rmd
add depth ranges to table 1

@@
   sample_tibble() %>%
-  # separate_wider_regex(sample,patterns=c(recovery_year="^[0-9]+","_",sample=".*"))  %>%
-  # select(sample,everything()) %>%
   unite("range",deployment_year,recovery_year,sep="--") %>%
@@
   ) %>%
-  count(island,range) %>%
+  group_by(island) %>%
+  mutate(depth = str_glue("{min(depth)}--{max(depth)} m")) %>%
+  ungroup() %>%
+  group_by(island,range) %>%
+  summarise( n = n(), depth = first(depth) ) %>%
+  ungroup() %>%
   pivot_wider(names_from="range",values_from="n")  %>%
-  select(island,order(colnames(.))) %>%
-  mutate(across(-island,~replace_na(.x,0)))
-print_table(arms_schedule,"arms_table","ARMS unit deployment schedule by island (number of units)",accuracy=1)
+  mutate(
+    across(-c(island,depth),~replace_na(.x,0)),
+    across(
+      everything(),
+      ~if_else(str_detect(island,'deep'),str_c("**",.x,"**"),as.character(.x))
+    )
+  ) %>%
+  select(island,`Depth range`=depth,order(colnames(.)))
+print_table(arms_schedule,"arms_table","ARMS unit deployment schedule by island (number of units). Mesophotic deployments apear in bold.",accuracy=1)
 ```

### 2025-03-14 manuscript.Rmd
reduced CAP analysis & extra cite to hurley 2016

@@
 `r str_to_sentence(english(spp['shallow']))` distinct taxa (including those collapsed to genus or family) were exclusive to shallow ARMS units and `r spp['deep']` were found only deep (`r format_list(deep_spp)`; [@tbl:species_depth_table_s]).
+For a more in-depth taxonomic treatment of the assemblage composition of mesophotic vs.
+shallow ARMS units, see Hurley et al.
+[-@hurleyAssessmentShallowMesophotic2016].
 
@@
 Although not a significant factor in our model, coral cover and species composition are thought to affect patterns of benthic assemblages.
-Due to the limited availability of datasets covering our entire study area, our model only included coral cover at the island scale, but previous work has shown that coral species composition does not vary considerably across the archipelago [@griggCommunityStructureSuccession1983].
+Due to the limited availability of datasets covering our entire study area, our model only included coral cover at the island scale, but previous work has shown that coral species composition does not vary considerably across the archipelago (aside from the presence of *Acropora* in the NWHI) [@griggCommunityStructureSuccession1983].
 Crab communities from Lalo (French Frigate Shoals), which was found to harbor the highest coral species diversity [@griggCommunityStructureSuccession1983], did not group separately from the rest of our samples ([@fig:fig_cap]).
@@
 Although our shallow-water ARMS units were deployed within a narrow depth range (`r dr(cc$crabs_shallow_untransformed)` m), depth still emerged as a significant and consistent driver of community variation among these units.
-Similar fine-scale depth patterns have been found in previous studies, both using molecular [e.g., @desouzaImportanceDepthTemperature2023; @westEDNAMetabarcodingSurvey2020] and traditional survey methods [e.g., @williamsBenthicCommunitiesTwo2013; @srinivasanDepthDistributionsCoral2003], but our results recover them alongside multiple other geographic and environmental gradients.
+Similar fine-scale depth patterns have been found in previous studies, both using molecular [e.g., @desouzaImportanceDepthTemperature2023; @westEDNAMetabarcodingSurvey2020] and traditional survey methods [e.g., @williamsBenthicCommunitiesTwo2013; @srinivasanDepthDistributionsCoral2003], but our results recover them alongside numerous other geographic and environmental gradients.
 Depth in and of itself is unlikely to affect marine community composition and oceanographic factors with well-known depth profiles (e.g., temperature, oxygen, nutrients, etc.) are not expected to vary dramatically across such a narrow depth range (\< 10 m), although internal waves can deliver variable temperatures across shallow reefs [@storlazziInternalTidesCan2020; @wyattHeatAccumulationCoral2020] and fine-scale habitat variability among sites may also influence the community differences we found [@steyaertRemoteReefCryptobenthic2022].
@@
 Shallow communities from O&#x02BB;ahu share more affinity with Hōlanikū (~2,200 km away) than they do with O&#x02BB;ahu communities at 90 m depth ([@fig:fig_cap]).
+One potential weakness of our dataset is the broad timescale over which ARMS units were deployed ([@tbl:arms_table]).
+However, the majority of our ARMS units were deployed over the periods of 2010--2012 or 2010--2013, with an overlap of two years.
+A reduced version of the CAP ordination using only those overlapping deployments retains the same overall patterns as the full ordination, with the primary difference being the more pronounced separation among mesophotic assemblages ([@fig:fig_cap_reduced]).
+
 The ordination also revealed two distinct groups within the mesophotic samples: one comprising the 30 m samples and the other the 60 m and 90 m samples.
@@
 ### Processing of satellite-derived environmental data
-Level-3 mapped (4 km grid-size) sea-surface temperature (SST) and chlorophyll-A (chl-A) data were retrieved from NASA's Physical Oceanography Distributed Active Archive Center [PO.DAAC; @nasaobpgMODISAquaGlobal2020] and Ocean Biology Distributed Active Archive Center [OB.DAAC; @nasagoddardspaceflightcenterModerateresolutionImagingSpectroradiometer2022] respectively.
+Level-3 mapped (4 km grid-size) sea-surface temperature (SST) and chlorophyll-A (chl-A) data were retrieved from NASA's Physical Oceanography Distributed Active Archive Center (PO.DAAC [@nasaobpgMODISAquaGlobal2020]) and Ocean Biology Distributed Active Archive Center (OB.DAAC [@nasagoddardspaceflightcenterModerateresolutionImagingSpectroradiometer2022]) respectively.
 Monthly summary datasets were downloaded for the period of 2009--2016 and averaged over ARMS unit soaking periods (e.g.
@@
 cat("{#fig:fig_pairwise}")
+```
+
+\newpage
+
+```{r cap-analysis-reduced,dependson="init",echo=FALSE,message=FALSE,warning=FALSE,results="hide"}
+# reduce crab data to only those deployed from 2010--2013
+crabs_reduced <- prune_samples(
+  with(sample_tibble(cc$crabs),deployment_year == 2010 & recovery_year %in% c(2012,2013)),
+  cc$crabs
+)
+# get sample data
+crabs_reduced_data <- sample_tibble(crabs_reduced)
+
+# get otu table
+otus <- otu_table(crabs_reduced) %>%
+  as.data.frame()
+# get sample data
+cd <- crabs_reduced_data %>%
+  mutate(sample_group = factor(sample_group,ordered=FALSE)) %>%
+  as.data.frame()
+
+# run CAP analysis, taking care of negative eigenvalues where necessary
+cap_depth <- CAPdiscrim(otus ~ sample_group, data=cd, dist="bray", axes=2, m=0, add=TRUE)
+# # calculate proportion of explained variance
+cap_var <- cap_depth$manova$Eigenvalues / sum(cap_depth$manova$Eigenvalues)
+cap_var <- list(x = cap_var[1], y = cap_var[2])
+```
+```{r fig-cap-reduced, fig.cap="Reduced dataset CAP ordination of crab communities by island group and depth zone.
+Samples were deployed only during the years 2010--2013", results="asis", dev="svg", fig.width=9, fig.height=6, dependson="cap-analysis-reduced"}
+# make the sample group look nice
+nice <- function(s) {
+  str_replace(s,"shallow","Shallow") %>%
+    str_replace("main","(MHI)") %>%
+    str_replace("nwhi","(NWHI)") %>%
+    str_replace("(^[0-9]+)","O‘ahu \\1") %>%
+    str_split("((?<=[0-9])(?=[a-zA-Z]))|(_)") %>%
+    map_chr(~str_c(.x,collapse=" "))
+}
+
+# create tibble of cap results and join in sample data
+pcap_reduced <- cap_depth$x %>%
+  as_tibble(rownames="sample") %>%
+  left_join(cd,by="sample") %>%
+  mutate(
+    sample_group = nice(sample_group),
+    sample_group = fct_reorder(sample_group,depth)
+  )
+
+# get centroids of each point cluster
+centroids_reduced <- pcap_reduced %>% group_by(sample_group) %>%
+  summarise(LD1=mean(LD1),LD2=mean(LD2))
+# make an oahu/not oahu factor
+pcap_reduced <- pcap_reduced %>%
+  mutate(
+    oahu = factor(
+      case_when(
+        island == "Oahu" ~ "oahu",
+        # island == "French Frigate Shoals" ~ "ffs",
+        .default = "noahu"
+      ),
+      levels = c("oahu","noahu")
+    )
+  )
+
+# plot the thing
+ggplot(pcap_reduced,aes(x=LD1,y=LD2,fill=sample_group)) +
+  stat_ellipse(geom="polygon",level=0.95,alpha=0.1,show.legend = FALSE) +
+  geom_point(mapping=aes(shape=oahu),color="black",size=5)  +
+  geom_label_repel(data=centroids_reduced,aes(label=sample_group), color="white", show.legend = FALSE) +
+  # scale_shape_manual(values=c("oahu" = 23, "ffs" = 22, "noahu" = 21),name="Island", labels=c("noahu"="Other islands","oahu"="O‘ahu","ffs"="Lalo (French Frigate Shoals)" )) +
+  scale_shape_manual(values=c("oahu" = 23, "noahu" = 21),name="Island", labels=c("noahu"="Other islands","oahu"="O‘ahu")) +
+  scale_fill_manual(values=sample_depth_pal,name="Sample group") +
+  guides(
+    fill=guide_legend(override.aes=list(shape=21)),
+    shape=guide_legend(override.aes=list(fill="black"))
+  ) +
+  xlab(str_glue("Linear discriminant 1 [{pc(cap_var$x,0.1)}]")) +
+  ylab(str_glue("Linear discriminant 2 [{pc(cap_var$y,0.1)}]")) +
+  plotz_theme()
+cat("{#fig:fig_cap_reduced}")
 ```

### 2025-03-14 manuscript.Rmd
more revision in the discussion

@@
 While depth was a significant driver of community composition among shallow-water ARMS units ([@fig:fig_dbrda; @tbl:rda_table_s]), its influence was considerably more striking when comparing shallow to mesophotic samples.
-In our CAP ordination, all shallow communities across the archipelago clustered separately from the deep samples $\geq$ 30 m at the single island of O&#x02BB;ahu ([@fig:fig_cap]).
+In our CAP ordination, all shallow assemblages across the archipelago clustered separately from the deep assemblages $\geq$ 30 m at the single island of O&#x02BB;ahu ([@fig:fig_cap]).
 Additionally, the two deepest sampled depths (60 m and 90 m) clustered apart from 30 m.
 According to this analysis, there was more community variation among samples spanning 90 m of depth on a single island than across ~2,500 km of shallow habitat experiencing multiple gradients of environmental, ecological, and anthropogenic influence.
-Shallow communities from O&#x02BB;ahu share more affinity with Hōlanikū (~2,200 km away) than they do with O&#x02BB;ahu communities at 90 m depth ([@fig:fig_cap]).
-One potential weakness of our dataset is the broad timescale over which ARMS units were deployed ([@tbl:arms_table]).
-However, the majority of our ARMS units were deployed over the periods of 2010--2012 or 2010--2013, with an overlap of two years.
-A reduced version of the CAP ordination using only those overlapping deployments retains the same overall patterns as the full ordination, with the primary difference being the more pronounced separation among mesophotic assemblages ([@fig:fig_cap_reduced]).
-
-The ordination also revealed two distinct groups within the mesophotic samples: one comprising the 30 m samples and the other the 60 m and 90 m samples.
-These two groups correspond with the upper and lower mesophotic zones, respectively [@kahngCommunityEcologyMesophotic2010; @rochaMesophoticCoralEcosystems2018; but see @pyleMesophoticCoralEcosystems2019].
-ARMS units from the upper mesophotic zone (~30--45 m), which has often been considered an extension of shallow reefs, lie closer to shallow O&#x02BB;ahu units.
-Conversely, the lower mesophotic (~60--150 m), which is primarily thought to constitute a unique community, can be seen in the more tightly- and separately-grouped cluster of units from 60 m and 90 m.
+This is further supported by the fact that average Euclidean distance in the CAP among O&#x02BB;ahu samples (including mesophotic) was significantly greater than the average distance among all shallow samples (MHI & NWHI).
+Shallow crab communities from O&#x02BB;ahu appear to share more affinity with Hōlanikū (~2,200 km distant) than they do with those at 90 m depth on the same reef slope ([@fig:fig_cap]).
+The CAP also revealed two groups within the mesophotic samples: one at 30 m and the other encompassing the 60--90 m depths.
+These depth zones correspond with the upper and lower mesophotic zones, respectively [@kahngCommunityEcologyMesophotic2010; @rochaMesophoticCoralEcosystems2018; but see @pyleMesophoticCoralEcosystems2019].
+The upper mesophotic zone has been found to constitute an extension of shallow reefs (with beta diversity patterns largely driven by nestedness), whereas the lower mesophotic harbors unique communities (with beta diversity driven primarily by turnover) [@rochaMesophoticCoralEcosystems2018; @slatteryFunctionStabilityMesophotic2024a; @bellwoodRegionalscaleAssemblyRules2001; @kahngCommunityEcologyMesophotic2010].
+The evident groupings in the CAP ordination match these trends, with 30 m samples lying closer to shallow O&#x02BB;ahu samples and 60--90 m samples forming a distinct cluster.
 This pattern of community transition from shallow to mesophotic reefs is consistent with what has been shown for various other taxonomic groups [@lesserGlobalCommunityBreaks2019; @rochaMesophoticCoralEcosystems2018], lending support to the hypothesis that there may be broad-scale depth-based assembly rules for coral reef communities.
@@
 
-Our depth transect was limited to O&#x02BB;ahu, but community and alpha diversity differences among depth zones at that single island appear to overwhelm differences observed among shallow-water coral reef habitat across the archipelago.
-Similar patterns were found at a local scale (tens of kilometers) on Hawai&#x02BB;i Island using environmental DNA (eDNA) [@hobanPlumbingDepthsEnvironmental2023].
-In that study, communities were found to have more affinity within depth zones across sites than among depth zones within sites.
-These results and the patterns we observed suggest that factors associated with the shallow-to-mesophotic depth gradient exert stronger ecological influence than those (including depth) that vary across the latitudinal, physical, and anthropogenic gradients throughout the Hawaiian Archipelago.
-
-Since we lack environmental data for mesophotic samples, we cannot make conclusions about the specific influences driving the differences we found.
+Since we lack environmental data at mesophotic depths, we cannot make conclusions about the specific influences driving the differences we found.
 However, light and temperature have been proposed as important bottom-up drivers of depth zonation in benthic communities [@kahngCommunityEcologyMesophotic2010; @lesserGlobalCommunityBreaks2019; @pyleComprehensiveInvestigationMesophotic2016; @pyleMesophoticCoralEcosystems2019].
@@
 In contrast, the lower depth boundaries of mesophotic ecosystems are largely defined by surface irradiance [@kahngCommunityEcologyMesophotic2010; @hindersteinThemeSectionMesophotic2010], so light seems a possible candidate as bottom-up abiotic factor influencing community structure across depths.
-Crab communities are unlikely to be directly affected by light variability.
-However, habitat-building species such as corals, algae, and sponges may be [@kahngCommunityEcologyMesophotic2010; @slatteryFunctionStabilityMesophotic2024a], and may in turn affect the structure of the communities dependent upon them [@hustonCompetitionCoexistenceEffects1994; @smithRoleHabitatComplexity2014; @sebensHabitatStructureCommunity1991].
+Crab assemblage structure is unlikely to be directly affected by light variability (although the behavior of individual crabs may).
+However, habitat-building species such as corals, algae, and sponges frequently are subject to light requirements [@kahngCommunityEcologyMesophotic2010; @slatteryFunctionStabilityMesophotic2024a], and may in turn affect the structure of the communities dependent upon them [@hustonCompetitionCoexistenceEffects1994; @smithRoleHabitatComplexity2014; @sebensHabitatStructureCommunity1991].
 If patterns of community structure are driven primarily from the bottom up by light, that may contribute to our observation of a significant depth effect over less than 10 m in our shallow samples.
+
+Two potential weaknesses of our dataset are the time periods over which ARMS units were deployed and the fact that mesophotic samples were obtained solely from the island of O&#x02BB;ahu.
+ARMS units were deployed across a broad range of years, but the majority were deployed during 2010--2012 or 2010--2013 (an overlap of two years) ([@tbl:arms_table]).
+A reduced version of the CAP using only these overlapping deployments retains the same (if not more pronounced) overall patterns as the full version ([@fig:fig_cap_reduced]), suggesting that the inclusion of previous or subsequent soaking periods did not adversely skew our analyses.
+The limitation of our depth transect to O&#x02BB;ahu means that we cannot make broad-scale conclusions about biodiversity patterns on deep reefs across the Hawaiian Islands.
+However, it remains the case that community and alpha diversity differences among depth zones at that single island appear to overwhelm differences observed among shallow-water coral reef habitat across the archipelago.
+Similar patterns were found at a local scale (tens of kilometers) on Hawai&#x02BB;i Island using environmental DNA (eDNA) [@hobanPlumbingDepthsEnvironmental2023].
+In that study, communities were found to have more affinity within depth zones across sites than among depth zones within sites.
+Thus, despite the limitations in our dataset, the patterns we observed suggest that factors associated with the shallow-to-mesophotic depth gradient exert stronger ecological influence than those (including depth) that vary across the latitudinal, physical, and anthropogenic gradients throughout the Hawaiian Archipelago.
 
@@
 Such differences should be tested in future studies to understand the specific physical and ecological drivers of coral reef depth zonation.
-The logistical complexity and time investment required for ARMS deployments make this challenging, but emerging approaches such as environmental DNA (eDNA) may offer a suitable alternative.
+The cost, logistical complexity and time investment required for ARMS deployments make this challenging, but emerging approaches such as environmental DNA (eDNA) may offer a suitable alternative.
 
@@
 Our best knowledge of the role of depth in structuring marine communities comes from the intertidal, and points to biotic factors (e.g., predation, competition) determining lower limits and abiotic factors (e.g., desiccation, thermal stress) setting upper limits of vertical species distributions [@connellEffectsCompetitionPredation1961; @connellInfluenceInterspecificCompetition1961; @dennyPhysicalProcessesThat2001; @paineIntertidalCommunityStructure1974].
-There are numerous examples from other study systems in which the relative contributions of biotic versus abiotic factors to structuring communities run in opposing directions across some habitat gradient [e.g., climate, soil type, elevation, etc.
-@bradyEvolutionaryEcologyPlant2005; @moore10PlantCompetition2011; @wishingradTemperateZoneIsolation2023].
+There are numerous examples from other study systems in which the relative contributions of biotic versus abiotic factors to structuring communities run in opposing directions across some habitat gradient (e.g., climate, soil type, elevation, etc.
+[@bradyEvolutionaryEcologyPlant2005; @moore10PlantCompetition2011; @wishingradTemperateZoneIsolation2023]).
 However, abiotic stressors on intertidal communities are obviously different from those on coral reefs, and generalizations about the mechanisms and ecological processes that structure coral reef communities remain a subject of considerable debate [@bellwoodEnvironmentalGeometricConstraints2005; @bellwoodSleepingFunctionalGroup2006; @carrBiodiversityPopulationRegulation2002; @connollyIndoPacificBiodiversityCoral2003; @dornelasCoralReefDiversity2006; @grahamImportanceStructuralComplexity2013; @komyakovaRelativeImportanceCoral2013].
@@
 
-By comparing shallow-water ARMS across the latitudinal, physical and anthropogenic gradients of the Hawaiian Archipelago, we were able to test the relative importance of the significant drivers of community structure.
-We additionally find that a 90 m depth transect at a single island shows greater divergence in canonical space than the full ~2,500 km gradient of the Hawaiian Archipelago for shallow-water ARMS.
+By comparing shallow-water ARMS across the latitudinal, physical and anthropogenic gradients of the Hawaiian Archipelago, we were able to test the relative importance of these factors in driving shallow crab assemblage structure.
+We additionally find that assemblages alkong a 90 m depth transect at a single island show greater divergence in canonical space than across the full ~2,500 km gradient of the Hawaiian Archipelago on shallow reefs.
 Our analyses corroborate previous work and show that depth is among the strongest drivers of community structure.
@@
 Hee, for discussion, support, and their efforts in the disassembly and sorting of ARMS samples on retrieval.
-Finally, we are grateful to two anonymous reviewers for constructive comments and suggestions on an earlier version of this paper.
 This is contribution #xxx from the Hawai&#x02BB;i Institute of Marine Biology and #xxx from the School of Ocean and Earth Science and Technology at the University of Hawai&#x02BB;i.

### 2025-03-14 manuscript.Rmd
little typos

@@
 Crab assemblage structure is unlikely to be directly affected by light variability (although the behavior of individual crabs may).
-However, habitat-building species such as corals, algae, and sponges frequently are subject to light requirements [@kahngCommunityEcologyMesophotic2010; @slatteryFunctionStabilityMesophotic2024a], and may in turn affect the structure of the communities dependent upon them [@hustonCompetitionCoexistenceEffects1994; @smithRoleHabitatComplexity2014; @sebensHabitatStructureCommunity1991].
+However, habitat-building species such as corals, algae, and sponges are frequently subject to light requirements [@kahngCommunityEcologyMesophotic2010; @slatteryFunctionStabilityMesophotic2024a], and may in turn affect the structure of the communities dependent upon them [@hustonCompetitionCoexistenceEffects1994; @smithRoleHabitatComplexity2014; @sebensHabitatStructureCommunity1991].
 If patterns of community structure are driven primarily from the bottom up by light, that may contribute to our observation of a significant depth effect over less than 10 m in our shallow samples.
@@
 By comparing shallow-water ARMS across the latitudinal, physical and anthropogenic gradients of the Hawaiian Archipelago, we were able to test the relative importance of these factors in driving shallow crab assemblage structure.
-We additionally find that assemblages alkong a 90 m depth transect at a single island show greater divergence in canonical space than across the full ~2,500 km gradient of the Hawaiian Archipelago on shallow reefs.
+We additionally find that assemblages along a 90 m depth transect at a single island show greater divergence in canonical space than across the full ~2,500 km gradient of the Hawaiian Archipelago on shallow reefs.
 Our analyses corroborate previous work and show that depth is among the strongest drivers of community structure.

### 2025-03-14 manuscript.Rmd
updates to abstract and supplemental CAP caption

@@
 
-Investigations into the environmental and anthropogenic factors that shape coral reef communities have historically been limited in both depth range and taxonomic scope.
+Investigations into the environmental and anthropogenic factors that shape coral reef communities have been historically limited both in depth range and taxonomic scope.
 Despite high biodiversity and the recognized importance of mesophotic habitats and due to logistical challenges, previous studies have focused largely on fishes and corals and the shallowest ~30 m of their depth ranges.
 Here, we examine the variability patterns of brachyuran crabs recovered from Autonomous Reef Monitoring Structures across the ~2,500 km Hawaiian Archipelago and ~90 m of depth.
-We examine the effects of a suite of environmental, ecological, and anthropogenic factors on shallow (8--17 m) crab assemblages and compare them with those sampled along a shallow-to-mesophotic depth gradient (12--90 m) on O&#x02BB;ahu.
-Shallow assemblages were significantly different between the Main and Northwest Hawaiian Islands and significantly associated with human impacts, chlorophyll-A, sea-surface temperature, island slope, the potential for larval immigration, and depth.
-However, we found more variation along the depth gradient on O&#x02BB;ahu than among the shallow assemblages across the entire archipelago.
+We examine the effects of a suite of environmental, ecological, and anthropogenic factors (latitude, longitude, site depth, deployment years, chlorophyll-A, sea surface temperature, island slope, coral cover, potential larval immigration, cumulative human impacts) on shallow (8--17 m) crab assemblages and compare them with those sampled along a shallow-to-mesophotic depth gradient (12--90 m) on O&#x02BB;ahu.
+Shallow assemblage structure was significantly different between the inhabited Main and relatively pristine Northwest Hawaiian Islands and significantly associated with human impacts, chlorophyll-A, sea-surface temperature, island slope, potential larval immigration, and depth.
+Despite this variation, community differences along the depth gradient on O&#x02BB;ahu alone were greater than among the shallow assemblages across the entire archipelago.
 This finding suggests that 90 m of depth is a stronger determinant of brachyuran assemblage structure than the latitudinal, environmental, or anthropogenic gradients across the entire span of the Hawaiian Islands.
@@
 cap_depth <- CAPdiscrim(otus ~ sample_group, data=cd, dist="bray", axes=2, m=0, add=TRUE)
-# # calculate proportion of explained variance
+# calculate proportion of explained variance
 cap_var <- cap_depth$manova$Eigenvalues / sum(cap_depth$manova$Eigenvalues)
@@
 ```
-```{r fig-cap-reduced, fig.cap="Reduced dataset CAP ordination of crab communities by island group and depth zone.
-Samples were deployed only during the years 2010--2013", results="asis", dev="svg", fig.width=9, fig.height=6, dependson="cap-analysis-reduced"}
+```{r fig-cap-reduced, fig.cap="CAP ordination of crab communities by island group and depth zone showing only samples deployed during the years 2010--2013.", results="asis", dev="svg", fig.width=9, fig.height=6, dependson="cap-analysis-reduced"}
 # make the sample group look nice
+# (possibly already defined above, but let's do it here just for good measure)
 nice <- function(s) {

### 2025-03-17 manuscript.Rmd
little updates and add quotes around suspected species complexes

@@
 
-A total of 89 ARMS units were deployed over the course of six years (2009--2016) in shallow (`r dr(cc$crabs_shallow_untransformed)` m depth) coral reefs in both the Main (MHI) and Northwest Hawaiian Islands (NWHI) ([@fig:fig_map; @tbl:arms_table]).
+A total of 89 ARMS units were deployed over the course of six years (2010--2016) in shallow (`r dr(cc$crabs_shallow_untransformed)` m depth) coral reefs in both the Main (MHI) and Northwest Hawaiian Islands (NWHI) ([@fig:fig_map; @tbl:arms_table]).
 All units included in this study were deployed in forereef habitat.
@@
 We recognize that the species-level taxonomy of Indo-Pacific brachyurans remains poorly resolved, particularly for abundant groups like the Chlorodiellinae [@lasleyjrIntegrativeSystematicRevision2014; @lasleyjrPhylogeneticRelationshipsUbiquitous2015], and some "species" are known to comprise complexes [e.g.
-*Perinia tumida*, @servisCharacterizingCoralReef2020; @plaisanceReefassociatedCrustaceanFauna2009].
+"*Perinia tumida*", @servisCharacterizingCoralReef2020; @plaisanceReefassociatedCrustaceanFauna2009].
 However, the prevalence of species complexes and cryptic diversity suggests that our assignments are likely to under- rather than overrepresent the true number of taxa.
-Thus, any significant patterns we detected likely persist in spite of the potential presence of cryptic diversity, and we have retained species names in cases where complexes are suspected.
+Thus, any significant patterns we detected likely persist in spite of the potential presence of cryptic diversity.
+Since our primary concern was categorical diversity of assemblages rather than accurate naming, we have retained species names in cases where complexes are suspected.
 
@@
 Continuous variables were scaled to unit variance.
-dbRDA analyses were conducted only for communities sampled from shallow (\< 30 m) ARMS units, since detailed driver variables were not available for deep ($\geq$ 30 m) samples.
+dbRDA analyses were conducted only for communities sampled from shallow (\< 30 m) ARMS units, since detailed environmental/ecological drivers were not available for deep ($\geq$ 30 m) samples.
 
@@
 Variation among island groups and sampling depths was further examined and visualized using canonical analysis of principal coordinates (CAP) based on discriminant analysis [@andersonCanonicalAnalysisPrincipal2003] as implemented in the `CAPdiscrim` function from `BiodiversityR` [@kindtTreeDiversityAnalysis2005].
-To explore differences in community variability among groups (shallow/deep, O&#x02BB;ahu/all shallow samples), we compared Euclidean distances among samples in the CAP ordination using Welch two-sample t-tests, assuming unequal variances.
+To explore differences in community variability among groups (shallow/deep, O&#x02BB;ahu/all shallow samples), we compared Euclidean distances among assemblage points in canonical space (from the CAP analysis) using Welch two-sample t-tests, assuming unequal variances.
 
@@
 For details regarding data import, preparation, and synthesis, see Supplemental material.
-Due to logistical limitations, environmental data were collected only for shallow samples (\< 30 m).
+Due to logistical and dataset limitations, environmental/ecological drivers were only available for shallow (\< 30 m) samples.
 
@@
 ![Representative photographs of the four most abundant crab species found across all deployed ARMS units (specimens shown were not collected as part of this study).
-(a) *Chlorodiella laevissima* (FLMNH 45490, Kāne&#x02BB;ohe, Hawai&#x02BB;i).
-(b) *Perinia tumida* (FLMNH 45652, Kāne&#x02BB;ohe, Hawai&#x02BB;i).
+(a) "*Chlorodiella laevissima*" (FLMNH 45490, Kāne&#x02BB;ohe, Hawai&#x02BB;i).
+(b) "*Perinia tumida*" (FLMNH 45652, Kāne&#x02BB;ohe, Hawai&#x02BB;i).
 (c) *Percnon abbreviatum* (FLMNH 45628, Kāne&#x02BB;ohe, Hawai&#x02BB;i).
@@
   # sum counts by island group, depth zone, and species
-  group_by(OTU,island_group,hawaiian_name,depth_zone,depth,family,genus,species) %>%
+  group_by(OTU,island_group,hawaiian_name,depth_zone,depth,family,genus,species,complex) %>%
   summarise(n=sum(Abundance)) %>%
@@
   left_join(crab_taxonomy_validated,by=c("OTU" = "otu")) %>%
-  select(hawaiian_name,island_group,depth_zone,depth,family.y,display_species,depth_exclusive) %>%
+  select(hawaiian_name,island_group,depth_zone,depth,family.y,display_species,depth_exclusive,complex=complex.y) %>%
+  mutate(display_species = as.character(case_when(
+    as.logical(complex) ~ str_glue('"{display_species}"'),
+    .default = display_species
+  ))) %>%
+  select(-complex) %>%
   rename(island = hawaiian_name,valid_species=display_species) %>%

### 2025-03-17 collapse_taxonomy.R

@@ -40,8 +40,8 @@ crabs_collapsed <- crabs %>%
   pivot_wider(names_from = "otu", values_from = "count", values_fn = sum)
 
 # make sure the taxonomy table is unique
-taxonomy_collapsed <- taxonomy_collapsed %>% select(otu,taxon_level:species) %>%
-  distinct(otu,taxon_level,family,genus,species)
+taxonomy_collapsed <- taxonomy_collapsed %>% select(otu,taxon_level:species,complex) %>%
+  distinct(otu,taxon_level,family,genus,species,complex)
 
 # save the new collapsed tables
 dir_create(here("data","generated"))

### 2025-03-17 crabs.R

@@ -374,9 +374,12 @@ sample_summary <- function(ps,top_n=5,taxonomy=NULL) {
   ss <- list() 
   
   # make otu lookup table
-  otu_lookup <- tax_table(ps) %>%
-    as.data.frame() %>%
-    as_tibble(rownames="otu") %>%
+  otu_lookup <- ps %>%
+    taxa_tibble(otu_col = "otu") %>%
+    mutate(species = as.character(case_when(
+      as.logical(complex) ~ str_glue('"{species}"'),
+      .default = species
+    ))) %>%
     select(otu,species) %>%
     deframe()
   
@@ -432,6 +435,7 @@ sample_summary <- function(ps,top_n=5,taxonomy=NULL) {
     slice(match(ss$top_otus,otu)) %>%
     mutate(
       disp = case_when(
+        as.logical(complex) & taxon_level == "species" ~ str_glue("\"{display_species}\" {authority}"),
         taxon_level == "species" ~ str_glue("{display_species} {authority}"),
         .default = display_species
       )

### 2025-03-17 manuscript.Rmd
strip no-longer-relevant comment

@@
 
-<!-- TODO: update number of different types to reflect family/genus/species level rather than species/morphospecies level -->
 The `r nsamples(cc$crabs_untransformed)` ARMS units deployed across the Hawaiian Islands (in `r format_range(cc$crab_data$depth)` m depth) yielded a total of `r num(sum(sample_sums(cc$crabs_untransformed)),1)` adult crabs.

### 2025-03-22 manuscript.Rmd
updates and reformat for sci. reports

@@
     keep_md: true
-tblPrefix: table
+tblPrefix: Table
 figPrefix:
-  - "figure"
-  - "figures"
+  - "Fig."
+  - "Figs."
 chapters: True
@@
 chapDelim: ""
+lofTitle: |
+	# Figure legends {-}
 header-includes:
@@
 ---
-<!--word count: Proceedings B publishes articles up to 12 pages, including figures and tables, with no more than 10 000 words.
-The word count includes the title page, author names, abstract, figure legends, tables and the reference list.
-As a rough guide, 10 000 words = 12 pages without additional figures; 850 words = 1 page and 2-4 small figures = 1 page.
-For example, an acceptable paper could be made up of 9 000 words and 4 small figures or 8500 words, 2 large figures and 4 small figures.-->
+<!--word count:
+Articles should ideally be no more than 11 typeset pages
+The main text should be no more than 4,500 words (not including Abstract, Methods, References and figure legends)
+The title should be no more than 20 words, should describe the main message of the article using a single scientifically accurate sentence, and should not contain puns or idioms
+The abstract should be no more than 200 words
+-->
 ### Keywords {-}
@@
 Investigations into the environmental and anthropogenic factors that shape coral reef communities have been historically limited both in depth range and taxonomic scope.
-Despite high biodiversity and the recognized importance of mesophotic habitats and due to logistical challenges, previous studies have focused largely on fishes and corals and the shallowest ~30 m of their depth ranges.
-Here, we examine the variability patterns of brachyuran crabs recovered from Autonomous Reef Monitoring Structures across the ~2,500 km Hawaiian Archipelago and ~90 m of depth.
+Despite high biodiversity and the recognized importance of mesophotic habitats and due to logistical challenges, previous studies have focused largely on conspicuous taxa such as fishes and corals and the shallowest ~30 m of their depth ranges.
+Here, we examine the variability patterns of brachyuran crab assemblages recovered from Autonomous Reef Monitoring Structures across the ~2,500 km Hawaiian Archipelago and ~90 m of depth.
 We examine the effects of a suite of environmental, ecological, and anthropogenic factors (latitude, longitude, site depth, deployment years, chlorophyll-A, sea surface temperature, island slope, coral cover, potential larval immigration, cumulative human impacts) on shallow (8--17 m) crab assemblages and compare them with those sampled along a shallow-to-mesophotic depth gradient (12--90 m) on O&#x02BB;ahu.
 Shallow assemblage structure was significantly different between the inhabited Main and relatively pristine Northwest Hawaiian Islands and significantly associated with human impacts, chlorophyll-A, sea-surface temperature, island slope, potential larval immigration, and depth.
-Despite this variation, community differences along the depth gradient on O&#x02BB;ahu alone were greater than among the shallow assemblages across the entire archipelago.
+Despite this spatial and environmental variation, community differences along the depth gradient on O&#x02BB;ahu alone were greater than among the shallow assemblages across the entire archipelago.
 This finding suggests that 90 m of depth is a stronger determinant of brachyuran assemblage structure than the latitudinal, environmental, or anthropogenic gradients across the entire span of the Hawaiian Islands.
@@
 
-# Materials and Methods {-}
+# Results {-}
+
+## Diversity summary {-}
+
+### All ARMS {-}
+
+The `r nsamples(cc$crabs_untransformed)` ARMS units deployed across the Hawaiian Islands (in `r format_range(cc$crab_data$depth)` m depth) yielded a total of `r num(sum(sample_sums(cc$crabs_untransformed)),1)` adult crabs.
+Units contained `r num(all_summary$abundance_mean,0.1)` $\pm$ `r num(all_summary$abundance_sd,0.1)` (mean $\pm$ SD) individuals, and abundance ranged from `r format_range(all_summary$abundance_range)`.
+Crabs were identified to `r sum(unlist(all_summary$types_breakdown))` distinct taxa (`r all_summary$types_breakdown$species` species-level identifications, `r all_summary$types_breakdown$genus` genus, `r all_summary$types_breakdown$family` family) encompassing `r all_summary$taxonomy_breakdown$genera` genera in `r all_summary$taxonomy_breakdown$families` families, which accounted for ~`r pc(all_summary$types_breakdown$species / hawaiian_crabs,0.1)` of the known crab diversity reported from Hawaiian waters for species-level IDs [@castroCatalogAnomuranBrachyuran2011].
+The four most abundant species were: `r with(all_summary,top_list(top,top_n,top_units,fig="fig_crabs"))`.
+There were `r all_summary$singleton_count` singleton taxa (taxa occurring in a single unit across all sites), accounting for ~`r pc(all_summary$singleton_count/ntaxa(cc$crabs_untransformed),0.1)` of total recovered crab taxa.
+For a breakdown of specific crab taxa recovered by island and deployment depth, see Supplementary [@tbl:tbl_supp_species].
+
+![Representative photographs of the four most abundant crab species found across all deployed ARMS units (specimens shown were not collected as part of this study).
+(a) "*Chlorodiella laevissima*" (FLMNH 45490, Kāne&#x02BB;ohe, Hawai&#x02BB;i).
+(b) "*Perinia tumida*" (FLMNH 45652, Kāne&#x02BB;ohe, Hawai&#x02BB;i).
+(c) *Percnon abbreviatum* (FLMNH 45628, Kāne&#x02BB;ohe, Hawai&#x02BB;i).
+(d) *Dynomene hispida* (FLMNH 69204, Manawai, Hawai&#x02BB;i).
+Photo credits: Florida Museum of Natural History, licensed [CC BY-NC 3.0](https://creativecommons.org/licenses/by-nc/3.0/)](data/images/4crabs.png){#fig:fig_crabs}
+
+<br><br>
+
+### Mesophotic ARMS {-}
+
+The `r nsamples(cc$crabs_deep_untransformed)` units deployed at mesophotic depths $\geq$ 30 m on O&#x02BB;ahu yielded fewer species and individuals compared to the complete dataset, with `r num(sum(sample_sums(cc$crabs_deep_untransformed)),1)` crabs overall, `r sum(unlist(deep_summary$types_breakdown))` distinct taxa (`r deep_summary$types_breakdown$species` species-level identifications, `r deep_summary$types_breakdown$genus` genus, `r deep_summary$types_breakdown$family` family), `r num(deep_summary$abundance_mean,0.1)` $\pm$ `r num(deep_summary$abundance_sd,0.1)` individuals per unit, and abundance ranging from `r format_range(deep_summary$abundance_range)`.
+The four most abundant taxa in the deep ARMS did not overlap with the most abundant taxa from the complete dataset: `r with(deep_summary,top_list(top,top_n,top_units))`.
+There were `r deep_summary$singleton_count` singletons, which proportionally accounted for ~`r pc(deep_summary$singleton_count/ntaxa(cc$crabs_deep_untransformed),0.1)` of species found on deep ARMS.
+
+### Comparisons {-}
+```{r depth-ranges, results="hide", dependson="taxonomy-validation"}
+f <- cc$crabs %>%
+  psmelt() %>%
+  group_by(depth_zone) %>%
+  filter(Abundance > 0) %>%
+  ungroup() %>%
+  group_by(family) %>%
+  filter(n_distinct(shallow_deep) == 1) %>%
+  ungroup()
+
+fam <- f %>%
+  distinct(family,shallow_deep) %>%
+  count(shallow_deep) %>%
+  deframe()
+
+deep_fams <- f %>%
+  filter(shallow_deep == "deep") %>%
+  pull(family) %>%
+  unique() %>%
+  sort()
+
+g <- cc$crabs %>%
+  psmelt() %>%
+  filter(!is.na(genus)) %>%
+  group_by(depth_zone) %>%
+  filter(Abundance > 0) %>%
+  ungroup() %>%
+  group_by(genus) %>%
+  filter(n_distinct(shallow_deep) == 1) %>%
+  ungroup()
+
+gen <- g %>%
+  distinct(genus,shallow_deep) %>%
+  count(shallow_deep) %>%
+  deframe()
+
+deep_gens <- g %>%
+  filter(shallow_deep == "deep") %>%
+  pull(genus) %>%
+  unique() %>%
+  sort()
+deep_gens <- str_glue("*{deep_gens}*")
+
+s <- cc$crabs %>%
+  psmelt() %>%
+  group_by(depth_zone) %>%
+  filter(Abundance > 0) %>%
+  ungroup() %>%
+  group_by(species) %>%
+  filter(n_distinct(shallow_deep) == 1) %>%
+  ungroup()
+
+spp <- s %>%
+  distinct(species,shallow_deep) %>%
+  count(shallow_deep) %>%
+  deframe()
+
+deep_spp <- s %>%
+  filter(shallow_deep == "deep") %>%
+  arrange(species) %>%
+  select(OTU) %>%
+  left_join(crab_taxonomy_validated,by=c("OTU" = "otu")) %>%
+  pull(display_species) %>%
+  unique()
+```
+Out of the `r all_summary$taxonomy_breakdown$families` families recovered, `r fam['shallow']` were exclusive to shallow ARMS recovered from `r dr(cc$crabs_shallow_untransformed)` m and `r fam['deep']` were exclusive to units $\geq$ 30 m (`r format_list(deep_fams)`; Supplementary [@tbl:tbl_supp_family_depth]).
+`r str_to_sentence(english(gen['shallow']))` genera were found exclusively shallow and `r gen['deep']` exclusively deep (`r format_list(deep_gens)`; Supplementary [@tbl:tbl_supp_genus_depth]).
+`r str_to_sentence(english(spp['shallow']))` distinct taxa (including those collapsed to genus or family) were exclusive to shallow ARMS units and `r spp['deep']` were found only deep (`r format_list(deep_spp)`; Supplementary [@tbl:tbl_supp_species_depth]).
+For a more in-depth taxonomic treatment of the assemblage composition of mesophotic vs.
+shallow ARMS units, see Hurley et al.
+[-@hurleyAssessmentShallowMesophotic2016].
+
+## Alpha diversity {-}
+
+### Summary {-}
+
+Overall mean species richness for the complete dataset was `r num(all_alpha$richness,0.1)` $\pm$ `r num(all_alpha$richness_sd,0.1)` species per unit (effective Simpson diversity: `r num(all_alpha$simpson,0.01)` $\pm$ `r num(all_alpha$simpson_sd,0.01)`).
+Overall mean species richness for shallow ARMS \< 30 m was `r num(shallow_alpha$richness,0.1)` $\pm$ `r num(shallow_alpha$richness_sd,0.1)` species per unit (effective Simpson diversity: `r num(shallow_alpha$simpson,0.01)` $\pm$ `r num(shallow_alpha$simpson_sd,0.01)`).
+Overall mean species richness for mesophotic ARMS $\geq$ 30 m was `r num(deep_alpha$richness,0.1)` $\pm$ `r num(deep_alpha$richness_sd,0.1)` species per unit (effective Simpson diversity: `r num(deep_alpha$simpson,0.01)` $\pm$ `r num(deep_alpha$simpson_sd,0.01)`).
+
+### Comparisons {-}
+
+For the complete dataset, there was no significant difference in observed species richness (`r k_test(all_alpha$k_r_island)`; Supplementary [@fig:fig_supp_alpha[]{}a]) or effective Simpson diversity (`r k_test(all_alpha$k_s_island,plain=TRUE)`; Supplementary [@fig:fig_supp_alpha[]{}c]) between the MHI and NWHI.
+Richness was not significantly different between shallow and deep samples (`r k_test(all_alpha$k_r_depth,plain=TRUE)`; Supplementary [@fig:fig_supp_alpha[]{}b]) but effective Simpson diversity was (`r k_test(all_alpha$k_s_depth,plain=TRUE)`; Supplementary [@fig:fig_supp_alpha[]{}d]).
+When comparing diversity across sampling depths for only sites in the MHI, richness was not significantly different (`r k_test(all_alpha$k_r_depth_main,plain=TRUE)`; Supplementary [@fig:fig_supp_alpha_main[]{}a]) but effective Simpson diversity was (`r k_test(all_alpha$k_s_depth_main,plain=TRUE)`; Supplementary [@fig:fig_supp_alpha_main[]{}c]).
+Neither richness nor Simpson diversity were significantly different between shallow and deep when comparisons were constrained to the island of O&#x02BB;ahu (Richness: `r k_test(all_alpha$k_r_depth_oahu,plain=TRUE)`, Supplementary [@fig:fig_supp_alpha_main[]{}b]; Simpson: `r k_test(all_alpha$k_s_depth_oahu,plain=TRUE)`, Supplementary [@fig:fig_supp_alpha_main[]{}d]).
+When only shallow arms \< 30 m were considered, there were no significant differences in alpha diversity between the MHI and NWHI for either index (richness: `r k_test(shallow_alpha$k_r_island,plain=TRUE)`, Supplementary [@fig:fig_supp_alpha2[]{}a]; effective Simpson: `r k_test(shallow_alpha$k_s_island,plain=TRUE)`, Supplementary [@fig:fig_supp_alpha2[]{}b]).
+
+<!-- ### Effects of environmental variables {-} -->
+## Brachyuran community structure {-}
+
+```{r dbrda-analysis, results="hide", dependson="init"}
+library(formula.tools)
+
+# set random seed for consistency
+set.seed(32767)
+
+reduce_model <- function(m,o,d,vif_limit=3,condition=character()) {
+  vifs <- vif.cca(m)
+  if (any(vifs > vif_limit)) {
+    oldterms <- rhs.vars(m$call$formula)
+    v <- names(vifs)[which.max(vifs)]
+    newterms <- grep(v,oldterms,value=TRUE,invert=TRUE)
+    f <- reformulate(newterms,response=quote(o))
+    newmod <- dbrda(f,data=d,distance="bray")
+    reduce_model(newmod,o,d,vif_limit,condition)
+  } else {
+    return(m)
+  }
+}
+
+# calculate full dbRDA model
+db_modl <- dbrda(
+  cc$crab_otus_shallow ~ depth +
+    chl + sst + slope + coral_cover +
+    larval_immigration + human_impact +
+    lat + lon +
+    Condition(recovery_year),
+  data=cc$crab_data_shallow,
+  # data=dd,
+  distance="bray"
+)
+
+# reduce model to terms with VIF <= 3
+db_modl <- reduce_model(db_modl,cc$crab_otus_shallow,cc$crab_data_shallow,vif_limit = 3)
+
+# calculate biplot vectors and magnitudes
+vectors <- db_modl %>%
+  scores(display="bp") %>%
+  as_tibble(rownames="term") %>%
+  inner_join(model_terms %>% select(variable,driver=`Environmental Driver`),by=c("term" = "variable")) %>%
+  mutate(size = sqrt(dbRDA1^2 + dbRDA2^2), driver=str_to_lower(driver)) %>%
+  arrange(desc(size))
+
+# get marginal term information and sort by descending f-statistic
+modl_perm <- anova(db_modl,by="margin",permutations = PERM) %>%
+  # as_tibble(rownames="term") %>%
+  # janitor::clean_names() %>%
+  broom::tidy() %>%
+  arrange(desc(statistic))
+
+# make a nicer display-worthy table for significant terms
+rda_table <- modl_perm %>%
+  inner_join(model_terms %>% select(variable,driver=`Environmental Driver`),by=c("term" = "variable")) %>%
+  select(-term) %>%
+  select(Term=driver,DF=df,`Sum of Squares`=SumOfSqs,`F-statistic`=statistic,`p-value`=p.value)
+
+# get number of model terms
+tt <- attr(db_modl$terms,"term.label")
+nterm <- length(grep("Condition",tt,ignore.case = FALSE,value=TRUE,invert = TRUE))
+
+# get top drivers for the text
+drivers <- rda_table %>%
+  slice(1:2) %>%
+  select(term=Term,stat=`F-statistic`,pval=`p-value`)  %>%
+  mutate(
+    term=str_to_lower(term),
+    term = str_replace_all(term,r'(\b[a-z]\b)',replacement=str_to_upper)
+  )
+
+all_terms <- english::english(nrow(model_terms))
+```
+### Environmental drivers {-}
+Environmental variables were strongly clustered according to island group (MHI vs NWHI) (Supplementary [@fig:fig_supp_pca]).
+The best dbRDA model (`r permanova(db_modl,stat="F")`) explained `r pc(RsquareAdj(db_modl)$adj.r.squared,0.01)` (adjusted $\mathrm{R}^2$) of the variability in crab communities across the archipelago ([@fig:fig_dbrda]).
+`r str_to_sentence(english(nterm))` of the `r str_to_lower(english(nrow(model_terms)))` environmental variables tested were significant and sufficiently uncorrelated.
+The two drivers contributing most to explained variation were `r drivers$term[1]` (F = `r num(drivers$stat[1],0.01)`) and `r drivers$term[2]` (F = `r num(drivers$stat[2],0.01)`) (Supplementary [@tbl:tbl_supp_rda]).
+These factors also produced the longest effect vectors in the dbRDA plot ([@fig:fig_dbrda]).
+
+```{r fig-dbrda, fig.cap="dbRDA plot showing significant drivers of crab community variation across the Hawaiian Islands.
+Colored ellipses represent 95% confidence intervals around island groups.", results="asis", fig.width=9, fig.height=6, dependson="rda-analysis"}
+termz <- model_terms %>%
+  select(variable,display=`Environmental Driver`)
+
+# make the initial ordination plot
+rda_plotz <- plot_ord(cc$crabs_shallow,db_modl,scale=2,color="island_group",labelsize=3,term_map=termz) +
+  stat_ellipse(geom="polygon",aes(group=island_group,fill=island_group),color=NA,alpha=0.1,show.legend = FALSE) +
+  geom_point(aes(fill=island_group),shape=21,color="black",size=3) +
+  plotz_theme() +
+  scale_fill_manual(name="Island group",values=island_group_pal,labels = c("Main","Northwest")) +
+  guides(color = "none")
+
+# calculate variation explained (constrained & unconstrained)
+eigs <- eigenvals(db_modl)
+constrained <- eigs[str_detect(names(eigs),"dbRDA")]
+constrained <- constrained / sum(constrained)
+unconstrained <- eigs/sum(eigs)
+
+# function to format axis titles
+splained <- function(name,constrained,total,accuracy=0.1) {
+  str_glue("{name} ({pc(constrained,accuracy)} constrained, {pc(total,accuracy)} total)")
+}
+
+axes <- str_glue_data(list(num=1:2),"dbRDA{num}")
+axis_var <- splained(axes,constrained[1:2],unconstrained[1:2])
+
+# put the arrows on top
+rda_plotz$layers <- rda_plotz$layers[c(1,2,5,6,3,4)]
+
+# plot the plotz
+rda_plotz +
+  xlab(axis_var[1]) + ylab(axis_var[2])
+cat("{#fig:fig_dbrda}")
+```
+
+<br><br>
+
+### Regional & depth comparisons {-}
+
+```{r permanova, dependson="init"}
+anova_region <- adonis2(cc$crab_dist_shallow ~ island_group,data=cc$crab_data_shallow,permutations=PERM)
+anova_island <- adonis2(cc$crab_dist_shallow ~ island,data=cc$crab_data_shallow,permutations=PERM)
+bd_region <- betadisper(cc$crab_dist_shallow,cc$crab_data_shallow$island_group)
+
+anova_depth <- adonis2(cc$crab_dist ~ depth_zone,data=cc$crab_data,permutations=PERM)
+anova_depth_oahu <- adonis2(cc$crab_dist_oahu ~ depth_zone,data=sample_tibble(cc$crabs_oahu),permutations=PERM)
+anova_depth2_oahu <- adonis2(cc$crab_dist_oahu ~ shallow_deep,data=sample_tibble(cc$crabs_oahu),permutations=PERM)
+bd_depth <- betadisper(cc$crab_dist,cc$crab_data$depth_zone)
+
+anova_both <- adonis2(cc$crab_dist ~ island_group + depth_zone, data=cc$crab_data,permutations = PERM, by="margin")
+effects <- broom::tidy(anova_both) %>%
+  filter(!term %in% c("Residual","Total")) %>%
+  pull(R2)
+effect_ratio <- max(effects)/min(effects)
+
+cd <- cc$crabs_deep %>%
+  sample_tibble()
+bd_deep <- betadisper(distance(cc$crabs_deep,"bray"),cd$depth_zone)
+```
+
+```{r cap-analysis,dependson="init",echo=FALSE,message=FALSE,warning=FALSE,results="hide"}
+# convert otu tables and sample data into appropriate format
+otus <- otu_table(cc$crabs) %>%
+  as.data.frame()
+cd <- cc$crab_data %>%
+  mutate(sample_group = factor(sample_group,ordered=FALSE)) %>%
+  as.data.frame()
+
+# run CAP analysis, taking care of negative eigenvalues where necessary
+cap_depth <- CAPdiscrim(otus ~ sample_group, data=cd, dist="bray", axes=2, m=0, add=TRUE)
+# # calculate proportion of explained variance
+cap_var <- cap_depth$manova$Eigenvalues / sum(cap_depth$manova$Eigenvalues)
+cap_var <- list(x = cap_var[1], y = cap_var[2])
+
+# make the sample group look nice
+nice <- function(s) {
+  str_replace(s,"shallow","Shallow") %>%
+    str_replace("main","(MHI)") %>%
+    str_replace("nwhi","(NWHI)") %>%
+    str_replace("(^[0-9]+)","O‘ahu \\1") %>%
+    str_split("((?<=[0-9])(?=[a-zA-Z]))|(_)") %>%
+    map_chr(~str_c(.x,collapse=" "))
+}
+
+# get all the stuff we need to plot
+pcap <- cap_depth$x %>%
+  as_tibble(rownames="sample") %>%
+  left_join(cd,by="sample") %>%
+  mutate(
+    sample_group = nice(sample_group),
+    sample_group = fct_reorder(sample_group,depth)
+  )
+
+shallow_cd <- pcap %>%
+  filter(shallow_deep == "shallow") %>%
+  select(sample,LD1,LD2) %>%
+  column_to_rownames("sample") %>%
+  centroid_dist(mean=FALSE)
+
+deep_cd <- pcap %>%
+  filter(shallow_deep == "deep") %>%
+  select(sample,LD1,LD2) %>%
+  column_to_rownames("sample") %>%
+  centroid_dist(mean=FALSE)
+
+centroid_t <- t.test(deep_cd,shallow_cd,alternative = "g",var.equal = FALSE)
+
+oahu_dist <- pcap %>%
+  filter(island == "Oahu") %>%
+  select(sample,LD1,LD2) %>%
+  column_to_rownames("sample") %>%
+  dist() %>%
+  c()
+
+shallow_dist <- pcap %>%
+  filter(shallow_deep == "shallow") %>%
+  select(sample,LD1,LD2) %>%
+  column_to_rownames("sample") %>%
+  dist() %>%
+  c()
+
+deep_dist <- pcap %>%
+  filter(shallow_deep == "deep") %>%
+  select(sample,LD1,LD2) %>%
+  column_to_rownames("sample") %>%
+  dist() %>%
+  c()
+
+distance_t <- t.test(deep_dist,shallow_dist,alternative = "g",var.equal = FALSE)
+oahu_t <- t.test(oahu_dist,shallow_dist,alternative = "g",var.equal = FALSE)
+```
+Shallow crab community structure was significantly different between island groups (MHI vs.
+NWHI) (`r permanova(anova_region,0.01)`) as well as among individual islands (`r permanova(anova_island,0.01)`; see Supplementary [@fig:fig_supp_pairwise] for results of pairwise comparisons).
+There was no significant difference in community dispersion between the MHI and NWHI (`r permdisp(bd_region)`).
+When mesophotic ($\geq$ 30 m) ARMS units were included, there were significant differences among depth zones (\< 30 m, 30 m, 60 m, 90 m) both for all samples (`r permanova(anova_depth,0.01)`) and samples restricted to O&#x02BB;ahu (`r permanova(anova_depth_oahu,0.01)`), and community dispersion (`r permdisp(bd_depth)`).
+However, community dispersion was not different among depth zones $\geq$ 30 m (`r permdisp(bd_deep)`).
+Although crab communities were significantly different among both depth zones and island groups for the complete dataset (`r permanova(anova_both,0.01)`), depth explained `r num(effect_ratio,0.1)` times more variability than island group (based on model $\mathrm{R}^2$).
+
+In the CAP analysis of island group and depth zone, ARMS units were correctly reassigned to sample groups `r pc(cap_depth$percent/100,0.1)` of the time.
+The first two linear discriminant axes explained `r pc(cap_var$x,0.1)` and `r pc(cap_var$y,0.1)` of the variation, respectively.
+Mesophotic samples (30--90 m) from O&#x02BB;ahu grouped separately from the shallow samples collected across the entire ~2,500 km expanse of the archipelago ([@fig:fig_cap]), whereas shallow samples from O&#x02BB;ahu primarily clustered with shallow samples from other islands ([@fig:fig_cap]).
+Variability among 30--90 m samples was apparently greater than among all shallow samples, with mean Euclidean distances between deep samples in the CAP ordination significantly greater than between shallow samples (`r t_test(distance_t)`; [@fig:fig_cap], Supplementary [@fig:fig_supp_euclid]).
+Similarly, variability among samples from O&#x02BB;ahu (including mesophotic depths) exceeded variability among all shallow samples combined (based on Euclidean distance among CAP points, as above; `r t_test(oahu_t,plain=TRUE)`).
+
+```{r fig-cap, fig.cap="CAP ordination of crab communities by island group and depth zone.", results="asis", fig.width=9, fig.height=6, dependson="cap-analysis"}
+centroids <- pcap %>% group_by(sample_group) %>%
+  summarise(LD1=mean(LD1),LD2=mean(LD2))
+pcap <- pcap %>%
+  mutate(
+    oahu = factor(
+      # if_else(island == "Oahu","oahu","noahu"),
+      case_when(
+        island == "Oahu" ~ "oahu",
+        island == "French Frigate Shoals" ~ "ffs",
+        .default = "noahu"
+      ),
+      levels = c("oahu","ffs","noahu")
+    )
+  )
+
+ggplot(pcap,aes(x=LD1,y=LD2,fill=sample_group)) +
+  stat_ellipse(geom="polygon",level=0.95,alpha=0.1,show.legend = FALSE) +
+  # geom_point(aes(shape=oahu),color="black",shape=21,size=5)  +
+  geom_point(mapping=aes(shape=oahu),color="black",size=5)  +
+  geom_label_repel(data=centroids,aes(label=sample_group), color="white", show.legend = FALSE) +
+  scale_shape_manual(values=c("oahu" = 23, "ffs" = 22, "noahu" = 21),name="Island", labels=c("noahu"="Other islands","oahu"="O‘ahu","ffs"="Lalo (French Frigate Shoals)" )) +
+  scale_fill_manual(values=sample_depth_pal,name="Sample group") +
+  guides(
+    fill=guide_legend(override.aes=list(shape=21)),
+    shape=guide_legend(override.aes=list(fill="black"))
+  ) +
+  xlab(str_glue("Linear discriminant 1 [{pc(cap_var$x,0.1)}]")) +
+  ylab(str_glue("Linear discriminant 2 [{pc(cap_var$y,0.1)}]")) +
+  plotz_theme()
+
+# centroids <- pcap %>% group_by(sample_group) %>%
+#   summarise(LD1=mean(LD1),LD2=mean(LD2))
+# pcap <- pcap %>%
+#   mutate(
+#     oahu = factor(
+#       if_else(island == "Oahu","oahu","noahu"),
+#       levels = c("oahu","noahu")
+#     )
+#   )
+
+# ggplot(pcap,aes(x=LD1,y=LD2,fill=sample_group)) +
+#   stat_ellipse(geom="polygon",level=0.95,alpha=0.1,show.legend = FALSE) +
+#   # geom_point(aes(shape=oahu),color="black",shape=21,size=5)  +
+#   geom_point(mapping=aes(shape=oahu),color="black",size=5)  +
+#   geom_label_repel(data=centroids,aes(label=sample_group), color="white", show.legend = FALSE) +
+#   scale_shape_manual(values=c("oahu" = 23, "noahu" = 21),name="Island", labels=c("noahu"="Other islands","oahu"="O‘ahu")) +
+#   scale_fill_manual(values=sample_depth_pal,name="Sample group") +
+#   guides(
+#     fill=guide_legend(override.aes=list(shape=21)),
+#     shape=guide_legend(override.aes=list(fill="black"))
+#   ) +
+#   xlab(str_glue("Linear discriminant 1 [{pc(cap_var$x,0.1)}]")) +
+#   ylab(str_glue("Linear discriminant 2 [{pc(cap_var$y,0.1)}]")) +
+#   plotz_theme()
+cat("{#fig:fig_cap}")
+```
+
+<br><br>
+
+<!-- TODO: add somewhere something like:
+although depth was significant in the shallows, its relative effect on community differences was less important than other factors (e.g., human impacts, chlorophyll, etc.).
+however, it clearly outpaced other covarying factors along the depth gradient
+-->
+# Discussion {-}
+The Hawaiian Archipelago is among the most remote oceanic island chains on the planet and encompasses a range of gradients (anthropogenic, biotic, abiotic), making it an ideal natural laboratory to study the impacts of various drivers of species assemblages across space.
+Human activities are considerably higher in the populated MHI than in the protected and remote NWHI [@selkoeEvaluatingAnthropogenicThreats2008; @selkoeMapHumanImpacts2009; @selkoeDNACoralReef2016].
+Although we did not consider the effects of individual impacts here, an index representing *cumulative* anthropogenic impacts was among the top drivers of shallow brachyuran community structure, similar to findings in previous work examining genetic diversity as it relates to species richness and habitat area, quality, and stability across the archipelago [@selkoeDNACoralReef2016].
+The next-most significant driver, chlorophyll-A, can be used as a proxy for primary productivity.
+Productivity gradients across the Hawaiian Archipelago coincide with differences in coral reef community composition [@selkoeDNACoralReef2016], endemic reef fish dominance [@friedlanderDominanceEndemicsReef2020], survival rates of top predators [@bakerEffectVariableOceanic2007], and island geomorphology [@goveNearislandBiologicalHotspots2016].
+Although not a significant factor in our model, coral cover and species composition are thought to affect patterns of benthic assemblages.
+Due to the limited availability of datasets covering our entire study area, our model only included an estimate of coral cover at the island scale, but previous work has shown that coral species composition does not vary considerably across the archipelago (aside from the presence of *Acropora* in the NWHI) [@griggCommunityStructureSuccession1983].
+Crab communities from Lalo (French Frigate Shoals), which was found to harbor the highest coral species diversity [@griggCommunityStructureSuccession1983], did not group separately from the rest of our samples ([@fig:fig_cap]).
+Given the prevalence in our model of significant factors known to be important influencers of marine communities in Hawai&#x02BB;i, it is not surprising that we find the same top drivers of community structure for brachyuran crabs, one of the primary taxonomic groups inhabiting coral reefs.
+What is surprising---given the maximum vertical separation among shallow ARMS units of `r diff(range(cc$crab_data_shallow_unscaled$depth))` m---is the consistent significance of depth as a driver of community composition across these gradients.
+
+Although our shallow-water ARMS units were deployed within a narrow depth range (`r dr(cc$crabs_shallow_untransformed)` m), depth still emerged as a significant and consistent driver of community variation among these units.
+Similar fine-scale depth patterns have been found in previous studies, both using molecular [e.g., @desouzaImportanceDepthTemperature2023; @westEDNAMetabarcodingSurvey2020] and traditional survey methods [e.g., @williamsBenthicCommunitiesTwo2013; @srinivasanDepthDistributionsCoral2003], but our results recover them alongside numerous other geographic and environmental gradients.
+Depth in and of itself is unlikely to affect marine community composition and oceanographic factors with well-known depth profiles (e.g., temperature, oxygen, nutrients, etc.) are not expected to vary dramatically across such a narrow depth range (\< 10 m), although internal waves can deliver variable temperatures across shallow reefs [@storlazziInternalTidesCan2020; @wyattHeatAccumulationCoral2020] and fine-scale habitat variability among sites may also influence the community differences we found [@steyaertRemoteReefCryptobenthic2022].
+Unfortunately, we lack high resolution environmental data at depth for individual deployments, and thus cannot unravel the co-varying factor(s) associated with the depth effects we observed for shallow assemblages.
+Regardless, given its significance across both narrow (\< 10 m) and wider (90 m) depth ranges sampled here, a factor associated with depth is clearly an important driver for cryptobenthic crab communities across the Hawaiian Archipelago.
+As brachyurans are a ubiquitous and important component of coral reef communities [@reynsPatternsProcessesBrachyuran1999; @lasleyjrPhylogeneticRelationshipsUbiquitous2015; @hazeriLatitudinalSpeciesDiversity2019], these findings are likely to apply more broadly as well.
+
+While depth was a significant driver of community composition among shallow-water ARMS units ([@fig:fig_dbrda], Supplementary [@tbl:tbl_supp_rda]), its influence was considerably more striking when comparing shallow to mesophotic habitat.
+In our CAP ordination, all shallow assemblages across the archipelago clustered separately from the deep assemblages $\geq$ 30 m at the single island of O&#x02BB;ahu ([@fig:fig_cap]).
+Additionally, the two deepest sampled depths (60 m and 90 m) clustered apart from 30 m.
+According to this analysis, there was more community variation among samples spanning 90 m of depth on a single island than across ~2,500 km of shallow habitat experiencing multiple gradients of environmental, ecological, and anthropogenic influence.
+This is further supported by the fact that the average Euclidean distance among O&#x02BB;ahu samples (including mesophotic) in canonical space was significantly greater than the average distance among all shallow samples (MHI & NWHI).
+Shallow crab communities from O&#x02BB;ahu appear to share more affinity with Hōlanikū (~2,200 km distant) than they do with those at 90 m depth on the same reef slope ([@fig:fig_cap]).
+The CAP also revealed two groups within the mesophotic samples: one at 30 m and the other encompassing the 60--90 m depths.
+These depth zones correspond with the upper and lower mesophotic zones, respectively [@kahngCommunityEcologyMesophotic2010; @rochaMesophoticCoralEcosystems2018; but see @pyleMesophoticCoralEcosystems2019].
+The upper mesophotic zone has been found to constitute an extension of shallow reefs (with beta diversity patterns largely driven by nestedness), whereas the lower mesophotic harbors unique communities (with beta diversity driven primarily by turnover) [@rochaMesophoticCoralEcosystems2018; @slatteryFunctionStabilityMesophotic2024a; @bellwoodRegionalscaleAssemblyRules2001; @kahngCommunityEcologyMesophotic2010].
+The evident groupings in the CAP ordination match these trends, with 30 m samples lying closer to shallow O&#x02BB;ahu samples and 60--90 m samples forming a distinct cluster.
+This pattern of community transition from shallow to mesophotic reefs is consistent with what has been shown for various other taxonomic groups [@lesserGlobalCommunityBreaks2019; @rochaMesophoticCoralEcosystems2018], lending support to the hypothesis that there may be broad-scale depth-based assembly rules for coral reef communities.
+
+In addition to community composition, mesophotic samples also exhibited alpha diversity differences.
+Effective Simpson diversity was significantly higher in deep communities (for comparisons among all islands and solely within the MHI, but not exclusively at O&#x02BB;ahu) (Supplementary [@fig:fig_supp_alpha; @fig:fig_supp_alpha_main]).
+This pattern is in contrast to observations that biodiversity often decreases with depth [@pyleComprehensiveInvestigationMesophotic2016; @kahngCommunityEcologyMesophotic2010], although the differences we found were slight and our findings may be influenced by differences in habitat availability across depth zones (especially of the habitat types ARMS are designed to mimic).
+
+Since we lack environmental data at mesophotic depths, we cannot make conclusions about the specific influences driving the differences we found.
+However, light and temperature have been proposed as important bottom-up drivers of depth zonation in benthic communities [@kahngCommunityEcologyMesophotic2010; @lesserGlobalCommunityBreaks2019; @pyleComprehensiveInvestigationMesophotic2016; @pyleMesophoticCoralEcosystems2019].
+Oceanographic conditions may be inconsistent close to shore, but the average mixed-layer depth calculated (using 0.8 &deg;C $\upDelta$T [@karaOptimalDefinitionOcean2000]) from CTD casts taken at station ALOHA (22.75&deg;N, 158.00&deg;W; within the latitudinal expanse of the MHI) was 84.6 $\pm$ 26 m over the course of the year (based on monthly averages from casts taken across the duration of ARMS unit deployments [@HawaiiOceanTimeseries]).
+Thus, given the assumption of similar temperature profiles near the O&#x02BB;ahu shoreline, temperature alone appears insufficient to explain the zonation patterns we observed---although due to its significant role as a driver of shallow community variation, it cannot be ruled out.
+In contrast, the lower depth boundaries of mesophotic ecosystems are largely defined by surface irradiance [@kahngCommunityEcologyMesophotic2010; @hindersteinThemeSectionMesophotic2010], so light seems a possible candidate as bottom-up abiotic factor influencing community structure across depths.
+Crab assemblage structure is unlikely to be directly affected by light variability (although the behavior of individual crabs may).
+However, habitat-building species such as corals, algae, and sponges are frequently subject to light requirements [@kahngCommunityEcologyMesophotic2010; @slatteryFunctionStabilityMesophotic2024a], and may in turn affect the structure of the communities dependent upon them [@hustonCompetitionCoexistenceEffects1994; @smithRoleHabitatComplexity2014; @sebensHabitatStructureCommunity1991].
+If patterns of community structure are driven primarily from the bottom up by light, that may contribute to our observation of a significant depth effect over less than 10 m in our shallow samples.
+
+Two potential weaknesses of our dataset are the time periods over which ARMS units were deployed and the fact that mesophotic samples were obtained solely from the island of O&#x02BB;ahu.
+ARMS units were deployed across a broad range of years, but the majority were deployed during 2010--2012 or 2010--2013 (an overlap of two years) ([@tbl:tbl_arms]).
+A reduced version of the CAP including only ARMS units deployed during these overlapping time periods retains the same (if not more pronounced) overall patterns as the full version (Supplementary [@fig:fig_supp_cap_reduced]), suggesting that the inclusion of previous or subsequent soaking periods did not adversely skew our analyses.
+The limitation of our depth transect to O&#x02BB;ahu means that we cannot make broad-scale conclusions about biodiversity patterns on deep reefs across the Hawaiian Islands.
+However, it remains the case that community and alpha diversity differences among depth zones at that single island appear to overwhelm differences observed among shallow-water coral reef habitat across the archipelago.
+Similar patterns were found at a local scale (tens of kilometers) on Hawai&#x02BB;i Island using environmental DNA (eDNA) [@hobanPlumbingDepthsEnvironmental2023].
+In that study, communities were found to have more affinity within depth zones across sites than among depth zones within sites.
+Thus, despite the limitations in our dataset, the patterns we observed suggest that factors associated with the shallow-to-mesophotic depth gradient exert stronger ecological influence than those (including depth) that vary across the latitudinal, physical, and anthropogenic gradients throughout the Hawaiian Archipelago.
+
+Further work to investigate drivers of depth zonation could take advantage of established gradients and use the Hawaiian Archipelago as a natural laboratory, by sampling depth transects across the archipelago.
+Coral reef biodiversity is generally thought to be structured at depth by a combination of the attenuation of light, a decrease in temperature, increasing salinity, and a reduction in trophic resources [@lesserEcologyMesophoticCoral2009; @kahngCommunityEcologyMesophotic2010; @slatteryFunctionStabilityMesophotic2024a].
+If mesophotic communities are found to be different from their shallow counterparts across multiple islands (ideally comparing sites experiencing similar irradiance), this suggests that they are vertically structured by a factor (e.g., light) outside of the numerous natural gradients that exist across the Hawaiian Islands.
+In contrast, other factors (such as temperature, dissolved inorganic nitrogen, particulate organic carbon, or productivity) expected to have more strongly-defined gradients across the regions [@slatteryFunctionStabilityMesophotic2024a], would lead to variable zonation patterns that should be detected in a spatial comparison of the MHI & NWHI.
+Such differences should be tested in future studies to understand the specific physical and ecological drivers of coral reef depth zonation.
+The cost, logistical complexity, and time investment required for ARMS deployments make this challenging, but emerging approaches such as environmental DNA (eDNA) may offer a suitable alternative.
+
+Marine communities are structured by a suite of factors, but there remains no consensus for either generalized mechanisms or the relative importance of drivers of community structure for coral reefs.
+Our best knowledge of the role of depth in structuring marine communities comes from the intertidal, and points to biotic factors (e.g., predation, competition) determining lower limits and abiotic factors (e.g., desiccation, thermal stress) setting upper limits of vertical species distributions [@connellEffectsCompetitionPredation1961; @connellInfluenceInterspecificCompetition1961; @dennyPhysicalProcessesThat2001; @paineIntertidalCommunityStructure1974].
+There are numerous examples from other study systems in which the relative contributions of biotic versus abiotic factors to structuring communities run in opposing directions across some habitat gradient (e.g., climate, soil type, elevation, etc.
+[@bradyEvolutionaryEcologyPlant2005; @moore10PlantCompetition2011; @wishingradTemperateZoneIsolation2023]).
+However, abiotic stressors on intertidal communities are obviously different from those on coral reefs, and generalizations about the mechanisms and ecological processes that structure coral reef communities remain a subject of considerable debate [@bellwoodEnvironmentalGeometricConstraints2005; @bellwoodSleepingFunctionalGroup2006; @carrBiodiversityPopulationRegulation2002; @connollyIndoPacificBiodiversityCoral2003; @dornelasCoralReefDiversity2006; @grahamImportanceStructuralComplexity2013; @komyakovaRelativeImportanceCoral2013].
+In the Hawaiian archipelago, we found significant differences in the composition of shallow coral reef crab communities between the populated high islands of the MHI and the uninhabited atolls of the NWHI.
+Moreover, these differences were driven by a combination of environmental (sea-surface temperature, chlorophyll-A concentration, depth), biotic (potential larval immigration), and anthropogenic (cumulative human impacts) factors.
+Despite the significance of these factors in driving shallow community variation across the archipelago, mesophotic assemblages were found to be distinct ([@fig:fig_cap]).
+Although there were variables we did not consider in our model, this suggests that vertical distributions are influenced by factors other than those driving spatial variation across the island chain.
+In light of this, and since mesophotic communities are generally less biodiverse than shallow reefs [@kahngCommunityEcologyMesophotic2010; @lesserEcologyMesophoticCoral2009; @pyleComprehensiveInvestigationMesophotic2016], it may be possible that the gradients driving depth distributions on coral reefs run opposite to those in the intertidal.
+Given the rapid changes in important factors like temperature, nutrients, and light throughout the upper ~100--150 m of the water column (particularly the diminishment of photosynthetically active radiation below 100 m), the *lower* limits of the coral reef mesophotic zone may be determined by abiotic factors that correlate with depth.
+Conversely, the *upper* limits of the mesophotic zone (the transition from shallow to mesophotic habitat), which occur at the higher end of the depth-biodiversity gradient [@kahngCommunityEcologyMesophotic2010; @lesserEcologyMesophoticCoral2009] may be driven by ecological interactions [@vellendEffectsDiversityDiversity2008].
+
+By comparing shallow-water ARMS across the latitudinal, physical and anthropogenic gradients of the Hawaiian Archipelago, we were able to test the relative importance of these factors in driving shallow crab assemblage structure.
+We additionally find that assemblages along a 90 m depth transect at a single island show greater divergence in canonical space than across the full ~2,500 km gradient of the Hawaiian Archipelago on shallow reefs.
+Our analyses corroborate previous work and show that depth is among the strongest drivers of community structure.
+We conclude that some factor(s) covarying with depth exert more influence on cryptobenthic coral reef community structure than those biotic and abiotic gradients that span the entire Hawaiian Archipelago in shallow-water environments.
+
+# Methods {-}
 
@@
 
-A total of 89 ARMS units were deployed over the course of six years (2010--2016) in shallow (`r dr(cc$crabs_shallow_untransformed)` m depth) coral reefs in both the Main (MHI) and Northwest Hawaiian Islands (NWHI) ([@fig:fig_map; @tbl:arms_table]).
+A total of 89 ARMS units were deployed over the course of six years (2010--2016) in shallow (`r dr(cc$crabs_shallow_untransformed)` m depth) coral reefs in both the Main (MHI) and Northwest Hawaiian Islands (NWHI) ([@fig:fig_map; @tbl:tbl_arms]).
 All units included in this study were deployed in forereef habitat.
 To explore the effects of depth, we also present data from an earlier study by Hurley et al.
-[-@hurleyAssessmentShallowMesophotic2016], in which 27 units were deployed off O&#x02BB;ahu from 2009--2012 in `r format_range(cc$crab_data %>% sample_tibble() %>% filter(island == "Oahu") %>% pull(depth))` m depth; nine shallow units were deployed at 12 m, and six each at depths of 30 m, 60 m, and 90 m ([@tbl:arms_table]).
+[-@hurleyAssessmentShallowMesophotic2016], in which 27 units were deployed off O&#x02BB;ahu from 2009--2012 in `r format_range(cc$crab_data %>% sample_tibble() %>% filter(island == "Oahu") %>% pull(depth))` m depth; nine shallow units were deployed at 12 m, and six each at depths of 30 m, 60 m, and 90 m ([@tbl:tbl_arms]).
 Shallow units were deployed according to NOAA-RAMP methodology, whereas units $\geq$ 30 m were deployed and recovered using technical mixed gas closed-circuit rebreather diving [see @hurleyAssessmentShallowMesophotic2016].
@@
 
-```{r arms-table,results="asis", dependson="init"}
+```{r tbl-arms,results="asis", dependson="init"}
 arms_schedule <- cc$crabs_untransformed %>%
@@
   select(island,`Depth range`=depth,order(colnames(.)))
-print_table(arms_schedule,"arms_table","ARMS unit deployment schedule by island (number of units). Mesophotic deployments apear in bold.",accuracy=1)
+print_table(arms_schedule,"tbl_arms","ARMS unit deployment schedule by island (number of units). Mesophotic deployments apear in bold.",accuracy=1)
 ```
@@
 
-We used distance-based redundancy analysis (dbRDA; a form of multivariate multiple regression) with `dbrda` to examine the relative contribution of environmental variables ([@tbl:env_table]) to shallow crab community variation across the islands, conditioning the recovery year as a random effect.
+We used distance-based redundancy analysis (dbRDA; a form of multivariate multiple regression) with `dbrda` to examine the relative contribution of environmental variables ([@tbl:tbl_env]) to shallow crab community variation across the islands, conditioning the recovery year as a random effect.
 From a model incorporating all potential drivers, we sequentially removed variables with the largest variance inflation factors (VIF; calculated using `vif.cca`) until all values were <3, indicating variables were sufficiently uncorrelated.
@@
 Potential drivers of brachyuran crab community structure were assembled from previous studies [@selkoeDNACoralReef2016; @halpernRecentPaceChange2019], NOAA's National Coral Reef Monitoring Program (NCRMP), the Pacific Islands Ocean Observing System (PacIOOS), and MODIS satellite data products compiled by NASA's Physical Oceanography/Ocean Biology Distributed Active Archive Centers [@nasagoddardspaceflightcenterModerateresolutionImagingSpectroradiometer2022; @nasaobpgMODISAquaGlobal2020].
-Variables were either intrinsic (site depth, latitude, longitude, island slope) or derived from remote sensing and/or modeling (chlorophyll-A, sea-surface temperature, coral cover, potential larval immigration, cumulative human impacts) ([@tbl:env_table; @tbl:arms_metadata_table]).
+Variables were either intrinsic (site depth, latitude, longitude, island slope) or derived from remote sensing and/or modeling (chlorophyll-A, sea-surface temperature, coral cover, potential larval immigration, cumulative human impacts) ([@tbl:tbl_env], Supplementary [@tbl:tbl_supp_arms_metadata]).
 Sea-surface temperature (SST) and chlorophyll-A values were calculated by averaging monthly means across the duration of the soaking period for each deployment using AQUA MODIS 4 km-resolution gridded datasets.
@@
 mat.].
-For details regarding data import, preparation, and synthesis, see Supplemental material.
+For details regarding data import, preparation, and synthesis, see Supplementary Information.
 Due to logistical and dataset limitations, environmental/ecological drivers were only available for shallow (\< 30 m) samples.
 
-```{r env-table,results="asis", dependson="init"}
-print_table(model_terms,"env_table","Potential environmental drivers of crab community structure across the Hawaiian Archipelago",hide="variable")
+```{r tbl-env,results="asis", dependson="init"}
+print_table(model_terms,"tbl_env","Potential environmental drivers of crab community structure across the Hawaiian Archipelago",hide="variable")
 ```
@@
 
-# Results {-}
-
-## Diversity summary {-}
-
-### All ARMS {-}
-
-The `r nsamples(cc$crabs_untransformed)` ARMS units deployed across the Hawaiian Islands (in `r format_range(cc$crab_data$depth)` m depth) yielded a total of `r num(sum(sample_sums(cc$crabs_untransformed)),1)` adult crabs.
-Units contained `r num(all_summary$abundance_mean,0.1)` $\pm$ `r num(all_summary$abundance_sd,0.1)` (mean $\pm$ SD) individuals, and abundance ranged from `r format_range(all_summary$abundance_range)`.
-Crabs were identified to `r sum(unlist(all_summary$types_breakdown))` distinct taxa (`r all_summary$types_breakdown$species` species-level identifications, `r all_summary$types_breakdown$genus` genus, `r all_summary$types_breakdown$family` family) encompassing `r all_summary$taxonomy_breakdown$genera` genera in `r all_summary$taxonomy_breakdown$families` families, which accounted for ~`r pc(all_summary$types_breakdown$species / hawaiian_crabs,0.1)` of the known crab diversity reported from Hawaiian waters for species-level IDs [@castroCatalogAnomuranBrachyuran2011].
-The four most abundant species were: `r with(all_summary,top_list(top,top_n,top_units,fig="fig_crabs"))`.
-There were `r all_summary$singleton_count` singleton taxa (taxa occurring in a single unit across all sites), accounting for ~`r pc(all_summary$singleton_count/ntaxa(cc$crabs_untransformed),0.1)` of total recovered crab taxa.
-For a breakdown of specific crab taxa recovered by island and deployment depth, see [@tbl:species_table_s].
-
-![Representative photographs of the four most abundant crab species found across all deployed ARMS units (specimens shown were not collected as part of this study).
-(a) "*Chlorodiella laevissima*" (FLMNH 45490, Kāne&#x02BB;ohe, Hawai&#x02BB;i).
-(b) "*Perinia tumida*" (FLMNH 45652, Kāne&#x02BB;ohe, Hawai&#x02BB;i).
-(c) *Percnon abbreviatum* (FLMNH 45628, Kāne&#x02BB;ohe, Hawai&#x02BB;i).
-(d) *Dynomene hispida* (FLMNH 69204, Manawai, Hawai&#x02BB;i).
-Photo credits: Florida Museum of Natural History, licensed [CC BY-NC 3.0](https://creativecommons.org/licenses/by-nc/3.0/)](data/images/4crabs.png){#fig:fig_crabs}
-
-<br><br>
-
-### Mesophotic ARMS {-}
-
-The `r nsamples(cc$crabs_deep_untransformed)` units deployed at mesophotic depths $\geq$ 30 m on O&#x02BB;ahu yielded fewer species and individuals compared to the complete dataset, with `r num(sum(sample_sums(cc$crabs_deep_untransformed)),1)` crabs overall, `r sum(unlist(deep_summary$types_breakdown))` distinct taxa (`r deep_summary$types_breakdown$species` species-level identifications, `r deep_summary$types_breakdown$genus` genus, `r deep_summary$types_breakdown$family` family), `r num(deep_summary$abundance_mean,0.1)` $\pm$ `r num(deep_summary$abundance_sd,0.1)` individuals per unit, and abundance ranging from `r format_range(deep_summary$abundance_range)`.
-The four most abundant taxa in the deep ARMS did not overlap with the most abundant taxa from the complete dataset: `r with(deep_summary,top_list(top,top_n,top_units))`.
-There were `r deep_summary$singleton_count` singletons, which proportionally accounted for ~`r pc(deep_summary$singleton_count/ntaxa(cc$crabs_deep_untransformed),0.1)` of species found on deep ARMS.
-
-### Comparisons {-}
-```{r depth-ranges, results="hide", dependson="taxonomy-validation"}
-f <- cc$crabs %>%
-  psmelt() %>%
-  group_by(depth_zone) %>%
-  filter(Abundance > 0) %>%
-  ungroup() %>%
-  group_by(family) %>%
-  filter(n_distinct(shallow_deep) == 1) %>%
-  ungroup()
-
-fam <- f %>%
-  distinct(family,shallow_deep) %>%
-  count(shallow_deep) %>%
-  deframe()
-
-deep_fams <- f %>%
-  filter(shallow_deep == "deep") %>%
-  pull(family) %>%
-  unique() %>%
-  sort()
-
-g <- cc$crabs %>%
-  psmelt() %>%
-  filter(!is.na(genus)) %>%
-  group_by(depth_zone) %>%
-  filter(Abundance > 0) %>%
-  ungroup() %>%
-  group_by(genus) %>%
-  filter(n_distinct(shallow_deep) == 1) %>%
-  ungroup()
-
-gen <- g %>%
-  distinct(genus,shallow_deep) %>%
-  count(shallow_deep) %>%
-  deframe()
-
-deep_gens <- g %>%
-  filter(shallow_deep == "deep") %>%
-  pull(genus) %>%
-  unique() %>%
-  sort()
-deep_gens <- str_glue("*{deep_gens}*")
-
-s <- cc$crabs %>%
-  psmelt() %>%
-  group_by(depth_zone) %>%
-  filter(Abundance > 0) %>%
-  ungroup() %>%
-  group_by(species) %>%
-  filter(n_distinct(shallow_deep) == 1) %>%
-  ungroup()
-
-spp <- s %>%
-  distinct(species,shallow_deep) %>%
-  count(shallow_deep) %>%
-  deframe()
-
-deep_spp <- s %>%
-  filter(shallow_deep == "deep") %>%
-  arrange(species) %>%
-  select(OTU) %>%
-  left_join(crab_taxonomy_validated,by=c("OTU" = "otu")) %>%
-  pull(display_species) %>%
-  unique()
-```
-Out of the `r all_summary$taxonomy_breakdown$families` families recovered, `r fam['shallow']` were exclusive to shallow ARMS recovered from `r dr(cc$crabs_shallow_untransformed)` m and `r fam['deep']` were exclusive to units $\geq$ 30 m (`r format_list(deep_fams)`; [@tbl:family_depth_table_s]).
-`r str_to_sentence(english(gen['shallow']))` genera were found exclusively shallow and `r gen['deep']` exclusively deep (`r format_list(deep_gens)`; [@tbl:genus_depth_table_s]).
-`r str_to_sentence(english(spp['shallow']))` distinct taxa (including those collapsed to genus or family) were exclusive to shallow ARMS units and `r spp['deep']` were found only deep (`r format_list(deep_spp)`; [@tbl:species_depth_table_s]).
-For a more in-depth taxonomic treatment of the assemblage composition of mesophotic vs.
-shallow ARMS units, see Hurley et al.
-[-@hurleyAssessmentShallowMesophotic2016].
-
-## Alpha diversity {-}
-
-### Summary {-}
-
-Overall mean species richness for the complete dataset was `r num(all_alpha$richness,0.1)` $\pm$ `r num(all_alpha$richness_sd,0.1)` species per unit (effective Simpson diversity: `r num(all_alpha$simpson,0.01)` $\pm$ `r num(all_alpha$simpson_sd,0.01)`).
-Overall mean species richness for shallow ARMS \< 30 m was `r num(shallow_alpha$richness,0.1)` $\pm$ `r num(shallow_alpha$richness_sd,0.1)` species per unit (effective Simpson diversity: `r num(shallow_alpha$simpson,0.01)` $\pm$ `r num(shallow_alpha$simpson_sd,0.01)`).
-Overall mean species richness for mesophotic ARMS $\geq$ 30 m was `r num(deep_alpha$richness,0.1)` $\pm$ `r num(deep_alpha$richness_sd,0.1)` species per unit (effective Simpson diversity: `r num(deep_alpha$simpson,0.01)` $\pm$ `r num(deep_alpha$simpson_sd,0.01)`).
-
-### Comparisons {-}
-
-For the complete dataset, there was no significant difference in observed species richness (`r k_test(all_alpha$k_r_island)`; [@fig:fig_alpha[]{}a]) or effective Simpson diversity (`r k_test(all_alpha$k_s_island,plain=TRUE)`; [@fig:fig_alpha[]{}c]) between the MHI and NWHI.
-Richness was not significantly different between shallow and deep samples (`r k_test(all_alpha$k_r_depth,plain=TRUE)`; [@fig:fig_alpha[]{}b]) but effective Simpson diversity was (`r k_test(all_alpha$k_s_depth,plain=TRUE)`; [@fig:fig_alpha[]{}d]).
-When comparing diversity across sampling depths for only sites in the MHI, richness was not significantly different (`r k_test(all_alpha$k_r_depth_main,plain=TRUE)`; [@fig:fig_alpha_main[]{}a]) but effective Simpson diversity was (`r k_test(all_alpha$k_s_depth_main,plain=TRUE)`; [@fig:fig_alpha_main[]{}c]).
-Neither richness nor Simpson diversity were significantly different between shallow and deep when comparisons were constrained to the island of O&#x02BB;ahu (Richness: `r k_test(all_alpha$k_r_depth_oahu,plain=TRUE)`, [@fig:fig_alpha_main[]{}b]; Simpson: `r k_test(all_alpha$k_s_depth_oahu,plain=TRUE)`, [@fig:fig_alpha_main[]{}d]).
-When only shallow arms \< 30 m were considered, there were no significant differences in alpha diversity between the MHI and NWHI for either index (richness: `r k_test(shallow_alpha$k_r_island,plain=TRUE)`, [@fig:fig_alpha2[]{}a]; effective Simpson: `r k_test(shallow_alpha$k_s_island,plain=TRUE)`, [@fig:fig_alpha2[]{}b]).
-
-<!-- ### Effects of environmental variables {-} -->
-## Brachyuran community structure {-}
-
-```{r dbrda-analysis, results="hide", dependson="init"}
-library(formula.tools)
-
-# set random seed for consistency
-set.seed(32767)
-
-reduce_model <- function(m,o,d,vif_limit=3,condition=character()) {
-  vifs <- vif.cca(m)
-  if (any(vifs > vif_limit)) {
-    oldterms <- rhs.vars(m$call$formula)
-    v <- names(vifs)[which.max(vifs)]
-    newterms <- grep(v,oldterms,value=TRUE,invert=TRUE)
-    f <- reformulate(newterms,response=quote(o))
-    newmod <- dbrda(f,data=d,distance="bray")
-    reduce_model(newmod,o,d,vif_limit,condition)
-  } else {
-    return(m)
-  }
-}
-
-# calculate full dbRDA model
-db_modl <- dbrda(
-  cc$crab_otus_shallow ~ depth +
-    chl + sst + slope + coral_cover +
-    larval_immigration + human_impact +
-    lat + lon +
-    Condition(recovery_year),
-  data=cc$crab_data_shallow,
-  # data=dd,
-  distance="bray"
-)
-
-# reduce model to terms with VIF <= 3
-db_modl <- reduce_model(db_modl,cc$crab_otus_shallow,cc$crab_data_shallow,vif_limit = 3)
-
-# calculate biplot vectors and magnitudes
-vectors <- db_modl %>%
-  scores(display="bp") %>%
-  as_tibble(rownames="term") %>%
-  inner_join(model_terms %>% select(variable,driver=`Environmental Driver`),by=c("term" = "variable")) %>%
-  mutate(size = sqrt(dbRDA1^2 + dbRDA2^2), driver=str_to_lower(driver)) %>%
-  arrange(desc(size))
-
-# get marginal term information and sort by descending f-statistic
-modl_perm <- anova(db_modl,by="margin",permutations = PERM) %>%
-  # as_tibble(rownames="term") %>%
-  # janitor::clean_names() %>%
-  broom::tidy() %>%
-  arrange(desc(statistic))
-
-# make a nicer display-worthy table for significant terms
-rda_table <- modl_perm %>%
-  inner_join(model_terms %>% select(variable,driver=`Environmental Driver`),by=c("term" = "variable")) %>%
-  select(-term) %>%
-  select(Term=driver,DF=df,`Sum of Squares`=SumOfSqs,`F-statistic`=statistic,`p-value`=p.value)
-
-# get number of model terms
-tt <- attr(db_modl$terms,"term.label")
-nterm <- length(grep("Condition",tt,ignore.case = FALSE,value=TRUE,invert = TRUE))
-
-# get top drivers for the text
-drivers <- rda_table %>%
-  slice(1:2) %>%
-  select(term=Term,stat=`F-statistic`,pval=`p-value`)  %>%
-  mutate(
-    term=str_to_lower(term),
-    term = str_replace_all(term,r'(\b[a-z]\b)',replacement=str_to_upper)
-  )
-
-all_terms <- english::english(nrow(model_terms))
-```
-### Environmental drivers {-}
-Environmental variables were strongly clustered according to island group (MHI vs NWHI) ([@fig:fig_pca]).
-The best dbRDA model (`r permanova(db_modl,stat="F")`) explained `r pc(RsquareAdj(db_modl)$adj.r.squared,0.01)` (adjusted $\mathrm{R}^2$) of the variability in crab communities across the archipelago ([@fig:fig_dbrda]).
-`r str_to_sentence(english(nterm))` of the `r str_to_lower(english(nrow(model_terms)))` environmental variables tested were significant and sufficiently uncorrelated.
-The two drivers contributing most to explained variation were `r drivers$term[1]` (F = `r num(drivers$stat[1],0.01)`) and `r drivers$term[2]` (F = `r num(drivers$stat[2],0.01)`) ([@tbl:rda_table_s]).
-These factors also produced the longest effect vectors in the dbRDA plot ([@fig:fig_dbrda]).
-
-```{r fig-dbrda, fig.cap="dbRDA plot showing significant drivers of crab community variation across the Hawaiian Islands.
-Colored ellipses represent 95% confidence intervals around island groups.", results="asis", fig.width=9, fig.height=6, dependson="rda-analysis"}
-termz <- model_terms %>%
-  select(variable,display=`Environmental Driver`)
-
-# make the initial ordination plot
-rda_plotz <- plot_ord(cc$crabs_shallow,db_modl,scale=2,color="island_group",labelsize=3,term_map=termz) +
-  stat_ellipse(geom="polygon",aes(group=island_group,fill=island_group),color=NA,alpha=0.1,show.legend = FALSE) +
-  geom_point(aes(fill=island_group),shape=21,color="black",size=3) +
-  plotz_theme() +
-  scale_fill_manual(name="Island group",values=island_group_pal,labels = c("Main","Northwest")) +
-  guides(color = "none")
-
-# calculate variation explained (constrained & unconstrained)
-eigs <- eigenvals(db_modl)
-constrained <- eigs[str_detect(names(eigs),"dbRDA")]
-constrained <- constrained / sum(constrained)
-unconstrained <- eigs/sum(eigs)
-
-# function to format axis titles
-splained <- function(name,constrained,total,accuracy=0.1) {
-  str_glue("{name} ({pc(constrained,accuracy)} constrained, {pc(total,accuracy)} total)")
-}
-
-axes <- str_glue_data(list(num=1:2),"dbRDA{num}")
-axis_var <- splained(axes,constrained[1:2],unconstrained[1:2])
-
-# put the arrows on top
-rda_plotz$layers <- rda_plotz$layers[c(1,2,5,6,3,4)]
-
-# plot the plotz
-rda_plotz +
-  xlab(axis_var[1]) + ylab(axis_var[2])
-cat("{#fig:fig_dbrda}")
-```
-
-<br><br>
-
-### Regional & depth comparisons {-}
-
-```{r permanova, dependson="init"}
-anova_region <- adonis2(cc$crab_dist_shallow ~ island_group,data=cc$crab_data_shallow,permutations=PERM)
-anova_island <- adonis2(cc$crab_dist_shallow ~ island,data=cc$crab_data_shallow,permutations=PERM)
-bd_region <- betadisper(cc$crab_dist_shallow,cc$crab_data_shallow$island_group)
-
-anova_depth <- adonis2(cc$crab_dist ~ depth_zone,data=cc$crab_data,permutations=PERM)
-anova_depth_oahu <- adonis2(cc$crab_dist_oahu ~ depth_zone,data=sample_tibble(cc$crabs_oahu),permutations=PERM)
-anova_depth2_oahu <- adonis2(cc$crab_dist_oahu ~ shallow_deep,data=sample_tibble(cc$crabs_oahu),permutations=PERM)
-bd_depth <- betadisper(cc$crab_dist,cc$crab_data$depth_zone)
-
-anova_both <- adonis2(cc$crab_dist ~ island_group + depth_zone, data=cc$crab_data,permutations = PERM, by="margin")
-effects <- broom::tidy(anova_both) %>%
-  filter(!term %in% c("Residual","Total")) %>%
-  pull(R2)
-effect_ratio <- max(effects)/min(effects)
-
-cd <- cc$crabs_deep %>%
-  sample_tibble()
-bd_deep <- betadisper(distance(cc$crabs_deep,"bray"),cd$depth_zone)
-```
-
-```{r cap-analysis,dependson="init",echo=FALSE,message=FALSE,warning=FALSE,results="hide"}
-# convert otu tables and sample data into appropriate format
-otus <- otu_table(cc$crabs) %>%
-  as.data.frame()
-cd <- cc$crab_data %>%
-  mutate(sample_group = factor(sample_group,ordered=FALSE)) %>%
-  as.data.frame()
-
-# run CAP analysis, taking care of negative eigenvalues where necessary
-cap_depth <- CAPdiscrim(otus ~ sample_group, data=cd, dist="bray", axes=2, m=0, add=TRUE)
-# # calculate proportion of explained variance
-cap_var <- cap_depth$manova$Eigenvalues / sum(cap_depth$manova$Eigenvalues)
-cap_var <- list(x = cap_var[1], y = cap_var[2])
-
-# make the sample group look nice
-nice <- function(s) {
-  str_replace(s,"shallow","Shallow") %>%
-    str_replace("main","(MHI)") %>%
-    str_replace("nwhi","(NWHI)") %>%
-    str_replace("(^[0-9]+)","O‘ahu \\1") %>%
-    str_split("((?<=[0-9])(?=[a-zA-Z]))|(_)") %>%
-    map_chr(~str_c(.x,collapse=" "))
-}
-
-# get all the stuff we need to plot
-pcap <- cap_depth$x %>%
-  as_tibble(rownames="sample") %>%
-  left_join(cd,by="sample") %>%
-  mutate(
-    sample_group = nice(sample_group),
-    sample_group = fct_reorder(sample_group,depth)
-  )
-
-shallow_cd <- pcap %>%
-  filter(shallow_deep == "shallow") %>%
-  select(sample,LD1,LD2) %>%
-  column_to_rownames("sample") %>%
-  centroid_dist(mean=FALSE)
-
-deep_cd <- pcap %>%
-  filter(shallow_deep == "deep") %>%
-  select(sample,LD1,LD2) %>%
-  column_to_rownames("sample") %>%
-  centroid_dist(mean=FALSE)
-
-centroid_t <- t.test(deep_cd,shallow_cd,alternative = "g",var.equal = FALSE)
-
-oahu_dist <- pcap %>%
-  filter(island == "Oahu") %>%
-  select(sample,LD1,LD2) %>%
-  column_to_rownames("sample") %>%
-  dist() %>%
-  c()
-
-shallow_dist <- pcap %>%
-  filter(shallow_deep == "shallow") %>%
-  select(sample,LD1,LD2) %>%
-  column_to_rownames("sample") %>%
-  dist() %>%
-  c()
-
-deep_dist <- pcap %>%
-  filter(shallow_deep == "deep") %>%
-  select(sample,LD1,LD2) %>%
-  column_to_rownames("sample") %>%
-  dist() %>%
-  c()
-
-distance_t <- t.test(deep_dist,shallow_dist,alternative = "g",var.equal = FALSE)
-oahu_t <- t.test(oahu_dist,shallow_dist,alternative = "g",var.equal = FALSE)
-```
-Shallow crab community structure was significantly different between island groups (MHI vs.
-NWHI) (`r permanova(anova_region,0.01)`) as well as among individual islands (`r permanova(anova_island,0.01)`; see [@fig:fig_pairwise] for results of pairwise comparisons).
-There was no significant difference in community dispersion between the MHI and NWHI (`r permdisp(bd_region)`).
-When mesophotic ($\geq$ 30 m) ARMS units were included, there were significant differences among depth zones (\< 30 m, 30 m, 60 m, 90 m) both for all samples (`r permanova(anova_depth,0.01)`) and samples restricted to O&#x02BB;ahu (`r permanova(anova_depth_oahu,0.01)`), and community dispersion (`r permdisp(bd_depth)`).
-However, community dispersion was not different among depth zones $\geq$ 30 m (`r permdisp(bd_deep)`).
-Although crab communities were significantly different among both depth zones and island groups for the complete dataset (`r permanova(anova_both,0.01)`), depth explained `r num(effect_ratio,0.1)` times more variability than island group (based on model $\mathrm{R}^2$).
-
-In the CAP analysis of island group and depth zone, ARMS units were correctly reassigned to sample groups `r pc(cap_depth$percent/100,0.1)` of the time.
-The first two linear discriminant axes explained `r pc(cap_var$x,0.1)` and `r pc(cap_var$y,0.1)` of the variation, respectively.
-Mesophotic samples (30--90 m) from O&#x02BB;ahu grouped separately from the shallow samples collected across the entire ~2,500 km expanse of the archipelago ([@fig:fig_cap]), whereas shallow samples from O&#x02BB;ahu primarily clustered with shallow samples from other islands ([@fig:fig_cap]).
-Variability among 30--90 m samples was apparently greater than among all shallow samples, with mean Euclidean distances between deep samples in the CAP ordination significantly greater than between shallow samples (`r t_test(distance_t)`; [@fig:fig_cap; @fig:fig_euclid]).
-Similarly, variability among samples from O&#x02BB;ahu (including mesophotic depths) exceeded variability among all shallow samples combined (based on Euclidean distance among CAP points, as above; `r t_test(oahu_t,plain=TRUE)`).
-
-```{r fig-cap, fig.cap="CAP ordination of crab communities by island group and depth zone.", results="asis", fig.width=9, fig.height=6, dependson="cap-analysis"}
-centroids <- pcap %>% group_by(sample_group) %>%
-  summarise(LD1=mean(LD1),LD2=mean(LD2))
-pcap <- pcap %>%
-  mutate(
-    oahu = factor(
-      # if_else(island == "Oahu","oahu","noahu"),
-      case_when(
-        island == "Oahu" ~ "oahu",
-        island == "French Frigate Shoals" ~ "ffs",
-        .default = "noahu"
-      ),
-      levels = c("oahu","ffs","noahu")
-    )
-  )
-
-ggplot(pcap,aes(x=LD1,y=LD2,fill=sample_group)) +
-  stat_ellipse(geom="polygon",level=0.95,alpha=0.1,show.legend = FALSE) +
-  # geom_point(aes(shape=oahu),color="black",shape=21,size=5)  +
-  geom_point(mapping=aes(shape=oahu),color="black",size=5)  +
-  geom_label_repel(data=centroids,aes(label=sample_group), color="white", show.legend = FALSE) +
-  scale_shape_manual(values=c("oahu" = 23, "ffs" = 22, "noahu" = 21),name="Island", labels=c("noahu"="Other islands","oahu"="O‘ahu","ffs"="Lalo (French Frigate Shoals)" )) +
-  scale_fill_manual(values=sample_depth_pal,name="Sample group") +
-  guides(
-    fill=guide_legend(override.aes=list(shape=21)),
-    shape=guide_legend(override.aes=list(fill="black"))
-  ) +
-  xlab(str_glue("Linear discriminant 1 [{pc(cap_var$x,0.1)}]")) +
-  ylab(str_glue("Linear discriminant 2 [{pc(cap_var$y,0.1)}]")) +
-  plotz_theme()
-
-# centroids <- pcap %>% group_by(sample_group) %>%
-#   summarise(LD1=mean(LD1),LD2=mean(LD2))
-# pcap <- pcap %>%
-#   mutate(
-#     oahu = factor(
-#       if_else(island == "Oahu","oahu","noahu"),
-#       levels = c("oahu","noahu")
-#     )
-#   )
-
-# ggplot(pcap,aes(x=LD1,y=LD2,fill=sample_group)) +
-#   stat_ellipse(geom="polygon",level=0.95,alpha=0.1,show.legend = FALSE) +
-#   # geom_point(aes(shape=oahu),color="black",shape=21,size=5)  +
-#   geom_point(mapping=aes(shape=oahu),color="black",size=5)  +
-#   geom_label_repel(data=centroids,aes(label=sample_group), color="white", show.legend = FALSE) +
-#   scale_shape_manual(values=c("oahu" = 23, "noahu" = 21),name="Island", labels=c("noahu"="Other islands","oahu"="O‘ahu")) +
-#   scale_fill_manual(values=sample_depth_pal,name="Sample group") +
-#   guides(
-#     fill=guide_legend(override.aes=list(shape=21)),
-#     shape=guide_legend(override.aes=list(fill="black"))
-#   ) +
-#   xlab(str_glue("Linear discriminant 1 [{pc(cap_var$x,0.1)}]")) +
-#   ylab(str_glue("Linear discriminant 2 [{pc(cap_var$y,0.1)}]")) +
-#   plotz_theme()
-cat("{#fig:fig_cap}")
-```
-
-<br><br>
-
-# Discussion {-}
-The Hawaiian Archipelago is among the most remote oceanic island chains on the planet and encompasses a range of gradients (anthropogenic, biotic, abiotic), making it an ideal natural laboratory to study the impacts of various drivers of species assemblages across space.
-Human activities are considerably higher in the populated MHI than in the protected and remote NWHI [@selkoeEvaluatingAnthropogenicThreats2008; @selkoeMapHumanImpacts2009; @selkoeDNACoralReef2016].
-Although we did not consider the effects of individual impacts here, an index representing *cumulative* anthropogenic impacts was among the top drivers of shallow brachyuran community structure, similar to findings in previous work examining genetic diversity as it relates to species richness and habitat area, quality, and stability across the archipelago [@selkoeDNACoralReef2016].
-The next-most significant driver, chlorophyll-A, can be used as a proxy for primary productivity.
-Productivity gradients across the Hawaiian Archipelago coincide with differences in coral reef community composition [@selkoeDNACoralReef2016], endemic reef fish dominance [@friedlanderDominanceEndemicsReef2020], survival rates of top predators [@bakerEffectVariableOceanic2007], and island geomorphology [@goveNearislandBiologicalHotspots2016].
-Although not a significant factor in our model, coral cover and species composition are thought to affect patterns of benthic assemblages.
-Due to the limited availability of datasets covering our entire study area, our model only included coral cover at the island scale, but previous work has shown that coral species composition does not vary considerably across the archipelago (aside from the presence of *Acropora* in the NWHI) [@griggCommunityStructureSuccession1983].
-Crab communities from Lalo (French Frigate Shoals), which was found to harbor the highest coral species diversity [@griggCommunityStructureSuccession1983], did not group separately from the rest of our samples ([@fig:fig_cap]).
-Given the prevalence in our model of significant factors known to be important influencers of marine communities in Hawai&#x02BB;i, it is not surprising that we find the same top drivers of community structure for brachyuran crabs, one of the primary taxonomic groups inhabiting coral reefs.
-What is surprising---given the maximum vertical separation among shallow ARMS units of `r diff(range(cc$crab_data_shallow_unscaled$depth))` m---is the consistent significance of depth as a driver of community composition across these gradients.
-
-Although our shallow-water ARMS units were deployed within a narrow depth range (`r dr(cc$crabs_shallow_untransformed)` m), depth still emerged as a significant and consistent driver of community variation among these units.
-Similar fine-scale depth patterns have been found in previous studies, both using molecular [e.g., @desouzaImportanceDepthTemperature2023; @westEDNAMetabarcodingSurvey2020] and traditional survey methods [e.g., @williamsBenthicCommunitiesTwo2013; @srinivasanDepthDistributionsCoral2003], but our results recover them alongside numerous other geographic and environmental gradients.
-Depth in and of itself is unlikely to affect marine community composition and oceanographic factors with well-known depth profiles (e.g., temperature, oxygen, nutrients, etc.) are not expected to vary dramatically across such a narrow depth range (\< 10 m), although internal waves can deliver variable temperatures across shallow reefs [@storlazziInternalTidesCan2020; @wyattHeatAccumulationCoral2020] and fine-scale habitat variability among sites may also influence the community differences we found [@steyaertRemoteReefCryptobenthic2022].
-Unfortunately, we lack high resolution environmental data at depth for individual deployments, and thus cannot unravel the co-varying factor(s) associated with the depth effects we observed for shallow assemblages.
-Regardless, given its significance across both narrow (\< 10 m) and wider (90 m) depth ranges sampled here, a factor associated with depth is clearly an important driver for cryptobenthic crab communities across the Hawaiian Archipelago.
-As brachyurans are a ubiquitous and important component of coral reef communities [@reynsPatternsProcessesBrachyuran1999; @lasleyjrPhylogeneticRelationshipsUbiquitous2015; @hazeriLatitudinalSpeciesDiversity2019], these findings are likely to apply more broadly as well.
-
-While depth was a significant driver of community composition among shallow-water ARMS units ([@fig:fig_dbrda; @tbl:rda_table_s]), its influence was considerably more striking when comparing shallow to mesophotic samples.
-In our CAP ordination, all shallow assemblages across the archipelago clustered separately from the deep assemblages $\geq$ 30 m at the single island of O&#x02BB;ahu ([@fig:fig_cap]).
-Additionally, the two deepest sampled depths (60 m and 90 m) clustered apart from 30 m.
-According to this analysis, there was more community variation among samples spanning 90 m of depth on a single island than across ~2,500 km of shallow habitat experiencing multiple gradients of environmental, ecological, and anthropogenic influence.
-This is further supported by the fact that average Euclidean distance in the CAP among O&#x02BB;ahu samples (including mesophotic) was significantly greater than the average distance among all shallow samples (MHI & NWHI).
-Shallow crab communities from O&#x02BB;ahu appear to share more affinity with Hōlanikū (~2,200 km distant) than they do with those at 90 m depth on the same reef slope ([@fig:fig_cap]).
-The CAP also revealed two groups within the mesophotic samples: one at 30 m and the other encompassing the 60--90 m depths.
-These depth zones correspond with the upper and lower mesophotic zones, respectively [@kahngCommunityEcologyMesophotic2010; @rochaMesophoticCoralEcosystems2018; but see @pyleMesophoticCoralEcosystems2019].
-The upper mesophotic zone has been found to constitute an extension of shallow reefs (with beta diversity patterns largely driven by nestedness), whereas the lower mesophotic harbors unique communities (with beta diversity driven primarily by turnover) [@rochaMesophoticCoralEcosystems2018; @slatteryFunctionStabilityMesophotic2024a; @bellwoodRegionalscaleAssemblyRules2001; @kahngCommunityEcologyMesophotic2010].
-The evident groupings in the CAP ordination match these trends, with 30 m samples lying closer to shallow O&#x02BB;ahu samples and 60--90 m samples forming a distinct cluster.
-This pattern of community transition from shallow to mesophotic reefs is consistent with what has been shown for various other taxonomic groups [@lesserGlobalCommunityBreaks2019; @rochaMesophoticCoralEcosystems2018], lending support to the hypothesis that there may be broad-scale depth-based assembly rules for coral reef communities.
-
-In addition to community composition, mesophotic samples also exhibited alpha diversity differences.
-Effective Simpson diversity was significantly higher in deep communities (for comparisons among all islands and solely within the MHI, but not exclusively at O&#x02BB;ahu) ([@fig:fig_alpha; @fig:fig_alpha_main]).
-This pattern is in contrast to observations that biodiversity often decreases with depth [@pyleComprehensiveInvestigationMesophotic2016; @kahngCommunityEcologyMesophotic2010], although the differences we found were slight and our findings may be influenced by differences in habitat availability across depth zones (especially of the habitat types ARMS are designed to mimic).
-
-Since we lack environmental data at mesophotic depths, we cannot make conclusions about the specific influences driving the differences we found.
-However, light and temperature have been proposed as important bottom-up drivers of depth zonation in benthic communities [@kahngCommunityEcologyMesophotic2010; @lesserGlobalCommunityBreaks2019; @pyleComprehensiveInvestigationMesophotic2016; @pyleMesophoticCoralEcosystems2019].
-Oceanographic conditions may be inconsistent close to shore, but the average mixed-layer depth calculated (using 0.8 &deg;C $\upDelta$T [@karaOptimalDefinitionOcean2000]) from CTD casts taken at station ALOHA (22.75&deg;N, 158.00&deg;W; within the latitudinal expanse of the MHI) was 84.6 $\pm$ 26 m over the course of the year (based on monthly averages from casts taken across the duration of ARMS unit deployments [@HawaiiOceanTimeseries]).
-Thus, given the assumption of similar temperature profiles near the O&#x02BB;ahu shoreline, temperature alone appears insufficient to explain the zonation patterns we observed---although due to its significant role as a driver of shallow community variation, it cannot be ruled out.
-In contrast, the lower depth boundaries of mesophotic ecosystems are largely defined by surface irradiance [@kahngCommunityEcologyMesophotic2010; @hindersteinThemeSectionMesophotic2010], so light seems a possible candidate as bottom-up abiotic factor influencing community structure across depths.
-Crab assemblage structure is unlikely to be directly affected by light variability (although the behavior of individual crabs may).
-However, habitat-building species such as corals, algae, and sponges are frequently subject to light requirements [@kahngCommunityEcologyMesophotic2010; @slatteryFunctionStabilityMesophotic2024a], and may in turn affect the structure of the communities dependent upon them [@hustonCompetitionCoexistenceEffects1994; @smithRoleHabitatComplexity2014; @sebensHabitatStructureCommunity1991].
-If patterns of community structure are driven primarily from the bottom up by light, that may contribute to our observation of a significant depth effect over less than 10 m in our shallow samples.
-
-Two potential weaknesses of our dataset are the time periods over which ARMS units were deployed and the fact that mesophotic samples were obtained solely from the island of O&#x02BB;ahu.
-ARMS units were deployed across a broad range of years, but the majority were deployed during 2010--2012 or 2010--2013 (an overlap of two years) ([@tbl:arms_table]).
-A reduced version of the CAP using only these overlapping deployments retains the same (if not more pronounced) overall patterns as the full version ([@fig:fig_cap_reduced]), suggesting that the inclusion of previous or subsequent soaking periods did not adversely skew our analyses.
-The limitation of our depth transect to O&#x02BB;ahu means that we cannot make broad-scale conclusions about biodiversity patterns on deep reefs across the Hawaiian Islands.
-However, it remains the case that community and alpha diversity differences among depth zones at that single island appear to overwhelm differences observed among shallow-water coral reef habitat across the archipelago.
-Similar patterns were found at a local scale (tens of kilometers) on Hawai&#x02BB;i Island using environmental DNA (eDNA) [@hobanPlumbingDepthsEnvironmental2023].
-In that study, communities were found to have more affinity within depth zones across sites than among depth zones within sites.
-Thus, despite the limitations in our dataset, the patterns we observed suggest that factors associated with the shallow-to-mesophotic depth gradient exert stronger ecological influence than those (including depth) that vary across the latitudinal, physical, and anthropogenic gradients throughout the Hawaiian Archipelago.
-
-Further work to investigate drivers of depth zonation could take advantage of established gradients and use the Hawaiian Archipelago as a natural laboratory, by sampling depth transects across the archipelago.
-Coral reef biodiversity is generally thought to be structured at depth by a combination of the attenuation of light, a decrease in temperature, increasing salinity, and a reduction in trophic resources [@lesserEcologyMesophoticCoral2009; @kahngCommunityEcologyMesophotic2010; @slatteryFunctionStabilityMesophotic2024a].
-If mesophotic communities are found to be different from their shallow counterparts across multiple islands (ideally comparing sites experiencing similar irradiance), this suggests that they are vertically structured by a factor (e.g., light) outside of the numerous natural gradients that exist across the Hawaiian Islands.
-In contrast, other factors (such as temperature, dissolved inorganic nitrogen, particulate organic carbon, or productivity) expected to have more strongly-defined gradients across the regions [@slatteryFunctionStabilityMesophotic2024a], would lead to variable zonation patterns that should be detected in our spatial comparison of the MHI & NWHI.
-Such differences should be tested in future studies to understand the specific physical and ecological drivers of coral reef depth zonation.
-The cost, logistical complexity and time investment required for ARMS deployments make this challenging, but emerging approaches such as environmental DNA (eDNA) may offer a suitable alternative.
-
-Marine communities are structured by a suite of factors, but there remains no consensus for either generalized mechanisms or the relative importance of drivers of community structure for coral reefs.
-Our best knowledge of the role of depth in structuring marine communities comes from the intertidal, and points to biotic factors (e.g., predation, competition) determining lower limits and abiotic factors (e.g., desiccation, thermal stress) setting upper limits of vertical species distributions [@connellEffectsCompetitionPredation1961; @connellInfluenceInterspecificCompetition1961; @dennyPhysicalProcessesThat2001; @paineIntertidalCommunityStructure1974].
-There are numerous examples from other study systems in which the relative contributions of biotic versus abiotic factors to structuring communities run in opposing directions across some habitat gradient (e.g., climate, soil type, elevation, etc.
-[@bradyEvolutionaryEcologyPlant2005; @moore10PlantCompetition2011; @wishingradTemperateZoneIsolation2023]).
-However, abiotic stressors on intertidal communities are obviously different from those on coral reefs, and generalizations about the mechanisms and ecological processes that structure coral reef communities remain a subject of considerable debate [@bellwoodEnvironmentalGeometricConstraints2005; @bellwoodSleepingFunctionalGroup2006; @carrBiodiversityPopulationRegulation2002; @connollyIndoPacificBiodiversityCoral2003; @dornelasCoralReefDiversity2006; @grahamImportanceStructuralComplexity2013; @komyakovaRelativeImportanceCoral2013].
-In the Hawaiian archipelago, we found significant differences in the composition of shallow coral reef crab communities between the populated high islands of the MHI and the uninhabited atolls of the NWHI.
-Moreover, these differences were driven by a combination of environmental (sea-surface temperature, chlorophyll-A concentration, depth), biotic (potential larval immigration), and anthropogenic (cumulative human impacts) factors.
-Despite the significance of these factors in driving shallow community variation across the archipelago, mesophotic assemblages were found to be distinct ([@fig:fig_cap]).
-Although there were variables we did not consider in our model, this suggests that vertical distributions are influenced by factors other than those driving spatial variation across the island chain.
-In light of this, and since mesophotic communities are generally less biodiverse than shallow reefs [@kahngCommunityEcologyMesophotic2010; @lesserEcologyMesophoticCoral2009; @pyleComprehensiveInvestigationMesophotic2016], it may be possible that the gradients driving depth distributions on coral reefs run opposite to those in the intertidal.
-Given the rapid changes in important factors like temperature, nutrients, and light throughout the upper ~100--150 m of the water column (particularly the diminishment of photosynthetically active radiation below 100 m), the *lower* limits of the coral reef mesophotic zone may be determined by abiotic factors that correlate with depth.
-Conversely, the *upper* limits of the mesophotic zone (the transition from shallow to mesophotic habitat), which occur at the higher end of the depth-biodiversity gradient [@kahngCommunityEcologyMesophotic2010; @lesserEcologyMesophoticCoral2009] may be driven by ecological interactions [@vellendEffectsDiversityDiversity2008].
-
-By comparing shallow-water ARMS across the latitudinal, physical and anthropogenic gradients of the Hawaiian Archipelago, we were able to test the relative importance of these factors in driving shallow crab assemblage structure.
-We additionally find that assemblages along a 90 m depth transect at a single island show greater divergence in canonical space than across the full ~2,500 km gradient of the Hawaiian Archipelago on shallow reefs.
-Our analyses corroborate previous work and show that depth is among the strongest drivers of community structure.
-We conclude that some factor(s) covarying with depth exert more influence on cryptobenthic coral reef community structure than those biotic and abiotic gradients that span the entire Hawaiian Archipelago in shallow-water environments.
-
-# Acknowledgements
+# References {-}
+
+<!-- show main bibliography -->
+<div class="multi-refs"></div>
+
+# Acknowledgements {-}
 We thank the NOAA Coral Reef Conservation Program for supporting the deployment and recovery of the shallow ARMS.
@@
 
-# Funding
+# Author contributions {-}
+Conceptualization: MLH, KKH, DJS, RJT; Data curation: MLH; Formal Analysis: MLH, KKH; Funding acquisition: KKH, DJS, RJT; Investigation: MLH, KKH, KR, DJS; Methodology: MLH, KKH, DJS, MAT; Project administration: KKH, DJS, RJT; Resources: KR, MAT, RJT; Software: MLH; Supervision: DJS, RJT; Validation: MLH, MAT, RJT; Visualization: MLH, KR; Writing -- original draft: MLH, KKH, RJT; Writing -- review & editing: MLH, KKH, KR, DJS, MAT, RJT
+
+# Data availability statement {-}
+All data and code used to generate this manuscript and its accompanying figures and tables can be found on github at [https://github.com/mhoban/arms_crabs](https://github.com/mhoban/arms_crabs).
+
+# Additional Information {-}
+
+## Funding {-}
 This work was funded through a combination of grants including NSF OCE-1260169, NSF GRFP DGE-1329626, and NSF OCE-2048457.
@@
 
-# References {-}
-
-<!-- show main bibliography -->
-<div class="multi-refs"></div>
+## Competing interests {-}
+The authors declare no competing interests.
+
+\listoffigures
 
@@
 
-# Supplemental material {label="S"}
+# Supplementary Information {label="S"}
 
@@
 
-```{r fig-alpha, fig.cap='Box plots of (a,b) species richness and (c,d) effective Simpson diversity for complete dataset (all sampled depths).
+```{r fig-supp-alpha, fig.cap='Box plots of (a,b) species richness and (c,d) effective Simpson diversity for complete dataset (all sampled depths).
 Asterisks indicate significant differences (K-W test).
@@
   theme(plot.tag = element_text(face="italic"))
-cat("{#fig:fig_alpha}")
+cat("{#fig:fig_supp_alpha}")
 ```
@@
 
-```{r fig-alpha-main, fig.cap='Box plots of species richness and effective Simpson diversity per depth zone for Main Hawaiian Islands (a,c) and O&#x02BB;ahu (b,d) only.
+```{r fig-supp-alpha-main, fig.cap='Box plots of species richness and effective Simpson diversity per depth zone for Main Hawaiian Islands (a,c) and O&#x02BB;ahu (b,d) only.
 Asterisks and notches as above.
@@
   theme(plot.tag = element_text(face="italic"))
-cat("{#fig:fig_alpha_main}")
+cat("{#fig:fig_supp_alpha_main}")
 ```
@@
 
-```{r fig-alpha2, fig.cap='Box plots of (a) species richness and (b) Effective Simpson diversity by island group for shallow samples only.
-Notches as in [@fig:fig_alpha].
-Both comparisons were non-significant.', results="asis", dev="svg", fig.width=8, fig.height=4, dependson="init"}
+```{r fig-supp-alpha2, fig.cap='Box plots of (a) species richness and (b) Effective Simpson diversity by island group for shallow samples only.
+Notches as above.
+Both comparisons were non-significant.', results="asis", dev="svg", fig.width=8, fig.height=4, dependson="fig-supp-alpha"}
 art <- shallow_alpha$richness_table %>%
@@
   theme(plot.tag = element_text(face="italic"))
-cat("{#fig:fig_alpha2}")
+cat("{#fig:fig_supp_alpha2}")
 ```
@@
 
-```{r fig-pca, fig.cap='Principal component analysis (PCA) of environmental variables by sampling site.
+```{r fig-supp-pca, fig.cap='Principal component analysis (PCA) of environmental variables by sampling site.
 Ellipses represent 95% confidence intervals around island groups.', results='asis', dev='svg', fig.width=8,fig.height=8, dependson='init'}
@@
   plotz_theme()
-cat("{#fig:fig_pca}")
+cat("{#fig:fig_supp_pca}")
 ```
@@
 
-```{r pairwise-island, fig.cap='Results of pairwise PERMANOVA tests by island for shallow crab communities (p-values adjusted according to the false discovery rate [@benjaminiControllingFalseDiscovery1995]).
+```{r fig-supp-pairwise-island, fig.cap='Results of pairwise PERMANOVA tests by island for shallow crab communities (p-values adjusted according to the false discovery rate [@benjaminiControllingFalseDiscovery1995]).
 Significant comparisons presented in bold.', results="asis", dev="svg", fig.width=12, fig.height=9, dependson="init"}
@@
   scale_y_discrete(limits=rev)
-cat("{#fig:fig_pairwise}")
+cat("{#fig:fig_supp_pairwise}")
 ```
@@
 ```
-```{r fig-cap-reduced, fig.cap="CAP ordination of crab communities by island group and depth zone showing only samples deployed during the years 2010--2013.", results="asis", dev="svg", fig.width=9, fig.height=6, dependson="cap-analysis-reduced"}
+```{r fig-supp-cap-reduced, fig.cap="CAP ordination of crab communities by island group and depth zone showing only samples deployed during the years 2010--2013.", results="asis", dev="svg", fig.width=9, fig.height=6, dependson="cap-analysis-reduced"}
 # make the sample group look nice
@@
   plotz_theme()
-cat("{#fig:fig_cap_reduced}")
+cat("{#fig:fig_supp_cap_reduced}")
 ```
@@
 
-```{r fig-euclid, fig.cap='Box plot of Euclidean distances between points in CAP ordination by depth zone.
+```{r fig-supp-euclid, fig.cap='Box plot of Euclidean distances between points in CAP ordination by depth zone.
 Asterisk indicates significant difference (Welch two-sample t-test).', results="asis", dev="svg", fig.width=6, fig.height=4, dependson="cap-analysis"}
@@
   ylab("Euclidean distance between CAP points")
-cat("{#fig:fig_euclid}")
+cat("{#fig:fig_supp_euclid}")
 ```
@@
 
-```{r arms-metadata-table, results="asis", dependson="init"}
+```{r tbl-supp-arms-metadata, results="asis", dependson="init"}
 region_map <- c(
@@
   janitor::clean_names(case="sentence")
-print_table(arms_metadata,"arms_metadata_table","ARMS unit metadata",accuracy=1,hdr_case=NA,format_numbers=FALSE)
+print_table(arms_metadata,"tbl_supp_arms_metadata","ARMS unit metadata",accuracy=1,hdr_case=NA,format_numbers=FALSE)
 ```
@@
 
-```{r species-table, results="asis", dependson="taxonomy-validation"}
+```{r tbl-supp-species, results="asis", dependson="taxonomy-validation"}
 # make table of all sampled taxa
@@
   })
-print_table(species_tbl,"species_table_s","List of crab taxa detected by island and ARMS unit deployment depth",hdr_case="none",accuracy=1)
+print_table(species_tbl,"tbl_supp_species","List of crab taxa detected by island and ARMS unit deployment depth",hdr_case="none",accuracy=1)
 ```
@@
 
-```{r family-depth-table, results="asis", dependson="taxonomy-validation"}
+```{r tbl-family-depth, results="asis", dependson="taxonomy-validation"}
 deep_map <- c(
@@
 
-print_table(crabbo,"family_depth_table_s","Brachyuran families exclusive to shallow or mesophotic habitat",hdr_case="none",accuracy=1)
+print_table(crabbo,"tbl_supp_family_depth","Brachyuran families exclusive to shallow or mesophotic habitat",hdr_case="none",accuracy=1)
 ```
@@
 
-```{r genus-depth-table, results="asis", dependson="taxonomy-validation"}
+```{r tbl-supp-genus-depth, results="asis", dependson="taxonomy-validation"}
 deep_map <- c(
@@
 
-print_table(crabbo,"genus_depth_table_s","Brachyuran genera exclusive to shallow or mesophotic habitat",hdr_case="none",accuracy=1)
+print_table(crabbo,"tbl_supp_genus_depth","Brachyuran genera exclusive to shallow or mesophotic habitat",hdr_case="none",accuracy=1)
 ```
@@
 
-```{r species-depth-table, results="asis", dependson="taxonomy-validation"}
+```{r tbl-supp-species-depth, results="asis", dependson="taxonomy-validation"}
 deep_map <- c(
@@
 
-print_table(crabbo,"species_depth_table_s","Brachyuran taxa exclusive to shallow or mesophotic habitat",hdr_case="none",accuracy=1)
+print_table(crabbo,"tbl_supp_species_depth","Brachyuran taxa exclusive to shallow or mesophotic habitat",hdr_case="none",accuracy=1)
 ```
@@
 
-```{r rda-table,results="asis",dependson="rda-analysis"}
-print_table(rda_table,"rda_table_s","Significant marginal terms in the dbRDA analysis of crab community variation across the Hawaiian Archipelago",hdr_case="none",accuracy=0.001)
+```{r tbl-supp-rda,results="asis",dependson="rda-analysis"}
+print_table(rda_table,"tbl_supp_rda","Significant marginal terms in the dbRDA analysis of crab community variation across the Hawaiian Archipelago",hdr_case="none",accuracy=0.001)
 ```

### 2025-03-22 manuscript.Rmd
abstract shortening and csl modifications

@@
 
-Investigations into the environmental and anthropogenic factors that shape coral reef communities have been historically limited both in depth range and taxonomic scope.
-Despite high biodiversity and the recognized importance of mesophotic habitats and due to logistical challenges, previous studies have focused largely on conspicuous taxa such as fishes and corals and the shallowest ~30 m of their depth ranges.
-Here, we examine the variability patterns of brachyuran crab assemblages recovered from Autonomous Reef Monitoring Structures across the ~2,500 km Hawaiian Archipelago and ~90 m of depth.
-We examine the effects of a suite of environmental, ecological, and anthropogenic factors (latitude, longitude, site depth, deployment years, chlorophyll-A, sea surface temperature, island slope, coral cover, potential larval immigration, cumulative human impacts) on shallow (8--17 m) crab assemblages and compare them with those sampled along a shallow-to-mesophotic depth gradient (12--90 m) on O&#x02BB;ahu.
-Shallow assemblage structure was significantly different between the inhabited Main and relatively pristine Northwest Hawaiian Islands and significantly associated with human impacts, chlorophyll-A, sea-surface temperature, island slope, potential larval immigration, and depth.
-Despite this spatial and environmental variation, community differences along the depth gradient on O&#x02BB;ahu alone were greater than among the shallow assemblages across the entire archipelago.
-This finding suggests that 90 m of depth is a stronger determinant of brachyuran assemblage structure than the latitudinal, environmental, or anthropogenic gradients across the entire span of the Hawaiian Islands.
+Investigations into the factors that shape coral reef communities have been historically limited in depth and taxonomic scope.
+Despite high biodiversity and the recognized importance of mesophotic habitats, and due to logistical challenges, studies have focused largely on conspicuous taxa such as fishes and corals and the shallowest ~30 m.
+Here, we examine the variability of crab assemblages from Autonomous Reef Monitoring Structures across Hawai&#x02BB;i and ~90 m depth.
+We examine the effects of environmental, ecological, and anthropogenic factors (latitude, longitude, depth, years, chlorophyll-A, temperature, slope, coral cover, potential larval immigration, human impacts) on shallow (8--17 m) crab assemblages and compare them with those sampled along a mesophotic depth gradient (12--90 m) on O&#x02BB;ahu.
+Shallow assemblages were significantly different between the inhabited Main and relatively pristine Northwestern Hawaiian Islands and associated with human impacts, chlorophyll-A, sea-surface temperature, slope, potential larval immigration, and depth.
+Despite this variability, differences along the depth gradient on O&#x02BB;ahu alone were greater than among the shallow assemblages across the entire archipelago.
+This finding suggests that 90 m of depth is a stronger determinant of brachyuran assemblage structure than the latitudinal, environmental, or anthropogenic gradients across the entire ~2,500 km span of the Hawaiian Islands.
 

### 2025-03-22 crabs.R
remove commented lines

@@ -8,7 +8,7 @@ source("collapse_taxonomy.R")
 
 # plot an ordination with various options
 plot_ord <- function(ps,ord,arrows=TRUE,sig=FALSE,by="term",perm=999,alpha=0.05,scale=1,labelsize=5,pointsize=3,term_map=NULL,...) {
-  
+
   # map classnames to axis names
   axes <- c(
     "dbrda" = "dbRDA",
@@ -21,11 +21,11 @@ plot_ord <- function(ps,ord,arrows=TRUE,sig=FALSE,by="term",perm=999,alpha=0.05,
     "CCA" = "CCA"
   )
   # figure out axis names and b0rk out if they don't exist
-  axis <- axes[class(ord)[1]] 
+  axis <- axes[class(ord)[1]]
   if (is.na(axis) | is.null(axis)) {
     stop("invalid analysis type")
   }
-  
+
   # get the model terms
   terms <- attr(ord$terms,"term.labels")
   if (arrows & sig) {
@@ -35,11 +35,11 @@ plot_ord <- function(ps,ord,arrows=TRUE,sig=FALSE,by="term",perm=999,alpha=0.05,
       filter(term != "Residual",`Pr(>F)` < alpha) %>%
       pull(term)
   }
-  
+
   # do the initial plot with whatever options were passed in
-  plotz <- plot_ordination(ps,ord,...) + 
+  plotz <- plot_ordination(ps,ord,...) +
     geom_point(size=pointsize)
-  
+
   # plot loading arrows, if requested
   if (arrows) {
     # regex to match term names
@@ -50,62 +50,62 @@ plot_ord <- function(ps,ord,arrows=TRUE,sig=FALSE,by="term",perm=999,alpha=0.05,
       # this is where we scale the arrow endpoints
       mutate(across(where(is.numeric),~.x*scale))
     if (sig) {
-      # if we only want significant terms, this is 
+      # if we only want significant terms, this is
       # where we filter for that
       arrowdf <- arrowdf %>%
         filter(str_detect(labels,pattern))
     }
-    
+
     arrowdf <- arrowdf %>%
       mutate(
         hjust = if_else(.data[[str_c(axis,"1")]] < 0,1,0),
         xnudge = if_else(.data[[str_c(axis,"1")]] < 0,-0.02,0.02)
       )
-    
+
     if (!is.null(term_map)) {
       arrowdf <- arrowdf %>%
         inner_join(term_map,by=c("labels" = "variable")) %>%
[…]

### 2025-03-22 manuscript.Rmd
rearrangements and updates for sci reports

@@
 
-The Hawaiian Archipelago, stretching across ~2,500 km in the North Pacific Ocean, provides a natural laboratory to examine drivers of metazoan community structure.
+The Hawaiian Archipelago, spanning ~2,500 km in the North Pacific Ocean, provides a natural laboratory to examine drivers of metazoan community structure.
 This chain of islands comprises two regions, the inhabited high islands of the main Hawaiian Islands (MHI) and the uninhabited low islands, atolls, and banks of the Northwestern Hawaiian Islands (NWHI) which reside within one of the world's largest marine protected areas---the Papahānaumokuākea Marine National Monument.
@@
 Such standardization can reduce variability in recovered communities that may arise due to local variation in habitat composition and availability [@ransomeImportanceStandardizationBiodiversity2017].
-We targeted brachyuran crabs, a ubiquitous group often used as an exemplar to assess coral reef biodiversity [@knowltonCoralReefBiodiversity2010; @plaisanceDiversityCoralReefs2011; @servisCharacterizingCoralReef2020; @plaisanceReefassociatedCrustaceanFauna2009] and comprising a notable proportion of the Hawaiian coral reef cryptofauna [284 reported species, @castroCatalogAnomuranBrachyuran2011].
+We targeted brachyuran crabs, a ubiquitous group often used as an exemplar to assess coral reef biodiversity [@knowltonCoralReefBiodiversity2010; @plaisanceDiversityCoralReefs2011; @servisCharacterizingCoralReef2020; @plaisanceReefassociatedCrustaceanFauna2009] and comprising a notable proportion of the Hawaiian coral reef cryptofauna (284 reported species [@castroCatalogAnomuranBrachyuran2011]).
 We used assemblages recovered from ARMS units to examine drivers of community composition across the Hawaiian Archipelago.
@@
 For a more in-depth taxonomic treatment of the assemblage composition of mesophotic vs.
-shallow ARMS units, see Hurley et al.
-[-@hurleyAssessmentShallowMesophotic2016].
+shallow ARMS units, see [@hurleyAssessmentShallowMesophotic2016].
 
@@
 ### Environmental drivers {-}
-Environmental variables were strongly clustered according to island group (MHI vs NWHI) (Supplementary [@fig:fig_supp_pca]).
+The PCA of environmental variables indicated strong clustering according to island group (Supplementary [@fig:fig_supp_pca]), with apparent orthogonal variation in variables associated with the MHI vs the NWHI.
 The best dbRDA model (`r permanova(db_modl,stat="F")`) explained `r pc(RsquareAdj(db_modl)$adj.r.squared,0.01)` (adjusted $\mathrm{R}^2$) of the variability in crab communities across the archipelago ([@fig:fig_dbrda]).
 `r str_to_sentence(english(nterm))` of the `r str_to_lower(english(nrow(model_terms)))` environmental variables tested were significant and sufficiently uncorrelated.
-The two drivers contributing most to explained variation were `r drivers$term[1]` (F = `r num(drivers$stat[1],0.01)`) and `r drivers$term[2]` (F = `r num(drivers$stat[2],0.01)`) (Supplementary [@tbl:tbl_supp_rda]).
+The two drivers contributing most to explained variation were `r drivers$term[1]` (F = `r num(drivers$stat[1],0.01)`) and `r drivers$term[2]` (F = `r num(drivers$stat[2],0.01)`) ([@tbl:tbl_rda]).
 These factors also produced the longest effect vectors in the dbRDA plot ([@fig:fig_dbrda]).
@@
 The first two linear discriminant axes explained `r pc(cap_var$x,0.1)` and `r pc(cap_var$y,0.1)` of the variation, respectively.
-Mesophotic samples (30--90 m) from O&#x02BB;ahu grouped separately from the shallow samples collected across the entire ~2,500 km expanse of the archipelago ([@fig:fig_cap]), whereas shallow samples from O&#x02BB;ahu primarily clustered with shallow samples from other islands ([@fig:fig_cap]).
-Variability among 30--90 m samples was apparently greater than among all shallow samples, with mean Euclidean distances between deep samples in the CAP ordination significantly greater than between shallow samples (`r t_test(distance_t)`; [@fig:fig_cap], Supplementary [@fig:fig_supp_euclid]).
-Similarly, variability among samples from O&#x02BB;ahu (including mesophotic depths) exceeded variability among all shallow samples combined (based on Euclidean distance among CAP points, as above; `r t_test(oahu_t,plain=TRUE)`).
+Mesophotic samples (30--90 m) from O&#x02BB;ahu were grouped separately in ordination space from all shallow samples across the archipelago ([@fig:fig_cap]).
+Shallow samples from O&#x02BB;ahu primarily clustered with shallow samples from other islands ([@fig:fig_cap]).
+Variability among 30--90 m samples was apparently greater than among all shallow samples, with mean Euclidean distances between deep samples in canonical space significantly greater than between shallow samples (`r t_test(distance_t)`; [@fig:fig_cap], Supplementary [@fig:fig_supp_euclid]).
+Similarly, variability among samples from O&#x02BB;ahu (including mesophotic depths) exceeded variability among all shallow samples combined (based on Euclidean distances as above; `r t_test(oahu_t,plain=TRUE)`).
 
@@
   scale_shape_manual(values=c("oahu" = 23, "ffs" = 22, "noahu" = 21),name="Island", labels=c("noahu"="Other islands","oahu"="O‘ahu","ffs"="Lalo (French Frigate Shoals)" )) +
-  scale_fill_manual(values=sample_depth_pal,name="Sample group") +
+  scale_fill_manual(values=sample_depth_pal,name="Sample grouping") +
   guides(
@@
 
-# centroids <- pcap %>% group_by(sample_group) %>%
-#   summarise(LD1=mean(LD1),LD2=mean(LD2))
-# pcap <- pcap %>%
-#   mutate(
-#     oahu = factor(
-#       if_else(island == "Oahu","oahu","noahu"),
-#       levels = c("oahu","noahu")
-#     )
-#   )
-
-# ggplot(pcap,aes(x=LD1,y=LD2,fill=sample_group)) +
-#   stat_ellipse(geom="polygon",level=0.95,alpha=0.1,show.legend = FALSE) +
-#   # geom_point(aes(shape=oahu),color="black",shape=21,size=5)  +
-#   geom_point(mapping=aes(shape=oahu),color="black",size=5)  +
-#   geom_label_repel(data=centroids,aes(label=sample_group), color="white", show.legend = FALSE) +
-#   scale_shape_manual(values=c("oahu" = 23, "noahu" = 21),name="Island", labels=c("noahu"="Other islands","oahu"="O‘ahu")) +
-#   scale_fill_manual(values=sample_depth_pal,name="Sample group") +
-#   guides(
-#     fill=guide_legend(override.aes=list(shape=21)),
-#     shape=guide_legend(override.aes=list(fill="black"))
-#   ) +
-#   xlab(str_glue("Linear discriminant 1 [{pc(cap_var$x,0.1)}]")) +
-#   ylab(str_glue("Linear discriminant 2 [{pc(cap_var$y,0.1)}]")) +
-#   plotz_theme()
 cat("{#fig:fig_cap}")
@@
 
-While depth was a significant driver of community composition among shallow-water ARMS units ([@fig:fig_dbrda], Supplementary [@tbl:tbl_supp_rda]), its influence was considerably more striking when comparing shallow to mesophotic habitat.
+While depth was a significant driver of community composition among shallow-water ARMS units ([@fig:fig_dbrda, @tbl:tbl_rda]), its influence was considerably more striking when comparing shallow to mesophotic habitat.
 In our CAP ordination, all shallow assemblages across the archipelago clustered separately from the deep assemblages $\geq$ 30 m at the single island of O&#x02BB;ahu ([@fig:fig_cap]).
@@
 [-@hurleyAssessmentShallowMesophotic2016], in which 27 units were deployed off O&#x02BB;ahu from 2009--2012 in `r format_range(cc$crab_data %>% sample_tibble() %>% filter(island == "Oahu") %>% pull(depth))` m depth; nine shallow units were deployed at 12 m, and six each at depths of 30 m, 60 m, and 90 m ([@tbl:tbl_arms]).
-Shallow units were deployed according to NOAA-RAMP methodology, whereas units $\geq$ 30 m were deployed and recovered using technical mixed gas closed-circuit rebreather diving [see @hurleyAssessmentShallowMesophotic2016].
+Shallow units were deployed according to NOAA-RAMP methodology, whereas units $\geq$ 30 m were deployed and recovered using technical mixed gas closed-circuit rebreather diving (see [@hurleyAssessmentShallowMesophotic2016]).
 Except when discussing specific depth zones, we will refer to any units deployed at `r dr(cc$crabs_shallow_untransformed)` m as "shallow" and those deployed $\geq$ 30 m as "deep".
@@
 cat("{#fig:fig_map}")
+```
+
+<br><br>
+
+ARMS units were retrieved by encapsulating them in-situ in a 100 μm mesh-lined crate to prevent escape of motile organisms.
+Once returned to the surface, they were placed in containers of aerated, filtered seawater, crates were removed, and units were systematically disassembled plate-by-plate (see [@lerayDNABarcodingMetabarcoding2015] for processing details).
+All adult brachyuran crabs ($\geq$ 2 mm) were photographed, identified, preserved in 95% ethanol, and retained in a specimen archive at the Hawai&#x02BB;i Institute of Marine Biology.
+Specimens were initially identified to family or genus using Poore & Ahyong [-@pooreMarineDecapodCrustacea2023] and additionally to species where possible according to expert advice from Scott Godwin (former curator of invertebrate zoology, Bishop Museum) and Gustav Paulay (curator of invertebrate zoology, Florida Museum of Natural History).
+Where species-level identifications were not possible, taxa were collapsed to family or genus (depending on individual identification).
+We recognize that the species-level taxonomy of Indo-Pacific brachyurans remains poorly resolved, particularly for abundant groups like the Chlorodiellinae [@lasleyjrIntegrativeSystematicRevision2014; @lasleyjrPhylogeneticRelationshipsUbiquitous2015], and some "species" are known to comprise complexes [e.g.
+"*Perinia tumida*", @servisCharacterizingCoralReef2020; @plaisanceReefassociatedCrustaceanFauna2009].
+However, the prevalence of species complexes and cryptic diversity suggests that our assignments are likely to under- rather than overrepresent the true number of taxa.
+Thus, any significant patterns we detected likely persist in spite of the potential presence of cryptic diversity.
+Since our primary concern was categorical diversity of assemblages rather than accurate naming, we have retained species names in cases where complexes are suspected.
+
+## Data analysis {-}
+
+Unless otherwise specified, all analyses were performed using `R` version `r R.version$major`.`r R.version$minor` [@rcoreteamLanguageEnvironmentStatistical2021].
+Community matrices were loaded and encapsulated into data objects using the R package `phyloseq` [@mcmurdiePhyloseqPackageReproducible2013] and multivariate community analyses were performed using the package `vegan` [@oksanenVeganCommunityEcology2022] on Bray-Curtis dissimilarities of Hellinger-transformed species abundance matrices [@legendreEcologicallyMeaningfulTransformations2001; @raoReviewCanonicalCoordinates1995].
+
+To examine patterns of crab diversity across the islands, we used `estimate_richness` to calculate species richness and effective Simpson diversity (effective Simpson = $\frac{1}{\sum_{i=1}^{R}p_i^2}$, where R is the total number of species, and $\mathrm{p}_i$ is the proportional abundance of each species).
+The effective Simpson index (the inverse of the "classic" Simpson entropy), which has units of species, is easier to interpret in an ecological context than an entropic non-linear index [@jostEntropyDiversity2006; @tuomistoUpdatedConsumerGuide2012].
+We used nonparametric Kruskal-Wallis ranked sum tests to compare diversity between shallow and mesophotic samples and between island groups (MHI vs.
+NWHI).
+Island group comparisons were performed both including and excluding mesophotic samples.
+
+To visualize how potential environmental drivers varied across the archipelago, we used a principal components analysis (PCA), plotting the results with points colored by island group (MHI or NWHI).
+We used distance-based redundancy analysis (dbRDA; a form of multivariate multiple regression) with `dbrda` to examine the relative contribution of environmental variables ([@tbl:tbl_env]) to shallow crab community variation across the islands, conditioning the recovery year as a random effect.
+From a model incorporating all potential drivers, we sequentially removed variables with the largest variance inflation factors (VIF; calculated using `vif.cca`) until all values were <3, indicating variables were sufficiently uncorrelated.
+Model significance was assessed using permutation tests with `r num(PERM,1)` permutations with the function `anova`.
+Continuous variables were scaled to unit variance.
+dbRDA analyses were conducted only for communities sampled from shallow (\< 30 m) ARMS units, since detailed environmental/ecological drivers were not available for deep ($\geq$ 30 m) samples.
+
+We assessed variation among groups using `betadisper` and `adonis2`.
+These functions implement analysis of multivariate homogeneity of group dispersions (PERMDISP) and multivariate permutational analysis of variance (PERMANOVA) respectively.
+PERMDISP is used to assess variability within groups and PERMANOVA is used to assess differences among groups.
+Significance was assessed with permutation tests in both cases.
+We used PERMANOVA to make the following comparisons: island group (MHI vs NWHI, shallow samples only), island (shallow samples only), depth zone only (all samples), depth zone only (O&#x02BB;ahu samples), island group and depth zone (all samples).
+Variation among island groups and sampling depths was further examined and visualized using canonical analysis of principal coordinates (CAP) based on discriminant analysis [@andersonCanonicalAnalysisPrincipal2003] as implemented in the `CAPdiscrim` function from `BiodiversityR` [@kindtTreeDiversityAnalysis2005].
+To explore differences in community variability among groups (shallow/deep, O&#x02BB;ahu/all shallow samples), we compared Euclidean distances among assemblage points in canonical space (from the CAP analysis) using Welch two-sample t-tests, assuming unequal variances.
+
+## Environmental variables {-}
+
+We assembled potential drivers of brachyuran crab community structure from previous studies [@selkoeDNACoralReef2016; @halpernRecentPaceChange2019], NOAA's National Coral Reef Monitoring Program (NCRMP), the Pacific Islands Ocean Observing System (PacIOOS), and MODIS satellite data products compiled by NASA's Physical Oceanography/Ocean Biology Distributed Active Archive Centers [@nasagoddardspaceflightcenterModerateresolutionImagingSpectroradiometer2022; @nasaobpgMODISAquaGlobal2020].
+Variables were either intrinsic (site depth, latitude, longitude, island slope) or derived from remote sensing and/or modeling (chlorophyll-A, sea-surface temperature, coral cover, potential larval immigration, cumulative human impacts) ([@tbl:tbl_env], Supplementary [@tbl:tbl_supp_arms_metadata]).
+Sea-surface temperature (SST) and chlorophyll-A values were calculated by averaging monthly means across the duration of the soaking period for each deployment using AQUA MODIS 4 km-resolution gridded datasets.
+The cumulative human impacts index, a dimensionless numerical value representing spatially-explicit cumulative anthropogenic impacts on various marine ecosystems, is based on a synthesis of a combination of "high resolution, annual data on the intensity of 14 human stressors and their impact on 21 marine ecosystems over 11 years (2003--2013)" [@halpernGlobalMapHuman2008; @halpernRecentPaceChange2019].
+The index of potential larval immigration was imported from [@selkoeDNACoralReef2016] and represents the potential of a given island to act as a destination for larval settlement.
+It is an "estimate of in-closeness centrality" from a biophysical model simulating larval dispersal in the Hawaiian Islands [@selkoeDNACoralReef2016 supp.
+mat.].
+For details regarding data import, preparation, and synthesis, see Supplementary Information.
+Due to logistical and dataset limitations, environmental/ecological drivers were only available for shallow (\< 30 m) samples.
+
+# References {-}
+
+<!-- show main bibliography -->
+<div class="multi-refs"></div>
+
+# Acknowledgements {-}
+We thank the NOAA Coral Reef Conservation Program for supporting the deployment and recovery of the shallow ARMS.
+We thank the UH Mānoa Dive Safety Office (D.
+Pence, K.
+Stender, J.
+Jones, and C.J.
+Bradley) for installation and retrieval of the mesophotic ARMS.
+We specially thank K.
+Flanagan from the UH dive program, whose support was instrumental in the development of the mesophotic ARMS project, and J.
+Copus, who provided expert mesophotic field support; both were taken too soon.
+May their memories be a blessing.
+We thank the members of the ToBo lab, particularly M.
+Iacchei, C.
+Ka&#x02BB;apu-Lyons, and Z.
+Hee, for discussion, support, and their efforts in the disassembly and sorting of ARMS samples on retrieval.
+This is contribution #xxx from the Hawai&#x02BB;i Institute of Marine Biology and #xxx from the School of Ocean and Earth Science and Technology at the University of Hawai&#x02BB;i.
+
+# Author contributions {-}
+Conceptualization: MLH, KKH, DJS, RJT; Data curation: MLH; Formal Analysis: MLH, KKH; Funding acquisition: KKH, DJS, RJT; Investigation: MLH, KKH, KR, DJS; Methodology: MLH, KKH, DJS, MAT; Project administration: KKH, DJS, RJT; Resources: KR, MAT, RJT; Software: MLH; Supervision: DJS, RJT; Validation: MLH, MAT, RJT; Visualization: MLH, KR; Writing -- original draft: MLH, KKH, RJT; Writing -- review & editing: MLH, KKH, KR, DJS, MAT, RJT
+
+# Data availability statement {-}
+All data and code used to generate this manuscript and its accompanying figures and tables can be found on github at [https://github.com/mhoban/arms_crabs](https://github.com/mhoban/arms_crabs).
+
+# Additional Information {-}
+
+## Funding {-}
+This work was funded through a combination of grants including NSF OCE-1260169, NSF GRFP DGE-1329626, and NSF OCE-2048457.
+In addition, the NOAA Coral Reef Conservation Program funded the Pacific Reef Assessment and Monitoring Program, which included the shallow ARMS deployment in the Hawaiian Archipelago.
+<!--
+Specimens used in this study were collected under Hawai&#x02BB;i Division of Aquatic Resources Special Activity Permit numbers SAP 2013-47, SAP 2012-63, SAP 2011-01, and SAP 2008-99.
+-->
+
+## Competing interests {-}
+The authors declare no competing interests.
+
+\listoffigures
+
+# Tables {-}
+```{r tbl-rda,results="asis",dependson="rda-analysis"}
+print_table(rda_table,"tbl_rda","Significant marginal terms in the dbRDA analysis of crab community variation across the Hawaiian Archipelago",hdr_case="none",accuracy=0.001)
 ```
@@
 
-ARMS units were retrieved by encapsulating them in-situ in a 100 μm mesh-lined crate to prevent escape of motile organisms.
-Once returned to the surface, they were placed in containers of aerated, filtered seawater, crates were removed, and units were systematically disassembled plate-by-plate [see @lerayDNABarcodingMetabarcoding2015 for processing details].
-All adult brachyuran crabs ($\geq$ 2 mm) were photographed, identified, preserved in 95% ethanol, and retained in a specimen archive at the Hawai&#x02BB;i Institute of Marine Biology.
-Specimens were initially identified to family or genus using Poore & Ahyong [-@pooreMarineDecapodCrustacea2023] and additionally to species where possible according to expert advice from Scott Godwin (former curator of invertebrate zoology, Bishop Museum) and Gustav Paulay (curator of invertebrate zoology, Florida Museum of Natural History).
-Where species-level identifications were not possible, taxa were collapsed to family or genus (depending on individual identification).
-We recognize that the species-level taxonomy of Indo-Pacific brachyurans remains poorly resolved, particularly for abundant groups like the Chlorodiellinae [@lasleyjrIntegrativeSystematicRevision2014; @lasleyjrPhylogeneticRelationshipsUbiquitous2015], and some "species" are known to comprise complexes [e.g.
-"*Perinia tumida*", @servisCharacterizingCoralReef2020; @plaisanceReefassociatedCrustaceanFauna2009].
-However, the prevalence of species complexes and cryptic diversity suggests that our assignments are likely to under- rather than overrepresent the true number of taxa.
-Thus, any significant patterns we detected likely persist in spite of the potential presence of cryptic diversity.
-Since our primary concern was categorical diversity of assemblages rather than accurate naming, we have retained species names in cases where complexes are suspected.
-
-## Data analysis {-}
-
-Unless otherwise specified, all analyses were performed using `R` version `r R.version$major`.`r R.version$minor` [@rcoreteamLanguageEnvironmentStatistical2021].
-Community matrices were loaded and encapsulated into data objects using the R package `phyloseq` [@mcmurdiePhyloseqPackageReproducible2013] and multivariate community analyses were performed using the package `vegan` [@oksanenVeganCommunityEcology2022] on Bray-Curtis dissimilarities of Hellinger-transformed species abundance matrices [@legendreEcologicallyMeaningfulTransformations2001; @raoReviewCanonicalCoordinates1995].
-
-To examine patterns of crab diversity across the islands, we used `estimate_richness` to calculate species richness and effective Simpson diversity (effective Simpson = $\frac{1}{\sum_{i=1}^{R}p_i^2}$, where R is the total number of species, and $\mathrm{p}_i$ is the proportional abundance of each species).
-The effective Simpson index (the inverse of the "classic" Simpson entropy), which has units of species, is easier to interpret in an ecological context than an entropic non-linear index [@jostEntropyDiversity2006; @tuomistoUpdatedConsumerGuide2012].
-We used nonparametric Kruskal-Wallis ranked sum tests to compare diversity between shallow and mesophotic samples and between island groups (MHI vs.
-NWHI).
-Island group comparisons were performed both including and excluding mesophotic samples.
-
-We used distance-based redundancy analysis (dbRDA; a form of multivariate multiple regression) with `dbrda` to examine the relative contribution of environmental variables ([@tbl:tbl_env]) to shallow crab community variation across the islands, conditioning the recovery year as a random effect.
-From a model incorporating all potential drivers, we sequentially removed variables with the largest variance inflation factors (VIF; calculated using `vif.cca`) until all values were <3, indicating variables were sufficiently uncorrelated.
-Model significance was assessed using permutation tests with `r num(PERM,1)` permutations with the function `anova`.
-Continuous variables were scaled to unit variance.
-dbRDA analyses were conducted only for communities sampled from shallow (\< 30 m) ARMS units, since detailed environmental/ecological drivers were not available for deep ($\geq$ 30 m) samples.
-
-We assessed variation among groups using `betadisper` and `adonis2`.
-These functions implement analysis of multivariate homogeneity of group dispersions (PERMDISP) and multivariate permutational analysis of variance (PERMANOVA) respectively.
-PERMDISP is used to assess variability within groups and PERMANOVA is used to assess differences among groups.
-Significance was assessed with permutation tests in both cases.
-We used PERMANOVA to make the following comparisons: island group (MHI vs NWHI, shallow samples only), island (shallow samples only), depth zone only (all samples), depth zone only (O&#x02BB;ahu samples), island group and depth zone (all samples).
-Variation among island groups and sampling depths was further examined and visualized using canonical analysis of principal coordinates (CAP) based on discriminant analysis [@andersonCanonicalAnalysisPrincipal2003] as implemented in the `CAPdiscrim` function from `BiodiversityR` [@kindtTreeDiversityAnalysis2005].
-To explore differences in community variability among groups (shallow/deep, O&#x02BB;ahu/all shallow samples), we compared Euclidean distances among assemblage points in canonical space (from the CAP analysis) using Welch two-sample t-tests, assuming unequal variances.
-
-## Environmental variables {-}
-
-Potential drivers of brachyuran crab community structure were assembled from previous studies [@selkoeDNACoralReef2016; @halpernRecentPaceChange2019], NOAA's National Coral Reef Monitoring Program (NCRMP), the Pacific Islands Ocean Observing System (PacIOOS), and MODIS satellite data products compiled by NASA's Physical Oceanography/Ocean Biology Distributed Active Archive Centers [@nasagoddardspaceflightcenterModerateresolutionImagingSpectroradiometer2022; @nasaobpgMODISAquaGlobal2020].
-Variables were either intrinsic (site depth, latitude, longitude, island slope) or derived from remote sensing and/or modeling (chlorophyll-A, sea-surface temperature, coral cover, potential larval immigration, cumulative human impacts) ([@tbl:tbl_env], Supplementary [@tbl:tbl_supp_arms_metadata]).
-Sea-surface temperature (SST) and chlorophyll-A values were calculated by averaging monthly means across the duration of the soaking period for each deployment using AQUA MODIS 4 km-resolution gridded datasets.
-The cumulative human impacts index, a dimensionless numerical value representing spatially-explicit cumulative anthropogenic impacts on various marine ecosystems, is based on a synthesis of a combination of "high resolution, annual data on the intensity of 14 human stressors and their impact on 21 marine ecosystems over 11 years (2003--2013)" [@halpernGlobalMapHuman2008; @halpernRecentPaceChange2019].
-The index of potential larval immigration was imported from [@selkoeDNACoralReef2016] and represents the potential of a given island to act as a destination for larval settlement.
-It is an "estimate of in-closeness centrality" from a biophysical model simulating larval dispersal in the Hawaiian Islands [@selkoeDNACoralReef2016 supp.
-mat.].
-For details regarding data import, preparation, and synthesis, see Supplementary Information.
-Due to logistical and dataset limitations, environmental/ecological drivers were only available for shallow (\< 30 m) samples.
-
 ```{r tbl-env,results="asis", dependson="init"}
@@
 ```
-
-<br><br>
-
-# References {-}
-
-<!-- show main bibliography -->
-<div class="multi-refs"></div>
-
-# Acknowledgements {-}
-We thank the NOAA Coral Reef Conservation Program for supporting the deployment and recovery of the shallow ARMS.
-We thank the UH Mānoa Dive Safety Office (D.
-Pence, K.
-Stender, J.
-Jones, and C.J.
-Bradley) for installation and retrieval of the mesophotic ARMS.
-We specially thank K.
-Flanagan from the UH dive program, whose support was instrumental in the development of the mesophotic ARMS project, and J.
-Copus, who provided expert mesophotic field support; both were taken too soon.
-May their memories be a blessing.
-We thank the members of the ToBo lab, particularly M.
-Iacchei, C.
-Ka&#x02BB;apu-Lyons, and Z.
-Hee, for discussion, support, and their efforts in the disassembly and sorting of ARMS samples on retrieval.
-This is contribution #xxx from the Hawai&#x02BB;i Institute of Marine Biology and #xxx from the School of Ocean and Earth Science and Technology at the University of Hawai&#x02BB;i.
-
-# Author contributions {-}
-Conceptualization: MLH, KKH, DJS, RJT; Data curation: MLH; Formal Analysis: MLH, KKH; Funding acquisition: KKH, DJS, RJT; Investigation: MLH, KKH, KR, DJS; Methodology: MLH, KKH, DJS, MAT; Project administration: KKH, DJS, RJT; Resources: KR, MAT, RJT; Software: MLH; Supervision: DJS, RJT; Validation: MLH, MAT, RJT; Visualization: MLH, KR; Writing -- original draft: MLH, KKH, RJT; Writing -- review & editing: MLH, KKH, KR, DJS, MAT, RJT
-
-# Data availability statement {-}
-All data and code used to generate this manuscript and its accompanying figures and tables can be found on github at [https://github.com/mhoban/arms_crabs](https://github.com/mhoban/arms_crabs).
-
-# Additional Information {-}
-
-## Funding {-}
-This work was funded through a combination of grants including NSF OCE-1260169, NSF GRFP DGE-1329626, and NSF OCE-2048457.
-In addition, the NOAA Coral Reef Conservation Program funded the Pacific Reef Assessment and Monitoring Program, which included the shallow ARMS deployment in the Hawaiian Archipelago.
-<!--
-Specimens used in this study were collected under Hawai&#x02BB;i Division of Aquatic Resources Special Activity Permit numbers SAP 2013-47, SAP 2012-63, SAP 2011-01, and SAP 2008-99.
--->
-
-## Competing interests {-}
-The authors declare no competing interests.
-
-\listoffigures
 
@@
 (p1 + p2) / (p3 + p4) &
-  plot_annotation(tag_levels = list(str_glue("({letters[1:4]})"))) &
-  theme(plot.tag = element_text(face="italic"))
+  plot_annotation(tag_levels = list(letters[1:4])) &
+  theme(plot.tag = element_text(face="bold"))
 cat("{#fig:fig_supp_alpha}")
@@
 (p1 + p3) / (p2 + p4) &
-  plot_annotation(tag_levels = list(str_glue("({letters[1:4]})"))) &
-  theme(plot.tag = element_text(face="italic"))
+  plot_annotation(tag_levels = list(letters[1:4])) &
+  theme(plot.tag = element_text(face="bold"))
 cat("{#fig:fig_supp_alpha_main}")
@@
 (p1 + p2) &
-  plot_annotation(tag_levels = list(str_glue("({letters[1:2]})"))) &
-  theme(plot.tag = element_text(face="italic"))
+  plot_annotation(tag_levels = list(letters[1:2])) &
+  theme(plot.tag = element_text(face="bold"))
 cat("{#fig:fig_supp_alpha2}")
@@
 ```{r fig-supp-pca, fig.cap='Principal component analysis (PCA) of environmental variables by sampling site.
-Ellipses represent 95% confidence intervals around island groups.', results='asis', dev='svg', fig.width=8,fig.height=8, dependson='init'}
+Ellipses show 95% confidence intervals around island groups.', results='asis', dev='svg', fig.width=8,fig.height=8, dependson='init'}
 cds <- cc$crab_data_shallow
 
+termz <- model_terms %>%
+  select(variable,display=`Environmental Driver`)
+
 pcdata <- cds %>%
-  select(where(is.numeric),-c(lon,lat),-closest_island) %>%
-  select(deployment_year,depth,chl,sst,slope:human_impact,-contains('radius'))
+  select(where(is.numeric),-closest_island) %>%
+  select(recovery_year,lon,lat,depth,chl,sst,slope:human_impact,-contains('radius'))
 
@@
 bp <- cds %>%
-  bind_cols(scores(pca))
+  bind_cols(scores(pca)) %>%
+	mutate(island_group = str_to_sentence(island_group))
 
@@
     xnudge = if_else(PC1 < 0,-0.02,0.02)
-  )
+  ) %>%
+  left_join(termz,by=c(label="variable")) %>%
+  mutate(
+    label = str_to_sentence( str_replace_all(label,"_"," ") ),
+    label = coalesce(display,label)
+  ) %>%
+  select(-display)
 
@@
   stat_ellipse(geom="polygon",aes(group=island_group,fill=island_group),color=NA,alpha=0.1,show.legend = FALSE) +
-  # geom_mark_ellipse(color="black",alpha=0.2,show.legend=FALSE, linewidth=0.07) +
   scale_fill_manual(values=island_group_pal,name="Island group") +
@@
 
+```{r fig-supp-euclid, fig.cap='Box plot of Euclidean distances between points in canonical space based on CAP analysis of crab assemblages by depth zone.
+Asterisk indicates significant difference (Welch two-sample t-test).', results="asis", dev="svg", fig.width=6, fig.height=4, dependson="cap-analysis"}
+dists <- bind_rows(
+  tibble(dist=shallow_dist,depth="Shallow"),
+  tibble(dist=deep_dist,depth="Deep")
+) %>%
+  mutate(depth = factor(depth,levels=c("Shallow","Deep")))
+
+if (distance_t$p.value < 0.05) {
+  which_max <- dists %>%
+    group_by(depth) %>%
+    summarise(d = mean(dist)) %>%
+    slice(which.max(d)) %>%
+    pull(depth) %>%
+    as.character()
+  max_d <- max(dists$dist) + (diff(range(dists$dist)))/10
+  sig <- tribble(
+    ~x,~y,
+    which_max,max_d
+  )
+} else {
+  sig <- tribble(~x,~y)
+}
+
+pointsize <- 5
+ggplot(dists,aes(x=depth,y=dist,fill=depth)) +
+  geom_boxplot(notch=TRUE) +
+  geom_point(aes(x,y),data=sig,shape=8,size=pointsize,stroke=1.5,fill="black") +
+  scale_fill_manual(values=sample_depth_pal) +
+  guides(fill="none") +
+  plotz_theme() +
+  xlab("Sampling depth") +
+  ylab("Euclidean distance")
+cat("{#fig:fig_supp_euclid}")
+```
+
+\newpage
+
 ```{r cap-analysis-reduced,dependson="init",echo=FALSE,message=FALSE,warning=FALSE,results="hide"}
@@
 ```
-```{r fig-supp-cap-reduced, fig.cap="CAP ordination of crab communities by island group and depth zone showing only samples deployed during the years 2010--2013.", results="asis", dev="svg", fig.width=9, fig.height=6, dependson="cap-analysis-reduced"}
+```{r fig-supp-cap-reduced, fig.cap="CAP ordination of crab communities by island group and depth zone including only samples deployed during the years 2010--2013.", results="asis", dev="svg", fig.width=9, fig.height=6, dependson="cap-analysis-reduced"}
 # make the sample group look nice
@@
   scale_shape_manual(values=c("oahu" = 23, "noahu" = 21),name="Island", labels=c("noahu"="Other islands","oahu"="O‘ahu")) +
-  scale_fill_manual(values=sample_depth_pal,name="Sample group") +
+  scale_fill_manual(values=sample_depth_pal,name="Sample grouping") +
   guides(
@@
 cat("{#fig:fig_supp_cap_reduced}")
-```
-
-\newpage
-
-```{r fig-supp-euclid, fig.cap='Box plot of Euclidean distances between points in CAP ordination by depth zone.
-Asterisk indicates significant difference (Welch two-sample t-test).', results="asis", dev="svg", fig.width=6, fig.height=4, dependson="cap-analysis"}
-dists <- bind_rows(
-  tibble(dist=shallow_dist,depth="Shallow"),
-  tibble(dist=deep_dist,depth="Deep")
-) %>%
-  mutate(depth = factor(depth,levels=c("Shallow","Deep")))
-
-if (distance_t$p.value < 0.05) {
-  which_max <- dists %>%
-    group_by(depth) %>%
-    summarise(d = mean(dist)) %>%
-    slice(which.max(d)) %>%
-    pull(depth) %>%
-    as.character()
-  max_d <- max(dists$dist) + (diff(range(dists$dist)))/10
-  sig <- tribble(
-    ~x,~y,
-    which_max,max_d
-  )
-} else {
-  sig <- tribble(~x,~y)
-}
-
-pointsize <- 5
-ggplot(dists,aes(x=depth,y=dist,fill=depth)) +
-  geom_boxplot(notch=TRUE) +
-  geom_point(aes(x,y),data=sig,shape=8,size=pointsize,stroke=1.5,fill="black") +
-  scale_fill_manual(values=sample_depth_pal) +
-  guides(fill="none") +
-  plotz_theme() +
-  xlab("Sampling depth") +
-  ylab("Euclidean distance between CAP points")
-cat("{#fig:fig_supp_euclid}")
-```
-
-\newpage
-
-```{r tbl-supp-arms-metadata, results="asis", dependson="init"}
-region_map <- c(
-  'main' = 'MHI',
-  'northwest' = 'NWHI',
-  'mce' = 'MHI (mesophotic)'
-)
-
-arms_metadata <- cc$crabs_untransformed %>%
-  sample_tibble() %>%
-  # separate_wider_regex(sample,patterns=c(recovery_year="^[0-9]+","_",sample=".*"))  %>%
-  # select(sample,everything()) %>%
-  mutate(
-    island = case_when(
-      str_remove_all(hawaiian_name,'[^a-zA-Z ]') != island ~ as.character(str_glue("{hawaiian_name} ({island})")),
-      TRUE ~ as.character(hawaiian_name)
-    ),
-    island = case_when(
-      island == "O‘ahu" ~ str_glue("{hawaiian_name} ({shallow_deep})"),
-      TRUE ~ island
-    ),
-    island = fct_reorder(island,-lat),
-    region = region_map[region],
-    lat = num(lat,0.0001),
-    lon = num(lon,0.0001)
-  ) %>%
-  select(sample,region,island,depth,deployment_year,recovery_year,lat,lon) %>%
-  janitor::clean_names(case="sentence")
-print_table(arms_metadata,"tbl_supp_arms_metadata","ARMS unit metadata",accuracy=1,hdr_case=NA,format_numbers=FALSE)
 ```
@@
 
-```{r tbl-supp-rda,results="asis",dependson="rda-analysis"}
-print_table(rda_table,"tbl_supp_rda","Significant marginal terms in the dbRDA analysis of crab community variation across the Hawaiian Archipelago",hdr_case="none",accuracy=0.001)
+```{r tbl-supp-arms-metadata, results="asis", dependson="init"}
+region_map <- c(
+  'main' = 'MHI',
+  'northwest' = 'NWHI',
+  'mce' = 'MHI (mesophotic)'
+)
+
+arms_metadata <- cc$crabs_untransformed %>%
+  sample_tibble() %>%
+  # separate_wider_regex(sample,patterns=c(recovery_year="^[0-9]+","_",sample=".*"))  %>%
+  # select(sample,everything()) %>%
+  mutate(
+    island = case_when(
+      str_remove_all(hawaiian_name,'[^a-zA-Z ]') != island ~ as.character(str_glue("{hawaiian_name} ({island})")),
+      TRUE ~ as.character(hawaiian_name)
+    ),
+    island = case_when(
+      island == "O‘ahu" ~ str_glue("{hawaiian_name} ({shallow_deep})"),
+      TRUE ~ island
+    ),
+    island = fct_reorder(island,-lat),
+    region = region_map[region],
+    lat = num(lat,0.0001),
+    lon = num(lon,0.0001)
+  ) %>%
+  select(sample,region,island,depth,deployment_year,recovery_year,lat,lon) %>%
+  janitor::clean_names(case="sentence")
+print_table(arms_metadata,"tbl_supp_arms_metadata","ARMS unit metadata",accuracy=1,hdr_case=NA,format_numbers=FALSE)
 ```

### 2025-03-22 manuscript.Rmd
show suspected complexes as cf. spp.

@@
 ![Representative photographs of the four most abundant crab species found across all deployed ARMS units (specimens shown were not collected as part of this study).
-(a) "*Chlorodiella laevissima*" (FLMNH 45490, Kāne&#x02BB;ohe, Hawai&#x02BB;i).
-(b) "*Perinia tumida*" (FLMNH 45652, Kāne&#x02BB;ohe, Hawai&#x02BB;i).
+(a) *Chlorodiella* cf.
+*laevissima* (FLMNH 45490, Kāne&#x02BB;ohe, Hawai&#x02BB;i).
+(b) *Perinia* cf.
+*tumida* (FLMNH 45652, Kāne&#x02BB;ohe, Hawai&#x02BB;i).
 (c) *Percnon abbreviatum* (FLMNH 45628, Kāne&#x02BB;ohe, Hawai&#x02BB;i).
@@
 We recognize that the species-level taxonomy of Indo-Pacific brachyurans remains poorly resolved, particularly for abundant groups like the Chlorodiellinae [@lasleyjrIntegrativeSystematicRevision2014; @lasleyjrPhylogeneticRelationshipsUbiquitous2015], and some "species" are known to comprise complexes [e.g.
-"*Perinia tumida*", @servisCharacterizingCoralReef2020; @plaisanceReefassociatedCrustaceanFauna2009].
+*Perinia* cf.
+*tumida*, @servisCharacterizingCoralReef2020; @plaisanceReefassociatedCrustaceanFauna2009].
 However, the prevalence of species complexes and cryptic diversity suggests that our assignments are likely to under- rather than overrepresent the true number of taxa.
@@
   mutate(display_species = as.character(case_when(
-    as.logical(complex) ~ str_glue('"{display_species}"'),
+    as.logical(complex) ~ cf_species(display_species),
     .default = display_species

### 2025-03-22 crabs.R

@@ -363,6 +363,15 @@ setup_crabs <- function() {
   return(cc)
 }
 
+# make a species be cf.
+cf_species <- function(spp,deitalicize=TRUE) {
+  if (deitalicize) {
+    str_replace(spp," ","* cf. *")
+  } else {
+    str_replace(spp," "," cf. ")
+  }
+}
+
 sample_summary <- function(ps,top_n=5,taxonomy=NULL) {
   ss <- list()
 
@@ -370,7 +379,7 @@ sample_summary <- function(ps,top_n=5,taxonomy=NULL) {
   otu_lookup <- ps %>%
     taxa_tibble(otu_col = "otu") %>%
     mutate(species = as.character(case_when(
-      as.logical(complex) ~ str_glue('"{species}"'),
+      as.logical(complex) ~ cf_species(species),
       .default = species
     ))) %>%
     select(otu,species) %>%
@@ -428,7 +437,7 @@ sample_summary <- function(ps,top_n=5,taxonomy=NULL) {
     slice(match(ss$top_otus,otu)) %>%
     mutate(
       disp = case_when(
-        as.logical(complex) & taxon_level == "species" ~ str_glue("\"{display_species}\" {authority}"),
+        as.logical(complex) & taxon_level == "species" ~ str_glue("{cf_species(display_species)} {authority}"),
         taxon_level == "species" ~ str_glue("{display_species} {authority}"),
         .default = display_species
       )

### 2025-03-22 manuscript.Rmd
little figure caption tweak

@@
 
-![Representative photographs of the four most abundant crab species found across all deployed ARMS units (specimens shown were not collected as part of this study).
+![Representative photographs of the four most abundant crab species found across all deployed ARMS units (pictured specimens were not collected as part of this study).
 (a) *Chlorodiella* cf.

### 2025-03-22 manuscript.Rmd
fix table 3 column widths and some citation details, switch back to custom pandoc-crossref

@@
 
-  table_str <- pander::pandoc.table.return(
+  table_str <- pandoc.table.return(
     tbl,
@@
     caption=NULL,
-    style="grid",
+    style="multiline",
     split.cells=cell_width,
@@
 
-We assembled potential drivers of brachyuran crab community structure from previous studies [@selkoeDNACoralReef2016; @halpernRecentPaceChange2019], NOAA's National Coral Reef Monitoring Program (NCRMP), the Pacific Islands Ocean Observing System (PacIOOS), and MODIS satellite data products compiled by NASA's Physical Oceanography/Ocean Biology Distributed Active Archive Centers [@nasagoddardspaceflightcenterModerateresolutionImagingSpectroradiometer2022; @nasaobpgMODISAquaGlobal2020].
+We assembled potential drivers of brachyuran crab community structure from previous studies [@selkoeDNACoralReef2016; @halpernRecentPaceChange2019], NOAA's National Coral Reef Monitoring Program (NCRMP), the Pacific Islands Ocean Observing System (PacIOOS), and MODIS satellite data products compiled by NASA's Physical Oceanography/Ocean Biology Distributed Active Archive Centers [@nasaCHL; @nasaSST].
 Variables were either intrinsic (site depth, latitude, longitude, island slope) or derived from remote sensing and/or modeling (chlorophyll-A, sea-surface temperature, coral cover, potential larval immigration, cumulative human impacts) ([@tbl:tbl_env], Supplementary [@tbl:tbl_supp_arms_metadata]).
@@
 
-<br><br>
+\newpage
 
@@
 
-<br><br>
+\newpage
 
 ```{r tbl-env,results="asis", dependson="init"}
-print_table(model_terms,"tbl_env","Potential environmental drivers of crab community structure across the Hawaiian Archipelago",hide="variable")
+mt <- model_terms %>%
+	mutate(
+		Description = str_wrap(Description,60),
+		# `Environmental Driver` = str_pad(`Environmental Driver`,30,pad="-",side="right"),
+		# `Data resolution` = str_pad(`Data resolution`,20,pad="-",side="right"),
+		Source = str_wrap(Source,30)
+	)
+print_table(mt,"tbl_env","Potential environmental drivers of crab community structure across the Hawaiian Archipelago",hide="variable")
 ```
@@
 ### Processing of satellite-derived environmental data
-Level-3 mapped (4 km grid-size) sea-surface temperature (SST) and chlorophyll-A (chl-A) data were retrieved from NASA's Physical Oceanography Distributed Active Archive Center (PO.DAAC [@nasaobpgMODISAquaGlobal2020]) and Ocean Biology Distributed Active Archive Center (OB.DAAC [@nasagoddardspaceflightcenterModerateresolutionImagingSpectroradiometer2022]) respectively.
+Level-3 mapped (4 km grid-size) sea-surface temperature (SST) and chlorophyll-A (chl-A) data were retrieved from NASA's Physical Oceanography Distributed Active Archive Center (PO.DAAC [@nasaSST]) and Ocean Biology Distributed Active Archive Center (OB.DAAC [@nasaCHL]) respectively.
 Monthly summary datasets were downloaded for the period of 2009--2016 and averaged over ARMS unit soaking periods (e.g.

### 2025-03-24 manuscript.Rmd
Rearrange tables, add taxa at all depth, hide sup. figs from list.

@@
   # Figure legends {-}
+lofItemTitle: |
+  Fig.
+lofItemTemplate: |
+  $$lofItemTitle$$ $$i$$$$listItemTitleDelim$$ $$t$$
 header-includes:
@@
   unique()
+
+ndepths <- cc$crab_data %>%
+  pull(depth_zone) %>%
+  n_distinct()
+
+everywhere <- cc$crabs_untransformed %>%
+  psmelt() %>%
+  filter(Abundance > 0) %>%
+  group_by(OTU,species) %>%
+  summarise(dp = n_distinct(depth_zone)) %>%
+  filter(dp == ndepths) %>%
+  arrange(species) %>%
+  left_join(crab_taxonomy_validated,by=c("OTU" = "otu")) %>%
+  mutate(display_species = as.character(case_when(
+    as.logical(complex) ~ cf_species(display_species),
+    .default = display_species
+  ))) %>%
+  pull(display_species)
 ```
@@
 `r str_to_sentence(english(spp['shallow']))` distinct taxa (including those collapsed to genus or family) were exclusive to shallow ARMS units and `r spp['deep']` were found only deep (`r format_list(deep_spp)`; Supplementary [@tbl:tbl_supp_species_depth]).
+Out of the `r ntaxa(cc$crabs)` total recovered taxa, only `r length(everywhere)` were found across all `r ndepths` depth zones (`r format_list(everywhere)`).
 For a more in-depth taxonomic treatment of the assemblage composition of mesophotic vs.
@@
 
-# References {-}
-
-<!-- show main bibliography -->
-<div class="multi-refs"></div>
-
-# Acknowledgements {-}
-We thank the NOAA Coral Reef Conservation Program for supporting the deployment and recovery of the shallow ARMS.
-We thank the UH Mānoa Dive Safety Office (D.
-Pence, K.
-Stender, J.
-Jones, and C.J.
-Bradley) for installation and retrieval of the mesophotic ARMS.
-We specially thank K.
-Flanagan from the UH dive program, whose support was instrumental in the development of the mesophotic ARMS project, and J.
-Copus, who provided expert mesophotic field support; both were taken too soon.
-May their memories be a blessing.
-We thank the members of the ToBo lab, particularly M.
-Iacchei, C.
-Ka&#x02BB;apu-Lyons, and Z.
-Hee, for discussion, support, and their efforts in the disassembly and sorting of ARMS samples on retrieval.
-This is contribution #xxx from the Hawai&#x02BB;i Institute of Marine Biology and #xxx from the School of Ocean and Earth Science and Technology at the University of Hawai&#x02BB;i.
-
-# Author contributions {-}
-Conceptualization: MLH, KKH, DJS, RJT; Data curation: MLH; Formal Analysis: MLH, KKH; Funding acquisition: KKH, DJS, RJT; Investigation: MLH, KKH, KR, DJS; Methodology: MLH, KKH, DJS, MAT; Project administration: KKH, DJS, RJT; Resources: KR, MAT, RJT; Software: MLH; Supervision: DJS, RJT; Validation: MLH, MAT, RJT; Visualization: MLH, KR; Writing -- original draft: MLH, KKH, RJT; Writing -- review & editing: MLH, KKH, KR, DJS, MAT, RJT
-
-# Data availability statement {-}
-All data and code used to generate this manuscript and its accompanying figures and tables can be found on github at [https://github.com/mhoban/arms_crabs](https://github.com/mhoban/arms_crabs).
-
-# Additional Information {-}
-
-## Funding {-}
-This work was funded through a combination of grants including NSF OCE-1260169, NSF GRFP DGE-1329626, and NSF OCE-2048457.
-In addition, the NOAA Coral Reef Conservation Program funded the Pacific Reef Assessment and Monitoring Program, which included the shallow ARMS deployment in the Hawaiian Archipelago.
-<!--
-Specimens used in this study were collected under Hawai&#x02BB;i Division of Aquatic Resources Special Activity Permit numbers SAP 2013-47, SAP 2012-63, SAP 2011-01, and SAP 2008-99.
--->
-
-## Competing interests {-}
-The authors declare no competing interests.
-
-\listoffigures
+\newpage
 
@@
 
+# References {-}
+
+<!-- show main bibliography -->
+<div class="multi-refs"></div>
+
+# Acknowledgements {-}
+We thank the NOAA Coral Reef Conservation Program for supporting the deployment and recovery of the shallow ARMS.
+We thank the UH Mānoa Dive Safety Office (D.
+Pence, K.
+Stender, J.
+Jones, and C.J.
+Bradley) for installation and retrieval of the mesophotic ARMS.
+We specially thank K.
+Flanagan from the UH dive program, whose support was instrumental in the development of the mesophotic ARMS project, and J.
+Copus, who provided expert mesophotic field support; both were taken too soon.
+May their memories be a blessing.
+We thank the members of the ToBo lab, particularly M.
+Iacchei, C.
+Ka&#x02BB;apu-Lyons, and Z.
+Hee, for discussion, support, and their efforts in the disassembly and sorting of ARMS samples on retrieval.
+This is contribution #xxx from the Hawai&#x02BB;i Institute of Marine Biology and #xxx from the School of Ocean and Earth Science and Technology at the University of Hawai&#x02BB;i.
+
+# Author contributions {-}
+Conceptualization: MLH, KKH, DJS, RJT; Data curation: MLH; Formal Analysis: MLH, KKH; Funding acquisition: KKH, DJS, RJT; Investigation: MLH, KKH, KR, DJS; Methodology: MLH, KKH, DJS, MAT; Project administration: KKH, DJS, RJT; Resources: KR, MAT, RJT; Software: MLH; Supervision: DJS, RJT; Validation: MLH, MAT, RJT; Visualization: MLH, KR; Writing -- original draft: MLH, KKH, RJT; Writing -- review & editing: MLH, KKH, KR, DJS, MAT, RJT
+
+# Data availability statement {-}
+All data and code used to generate this manuscript and its accompanying figures and tables can be found on github at [https://github.com/mhoban/arms_crabs](https://github.com/mhoban/arms_crabs).
+
+# Additional Information {-}
+
+## Funding {-}
+This work was funded through a combination of grants including NSF OCE-1260169, NSF GRFP DGE-1329626, and NSF OCE-2048457.
+In addition, the NOAA Coral Reef Conservation Program funded the Pacific Reef Assessment and Monitoring Program, which included the shallow ARMS deployment in the Hawaiian Archipelago.
+<!--
+Specimens used in this study were collected under Hawai&#x02BB;i Division of Aquatic Resources Special Activity Permit numbers SAP 2013-47, SAP 2012-63, SAP 2011-01, and SAP 2008-99.
+-->
+
+## Competing interests {-}
+The authors declare no competing interests.
+
+\listoffigures
+
 \newpage
@@
   theme(plot.tag = element_text(face="bold"))
-cat("{#fig:fig_supp_alpha}")
+cat("{#fig:fig_supp_alpha hidden=\"y\"}")
 ```
@@
   theme(plot.tag = element_text(face="bold"))
-cat("{#fig:fig_supp_alpha_main}")
+cat("{#fig:fig_supp_alpha_main hidden=\"y\"}")
 ```
@@
   theme(plot.tag = element_text(face="bold"))
-cat("{#fig:fig_supp_alpha2}")
+cat("{#fig:fig_supp_alpha2 hidden=\"y\"}")
 ```
@@
   plotz_theme()
-cat("{#fig:fig_supp_pca}")
+cat("{#fig:fig_supp_pca hidden=\"y\"}")
 ```
@@
   scale_y_discrete(limits=rev)
-cat("{#fig:fig_supp_pairwise}")
+cat("{#fig:fig_supp_pairwise hidden=\"y\"}")
 ```
@@
   ylab("Euclidean distance")
-cat("{#fig:fig_supp_euclid}")
+cat("{#fig:fig_supp_euclid hidden=\"y\"}")
 ```
@@
   plotz_theme()
-cat("{#fig:fig_supp_cap_reduced}")
+cat("{#fig:fig_supp_cap_reduced hidden=\"y\"}")
 ```

### 2025-03-24 manuscript.Rmd
last few tweaks before submission to scientific reports

@@
 ### Environmental drivers {-}
-The PCA of environmental variables indicated strong clustering according to island group (Supplementary [@fig:fig_supp_pca]), with apparent orthogonal variation in variables associated with the MHI vs the NWHI.
+The PCA of environmental variables indicated strong clustering according to island group (Supplementary [@fig:fig_supp_pca]), with apparently orthogonal variation in variables associated with the MHI vs the NWHI.
 The best dbRDA model (`r permanova(db_modl,stat="F")`) explained `r pc(RsquareAdj(db_modl)$adj.r.squared,0.01)` (adjusted $\mathrm{R}^2$) of the variability in crab communities across the archipelago ([@fig:fig_dbrda]).
@@
 Although our shallow-water ARMS units were deployed within a narrow depth range (`r dr(cc$crabs_shallow_untransformed)` m), depth still emerged as a significant and consistent driver of community variation among these units.
-Similar fine-scale depth patterns have been found in previous studies, both using molecular [e.g., @desouzaImportanceDepthTemperature2023; @westEDNAMetabarcodingSurvey2020] and traditional survey methods [e.g., @williamsBenthicCommunitiesTwo2013; @srinivasanDepthDistributionsCoral2003], but our results recover them alongside numerous other geographic and environmental gradients.
+Similar fine-scale depth patterns have been found in previous studies, both using molecular [e.g., @desouzaImportanceDepthTemperature2023; @westEDNAMetabarcodingSurvey2020] and traditional survey methods [e.g., @williamsBenthicCommunitiesTwo2013; @srinivasanDepthDistributionsCoral2003], but our results recover them alongside numerous other geographic and environmental gradients and across a broad geographic scale.
 Depth in and of itself is unlikely to affect marine community composition and oceanographic factors with well-known depth profiles (e.g., temperature, oxygen, nutrients, etc.) are not expected to vary dramatically across such a narrow depth range (\< 10 m), although internal waves can deliver variable temperatures across shallow reefs [@storlazziInternalTidesCan2020; @wyattHeatAccumulationCoral2020] and fine-scale habitat variability among sites may also influence the community differences we found [@steyaertRemoteReefCryptobenthic2022].
@@
 The evident groupings in the CAP ordination match these trends, with 30 m samples lying closer to shallow O&#x02BB;ahu samples and 60--90 m samples forming a distinct cluster.
-This pattern of community transition from shallow to mesophotic reefs is consistent with what has been shown for various other taxonomic groups [@lesserGlobalCommunityBreaks2019; @rochaMesophoticCoralEcosystems2018], lending support to the hypothesis that there may be broad-scale depth-based assembly rules for coral reef communities.
+This pattern of community transition from shallow to mesophotic reefs is consistent with what has been shown for various other taxonomic groups [@lesserGlobalCommunityBreaks2019; @rochaMesophoticCoralEcosystems2018] and subsets of entire communities [@hobanPlumbingDepthsEnvironmental2023], lending support to the hypothesis that there may be broad-scale depth-based assembly rules for coral reef communities.
 
@@
 In contrast, the lower depth boundaries of mesophotic ecosystems are largely defined by surface irradiance [@kahngCommunityEcologyMesophotic2010; @hindersteinThemeSectionMesophotic2010], so light seems a possible candidate as bottom-up abiotic factor influencing community structure across depths.
-Crab assemblage structure is unlikely to be directly affected by light variability (although the behavior of individual crabs may).
+Crab assemblage structure overall is unlikely to be directly affected by light variability (diurnal behavior patterns of individual species notwithstanding).
 However, habitat-building species such as corals, algae, and sponges are frequently subject to light requirements [@kahngCommunityEcologyMesophotic2010; @slatteryFunctionStabilityMesophotic2024a], and may in turn affect the structure of the communities dependent upon them [@hustonCompetitionCoexistenceEffects1994; @smithRoleHabitatComplexity2014; @sebensHabitatStructureCommunity1991].
@@
 Two potential weaknesses of our dataset are the time periods over which ARMS units were deployed and the fact that mesophotic samples were obtained solely from the island of O&#x02BB;ahu.
-ARMS units were deployed across a broad range of years, but the majority were deployed during 2010--2012 or 2010--2013 (an overlap of two years) ([@tbl:tbl_arms]).
-A reduced version of the CAP including only ARMS units deployed during these overlapping time periods retains the same (if not more pronounced) overall patterns as the full version (Supplementary [@fig:fig_supp_cap_reduced]), suggesting that the inclusion of previous or subsequent soaking periods did not adversely skew our analyses.
+Our ARMS units were deployed across a broad range of years, but the majority were deployed during 2010--2012 or 2010--2013 (an overlap of two years) ([@tbl:tbl_arms]).
+A reduced version of the CAP ordination filtered to include only ARMS units deployed during these overlapping time periods retains the same (if not more pronounced) overall patterns as the full version (Supplementary [@fig:fig_supp_cap_reduced]), suggesting that the inclusion of prior or subsequent soaking periods did not adversely skew our analyses.
 The limitation of our depth transect to O&#x02BB;ahu means that we cannot make broad-scale conclusions about biodiversity patterns on deep reefs across the Hawaiian Islands.
-However, it remains the case that community and alpha diversity differences among depth zones at that single island appear to overwhelm differences observed among shallow-water coral reef habitat across the archipelago.
+However, it remains the case that community and alpha diversity differences among depth zones at that single island appear to overwhelm differences observed among shallow-water coral reef habitat across the archipelago and depth appears to be the strongest organizing factor in canonical space regardless of where or when the communities were sampled.
 Similar patterns were found at a local scale (tens of kilometers) on Hawai&#x02BB;i Island using environmental DNA (eDNA) [@hobanPlumbingDepthsEnvironmental2023].
@@
 Such differences should be tested in future studies to understand the specific physical and ecological drivers of coral reef depth zonation.
-The cost, logistical complexity, and time investment required for ARMS deployments make this challenging, but emerging approaches such as environmental DNA (eDNA) may offer a suitable alternative.
+The cost, logistical complexity, and time investment required for ARMS deployments (particularly at mesophotic depths) make this challenging, but emerging approaches such as environmental DNA (eDNA) may offer a suitable alternative.
 

### 2025-03-24 manuscript.Rmd

@@
 library(ggtext)
+source("lib/CAPdiscrim.R")
 

### 2025-05-18 manuscript.Rmd
various reviewer suggestions

@@
 We examine the effects of environmental, ecological, and anthropogenic factors (latitude, longitude, depth, years, chlorophyll-A, temperature, slope, coral cover, potential larval immigration, human impacts) on shallow (8--17 m) crab assemblages and compare them with those sampled along a mesophotic depth gradient (12--90 m) on O&#x02BB;ahu.
-Shallow assemblages were significantly different between the inhabited Main and relatively pristine Northwestern Hawaiian Islands and associated with human impacts, chlorophyll-A, sea-surface temperature, slope, potential larval immigration, and depth.
+Shallow assemblages were significantly different between the inhabited Main and relatively pristine Northwestern Hawaiian Islands and associated with sea-surface temperature, chlorophyll-A, depth, slope, potential larval immigration, and human impacts.
 Despite this variability, differences along the depth gradient on O&#x02BB;ahu alone were greater than among the shallow assemblages across the entire archipelago.
@@
 While the structure and composition of coral reef communities is influenced by a suite of environmental and anthropogenic factors, the nature and makeup of those factors remain poorly resolved [@hughesCommunityStructureDiversity1989; @smithEffectsTopBottom2010; @brandlCoralReefEcosystem2019].
-One aspect of coral reef community structure that is thought to be important but remains under-investigated is depth.
+One aspect of coral reef community structure thought to be important but remains under-investigated is depth.
 Most studies to date have focused on the upper ~30 m of the water column, yet coral reefs and associated communities are known to extend beyond 150 m [@pyleExploringDeepCoral1996; @pyleMesophoticCoralEcosystems2019], with photosymbiotic corals reported to 170 m [@rouzeSymbioticAssociationsDeepest2021].
@@
 To date, the majority of studies that have examined community structure across the Hawaiian archipelago and/or along a depth gradient have focused on conspicuous taxa such as corals and fishes [e.g., @benfieldComparisonCoralReef2008; @caballero-aragonWaveExposureTemperature2023; @kaneTrophicDesignationLive2017; @langeWaveExposureShapes2021; @vanlierHabitatConnectivityComplexity2018].
-However, these taxa make up a minority of coral reef biodiversity and thus may not be indicative of community patterns overall [@fautinOverviewMarineBiodiversity2010].
+One previous study [@hurleyAssessmentShallowMesophotic2016] found a significant depth gradient on the island of O&#x02BB;ahu for brachyuran crabs.
+However, the most frequently-targeted taxa make up a minority of coral reef biodiversity and thus may not be indicative of community patterns overall [@fautinOverviewMarineBiodiversity2010].
 Furthermore, studies on these taxa have also been criticized for confounding multiple potentially covarying factors (such as habitat area and/or complexity) across the measured gradients [e.g., @bellwoodRegionalscaleAssemblyRules2001; @grahamEffectsMarineReserve2003; @connollyCommunityStructureCorals2005; @ewersConfoundingFactorsDetection2006; @fletcherRoleHabitatArea2007].
@@
 Such standardization can reduce variability in recovered communities that may arise due to local variation in habitat composition and availability [@ransomeImportanceStandardizationBiodiversity2017].
-We targeted brachyuran crabs, a ubiquitous group often used as an exemplar to assess coral reef biodiversity [@knowltonCoralReefBiodiversity2010; @plaisanceDiversityCoralReefs2011; @servisCharacterizingCoralReef2020; @plaisanceReefassociatedCrustaceanFauna2009] and comprising a notable proportion of the Hawaiian coral reef cryptofauna (284 reported species [@castroCatalogAnomuranBrachyuran2011]).
+This study builds on on previous work targeting brachyuran crabs [@hurleyAssessmentShallowMesophotic2016], a ubiquitous group often used as an exemplar to assess coral reef biodiversity [@knowltonCoralReefBiodiversity2010; @plaisanceDiversityCoralReefs2011; @servisCharacterizingCoralReef2020; @plaisanceReefassociatedCrustaceanFauna2009] and comprising a notable proportion of the Hawaiian coral reef cryptofauna (284 reported species [@castroCatalogAnomuranBrachyuran2011]).
 We used assemblages recovered from ARMS units to examine drivers of community composition across the Hawaiian Archipelago.
@@
 mesophotic), and examined the effects of a suite of possible environmental and anthropogenic drivers of community structure on shallow reefs (including fine-scale depth differences), to test hypotheses about their relative importance.
-
-<!--
-Much of our understanding of biotic zonation and community structure in the ocean originates with classic studies from the intertidal [e.g., @connellInfluenceInterspecificCompetition1961; @connellEffectsCompetitionPredation1961; @paineIntertidalCommunityStructure1974].
-While there are no shortages of exceptions, a long history of studies point towards a generalization that lower bathymetric limits are regulated by ecological interactions whereas upper limits of intertidal communities are more often determined by abiotic factors [@stephenson1972life; @mengeRockyIntertidalCommunities2001; @bertnessTestingRelativeContribution1999; @dennyPhysicalProcessesThat2001].
-However, most of these models describe vertical zonation at specific localities rather than community variation over larger spatial scales [but see @birdTideWaveMarks2013; @bishop-taylorTidesModellingElevation2019] and may not apply to coral reefs, which are subject to more complex community dynamics [@abramsRecruitmentLotteriesCoexistence1984; @bussCompetitiveNetworksNontransitive1979; @hixonPredationPreyRefuges1993; @hubbellUnifiedNeutralTheory2011; @moraEcologyFishesCoral2015; @saleStockrecruitRelationshipsRegional1982; @sheppardBiologyCoralReefs2017].
-While there is little doubt that ecological interactions are important on coral reefs, the suite of factors and which play the most important role in structuring coral reef communities remains a subject of considerable debate [@bellwoodRegionalscaleAssemblyRules2001; @connollyUnifiedModelExplains2017; @hubbellUnifiedNeutralTheory2011; @jacksonAdaptationDiversityReef1991; @mumbyCoralReefHabitats2008; @saleMaintenanceHighDiversity1977].
--->
 
@@
 ![Representative photographs of the four most abundant crab species found across all deployed ARMS units (pictured specimens were not collected as part of this study).
+Individual scale information was not available, but all specimens were <3.5 cm in carapace width.
 (a) *Chlorodiella* cf.
@@
 db_modl <- dbrda(
-  cc$crab_otus_shallow ~ depth +
+  otu_table(cc$crabs_shallow) ~ depth +
     chl + sst + slope + coral_cover +
@@
     lat + lon +
-    Condition(recovery_year),
-  data=cc$crab_data_shallow,
-  # data=dd,
+    Condition(soak_time),
+  data=sample_tibble(cc$crabs_shallow),
   distance="bray"
@@
 # reduce model to terms with VIF <= 3
-db_modl <- reduce_model(db_modl,cc$crab_otus_shallow,cc$crab_data_shallow,vif_limit = 3)
+db_modl <- reduce_model(db_modl,otu_table(cc$crabs_shallow),sample_tibble(cc$crabs_shallow),vif_limit = 3)
 
@@
 
-# get marginal term information and sort by descending f-statistic
-modl_perm <- anova(db_modl,by="margin",permutations = PERM) %>%
-  # as_tibble(rownames="term") %>%
-  # janitor::clean_names() %>%
+# get term information and sort by descending f-statistic
+modl_perm <- anova(db_modl,by="term",permutations = PERM) %>%
   broom::tidy() %>%
@@
 ```{r fig-dbrda, fig.cap="dbRDA plot showing significant drivers of crab community variation across the Hawaiian Islands.
-Colored ellipses represent 95% confidence intervals around island groups.", results="asis", fig.width=9, fig.height=6, dependson="rda-analysis"}
+Colored ellipses represent 95% confidence intervals around island groups.", results="asis", fig.width=9, fig.height=6, dependson="dbrda-analysis"}
 termz <- model_terms %>%
@@
 
+plotz_info <- ggplot_build(rda_plotz)
+xr <- plotz_info$layout$panel_scales_x[[1]]$range$range
+yr <- plotz_info$layout$panel_scales_y[[1]]$range$range
+
 # calculate variation explained (constrained & unconstrained)
@@
 
+rsq <- round(RsquareAdj(db_modl)$adj.r.squared,2)
+rsq <- str_glue("Adjusted~model~R^2:~{rsq}")
+
 # plot the plotz
 rda_plotz +
+  annotate(geom = "text", x = xr[1], y = yr[1], label = rsq, parse=TRUE,hjust=0) +
   xlab(axis_var[1]) + ylab(axis_var[2])
@@
 There was no significant difference in community dispersion between the MHI and NWHI (`r permdisp(bd_region)`).
-When mesophotic ($\geq$ 30 m) ARMS units were included, there were significant differences among depth zones (\< 30 m, 30 m, 60 m, 90 m) both for all samples (`r permanova(anova_depth,0.01)`) and samples restricted to O&#x02BB;ahu (`r permanova(anova_depth_oahu,0.01)`), and community dispersion (`r permdisp(bd_depth)`).
+When mesophotic ($\geq$ 30 m) ARMS units were included, there were significant differences among depth zones (\< 30 m, 30 m, 60 m, 90 m) both for all islands (`r permanova(anova_depth,0.01)`) and samples restricted to O&#x02BB;ahu (`r permanova(anova_depth_oahu,0.01)`), and community dispersion (`r permdisp(bd_depth)`).
 However, community dispersion was not different among depth zones $\geq$ 30 m (`r permdisp(bd_deep)`).
@@
 Shallow samples from O&#x02BB;ahu primarily clustered with shallow samples from other islands ([@fig:fig_cap]).
+A CAP ordination including only samples from O&#x02BB;ahu shows a similar pattern of depth zonation ([@fig:fig_supp_cap_oahu]).
 Variability among 30--90 m samples was apparently greater than among all shallow samples, with mean Euclidean distances between deep samples in canonical space significantly greater than between shallow samples (`r t_test(distance_t)`; [@fig:fig_cap], Supplementary [@fig:fig_supp_euclid]).
@@
 Human activities are considerably higher in the populated MHI than in the protected and remote NWHI [@selkoeEvaluatingAnthropogenicThreats2008; @selkoeMapHumanImpacts2009; @selkoeDNACoralReef2016].
-Although we did not consider the effects of individual impacts here, an index representing *cumulative* anthropogenic impacts was among the top drivers of shallow brachyuran community structure, similar to findings in previous work examining genetic diversity as it relates to species richness and habitat area, quality, and stability across the archipelago [@selkoeDNACoralReef2016].
-The next-most significant driver, chlorophyll-A, can be used as a proxy for primary productivity.
+Although we did not consider the effects of individual impacts here, an index representing *cumulative* anthropogenic impacts was among the significant drivers of shallow brachyuran community structure, similar to findings in previous work examining genetic diversity as it relates to species richness and habitat area, quality, and stability across the archipelago [@selkoeDNACoralReef2016].
+The second most significant driver, chlorophyll-A, can be used as a proxy for primary productivity.
 Productivity gradients across the Hawaiian Archipelago coincide with differences in coral reef community composition [@selkoeDNACoralReef2016], endemic reef fish dominance [@friedlanderDominanceEndemicsReef2020], survival rates of top predators [@bakerEffectVariableOceanic2007], and island geomorphology [@goveNearislandBiologicalHotspots2016].
@@
 Given the prevalence in our model of significant factors known to be important influencers of marine communities in Hawai&#x02BB;i, it is not surprising that we find the same top drivers of community structure for brachyuran crabs, one of the primary taxonomic groups inhabiting coral reefs.
-What is surprising---given the maximum vertical separation among shallow ARMS units of `r diff(range(cc$crab_data_shallow_unscaled$depth))` m---is the consistent significance of depth as a driver of community composition across these gradients.
+What is surprising, given the `r diff(range(cc$crab_data_shallow_unscaled$depth))` m maximum depth difference among shallow ARMS, is the consistent significance of depth as a driver of community composition across these gradients.
 
@@
 Similar fine-scale depth patterns have been found in previous studies, both using molecular [e.g., @desouzaImportanceDepthTemperature2023; @westEDNAMetabarcodingSurvey2020] and traditional survey methods [e.g., @williamsBenthicCommunitiesTwo2013; @srinivasanDepthDistributionsCoral2003], but our results recover them alongside numerous other geographic and environmental gradients and across a broad geographic scale.
-Depth in and of itself is unlikely to affect marine community composition and oceanographic factors with well-known depth profiles (e.g., temperature, oxygen, nutrients, etc.) are not expected to vary dramatically across such a narrow depth range (\< 10 m), although internal waves can deliver variable temperatures across shallow reefs [@storlazziInternalTidesCan2020; @wyattHeatAccumulationCoral2020] and fine-scale habitat variability among sites may also influence the community differences we found [@steyaertRemoteReefCryptobenthic2022].
+Depth in and of itself is unlikely to affect marine community composition and oceanographic factors with well-known depth profiles (e.g., temperature, oxygen, nutrients, etc.) are not expected to vary dramatically across such a narrow depth range (\< 10 m) in Hawai&#x02BB;i, although internal waves can deliver variable temperatures across shallow reefs [@storlazziInternalTidesCan2020; @wyattHeatAccumulationCoral2020] and fine-scale habitat variability among sites may also influence the community differences we found [@steyaertRemoteReefCryptobenthic2022].
 Unfortunately, we lack high resolution environmental data at depth for individual deployments, and thus cannot unravel the co-varying factor(s) associated with the depth effects we observed for shallow assemblages.
@@
 The evident groupings in the CAP ordination match these trends, with 30 m samples lying closer to shallow O&#x02BB;ahu samples and 60--90 m samples forming a distinct cluster.
-This pattern of community transition from shallow to mesophotic reefs is consistent with what has been shown for various other taxonomic groups [@lesserGlobalCommunityBreaks2019; @rochaMesophoticCoralEcosystems2018] and subsets of entire communities [@hobanPlumbingDepthsEnvironmental2023], lending support to the hypothesis that there may be broad-scale depth-based assembly rules for coral reef communities.
+This pattern of community transition from shallow to mesophotic reefs is consistent with what has been shown for various other taxonomic groups [@lesserGlobalCommunityBreaks2019; @rochaMesophoticCoralEcosystems2018; @johnsonDepthPartitioningMesophotic2025] and subsets of entire communities [@hobanPlumbingDepthsEnvironmental2023], lending support to the hypothesis that there may be broad-scale depth-based assembly rules for coral reef communities.
 
@@
 A reduced version of the CAP ordination filtered to include only ARMS units deployed during these overlapping time periods retains the same (if not more pronounced) overall patterns as the full version (Supplementary [@fig:fig_supp_cap_reduced]), suggesting that the inclusion of prior or subsequent soaking periods did not adversely skew our analyses.
+Aditionally, soak time differed by one year between O&#x02BB;ahu mesophotic ARMS units and those deployed across the rest of the archipelago.
+This difference may potentially bias results, but since CAP ordinations show differences despite the extra year that shallow units spent in the water, we feel that the patterns we observed reflect a true separation.
 The limitation of our depth transect to O&#x02BB;ahu means that we cannot make broad-scale conclusions about biodiversity patterns on deep reefs across the Hawaiian Islands.
@@
 In the Hawaiian archipelago, we found significant differences in the composition of shallow coral reef crab communities between the populated high islands of the MHI and the uninhabited atolls of the NWHI.
-Moreover, these differences were driven by a combination of environmental (sea-surface temperature, chlorophyll-A concentration, depth), biotic (potential larval immigration), and anthropogenic (cumulative human impacts) factors.
+Moreover, these differences were driven by a combination of environmental (sea-surface temperature, chlorophyll-A concentration, depth, island slope), biotic (potential larval immigration), and anthropogenic (cumulative human impacts) factors.
 Despite the significance of these factors in driving shallow community variation across the archipelago, mesophotic assemblages were found to be distinct ([@fig:fig_cap]).
@@
 
-![Autonomous reef monitoring structures (ARMS) deployed in the Hawaiian Archipelago.](data/images/arms_example.png){#fig:fig_arms}
+![Autonomous reef monitoring structures (ARMS) deployed in the Hawaiian Archipelago.
+Photo credits: the authors and NOAA ([uncopyrighted](https://www.omao.noaa.gov/image-licensing-usage-info)).](data/images/arms_example.png){#fig:fig_arms}
 
@@
 A total of 89 ARMS units were deployed over the course of six years (2010--2016) in shallow (`r dr(cc$crabs_shallow_untransformed)` m depth) coral reefs in both the Main (MHI) and Northwest Hawaiian Islands (NWHI) ([@fig:fig_map; @tbl:tbl_arms]).
-All units included in this study were deployed in forereef habitat.
+All units included in this study were deployed in forereef habitat and had a soaking period of three years (2010--2013 or 2013--2016).
 To explore the effects of depth, we also present data from an earlier study by Hurley et al.
@@
 Shallow units were deployed according to NOAA-RAMP methodology, whereas units $\geq$ 30 m were deployed and recovered using technical mixed gas closed-circuit rebreather diving (see [@hurleyAssessmentShallowMesophotic2016]).
+Units from this study soaked for two years during 2009--2011 or 2010--2012.
 Except when discussing specific depth zones, we will refer to any units deployed at `r dr(cc$crabs_shallow_untransformed)` m as "shallow" and those deployed $\geq$ 30 m as "deep".
@@
   geom_sf(data=pmnm,color=NA,fill="#6ed8f6",alpha=0.2) +
-  geom_sf(data=crab_sites %>% st_transform(transformer),aes(fill=island),size=3,shape=21,color="black") +
+  geom_sf(data=crab_sites %>% st_transform(transformer),aes(fill=island),size=3,shape=21,color="black",alpha=0.7) +
   scale_fill_manual(values=island_pal,name="Island") +
@@
   ylab("Latitude") +
-  theme(legend.text = element_text(size=10),legend.title = element_text(size=12))
+  theme(legend.text = element_text(size=10),legend.title = element_text(size=12)) +
+	guides(fill = guide_legend(override.aes = list(alpha = 1)))
 cat("{#fig:fig_map}")
@@
 To visualize how potential environmental drivers varied across the archipelago, we used a principal components analysis (PCA), plotting the results with points colored by island group (MHI or NWHI).
-We used distance-based redundancy analysis (dbRDA; a form of multivariate multiple regression) with `dbrda` to examine the relative contribution of environmental variables ([@tbl:tbl_env]) to shallow crab community variation across the islands, conditioning the recovery year as a random effect.
+We used distance-based redundancy analysis (dbRDA; a form of multivariate multiple regression) with `dbrda` to examine the relative contribution of environmental variables ([@tbl:tbl_env]) to shallow crab community variation across the islands, conditioning the ARMS unit soak time as a random effect.
 From a model incorporating all potential drivers, we sequentially removed variables with the largest variance inflation factors (VIF; calculated using `vif.cca`) until all values were <3, indicating variables were sufficiently uncorrelated.
@@
 # Tables {-}
-```{r tbl-rda,results="asis",dependson="rda-analysis"}
-print_table(rda_table,"tbl_rda","Significant marginal terms in the dbRDA analysis of crab community variation across the Hawaiian Archipelago",hdr_case="none",accuracy=0.001)
+```{r tbl-rda,results="asis",dependson="dbrda-analysis"}
+print_table(rda_table,"tbl_rda","Significant terms in the dbRDA analysis of crab community variation across the Hawaiian Archipelago",hdr_case="none",accuracy=0.001)
 ```
@@
   plot_annotation(tag_levels = list(letters[1:4])) &
-  theme(plot.tag = element_text(face="bold"))
+  theme(plot.tag = element_text(face="bold", size=16))
 cat("{#fig:fig_supp_alpha hidden=\"y\"}")
@@
   plot_annotation(tag_levels = list(letters[1:4])) &
-  theme(plot.tag = element_text(face="bold"))
+  theme(plot.tag = element_text(face="bold", size = 16))
 cat("{#fig:fig_supp_alpha_main hidden=\"y\"}")
@@
   plot_annotation(tag_levels = list(letters[1:2])) &
-  theme(plot.tag = element_text(face="bold"))
+  theme(plot.tag = element_text(face="bold", size = 16))
 cat("{#fig:fig_supp_alpha2 hidden=\"y\"}")
@@
 cat("{#fig:fig_supp_pairwise hidden=\"y\"}")
+```
+
+\newpage
+
+```{r cap-analysis-oahu,dependson="init",echo=FALSE,message=FALSE,warning=FALSE,results="hide"}
+# reduce crab data to only those deployed from 2010--2013
+crabs_oahu <- prune_samples(
+  with(sample_tibble(cc$crabs),island == "Oahu"),
+  cc$crabs
+)
+# get sample data
+crabs_oahu_data <- sample_tibble(crabs_oahu)
+
+# get otu table
+otus <- otu_table(crabs_oahu) %>%
+  as.data.frame()
+# get sample data
+cd <- crabs_oahu_data %>%
+  mutate(depth_zone = factor(depth_zone,ordered=FALSE)) %>%
+  as.data.frame()
+
+# run CAP analysis, taking care of negative eigenvalues where necessary
+cap_depth <- CAPdiscrim(otus ~ depth_zone, data=cd, dist="bray", axes=2, m=0, add=TRUE)
+# calculate proportion of explained variance
+cap_var <- cap_depth$manova$Eigenvalues / sum(cap_depth$manova$Eigenvalues)
+cap_var <- list(x = cap_var[1], y = cap_var[2])
+```
+```{r fig-supp-cap-oahu, fig.cap="CAP ordination of crab communities by depth zone for the island of O&#x02BB;ahu.", results="asis", dev="svg", fig.width=9, fig.height=6, dependson="cap-analysis-oahu"}
+# make the sample group look nice
+# (possibly already defined above, but let's do it here just for good measure)
+nice <- function(s) {
+  str_replace(s,"shallow","Shallow") %>%
+    str_replace("main","(MHI)") %>%
+    str_replace("nwhi","(NWHI)") %>%
+    str_replace("(^[0-9]+)","O‘ahu \\1") %>%
+    str_split("((?<=[0-9])(?=[a-zA-Z]))|(_)") %>%
+    map_chr(~str_c(.x,collapse=" "))
+}
+
+# create tibble of cap results and join in sample data
+pcap_oahu <- cap_depth$x %>%
+  as_tibble(rownames="sample") %>%
+  left_join(cd,by="sample") %>%
+  mutate(
+    depth_zone = nice(depth_zone),
+    depth_zone = fct_reorder(depth_zone,depth)
+  )
+
+# get centroids of each point cluster
+centroids_reduced <- pcap_oahu %>% group_by(depth_zone) %>%
+  summarise(LD1=mean(LD1),LD2=mean(LD2))
+# make an oahu/not oahu factor
+pcap_oahu <- pcap_oahu %>%
+  mutate(
+    oahu = factor(
+      case_when(
+        island == "Oahu" ~ "oahu",
+        # island == "French Frigate Shoals" ~ "ffs",
+        .default = "noahu"
+      ),
+      levels = c("oahu","noahu")
+    )
+  )
+
+# plot the thing
+ggplot(pcap_oahu,aes(x=LD1,y=LD2,fill=depth_zone)) +
+  stat_ellipse(geom="polygon",level=0.95,alpha=0.1,show.legend = FALSE) +
+  geom_point(mapping=aes(shape=oahu),color="black",size=5)  +
+  geom_label_repel(data=centroids_reduced,aes(label=depth_zone), color="white", show.legend = FALSE) +
+  # scale_shape_manual(values=c("oahu" = 23, "ffs" = 22, "noahu" = 21),name="Island", labels=c("noahu"="Other islands","oahu"="O‘ahu","ffs"="Lalo (French Frigate Shoals)" )) +
+  scale_shape_manual(values=c("oahu" = 23, "noahu" = 21),name="Island", labels=c("noahu"="Other islands","oahu"="O‘ahu")) +
+  scale_fill_manual(values=sample_depth_pal,name="Sample grouping") +
+  guides(
+    fill=guide_legend(override.aes=list(shape=21)),
+    shape=guide_legend(override.aes=list(fill="black"))
+  ) +
+  xlab(str_glue("Linear discriminant 1 [{pc(cap_var$x,0.1)}]")) +
+  ylab(str_glue("Linear discriminant 2 [{pc(cap_var$y,0.1)}]")) +
+  plotz_theme()
+cat("{#fig:fig_supp_cap_oahu hidden=\"y\"}")
 ```

### 2025-05-19 manuscript.Rmd
move multivariate stats details to supplemental table

@@
 
+# get the p-value from an anova output
+apval <- function(aaa, modl_term = NULL) {
+  if (!is(aaa,'anova')) {
+    stop("wrong thing")
+  }
+  m <- suppressWarnings(broom::tidy(aaa)) %>%
+    filter( !term %in% c("Total","Residual") )
+  if (!is.null(modl_term)) {
+    m <- m %>% filter(term == modl_term)
+  }
+  if (nrow(m) > 0) {
+    pval(m$p.value[1])
+  } else {
+    stop("nothing in the model anymore")
+  }
+}
+
 # format permdisp test results
@@
 ```
-Shallow crab community structure was significantly different between island groups (MHI vs.
-NWHI) (`r permanova(anova_region,0.01)`) as well as among individual islands (`r permanova(anova_island,0.01)`; see Supplementary [@fig:fig_supp_pairwise] for results of pairwise comparisons).
-There was no significant difference in community dispersion between the MHI and NWHI (`r permdisp(bd_region)`).
-When mesophotic ($\geq$ 30 m) ARMS units were included, there were significant differences among depth zones (\< 30 m, 30 m, 60 m, 90 m) both for all islands (`r permanova(anova_depth,0.01)`) and samples restricted to O&#x02BB;ahu (`r permanova(anova_depth_oahu,0.01)`), and community dispersion (`r permdisp(bd_depth)`).
-However, community dispersion was not different among depth zones $\geq$ 30 m (`r permdisp(bd_deep)`).
-Although crab communities were significantly different among both depth zones and island groups for the complete dataset (`r permanova(anova_both,0.01)`), depth explained `r num(effect_ratio,0.1)` times more variability than island group (based on model $\mathrm{R}^2$).
+Shallow crab assemblages were significantly different between island groups (MHI vs.
+NWHI; `r apval(anova_region)`) and among individual islands (`r apval(anova_island)`; see Supplementary [@fig:fig_supp_pairwise] for results of pairwise comparisons).
+There was no significant difference in community dispersion between the MHI and NWHI (`r apval(anova(bd_region,perm=PERM))`) for shallow crab assemblages.
+Crab assemblages were significantly different among depth zones (\< 30 m, 30 m, 60 m, 90 m) both when comparing shallow assemblages at all islands to deep assemblages on O&#x02BB;ahu (`r apval(anova_depth)`) and when comparing assemblages on O&#x02BB;ahu alone (`r apval(anova_depth_oahu)`).
+Community dispersion was significantly different among depth zones for comparison among all islands (`r apval(anova(bd_depth,perm=PERM))`), however community dispersion was not different among depth zones $\geq$ 30 m (`r apval(anova(bd_deep,perm=PERM))`).
+Although crab assemblages were significantly different among both depth zones (`r apval(anova_both,"depth_zone")`) and island groups (`r apval(anova_both,"island_group")`) in a combined model the complete dataset, depth explained `r num(effect_ratio,0.1)` times more variability than island group (based on model term $\mathrm{R}^2$).
+For detailed descriptions of model results, see Supplementary [@tbl:tbl_supp_stats].
 
@@
 
+```{r tbl-supp-stats, results="asis", dependson="permanova"}
+suppressWarnings({
+  all_terms <- bind_rows(
+    broom::tidy(anova_island) %>%
+      mutate(model = "crabs_shallow ~ island"),
+    broom::tidy(anova_region) %>%
+      mutate(model = "crabs_shallow ~ island_group"),
+    broom::tidy(anova_depth) %>%
+      mutate(model = "crabs_all ~ depth_zone"),
+    broom::tidy(anova_depth_oahu) %>%
+      mutate(model = "crabs_oahu ~ depth_zone"),
+    broom::tidy(anova_both) %>%
+      mutate(model = "crabs_all ~ island_group + depth_zone"),
+    broom::tidy(anova(bd_region,PERM)) %>%
+      rename(SumOfSqs=sumsq,R2=meansq) %>%
+      mutate(term = replace(term,which(term == "Groups"),"island_group")) %>%
+      mutate(model = "crabs_shallow ~ island_group\n\n(permdisp)"),
+    broom::tidy(anova(bd_depth,PERM)) %>%
+      rename(SumOfSqs=sumsq,R2=meansq) %>%
+      mutate(term = replace(term,which(term == "Groups"),"depth_zone")) %>%
+      mutate(model = "crabs_all ~ depth_zone\n\n(permdisp)"),
+    broom::tidy(anova(bd_deep,PERM)) %>%
+      rename(SumOfSqs=sumsq,R2=meansq) %>%
+      mutate(term = replace(term,which(term == "Groups"),"depth_zone")) %>%
+      mutate(model = "crabs_deep ~ depth_zone\n\n(permdisp)")
+  ) %>%
+    filter(!term %in% c("Residuals","Residual","Total")) %>%
+		mutate(df = as.character(df),Permutations=as.character(PERM),p.value = pval(p.value)) %>%
+    select(model,everything()) %>%
+    rename(Model = model, Term = term, `DF` = df, `SS`=SumOfSqs, `R^2^`=R2, `Pseudo-F`=statistic,`p-value`=p.value)
+    # rename(Model = model, Term = term, `   DF   ` = df, `Sum of Squares`=SumOfSqs, `$\\mathrm{R}^2$`=R2, `Pseudo-F`=statistic,`p-value`=p.value)
+})
+print_table(all_terms,"tbl_supp_stats","Detailed results of multivariate statistical comparisons",hdr_case="none",accuracy=0.001)
+```
+
+\newpage
+
 ```{r tbl-supp-arms-metadata, results="asis", dependson="init"}

### 2025-05-19 manuscript.Rmd
comment about MCE management

@@
 We additionally find that assemblages along a 90 m depth transect at a single island show greater divergence in canonical space than across the full ~2,500 km gradient of the Hawaiian Archipelago on shallow reefs.
-Our analyses corroborate previous work and show that depth is among the strongest drivers of community structure.
+Our analyses corroborate previous work and show that depth is among the strongest drivers of community structure, highlighting the fact that mesophotic ecosystems constitute distinct habitats and should be considered as such when formulating management and conservation strategies.
 We conclude that some factor(s) covarying with depth exert more influence on cryptobenthic coral reef community structure than those biotic and abiotic gradients that span the entire Hawaiian Archipelago in shallow-water environments.

### 2025-05-19 manuscript.Rmd
modify title

@@
 ---
-title: 'Cryptobenthic reef assemblages are more distinct across a 90 m depth gradient than 2,500 km of shallow marine habitat in the Hawaiian Archipelago'
+title: 'A cryptobenthic reef assemblage is more distinct across a 90 m depth gradient than 2,500 km of shallow marine habitat in the Hawaiian Archipelago'
 author:

### 2025-06-12 manuscript.Rmd
various revisions

@@
 ---
-title: 'A cryptobenthic reef assemblage is more distinct across a 90 m depth gradient than 2,500 km of shallow marine habitat in the Hawaiian Archipelago'
+title: 'Cryptobenthic crab assemblages are more distinct across a 90 m depth gradient than 2,500 km of shallow marine habitat in the Hawaiian Archipelago'
 author:
@@
 ---
+<!-- TODO: specify depth zones shallow, 30m, 60m, 90 m -->
 <!--word count:
@@
 
-Investigations into the factors that shape coral reef communities have been historically limited in depth and taxonomic scope.
-Despite high biodiversity and the recognized importance of mesophotic habitats, and due to logistical challenges, studies have focused largely on conspicuous taxa such as fishes and corals and the shallowest ~30 m.
-Here, we examine the variability of crab assemblages from Autonomous Reef Monitoring Structures across Hawai&#x02BB;i and ~90 m depth.
-We examine the effects of environmental, ecological, and anthropogenic factors (latitude, longitude, depth, years, chlorophyll-A, temperature, slope, coral cover, potential larval immigration, human impacts) on shallow (8--17 m) crab assemblages and compare them with those sampled along a mesophotic depth gradient (12--90 m) on O&#x02BB;ahu.
-Shallow assemblages were significantly different between the inhabited Main and relatively pristine Northwestern Hawaiian Islands and associated with sea-surface temperature, chlorophyll-A, depth, slope, potential larval immigration, and human impacts.
+Investigations into factors that shape coral reef communities have been historically limited in depth and taxonomic scope.
+Despite high biodiversity and the recognized importance of mesophotic habitats, and due to logistical challenges, studies have focused largely on conspicuous taxa such as fishes and corals and shallow habitat &lt;30 m.
+Here, we examined the variability of crab assemblages from Autonomous Reef Monitoring Structures deployed on shallow reefs across the Hawaiian Islands and a mesophotic depth gradient on O&#x02BB;ahu.
+We tested the effects of environmental, ecological, and anthropogenic factors<!-- (latitude, longitude, depth, years, chlorophyll-A, temperature, slope, coral cover, potential larval immigration, human impacts)--> on shallow (8--17 m) crab assemblages.
+We additionally compared shallow assemblages to those sampled along a depth gradient (12--90 m) on O&#x02BB;ahu.
+Shallow assemblages were significantly different between the inhabited Main and the relatively pristine uninhabited Northwestern Hawaiian Islands and were associated with sea-surface temperature, chlorophyll-A, depth, island slope, potential larval immigration, and human impacts.
 Despite this variability, differences along the depth gradient on O&#x02BB;ahu alone were greater than among the shallow assemblages across the entire archipelago.
@@
 While the structure and composition of coral reef communities is influenced by a suite of environmental and anthropogenic factors, the nature and makeup of those factors remain poorly resolved [@hughesCommunityStructureDiversity1989; @smithEffectsTopBottom2010; @brandlCoralReefEcosystem2019].
-One aspect of coral reef community structure thought to be important but remains under-investigated is depth.
-Most studies to date have focused on the upper ~30 m of the water column, yet coral reefs and associated communities are known to extend beyond 150 m [@pyleExploringDeepCoral1996; @pyleMesophoticCoralEcosystems2019], with photosymbiotic corals reported to 170 m [@rouzeSymbioticAssociationsDeepest2021].
+One aspect of coral reef community structure thought to be important that remains under-investigated is depth (although interest in this area has grown considerably in recent years [@radiceRecentTrendsBiases2024]).
+Many studies to date have focused on the upper ~30 m of the water column, yet coral reefs and associated communities are known to extend beyond 150 m [@pyleExploringDeepCoral1996; @pyleMesophoticCoralEcosystems2019], with photosymbiotic corals reported to 170 m [@rouzeSymbioticAssociationsDeepest2021].
 As much as two thirds of coral reef ecosystems lie in this under-explored zone [@hindersteinThemeSectionMesophotic2010].
 These habitats, known as mesophotic coral ecosystems (MCEs), are typically defined as those extending from ~30--150 m depth [@hindersteinThemeSectionMesophotic2010; @kahngCommunityEcologyMesophotic2010].
-They have been hypothesized as possible refugia from climate change impacts on shallow reefs [@glynnCoralReefBleaching1996; @bongaertsAssessingDeepReef2010; @smithDepthRefugiumCatastrophic2014; @holsteinFertileFathomsDeep2015; @muirSpeciesIdentityDepth2017], although a growing number of studies have found evidence that mesophotic reefs are unique and uniquely threatened, challenging this notion [@diazLightTemperatureDrive2023; @diazMesophoticCoralBleaching2023; @hobanPlumbingDepthsEnvironmental2023; @hurleyAssessmentShallowMesophotic2016; @kaneHighLevelsMesophotic2014; @pinheiroDeepReefFishes2019; @rochaMesophoticCoralEcosystems2018; @montgomeryCommunitySimilaritySpecies2021].
+They have been hypothesized as possible refugia from climate change impacts on shallow reefs [@glynnCoralReefBleaching1996; @bongaertsAssessingDeepReef2010; @smithDepthRefugiumCatastrophic2014; @holsteinFertileFathomsDeep2015; @muirSpeciesIdentityDepth2017], although a growing number of studies have found evidence that MCEs are unique and uniquely threatened, challenging this notion [@diazLightTemperatureDrive2023; @diazMesophoticCoralBleaching2023; @hobanPlumbingDepthsEnvironmental2023; @hurleyAssessmentShallowMesophotic2016; @kaneHighLevelsMesophotic2014; @pinheiroDeepReefFishes2019; @rochaMesophoticCoralEcosystems2018; @montgomeryCommunitySimilaritySpecies2021].
 
@@
 This chain of islands comprises two regions, the inhabited high islands of the main Hawaiian Islands (MHI) and the uninhabited low islands, atolls, and banks of the Northwestern Hawaiian Islands (NWHI) which reside within one of the world's largest marine protected areas---the Papahānaumokuākea Marine National Monument.
-The island chain is subject to considerable anthropogenic and physical gradients driving differences in temperature, coral cover, human population, and anthropogenic impact, among others, all of which influence coral reef communities [@selkoeMapHumanImpacts2009].
+The island chain is subject to considerable anthropogenic and physical gradients driving differences in temperature, coral cover, human population, and anthropogenic impact, among others, all of which influence coral reefs and associated communities [@selkoeMapHumanImpacts2009].
 Nearshore drop-offs along the island chain also allow access to deeper reef habitat, enabling the examination of depth influences on community structure.
@@
 However, the most frequently-targeted taxa make up a minority of coral reef biodiversity and thus may not be indicative of community patterns overall [@fautinOverviewMarineBiodiversity2010].
-Furthermore, studies on these taxa have also been criticized for confounding multiple potentially covarying factors (such as habitat area and/or complexity) across the measured gradients [e.g., @bellwoodRegionalscaleAssemblyRules2001; @grahamEffectsMarineReserve2003; @connollyCommunityStructureCorals2005; @ewersConfoundingFactorsDetection2006; @fletcherRoleHabitatArea2007].
+Furthermore, some studies examining environmental effects on reef assemblages have been criticized for confounding multiple potentially covarying factors (such as habitat area and/or complexity) across the measured gradients [e.g., @bellwoodRegionalscaleAssemblyRules2001; @grahamEffectsMarineReserve2003; @connollyCommunityStructureCorals2005; @ewersConfoundingFactorsDetection2006; @fletcherRoleHabitatArea2007].
 
@@
 
-Given the strong reported differences between deep and shallow reefs across taxa and the paucity of research on drivers of cryptic coral reef community variation, we sought to answer the following four research questions: (i) How do shallow reef brachyuran crab assemblages vary along the natural and anthropogenic gradients of the Hawaiian Archipelago?
+Given the strong reported differences between deep and shallow reefs across taxa and the paucity of research on drivers of cryptic coral reef community variation, we sought to answer the following four research questions: (i) How do shallow reef brachyuran crab assemblages vary in response to the natural and anthropogenic gradients of the Hawaiian Archipelago?
 (ii) Do shallow reef brachyuran crab assemblages differ between the Main and Northwestern Hawaiian Islands?
@@
   n_distinct()
+
+dz <- cc$crab_data %>%
+  distinct(depth_zone,depth) %>%
+  mutate(depth_zone = fct_reorder(depth_zone,depth)) %>%
+  pull(depth_zone) %>%
+  levels()
 
@@
 `r str_to_sentence(english(spp['shallow']))` distinct taxa (including those collapsed to genus or family) were exclusive to shallow ARMS units and `r spp['deep']` were found only deep (`r format_list(deep_spp)`; Supplementary [@tbl:tbl_supp_species_depth]).
-Out of the `r ntaxa(cc$crabs)` total recovered taxa, only `r length(everywhere)` were found across all `r ndepths` depth zones (`r format_list(everywhere)`).
+Out of the `r ntaxa(cc$crabs)` total recovered taxa, only `r length(everywhere)` were found across all `r ndepths` depth zones (`r format_list(dz)`; `r format_list(everywhere)`).
 For a more in-depth taxonomic treatment of the assemblage composition of mesophotic vs.
@@
 
-```{r fig-cap, fig.cap="CAP ordination of crab communities by island group and depth zone.", results="asis", fig.width=9, fig.height=6, dependson="cap-analysis"}
+```{r fig-cap, fig.cap="Canonical analysis of principal coordinates (CAP) ordination of crab assemblages by island group and depth zone.", results="asis", fig.width=9, fig.height=6, dependson="cap-analysis"}
 centroids <- pcap %>% group_by(sample_group) %>%
@@
 A reduced version of the CAP ordination filtered to include only ARMS units deployed during these overlapping time periods retains the same (if not more pronounced) overall patterns as the full version (Supplementary [@fig:fig_supp_cap_reduced]), suggesting that the inclusion of prior or subsequent soaking periods did not adversely skew our analyses.
-Aditionally, soak time differed by one year between O&#x02BB;ahu mesophotic ARMS units and those deployed across the rest of the archipelago.
+Additionally, soak time differed by one year between O&#x02BB;ahu mesophotic ARMS units and those deployed across the rest of the archipelago.
 This difference may potentially bias results, but since CAP ordinations show differences despite the extra year that shallow units spent in the water, we feel that the patterns we observed reflect a true separation.
@@
 A total of 89 ARMS units were deployed over the course of six years (2010--2016) in shallow (`r dr(cc$crabs_shallow_untransformed)` m depth) coral reefs in both the Main (MHI) and Northwest Hawaiian Islands (NWHI) ([@fig:fig_map; @tbl:tbl_arms]).
-All units included in this study were deployed in forereef habitat and had a soaking period of three years (2010--2013 or 2013--2016).
+All shallow units included in this study were deployed in forereef habitat and had a soaking period of three years (2010--2013 or 2013--2016).
 To explore the effects of depth, we also present data from an earlier study by Hurley et al.

### 2025-06-14 manuscript.Rmd
more revision updatres

@@
 ---
-<!-- TODO: specify depth zones shallow, 30m, 60m, 90 m -->
 <!--word count:
@@
 
-Investigations into factors that shape coral reef communities have been historically limited in depth and taxonomic scope.
-Despite high biodiversity and the recognized importance of mesophotic habitats, and due to logistical challenges, studies have focused largely on conspicuous taxa such as fishes and corals and shallow habitat &lt;30 m.
+Investigations into factors that shape coral reef communities have been historically limited in sampling depth and taxonomic scope.
+Despite high biodiversity and the recognized importance of mesophotic habitats, studies have focused largely on conspicuous taxa such as fishes and corals and shallow habitat &lt;30 m.
 Here, we examined the variability of crab assemblages from Autonomous Reef Monitoring Structures deployed on shallow reefs across the Hawaiian Islands and a mesophotic depth gradient on O&#x02BB;ahu.
-We tested the effects of environmental, ecological, and anthropogenic factors<!-- (latitude, longitude, depth, years, chlorophyll-A, temperature, slope, coral cover, potential larval immigration, human impacts)--> on shallow (8--17 m) crab assemblages.
+We tested the effects of environmental, ecological, and anthropogenic factors on shallow (8--17 m) crab assemblages.
 We additionally compared shallow assemblages to those sampled along a depth gradient (12--90 m) on O&#x02BB;ahu.
@@
 While the structure and composition of coral reef communities is influenced by a suite of environmental and anthropogenic factors, the nature and makeup of those factors remain poorly resolved [@hughesCommunityStructureDiversity1989; @smithEffectsTopBottom2010; @brandlCoralReefEcosystem2019].
-One aspect of coral reef community structure thought to be important that remains under-investigated is depth (although interest in this area has grown considerably in recent years [@radiceRecentTrendsBiases2024]).
+One aspect of coral reef community structure thought to be important that remains under-investigated is depth (although interest in this area has grown substantially in recent years [@radiceRecentTrendsBiases2024]).
 Many studies to date have focused on the upper ~30 m of the water column, yet coral reefs and associated communities are known to extend beyond 150 m [@pyleExploringDeepCoral1996; @pyleMesophoticCoralEcosystems2019], with photosymbiotic corals reported to 170 m [@rouzeSymbioticAssociationsDeepest2021].
 As much as two thirds of coral reef ecosystems lie in this under-explored zone [@hindersteinThemeSectionMesophotic2010].
-These habitats, known as mesophotic coral ecosystems (MCEs), are typically defined as those extending from ~30--150 m depth [@hindersteinThemeSectionMesophotic2010; @kahngCommunityEcologyMesophotic2010].
-They have been hypothesized as possible refugia from climate change impacts on shallow reefs [@glynnCoralReefBleaching1996; @bongaertsAssessingDeepReef2010; @smithDepthRefugiumCatastrophic2014; @holsteinFertileFathomsDeep2015; @muirSpeciesIdentityDepth2017], although a growing number of studies have found evidence that MCEs are unique and uniquely threatened, challenging this notion [@diazLightTemperatureDrive2023; @diazMesophoticCoralBleaching2023; @hobanPlumbingDepthsEnvironmental2023; @hurleyAssessmentShallowMesophotic2016; @kaneHighLevelsMesophotic2014; @pinheiroDeepReefFishes2019; @rochaMesophoticCoralEcosystems2018; @montgomeryCommunitySimilaritySpecies2021].
+These habitats, known as mesophotic coral ecosystems (MCEs), are typically defined as extending from ~30--150 m depth [@hindersteinThemeSectionMesophotic2010; @kahngCommunityEcologyMesophotic2010].
+They have been proposed as possible refugia from climate change impacts on shallow reefs [@glynnCoralReefBleaching1996; @bongaertsAssessingDeepReef2010; @smithDepthRefugiumCatastrophic2014; @holsteinFertileFathomsDeep2015; @muirSpeciesIdentityDepth2017], although a growing number of studies have found evidence that MCEs are unique and uniquely threatened, challenging this notion [@diazLightTemperatureDrive2023; @diazMesophoticCoralBleaching2023; @hobanPlumbingDepthsEnvironmental2023; @hurleyAssessmentShallowMesophotic2016; @kaneHighLevelsMesophotic2014; @pinheiroDeepReefFishes2019; @rochaMesophoticCoralEcosystems2018; @montgomeryCommunitySimilaritySpecies2021].
 
 The Hawaiian Archipelago, spanning ~2,500 km in the North Pacific Ocean, provides a natural laboratory to examine drivers of metazoan community structure.
-This chain of islands comprises two regions, the inhabited high islands of the main Hawaiian Islands (MHI) and the uninhabited low islands, atolls, and banks of the Northwestern Hawaiian Islands (NWHI) which reside within one of the world's largest marine protected areas---the Papahānaumokuākea Marine National Monument.
-The island chain is subject to considerable anthropogenic and physical gradients driving differences in temperature, coral cover, human population, and anthropogenic impact, among others, all of which influence coral reefs and associated communities [@selkoeMapHumanImpacts2009].
+This chain of islands comprises two regions, the inhabited high islands of the main Hawaiian Islands (MHI) and the uninhabited low islands, atolls, and banks of the Northwestern Hawaiian Islands (NWHI) which reside within one of the world's largest marine protected areas---Papahānaumokuākea Marine National Monument.
+The island chain is subject to various anthropogenic and physical gradients driving differences in temperature, coral cover, human population, and anthropogenic impact, among others, all of which influence coral reefs and associated communities [@selkoeMapHumanImpacts2009].
 Nearshore drop-offs along the island chain also allow access to deeper reef habitat, enabling the examination of depth influences on community structure.
@@
 
-Here, we used autonomous reef monitoring structures (ARMS)---standardized collection devices designed to mimic reef structural complexity---to sample cryptic communities while controlling for habitat [@zimmermanArtificialReefMatrix2004; @brainardAutonomousReefMonitoring2009].
+Here, we used autonomous reef monitoring structures (ARMS)---standardized collection devices designed to mimic reef structural complexity---to sample cryptic brachyuran crab communities while controlling for habitat [@zimmermanArtificialReefMatrix2004; @brainardAutonomousReefMonitoring2009].
 Such standardization can reduce variability in recovered communities that may arise due to local variation in habitat composition and availability [@ransomeImportanceStandardizationBiodiversity2017].
 This study builds on on previous work targeting brachyuran crabs [@hurleyAssessmentShallowMesophotic2016], a ubiquitous group often used as an exemplar to assess coral reef biodiversity [@knowltonCoralReefBiodiversity2010; @plaisanceDiversityCoralReefs2011; @servisCharacterizingCoralReef2020; @plaisanceReefassociatedCrustaceanFauna2009] and comprising a notable proportion of the Hawaiian coral reef cryptofauna (284 reported species [@castroCatalogAnomuranBrachyuran2011]).
-We used assemblages recovered from ARMS units to examine drivers of community composition across the Hawaiian Archipelago.
-We compared crab communities among islands and regions (MHI vs.
+We used crab assemblages recovered from ARMS units to examine drivers of community composition across the Hawaiian Archipelago.
+We compared assemblages among islands and regions (MHI vs.
 NWHI), among major depth zones (shallow vs.
@@
 The `r nsamples(cc$crabs_untransformed)` ARMS units deployed across the Hawaiian Islands (in `r format_range(cc$crab_data$depth)` m depth) yielded a total of `r num(sum(sample_sums(cc$crabs_untransformed)),1)` adult crabs.
-Units contained `r num(all_summary$abundance_mean,0.1)` $\pm$ `r num(all_summary$abundance_sd,0.1)` (mean $\pm$ SD) individuals, and abundance ranged from `r format_range(all_summary$abundance_range)`.
+Units contained `r num(all_summary$abundance_mean,0.1)` $\pm$ `r num(all_summary$abundance_sd,0.1)` (mean $\pm$ SD) individuals, and abundance ranged from `r format_range(all_summary$abundance_range)` individuals per unit.
 Crabs were identified to `r sum(unlist(all_summary$types_breakdown))` distinct taxa (`r all_summary$types_breakdown$species` species-level identifications, `r all_summary$types_breakdown$genus` genus, `r all_summary$types_breakdown$family` family) encompassing `r all_summary$taxonomy_breakdown$genera` genera in `r all_summary$taxonomy_breakdown$families` families, which accounted for ~`r pc(all_summary$types_breakdown$species / hawaiian_crabs,0.1)` of the known crab diversity reported from Hawaiian waters for species-level IDs [@castroCatalogAnomuranBrachyuran2011].
@@
 
-![Representative photographs of the four most abundant crab species found across all deployed ARMS units (pictured specimens were not collected as part of this study).
-Individual scale information was not available, but all specimens were <3.5 cm in carapace width.
+![Photographs of the four most abundant crab species found across all deployed ARMS units (pictured are representative museum specimens from Hawai&#x02BB;i, not collected for this study).
+Specimen size data was not retained, but all individuals were &lt;3.5 cm in carapace width.
 (a) *Chlorodiella* cf.
@@
 
-The `r nsamples(cc$crabs_deep_untransformed)` units deployed at mesophotic depths $\geq$ 30 m on O&#x02BB;ahu yielded fewer species and individuals compared to the complete dataset, with `r num(sum(sample_sums(cc$crabs_deep_untransformed)),1)` crabs overall, `r sum(unlist(deep_summary$types_breakdown))` distinct taxa (`r deep_summary$types_breakdown$species` species-level identifications, `r deep_summary$types_breakdown$genus` genus, `r deep_summary$types_breakdown$family` family), `r num(deep_summary$abundance_mean,0.1)` $\pm$ `r num(deep_summary$abundance_sd,0.1)` individuals per unit, and abundance ranging from `r format_range(deep_summary$abundance_range)`.
+The `r nsamples(cc$crabs_deep_untransformed)` units deployed at mesophotic depths $\geq$ 30 m on O&#x02BB;ahu <!--yielded fewer species and individuals compared to the complete dataset, with --> contained `r num(sum(sample_sums(cc$crabs_deep_untransformed)),1)` crabs overall, `r sum(unlist(deep_summary$types_breakdown))` distinct taxa (`r deep_summary$types_breakdown$species` species-level identifications, `r deep_summary$types_breakdown$genus` genus, `r deep_summary$types_breakdown$family` family), `r num(deep_summary$abundance_mean,0.1)` $\pm$ `r num(deep_summary$abundance_sd,0.1)` individuals per unit, and abundance ranging from `r format_range(deep_summary$abundance_range)`.
 The four most abundant taxa in the deep ARMS did not overlap with the most abundant taxa from the complete dataset: `r with(deep_summary,top_list(top,top_n,top_units))`.
@@
 `r str_to_sentence(english(spp['shallow']))` distinct taxa (including those collapsed to genus or family) were exclusive to shallow ARMS units and `r spp['deep']` were found only deep (`r format_list(deep_spp)`; Supplementary [@tbl:tbl_supp_species_depth]).
-Out of the `r ntaxa(cc$crabs)` total recovered taxa, only `r length(everywhere)` were found across all `r ndepths` depth zones (`r format_list(dz)`; `r format_list(everywhere)`).
+Out of the `r ntaxa(cc$crabs)` total taxa, `r length(everywhere)` were found across all `r ndepths` depth zones (`r format_list(dz)`; `r format_list(everywhere)`).
 For a more in-depth taxonomic treatment of the assemblage composition of mesophotic vs.
 shallow ARMS units, see [@hurleyAssessmentShallowMesophotic2016].
-
+<!--
 ## Alpha diversity {-}
@@
 When only shallow arms \< 30 m were considered, there were no significant differences in alpha diversity between the MHI and NWHI for either index (richness: `r k_test(shallow_alpha$k_r_island,plain=TRUE)`, Supplementary [@fig:fig_supp_alpha2[]{}a]; effective Simpson: `r k_test(shallow_alpha$k_s_island,plain=TRUE)`, Supplementary [@fig:fig_supp_alpha2[]{}b]).
-
-<!-- ### Effects of environmental variables {-} -->
+-->
+
 ## Brachyuran community structure {-}
@@
 ### Environmental drivers {-}
-The PCA of environmental variables indicated strong clustering according to island group (Supplementary [@fig:fig_supp_pca]), with apparently orthogonal variation in variables associated with the MHI vs the NWHI.
+The PCA of environmental variables indicated strong clustering according to island group (Supplementary [@fig:fig_supp_pca]), with orthogonal variation in variables associated with the MHI vs the NWHI.
 The best dbRDA model (`r permanova(db_modl,stat="F")`) explained `r pc(RsquareAdj(db_modl)$adj.r.squared,0.01)` (adjusted $\mathrm{R}^2$) of the variability in crab communities across the archipelago ([@fig:fig_dbrda]).
@@
 ```{r fig-dbrda, fig.cap="dbRDA plot showing significant drivers of crab community variation across the Hawaiian Islands.
-Colored ellipses represent 95% confidence intervals around island groups.", results="asis", fig.width=9, fig.height=6, dependson="dbrda-analysis"}
+Colored ellipses represent 95% confidence intervals around island groups (Main vs.
+Northwestern Hawaiian Islands).", results="asis", fig.width=9, fig.height=6, dependson="dbrda-analysis"}
 termz <- model_terms %>%
@@
   pull(R2)
+cumulative_effect <- cumsum(effects)[2]
 effect_ratio <- max(effects)/min(effects)
@@
 Community dispersion was significantly different among depth zones for comparison among all islands (`r apval(anova(bd_depth,perm=PERM))`), however community dispersion was not different among depth zones $\geq$ 30 m (`r apval(anova(bd_deep,perm=PERM))`).
-Although crab assemblages were significantly different among both depth zones (`r apval(anova_both,"depth_zone")`) and island groups (`r apval(anova_both,"island_group")`) in a combined model the complete dataset, depth explained `r num(effect_ratio,0.1)` times more variability than island group (based on model term $\mathrm{R}^2$).
-For detailed descriptions of model results, see Supplementary [@tbl:tbl_supp_stats].
+Crab assemblages were significantly different among both depth zones (`r apval(anova_both,"depth_zone")`) and island groups (`r apval(anova_both,"island_group")`) in a combined model of the complete dataset.
+Although cumulative explanatory power was low (combined $\mathrm{R}^2$: `r num(cumulative_effect,0.01)`), depth ($\mathrm{R}^2=`r num(effects[2],0.001)`$) explained `r num(effect_ratio,0.1)` times more variability than island group ($\mathrm{R}^2=`r num(effects[1],0.001)`$).
+For detailed statistical results, see Supplementary [@tbl:tbl_supp_stats].
 
@@
 The first two linear discriminant axes explained `r pc(cap_var$x,0.1)` and `r pc(cap_var$y,0.1)` of the variation, respectively.
-Mesophotic samples (30--90 m) from O&#x02BB;ahu were grouped separately in ordination space from all shallow samples across the archipelago ([@fig:fig_cap]).
+Mesophotic samples (30--90 m) from O&#x02BB;ahu were grouped separately in ordination space from shallow samples across the archipelago (including O&#x02BB;ahu; [@fig:fig_cap]).
 Shallow samples from O&#x02BB;ahu primarily clustered with shallow samples from other islands ([@fig:fig_cap]).
-A CAP ordination including only samples from O&#x02BB;ahu shows a similar pattern of depth zonation ([@fig:fig_supp_cap_oahu]).
-Variability among 30--90 m samples was apparently greater than among all shallow samples, with mean Euclidean distances between deep samples in canonical space significantly greater than between shallow samples (`r t_test(distance_t)`; [@fig:fig_cap], Supplementary [@fig:fig_supp_euclid]).
-Similarly, variability among samples from O&#x02BB;ahu (including mesophotic depths) exceeded variability among all shallow samples combined (based on Euclidean distances as above; `r t_test(oahu_t,plain=TRUE)`).
+A CAP ordination limited to O&#x02BB;ahu shows similar patterns of depth zonation (Supplementary [@fig:fig_supp_cap_oahu]).
+Variability in canonical space was significantly greater among deep (30--90 m) than among shallow samples ($\le$ 30 m) (Welch two-sample t-test of mean Euclidean distances:  `r t_test(distance_t,plain=TRUE)`; [@fig:fig_cap], Supplementary [@fig:fig_supp_euclid]).
+Similarly, variability within O&#x02BB;ahu alone (when mesophotic samples were included) exceeded the variability among all shallow samples combined (`r t_test(oahu_t,plain=TRUE)`).
 
@@
 
-<!-- TODO: add somewhere something like:
-although depth was significant in the shallows, its relative effect on community differences was less important than other factors (e.g., human impacts, chlorophyll, etc.).
-however, it clearly outpaced other covarying factors along the depth gradient
--->
 # Discussion {-}
 The Hawaiian Archipelago is among the most remote oceanic island chains on the planet and encompasses a range of gradients (anthropogenic, biotic, abiotic), making it an ideal natural laboratory to study the impacts of various drivers of species assemblages across space.
+We investigated the effects of a suite of factors on shallow brachyuran assemblage structure across the Hawaiian Archipelago and compared shallow assemblages to those sampled along a mesophotic depth gradient on O&#x02BB;ahu.
+We found that crab assemblages were more distinct across 90 m of depth on O&#x02BB;ahu alone than across the ~2,500 km archipelago and the influence of diverse significant environmental, anthropogenic, and ecological drivers.
+
 Human activities are considerably higher in the populated MHI than in the protected and remote NWHI [@selkoeEvaluatingAnthropogenicThreats2008; @selkoeMapHumanImpacts2009; @selkoeDNACoralReef2016].
@@
 Regardless, given its significance across both narrow (\< 10 m) and wider (90 m) depth ranges sampled here, a factor associated with depth is clearly an important driver for cryptobenthic crab communities across the Hawaiian Archipelago.
-As brachyurans are a ubiquitous and important component of coral reef communities [@reynsPatternsProcessesBrachyuran1999; @lasleyjrPhylogeneticRelationshipsUbiquitous2015; @hazeriLatitudinalSpeciesDiversity2019], these findings are likely to apply more broadly as well.
+As brachyurans are a ubiquitous and important component of coral reef communities [@reynsPatternsProcessesBrachyuran1999; @lasleyjrPhylogeneticRelationshipsUbiquitous2015; @hazeriLatitudinalSpeciesDiversity2019], these findings may apply more broadly as well.
 
 While depth was a significant driver of community composition among shallow-water ARMS units ([@fig:fig_dbrda, @tbl:tbl_rda]), its influence was considerably more striking when comparing shallow to mesophotic habitat.
-In our CAP ordination, all shallow assemblages across the archipelago clustered separately from the deep assemblages $\geq$ 30 m at the single island of O&#x02BB;ahu ([@fig:fig_cap]).
+In our CAP ordination, shallow assemblages spanning the archipelago clustered separately from the deep assemblages $\geq$ 30 m at the single island of O&#x02BB;ahu ([@fig:fig_cap]).
 Additionally, the two deepest sampled depths (60 m and 90 m) clustered apart from 30 m.
-According to this analysis, there was more community variation among samples spanning 90 m of depth on a single island than across ~2,500 km of shallow habitat experiencing multiple gradients of environmental, ecological, and anthropogenic influence.
+According to this analysis, there was more variation among samples spanning 90 m of depth on a single island than across ~2,500 km of shallow habitat experiencing multiple gradients of environmental, ecological, and anthropogenic influence.
 This is further supported by the fact that the average Euclidean distance among O&#x02BB;ahu samples (including mesophotic) in canonical space was significantly greater than the average distance among all shallow samples (MHI & NWHI).
-Shallow crab communities from O&#x02BB;ahu appear to share more affinity with Hōlanikū (~2,200 km distant) than they do with those at 90 m depth on the same reef slope ([@fig:fig_cap]).
+Shallow crab communities from O&#x02BB;ahu appear to share more affinity with Lalo (~900 km distant) than they do with those at 90 m depth on the same reef slope ([@fig:fig_cap]).
 The CAP also revealed two groups within the mesophotic samples: one at 30 m and the other encompassing the 60--90 m depths.
@@
 The evident groupings in the CAP ordination match these trends, with 30 m samples lying closer to shallow O&#x02BB;ahu samples and 60--90 m samples forming a distinct cluster.
-This pattern of community transition from shallow to mesophotic reefs is consistent with what has been shown for various other taxonomic groups [@lesserGlobalCommunityBreaks2019; @rochaMesophoticCoralEcosystems2018; @johnsonDepthPartitioningMesophotic2025] and subsets of entire communities [@hobanPlumbingDepthsEnvironmental2023], lending support to the hypothesis that there may be broad-scale depth-based assembly rules for coral reef communities.
-
-In addition to community composition, mesophotic samples also exhibited alpha diversity differences.
+This pattern of community transition from shallow to mesophotic reefs is consistent with what has been shown for various other taxonomic groups [@lesserGlobalCommunityBreaks2019; @rochaMesophoticCoralEcosystems2018; @johnsonDepthPartitioningMesophotic2025] and subsets of entire communities [@hobanPlumbingDepthsEnvironmental2023], supporting the possibility that there may be broad-scale depth-based assembly rules for coral reef communities.
+
+<!-- In addition to community composition, mesophotic samples also exhibited alpha diversity differences.
 Effective Simpson diversity was significantly higher in deep communities (for comparisons among all islands and solely within the MHI, but not exclusively at O&#x02BB;ahu) (Supplementary [@fig:fig_supp_alpha; @fig:fig_supp_alpha_main]).
 This pattern is in contrast to observations that biodiversity often decreases with depth [@pyleComprehensiveInvestigationMesophotic2016; @kahngCommunityEcologyMesophotic2010], although the differences we found were slight and our findings may be influenced by differences in habitat availability across depth zones (especially of the habitat types ARMS are designed to mimic).
+-->
 
@@
 Thus, given the assumption of similar temperature profiles near the O&#x02BB;ahu shoreline, temperature alone appears insufficient to explain the zonation patterns we observed---although due to its significant role as a driver of shallow community variation, it cannot be ruled out.
-In contrast, the lower depth boundaries of mesophotic ecosystems are largely defined by surface irradiance [@kahngCommunityEcologyMesophotic2010; @hindersteinThemeSectionMesophotic2010], so light seems a possible candidate as bottom-up abiotic factor influencing community structure across depths.
+In contrast, the lower depth boundaries of mesophotic ecosystems are largely defined by surface irradiance [@kahngCommunityEcologyMesophotic2010; @hindersteinThemeSectionMesophotic2010], so light may exert a bottom-up influence on community structure across depths.
 Crab assemblage structure overall is unlikely to be directly affected by light variability (diurnal behavior patterns of individual species notwithstanding).
@@
 
-Two potential weaknesses of our dataset are the time periods over which ARMS units were deployed and the fact that mesophotic samples were obtained solely from the island of O&#x02BB;ahu.
+Two potential weaknesses of our dataset are the time periods over which ARMS units were deployed and the fact that mesophotic samples were only available from O&#x02BB;ahu.
 Our ARMS units were deployed across a broad range of years, but the majority were deployed during 2010--2012 or 2010--2013 (an overlap of two years) ([@tbl:tbl_arms]).
@@
 The limitation of our depth transect to O&#x02BB;ahu means that we cannot make broad-scale conclusions about biodiversity patterns on deep reefs across the Hawaiian Islands.
-However, it remains the case that community and alpha diversity differences among depth zones at that single island appear to overwhelm differences observed among shallow-water coral reef habitat across the archipelago and depth appears to be the strongest organizing factor in canonical space regardless of where or when the communities were sampled.
+However, it remains the case that differences among depth zones at that single island appear to overwhelm differences observed among shallow-water coral reef habitat across the archipelago and depth appears to be the strongest organizing factor in canonical space regardless of where or when the communities were sampled.
 Similar patterns were found at a local scale (tens of kilometers) on Hawai&#x02BB;i Island using environmental DNA (eDNA) [@hobanPlumbingDepthsEnvironmental2023].
@@
 Although there were variables we did not consider in our model, this suggests that vertical distributions are influenced by factors other than those driving spatial variation across the island chain.
+Our data was collected using ARMS, which in mimicking the reef matrix are biased toward recruiting a specific subset of fauna that prefer cryptic habitats.
+However, the depth zonation patterns that we observed match those observed for multiple other taxa and communities more broadly [@diazLightTemperatureDrive2023; @diazMesophoticCoralBleaching2023; @hobanPlumbingDepthsEnvironmental2023; @hurleyAssessmentShallowMesophotic2016; @kaneHighLevelsMesophotic2014; @pinheiroDeepReefFishes2019; @rochaMesophoticCoralEcosystems2018; @montgomeryCommunitySimilaritySpecies2021].
 In light of this, and since mesophotic communities are generally less biodiverse than shallow reefs [@kahngCommunityEcologyMesophotic2010; @lesserEcologyMesophoticCoral2009; @pyleComprehensiveInvestigationMesophotic2016], it may be possible that the gradients driving depth distributions on coral reefs run opposite to those in the intertidal.
@@
 
-![Autonomous reef monitoring structures (ARMS) deployed in the Hawaiian Archipelago.
+![Example photographs of autonomous reef monitoring structures (ARMS) deployed in the Hawaiian Archipelago.
 Photo credits: the authors and NOAA ([uncopyrighted](https://www.omao.noaa.gov/image-licensing-usage-info)).](data/images/arms_example.png){#fig:fig_arms}
@@
 Units from this study soaked for two years during 2009--2011 or 2010--2012.
-Except when discussing specific depth zones, we will refer to any units deployed at `r dr(cc$crabs_shallow_untransformed)` m as "shallow" and those deployed $\geq$ 30 m as "deep".
+When discussing assemblage depths, we will refer either to one of four specific depth zones (&lt; 30 m or "shallow", 30 m, 60 m, 90 m) or units deployed at `r dr(cc$crabs_shallow_untransformed)` m as "shallow" and $\geq$ 30 m as "deep".
 
@@
 
-To examine patterns of crab diversity across the islands, we used `estimate_richness` to calculate species richness and effective Simpson diversity (effective Simpson = $\frac{1}{\sum_{i=1}^{R}p_i^2}$, where R is the total number of species, and $\mathrm{p}_i$ is the proportional abundance of each species).
+<!--To examine patterns of crab diversity across the islands, we used `estimate_richness` to calculate species richness and effective Simpson diversity (effective Simpson = $\frac{1}{\sum_{i=1}^{R}p_i^2}$, where R is the total number of species, and $\mathrm{p}_i$ is the proportional abundance of each species).
 The effective Simpson index (the inverse of the "classic" Simpson entropy), which has units of species, is easier to interpret in an ecological context than an entropic non-linear index [@jostEntropyDiversity2006; @tuomistoUpdatedConsumerGuide2012].
@@
 NWHI).
-Island group comparisons were performed both including and excluding mesophotic samples.
+Island group comparisons were performed both including and excluding mesophotic samples.-->
 
@@
 
+<!--
 ```{r fig-supp-alpha, fig.cap='Box plots of (a,b) species richness and (c,d) effective Simpson diversity for complete dataset (all sampled depths).
@@
 \newpage
-
+-->
 ```{r fig-supp-pca, fig.cap='Principal component analysis (PCA) of environmental variables by sampling site.

### 2025-06-14 manuscript.Rmd
small typos

@@
 Such standardization can reduce variability in recovered communities that may arise due to local variation in habitat composition and availability [@ransomeImportanceStandardizationBiodiversity2017].
-This study builds on on previous work targeting brachyuran crabs [@hurleyAssessmentShallowMesophotic2016], a ubiquitous group often used as an exemplar to assess coral reef biodiversity [@knowltonCoralReefBiodiversity2010; @plaisanceDiversityCoralReefs2011; @servisCharacterizingCoralReef2020; @plaisanceReefassociatedCrustaceanFauna2009] and comprising a notable proportion of the Hawaiian coral reef cryptofauna (284 reported species [@castroCatalogAnomuranBrachyuran2011]).
+This study builds on previous work targeting brachyuran crabs [@hurleyAssessmentShallowMesophotic2016], a ubiquitous group often used as an exemplar to assess coral reef biodiversity [@knowltonCoralReefBiodiversity2010; @plaisanceDiversityCoralReefs2011; @servisCharacterizingCoralReef2020; @plaisanceReefassociatedCrustaceanFauna2009] and comprising a notable proportion of the Hawaiian coral reef cryptofauna (284 reported species [@castroCatalogAnomuranBrachyuran2011]).
 We used crab assemblages recovered from ARMS units to examine drivers of community composition across the Hawaiian Archipelago.
@@
   select(island,`Depth range`=depth,order(colnames(.)))
-print_table(arms_schedule,"tbl_arms","ARMS unit deployment schedule by island (number of units). Mesophotic deployments apear in bold.",accuracy=1)
+print_table(arms_schedule,"tbl_arms","ARMS unit deployment schedule by island (number of units). Mesophotic deployments appear in bold.",accuracy=1)
 ```

### 2025-06-17 manuscript.Rmd
more revisions

@@
 
+# make a format-compatible comment in the output
+cmt <- function(highlight, comment,  author = "mykle h", time = now(),id = "0")
+{
+  time <- format_ISO8601(time,usetz = "Z")
+	return(
+		sprintf(
+			'[%s]{.comment-start id="%s" author="%s" date="%s"}%s[]{.comment-end id="%s"}',
+			comment,
+			id,
+			author,
+			time,
+			highlight,
+			id
+		)
+	)
+}
+
 # do an english comma-separated list
@@
         island == "Oahu" ~ "oahu",
-        island == "French Frigate Shoals" ~ "ffs",
+        # island == "French Frigate Shoals" ~ "ffs",
         .default = "noahu"
@@
   geom_label_repel(data=centroids,aes(label=sample_group), color="white", show.legend = FALSE) +
-  scale_shape_manual(values=c("oahu" = 23, "ffs" = 22, "noahu" = 21),name="Island", labels=c("noahu"="Other islands","oahu"="O‘ahu","ffs"="Lalo (French Frigate Shoals)" )) +
-  scale_fill_manual(values=sample_depth_pal,name="Sample grouping") +
+  # scale_shape_manual(values=c("oahu" = 23, "ffs" = 22, "noahu" = 21),name="Island", labels=c("noahu"="Other islands","oahu"="O‘ahu","ffs"="Lalo (French Frigate Shoals)" )) +
+  scale_shape_manual(values=c("oahu" = 23, "noahu" = 21),name="Island", labels=c("noahu"="Other islands","oahu"="O‘ahu")) +
+  scale_fill_manual(values=sample_depth_pal,name="Depth/region") +
   guides(
@@
 
+`r cmt("Three of the factors we found to be significantly associated with shallow crab assemblage variation correspond to well-known gradients across the islands: sea-surface temperature, chlorophyll-A, and human impacts","Is there a citation for these gradients existing across the islands?
+I feel like there must be but I don't know it off hand.
+Or is it even necessary because they're cited below?")`.
+While `r cmt("few studies","is this true?
+could probably use more citations here")` have examined the role of sea-surface temperature (SST) in structuring reef communities across the span of the Hawaiian Islands, `r cmt("it was found","probably some more examples here, too")` to be an important factor influencing intertidal communities within the MHI [@wardSpatialVariationAntecedent2023].
+Chlorophyll-A can be used as a proxy for primary productivity.
+Productivity gradients across the Hawaiian Archipelago coincide with differences in coral reef community composition [@selkoeDNACoralReef2016], endemic reef fish dominance [@friedlanderDominanceEndemicsReef2020], survival rates of top predators [@bakerEffectVariableOceanic2007], and island geomorphology [@goveNearislandBiologicalHotspots2016].
 Human activities are considerably higher in the populated MHI than in the protected and remote NWHI [@selkoeEvaluatingAnthropogenicThreats2008; @selkoeMapHumanImpacts2009; @selkoeDNACoralReef2016].
-Although we did not consider the effects of individual impacts here, an index representing *cumulative* anthropogenic impacts was among the significant drivers of shallow brachyuran community structure, similar to findings in previous work examining genetic diversity as it relates to species richness and habitat area, quality, and stability across the archipelago [@selkoeDNACoralReef2016].
-The second most significant driver, chlorophyll-A, can be used as a proxy for primary productivity.
-Productivity gradients across the Hawaiian Archipelago coincide with differences in coral reef community composition [@selkoeDNACoralReef2016], endemic reef fish dominance [@friedlanderDominanceEndemicsReef2020], survival rates of top predators [@bakerEffectVariableOceanic2007], and island geomorphology [@goveNearislandBiologicalHotspots2016].
-Although not a significant factor in our model, coral cover and species composition are thought to affect patterns of benthic assemblages.
+Although we did not consider the effects of individual anthropogenic impacts here, an index representing cumulative impacts was a significant driver of shallow brachyuran community structure, similar to findings in previous work examining the relationships of genetic diversity to species richness and habitat area, quality, and stability across the archipelago [@selkoeDNACoralReef2016].
+Thus, given the prevalence in our model of significant factors known to be important influencers of marine communities in Hawai&#x02BB;i, it is not surprising that we find the same top drivers of community structure for brachyuran crabs, one of the primary taxonomic groups inhabiting coral reefs.
+What is surprising, given the `r diff(range(cc$crab_data_shallow_unscaled$depth))` m maximum depth difference among shallow ARMS, is the consistent significance of depth as a driver of community composition across these gradients.
+
+<!--
+`r cmt("Although it was not a significant factor in our model, coral cover and species composition are thought to affect patterns of benthic assemblages","This bit about coral cover/composition was added in response to a previous reviewer.
+It feels a bit awkward here but not important enough to warrant its own paragraph.
+Do you have any thoughts?")`.
 Due to the limited availability of datasets covering our entire study area, our model only included an estimate of coral cover at the island scale, but previous work has shown that coral species composition does not vary considerably across the archipelago (aside from the presence of *Acropora* in the NWHI) [@griggCommunityStructureSuccession1983].
 Crab communities from Lalo (French Frigate Shoals), which was found to harbor the highest coral species diversity [@griggCommunityStructureSuccession1983], did not group separately from the rest of our samples ([@fig:fig_cap]).
-Given the prevalence in our model of significant factors known to be important influencers of marine communities in Hawai&#x02BB;i, it is not surprising that we find the same top drivers of community structure for brachyuran crabs, one of the primary taxonomic groups inhabiting coral reefs.
-What is surprising, given the `r diff(range(cc$crab_data_shallow_unscaled$depth))` m maximum depth difference among shallow ARMS, is the consistent significance of depth as a driver of community composition across these gradients.
-
-Although our shallow-water ARMS units were deployed within a narrow depth range (`r dr(cc$crabs_shallow_untransformed)` m), depth still emerged as a significant and consistent driver of community variation among these units.
-Similar fine-scale depth patterns have been found in previous studies, both using molecular [e.g., @desouzaImportanceDepthTemperature2023; @westEDNAMetabarcodingSurvey2020] and traditional survey methods [e.g., @williamsBenthicCommunitiesTwo2013; @srinivasanDepthDistributionsCoral2003], but our results recover them alongside numerous other geographic and environmental gradients and across a broad geographic scale.
-Depth in and of itself is unlikely to affect marine community composition and oceanographic factors with well-known depth profiles (e.g., temperature, oxygen, nutrients, etc.) are not expected to vary dramatically across such a narrow depth range (\< 10 m) in Hawai&#x02BB;i, although internal waves can deliver variable temperatures across shallow reefs [@storlazziInternalTidesCan2020; @wyattHeatAccumulationCoral2020] and fine-scale habitat variability among sites may also influence the community differences we found [@steyaertRemoteReefCryptobenthic2022].
+-->
+
+Depth consistently emerged as a significant driver of brachyuran assemblage variation among shallow-water ARMS units, despite the narrow depth range (`r dr(cc$crabs_shallow_untransformed)` m) within which they were deployed.<!-- Although our shallow-water ARMS units were deployed within a narrow depth range (`r dr(cc$crabs_shallow_untransformed)` m), depth consistently emerged as a significant driver of community variation among these units.--> Fine-scale depth patterns have been found in previous studies, both using molecular [@desouzaImportanceDepthTemperature2023; @westEDNAMetabarcodingSurvey2020] and traditional survey methods [@williamsBenthicCommunitiesTwo2013; @srinivasanDepthDistributionsCoral2003], but we recovered them despite numerous other latitudinal and environmental gradients and across a broad spatial scale.
+Depth in and of itself is unlikely to affect marine community composition and oceanographic factors with well-known depth profiles (e.g., temperature, oxygen, nutrients, etc.) are not expected to vary dramatically across such a narrow depth range (\< 10 m) in Hawai&#x02BB;i, although internal waves can deliver variable temperatures across shallow reefs [@storlazziInternalTidesCan2020; @wyattHeatAccumulationCoral2020] and fine-scale habitat variability may also exert an influence [@steyaertRemoteReefCryptobenthic2022].
 Unfortunately, we lack high resolution environmental data at depth for individual deployments, and thus cannot unravel the co-varying factor(s) associated with the depth effects we observed for shallow assemblages.
@@
 While depth was a significant driver of community composition among shallow-water ARMS units ([@fig:fig_dbrda, @tbl:tbl_rda]), its influence was considerably more striking when comparing shallow to mesophotic habitat.
-In our CAP ordination, shallow assemblages spanning the archipelago clustered separately from the deep assemblages $\geq$ 30 m at the single island of O&#x02BB;ahu ([@fig:fig_cap]).
-Additionally, the two deepest sampled depths (60 m and 90 m) clustered apart from 30 m.
-According to this analysis, there was more variation among samples spanning 90 m of depth on a single island than across ~2,500 km of shallow habitat experiencing multiple gradients of environmental, ecological, and anthropogenic influence.
-This is further supported by the fact that the average Euclidean distance among O&#x02BB;ahu samples (including mesophotic) in canonical space was significantly greater than the average distance among all shallow samples (MHI & NWHI).
-Shallow crab communities from O&#x02BB;ahu appear to share more affinity with Lalo (~900 km distant) than they do with those at 90 m depth on the same reef slope ([@fig:fig_cap]).
-The CAP also revealed two groups within the mesophotic samples: one at 30 m and the other encompassing the 60--90 m depths.
+In the CAP ordination, shallow assemblages spanning the archipelago clustered separately from the deep assemblages $\geq$ 30 m at the single island of O&#x02BB;ahu ([@fig:fig_cap])<!-- and within O&#x02BB;ahu the two deepest depths (60 m and 90 m) clustered apart from 30 m-->.
+Further, the average<!-- Euclidean--> distance among O&#x02BB;ahu samples (including 30--90 m) in canonical space was significantly greater than among all shallow samples (MHI and NWHI).
+Thus, there was more variation (both visually and statistically) among assemblages spanning 90 m of depth on a single island than across ~2,500 km of shallow habitat experiencing multiple gradients of environmental, ecological, and anthropogenic influence.<!--This is further supported by the fact that  Shallow crab communities from O&#x02BB;ahu appear to share more affinity with Lalo (~900 km distant) than they do with those at 90 m depth on the same reef slope ([@fig:fig_cap]).--> The CAP ordination also revealed two groups within the mesophotic samples: one at 30 m and the other encompassing the 60--90 m depths.
 These depth zones correspond with the upper and lower mesophotic zones, respectively [@kahngCommunityEcologyMesophotic2010; @rochaMesophoticCoralEcosystems2018; but see @pyleMesophoticCoralEcosystems2019].
-The upper mesophotic zone has been found to constitute an extension of shallow reefs (with beta diversity patterns largely driven by nestedness), whereas the lower mesophotic harbors unique communities (with beta diversity driven primarily by turnover) [@rochaMesophoticCoralEcosystems2018; @slatteryFunctionStabilityMesophotic2024a; @bellwoodRegionalscaleAssemblyRules2001; @kahngCommunityEcologyMesophotic2010].
-The evident groupings in the CAP ordination match these trends, with 30 m samples lying closer to shallow O&#x02BB;ahu samples and 60--90 m samples forming a distinct cluster.
+The upper mesophotic zone has been found to constitute an extension of shallow reefs (with beta diversity patterns largely driven by nestedness), while the lower mesophotic harbors unique communities (with beta diversity driven primarily by turnover) [@rochaMesophoticCoralEcosystems2018; @slatteryFunctionStabilityMesophotic2024a; @bellwoodRegionalscaleAssemblyRules2001; @kahngCommunityEcologyMesophotic2010].
+Groupings in ordination space match these trends, with 30 m samples lying closer to shallow O&#x02BB;ahu samples and 60--90 m samples forming a distinct cluster.
 This pattern of community transition from shallow to mesophotic reefs is consistent with what has been shown for various other taxonomic groups [@lesserGlobalCommunityBreaks2019; @rochaMesophoticCoralEcosystems2018; @johnsonDepthPartitioningMesophotic2025] and subsets of entire communities [@hobanPlumbingDepthsEnvironmental2023], supporting the possibility that there may be broad-scale depth-based assembly rules for coral reef communities.
@@
 However, light and temperature have been proposed as important bottom-up drivers of depth zonation in benthic communities [@kahngCommunityEcologyMesophotic2010; @lesserGlobalCommunityBreaks2019; @pyleComprehensiveInvestigationMesophotic2016; @pyleMesophoticCoralEcosystems2019].
-Oceanographic conditions may be inconsistent close to shore, but the average mixed-layer depth calculated (using 0.8 &deg;C $\upDelta$T [@karaOptimalDefinitionOcean2000]) from CTD casts taken at station ALOHA (22.75&deg;N, 158.00&deg;W; within the latitudinal expanse of the MHI) was 84.6 $\pm$ 26 m over the course of the year (based on monthly averages from casts taken across the duration of ARMS unit deployments [@HawaiiOceanTimeseries]).
+Oceanographic conditions may be inconsistent close to shore, but the average mixed-layer depth (calculated using the 0.8 &deg;C $\upDelta$T criterion [@karaOptimalDefinitionOcean2000]) based on CTD casts taken at station ALOHA (22.75&deg;N, 158.00&deg;W; within the latitudinal expanse of the MHI) was 84.6 $\pm$ 26 m over the course of the year (based on monthly averages from casts taken across the duration of ARMS unit deployments [@HawaiiOceanTimeseries]).
 Thus, given the assumption of similar temperature profiles near the O&#x02BB;ahu shoreline, temperature alone appears insufficient to explain the zonation patterns we observed---although due to its significant role as a driver of shallow community variation, it cannot be ruled out.
@@
 A reduced version of the CAP ordination filtered to include only ARMS units deployed during these overlapping time periods retains the same (if not more pronounced) overall patterns as the full version (Supplementary [@fig:fig_supp_cap_reduced]), suggesting that the inclusion of prior or subsequent soaking periods did not adversely skew our analyses.
-Additionally, soak time differed by one year between O&#x02BB;ahu mesophotic ARMS units and those deployed across the rest of the archipelago.
-This difference may potentially bias results, but since CAP ordinations show differences despite the extra year that shallow units spent in the water, we feel that the patterns we observed reflect a true separation.
+Soak time differed by one year between O&#x02BB;ahu mesophotic ARMS units and those deployed across the rest of the archipelago.
+This difference may potentially bias results, but due to the distinct clustering in the CAP ordination despite the extra year that shallow units spent in the water, we feel that the patterns we observed reflect a true separation.
 The limitation of our depth transect to O&#x02BB;ahu means that we cannot make broad-scale conclusions about biodiversity patterns on deep reefs across the Hawaiian Islands.
-However, it remains the case that differences among depth zones at that single island appear to overwhelm differences observed among shallow-water coral reef habitat across the archipelago and depth appears to be the strongest organizing factor in canonical space regardless of where or when the communities were sampled.
+However, it remains the case that differences among depth zones at that single island outpace differences among shallow-water coral reef habitat across the archipelago and depth appears to be the strongest organizing factor in canonical space regardless of where or when the communities were sampled.
 Similar patterns were found at a local scale (tens of kilometers) on Hawai&#x02BB;i Island using environmental DNA (eDNA) [@hobanPlumbingDepthsEnvironmental2023].
 In that study, communities were found to have more affinity within depth zones across sites than among depth zones within sites.
-Thus, despite the limitations in our dataset, the patterns we observed suggest that factors associated with the shallow-to-mesophotic depth gradient exert stronger ecological influence than those (including depth) that vary across the latitudinal, physical, and anthropogenic gradients throughout the Hawaiian Archipelago.
+Despite the limitations in our dataset, the patterns we observed suggest that factors associated with the shallow-to-mesophotic depth gradient exert stronger ecological influence than those that vary across the latitudinal, physical, and anthropogenic gradients throughout the Hawaiian Archipelago.
 
@@
 Our data was collected using ARMS, which in mimicking the reef matrix are biased toward recruiting a specific subset of fauna that prefer cryptic habitats.
-However, the depth zonation patterns that we observed match those observed for multiple other taxa and communities more broadly [@diazLightTemperatureDrive2023; @diazMesophoticCoralBleaching2023; @hobanPlumbingDepthsEnvironmental2023; @hurleyAssessmentShallowMesophotic2016; @kaneHighLevelsMesophotic2014; @pinheiroDeepReefFishes2019; @rochaMesophoticCoralEcosystems2018; @montgomeryCommunitySimilaritySpecies2021].
+However, the depth zonation patterns that we observed match those observed for other taxa and communities [@diazLightTemperatureDrive2023; @diazMesophoticCoralBleaching2023; @hobanPlumbingDepthsEnvironmental2023; @hurleyAssessmentShallowMesophotic2016; @kaneHighLevelsMesophotic2014; @pinheiroDeepReefFishes2019; @rochaMesophoticCoralEcosystems2018; @montgomeryCommunitySimilaritySpecies2021].
 In light of this, and since mesophotic communities are generally less biodiverse than shallow reefs [@kahngCommunityEcologyMesophotic2010; @lesserEcologyMesophoticCoral2009; @pyleComprehensiveInvestigationMesophotic2016], it may be possible that the gradients driving depth distributions on coral reefs run opposite to those in the intertidal.
-Given the rapid changes in important factors like temperature, nutrients, and light throughout the upper ~100--150 m of the water column (particularly the diminishment of photosynthetically active radiation below 100 m), the *lower* limits of the coral reef mesophotic zone may be determined by abiotic factors that correlate with depth.
-Conversely, the *upper* limits of the mesophotic zone (the transition from shallow to mesophotic habitat), which occur at the higher end of the depth-biodiversity gradient [@kahngCommunityEcologyMesophotic2010; @lesserEcologyMesophoticCoral2009] may be driven by ecological interactions [@vellendEffectsDiversityDiversity2008].
+Given the rapid changes in important factors like temperature, nutrients, and light throughout the upper ~100--150 m of the water column (particularly the diminishment of photosynthetically active radiation below 100 m), the *lower* limits of the coral reef mesophotic zone (the transition from reef-associated to slope habitat) may be determined by abiotic factors that correlate with depth.
+Conversely, the *upper* limits of the mesophotic zone (the transition from shallow reef to mesophotic habitat), which occur at the higher end of the depth-biodiversity gradient [@kahngCommunityEcologyMesophotic2010; @lesserEcologyMesophoticCoral2009] may be driven by ecological interactions [@vellendEffectsDiversityDiversity2008].
 
@@
   scale_shape_manual(values=c("oahu" = 23, "noahu" = 21),name="Island", labels=c("noahu"="Other islands","oahu"="O‘ahu")) +
-  scale_fill_manual(values=sample_depth_pal,name="Sample grouping") +
+  scale_fill_manual(values=sample_depth_pal,name="Depth/region") +
   guides(
-    fill=guide_legend(override.aes=list(shape=21)),
-    shape=guide_legend(override.aes=list(fill="black"))
+    fill=guide_legend(override.aes=list(shape=21),order=2),
+    shape=guide_legend(override.aes=list(fill="black"),order=1)
   ) +

### 2025-06-30 collapse_taxonomy.R
change sp. to spp.

@@ -6,20 +6,20 @@ taxonomy_collapsed <- taxonomy %>%
       # fix two misspellings
       species == "Charybdis (Charybdis) hawaiensis" ~ "Charybdis Charybdis hawaiensis",
       species == "Catoptrus inequalis" ~ "Catoptrus inaequalis",
-      # make morphospecies into just "genus/family sp."
-      str_detect(species,"sp[0-9]+") ~ str_c(coalesce(genus,family)," sp."),
+      # make morphospecies into just "genus/family spp."
+      str_detect(species,"sp[0-9]+") ~ str_c(coalesce(genus,family)," spp."),
       .default = species
     ),
     # reset known taxon level
     taxon_level = case_when(
-      species == str_glue("{genus} sp.") ~ "genus",
-      species == str_glue("{family} sp.") ~ "family",
+      species == str_glue("{genus} spp.") ~ "genus",
+      species == str_glue("{family} spp.") ~ "family",
       .default = taxon_level
     )
   ) %>%
   # assign them new otu IDs
   group_by(species) %>%
-  mutate( old_otu = otu,  otu = str_replace_all(str_replace(species,fixed(" sp."),"")," ","_")) %>%
+  mutate( old_otu = otu,  otu = str_replace_all(str_replace(species,fixed(" spp."),"")," ","_")) %>%
   ungroup() %>%
   select(-ends_with("otu"),ends_with("otu")) %>%
   arrange(family,genus,species)

### 2025-06-30 manuscript.Rmd

@@
     keep_md: true
+link-citations: true
+linkReferences: true
 tblPrefix: Table
@@
 multiref_no_duplicates: true
+fig_caption: true
 ---
@@
 ```{r setup, include=FALSE}
+knitr::opts_knit$set( eval.after = "fig.cap" )
 knitr::opts_chunk$set(
@@
   "Shallow (MHI)" = "#ffb400",
+  "Shallow (O‘ahu)" = "#ffb400",
   "shallow_main" = "#ffb400",
@@
       taxon_level == "species" ~ str_glue("*{species}*"),
-      taxon_level == "genus" ~ str_glue("*{genus}* sp."),
-      taxon_level == "family" ~ str_glue("{family} sp."),
+      taxon_level == "genus" ~ str_glue("*{genus}* spp."),
+      taxon_level == "family" ~ str_glue("{family} spp."),
       .default = species
@@
 
-Investigations into factors that shape coral reef communities have been historically limited in sampling depth and taxonomic scope.
-Despite high biodiversity and the recognized importance of mesophotic habitats, studies have focused largely on conspicuous taxa such as fishes and corals and shallow habitat &lt;30 m.
+Despite high biodiversity and the recognized importance of mesophotic habitats, most studies of coral reef community structure have focused on conspicuous taxa such as fishes and corals in shallow habitat <30 m.
 Here, we examined the variability of crab assemblages from Autonomous Reef Monitoring Structures deployed on shallow reefs across the Hawaiian Islands and a mesophotic depth gradient on O&#x02BB;ahu.
-We tested the effects of environmental, ecological, and anthropogenic factors on shallow (8--17 m) crab assemblages.
-We additionally compared shallow assemblages to those sampled along a depth gradient (12--90 m) on O&#x02BB;ahu.
-Shallow assemblages were significantly different between the inhabited Main and the relatively pristine uninhabited Northwestern Hawaiian Islands and were associated with sea-surface temperature, chlorophyll-A, depth, island slope, potential larval immigration, and human impacts.
-Despite this variability, differences along the depth gradient on O&#x02BB;ahu alone were greater than among the shallow assemblages across the entire archipelago.
-This finding suggests that 90 m of depth is a stronger determinant of brachyuran assemblage structure than the latitudinal, environmental, or anthropogenic gradients across the entire ~2,500 km span of the Hawaiian Islands.
+We tested the effects of environmental, ecological, and anthropogenic factors on shallow (`r dr(cc$crabs_shallow_untransformed)` m) crab assemblages.
+These assemblages were significantly different between the densely inhabited high islands (Main Hawaiian Islands) and the relatively pristine uninhabited atolls of the Northwestern Hawaiian Islands.
+Drivers of these differences include sea-surface temperature, chlorophyll-A, depth, island slope, potential larval immigration, and human impacts.
+We then compared shallow assemblages to those sampled along a depth gradient (`r dr(cc$crabs_oahu)` m) on O&#x02BB;ahu.
+Despite considerable variability among sites distributed across nearly 10 degrees of latitude, differences along the depth gradient on O&#x02BB;ahu alone were greater than among the shallow assemblages across the entire archipelago.
+This finding suggests that 90 m of depth is a stronger driver of brachyuran assemblage structure than the latitudinal, environmental, and anthropogenic gradients across the entire ~2,500 km span of the Hawaiian Islands.
 
@@
 Coral reefs are among the most diverse habitats in the world, and they face increasing threats of degradation, including climate-driven bleaching and phase shifts to alternate ecological regimes [@grahamManagingResilienceReverse2013; @norstromAlternativeStatesCoral2009; @pandolfiGlobalTrajectoriesLongterm2003].
-In the face of these threats, understanding drivers of metazoan community structure in coral reef ecosystems is critically important [@hughesRisingChallengeSustaining2010; @pageSeekingResistanceCoral2019], particularly as the search for criteria to prioritize climate refugia such as upwelling zones or mesophotic reefs intensifies.
+In the face of these threats, understanding drivers of metazoan community structure in coral reef ecosystems is critically important [@hughesRisingChallengeSustaining2010; @pageSeekingResistanceCoral2019], particularly as the search for criteria to prioritize climate refugia such as upwelling zones or deep water reefs intensifies.
 While the structure and composition of coral reef communities is influenced by a suite of environmental and anthropogenic factors, the nature and makeup of those factors remain poorly resolved [@hughesCommunityStructureDiversity1989; @smithEffectsTopBottom2010; @brandlCoralReefEcosystem2019].
@@
 Many studies to date have focused on the upper ~30 m of the water column, yet coral reefs and associated communities are known to extend beyond 150 m [@pyleExploringDeepCoral1996; @pyleMesophoticCoralEcosystems2019], with photosymbiotic corals reported to 170 m [@rouzeSymbioticAssociationsDeepest2021].
-As much as two thirds of coral reef ecosystems lie in this under-explored zone [@hindersteinThemeSectionMesophotic2010].
+As much as two thirds of coral reef ecosystems lie below typical scuba depths in this under-explored zone [@hindersteinThemeSectionMesophotic2010].
 These habitats, known as mesophotic coral ecosystems (MCEs), are typically defined as extending from ~30--150 m depth [@hindersteinThemeSectionMesophotic2010; @kahngCommunityEcologyMesophotic2010].
@@
 One previous study [@hurleyAssessmentShallowMesophotic2016] found a significant depth gradient on the island of O&#x02BB;ahu for brachyuran crabs.
-However, the most frequently-targeted taxa make up a minority of coral reef biodiversity and thus may not be indicative of community patterns overall [@fautinOverviewMarineBiodiversity2010].
-Furthermore, some studies examining environmental effects on reef assemblages have been criticized for confounding multiple potentially covarying factors (such as habitat area and/or complexity) across the measured gradients [e.g., @bellwoodRegionalscaleAssemblyRules2001; @grahamEffectsMarineReserve2003; @connollyCommunityStructureCorals2005; @ewersConfoundingFactorsDetection2006; @fletcherRoleHabitatArea2007].
-
+However, the most conspicuous and frequently-studied taxa make up a minority of coral reef biodiversity and thus may not be indicative of community patterns overall [@fautinOverviewMarineBiodiversity2010].
+Furthermore, some studies examining environmental effects on reef assemblages have been criticized for confounding multiple potentially covarying factors (such as habitat area and/or complexity) across the measured gradients [@bellwoodRegionalscaleAssemblyRules2001; @grahamEffectsMarineReserve2003; @connollyCommunityStructureCorals2005; @ewersConfoundingFactorsDetection2006; @fletcherRoleHabitatArea2007].
 Here, we used autonomous reef monitoring structures (ARMS)---standardized collection devices designed to mimic reef structural complexity---to sample cryptic brachyuran crab communities while controlling for habitat [@zimmermanArtificialReefMatrix2004; @brainardAutonomousReefMonitoring2009].
@@
 
-## Diversity summary {-}
+## Brachyuran diversity summary {-}
 
@@
 
-```{r fig-dbrda, fig.cap="dbRDA plot showing significant drivers of crab community variation across the Hawaiian Islands.
-Colored ellipses represent 95% confidence intervals around island groups (Main vs.
-Northwestern Hawaiian Islands).", results="asis", fig.width=9, fig.height=6, dependson="dbrda-analysis"}
+<!-- ```{r fig-dbrda, fig.cap="dbRDA plot showing significant drivers of crab community variation among shallow-water (`r dr(cc$crabs_shallow_untransformed)` m) sites  across the Hawaiian Archipelago.
+Colored ellipses represent 95% confidence intervals around island groups (Main Hawaiian Islands vs.
+Northwestern Hawaiian Islands).", results="asis", fig.width=9, fig.height=6, dependson="dbrda-analysis"} -->
+```{r fig-dbrda, fig.cap=str_glue("dbRDA plot showing significant drivers of crab community variation among shallow-water ({dr(cc$crabs_shallow_untransformed)} m) sites  across the Hawaiian Archipelago.
+Colored ellipses represent 95% confidence intervals around island groups (Main Hawaiian Islands vs.
+Northwestern Hawaiian Islands)."), results="asis", fig.width=9, fig.height=6, dependson="dbrda-analysis"}
 termz <- model_terms %>%
@@
 
-```{r fig-cap, fig.cap="Canonical analysis of principal coordinates (CAP) ordination of crab assemblages by island group and depth zone.", results="asis", fig.width=9, fig.height=6, dependson="cap-analysis"}
+```{r fig-cap, fig.cap="Canonical analysis of principal coordinates (CAP) ordination of crab assemblages by island group and depth zone.
+Shallow samples are the same as those plotted with environmental drivers in @fig:fig_dbrda, to which the mesophotic samples (30 m, 60 m, and 90 m) are added here for comparison.
+Depth transect samples collected on the island of O&#x02BB;ahu are plotted with a diamond (and separately as Supplementary @fig:fig_supp_cap_oahu) for direct comparison relative to archipelago-wide shallow samples represented by a circle.", results="asis", fig.width=9, fig.height=6, dependson="cap-analysis"}
 centroids <- pcap %>% group_by(sample_group) %>%
@@
 
+# add a "Shallow (O‘ahu)" level to the sample_group column
+
+newgrp <- "Shallow (O‘ahu)"
+oldgrp <- 'Shallow (MHI)'
+ll <- levels(pcap$sample_group)
+after <- which(ll == oldgrp)
+
+# give the new group the same color as the other one
+# sample_depth_pal[newgrp] <- sample_depth_pal[oldgrp]
+
+# add the new factor level and make sure levels are appropriately ordered
+# also, add a shape column that we'll use below
+pcap <- pcap %>%
+  mutate(
+    sample_group = case_when(
+      island == "Oahu" & depth_zone == "shallow" ~ newgrp,
+      .default = sample_group
+    ),
+    sample_group = fct_reorder(sample_group,depth),
+    shape = case_when(
+      island == "Oahu" ~ 23,
+      .default = 21
+    )
+  )
+
+# get the new list of factor levels
+depth_groups <- levels(pcap$sample_group)
+
+# get a vector of shapes in the correct order
+shapes <- pcap %>%
+  distinct(sample_group,shape) %>%
+  arrange(match(sample_group,depth_groups)) %>%
+  deframe
+
+# plot the cap
 ggplot(pcap,aes(x=LD1,y=LD2,fill=sample_group)) +
@@
   # scale_shape_manual(values=c("oahu" = 23, "ffs" = 22, "noahu" = 21),name="Island", labels=c("noahu"="Other islands","oahu"="O‘ahu","ffs"="Lalo (French Frigate Shoals)" )) +
-  scale_shape_manual(values=c("oahu" = 23, "noahu" = 21),name="Island", labels=c("noahu"="Other islands","oahu"="O‘ahu")) +
+  scale_shape_manual(values=c("oahu" = 23, "noahu" = 21),name="Island", labels=c("noahu"="Other islands","oahu"="O‘ahu"),guide="none") +
   scale_fill_manual(values=sample_depth_pal,name="Depth/region") +
   guides(
-    fill=guide_legend(override.aes=list(shape=21)),
-    shape=guide_legend(override.aes=list(fill="black"))
+    fill=guide_legend(override.aes=list(shape=shapes))
   ) +
@@
 
-`r cmt("Three of the factors we found to be significantly associated with shallow crab assemblage variation correspond to well-known gradients across the islands: sea-surface temperature, chlorophyll-A, and human impacts","Is there a citation for these gradients existing across the islands?
-I feel like there must be but I don't know it off hand.
-Or is it even necessary because they're cited below?")`.
-While `r cmt("few studies","is this true?
-could probably use more citations here")` have examined the role of sea-surface temperature (SST) in structuring reef communities across the span of the Hawaiian Islands, `r cmt("it was found","probably some more examples here, too")` to be an important factor influencing intertidal communities within the MHI [@wardSpatialVariationAntecedent2023].
+Three of the factors we found to be significantly associated with shallow crab assemblage variation correspond to well-known gradients across the islands: sea-surface temperature, chlorophyll-A, and human impacts.
+While few studies have examined the role of sea-surface temperature (SST) in structuring reef communities across the span of the Hawaiian Islands, it was found to be an important factor influencing intertidal communities within the MHI [@wardSpatialVariationAntecedent2023].
 Chlorophyll-A can be used as a proxy for primary productivity.
@@
 
-Further work to investigate drivers of depth zonation could take advantage of established gradients and use the Hawaiian Archipelago as a natural laboratory, by sampling depth transects across the archipelago.
+Further work to investigate drivers of depth zonation could take advantage of established gradients across the Hawaiian Archipelago by sampling depth transects across the archipelago.
 Coral reef biodiversity is generally thought to be structured at depth by a combination of the attenuation of light, a decrease in temperature, increasing salinity, and a reduction in trophic resources [@lesserEcologyMesophoticCoral2009; @kahngCommunityEcologyMesophotic2010; @slatteryFunctionStabilityMesophotic2024a].
@@
 # Author contributions {-}
-Conceptualization: MLH, KKH, DJS, RJT; Data curation: MLH; Formal Analysis: MLH, KKH; Funding acquisition: KKH, DJS, RJT; Investigation: MLH, KKH, KR, DJS; Methodology: MLH, KKH, DJS, MAT; Project administration: KKH, DJS, RJT; Resources: KR, MAT, RJT; Software: MLH; Supervision: DJS, RJT; Validation: MLH, MAT, RJT; Visualization: MLH, KR; Writing -- original draft: MLH, KKH, RJT; Writing -- review & editing: MLH, KKH, KR, DJS, MAT, RJT
+Conceptualization: MLH, KKH, DJS, RJT; Data curation: MLH; Formal Analysis: MLH, KKH; Funding acquisition: KKH, DJS, RJT; Investigation: MLH, KKH, KR, DJS, RJT; Methodology: MLH, KKH, DJS, MAT, RJT; Project administration: KKH, DJS, RJT; Resources: KR, MAT, RJT; Software: MLH; Supervision: DJS, RJT; Validation: MLH, MAT, RJT; Visualization: MLH, KR; Writing -- original draft: MLH, KKH, RJT; Writing -- review & editing: MLH, KKH, KR, DJS, MAT, RJT
 
@@
 nice <- function(s) {
-  str_replace(s,"shallow","Shallow") %>%
+  str_replace(s,"shallow","Shallow (O‘ahu)") %>%
     str_replace("main","(MHI)") %>%
@@
   stat_ellipse(geom="polygon",level=0.95,alpha=0.1,show.legend = FALSE) +
-  geom_point(mapping=aes(shape=oahu),color="black",size=5)  +
+  geom_point(mapping=aes(shape=oahu),color="black",size=5,shape=23)  +
   geom_label_repel(data=centroids_reduced,aes(label=depth_zone), color="white", show.legend = FALSE) +
   # scale_shape_manual(values=c("oahu" = 23, "ffs" = 22, "noahu" = 21),name="Island", labels=c("noahu"="Other islands","oahu"="O‘ahu","ffs"="Lalo (French Frigate Shoals)" )) +
-  scale_shape_manual(values=c("oahu" = 23, "noahu" = 21),name="Island", labels=c("noahu"="Other islands","oahu"="O‘ahu")) +
+  # scale_shape_manual(values=c("oahu" = 23, "noahu" = 21),name="Island", labels=c("noahu"="Other islands","oahu"="O‘ahu")) +
   scale_fill_manual(values=sample_depth_pal,name="Depth/region") +
   guides(
-    fill=guide_legend(override.aes=list(shape=21),order=2),
+    fill=guide_legend(override.aes=list(shape=23),order=2),
     shape=guide_legend(override.aes=list(fill="black"),order=1)
@@
 
+newgrp <- "Shallow (O‘ahu)"
+oldgrp <- 'Shallow (MHI)'
+ll <- levels(pcap_reduced$sample_group)
+after <- which(ll == oldgrp)
+
+# give the new group the same color as the other one
+# sample_depth_pal[newgrp] <- sample_depth_pal[oldgrp]
+
+# add the new factor level and make sure levels are appropriately ordered
+# also, add a shape column that we'll use below
+pcap_reduced <- pcap_reduced %>%
+  mutate(
+    sample_group = case_when(
+      island == "Oahu" & depth_zone == "shallow" ~ newgrp,
+      .default = sample_group
+    ),
+    dz=str_replace(depth_zone,"^([0-9]+)","z\\1"),
+    dz=str_c(dz,region,depth)
+  ) %>%
+  arrange(dz) %>%
+  mutate(
+    # sample_group = fct_reorder(sample_group,as.character(sample_group)),
+    sample_group = fct_inorder(sample_group),
+    shape = case_when(
+      island == "Oahu" ~ 23,
+      .default = 21
+    )
+  )
+
+# get the new list of factor levels
+depth_groups <- levels(pcap_reduced$sample_group)
+
+# get a vector of shapes in the correct order
+shapes <- pcap_reduced %>%
+  distinct(sample_group,shape) %>%
+  arrange(match(sample_group,depth_groups)) %>%
+  deframe
+
 # plot the thing
@@
   # scale_shape_manual(values=c("oahu" = 23, "ffs" = 22, "noahu" = 21),name="Island", labels=c("noahu"="Other islands","oahu"="O‘ahu","ffs"="Lalo (French Frigate Shoals)" )) +
-  scale_shape_manual(values=c("oahu" = 23, "noahu" = 21),name="Island", labels=c("noahu"="Other islands","oahu"="O‘ahu")) +
+  scale_shape_manual(values=c("oahu" = 23, "noahu" = 21),name="Island", labels=c("noahu"="Other islands","oahu"="O‘ahu"),guide="none") +
   scale_fill_manual(values=sample_depth_pal,name="Depth/region") +
   guides(
-    fill=guide_legend(override.aes=list(shape=21),order=2),
-    shape=guide_legend(override.aes=list(fill="black"),order=1)
+    fill=guide_legend(override.aes=list(shape=shapes))
+    # shape=guide_legend(override.aes=list(fill="black"),order=1)
   ) +

### 2025-07-03 manuscript.Rmd
fix cap figure confidence ellipses

@@
 Depth transect samples collected on the island of O&#x02BB;ahu are plotted with a diamond (and separately as Supplementary @fig:fig_supp_cap_oahu) for direct comparison relative to archipelago-wide shallow samples represented by a circle.", results="asis", fig.width=9, fig.height=6, dependson="cap-analysis"}
-centroids <- pcap %>% group_by(sample_group) %>%
-  summarise(LD1=mean(LD1),LD2=mean(LD2))
-pcap <- pcap %>%
+pcap2 <- pcap %>%
   mutate(
@@
 
+centroids <- pcap2 %>% group_by(sample_group) %>%
+  summarise(LD1=mean(LD1),LD2=mean(LD2))
+
 # add a "Shallow (O‘ahu)" level to the sample_group column
@@
 oldgrp <- 'Shallow (MHI)'
-ll <- levels(pcap$sample_group)
+ll <- levels(pcap2$sample_group)
 after <- which(ll == oldgrp)
@@
 # also, add a shape column that we'll use below
-pcap <- pcap %>%
+pcap2 <- pcap2 %>%
   mutate(
+    old_sample_group = sample_group,
     sample_group = case_when(
@@
     ),
-    sample_group = fct_reorder(sample_group,depth),
+    dz=str_replace(depth_zone,"^([0-9]+)","z\\1"),
+    dz=str_c(dz,region,depth)
+  ) %>%
+  arrange(dz) %>%
+  mutate(
+    sample_group = fct_inorder(sample_group),
     shape = case_when(
@@
 # get the new list of factor levels
-depth_groups <- levels(pcap$sample_group)
+depth_groups <- levels(pcap2$sample_group)
 
 # get a vector of shapes in the correct order
-shapes <- pcap %>%
+shapes <- pcap2 %>%
   distinct(sample_group,shape) %>%
   arrange(match(sample_group,depth_groups)) %>%
-  deframe
+  deframe()
+
+centroids <- pcap2 %>% group_by(old_sample_group) %>%
+  summarise(LD1=mean(LD1),LD2=mean(LD2)) %>%
+  mutate(old_sample_group = as.character(old_sample_group)) %>%
+  rename(sample_group=old_sample_group)
 
 # plot the cap
-ggplot(pcap,aes(x=LD1,y=LD2,fill=sample_group)) +
-  stat_ellipse(geom="polygon",level=0.95,alpha=0.1,show.legend = FALSE) +
-  # geom_point(aes(shape=oahu),color="black",shape=21,size=5)  +
+ggplot(pcap2,aes(x=LD1,y=LD2,fill=sample_group)) +
   geom_point(mapping=aes(shape=oahu),color="black",size=5)  +
-  geom_label_repel(data=centroids,aes(label=sample_group), color="white", show.legend = FALSE) +
-  # scale_shape_manual(values=c("oahu" = 23, "ffs" = 22, "noahu" = 21),name="Island", labels=c("noahu"="Other islands","oahu"="O‘ahu","ffs"="Lalo (French Frigate Shoals)" )) +
+  stat_ellipse(mapping=aes(x=LD1,y=LD2,fill=old_sample_group),inherit.aes = FALSE,geom="polygon",level=0.95,alpha=0.1,show.legend = FALSE) +
   scale_shape_manual(values=c("oahu" = 23, "noahu" = 21),name="Island", labels=c("noahu"="Other islands","oahu"="O‘ahu"),guide="none") +
-  scale_fill_manual(values=sample_depth_pal,name="Depth/region") +
+  scale_fill_manual(values=sample_depth_pal,name="Depth/region",guide="none") + 
+  # geom_label_repel(data=centroids,aes(x=LD1,y=LD2,fill=sample_group,label=sample_group),inherit.aes = FALSE, color="white", show.legend = FALSE) +
+  geom_label_repel(data=centroids,mapping=aes(x=LD1,LD2,label=sample_group,fill=sample_group),color="white",inherit.aes = FALSE,show.legend = FALSE) + 
   guides(
